---
output: github_document
---

<!-- README.md is generated from README.Rmd using [duckknit](https://github.com/rundel/duckknit). Please edit that file. -->

```{r, include = FALSE}
library(duckknit)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# Rewrite duckknit's ``` duckdb fences to ``` sql so GitHub applies
# SQL syntax highlighting in the rendered README.
knitr::knit_hooks$set(document = function(x) {
  gsub("(?m)^``` duckdb", "``` sql", x, perl = TRUE)
})

# Use duckknit's duckdb engine for sql chunks.
knitr::knit_engines$set(sql = knitr::knit_engines$get("duckdb"))

# Pre-start the default duckknit session with -unsigned so unsigned extensions
# can be loaded. duckknit_start_session() only accepts a db path, not extra
# CLI flags, so we launch the process directly and register it in duckknit's
# internal session table.
options(duckknit.duckdb = "/usr/local/bin/duckdb152")
if (!requireNamespace("Rduckhts", quietly = TRUE)) {
  stop("Install the bootstrapped Rduckhts package before rendering README.md")
}
Rduckhts::setup_hts_env()
extension_path <- system.file(
  "duckhts_extension", "build", "duckhts.duckdb_extension",
  package = "Rduckhts", mustWork = TRUE
)
local({
  ns  <- asNamespace("duckknit")
  cmd <- duckknit:::find_duckdb()
  proc <- processx::process$new(
    command = cmd,
    args = c("-unsigned", ":memory:"),
    stdin = "|", stdout = "|", stderr = "|",
    cleanup = TRUE, cleanup_tree = TRUE
  )
  unlockBinding(".sessions", ns)
  ns$.sessions[["readme"]] <- proc
  lockBinding(".sessions", ns)
  unlockBinding(".state", ns)
  ns$.state$last_session <- "readme"
  lockBinding(".state", ns)

  load_sql <- sprintf(
    "SET allow_extensions_metadata_mismatch=true;\nLOAD '%s';",
    gsub("'", "''", extension_path, fixed = TRUE)
  )
  load_result <- duckknit:::duckknit_exec(proc, load_sql)
  if (nzchar(trimws(load_result$stderr))) {
    stop(load_result$stderr, call. = FALSE)
  }
})
```

# DuckHTS

[![CRAN Status](https://www.r-pkg.org/badges/version/Rduckhts)](https://cran.r-project.org/package=Rduckhts)
[![R-universe version](https://RGenomicsETL.r-universe.dev/Rduckhts/badges/version)](https://RGenomicsETL.r-universe.dev/Rduckhts)

A [DuckDB](https://duckdb.org/) extension for reading high-throughput sequencing (HTS) file formats and running interval-oriented genomics workflows directly in SQL.

DuckHTS brings sequencing alignments, variants, reference sequences, annotations, and interval text files into DuckDB through [htslib](https://github.com/samtools/htslib). Query VCF, BCF, BAM, CRAM, FASTA, FASTQ, GTF, GFF, BED, and tabix-indexed files with standard SQL, then compose metadata inspection, compression/indexing, sequence utilities, and interval/reference helpers in the same session.

It covers the sequencing side of the stack: raw/reference data access, annotation parsing, BGZF/tabix utilities, and building blocks for coverage, CNV, QC, and downstream export workflows.

The project thesis and ownership boundaries are documented in [`ARCHITECTURE.md`](ARCHITECTURE.md). Current implementation contracts are indexed in [`design/README.md`](design/README.md); the DuckVEP transcript-model build and consequence-engine dataflow are described in [`design/duckvep.md`](design/duckvep.md).

> **Part of a growing genomics stack in DuckDB.**
> DuckHTS covers sequencing and reference-oriented formats. [PlinkingDuck](https://github.com/teaguesterling/plinking_duck) extends the same idea to PLINK genotype workflows and explicitly builds on DuckHTS for the sequencing side of the stack. See also its documentation: https://plinking-duck.readthedocs.io . Together, these tools are pushing DuckDB toward a broader genomics analysis environment built around portable file readers, SQL-native analytics, and composable local workflows. After all, most of Genomics File Formats are Tables/Arrays in disguise.


This extension does not support MSVC builds (windows_amd64/windows_arm64). Use MinGW/RTools for Windows.

## Functions

<details>
<summary>Show generated function catalog</summary>

```{r function-catalog, echo=FALSE, results='asis'}
cat(paste(readLines("r/Rduckhts/inst/function_catalog/functions.md"), collapse = "\n"))
```

</details>

`read_fastq` with `mate_path` requires exact QNAME pairing. `read_bam` supports typed `standard_tags` and `auxiliary_tags` maps. `read_tabix` supports header-aware parsing (`header`, `header_names`) and optional type inference (`auto_detect`, `column_types`). Region lists in comma-separated form are supported by `read_bam`, `read_bcf`, `read_fasta`, `read_gff`, `read_gtf`, and `read_tabix`. Indexed `read_bam`, `read_bcf`, `read_gff`, `read_gtf`, and `read_tabix` multi-region queries use htslib iterators that emit a matching record once when requested regions overlap. `read_fasta` retains its separate per-region sequence-row contract.

## Examples

The examples below run directly against bundled local test files through the DuckDB CLI and load the built extension from `build/release/duckhts.duckdb_extension`.

### Core readers

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT CHROM, POS, REF, ALT, SAMPLE_ID
FROM read_bcf('test/data/formatcols.vcf.gz', tidy_format := true)
LIMIT 3;

SELECT count(*) AS n
FROM read_bam('test/data/range.bam', region := 'CHROMOSOME_I:1-1000');

SELECT *
FROM fasta_index('test/data/ce.fa');

SELECT NAME, length(SEQUENCE) AS seq_length
FROM read_fasta('test/data/ce.fa', region := 'CHROMOSOME_I:1-25');

SELECT NAME, MATE, PAIR_ID
FROM read_fastq('test/data/interleaved.fq', interleaved := true)
LIMIT 3;
```

### Variant normalization

`duckhts_bcftools_norm(...)` applies bcftools-style FASTA-backed allele normalization to a regular table or derived relation while preserving the original columns. In split mode, multiallelic rows are expanded first and then normalized one ALT at a time.

```{sql eval=TRUE, collapse=FALSE, comment=""}
CREATE OR REPLACE TEMP TABLE readme_norm AS
SELECT *
FROM (VALUES
  ('chrS', 2, 'T', 'TT,TTT'),
  ('chrS', 2, 'T', '*,TT')
) AS t(chrom, pos, ref, alt);

SELECT chrom, pos, ref, alt, alt_index,
       pos_normed, ref_normed, alt_normed, norm_status
FROM duckhts_bcftools_norm(
  'readme_norm',
  'test/data/liftover_repeat_src.fa',
  split_multiallelic := true
)
ORDER BY alt, alt_index;
```

### VariantKey + RegionKey

DuckHTS vendors the official VariantKey / RegionKey C API and exposes SQL helpers that mirror bcftools `%VKX`-style VariantKey output on VCF rows. `variantkey(...)` accepts 1-based VCF `POS`, while `regionkey(...)` uses 0-based half-open interval semantics. Large, ambiguous, and symbolic alleles still encode through the official hashed nonreversible VariantKey mode, but those keys do not encode `END`, `SVLEN`, mate breakend coordinates, or other SV metadata; use RegionKey explicitly for span-oriented interval work. See Nicola Asuni (2018) <https://doi.org/10.1101/473744>.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT variantkey_hex(variantkey('1', 324684, 'C', 'G')) AS vkx,
       reverse_variantkey(parse_variantkey_hex('08027a2588b00000')) AS reversed;

SELECT regionkey_hex(regionkey('X', 1007, 1807, 1)) AS rkx,
       are_overlapping_regionkeys(
         regionkey('X', 1007, 1807, 1),
         parse_regionkey_hex('b80001f78000387a')
       ) AS overlaps;
```

### Interval + reference helpers

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT chrom, start, "end", name, block_count
FROM read_bed('test/data/targets.bed');

SELECT chrom, start, "end", pct_gc, num_a, num_c, num_g, num_t
FROM fasta_nuc('test/data/ce.fa', bed_path := 'test/data/targets.bed')
ORDER BY chrom, start;

SELECT chrom, start, "end", seq_len, pct_gc
FROM fasta_nuc('test/data/ce.fa', bin_width := 10, region := 'CHROMOSOME_I:1-20');
```

### cgranges registry entry points

`duckhts_cgranges_*` exposes a session-scoped immutable interval index for
native overlap queries. You can either build it row-wise with
`duckhts_cgranges_create(...)` + `duckhts_cgranges_add(...)`, or bulk-load it
from SQL with `duckhts_cgranges_from_query(...)`. For row-preserving filters or
count annotations over provider rows, use the vectorized scalar helpers
`duckhts_cgranges_has_overlap(...)` and
`duckhts_cgranges_count_overlaps(...)` directly in queries over `read_bed(...)`,
`read_bam(...)`, `read_bcf(...)`, or regular tables. For streaming
one-row-per-hit expansion while keeping provider columns, use
`duckhts_cgranges_overlaps_list(...)` with `UNNEST(...)`. The older
`duckhts_cgranges_overlaps_bulk(...)` table function still accepts a probe query
and emits matching indexed intervals in one table-function call; that bulk query
runs on the extension-owned helper connection, so use a regular table or view
rather than a temp table.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT duckhts_cgranges_create('readme_idx');
SELECT duckhts_cgranges_add('readme_idx', 'chr1', 10, 20, 'a');
SELECT duckhts_cgranges_add('readme_idx', 'chr1', 30, 40, 'b');
SELECT duckhts_cgranges_index('readme_idx');

SELECT interval_ordinal, label, interval_chrom, interval_start, interval_end
FROM duckhts_cgranges_overlaps('readme_idx', 'chr1', 35, 36, query_row_id := 7);

SELECT duckhts_cgranges_from_query(
  'readme_qry_idx',
  'SELECT * FROM (VALUES (''chr2'', 100, 110, ''alpha''), (''chr2'', 150, 170, ''beta'')) AS t(chrom, start, "end", label)',
  'chrom', 'start', 'end', 'label'
);
SELECT duckhts_cgranges_index('readme_qry_idx');

SELECT interval_ordinal, label, interval_chrom, interval_start, interval_end
FROM duckhts_cgranges_overlaps('readme_qry_idx', 'chr2', 140, 170, mode := 'contain');

CREATE TABLE readme_probes AS
SELECT * FROM (VALUES
  (10, 'chr2', 100, 105),
  (20, 'chr2', 160, 161),
  (30, 'chr2', 500, 510)
) AS t(probe_id, chrom, start, "end");

SELECT
  p.probe_id,
  hit.interval_ordinal,
  hit.label,
  hit.label_type,
  hit.interval_chrom,
  hit.interval_start,
  hit.interval_end
FROM readme_probes AS p
CROSS JOIN UNNEST(
  duckhts_cgranges_overlaps_list('readme_qry_idx', p.chrom, p.start, p."end")
) AS u(hit)
ORDER BY p.probe_id, hit.interval_ordinal;

SELECT query_row_id, interval_ordinal, label, interval_chrom, interval_start, interval_end
FROM duckhts_cgranges_overlaps_bulk(
  'readme_qry_idx',
  'SELECT probe_id, chrom, start, "end" FROM readme_probes',
  'chrom', 'start', 'end',
  query_row_id_col := 'probe_id'
)
ORDER BY query_row_id, interval_ordinal;

SELECT duckhts_cgranges_destroy('readme_idx');
SELECT duckhts_cgranges_destroy('readme_qry_idx');
DROP TABLE readme_probes;
```

### Fixed-bin native counting

`bam_bin_counts()` does fixed-width read-start binning directly in native code.
This is the counting primitive used for WisecondorX-style workflows: duplicate
handling is explicit via `rmdup`, and optional `stats := 'gc,mq'` adds one-pass
GC and MAPQ summaries on the same scan.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT
  bin_id,
  count_total,
  count_fwd,
  count_rev,
  count_pre,
  printf('%.2f', gc_perc_pre) AS gc_pre,
  printf('%.2f', gc_perc_post) AS gc_post,
  printf('%.1f', mean_mapq_post) AS mean_mapq_post
FROM bam_bin_counts(
  'test/data/fixture_mixed.cram',
  5000,
  reference := 'test/data/fixture_ref.fa',
  rmdup := 'streaming',
  stats := 'gc,mq'
)
ORDER BY bin_id;
```

### Mosdepth-compatible coverage outputs

`duckhts_mosdepth()` writes mosdepth-style output files directly from indexed
BAM/CRAM input. The example below writes windowed fragment coverage and then
reads back the generated BED.gz output.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT success, summary_path, regions_path
FROM duckhts_mosdepth(
  '/tmp/duckhts_readme_mosdepth',
  'test/data/range.bam',
  chrom := 'CHROMOSOME_II',
  by := '1000',
  no_per_base := TRUE,
  fragment_mode := TRUE,
  use_median := TRUE,
  overwrite := TRUE
);

SELECT
  column0 AS chrom,
  CAST(column1 AS BIGINT) AS start,
  CAST(column2 AS BIGINT) AS "end",
  CAST(column3 AS DOUBLE) AS depth
FROM read_csv(
  '/tmp/duckhts_readme_mosdepth.regions.bed.gz',
  delim := '\t',
  header := FALSE,
  compression := 'gzip'
)
LIMIT 3;
```

### Polygenic risk scoring

`bcftools_score` computes per-sample polygenic risk scores (PRS) from a VCF/BCF and one or more GWAS summary statistics files, mirroring the `bcftools +score` plugin API.

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Hard-call (GT) PRS — PLINK summary format
-- S1: 0×0.5  + 1×(−0.2) + 2×1.0 = 1.8
-- S2: 1×0.5  + 2×(−0.2) + 0×1.0 = 0.1
SELECT SAMPLE, round(score_summary, 3) AS prs
FROM bcftools_score(
  'test/data/score_input.vcf',
  'test/data/score_summary.tsv',
  use := 'GT',
  columns := 'PLINK'
);

-- Multi-PRS TSV/SSF scoring: multiple summary files in one genotype scan
SELECT SAMPLE,
       round(score_summary, 3) AS prs_a,
       round(score_summary_na, 3) AS prs_b
FROM bcftools_score(
  'test/data/score_input.vcf',
  ['test/data/score_summary.tsv', 'test/data/score_summary_na.tsv'],
  use := 'GT',
  columns := 'PLINK'
);

-- Dosage-based PRS (DS field) — fractional allele dosages from imputed data
-- S1: 0.1×0.5 + 0.8×(−0.2) + 1.8×1.0 = 1.69
-- S2: 1.0×0.5 + 1.9×(−0.2) + 0.2×1.0 = 0.32
SELECT SAMPLE, round(score_summary, 3) AS prs_ds
FROM bcftools_score(
  'test/data/score_dosage.vcf',
  'test/data/score_summary.tsv',
  use := 'DS',
  columns := 'PLINK'
);

-- GWAS-VCF multi-PRS: each FORMAT sample column becomes a separate PRS track
SELECT SAMPLE, round(PRS_A, 3) AS prs_a, round(PRS_B, 3) AS prs_b
FROM bcftools_score(
  'test/data/score_input.vcf',
  'test/data/score_gwas_summary.vcf',
  use := 'GT'
);
```

### Liftover score-style rows

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT src_chrom, src_pos, dest_chrom, dest_pos, dest_ref, dest_alt,
       mapped, reverse_complemented, reject_reason, note
FROM duckdb_liftover(
  '(VALUES
     (''chrF'', 2, ''C'', ''T''),
     (''chrR'', 2, ''A'', ''G''),
     (''chrF'', 11, ''A'', ''T'')
   ) AS t(chrom, pos, ref, alt)',
  'chrom',
  'pos',
  ref_col := 'ref',
  alt_col := 'alt',
  chain_path := 'test/data/liftover.chain',
  dst_fasta_ref := 'test/data/liftover_dst.fa',
  src_fasta_ref := 'test/data/liftover_src.fa'
);
```

### SIMD dispatch flow

DuckHTS uses explicit runtime SIMD dispatch for byte-oriented helper kernels,
starting with `seq_gc_content(...)`. `scalar` is always available and is the
portable baseline. Optional platform backends such as `avx2` or `avx512`
should be checked with `duckhts_simd_backend_available(...)` before being requested.
The `auto` policy resolves each logical kernel independently from the current
compiled-and-CPU-supported capability mask; use `duckhts_simd_kernel_info()` for
the per-kernel result and `SELECT backend FROM duckhts_simd_set_backend('auto')`
to return to runtime auto-detection.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT backend, selectable, compiled, cpu_supported, available, selected
FROM duckhts_simd_info();

SELECT kernel, selected_backend, scalar_fallback
FROM duckhts_simd_kernel_info();

SELECT backend AS selected_backend FROM duckhts_simd_set_backend('scalar');

SELECT
  duckhts_simd_requested_backend() AS requested_backend,
  duckhts_simd_backend() AS selected_backend,
  printf('%.3f', seq_gc_content('ACGTNNacgtnn')) AS gc_content;

SELECT backend IS NOT NULL AS restored_auto FROM duckhts_simd_set_backend('auto');
```

### Sequence utilities

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT
  NAME,
  seq_hash_2bit(substr(SEQUENCE, 1, 12)) AS hash_2bit_prefix,
  seq_encode_4bit(substr(SEQUENCE, 1, 16)) AS codes,
  seq_decode_4bit(seq_encode_4bit(substr(SEQUENCE, 1, 16))) AS roundtrip
FROM read_fasta('test/data/ce.fa')
LIMIT 2;

SELECT
  NAME,
  MATE,
  seq_encode_4bit(substr(SEQUENCE, 1, 12)) AS codes,
  seq_decode_4bit(seq_encode_4bit(substr(SEQUENCE, 1, 12))) AS roundtrip
FROM read_fastq('test/data/interleaved.fq', interleaved := true)
LIMIT 2;
```

### FASTQ quality decoding and fused QC

`read_fastq()` separates input interpretation from output representation:

- `input_quality_encoding` tells DuckHTS how to decode FASTQ ASCII into numeric qualities. The default is modern `phred33`. Use `phred64`, `solexa64`, or `auto` only for legacy files.
- `quality_representation := 'phred'` returns canonical numeric qualities as `UTINYINT[]`.
- `quality_representation := 'string'` returns canonical Phred+33 text. For legacy inputs this means decode first, then re-encode as modern FASTQ text.

This makes the flow explicit:

1. FASTQ text input is decoded according to `input_quality_encoding`.
2. DuckHTS normalizes to numeric Phred values internally.
3. Output is either raw numeric quality arrays (`phred`) or canonical Phred+33 text (`string`).

For BAM/CRAM, qualities are already stored as numeric values, so there is no FASTQ text-encoding ambiguity on input.

Use `duckhts_fastq_qc(...)` for global and per-cycle quality-control reductions. It consumes
the projected sequence and canonical quality strings in one bounded aggregate instead of
creating one SQL row per base. Expand only the compact cycle result when plotting or joining
per-cycle statistics.

```{sql eval=TRUE, collapse=FALSE, comment=""}
WITH q AS (
  SELECT duckhts_fastq_qc(SEQUENCE, QUALITY) AS qc
  FROM read_fastq('test/data/r1.fq')
)
SELECT
  qc.reads,
  qc.bases,
  qc.q30_bases,
  qc.max_read_length
FROM q;

WITH q AS (
  SELECT duckhts_fastq_qc(SEQUENCE, QUALITY) AS qc
  FROM read_fastq('test/data/r1.fq')
)
SELECT cycle.cycle, cycle.bases, cycle.quality_sum
FROM q, UNNEST(qc.cycles) AS u(cycle)
ORDER BY cycle.cycle
LIMIT 5;
```

Use numeric quality arrays when the query genuinely needs the full quality histogram:

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT *
FROM detect_quality_encoding('test/data/legacy_phred64.fq');

WITH q AS (
  SELECT NAME, QUALITY
  FROM read_fastq(
    'test/data/r1.fq',
    quality_representation := 'phred'
  )
),
expanded AS (
  SELECT
    NAME,
    generate_subscripts(QUALITY, 1) AS pos,
    unnest(QUALITY) AS q
  FROM q
)
SELECT pos, q AS phred, count(*) AS n_reads
FROM expanded
GROUP BY pos, phred
ORDER BY pos, phred
LIMIT 12;
```

### Metadata + export/index helpers

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT idx, raw
FROM read_hts_header('test/data/formatcols.vcf.gz', mode := 'raw')
LIMIT 3;

SELECT seqname, tid, index_type, chunk_beg_vo, chunk_end_vo
FROM read_hts_index_spans('test/data/formatcols.vcf.gz')
LIMIT 3;

SELECT index_type, octet_length(raw) AS raw_bytes
FROM read_hts_index_raw('test/data/formatcols.vcf.gz');

COPY (
  SELECT chrom, start, "end", name
  FROM read_bed('test/data/targets.bed')
) TO '/tmp/duckhts_readme_targets.bed' (FORMAT CSV, DELIMITER '\t', HEADER FALSE);

SELECT success, output_path, bytes_out
FROM bgzip('/tmp/duckhts_readme_targets.bed',
           output_path := '/tmp/duckhts_readme_targets.bed.gz',
           keep := TRUE,
           overwrite := TRUE);

SELECT success, output_path, bytes_out
FROM bgunzip('/tmp/duckhts_readme_targets.bed.gz',
             output_path := '/tmp/duckhts_readme_targets.roundtrip.bed',
             keep := TRUE,
             overwrite := TRUE);

SELECT success, index_format, index_path
FROM bam_index('test/data/range.bam',
               index_path := '/tmp/duckhts_readme_range.bam.bai',
               threads := 1);

SELECT success, index_format, index_path
FROM bcf_index('test/data/vcf_file.bcf',
               index_path := '/tmp/duckhts_readme_vcf_file.bcf.csi',
               threads := 1);

SELECT success, index_format, index_path
FROM tabix_index('/tmp/duckhts_readme_targets.bed.gz',
                 preset := 'bed',
                 index_path := '/tmp/duckhts_readme_targets.bed.gz.tbi');
```

### Multi-file queries

`hts_union_query` builds a `UNION ALL BY NAME` across files matching a glob pattern.
Because DuckDB's `query()` cannot accept subquery expressions, use the
`SET VARIABLE` + `getvariable()` pattern:

```{sql eval=TRUE, collapse=FALSE, comment=""}
SET VARIABLE q = hts_union_query('read_fastq', 'test/data/r*.fq');
SELECT filename, count(*) AS n
FROM query(getvariable('q'))
GROUP BY ALL
ORDER BY filename;
```

Per-file parameters can be passed as the third argument (SQL literal):

```{sql eval=TRUE, collapse=FALSE, comment=""}
SET VARIABLE q = hts_union_query('read_bam', 'test/data/range.bam',
                                  'region := ''CHROMOSOME_I:1-1000''');
SELECT count(*) AS n FROM query(getvariable('q'));
```

## Remote URLs and HTS_PATH

Remote URLs (S3/GCS/HTTP/S) can work in two htslib build modes:

1. Dynamic plugin mode (`ENABLE_PLUGINS`): remote handlers are loaded from `HTS_PATH`.
2. Static-handler mode (plugins disabled): handlers are compiled into `libhts` and `HTS_PATH` is not needed.

Use `HTS_PATH` only when you want dynamic plugin discovery (for example, to point at an external htslib plugin directory). `Rduckhts` users only need to call the public helper before the first HTS file is opened:

```r
library(Rduckhts)
setup_hts_env()
```

Example (works in static-handler mode and plugin mode):

```{sql, eval=TRUE, collapse=FALSE, comment=""}
SELECT CHROM, COUNT(*) AS n
FROM read_bcf('s3://1000genomes-dragen-v3.7.6/data/cohorts/gvcf-genotyper-dragen-3.7.6/hg19/3202-samples-cohort/3202_samples_cohort_gg_chr22.vcf.gz',
              region := 'chr22:16050000-16050500')
GROUP BY CHROM;
```

For a direct DuckDB CLI process, set `HTS_PATH` explicitly before its first HTS read, for example:

```bash
export HTS_PATH=$(Rscript --quiet -e 'Rduckhts::setup_hts_env(); cat(Sys.getenv("HTS_PATH"),sep="")')
```

If htslib has already opened a file in the process, restart the process after changing `HTS_PATH`; plugin discovery has already occurred.

If you don’t have htslib plugins installed locally, download the prebuilt binaries from the r-universe-binaries GitHub release and point `HTS_PATH` at the extracted htslib/libexec/htslib directory inside the package bundle.
https://github.com/RGenomicsETL/duckhts/releases/tag/r-universe-binaries

### Browser wasm/webR HTTP backend

For browser wasm/webR builds, DuckHTS does **not** use htslib `libcurl` for remote `http`/`https` access.

- The webR side-module path disables htslib `libcurl`/`S3`/`GCS` features because socket-based libcurl calls from a wasm side module are not reliable in the current webR runtime model.
- DuckHTS registers a package-owned htslib `hFILE` scheme handler implemented in `src/wasm_http_hfile.c` for `http` and `https`.
- This backend uses synchronous `XMLHttpRequest` from the worker for range reads, index probes, and seek behavior.

Browser constraints still apply:

- Remote hosts must allow CORS for the main file **and** index sidecars (`.tbi`/`.csi`), including range requests.
- Behavior can vary by browser and by server-side CORS policy changes over time.
- `ALL_PROXY` / websocket proxy settings do not affect this XHR backend.

Optional header/auth configuration for browser wasm can be provided from JavaScript before loading/querying:

```js
Module.duckhtsWasmHttpConfig = {
  headers: {
    Authorization: "Bearer <short-lived-token>",
    "X-Request-Source": "webr-local"
  },
  allowHosts: ["ftp.ebi.ac.uk", ".s3.amazonaws.com"],
  enforceHostAllowlist: true,
  withCredentials: false,
  allowInsecureAuth: false
};
```

Security behavior of this config:

- Headers are only attached when the URL hostname matches `allowHosts`.
- Requests are blocked for non-matching hosts when `enforceHostAllowlist: true`.
- `Authorization` is blocked on non-HTTPS URLs unless `allowInsecureAuth: true` is set explicitly.
- Credentials/cookies are only sent when `withCredentials: true` is set.

### Browser wasm/duckdb-wasm local setup

DuckHTS is also intended to run as a generic DuckDB community extension in browser wasm hosts (not only webR).

Use the local duckdb-wasm setup to exercise that path end-to-end:

```bash
./scripts/start_duckdb_wasm_local_test.sh
```

This setup uses a Docker-only build path via `scripts/docker/duckdb-wasm-local.Dockerfile`.

The container pre-installs cache-friendly, pinned wasm build dependencies (`emsdk`, `vcpkg`) so repeated local runs do minimal setup work.

Builds run in an isolated mirror worktree (`.duckdb_wasm_docker_work`) and copy back only the wasm extension artifact, so your host native `build/` and `cmake_build/` trees remain available for normal native development and testing.

Then open:

```text
http://127.0.0.1:8001/scripts/duckdb-wasm-local-test.html
```

This setup loads `duckhts.duckdb_extension.wasm` in duckdb-wasm, runs local HTTP reader smoke tests, and lets you set/clear `Module.duckhtsWasmHttpConfig` directly in the browser host runtime.

The setup stages `duckdb-browser.mjs`, `duckdb-browser-eh.worker.js`, and `duckdb-eh.wasm` at the site root for same-origin runtime loading, while using an import map for `apache-arrow` resolution.

### S3 credentials and configuration

The htslib S3 plugin supports credentials embedded in the URL or provided via environment variables or standard credentials files. For AWS-style credentials, the most common variables are:

- `AWS_ACCESS_KEY_ID`
- `AWS_SECRET_ACCESS_KEY`
- `AWS_SESSION_TOKEN` (optional, for temporary credentials)
- `AWS_DEFAULT_REGION`
- `AWS_PROFILE` / `AWS_DEFAULT_PROFILE`
- `AWS_SHARED_CREDENTIALS_FILE` (override credentials file location)

You can also configure htslib-specific settings like `HTS_S3_ADDRESS_STYLE`, `HTS_S3_HOST`, and `HTS_S3_S3CFG` for non-default S3 endpoints or path-style access.

See the htslib S3 plugin documentation for full details, URL syntax, and short‑lived credentials support:
https://www.htslib.org/doc/htslib-s3-plugin.html

## Development

This README is rendered with [`duckknit`](https://github.com/rundel/duckknit) to execute SQL snippets in a persistent DuckDB session.

### Environment setup

Run the one-time configure step to create the Python venv and detect platform settings:

```bash
make configure
```

Note: MSVC builds (windows_amd64/windows_arm64) are not supported. Use MinGW/RTools for Windows.

### Prerequisites

- C compiler (GCC or Clang)
- CMake ≥ 3.5
- Make
- Python 3 + venv
- R with the `rmarkdown` and [`duckknit`](https://github.com/rundel/duckknit) packages.
- Git
- [htslib](https://github.com/samtools/htslib) build dependencies: zlib, libbz2, liblzma, libdeflate, libcurl, libcrypto (OpenSSL)

On Debian/Ubuntu:

```bash
sudo apt install build-essential cmake python3 python3-venv git \
    zlib1g-dev libbz2-dev liblzma-dev libdeflate-dev libcurl4-openssl-dev libssl-dev
```

On macOS:

```bash
brew install cmake htslib xz libdeflate
```

### Vendor [htslib](https://github.com/samtools/htslib)

```bash
./scripts/vendor_htslib.sh
```

This downloads and verifies [htslib](https://github.com/samtools/htslib) 1.24 into `third_party/htslib/`.

### Build

```bash
make configure    # one-time setup (Python venv, platform detection)
make release      # build optimised extension
```

The build runs [htslib](https://github.com/samtools/htslib)'s Makefile (`make lib-static`) in-tree.

The extension binary is written to `build/release/duckhts.duckdb_extension`.

### Debug build

```bash
make debug
```

## Loading

```sql
-- Unsigned extensions must be loaded with -unsigned flag:
-- duckdb -unsigned

LOAD '/path/to/duckhts.duckdb_extension';
```

## Testing

SQL tests live in `test/sql/` using [DuckDB](https://duckdb.org/)'s SQLLogicTest format.

Before running tests for the first time, prepare the indexed test data:

```bash
./test/scripts/prepare_test_data.sh   # requires samtools, bcftools, bgzip, tabix
```

This copies files from the vendored [htslib](https://github.com/samtools/htslib) test suite into `test/data/` and
builds the required indexes (BAI, CSI, TBI, FAI) so region queries work.

Then run:

```bash
make test_release
```

## R demo

The R package lives under `r/Rduckhts` and provides helpers to load the extension
and create [DuckDB](https://duckdb.org/) tables from HTS files. See its README for R-specific usage:
[r/Rduckhts/README.Rmd](r/Rduckhts/README.Rmd).

```{r eval=TRUE}
library(DBI)
library(duckdb)

drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
con <- dbConnect(drv, dbdir = ":memory:")
ext_path <- normalizePath("build/release/duckhts.duckdb_extension", mustWork = FALSE)
dbExecute(con, sprintf("LOAD '%s'", ext_path))

dbGetQuery(con, "
  SELECT *
  FROM read_bcf('test/data/formatcols.vcf.gz', tidy_format := true)
  LIMIT 5
")

parquet_path <- tempfile(fileext = ".parquet")
dbExecute(con, sprintf(
  "COPY (SELECT * FROM read_bcf('test/data/formatcols.vcf.gz', tidy_format := true)) TO '%s' (FORMAT PARQUET)",
  parquet_path
))
file.exists(parquet_path)

dbGetQuery(con, "
  SELECT NAME, SEQUENCE, QUALITY, MATE, PAIR_ID
  FROM read_fastq('test/data/r1.fq', mate_path := 'test/data/r2.fq')
  LIMIT 5
")

dbGetQuery(con, "
  SELECT QNAME, RNAME, POS, READ_GROUP_ID, SAMPLE_ID
  FROM read_bam('test/data/rg.sam.gz')
  LIMIT 5
")

dbGetQuery(con, "
  SELECT idx, raw
  FROM read_hts_header('test/data/formatcols.vcf.gz', mode := 'raw')
  LIMIT 3
")

dbGetQuery(con, "
  SELECT seqname, tid, index_type, chunk_beg_vo, chunk_end_vo
  FROM read_hts_index_spans('test/data/formatcols.vcf.gz')
  LIMIT 3
")

dbGetQuery(con, "
  SELECT index_type, octet_length(raw) AS raw_bytes
  FROM read_hts_index_raw('test/data/formatcols.vcf.gz')
")
```

### SAMtags + auxiliary tags

Standard SAMtags can be surfaced as typed columns and non-standard tags
captured in a map for ad hoc access:

```{r eval=TRUE}
dbGetQuery(con, "
  SELECT RG, NM, map_extract(AUXILIARY_TAGS, 'XZ') AS XZ
  FROM read_bam('test/data/aux_tags.sam.gz', standard_tags := true, auxiliary_tags := true)
  LIMIT 1
")
```

### GFF/GTF annotation attributes

DuckHTS reads GFF3 with `read_gff(...)`, GTF with `read_gtf(...)`, and generic
GFF/GTF-like tabix text with `read_tabix(...)`. `read_gff(..., strict := true)`
performs GFF3 structural validation. Attribute decoding can be scalar and raw
for legacy convenience (`attributes_map`), grouped and lossless for multi-values
(`attributes_list` as `MAP(VARCHAR, VARCHAR[])`), or exact parser-style pairs (`attributes_pairs` as
`LIST<STRUCT(key, value, idx)>`).

The GFF3 implementation is audited against [GFFBase](https://github.com/Kuanhao-Chao/gffbase)
in [`benchmarks/benchmark_gffbase_conformance.md`](benchmarks/benchmark_gffbase_conformance.md),
including Rust/Python parser parity checks, local conformance cases, server
specs, and human-scale GENCODE timing.

```{r eval=TRUE}
dbGetQuery(con, "
  SELECT
    seqname,
    feature,
    start,
    \"end\",
    list_extract(map_extract_value(attributes_list, 'Dbxref'), 1) AS first_dbxref,
    list_count(attributes_pairs) AS n_attr_pairs
  FROM read_gff('test/data/gff_attrs.gff3',
                strict := true,
                attributes_list := true,
                attributes_pairs := true)
")

dbGetQuery(con, "
  SELECT p.key, p.value, p.idx
  FROM read_gff('test/data/gff_attrs.gff3', attributes_pairs := true),
       UNNEST(attributes_pairs) AS u(p)
  WHERE p.key IN ('Alias', 'Note')
  ORDER BY p.key, p.idx
")

dbGetQuery(con, "
  SELECT list_extract(map_extract_value(attributes_list, 'note'), 1) AS note
  FROM read_gtf('test/data/gtf_attrs.gtf', attributes_list := true)
")
```

### Tabix text files

```{r eval=TRUE}
dbGetQuery(con, "
  SELECT column0, column1
  FROM read_tabix('test/data/meta_tabix.tsv.gz')
  LIMIT 2
")

dbGetQuery(con, "
  SELECT chrom, pos
  FROM read_tabix('test/data/header_tabix.tsv.gz', header := true)
  LIMIT 2
")

dbGetQuery(con, "
  SELECT typeof(column1) AS column1_type
  FROM read_tabix('test/data/meta_tabix.tsv.gz', auto_detect := true)
  LIMIT 1
")

dbGetQuery(con, "
  SELECT pos + 1 AS pos_plus_one
  FROM read_tabix('test/data/header_tabix.tsv.gz', header := true,
                  column_types := ['VARCHAR','BIGINT','VARCHAR'])
  LIMIT 1
")

dbDisconnect(con, shutdown = TRUE)
```

## Project Structure

```
src/
  duckhts.c              # Extension entry point and SQL registration
  bcf_reader.c           # VCF/BCF reader (read_bcf)
  bam_reader.c           # SAM/BAM/CRAM reader (read_bam)
  seq_reader.c           # FASTA/FASTQ reader (read_fasta, read_fastq)
  tabix_reader.c         # Tabix/GTF/GFF reader (read_tabix, read_gtf, read_gff)
  vep_parser.c           # VEP/CSQ annotation parser
  kmer_udf.c             # Sequence utilities, including SIMD-routed seq_gc_content
  ......
  simd/                  # Runtime SIMD dispatch and scalar/AVX/NEON/wasm backends
    duckhts_simd_dispatch.c
    duckhts_simd_scalar.c
    duckhts_simd_avx2.c
    duckhts_simd_avx512.c
    duckhts_simd_neon.c
    duckhts_simd_wasm_simd128.c
  include/
    duckhts_simd.h
    duckhts_simd_internal.h
    duckhts_simd_kernels.def
    vcf_types.h
    vep_parser.h
third_party/
  htslib/                # Vendored htslib 1.24 (built automatically)
benchmarks/
  benchmark_simd_seq_gc.Rmd  # Scalar-vs-runtime-SIMD benchmark report
test/
  sql/                   # SQL logic tests, including SIMD conformance
duckdb_capi/
  duckdb.h               # DuckDB C API headers
  duckdb_extension.h
r/
  Rduckhts/              # R package with bundled extension sources
```

## References

- DuckDB: https://duckdb.org/
- DuckDB Extension API: https://duckdb.org/docs/extensions/overview
- DuckDB extension template (C): https://github.com/duckdb/extension-template-c
- htslib: https://github.com/samtools/htslib
- RBCFTools: https://github.com/RGenomicsETL/RBCFTools
- duckknit: https://github.com/rundel/duckknit

## License

MIT
