---
output: github_document
---

<!-- README.md is generated from README.Rmd using [duckknit](https://github.com/rundel/duckknit). Please edit that file. -->

```{r, include = FALSE}
library(duckknit)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# Rewrite duckknit's ``` duckdb fences to ``` sql so GitHub applies
# SQL syntax highlighting in the rendered README.
knitr::knit_hooks$set(document = function(x) {
  gsub("(?m)^``` duckdb", "``` sql", x, perl = TRUE)
})

# Use duckknit's duckdb engine for sql chunks.
knitr::knit_engines$set(sql = knitr::knit_engines$get("duckdb"))

# Pre-start the default duckknit session with -unsigned so unsigned extensions
# can be loaded. duckknit_start_session() only accepts a db path, not extra
# CLI flags, so we launch the process directly and register it in duckknit's
# internal session table.
options(duckknit.duckdb = "/usr/local/bin/duckdb152")
if (!requireNamespace("Rduckhts", quietly = TRUE)) {
  stop("Install the bootstrapped Rduckhts package before rendering README.md")
}
Rduckhts::setup_hts_env()
extension_path <- system.file(
  "duckhts_extension", "build", "duckhts.duckdb_extension",
  package = "Rduckhts", mustWork = TRUE
)
local({
  ns  <- asNamespace("duckknit")
  cmd <- duckknit:::find_duckdb()
  proc <- processx::process$new(
    command = cmd,
    args = c("-unsigned", ":memory:"),
    stdin = "|", stdout = "|", stderr = "|",
    cleanup = TRUE, cleanup_tree = TRUE
  )
  unlockBinding(".sessions", ns)
  ns$.sessions[["readme"]] <- proc
  lockBinding(".sessions", ns)
  unlockBinding(".state", ns)
  ns$.state$last_session <- "readme"
  lockBinding(".state", ns)

  load_sql <- sprintf(
    "SET allow_extensions_metadata_mismatch=true;\nLOAD '%s';",
    gsub("'", "''", extension_path, fixed = TRUE)
  )
  load_result <- duckknit:::duckknit_exec(proc, load_sql)
  if (nzchar(trimws(load_result$stderr))) {
    stop(load_result$stderr, call. = FALSE)
  }
})
```

# DuckHTS

[![CRAN Status](https://www.r-pkg.org/badges/version/Rduckhts)](https://cran.r-project.org/package=Rduckhts)
[![R-universe version](https://RGenomicsETL.r-universe.dev/Rduckhts/badges/version)](https://RGenomicsETL.r-universe.dev/Rduckhts)

Read VCF, BCF, BAM, CRAM, FASTA, FASTQ, BigWig, GTF, GFF, BED, and tabix-indexed
files directly in [DuckDB](https://duckdb.org/). DuckHTS uses
[htslib](https://github.com/samtools/htslib) for HTS formats and provides SQL
functions for consequence annotation, intervals, coverage, sequence operations,
compression, indexing, and export.

## Functions

<details>
<summary>Show generated function catalog</summary>

```{r function-catalog, echo=FALSE, results='asis'}
cat(paste(readLines("r/Rduckhts/inst/function_catalog/functions.md"), collapse = "\n"))
```

</details>

`read_fastq` with `mate_path` requires exact QNAME pairing. `read_bam` supports typed `standard_tags` and `auxiliary_tags` maps. `read_tabix` supports header-aware parsing (`header`, `header_names`) and optional type inference (`auto_detect`, `column_types`). Region lists in comma-separated form are supported by `read_bam`, `read_bcf`, `read_fasta`, `read_bigwig`, `read_gff`, `read_gtf`, and `read_tabix`. Indexed `read_bam`, `read_bcf`, `read_bigwig`, `read_gff`, `read_gtf`, and `read_tabix` multi-region queries emit a matching record once when requested regions overlap. `read_fasta` retains its separate per-region sequence-row contract.

## Examples

The examples below run directly against bundled local test files through the DuckDB CLI and load the built extension from `build/release/duckhts.duckdb_extension`.

### Core readers

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT CHROM, POS, REF, ALT, SAMPLE_ID
FROM read_bcf('test/data/formatcols.vcf.gz', tidy_format := true)
LIMIT 3;

SELECT count(*) AS n
FROM read_bam('test/data/range.bam', region := 'CHROMOSOME_I:1-1000');

SELECT *
FROM fasta_index('test/data/ce.fa');

SELECT NAME, length(SEQUENCE) AS seq_length
FROM read_fasta('test/data/ce.fa', region := 'CHROMOSOME_I:1-25');

SELECT NAME, MATE, PAIR_ID
FROM read_fastq('test/data/interleaved.fq', interleaved := true)
LIMIT 3;

SELECT CHROM, START0, END0, round(VALUE::DOUBLE, 1) AS VALUE
FROM read_bigwig(
  'third_party/libBigWig/test/test.bw',
  region := '1:1-150,10:201-300'
)
ORDER BY CHROM, START0;
```

### BigWig signal tracks

`read_bigwig()` returns the intervals physically stored in a BigWig as
zero-based, half-open `(CHROM, START0, END0, VALUE)` rows. Its optional
`region` uses the same one-based inclusive, comma-separated syntax as the
indexed HTS readers; overlapping requests are merged and do not duplicate a
stored interval. Local files, native HTTP/S3 paths, and browser HTTP use the
same htslib `hFILE` transport already used by DuckHTS. A full scan distributes
nonempty contigs across DuckDB workers; a multi-region scan distributes merged
ranges. `blocks_per_iteration` controls indexed block batching inside a worker,
not the number of workers.

This query reads a real 100 kb slice of the UCSC GRCh38 phyloP 100-way track
rather than converting it to an intermediate text file:

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT count(*) AS stored_intervals,
       min(VALUE)::DOUBLE AS minimum,
       max(VALUE)::DOUBLE AS maximum
FROM read_bigwig(
  'https://hgdownload.soe.ucsc.edu/goldenPath/hg38/phyloP100way/hg38.phyloP100way.bw',
  region := 'chr22:20000000-20099999'
);
```

### Variant normalization

`duckhts_bcftools_norm(...)` applies bcftools-style FASTA-backed allele normalization to a regular table or derived relation while preserving the original columns. In split mode, multiallelic rows are expanded first and then normalized one ALT at a time.

```{sql eval=TRUE, collapse=FALSE, comment=""}
CREATE OR REPLACE TEMP TABLE readme_norm AS
SELECT *
FROM (VALUES
  ('chrS', 2, 'T', 'TT,TTT'),
  ('chrS', 2, 'T', '*,TT')
) AS t(chrom, pos, ref, alt);

SELECT chrom, pos, ref, alt, alt_index,
       pos_normed, ref_normed, alt_normed, norm_status
FROM duckhts_bcftools_norm(
  'readme_norm',
  'test/data/liftover_repeat_src.fa',
  split_multiallelic := true
)
ORDER BY alt, alt_index;
```

### VariantKey + RegionKey

DuckHTS vendors the official VariantKey / RegionKey C API and exposes SQL helpers that mirror bcftools `%VKX`-style VariantKey output on VCF rows. `variantkey(...)` accepts 1-based VCF `POS`, while `regionkey(...)` uses 0-based half-open interval semantics. Large, ambiguous, and symbolic alleles still encode through the official hashed nonreversible VariantKey mode, but those keys do not encode `END`, `SVLEN`, mate breakend coordinates, or other SV metadata; use RegionKey explicitly for span-oriented interval work. See Nicola Asuni (2018) <https://doi.org/10.1101/473744>.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT variantkey_hex(variantkey('1', 324684, 'C', 'G')) AS vkx,
       reverse_variantkey(parse_variantkey_hex('08027a2588b00000')) AS reversed;

SELECT regionkey_hex(regionkey('X', 1007, 1807, 1)) AS rkx,
       are_overlapping_regionkeys(
         regionkey('X', 1007, 1807, 1),
         parse_regionkey_hex('b80001f78000387a')
       ) AS overlaps;
```

### DuckVEP: a resident Ensembl VEP consequence engine

#### Design and data ownership

DuckVEP compiles Ensembl release relations into an immutable, shared transcript
and regulation model. Sorted variant alleles then pass through a native C
candidate sweep, transcript projection, consequence classifier, sequence editor,
NMD evaluator, and optional HGVS renderer. Stable transcript/gene identifiers and
supplementary annotations remain ordinary DuckDB relations and are joined only
after the compact consequence rows have been filtered.

That last sentence is the point of the architecture, not an implementation
detail. DuckVEP is one fast relation producer inside DuckDB. Anything DuckDB can
read--Parquet or GeoParquet, VCF/BCF, CSV/TSV, JSON, an attached DuckLake, a local
DuckDB, HTTP/S3 objects, or a supported external database--can be an annotation
provider without a new DuckHTS file format or C plugin. Exact allele sources use
collision-safe VariantKey joins, gene sources use stable identifiers, and interval
sources use ordinary inequality joins, bounded key prefilters, or cgranges. DuckDB
can execute two-inequality overlap predicates with its native
[IEJoin](https://duckdb.org/2022/05/27/iejoin) implementation and can spill large
join state rather than requiring every provider to live in the resident C model.

The implementation targets executable Ensembl VEP 116 semantics rather than a
new interpretation of the Sequence Ontology rules. The current differential
ledger covers GRCh38, GRCh37, *P. falciparum*, exact structural variants, paired
breakends, regulation/motif features, and variant-induced NMD. See the rendered
[conformance report](benchmarks/duckvep_conformance.md), the
[throughput report](benchmarks/duckvep_throughput.md), and the implementation
contract in [design/duckvep.md](design/duckvep.md).

#### Building a receipted model

The model compiler reads ordinary DuckDB relations. Stage the exact Ensembl core
and funcgen release tables from their MySQL dump files, or attach a read-only
Ensembl MySQL database through DuckDB's `mysql` extension. The compiler is not a
downloader: source retrieval and dump-to-table typing are explicit staging steps,
while the same validation and model receipt apply to either transport.

```sql
-- Reference chunks are zero-based, half-open and sequence-bearing.
CREATE TABLE grch38_reference_chunks AS
SELECT chrom, start, "end", seq
FROM fasta_nuc(
  'Homo_sapiens.GRCh38.dna.primary_assembly.fa',
  bin_width := 1048576,
  include_seq := TRUE
);

-- ensembl_core and ensembl_funcgen are schemas containing the release-116
-- tables named in design/duckvep.md.
CREATE TABLE grch38_regions AS
SELECT * FROM duckvep_ensembl_regions(
  'ensembl_core', 'grch38_reference_chunks', 'GRCh38'
);

CREATE TABLE grch38_transcripts AS
SELECT * FROM duckvep_ensembl_transcripts(
  'ensembl_core', 'grch38_reference_chunks', 'GRCh38'
);

CREATE TABLE grch38_regulation AS
SELECT * FROM duckvep_ensembl_regulation_features(
  'ensembl_funcgen', 'grch38_regions'
);

CREATE TABLE grch38_receipt AS
SELECT * FROM duckvep_model_receipt(
  'grch38_regions', 'grch38_transcripts',
  'Ensembl', '116', 'GRCh38',
  source_manifest_sha256, reference_sha256,
  'VEP 116 core transcript selection',
  regulation_features_table := 'grch38_regulation'
);
```

`duckvep_ensembl_transcripts(...)` applies the pinned VEP core-transcript
selection, reconstructs ranked exon/cDNA/CDS geometry and strand-correct sequence,
loads the sequence-region genetic code (including mitochondrial code), prepares
complete transcript flanks, mature-miRNA segments, and supported Ensembl sequence
edits, and retains MANE/GENCODE/CCDS and stable identifiers as cold relational
columns. GRCh37 is built from the separate release-116 GRCh37 core schema and is
intentionally MANE-free. Regulation and motif features are compiled from the
matching funcgen release, not inferred from transcript overlap.

Publication is separate from compilation. Persist the prepared relations and
load their sorted hot projections into a named resident model with
`duckvep_model_load(...)`. `reference_fasta` enables reference validation and
VEP-compatible three-prime HGVS shifting; `transcript_coverage_complete := TRUE`
permits a no-transcript hit to become a supported `intergenic_variant`.

```sql
SELECT loaded
FROM duckvep_model_load(
  'human_116_grch38',
  'SELECT seq_region, sequence_length, seq_region_name
     FROM grch38_regions ORDER BY seq_region',
  'SELECT transcript_index, seq_region, transcript_start, transcript_end,
          strand, gene_index, transcript_flags, cds_start, cds_end,
          cds_sequence, codon_table, pre_cds_sequence, post_cds_sequence
     FROM grch38_transcripts
    ORDER BY seq_region, transcript_start, transcript_index',
  'SELECT transcript_index, exon.exon_start, exon.exon_end,
          exon.exon_cdna_start, exon.exon_cdna_end, exon.phase, exon.end_phase
     FROM grch38_transcripts,
          LATERAL unnest(exons) AS u(exon)
    ORDER BY transcript_index, exon.exon_cdna_start',
  mature_mirna_query :=
    'SELECT transcript_index, region.mature_mirna_start,
            region.mature_mirna_end
       FROM grch38_transcripts,
            LATERAL unnest(mature_mirna_regions) AS u(region)
      ORDER BY transcript_index, region.mature_mirna_start',
  peptide_edit_query :=
    'SELECT transcript_index, edit.protein_position,
            edit.alternate_amino_acid
       FROM grch38_transcripts,
            LATERAL unnest(peptide_edits) AS u(edit)
      ORDER BY transcript_index, edit.protein_position',
  interval_feature_query :=
    'SELECT regulation_feature_index, seq_region, feature_start, feature_end,
            feature_kind
       FROM grch38_regulation
      ORDER BY seq_region, feature_start, regulation_feature_index',
  reference_fasta := 'reference/Homo_sapiens.GRCh38.dna.primary_assembly.fa',
  transcript_coverage_complete := TRUE
);
```

#### Why sorted input becomes a sweep

The resident transcript arrays are sorted by `(seq_region, transcript_start)`.
For the first allele in a DuckDB vector, cgranges supplies a complete seed set.
As later alleles advance in coordinate order, the worker admits newly reachable
transcripts from the front of the sorted array and retires transcripts whose end
plus the requested flank is behind the allele. It classifies only that active set:

```text
transcripts ordered by start:  T1------T2----------T3-----T4---->
sorted allele stream:               v1  v2       v3    v4       >

seed once -> admit starts now reachable -> retire expired ends
          -> project/classify active transcripts -> advance
```

This is not a 5,000-record or 5,000-base buffer. The default 5,000 bases are only
the VEP-compatible upstream/downstream candidate distances; zero, 10,000, 50,000,
or another supported `UINTEGER` distance uses the same algorithm. Long alleles are
not clipped to that window. Transcript and regulation/motif sweeps own different
active sets because interval features do not use transcript flanks. The immutable
model is shared read-only across workers; each worker owns its candidate arrays,
exon cursors, faidx handle, reference window, edit scratch, and output builder.
Consequently memory grows with the shared model plus bounded per-worker workspace,
not with a copy of the model per thread. The current scalar surface reseeds at each
DuckDB vector; a future stateful table-function surface can preserve the frontier
across vectors without changing the consequence authority.

#### Canonical event relation and output choices

`duckvep_annotate(...)` consumes one coordinate-ordered row per ALT allele. Its
canonical columns cover literal small variants, exact span SVs, and paired
breakends in one relation. Genotypes, phase sets, imprecision intervals, raw ALT,
and sample metadata stay on the source row and join back by `event_index`.

| `rich` | `hgvs` | Result |
|:--|:--|:--|
| `FALSE` | `FALSE` | Compact masks, codes, ordinals, cDNA/CDS/protein positions, amino-acid bytes and NMD codes |
| `TRUE` | `FALSE` | Compact fields plus Consequence, IMPACT, region, amino-acid, NMD and audit text from the same native pass |
| `FALSE` | `TRUE` | Compact fields plus independent-event HGVSc/HGVSn/HGVSp |
| `TRUE` | `TRUE` | Rich consequence and HGVS from one fused candidate-discovery pass |

The transcript window defaults to 5,000 bases in each direction, matching the
VEP default. It is an explicit API parameter rather than a fixed buffer: either
direction may be zero or any supported `UINTEGER` distance. Audit columns named
`duckvep_status` and `duckvep_reason` explain unresolved computation (for example
missing sequence or a reference mismatch); they are not Ensembl CSQ fields.

#### Real human annotation and supplementary providers

The following is the production shape, using the public HG002 40x PCR-free
[DeepVariant GRCh38 WGS callset](https://storage.googleapis.com/brain-genomics-public/research/sequencing/grch38/vcf/hiseqx/wgs_pcr_free/40x/HG002.hiseqx.pcr-free.40x.deepvariant-v1.0.grch38.vcf.gz),
an Ensembl 116 GRCh38 model, dated ClinVar/ClinvArbitration, AlphaMissense v2,
gnomAD v2.1.1 gene constraint, and an Ensembl regulatory Parquet relation. This
callset contains chromosomes 1--22, X, Y, and MT; it is an annotation workload,
not a GIAB truth set. The provider files are ordinary typed Parquet; equivalent
HTTP/S3 paths or DuckLake tables can replace them. Run the statements in one
DuckDB CLI session so `.timer on` reports model load, VCF staging, the consequence
sweep, provider joins, and final Parquet materialization separately.

##### Build the provider relations once

The `providers/*.parquet` paths below are not special DuckHTS files. They are
release-specific projections built with DuckDB from the declared upstream
artifacts:

| Provider relation | Upstream artifact |
|:--|:--|
| `clinvar_20260706_grch38_keyed.parquet` | [ClinVar GRCh38 VCF](https://ftp.ncbi.nlm.nih.gov/pub/clinvar/vcf_GRCh38/), dated 2026-07-06 |
| `clinvarbitration_grch38_keyed.parquet` | [ClinvArbitration record 16792026](https://zenodo.org/records/16792026), whose release contract declares GRCh38 |
| `alphamissense_hg38_variantkey.parquet` | [AlphaMissense v2 GRCh38](https://zenodo.org/records/8360242) |
| UCSC `hg38.phyloP100way.bw` | [GRCh38 phyloP 100-way BigWig](https://hgdownload.soe.ucsc.edu/goldenPath/hg38/phyloP100way/) |
| `ensembl116_grch38_regulatory.parquet` | `grch38_regulation`, compiled above from the Ensembl 116 funcgen tables |
| `gnomad_v211_constraint_gene.parquet` | [gnomAD v2.1.1 gene constraint](https://gnomad.broadinstitute.org/downloads#v2-constraint) |

VariantKey is an exact key over a normalized allele representation. Normalize
arbitrary VCF inputs with `duckhts_bcftools_norm(...)` or an equivalent pinned
normalizer before building either side of the join; keep the untouched record,
genotype arrays and normalization lineage in their source relations.

ClinVar is already VCF, so `read_bcf()` supplies typed INFO fields while ALT
ordinal expansion produces the allele relation.

```sql
COPY (
  WITH alleles AS (
    SELECT duckhts_contig_key(CHROM) AS chrom, POS::BIGINT AS pos,
           upper(REF) AS ref, upper(a.alt) AS alt,
           INFO_ALLELEID AS allele_id,
           INFO_CLNSIG AS clinical_significance
    FROM read_bcf('source/clinvar_20260706.vcf.gz', scan_mode := 'sequential')
    CROSS JOIN unnest(ALT) AS a(alt)
  ), keyed AS (
    SELECT *, variantkey(chrom, pos, ref, alt) AS vk FROM alleles
  )
  SELECT chrom, pos, ref, alt, vk,
         mod(extract_variantkey_refalt(vk), 2) = 1 AS is_hash,
         allele_id, clinical_significance
  FROM keyed WHERE vk IS NOT NULL
  ORDER BY vk
) TO 'providers/clinvar_20260706_grch38_keyed.parquet'
  (FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 122880);
```

ClinvArbitration publishes an aggregate decision table. Its assembly is a
declared GRCh38 release fact; it is not inferred from `chr`-prefixed strings.

```sql
COPY (
  WITH source AS (
    SELECT duckhts_contig_key(contig) AS chrom,
           try_cast(position AS BIGINT) AS pos,
           upper(reference) AS ref, upper(alternate) AS alt,
           clinical_significance,
           try_cast(gold_stars AS UTINYINT) AS gold_stars,
           try_cast(allele_id AS BIGINT) AS allele_id
    FROM read_csv_auto(
      'source/clinvarbitration_16792026.tsv',
      delim := '\t', header := TRUE, all_varchar := TRUE
    )
  ), keyed AS (
    SELECT *, variantkey(chrom, pos, ref, alt) AS vk FROM source
  )
  SELECT *, mod(extract_variantkey_refalt(vk), 2) = 1 AS is_hash
  FROM keyed WHERE vk IS NOT NULL
  ORDER BY vk
) TO 'providers/clinvarbitration_grch38_keyed.parquet'
  (FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 122880);
```

The official AlphaMissense `.tsv.gz` is BGZF-compressed. DuckHTS can build its
coordinate index directly, then use the same file for indexed regional queries
or, as below, stream all 71.7 million rows once to build the reusable pack.

```sql
SELECT * FROM tabix_index(
  'source/AlphaMissense_hg38.tsv.gz',
  preset := 'gff', seq_col := 1, start_col := 2, end_col := 2,
  comment_char := '#', skip_lines := 4, threads := 4
);

COPY (
  WITH keyed AS (
    SELECT variantkey(duckhts_contig_key(chrom), pos, ref, alt) AS vk,
           split_part(transcript_id, '.', 1) AS transcript_stable_id,
           try_cast(split_part(transcript_id, '.', 2) AS UINTEGER)
             AS transcript_version,
           transcript_id, protein_variant, uniprot_id,
           am_pathogenicity, am_class
    FROM read_tabix(
      'source/AlphaMissense_hg38.tsv.gz',
      header_names := [
        'chrom', 'pos', 'ref', 'alt', 'genome', 'uniprot_id',
        'transcript_id', 'protein_variant', 'am_pathogenicity', 'am_class'
      ],
      column_types := [
        'VARCHAR', 'BIGINT', 'VARCHAR', 'VARCHAR', 'VARCHAR', 'VARCHAR',
        'VARCHAR', 'VARCHAR', 'DOUBLE', 'VARCHAR'
      ],
      scan_mode := 'sequential'
    )
  )
  SELECT vk, transcript_stable_id, transcript_version,
         transcript_id, protein_variant, uniprot_id,
         am_pathogenicity, am_class
  FROM keyed WHERE vk IS NOT NULL
  ORDER BY vk, transcript_stable_id, transcript_version, protein_variant
) TO 'providers/alphamissense_hg38_variantkey.parquet'
  (FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 122880);
```

The Ensembl provider is projected from the same release-116 funcgen relation
that supplies the resident core features. Identifiers and SO metadata remain in
Parquet; only compact feature geometry enters the C model.

```sql
COPY (
  WITH source AS (
    SELECT f.regulation_feature_index, f.stable_id, f.feature_class,
           coalesce(f.feature_so_term, f.feature_class) AS so_term,
           r.seq_region_name AS chrom,
           f.feature_start::BIGINT - 1 AS start0,
           f.feature_end::BIGINT AS end0
    FROM grch38_regulation f
    JOIN grch38_regions r USING (seq_region)
  ), keyed AS (
    SELECT *, regionkey(chrom, start0, end0) AS rk FROM source
  )
  SELECT *, rk >> 31 AS rk_chrom_start,
         regionkey(chrom, end0, end0) >> 31 AS rk_chrom_end
  FROM keyed WHERE rk IS NOT NULL
  ORDER BY rk
) TO 'providers/ensembl116_grch38_regulatory.parquet'
  (FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 122880);
```

Gene constraint joins by stable Ensembl gene identifier and therefore needs no
genomic key.

```sql
COPY (
  SELECT gene_id, gene AS gene_symbol, transcript,
         try_cast(pLI AS DOUBLE) AS pLI,
         try_cast(oe_lof_upper AS DOUBLE) AS oe_lof_upper
  FROM read_csv_auto(
    'source/gnomad.v2.1.1.lof_metrics.by_gene.txt.bgz',
    delim := '\t', header := TRUE, nullstr := 'NA'
  )
  WHERE gene_id IS NOT NULL
) TO 'providers/gnomad_v211_constraint_gene.parquet'
  (FORMAT PARQUET, COMPRESSION ZSTD);
```

The resulting files are reusable annotation-pack relations. Sorting exact
providers by `vk` and interval providers by `rk` gives Parquet zonemaps useful
ordering information; DuckDB still performs the final collision or overlap
predicate. The full provider preparation and join measurements are in the
[supplementary annotation benchmark](benchmarks/benchmark_variantkey_join_overlap.md).

```{r duckvep-human-evidence, include=FALSE}
human_evidence <- read.csv(
  "benchmarks/data/duckvep_human_annotation.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character")
)
stopifnot(
  nrow(human_evidence) == 13L,
  length(unique(human_evidence$input_alleles)) == 1L,
  length(unique(human_evidence$threads)) == 1L
)
human_fingerprint <- read.csv(
  "benchmarks/data/duckvep_human_annotation_fingerprint.csv",
  stringsAsFactors = FALSE,
  colClasses = "character"
)
stopifnot(
  nrow(human_fingerprint) == 2L,
  length(unique(human_fingerprint$rows)) == 1L,
  length(unique(human_fingerprint$xor64)) == 1L,
  length(unique(human_fingerprint$sum64)) == 1L
)
human_measure <- function(stage, denominator) {
  row <- human_evidence[human_evidence$stage == stage, , drop = FALSE]
  stopifnot(nrow(row) == 1L)
  sprintf(
    "%.3f seconds; %s %s",
    row$seconds,
    format(row$result_rows, big.mark = ",", scientific = FALSE),
    denominator
  )
}
```

##### Session and resident model

`.timer on` makes the DuckDB CLI report every following statement separately.
Keep the thread count and memory limit fixed for the complete run. Execute the
`duckvep_model_load(...)` statement above in this session before staging the
alleles; the checked run loaded the full Ensembl 116 model in
`r human_measure("model_load", "transcripts")`.

```sql
.timer on
SET threads = 4;
SET memory_limit = '4GB';
SET temp_directory = 'case/duckdb-tmp';
```

##### Canonical allele relation

The input VCF is a record relation, while the consequence engine accepts an
event relation with one row per ALT. `record_index` and the one-based
`alt_index` preserve the route back to the original record and its genotype
arrays. This small-variant lane admits literal nucleotide alleles only; exact
SVs and paired breakends use the same event schema through the typed columns
that are `NULL` below.

```sql
-- Preserve the original VCF record and ALT numbering before normalization.
-- The complete sample/FORMAT columns may stay in a wider source relation and
-- join back through (record_index, alt_index).
CREATE TABLE case_records AS
SELECT row_number() OVER ()::UBIGINT AS record_index,
       CHROM AS source_chrom, duckhts_contig_key(CHROM) AS chrom,
       POS::BIGINT AS pos, REF AS ref, ALT AS alt
FROM read_bcf(
  'case/HG002.hiseqx.pcr-free.40x.deepvariant-v1.0.grch38.vcf.gz',
  scan_mode := 'sequential', decompression_threads := 0
);
```

```sql
-- Split ALT with its original one-based ordinal, validate REF against the
-- release-matched FASTA, and left-align/minimize before any exact key is made.
CREATE TABLE case_normalized AS
SELECT record_index, alt_index, source_chrom, chrom,
       pos_normed AS pos, upper(ref_normed) AS ref,
       upper(alt_normed) AS alt, norm_status
FROM duckhts_bcftools_norm(
  'case_records',
  'reference/Homo_sapiens.GRCh38.dna.primary_assembly.fa',
  split_multiallelic := TRUE
)
WHERE pos_normed IS NOT NULL
  AND regexp_full_match(ref_normed, '[ACGTNacgtn]+')
  AND regexp_full_match(alt_normed, '[ACGTNacgtn]+')
  AND upper(ref_normed) <> upper(alt_normed);
```

```sql
CREATE TABLE case_alleles AS
WITH interpreted AS (
  -- This calls the same event interpreter as the consequence kernel. It is
  -- not another SQL trim implementation.
  SELECT n.*, duckvep_allele_geometry(pos, ref, alt) AS geometry
  FROM case_normalized n
), keyed AS (
  SELECT i.*, r.seq_region,
         variantkey(chrom, pos, ref, alt) AS vk,
         regionkey(chrom, geometry.feature_start0,
                   geometry.feature_end0) AS feature_rk
  FROM interpreted i
  JOIN grch38_regions r ON r.seq_region_name = i.chrom
)
SELECT row_number() OVER (
         ORDER BY seq_region, pos, record_index, alt_index
       )::UBIGINT AS event_index,
       record_index, alt_index, seq_region, chrom,
       pos::UBIGINT AS position, ref AS reference, alt AS alternate,
       vk, mod(extract_variantkey_refalt(vk), 2) = 1 AS is_hash,
       geometry.kind_code, geometry.interbase,
       geometry.raw_start0, geometry.raw_end0,
       geometry.feature_start0, geometry.feature_end0,
       geometry.edit_start0, geometry.edit_end0,
       geometry.insertion_boundary0,
       feature_rk >> 31 AS feature_rk_chrom_start,
       regionkey(chrom, geometry.feature_end0,
                 geometry.feature_end0) >> 31 AS feature_rk_chrom_end,
       NULL::UBIGINT AS end_position,
       NULL::VARCHAR AS structural_type,
       NULL::VARCHAR AS copy_change,
       NULL::UINTEGER AS mate_seq_region,
       NULL::UBIGINT AS mate_position
FROM keyed
WHERE vk IS NOT NULL AND feature_rk IS NOT NULL
-- Coordinate order is the sweep contract. Normalization can move POS, so sort
-- only after normalization and event interpretation.
ORDER BY seq_region, position, event_index;
```

Measured VCF decoding, ALT expansion, normalization, keying and ordering:
`r human_measure("vcf_decode_canonicalize_sort", "alleles staged")`.

##### Consequence and HGVS

The resident model contains only the hot transcript, exon, sequence and core
feature fields. Stable identifiers and supplementary provider payloads remain
in DuckDB and are joined later. `hgvs` and `rich` change output projection; they
do not cause a second candidate-discovery pass.

Do not retain the complete result of `duckvep_annotate(...)` before writing it:
this WGS emits 88 million rows. The final query below consumes the annotator
directly. In isolation, the same four-thread rich/HGVS stream writes all rows in
`r human_measure("DuckVEP_rich_HGVS_regulatory_parallel_stream_4GB", "annotation rows")`.

##### Exact allele providers

ClinVar and ClinvArbitration are allele-level providers. VariantKey can encode
short alleles reversibly or store a hash for longer alleles. The `is_hash`
equality prevents mixing those representations; hashed matches are then checked
against the normalized literal contig, position, REF and ALT so a collision
cannot become clinical evidence. Lists of structs keep each provider record
intact instead of independently aggregating fields that belong together.

```sql
CREATE TABLE case_clinvar AS
SELECT q.event_index,
       list_distinct(list(struct_pack(
         allele_id := p.allele_id,
         clinical_significance := p.clinical_significance
       ))) AS clinvar_records
FROM case_alleles q
JOIN read_parquet('providers/clinvar_20260706_grch38_keyed.parquet') p
  ON q.vk = p.vk AND q.is_hash = p.is_hash
 AND (NOT q.is_hash OR
      (q.chrom = p.chrom AND q.position = p.pos AND
       q.reference = p.ref AND q.alternate = p.alt))
GROUP BY q.event_index;
```

Measured dated ClinVar join:
`r human_measure("ClinVar_exact_collision_safe", "matched alleles")`.

```sql
CREATE TABLE case_clinvarbitration AS
SELECT q.event_index,
       list_distinct(list(struct_pack(
         allele_id := p.allele_id,
         classification := p.clinical_significance,
         gold_stars := p.gold_stars
       ))) AS clinvarbitration_records
FROM case_alleles q
JOIN read_parquet('providers/clinvarbitration_grch38_keyed.parquet') p
  ON q.vk = p.vk AND q.is_hash = p.is_hash
 AND (NOT q.is_hash OR
      (q.chrom = p.chrom AND q.position = p.pos AND
       q.reference = p.ref AND q.alternate = p.alt))
GROUP BY q.event_index;
```

Measured ClinvArbitration join:
`r human_measure("ClinvArbitration_exact_collision_safe", "matched alleles")`.

AlphaMissense contains single-nucleotide substitutions, whose VariantKeys are
reversible, so hashed events are excluded rather than subjected to a literal
fallback comparison. Scores are transcript-specific: the provider pack retains
the versioned transcript accession and the join resolves it to the same resident
`transcript_index` emitted by DuckVEP. Taking the maximum score across all
transcripts would attach one transcript's prediction to unrelated transcript
consequence rows.

```sql
CREATE TABLE case_alphamissense AS
SELECT q.event_index, t.transcript_index,
       list_distinct(list(struct_pack(
         transcript_id := p.transcript_id,
         protein_variant := p.protein_variant,
         uniprot_id := p.uniprot_id,
         pathogenicity := p.am_pathogenicity,
         classification := p.am_class
       ))) AS alphamissense_records
FROM case_alleles q
JOIN read_parquet('providers/alphamissense_hg38_variantkey.parquet') p
  ON q.vk = p.vk
JOIN grch38_transcripts t
  ON t.transcript_stable_id = p.transcript_stable_id
 AND t.transcript_version = p.transcript_version
WHERE NOT q.is_hash
GROUP BY q.event_index, t.transcript_index;
```

Measured AlphaMissense join:
`r human_measure("AlphaMissense_exact", "matched alleles")`.

##### Dense conservation signal

BigWig is already an indexed, compressed genomic signal store, so conservation
does not need a bespoke DuckVEP cache. Read only the case ranges, preserve
missing bases explicitly, and reduce overlaps in SQL. Conservation over a
replacement or deletion uses the minimized affected reference span, not its VCF
padding anchor. A pure insertion affects no reference base; report its immediate
left and right reference flanks separately instead of fabricating a one-base
mean. This example uses the real UCSC phyloP 100-way track and a chromosome-22
slice; production orchestration can construct comma-separated batches from the
distinct case coordinates and combine them with `UNION ALL BY NAME`.

```sql
CREATE TABLE case_phylop_chr22 AS
WITH signal AS (
  SELECT duckhts_contig_key(CHROM) AS chrom,
         START0::BIGINT AS start0, END0::BIGINT AS end0,
         VALUE::DOUBLE AS value
  FROM read_bigwig(
    'https://hgdownload.soe.ucsc.edu/goldenPath/hg38/phyloP100way/hg38.phyloP100way.bw',
    region := 'chr22:20000000-20099999'
  )
), requests AS (
  SELECT event_index, 'span' AS request_kind,
         edit_start0 AS start0, edit_end0 AS end0
  FROM case_alleles q
  WHERE q.chrom = '22'
    AND NOT interbase
    AND edit_start0 < 20099999 AND edit_end0 > 19999999
  UNION ALL
  SELECT event_index, 'left', insertion_boundary0 - 1, insertion_boundary0
  FROM case_alleles q
  WHERE q.chrom = '22' AND interbase AND insertion_boundary0 > 0
    AND insertion_boundary0 BETWEEN 20000000 AND 20099999
  UNION ALL
  SELECT event_index, 'right', insertion_boundary0, insertion_boundary0 + 1
  FROM case_alleles q
  WHERE q.chrom = '22' AND interbase
    AND insertion_boundary0 BETWEEN 19999999 AND 20099998
), overlaps AS (
  SELECT r.event_index, r.request_kind, r.end0 - r.start0 AS requested_bases,
         greatest(0, least(r.end0, s.end0) - greatest(r.start0, s.start0))
           AS observed_bases, s.value
  FROM requests r
  LEFT JOIN signal s
    ON r.start0 < s.end0 AND r.end0 > s.start0
)
SELECT event_index,
       sum(observed_bases) FILTER (WHERE request_kind = 'span')
         AS observed_bases,
       max(requested_bases) FILTER (WHERE request_kind = 'span')
         AS requested_bases,
       sum(observed_bases * value) FILTER (WHERE request_kind = 'span') /
         nullif(sum(observed_bases) FILTER (WHERE request_kind = 'span'), 0)
         AS phyloP_mean,
       min(value) FILTER (WHERE request_kind = 'span') AS phyloP_min,
       max(value) FILTER (WHERE request_kind = 'span') AS phyloP_max,
       max(value) FILTER (
         WHERE request_kind = 'left' AND observed_bases = 1
       ) AS insertion_left_phyloP,
       max(value) FILTER (
         WHERE request_kind = 'right' AND observed_bases = 1
       ) AS insertion_right_phyloP
FROM overlaps
GROUP BY event_index;
```

The overlap width weights run-length intervals correctly. `NULL` means the
requested span or flank had no observed phyloP value; it is not silently
converted to zero. Repeated cohort workloads may explicitly materialize release-labelled,
coordinate-sorted Parquet or DuckLake tiles, but BigWig remains a valid provider
and precision authority without such conversion.

##### Interval provider

For RegionKey-supported contigs, shifting a RegionKey right by 31 bits produces
an exact ordered `(chromosome code, start)` value. Constructing the same value
at the half-open end gives an exact `(chromosome code, end)` value. The two
inequalities below therefore express both chromosome identity and half-open
overlap, and DuckDB plans them as
[IEJoin](https://duckdb.org/2022/05/27/iejoin).

This example deliberately uses DuckVEP's VEP-feature coordinates. For an
insertion they are an empty interval at the interbase site, so the strict range
inequalities require both adjacent reference positions to lie inside the
provider interval, matching VEP's core-feature test. A provider whose contract
is “affected reference bases” should instead use `edit_start0/edit_end0` and an
explicit insertion policy, as the phyloP query does above. The uploaded
`raw_start0/raw_end0` span is provenance, not an overlap default.

Do not add `q.chrom = p.chrom` to this plan. DuckDB then chooses a chromosome
hash join and tests the range predicates across enormous same-chromosome
candidate sets. On this workload that physical-plan change was more than two
orders of magnitude slower. Use `EXPLAIN` to confirm `IE_JOIN`. For accessions,
patches, or arbitrary species contigs outside RegionKey's encoding domain, use
`duckhts_cgranges_overlaps_bulk(...)` instead; it preserves literal contig
identity and has its own measured memory/speed tradeoff.

```sql
CREATE TABLE case_interval AS
SELECT q.event_index,
       list_distinct(list(struct_pack(
         regulation_feature_index := p.regulation_feature_index,
         stable_id := p.stable_id,
         so_term := p.so_term
       ))) AS interval_records
FROM case_alleles q
JOIN read_parquet('providers/ensembl116_grch38_regulatory.parquet') p
  ON q.feature_rk_chrom_start < p.rk_chrom_end
 AND q.feature_rk_chrom_end > p.rk_chrom_start
GROUP BY q.event_index;
```

Measured Ensembl regulatory interval join:
`r human_measure("Ensembl_regulatory_RegionKey_IEJoin_stream", "matched alleles")`.
The complete cgranges build, bulk query, aggregation, and write takes
`r human_measure("Ensembl_regulatory_cgranges_bulk_stream", "matched alleles")`.

##### Streamed result

Transcript and regulation-feature indices are compact model-local ordinals.
Resolve them to stable identifiers only after annotation, then attach gene-level
and event-level payloads. The interval query above demonstrates and measures a
general relational overlap provider; it is not needed to rediscover the same
Ensembl core feature already selected by the native sweep. Core feature metadata
joins directly by `regulation_feature_index`, which cannot attach a nearby but
different feature. This keeps provider strings out of the shared C model and lets
DuckDB project or omit them normally.

The query streams annotation into the joins and uses one Parquet writer per
DuckDB worker. `event_index` is the stable identity and ordering key; forcing a
global `ORDER BY` or a single output file serializes the writer and requires a
large sort/materialization. Query the resulting Parquet dataset with an
explicit `ORDER BY` only when a consumer actually requires ordered rows.

```sql
COPY (
  SELECT q.chrom, q.position, q.reference, q.alternate,
         a.consequence, a.impact, a.cdna_position, a.cds_position,
         a.protein_position, a.reference_amino_acid,
         a.alternate_amino_acid, a.nmd_prediction,
         a.transcript_hgvs, a.protein_hgvs,
         t.transcript_stable_id, t.gene_stable_id, t.mane_select_refseq,
         rf.stable_id AS regulation_stable_id,
         rf.feature_so_term AS regulation_so_term,
         cv.clinvar_records, ca.clinvarbitration_records,
         am.alphamissense_records,
         gc.pLI, gc.oe_lof_upper,
         a.overlap_object, a.duckvep_status, a.duckvep_reason
  FROM duckvep_annotate(
    'case_alleles', 'human_116_grch38',
    hgvs := TRUE, rich := TRUE,
    upstream_distance := 5000, downstream_distance := 5000
  ) a
  JOIN case_alleles q USING (event_index)
  LEFT JOIN grch38_transcripts t USING (transcript_index)
  LEFT JOIN grch38_regulation rf USING (regulation_feature_index)
  LEFT JOIN read_parquet('providers/gnomad_v211_constraint_gene.parquet') gc
    ON gc.gene_id = t.gene_stable_id
  LEFT JOIN case_clinvar cv USING (event_index)
  LEFT JOIN case_clinvarbitration ca USING (event_index)
  LEFT JOIN case_alphamissense am
    ON am.event_index = a.event_index
   AND am.transcript_index = a.transcript_index
) TO 'case/HG002.deepvariant-v1.0.duckvep'
  (FORMAT PARQUET, COMPRESSION ZSTD, PER_THREAD_OUTPUT TRUE);
```

Measured model load, fused consequence/HGVS, all late joins, and four-writer
ZSTD Parquet output:
`r human_measure("compose_all_providers_parallel_stream_4GB", "annotation rows written")`.

##### Whole-genome measurement

The checked real-data integration run retained
`r format(unique(human_evidence$input_alleles), big.mark = ",")` alleles from
chromosomes 1--22, X, Y, and MT and emitted
`r format(human_evidence$result_rows[human_evidence$stage == "compose_all_providers_parallel_stream_4GB"], big.mark = ",")`
consequence/core-feature rows. With a 4 GB DuckDB memory limit, four unpinned
workers, warm input pages, and four Parquet writers, the model-load plus
annotation/composition stream completed in
`r human_measure("compose_all_providers_parallel_stream_4GB", "annotation rows written")`.
GNU `time -v` recorded
`r sprintf("%.2f GiB", human_evidence$max_rss_kb[human_evidence$stage == "compose_all_providers_parallel_stream_4GB"] / 1024^2)`
peak process RSS. That high-water mark includes the transient full-model load;
the immutable C model is allocated outside DuckDB's buffer-manager limit. The
checked full-row fingerprint matches the reference output across all
88,392,840 rows.

```{r duckvep-human-measured-results, echo=FALSE}
measured_stages <- c(
  "Ensembl_regulatory_RegionKey_IEJoin_stream",
  "Ensembl_regulatory_cgranges_bulk_stream",
  "DuckVEP_rich_HGVS_regulatory_parallel_stream_4GB",
  "compose_all_providers_parallel_stream_4GB"
)
measured <- human_evidence[
  match(measured_stages, human_evidence$stage),
  c("stage", "result_rows", "seconds", "max_rss_kb", "output_bytes")
]
stopifnot(!anyNA(measured$stage))
measured$stage <- c(
  "RegionKey IEJoin + interval result",
  "cgranges build/query + interval result",
  "rich + HGVS + core regulation, four writers",
  "rich + HGVS + all providers, four writers"
)
measured$result_rows <- format(
  measured$result_rows, big.mark = ",", scientific = FALSE
)
measured$max_rss_kb <- sprintf("%.2f", measured$max_rss_kb / 1024^2)
measured$output_bytes <- format(
  measured$output_bytes, big.mark = ",", scientific = FALSE
)
names(measured) <- c(
  "measured stage", "rows", "seconds", "peak RSS (GiB)", "output bytes"
)
knitr::kable(measured, row.names = FALSE)
```

The exact-source benchmark separately measures the complete 4.0-million-allele
GIAB probe against ClinVar, ClinvArbitration, AlphaMissense, REVEL, gene constraint,
and interval providers, including one-, four-, and twelve-provider plans, Parquet
row-group pruning, materialized output, and peak RSS. See the
[supplementary annotation benchmark](benchmarks/benchmark_variantkey_join_overlap.md).

#### Bundled offline lifecycle model

The deterministic fixture below remains deliberately small so every README render
can execute it offline. It checks the public load/annotate/drop lifecycle; it is not
the human performance or biological example.

```{sql eval=TRUE, collapse=FALSE, comment=""}
CREATE OR REPLACE TABLE readme_duckvep_regions AS
SELECT 1::UINTEGER AS seq_region, 50000::UBIGINT AS sequence_length,
       '11'::VARCHAR AS seq_region_name;

CREATE OR REPLACE TABLE readme_duckvep_transcripts AS
SELECT 0::UINTEGER AS transcript_index, 1::UINTEGER AS seq_region,
       100::UBIGINT AS transcript_start, 150::UBIGINT AS transcript_end,
       1::TINYINT AS strand, 0::UINTEGER AS gene_index,
       0::UBIGINT AS transcript_flags, NULL::UBIGINT AS cds_start,
       NULL::UBIGINT AS cds_end, NULL::BLOB AS cds_sequence,
       NULL::UTINYINT AS codon_table, NULL::BLOB AS pre_cds_sequence,
       NULL::BLOB AS post_cds_sequence;

CREATE OR REPLACE TABLE readme_duckvep_exons AS
SELECT 0::UINTEGER AS transcript_index, 100::UBIGINT AS exon_start,
       150::UBIGINT AS exon_end, 1::UBIGINT AS exon_cdna_start,
       51::UBIGINT AS exon_cdna_end, -1::TINYINT AS phase,
       -1::TINYINT AS end_phase;

SELECT loaded
FROM duckvep_model_load(
  'readme',
  'SELECT * FROM readme_duckvep_regions ORDER BY seq_region',
  'SELECT * FROM readme_duckvep_transcripts ORDER BY seq_region, transcript_start, transcript_index',
  'SELECT * FROM readme_duckvep_exons ORDER BY transcript_index, exon_cdna_start',
  reference_fasta := 'test/data/fixture_ref.fa'
);

CREATE OR REPLACE TABLE readme_duckvep_events AS
SELECT * FROM (VALUES
  (1::UBIGINT, 1::UINTEGER, 124::UBIGINT, 'A'::VARCHAR, 'G'::VARCHAR,
   NULL::UBIGINT, NULL::VARCHAR, NULL::VARCHAR,
   NULL::UINTEGER, NULL::UBIGINT)
) AS e(event_index, seq_region, position, reference, alternate,
       end_position, structural_type, copy_change,
       mate_seq_region, mate_position);

SELECT a.event_index, a.transcript_index, a.consequence, a.impact,
       a.cdna_position, a.transcript_hgvs, a.protein_hgvs,
       a.duckvep_status
FROM duckvep_annotate(
  'readme_duckvep_events', 'readme', hgvs := true,
  upstream_distance := 0, downstream_distance := 0, rich := true
) AS a
ORDER BY a.event_index, a.transcript_index;

SELECT duckvep_model_drop('readme') AS dropped;
```

#### Supplementary annotation composition

DuckHTS does not define a closed supplementary-annotation plugin API because
DuckDB already supplies the larger interface: a relation. For high-cardinality
pipelines, filter compact integer masks/codes first and decode selected SO bits
with `duckvep_so_terms()` afterward. Join `transcript_index`, `gene_index`, and
`regulation_feature_index` to the prepared model relations for stable IDs and
cold metadata. Exact or positional resources such as ClinVar, population
frequencies, calibrated pathogenicity scores, and splice scores use collision-safe
allele keys; protein/domain and regulatory tracks use half-open interval overlap;
constraint, PanelApp, OMIM, and phenotype resources use stable gene/disease IDs.

Those providers may be local Parquet, partitioned S3 Parquet, DuckLake snapshots,
attached DuckDB databases, tabix-indexed files, or any other source for which
DuckDB has a scanner. The optimizer can push columns and predicates into scans,
prune sorted Parquet row groups, plan several equality joins together, execute
two-inequality overlaps with
[IEJoin](https://duckdb.org/2022/05/27/iejoin), and spill join blocks when required.
cgranges remains useful for a repeatedly queried immutable interval registry;
it is an execution choice, not a provider format. This makes an annotation-pack
release a set of versioned relations and provenance, while a new case or cohort
is only the delta that must be scanned, annotated, and joined.

#### Testing and conformance infrastructure

The test system is deliberately several independent authorities rather than one
large collection of expected strings:

1. Pure-C unit and property oracles compare the optimized sweep, projection,
   edit, translation, consequence, HGVS, SV/BND, and workspace paths with slower
   reference implementations. ASan and UBSan execute the same properties.
2. Mass statistical campaigns use a declared seed and explicit rare-state strata:
   strand, exon rank, splice offset, CDS/UTR/start/stop geometry, phase, length
   change and modulo three, incomplete codons, NMD thresholds, HGVS shift limits,
   structural containment, BND orientation, and transcript-window distances.
   Coverage counters are part of the result. A passing campaign with a required
   zero-count state fails, and every counterexample is retained and minimized into
   a fixed witness.
3. The generated finite rule program is exhaustively checked over its Boolean
   consequence-predicate quotient. A transition manifest then names the wider
   geometry/state paths and the fixed, statistical, or executable evidence that
   currently covers each one. This is a compact proof obligation, not a claim that
   sampling proves an unobserved state.
4. Executable differentials run the pinned VEP 116/BioPerl environment on the
   identical allele relation and full-outer-join unique allele/object pairs.
   Consequence, NMD, HGVSc, HGVSn, HGVSp, unresolved rows, and missing/extra objects
   remain separate fatal metrics; there is no release option that permits HGVS
   discordance.
5. Public corpora search different state distributions: complete/dense ClinVar,
   whole GIAB, selected GRCh37, non-human codon tables, regulation/motif, exact SV,
   paired BND, and HPRC/pangenome-derived long alleles. A sparse WGS average cannot
   replace a dense transcript/HGVS or generated structural campaign.

The repository mirrors the relevant Ensembl VEP and Ensembl Variation self-test
cases with exact source lineage. The optional `{targets}` project orchestrates
coarse campaign invalidation, branching, retry, and retained artifacts; it does
not replace the biological comparison logic. Perl/BioPerl plus micromamba provide
the pinned oracle environment, and [`blit`](https://github.com/WangLabCSU/blit) is the
non-interactive process runner. Cache-backed campaigns create one acquisition
receipt and cheaply revalidate the complete installed-file inventory before reuse,
instead of hashing tens of gigabytes on every invocation. See
[the corpus workflow](design/duckvep_corpus_workflow.md) for exact commands,
artifact layers, large-corpus derivation, second-machine reproduction, and the
procedure for moving from VEP 116 to the next release.

#### Measured performance and FastVEP comparison

```{r duckvep-readme-benchmark-evidence, include=FALSE}
format_count <- function(x) {
  format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE)
}

surface_history <- read.csv(
  "benchmarks/data/duckvep_throughput.csv",
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character")
)
surface_modes <- c("compact", "rich", "hgvs", "rich_hgvs")
surface_threads <- c(1L, 4L)
surface_expected <- as.vector(outer(
  surface_modes,
  surface_threads,
  FUN = function(mode, threads) paste(mode, threads, sep = ":")
))
surface_candidates <- surface_history[
  surface_history$workload ==
    "ensembl116_grch38_giab_hg002_v4_2_1_full_literal_public_relation" &
    surface_history$output_mode %in% surface_modes &
    surface_history$threads %in% surface_threads &
    surface_history$input_partitions == 1L &
    surface_history$transcript_distance == 5000L &
    surface_history$regulation_features == 1383580L &
    !is.na(surface_history$corpus_source_alt_count),
  ,
  drop = FALSE
]
surface_groups <- split(surface_candidates, surface_candidates$source_revision)
surface_complete <- names(Filter(
  function(x) {
    nrow(x) == length(surface_expected) &&
      setequal(
        paste(x$output_mode, x$threads, sep = ":"),
        surface_expected
      )
  },
  surface_groups
))
surface_complete_in_order <- unique(surface_candidates$source_revision[
  surface_candidates$source_revision %in% surface_complete
])
stopifnot(length(surface_complete_in_order) > 0L)
surface_revision <- tail(surface_complete_in_order, 1L)
surface_evidence <- surface_candidates[
  surface_candidates$source_revision == surface_revision,
  ,
  drop = FALSE
]
stopifnot(
  nrow(surface_evidence) == 8L,
  length(unique(surface_evidence$variants)) == 1L,
  length(unique(surface_evidence$annotated_rows)) == 1L,
  length(unique(surface_evidence$corpus_source_alt_count)) == 1L,
  length(unique(surface_evidence$regulation_features)) == 1L
)
surface_rate <- function(mode, threads) {
  row <- surface_evidence[
    surface_evidence$output_mode == mode &
      surface_evidence$threads == threads,
    ,
    drop = FALSE
  ]
  stopifnot(nrow(row) == 1L)
  format_count(row$variants_per_second[[1L]])
}

time_value <- function(lines, label) {
  match <- grep(paste0("^\\s*", label, ":"), lines, value = TRUE)
  stopifnot(length(match) == 1L)
  fields <- strsplit(match, ": ", fixed = TRUE)[[1L]]
  stopifnot(length(fields) == 2L)
  trimws(fields[[2L]])
}
elapsed_seconds <- function(value) {
  fields <- as.numeric(strsplit(value, ":", fixed = TRUE)[[1L]])
  stopifnot(length(fields) %in% c(2L, 3L), !anyNA(fields))
  if (length(fields) == 2L) {
    fields[[1L]] * 60 + fields[[2L]]
  } else {
    fields[[1L]] * 3600 + fields[[2L]] * 60 + fields[[3L]]
  }
}
read_elapsed <- function(path) {
  elapsed_seconds(time_value(
    readLines(path, warn = FALSE),
    "Elapsed \\(wall clock\\) time \\(h:mm:ss or m:ss\\)"
  ))
}
fast_data <- "benchmarks/data/duckvep_fastvep"
duck_one <- median(vapply(
  c(
    "duckvep_giab_threads1.time",
    "duckvep_giab_threads1_run2.time",
    "duckvep_giab_threads1_run3.time"
  ),
  function(x) read_elapsed(file.path(fast_data, x)),
  numeric(1)
))
fast_one <- median(vapply(
  c(
    "fastvep_giab_threads1.time",
    "fastvep_giab_threads1_run2.time",
    "fastvep_giab_threads1_run3.time"
  ),
  function(x) read_elapsed(file.path(fast_data, x)),
  numeric(1)
))
duck_four <- median(vapply(
  c(
    "duckvep_giab_threads4.time",
    "duckvep_giab_threads4_run2.time",
    "duckvep_giab_threads4_run3.time"
  ),
  function(x) read_elapsed(file.path(fast_data, x)),
  numeric(1)
))
fast_four <- read_elapsed(file.path(fast_data, "fastvep_giab_threads4.time"))

surface_workload <- data.frame(
  measure = c(
    "source ALT alleles",
    "model-addressable literal alleles",
    "annotation rows",
    "resident regulatory/motif features",
    "transcript flank distance",
    "source revision"
  ),
  value = c(
    format_count(unique(surface_evidence$corpus_source_alt_count)),
    format_count(unique(surface_evidence$variants)),
    format_count(unique(surface_evidence$annotated_rows)),
    format_count(unique(surface_evidence$regulation_features)),
    paste0(format_count(unique(surface_evidence$transcript_distance)), " bases"),
    substr(surface_revision, 1L, 8L)
  )
)
surface_rates <- data.frame(
  projection = c("compact", "rich", "HGVS", "rich + HGVS"),
  `one pinned core` = vapply(
    surface_modes,
    surface_rate,
    character(1),
    threads = 1L
  ),
  `four pinned cores` = vapply(
    surface_modes,
    surface_rate,
    character(1),
    threads = 4L
  ),
  check.names = FALSE
)
fast_comparison <- data.frame(
  `pinned physical cores` = c(1L, 4L),
  `DuckVEP seconds` = sprintf("%.2f", c(duck_one, duck_four)),
  `FastVEP seconds` = sprintf("%.2f", c(fast_one, fast_four)),
  `FastVEP / DuckVEP` = sprintf(
    "%.2fx",
    c(fast_one / duck_one, fast_four / duck_four)
  ),
  check.names = FALSE
)
```

The resident-engine benchmark uses the complete GIAB HG002 v4.2.1 input and the
complete Ensembl 116 GRCh38 transcript and core-feature model.

```{r duckvep-readme-workload-table, echo=FALSE}
knitr::kable(surface_workload, row.names = FALSE)
```

Throughput is input alleles per second:

```{r duckvep-readme-throughput-table, echo=FALSE}
knitr::kable(surface_rates, row.names = FALSE, align = c("l", "r", "r"))
```

Input staging and model load are outside this resident-engine measurement. Each
pass validates the complete output relation, and a full-row fingerprint checks
one-core/four-core equality for every projection. Commands, CPU affinity and the
complete evidence are in
[benchmarks/duckvep_throughput.md](benchmarks/duckvep_throughput.md).

The end-to-end comparison includes VCF decoding, DuckVEP's coordinate sort,
annotation, and uncompressed tabular output for both DuckVEP and the current
native-compiled [FastVEP](https://github.com/Huang-lab/fastVEP) checkout.

```{r duckvep-readme-fastvep-table, echo=FALSE}
knitr::kable(fast_comparison, row.names = FALSE, align = c("r", "r", "r", "r"))
```

The tools emit different native tabular schemas. The comparison therefore keeps
the row and byte denominators and separately checks each result against VEP.
The illustrated
[*DuckVEP: the fastest Ensembl VEP-compatible consequence predictor in the
West?*](benchmarks/benchmark_duckvep_fastvep.md) report gives the build flags,
source revisions, cache construction, commands, speed, memory, and output
evidence.

#### Current scope and explicit gaps

The released claim is independent-event consequence and HGVS parity for the tested VEP
116 state space. Genotypes and phase sets are preserved by the HTS relations, but combined
phased transcript edit sets, haplotype consequences, and haplotype HGVS remain separate
work. Exact span SVs and paired BNDs are implemented; confidence-interval interpretation,
STR-specific classification, and arbitrary producer-specific symbolic records are not
silently guessed. GRCh37 uses its native Ensembl release-116 transcript model and carries
no invented native MANE designation; any MANE projection back to GRCh37 must be a separate
provenance-labelled mapping.

The resident kernel accepts any transcript catalog that satisfies its validated model
contract. The supplied builder currently compiles the Ensembl core transcript set, not
VEP's `--refseq` or `--merged` sets. It retains GENCODE Basic and Primary flags for SQL
filtering but does not prefilter the resident model to either set. The builder withholds
sequence for source rows marked with transcript-level sequence corrections. A model with
inserted or deleted transcript bases relative to its genomic exons also needs a richer
coordinate map than the current resident interface. Mitochondrial genetic codes and
ordinary MT coordinates are supported; events, transcripts, or features that cross a
circular sequence origin are not.

Supplementary clinical/population scores are deliberately SQL relations rather than
hard-wired C fields. The measured joins establish the storage and execution primitives,
not a universal licensed annotation pack. Broader species/codon-table campaigns, mapped
GRCh37 MANE, combined haplotypes, and the remaining imprecise-SV/STR states stay visible in
the issue tracker and are not included in the 1.5.0 parity claim.

### Interval + reference helpers

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT chrom, start, "end", name, block_count
FROM read_bed('test/data/targets.bed');

SELECT chrom, start, "end", pct_gc, num_a, num_c, num_g, num_t
FROM fasta_nuc('test/data/ce.fa', bed_path := 'test/data/targets.bed')
ORDER BY chrom, start;

SELECT chrom, start, "end", seq_len, pct_gc
FROM fasta_nuc('test/data/ce.fa', bin_width := 10, region := 'CHROMOSOME_I:1-20');
```

### cgranges registry entry points

`duckhts_cgranges_*` exposes a session-scoped immutable interval index for
native overlap queries. You can either build it row-wise with
`duckhts_cgranges_create(...)` + `duckhts_cgranges_add(...)`, or bulk-load it
from SQL with `duckhts_cgranges_from_query(...)`. For row-preserving filters or
count annotations over provider rows, use the vectorized scalar helpers
`duckhts_cgranges_has_overlap(...)` and
`duckhts_cgranges_count_overlaps(...)` directly in queries over `read_bed(...)`,
`read_bam(...)`, `read_bcf(...)`, or regular tables. For streaming
one-row-per-hit expansion while keeping provider columns, use
`duckhts_cgranges_overlaps_list(...)` with `UNNEST(...)`. The older
`duckhts_cgranges_overlaps_bulk(...)` table function still accepts a probe query
and emits matching indexed intervals in one table-function call; that bulk query
runs on the extension-owned helper connection, so use a regular table or view
rather than a temp table.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT duckhts_cgranges_create('readme_idx');
SELECT duckhts_cgranges_add('readme_idx', 'chr1', 10, 20, 'a');
SELECT duckhts_cgranges_add('readme_idx', 'chr1', 30, 40, 'b');
SELECT duckhts_cgranges_index('readme_idx');

SELECT interval_ordinal, label, interval_chrom, interval_start, interval_end
FROM duckhts_cgranges_overlaps('readme_idx', 'chr1', 35, 36, query_row_id := 7);

SELECT duckhts_cgranges_from_query(
  'readme_qry_idx',
  'SELECT * FROM (VALUES (''chr2'', 100, 110, ''alpha''), (''chr2'', 150, 170, ''beta'')) AS t(chrom, start, "end", label)',
  'chrom', 'start', 'end', 'label'
);
SELECT duckhts_cgranges_index('readme_qry_idx');

SELECT interval_ordinal, label, interval_chrom, interval_start, interval_end
FROM duckhts_cgranges_overlaps('readme_qry_idx', 'chr2', 140, 170, mode := 'contain');

CREATE TABLE readme_probes AS
SELECT * FROM (VALUES
  (10, 'chr2', 100, 105),
  (20, 'chr2', 160, 161),
  (30, 'chr2', 500, 510)
) AS t(probe_id, chrom, start, "end");

SELECT
  p.probe_id,
  hit.interval_ordinal,
  hit.label,
  hit.label_type,
  hit.interval_chrom,
  hit.interval_start,
  hit.interval_end
FROM readme_probes AS p
CROSS JOIN UNNEST(
  duckhts_cgranges_overlaps_list('readme_qry_idx', p.chrom, p.start, p."end")
) AS u(hit)
ORDER BY p.probe_id, hit.interval_ordinal;

SELECT query_row_id, interval_ordinal, label, interval_chrom, interval_start, interval_end
FROM duckhts_cgranges_overlaps_bulk(
  'readme_qry_idx',
  'SELECT probe_id, chrom, start, "end" FROM readme_probes',
  'chrom', 'start', 'end',
  query_row_id_col := 'probe_id'
)
ORDER BY query_row_id, interval_ordinal;

SELECT duckhts_cgranges_destroy('readme_idx');
SELECT duckhts_cgranges_destroy('readme_qry_idx');
DROP TABLE readme_probes;
```

### Fixed-bin native counting

`bam_bin_counts()` does fixed-width read-start binning directly in native code.
This is the counting primitive used for WisecondorX-style workflows: duplicate
handling is explicit via `rmdup`, and optional `stats := 'gc,mq'` adds one-pass
GC and MAPQ summaries on the same scan.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT
  bin_id,
  count_total,
  count_fwd,
  count_rev,
  count_pre,
  printf('%.2f', gc_perc_pre) AS gc_pre,
  printf('%.2f', gc_perc_post) AS gc_post,
  printf('%.1f', mean_mapq_post) AS mean_mapq_post
FROM bam_bin_counts(
  'test/data/fixture_mixed.cram',
  5000,
  reference := 'test/data/fixture_ref.fa',
  rmdup := 'streaming',
  stats := 'gc,mq'
)
ORDER BY bin_id;
```

### Mosdepth-compatible coverage outputs

`duckhts_mosdepth()` writes mosdepth-style output files directly from indexed
BAM/CRAM input. The example below writes windowed fragment coverage and then
reads back the generated BED.gz output.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT success, summary_path, regions_path
FROM duckhts_mosdepth(
  '/tmp/duckhts_readme_mosdepth',
  'test/data/range.bam',
  chrom := 'CHROMOSOME_II',
  by := '1000',
  no_per_base := TRUE,
  fragment_mode := TRUE,
  use_median := TRUE,
  overwrite := TRUE
);

SELECT
  column0 AS chrom,
  CAST(column1 AS BIGINT) AS start,
  CAST(column2 AS BIGINT) AS "end",
  CAST(column3 AS DOUBLE) AS depth
FROM read_csv(
  '/tmp/duckhts_readme_mosdepth.regions.bed.gz',
  delim := '\t',
  header := FALSE,
  compression := 'gzip'
)
LIMIT 3;
```

### Polygenic risk scoring

`bcftools_score` computes per-sample polygenic risk scores (PRS) from a VCF/BCF and one or more GWAS summary statistics files, mirroring the `bcftools +score` plugin API.

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Hard-call (GT) PRS — PLINK summary format
-- S1: 0×0.5  + 1×(−0.2) + 2×1.0 = 1.8
-- S2: 1×0.5  + 2×(−0.2) + 0×1.0 = 0.1
SELECT SAMPLE, round(score_summary, 3) AS prs
FROM bcftools_score(
  'test/data/score_input.vcf',
  'test/data/score_summary.tsv',
  use := 'GT',
  columns := 'PLINK'
);

-- Multi-PRS TSV/SSF scoring: multiple summary files in one genotype scan
SELECT SAMPLE,
       round(score_summary, 3) AS prs_a,
       round(score_summary_na, 3) AS prs_b
FROM bcftools_score(
  'test/data/score_input.vcf',
  ['test/data/score_summary.tsv', 'test/data/score_summary_na.tsv'],
  use := 'GT',
  columns := 'PLINK'
);

-- Dosage-based PRS (DS field) — fractional allele dosages from imputed data
-- S1: 0.1×0.5 + 0.8×(−0.2) + 1.8×1.0 = 1.69
-- S2: 1.0×0.5 + 1.9×(−0.2) + 0.2×1.0 = 0.32
SELECT SAMPLE, round(score_summary, 3) AS prs_ds
FROM bcftools_score(
  'test/data/score_dosage.vcf',
  'test/data/score_summary.tsv',
  use := 'DS',
  columns := 'PLINK'
);

-- GWAS-VCF multi-PRS: each FORMAT sample column becomes a separate PRS track
SELECT SAMPLE, round(PRS_A, 3) AS prs_a, round(PRS_B, 3) AS prs_b
FROM bcftools_score(
  'test/data/score_input.vcf',
  'test/data/score_gwas_summary.vcf',
  use := 'GT'
);
```

### Liftover score-style rows

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT src_chrom, src_pos, dest_chrom, dest_pos, dest_ref, dest_alt,
       mapped, reverse_complemented, reject_reason, note
FROM duckdb_liftover(
  '(VALUES
     (''chrF'', 2, ''C'', ''T''),
     (''chrR'', 2, ''A'', ''G''),
     (''chrF'', 11, ''A'', ''T'')
   ) AS t(chrom, pos, ref, alt)',
  'chrom',
  'pos',
  ref_col := 'ref',
  alt_col := 'alt',
  chain_path := 'test/data/liftover.chain',
  dst_fasta_ref := 'test/data/liftover_dst.fa',
  src_fasta_ref := 'test/data/liftover_src.fa'
);
```

### SIMD dispatch flow

DuckHTS uses explicit runtime SIMD dispatch for byte-oriented helper kernels,
starting with `seq_gc_content(...)`. `scalar` is always available and is the
portable baseline. Optional platform backends such as `avx2` or `avx512`
should be checked with `duckhts_simd_backend_available(...)` before being requested.
The `auto` policy resolves each logical kernel independently from the current
compiled-and-CPU-supported capability mask; use `duckhts_simd_kernel_info()` for
the per-kernel result and `SELECT backend FROM duckhts_simd_set_backend('auto')`
to return to runtime auto-detection.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT backend, selectable, compiled, cpu_supported, available, selected
FROM duckhts_simd_info();

SELECT kernel, selected_backend, scalar_fallback
FROM duckhts_simd_kernel_info();

SELECT backend AS selected_backend FROM duckhts_simd_set_backend('scalar');

SELECT
  duckhts_simd_requested_backend() AS requested_backend,
  duckhts_simd_backend() AS selected_backend,
  printf('%.3f', seq_gc_content('ACGTNNacgtnn')) AS gc_content;

SELECT backend IS NOT NULL AS restored_auto FROM duckhts_simd_set_backend('auto');
```

### Sequence utilities

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT
  NAME,
  seq_hash_2bit(substr(SEQUENCE, 1, 12)) AS hash_2bit_prefix,
  seq_encode_4bit(substr(SEQUENCE, 1, 16)) AS codes,
  seq_decode_4bit(seq_encode_4bit(substr(SEQUENCE, 1, 16))) AS roundtrip
FROM read_fasta('test/data/ce.fa')
LIMIT 2;

SELECT
  NAME,
  MATE,
  seq_encode_4bit(substr(SEQUENCE, 1, 12)) AS codes,
  seq_decode_4bit(seq_encode_4bit(substr(SEQUENCE, 1, 12))) AS roundtrip
FROM read_fastq('test/data/interleaved.fq', interleaved := true)
LIMIT 2;
```

### FASTQ quality decoding and fused QC

`read_fastq()` separates input interpretation from output representation:

- `input_quality_encoding` tells DuckHTS how to decode FASTQ ASCII into numeric qualities. The default is modern `phred33`. Use `phred64`, `solexa64`, or `auto` only for legacy files.
- `quality_representation := 'phred'` returns canonical numeric qualities as `UTINYINT[]`.
- `quality_representation := 'string'` returns canonical Phred+33 text. For legacy inputs this means decode first, then re-encode as modern FASTQ text.

This makes the flow explicit:

1. FASTQ text input is decoded according to `input_quality_encoding`.
2. DuckHTS normalizes to numeric Phred values internally.
3. Output is either raw numeric quality arrays (`phred`) or canonical Phred+33 text (`string`).

For BAM/CRAM, qualities are already stored as numeric values, so there is no FASTQ text-encoding ambiguity on input.

Use `duckhts_fastq_qc(...)` for global and per-cycle quality-control reductions. It consumes
the projected sequence and canonical quality strings in one bounded aggregate instead of
creating one SQL row per base. Expand only the compact cycle result when plotting or joining
per-cycle statistics.

```{sql eval=TRUE, collapse=FALSE, comment=""}
WITH q AS (
  SELECT duckhts_fastq_qc(SEQUENCE, QUALITY) AS qc
  FROM read_fastq('test/data/r1.fq')
)
SELECT
  qc.reads,
  qc.bases,
  qc.q30_bases,
  qc.max_read_length
FROM q;

WITH q AS (
  SELECT duckhts_fastq_qc(SEQUENCE, QUALITY) AS qc
  FROM read_fastq('test/data/r1.fq')
)
SELECT cycle.cycle, cycle.bases, cycle.quality_sum
FROM q, UNNEST(qc.cycles) AS u(cycle)
ORDER BY cycle.cycle
LIMIT 5;
```

Use numeric quality arrays when the query genuinely needs the full quality histogram:

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT *
FROM detect_quality_encoding('test/data/legacy_phred64.fq');

WITH q AS (
  SELECT NAME, QUALITY
  FROM read_fastq(
    'test/data/r1.fq',
    quality_representation := 'phred'
  )
),
expanded AS (
  SELECT
    NAME,
    generate_subscripts(QUALITY, 1) AS pos,
    unnest(QUALITY) AS q
  FROM q
)
SELECT pos, q AS phred, count(*) AS n_reads
FROM expanded
GROUP BY pos, phred
ORDER BY pos, phred
LIMIT 12;
```

### Metadata + export/index helpers

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT idx, raw
FROM read_hts_header('test/data/formatcols.vcf.gz', mode := 'raw')
LIMIT 3;

SELECT seqname, tid, index_type, chunk_beg_vo, chunk_end_vo
FROM read_hts_index_spans('test/data/formatcols.vcf.gz')
LIMIT 3;

SELECT index_type, octet_length(raw) AS raw_bytes
FROM read_hts_index_raw('test/data/formatcols.vcf.gz');

COPY (
  SELECT chrom, start, "end", name
  FROM read_bed('test/data/targets.bed')
) TO '/tmp/duckhts_readme_targets.bed' (FORMAT CSV, DELIMITER '\t', HEADER FALSE);

SELECT success, output_path, bytes_out
FROM bgzip('/tmp/duckhts_readme_targets.bed',
           output_path := '/tmp/duckhts_readme_targets.bed.gz',
           keep := TRUE,
           overwrite := TRUE);

SELECT success, output_path, bytes_out
FROM bgunzip('/tmp/duckhts_readme_targets.bed.gz',
             output_path := '/tmp/duckhts_readme_targets.roundtrip.bed',
             keep := TRUE,
             overwrite := TRUE);

SELECT success, index_format, index_path
FROM bam_index('test/data/range.bam',
               index_path := '/tmp/duckhts_readme_range.bam.bai',
               threads := 1);

SELECT success, index_format, index_path
FROM bcf_index('test/data/vcf_file.bcf',
               index_path := '/tmp/duckhts_readme_vcf_file.bcf.csi',
               threads := 1);

SELECT success, index_format, index_path
FROM tabix_index('/tmp/duckhts_readme_targets.bed.gz',
                 preset := 'bed',
                 index_path := '/tmp/duckhts_readme_targets.bed.gz.tbi');
```

### Multi-file queries

`hts_union_query` builds a `UNION ALL BY NAME` across files matching a glob pattern.
Because DuckDB's `query()` cannot accept subquery expressions, use the
`SET VARIABLE` + `getvariable()` pattern:

```{sql eval=TRUE, collapse=FALSE, comment=""}
SET VARIABLE q = hts_union_query('read_fastq', 'test/data/r*.fq');
SELECT filename, count(*) AS n
FROM query(getvariable('q'))
GROUP BY ALL
ORDER BY filename;
```

Per-file parameters can be passed as the third argument (SQL literal):

```{sql eval=TRUE, collapse=FALSE, comment=""}
SET VARIABLE q = hts_union_query('read_bam', 'test/data/range.bam',
                                  'region := ''CHROMOSOME_I:1-1000''');
SELECT count(*) AS n FROM query(getvariable('q'));
```

## Remote URLs and HTS_PATH

Remote URLs (S3/GCS/HTTP/S) can work in two htslib build modes:

1. Dynamic plugin mode (`ENABLE_PLUGINS`): remote handlers are loaded from `HTS_PATH`.
2. Static-handler mode (plugins disabled): handlers are compiled into `libhts` and `HTS_PATH` is not needed.

Use `HTS_PATH` only when you want dynamic plugin discovery (for example, to point at an external htslib plugin directory). `Rduckhts` users only need to call the public helper before the first HTS file is opened:

```r
library(Rduckhts)
setup_hts_env()
```

Example (works in static-handler mode and plugin mode):

```{sql, eval=TRUE, collapse=FALSE, comment=""}
SELECT CHROM, COUNT(*) AS n
FROM read_bcf('s3://1000genomes-dragen-v3.7.6/data/cohorts/gvcf-genotyper-dragen-3.7.6/hg19/3202-samples-cohort/3202_samples_cohort_gg_chr22.vcf.gz',
              region := 'chr22:16050000-16050500')
GROUP BY CHROM;
```

For a direct DuckDB CLI process, set `HTS_PATH` explicitly before its first HTS read, for example:

```bash
export HTS_PATH=$(Rscript --quiet -e 'Rduckhts::setup_hts_env(); cat(Sys.getenv("HTS_PATH"),sep="")')
```

If htslib has already opened a file in the process, restart the process after changing `HTS_PATH`; plugin discovery has already occurred.

If you don’t have htslib plugins installed locally, download the prebuilt binaries from the r-universe-binaries GitHub release and point `HTS_PATH` at the extracted htslib/libexec/htslib directory inside the package bundle.
https://github.com/RGenomicsETL/duckhts/releases/tag/r-universe-binaries

### Browser wasm/webR HTTP backend

For browser wasm/webR builds, DuckHTS does **not** use htslib `libcurl` for remote `http`/`https` access.

- The webR side-module path disables htslib `libcurl`/`S3`/`GCS` features because socket-based libcurl calls from a wasm side module are not reliable in the current webR runtime model.
- DuckHTS registers a package-owned htslib `hFILE` scheme handler implemented in `src/wasm_http_hfile.c` for `http` and `https`.
- This backend uses synchronous `XMLHttpRequest` from the worker for range reads, index probes, and seek behavior.

Browser constraints still apply:

- Remote hosts must allow CORS for the main file **and** index sidecars (`.tbi`/`.csi`), including range requests.
- Behavior can vary by browser and by server-side CORS policy changes over time.
- `ALL_PROXY` / websocket proxy settings do not affect this XHR backend.

Optional header/auth configuration for browser wasm can be provided from JavaScript before loading/querying:

```js
Module.duckhtsWasmHttpConfig = {
  headers: {
    Authorization: "Bearer <short-lived-token>",
    "X-Request-Source": "webr-local"
  },
  allowHosts: ["ftp.ebi.ac.uk", ".s3.amazonaws.com"],
  enforceHostAllowlist: true,
  withCredentials: false,
  allowInsecureAuth: false
};
```

Security behavior of this config:

- Headers are only attached when the URL hostname matches `allowHosts`.
- Requests are blocked for non-matching hosts when `enforceHostAllowlist: true`.
- `Authorization` is blocked on non-HTTPS URLs unless `allowInsecureAuth: true` is set explicitly.
- Credentials/cookies are only sent when `withCredentials: true` is set.

### Browser wasm/duckdb-wasm local setup

DuckHTS is also intended to run as a generic DuckDB community extension in browser wasm hosts (not only webR).

Use the local duckdb-wasm setup to exercise that path end-to-end:

```bash
./scripts/start_duckdb_wasm_local_test.sh
```

This setup uses a Docker-only build path via `scripts/docker/duckdb-wasm-local.Dockerfile`.

The container pre-installs cache-friendly, pinned wasm build dependencies (`emsdk`, `vcpkg`) so repeated local runs do minimal setup work.

Builds run in an isolated mirror worktree (`.duckdb_wasm_docker_work`) and copy back only the wasm extension artifact, so your host native `build/` and `cmake_build/` trees remain available for normal native development and testing.

Then open:

```text
http://127.0.0.1:8001/scripts/duckdb-wasm-local-test.html
```

This setup loads `duckhts.duckdb_extension.wasm` in duckdb-wasm, runs local HTTP
reader checks, and lets you set/clear `Module.duckhtsWasmHttpConfig` directly in
the browser host runtime.

The setup stages `duckdb-browser.mjs`, `duckdb-browser-eh.worker.js`, and `duckdb-eh.wasm` at the site root for same-origin runtime loading, while using an import map for `apache-arrow` resolution.

### S3 credentials and configuration

The htslib S3 plugin supports credentials embedded in the URL or provided via environment variables or standard credentials files. For AWS-style credentials, the most common variables are:

- `AWS_ACCESS_KEY_ID`
- `AWS_SECRET_ACCESS_KEY`
- `AWS_SESSION_TOKEN` (optional, for temporary credentials)
- `AWS_DEFAULT_REGION`
- `AWS_PROFILE` / `AWS_DEFAULT_PROFILE`
- `AWS_SHARED_CREDENTIALS_FILE` (override credentials file location)

You can also configure htslib-specific settings like `HTS_S3_ADDRESS_STYLE`, `HTS_S3_HOST`, and `HTS_S3_S3CFG` for non-default S3 endpoints or path-style access.

See the htslib S3 plugin documentation for full details, URL syntax, and short‑lived credentials support:
https://www.htslib.org/doc/htslib-s3-plugin.html

## Development

This README is rendered with [`duckknit`](https://github.com/rundel/duckknit) to execute SQL snippets in a persistent DuckDB session.

### Environment setup

Run the one-time configure step to create the Python venv and detect platform settings:

```bash
make configure
```

Note: MSVC builds (windows_amd64/windows_arm64) are not supported. Use MinGW/RTools for Windows.

### Prerequisites

Building the extension requires:

- C compiler (GCC or Clang)
- CMake ≥ 3.5
- Make
- Python 3 + venv
- Git
- [htslib](https://github.com/samtools/htslib) build dependencies: zlib, libbz2, liblzma, libdeflate, libcurl, libcrypto (OpenSSL)

Rendering the root documentation additionally requires R with `rmarkdown` and
[`duckknit`](https://github.com/rundel/duckknit). Rebuilding DuckVEP's executable
VEP conformance evidence additionally requires R with `{targets}`, Perl/BioPerl,
micromamba for the pinned oracle environment, and
[`blit`](https://github.com/WangLabCSU/blit) for non-interactive process execution.
Those conformance tools are not runtime or extension-build dependencies.

On Debian/Ubuntu:

```bash
sudo apt install build-essential cmake python3 python3-venv git \
    zlib1g-dev libbz2-dev liblzma-dev libdeflate-dev libcurl4-openssl-dev libssl-dev
```

On macOS:

```bash
brew install cmake htslib xz libdeflate
```

### Vendor [htslib](https://github.com/samtools/htslib)

```bash
./scripts/vendor_htslib.sh
```

This downloads and verifies [htslib](https://github.com/samtools/htslib) 1.24 into `third_party/htslib/`.

### Build

```bash
make configure    # one-time setup (Python venv, platform detection)
make release      # build optimised extension
```

The build runs [htslib](https://github.com/samtools/htslib)'s Makefile (`make lib-static`) in-tree.

The extension binary is written to `build/release/duckhts.duckdb_extension`.

### Debug build

```bash
make debug
```

## Loading

```sql
-- Unsigned extensions must be loaded with -unsigned flag:
-- duckdb -unsigned

LOAD '/path/to/duckhts.duckdb_extension';
```

## Testing

SQL tests live in `test/sql/` using [DuckDB](https://duckdb.org/)'s SQLLogicTest format.

Before running tests for the first time, prepare the indexed test data:

```bash
./test/scripts/prepare_test_data.sh   # requires samtools, bcftools, bgzip, tabix
```

This copies files from the vendored [htslib](https://github.com/samtools/htslib) test suite into `test/data/` and
builds the required indexes (BAI, CSI, TBI, FAI) so region queries work.

Then run:

```bash
make test_release
```

## References

- DuckDB: https://duckdb.org/
- DuckDB Extension API: https://duckdb.org/docs/extensions/overview
- DuckDB extension template (C): https://github.com/duckdb/extension-template-c
- htslib: https://github.com/samtools/htslib
- RBCFTools: https://github.com/RGenomicsETL/RBCFTools
- duckknit: https://github.com/rundel/duckknit

## License

MIT
