---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

knitr::knit_engines$set(sql = function(options) {
  code <- paste(options$code, collapse = "\n")
  load_extension <- is.null(options$load_extension) || isTRUE(options$load_extension)
  if (load_extension) {
    extension_path <- normalizePath("build/release/duckhts.duckdb_extension", mustWork = FALSE)
    extension_path <- gsub("'", "''", extension_path, fixed = TRUE)
    code <- paste(sprintf("LOAD '%s';", extension_path), code, sep = "\n")
  }
  if (isTRUE(options$eval)) {
    out <- system2("duckdb", "-unsigned", input = code, stdout = TRUE, stderr = TRUE)
    status <- attr(out, "status")
    if (!is.null(status) && status != 0) {
      stop(paste(out, collapse = "\n"))
    }
    knitr::engine_output(options, code, out)
  } else {
    knitr::engine_output(options, code, NULL)
  }
})

```
# DuckHTS

A [DuckDB](https://duckdb.org/) extension for reading high-throughput sequencing (HTS) file formats and running interval-oriented genomics workflows directly in SQL.

DuckHTS brings sequencing alignments, variants, reference sequences, annotations, and interval text files into DuckDB through [htslib](https://github.com/samtools/htslib). Query VCF, BCF, BAM, CRAM, FASTA, FASTQ, GTF, GFF, BED, and tabix-indexed files with standard SQL, then compose metadata inspection, compression/indexing, sequence utilities, and interval/reference helpers in the same session.

It covers the sequencing side of the stack: raw/reference data access, annotation parsing, BGZF/tabix utilities, and building blocks for coverage, CNV, QC, and downstream export workflows.

> **Part of a growing genomics stack in DuckDB.**
> DuckHTS covers sequencing and reference-oriented formats. [PlinkingDuck](https://github.com/teaguesterling/plinking_duck) extends the same idea to PLINK genotype workflows and explicitly builds on DuckHTS for the sequencing side of the stack. See also its documentation: https://plinking-duck.readthedocs.io . Together, these tools are pushing DuckDB toward a broader genomics analysis environment built around portable file readers, SQL-native analytics, and composable local workflows. After all, most of Genomics File Formats are Tables/Arrays in disguise.


This extention does not support MSVC builds (windows_amd64/windows_arm64). Use MinGW/RTools for Windows.

## Functions

```{r function-catalog, echo=FALSE, results='asis'}
cat(paste(readLines("r/Rduckhts/inst/function_catalog/functions.md"), collapse = "\n"))
```

`read_fastq` with `mate_path` requires exact QNAME pairing. `read_bam` supports typed `standard_tags` and `auxiliary_tags` maps. `read_tabix` supports header-aware parsing (`header`, `header_names`) and optional type inference (`auto_detect`, `column_types`). Region lists in comma-separated form are supported by `read_bam`, `read_bcf`, `read_fasta`, `read_gff`, `read_gtf`, and `read_tabix`. `read_bam` multi-region queries are deduplicated by htslib, while `read_bcf`/`read_fasta`/`read_gff`/`read_gtf`/`read_tabix` chain regions and can return duplicates for overlaps.

## Examples

The examples below run directly against bundled local test files through the DuckDB CLI and load the built extension from `build/release/duckhts.duckdb_extension`.

### Core readers

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT CHROM, POS, REF, ALT, SAMPLE_ID
FROM read_bcf('test/data/formatcols.vcf.gz', tidy_format := true)
LIMIT 3;

SELECT count(*) AS n
FROM read_bam('test/data/range.bam', region := 'CHROMOSOME_I:1-1000');

SELECT *
FROM fasta_index('test/data/ce.fa');

SELECT NAME, length(SEQUENCE) AS seq_length
FROM read_fasta('test/data/ce.fa', region := 'CHROMOSOME_I:1-25');

SELECT NAME, MATE, PAIR_ID
FROM read_fastq('test/data/interleaved.fq', interleaved := true)
LIMIT 3;
```

### Interval + reference helpers

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT chrom, start, "end", name, block_count
FROM read_bed('test/data/targets.bed');

SELECT chrom, start, "end", pct_gc, num_a, num_c, num_g, num_t
FROM fasta_nuc('test/data/ce.fa', bed_path := 'test/data/targets.bed')
ORDER BY chrom, start;

SELECT chrom, start, "end", seq_len, pct_gc
FROM fasta_nuc('test/data/ce.fa', bin_width := 10, region := 'CHROMOSOME_I:1-20');
```

### Polygenic risk scoring

`bcftools_score` computes per-sample polygenic risk scores (PRS) from a VCF/BCF and a GWAS summary statistics file, mirroring the `bcftools +score` plugin API.

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Hard-call (GT) PRS — PLINK summary format
-- S1: 0×0.5  + 1×(−0.2) + 2×1.0 = 1.8
-- S2: 1×0.5  + 2×(−0.2) + 0×1.0 = 0.1
SELECT SAMPLE, round(score_summary, 3) AS prs
FROM bcftools_score(
  'test/data/score_input.vcf',
  'test/data/score_summary.tsv',
  use := 'GT',
  columns := 'PLINK'
);

-- Dosage-based PRS (DS field) — fractional allele dosages from imputed data
-- S1: 0.1×0.5 + 0.8×(−0.2) + 1.8×1.0 = 1.69
-- S2: 1.0×0.5 + 1.9×(−0.2) + 0.2×1.0 = 0.32
SELECT SAMPLE, round(score_summary, 3) AS prs_ds
FROM bcftools_score(
  'test/data/score_dosage.vcf',
  'test/data/score_summary.tsv',
  use := 'DS',
  columns := 'PLINK'
);

-- GWAS-VCF multi-PRS: each FORMAT sample column becomes a separate PRS track
SELECT SAMPLE, round(PRS_A, 3) AS prs_a, round(PRS_B, 3) AS prs_b
FROM bcftools_score(
  'test/data/score_input.vcf',
  'test/data/score_gwas_summary.vcf',
  use := 'GT'
);
```

### Liftover score-style rows

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT src_chrom, src_pos, dest_chrom, dest_pos, dest_ref, dest_alt,
       mapped, reverse_complemented, reject_reason, note
FROM duckdb_liftover(
  '(VALUES
     (''chrF'', 2, ''C'', ''T''),
     (''chrR'', 2, ''A'', ''G''),
     (''chrF'', 11, ''A'', ''T'')
   ) AS t(chrom, pos, ref, alt)',
  'chrom',
  'pos',
  ref_col := 'ref',
  alt_col := 'alt',
  chain_path := 'test/data/liftover.chain',
  dst_fasta_ref := 'test/data/liftover_dst.fa',
  src_fasta_ref := 'test/data/liftover_src.fa'
);
```

### Sequence utilities

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT
  NAME,
  seq_hash_2bit(substr(SEQUENCE, 1, 12)) AS hash_2bit_prefix,
  seq_encode_4bit(substr(SEQUENCE, 1, 16)) AS codes,
  seq_decode_4bit(seq_encode_4bit(substr(SEQUENCE, 1, 16))) AS roundtrip
FROM read_fasta('test/data/ce.fa')
LIMIT 2;

SELECT
  NAME,
  MATE,
  seq_encode_4bit(substr(SEQUENCE, 1, 12)) AS codes,
  seq_decode_4bit(seq_encode_4bit(substr(SEQUENCE, 1, 12))) AS roundtrip
FROM read_fastq('test/data/interleaved.fq', interleaved := true)
LIMIT 2;
```

### FASTQ quality decoding and per-position histograms

`read_fastq()` separates input interpretation from output representation:

- `input_quality_encoding` tells DuckHTS how to decode FASTQ ASCII into numeric qualities. The default is modern `phred33`. Use `phred64`, `solexa64`, or `auto` only for legacy files.
- `quality_representation := 'phred'` returns canonical numeric qualities as `UTINYINT[]`.
- `quality_representation := 'string'` returns canonical Phred+33 text. For legacy inputs this means decode first, then re-encode as modern FASTQ text.

This makes the flow explicit:

1. FASTQ text input is decoded according to `input_quality_encoding`.
2. DuckHTS normalizes to numeric Phred values internally.
3. Output is either raw numeric quality arrays (`phred`) or canonical Phred+33 text (`string`).

For BAM/CRAM, qualities are already stored as numeric values, so there is no FASTQ text-encoding ambiguity on input.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT *
FROM detect_quality_encoding('test/data/legacy_phred64.fq');

WITH q AS (
  SELECT NAME, QUALITY
  FROM read_fastq(
    'test/data/r1.fq',
    quality_representation := 'phred'
  )
),
expanded AS (
  SELECT
    NAME,
    generate_subscripts(QUALITY, 1) AS pos,
    unnest(QUALITY) AS q
  FROM q
)
SELECT pos, q AS phred, count(*) AS n_reads
FROM expanded
GROUP BY pos, phred
ORDER BY pos, phred
LIMIT 12;
```

### Metadata + export/index helpers

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT idx, raw
FROM read_hts_header('test/data/formatcols.vcf.gz', mode := 'raw')
LIMIT 3;

SELECT seqname, tid, index_type, chunk_beg_vo, chunk_end_vo
FROM read_hts_index_spans('test/data/formatcols.vcf.gz')
LIMIT 3;

SELECT index_type, octet_length(raw) AS raw_bytes
FROM read_hts_index_raw('test/data/formatcols.vcf.gz');

COPY (
  SELECT chrom, start, "end", name
  FROM read_bed('test/data/targets.bed')
) TO '/tmp/duckhts_readme_targets.bed' (FORMAT CSV, DELIMITER '\t', HEADER FALSE);

SELECT success, output_path, bytes_out
FROM bgzip('/tmp/duckhts_readme_targets.bed',
           output_path := '/tmp/duckhts_readme_targets.bed.gz',
           keep := TRUE,
           overwrite := TRUE);

SELECT success, index_format, index_path
FROM tabix_index('/tmp/duckhts_readme_targets.bed.gz',
                 preset := 'bed',
                 index_path := '/tmp/duckhts_readme_targets.bed.gz.tbi');
```

## Remote URLs and HTS_PATH

Remote URLs (S3/GCS/HTTP/S) can work in two htslib build modes:

1. Dynamic plugin mode (`ENABLE_PLUGINS`): remote handlers are loaded from `HTS_PATH`.
2. Static-handler mode (plugins disabled): handlers are compiled into `libhts` and `HTS_PATH` is not needed.

Use `HTS_PATH` only when you want dynamic plugin discovery (for example, to point at an external htslib plugin directory).

Example (works in static-handler mode and plugin mode):

```{sql, eval=TRUE, collapse=FALSE, comment=""}
SELECT CHROM, COUNT(*) AS n
FROM read_bcf('s3://1000genomes-dragen-v3.7.6/data/cohorts/gvcf-genotyper-dragen-3.7.6/hg19/3202-samples-cohort/3202_samples_cohort_gg_chr22.vcf.gz',
              region := 'chr22:16050000-16050500')
GROUP BY CHROM;
```

If you need dynamic plugin mode, set `HTS_PATH` before loading the extension, for example:

```bash
export HTS_PATH=$(Rscript --quiet -e 'cat(Rduckhts:::duckhts_htslib_plugins_dir(),sep="")')
```

If `HTS_PATH` is changed after loading, restart the session and reload the extension.

If you don’t have htslib plugins installed locally, download the prebuilt binaries from the r-universe-binaries GitHub release and point `HTS_PATH` at the extracted htslib/libexec/htslib directory inside the package bundle.
https://github.com/RGenomicsETL/duckhts/releases/tag/r-universe-binaries

### S3 credentials and configuration

The htslib S3 plugin supports credentials embedded in the URL or provided via environment variables or standard credentials files. For AWS-style credentials, the most common variables are:

- `AWS_ACCESS_KEY_ID`
- `AWS_SECRET_ACCESS_KEY`
- `AWS_SESSION_TOKEN` (optional, for temporary credentials)
- `AWS_DEFAULT_REGION`
- `AWS_PROFILE` / `AWS_DEFAULT_PROFILE`
- `AWS_SHARED_CREDENTIALS_FILE` (override credentials file location)

You can also configure htslib-specific settings like `HTS_S3_ADDRESS_STYLE`, `HTS_S3_HOST`, and `HTS_S3_S3CFG` for non-default S3 endpoints or path-style access.

See the htslib S3 plugin documentation for full details, URL syntax, and short‑lived credentials support:
https://www.htslib.org/doc/htslib-s3-plugin.html

## Building

### Environment setup

Run the one-time configure step to create the Python venv and detect platform settings:

```bash
make configure
```

Note: MSVC builds (windows_amd64/windows_arm64) are not supported. Use MinGW/RTools for Windows.

### Prerequisites

- C compiler (GCC or Clang)
- CMake ≥ 3.5
- Make
- Python 3 + venv
- Git
- [htslib](https://github.com/samtools/htslib) build dependencies: zlib, libbz2, liblzma, libdeflate, libcurl, libcrypto (OpenSSL)

On Debian/Ubuntu:

```bash
sudo apt install build-essential cmake python3 python3-venv git \
    zlib1g-dev libbz2-dev liblzma-dev libdeflate-dev libcurl4-openssl-dev libssl-dev
```

On macOS:

```bash
brew install cmake htslib xz libdeflate
```

### Vendor [htslib](https://github.com/samtools/htslib)

```bash
./scripts/vendor_htslib.sh
```

This downloads and verifies [htslib](https://github.com/samtools/htslib) 1.23 into `third_party/htslib/`.

### Build

```bash
make configure    # one-time setup (Python venv, platform detection)
make release      # build optimised extension
```

The build runs [htslib](https://github.com/samtools/htslib)'s Makefile (`make lib-static`) in-tree.

The extension binary is written to `build/release/duckhts.duckdb_extension`.

### Debug build

```bash
make debug
```

## Loading

```sql
-- Unsigned extensions must be loaded with -unsigned flag:
-- duckdb -unsigned

LOAD '/path/to/duckhts.duckdb_extension';
```

## Testing

SQL tests live in `test/sql/` using [DuckDB](https://duckdb.org/)'s SQLLogicTest format.

Before running tests for the first time, prepare the indexed test data:

```bash
./test/scripts/prepare_test_data.sh   # requires samtools, bcftools, bgzip, tabix
```

This copies files from the vendored [htslib](https://github.com/samtools/htslib) test suite into `test/data/` and
builds the required indexes (BAI, CSI, TBI, FAI) so region queries work.

Then run:

```bash
make test_release
```

## R demo

The R package lives under `r/Rduckhts` and provides helpers to load the extension
and create [DuckDB](https://duckdb.org/) tables from HTS files. See its README for R-specific usage:
[r/Rduckhts/README.Rmd](r/Rduckhts/README.Rmd).

```{r eval=TRUE}
library(DBI)
library(duckdb)

drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
con <- dbConnect(drv, dbdir = ":memory:")
ext_path <- normalizePath("build/release/duckhts.duckdb_extension", mustWork = FALSE)
dbExecute(con, sprintf("LOAD '%s'", ext_path))

dbGetQuery(con, "
  SELECT *
  FROM read_bcf('test/data/formatcols.vcf.gz', tidy_format := true)
  LIMIT 5
")

parquet_path <- tempfile(fileext = ".parquet")
dbExecute(con, sprintf(
  "COPY (SELECT * FROM read_bcf('test/data/formatcols.vcf.gz', tidy_format := true)) TO '%s' (FORMAT PARQUET)",
  parquet_path
))
file.exists(parquet_path)

dbGetQuery(con, "
  SELECT NAME, SEQUENCE, QUALITY, MATE, PAIR_ID
  FROM read_fastq('test/data/r1.fq', mate_path := 'test/data/r2.fq')
  LIMIT 5
")

dbGetQuery(con, "
  SELECT QNAME, RNAME, POS, READ_GROUP_ID, SAMPLE_ID
  FROM read_bam('test/data/rg.sam.gz')
  LIMIT 5
")

dbGetQuery(con, "
  SELECT idx, raw
  FROM read_hts_header('test/data/formatcols.vcf.gz', mode := 'raw')
  LIMIT 3
")

dbGetQuery(con, "
  SELECT seqname, tid, index_type, chunk_beg_vo, chunk_end_vo
  FROM read_hts_index_spans('test/data/formatcols.vcf.gz')
  LIMIT 3
")

dbGetQuery(con, "
  SELECT index_type, octet_length(raw) AS raw_bytes
  FROM read_hts_index_raw('test/data/formatcols.vcf.gz')
")
```

### SAMtags + auxiliary tags

Standard SAMtags can be surfaced as typed columns and non-standard tags
captured in a map for ad hoc access:

```{r eval=TRUE}
dbGetQuery(con, "
  SELECT RG, NM, map_extract(AUXILIARY_TAGS, 'XZ') AS XZ
  FROM read_bam('test/data/aux_tags.sam.gz', standard_tags := true, auxiliary_tags := true)
  LIMIT 1
")

dbGetQuery(con, "
  SELECT seqname, feature, start, \"end\", attributes_map
  FROM read_gff('test/data/gff_file.gff.gz', attributes_map := true)
  WHERE feature = 'gene'
  LIMIT 5
")

dbGetQuery(con, "
  SELECT column0, column1
  FROM read_tabix('test/data/meta_tabix.tsv.gz')
  LIMIT 2
")

dbGetQuery(con, "
  SELECT chrom, pos
  FROM read_tabix('test/data/header_tabix.tsv.gz', header := true)
  LIMIT 2
")

dbGetQuery(con, "
  SELECT typeof(column1) AS column1_type
  FROM read_tabix('test/data/meta_tabix.tsv.gz', auto_detect := true)
  LIMIT 1
")

dbGetQuery(con, "
  SELECT pos + 1 AS pos_plus_one
  FROM read_tabix('test/data/header_tabix.tsv.gz', header := true,
                  column_types := ['VARCHAR','BIGINT','VARCHAR'])
  LIMIT 1
")

dbDisconnect(con, shutdown = TRUE)
```

## Project Structure

```
src/
  duckhts.c          # Extension entry point
  bcf_reader.c       # VCF/BCF reader (read_bcf)
  bam_reader.c       # SAM/BAM/CRAM reader (read_bam)
  seq_reader.c       # FASTA/FASTQ reader (read_fasta, read_fastq)
  tabix_reader.c     # Tabix/GTF/GFF reader (read_tabix, read_gtf, read_gff)
  vep_parser.c       # VEP/CSQ annotation parser
  include/
    vcf_types.h
    vep_parser.h
third_party/
  htslib/            # Vendored htslib 1.23 (built automatically)
test/
  sql/               # SQL logic tests
duckdb_capi/
  duckdb.h           # DuckDB C API headers
  duckdb_extension.h
r/
  Rduckhts/          # R package harness
```

## References

- DuckDB: https://duckdb.org/
- DuckDB Extension API: https://duckdb.org/docs/extensions/overview
- DuckDB extension template (C): https://github.com/duckdb/extension-template-c
- htslib: https://github.com/samtools/htslib
- RBCFTools: https://github.com/RGenomicsETL/RBCFTools

## License

MIT
