---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
.readme_tempdir <- tempdir()
.readme_package_path <- find.package("Rduckhts")
.readme_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  x <- gsub(.readme_tempdir, "<tempdir>", x, fixed = TRUE)
  x <- gsub(.readme_package_path, "<Rduckhts>", x, fixed = TRUE)
  x <- gsub("<tempdir>/[^[:space:]]+", "<tempfile>", x)
  x <- gsub("[[:blank:]]+(?=\\n|$)", "", x, perl = TRUE)
  .readme_output_hook(x, options)
})
```

# Rduckhts

[![CRAN Status](https://www.r-pkg.org/badges/version/Rduckhts)](https://cran.r-project.org/package=Rduckhts)
[![R-universe version](https://RGenomicsETL.r-universe.dev/Rduckhts/badges/version)](https://RGenomicsETL.r-universe.dev/Rduckhts)

Query sequencing files from R with SQL. Rduckhts bundles
[DuckHTS](https://github.com/RGenomicsETL/duckhts), a DuckDB extension built on
htslib. It reads the following formats as DuckDB tables:

- VCF/BCF
- BAM/CRAM
- FASTA/FASTQ
- GFF/GTF
- GenBank
- BED
- BigWig
- tabix-indexed text

Indexed files answer region queries without a full scan, and remote files are
read over HTTPS and S3. Results stay in DuckDB until you collect them. htslib
is bundled, so there is nothing else to install.

## Installation

```r
install.packages("Rduckhts")

# Development version
install.packages("Rduckhts", repos = c("https://rgenomicsetl.r-universe.dev",
                                       "https://cloud.r-project.org"))
```

Rduckhts needs the `duckdb` R package, version 1.5.0 or newer.

## Quick start

Every reader has the same shape, `rduckhts_<format>(con, table_name, path, ...)`.
It creates a DuckDB table that you query with SQL:

```{r quick-start}
library(DBI)
library(Rduckhts)

con <- rduckhts_connect()

bcf <- system.file("extdata", "vcf_file.bcf", package = "Rduckhts")
rduckhts_bcf(con, "variants", bcf)

dbGetQuery(con, "
  SELECT CHROM, count(*) AS variants, min(POS) AS first, max(POS) AS last
  FROM variants GROUP BY CHROM ORDER BY CHROM
")
```

## Common tasks

### Read one region of an indexed file

Pass `region` to any indexed reader. Only the matching records are decoded:

```{r region}
rduckhts_bcf(con, "variants_region", bcf, region = "1:3000150-3000151")
dbGetQuery(con, "SELECT CHROM, POS, REF, ALT FROM variants_region")

bam <- system.file("extdata", "range.bam", package = "Rduckhts")
rduckhts_bam(con, "reads", bam, region = "CHROMOSOME_I:1-1000")
dbGetQuery(con, "SELECT QNAME, FLAG, POS, MAPQ, CIGAR FROM reads")
```

### Fetch requested alleles, exactly

`rduckhts_geno_sites()` takes a table of requested alleles and returns each
one's genotypes from a single indexed scan. A requested ALT inside a
multi-allelic record comes back with its index in `ALT`. Requests without an
exact match stay in the result with a `match_status` and no fabricated
genotype.

```{r geno-sites}
geno <- system.file("extdata", "geno_sites.bcf", package = "Rduckhts")
dbWriteTable(con, "requests", temporary = TRUE, data.frame(
  request_id = 1:4, build = "GRCh38", chrom = c("chr1", "chr1", "chr1", "chr3"),
  pos = c(100L, 200L, 300L, 10L), ref = c("A", "T", "G", "C"),
  alt = c("G", "C", "A", "T")
))

fetched <- rduckhts_geno_sites(con, geno, "requests", source_build = "GRCh38",
                               format_fields = "DS")
fetched <- fetched[order(fetched$request_id), ]
rownames(fetched) <- NULL
fetched[, c("request_id", "chrom", "pos", "alt", "alt_index", "match_status")]
```

The test file holds two records at chr1:100, so request 1 matches twice, once
per record. Request 3's REF matches but its ALT is not at that site. Request 4's
position is not in the file.

### Keep annotation attributes as columns

GFF, GTF and GenBank readers can promote chosen attributes to typed columns:

```{r gff}
gff <- system.file("extdata", "gff_attrs.gff3", package = "Rduckhts")
rduckhts_gff(con, "features", gff, attributes = c("ID", "Dbxref"))
dbGetQuery(con, "SELECT seqname, feature, start, \"end\", ID, Dbxref FROM features")
```

### Read remote files

Remote paths work wherever a local path does, once `setup_hts_env()` has
pointed htslib at its bundled network plugins. Call it once per session before
the first remote read. This reads a 100 kb slice of the UCSC GRCh38 phyloP
track and one region of a 1000 Genomes cohort VCF on S3, fetching only the
index and the blocks that the region needs:

```{r remote}
setup_hts_env()

phylop <- "https://hgdownload.soe.ucsc.edu/goldenPath/hg38/phyloP100way/hg38.phyloP100way.bw"
rduckhts_bigwig(con, "conservation", phylop, region = "chr22:20000000-20099999")
dbGetQuery(con, "
  SELECT count(*) AS intervals, round(min(VALUE), 3) AS min, round(max(VALUE), 3) AS max
  FROM conservation
")

cohort <- paste0(
  "s3://1000genomes-dragen-v3.7.6/data/cohorts/gvcf-genotyper-dragen-3.7.6/",
  "hg19/3202-samples-cohort/3202_samples_cohort_gg_chr22.vcf.gz"
)
rduckhts_bcf(con, "cohort", cohort, region = "chr22:16050000-16050500")
dbGetQuery(con, "SELECT count(*) AS variants FROM cohort")
```

S3 access needs a build with htslib's S3 support; Windows (Rtools) builds do
not include it.

### Write Parquet

Converters stream a file into Parquet and keep its header in the Parquet
metadata:

```{r parquet}
parquet <- tempfile(fileext = ".parquet")
rduckhts_bcf_convert_parquet(con, bcf, parquet, overwrite = TRUE)
dbGetQuery(con, sprintf("SELECT count(*) AS rows FROM read_parquet('%s')", parquet))
```

### Work with a database file

`rduckhts_connect(dbdir = "file.duckdb")` works with database files too, even
read-only ones, and never changes the file's catalog.

- **Other connections:** for another DBI or pool connection, call
  `rduckhts_load(con)`, then `rduckhts_install_macros(con)`.
- **SQL functions:** the extension's functions (`read_bcf()`, `read_bam()` and
  the rest) are available directly on the connection.

```{r cleanup, include = FALSE}
unlink(parquet)
dbDisconnect(con, shutdown = TRUE)
```

## Learn more

- **Function reference:** [reference](https://rgenomicsetl.github.io/duckhts/Rduckhts/reference/),
  or `rduckhts_functions()` in R for the SQL catalog.
- **[Worked examples](https://rgenomicsetl.github.io/duckhts/Rduckhts/articles/workflows.html):**
  - variant normalization and liftover;
  - polygenic scores;
  - Somalier-style relatedness;
  - coverage and bin counts;
  - FASTQ quality control;
  - interval indexes;
  - compression and indexing;
  - multi-file reading.
- **[Browser and webR](https://rgenomicsetl.github.io/duckhts/Rduckhts/articles/browser.html):**
  remote access and request settings in wasm builds.
- **SQL, the DuckDB command line and Python:** the
  [DuckHTS README](https://github.com/RGenomicsETL/duckhts).
- **Consequence annotation:** the
  [DuckVEP](https://github.com/RGenomicsETL/DuckVEP) extension and its
  Rduckvep package.
- **Building another package against the bundled htslib:** `rduckhts_htslib_config()`.

## License

GPL (>= 2). The bundled DuckHTS extension is MIT-licensed, and other bundled
third-party code keeps its own licence; see `inst/COPYRIGHT`.

## Credits

The GenBank reader and FASTA converter were contributed by
[Ryan Ward](https://github.com/ryandward) of [Nurture Bio](https://github.com/Nurture-Bio).
Thanks to all [contributors](https://github.com/RGenomicsETL/duckhts/graphs/contributors),
and to the htslib, bcftools, DuckDB and RBCFTools projects that DuckHTS builds on.
