---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Rduckhts: DuckDB HTS File Reader Extension for R

[![CRAN Status](https://www.r-pkg.org/badges/version/Rduckhts)](https://cran.r-project.org/package=Rduckhts)[![R-universe version](https://RGenomicsETL.r-universe.dev/Rduckhts/badges/version)](https://RGenomicsETL.r-universe.dev/Rduckhts)

`Rduckhts` provides an R interface to a [DuckDB](https://duckdb.org/) `HTS` (High Throughput Sequencing) file reader extension. This enables reading common bioinformatics file formats such as `VCF`/`BCF`, `SAM`/`BAM`/`CRAM`, `FASTA`, `FASTQ`, `GFF`, `GTF`, and tabix-indexed files directly from `R` using `SQL` queries via  [`duckhts`](https://github.com/RGenomicsETL/duckhts).

## How it works

Following [RBCFTools](https://github.com/RGenomicsETL/RBCFTools), tables are created and returned instead of data frames. `VCF`/`BCF`, `SAM`/`BAM`/`CRAM`, `FASTA`, `FASTQ`, `GFF`, `GTF`, and `tabix` formats can be queried. We support region queries for indexed files, and we target Linux, macOS, and RTools. [`htslib`](https://github.com/samtools/htslib) 1.23 is bundled so build dependencies stay minimal. The extensnion is built by adapting the generic extension infracstructure by using only makefiles unlike  the submitted communtity extension [`duckhts`](https://github.com/RGenomicsETL/duckhts).

## Installation

The package can be installed from r-universe

```r
# Install 'Rduckhts' in R:
install.packages('Rduckhts', repos = c('https://rgenomicsetl.r-universe.dev', 'https://cloud.r-project.org'))
# When on CRAN
install.packages("Rduckhts")
```

## System Requirements

Installation requires `htslib` dependencies such ad zlib and libbz2, and  optionally for full functionally liblzma, libcurl, and openssl. The package requires GNU make. On Windows's Rtools, `htslib` plugins are not enabled. 

## Browser wasm/webR networking and setup

In browser wasm/webR builds, remote `http`/`https` access does not use htslib `libcurl`.

- The bundled extension enables a custom htslib `hFILE` backend in `src/wasm_http_hfile.c` for `http` and `https`.
- This backend uses synchronous worker-side XHR for range reads and tabix index access.
- htslib `libcurl`/`S3`/`GCS` paths are intentionally disabled for wasm in the package build.

What this means in practice:

- Same-origin URLs are the simplest setup and work well for local browser testing.
- Remote URLs work only when browser CORS policy allows them.
- CORS must allow both the primary file and index sidecars (`.tbi`/`.csi`), including range requests.
- htslib may probe a `.csi` sidecar before falling back to `.tbi`; a `.csi` `404` is not a regression if the `.tbi` path succeeds.
- `ALL_PROXY` and websocket proxy settings do not affect this wasm XHR backend.
- The simple local `python3 -m http.server` setup ignores HTTP `Range`; in that case the backend warns and falls back to full-object fetch + local byte slicing. This is acceptable for small smoke tests, but production servers should support `Range`.

### `Module.duckhtsWasmHttpConfig`

The browser HTTP backend reads optional request/auth settings from `Module.duckhtsWasmHttpConfig`.

- In plain browser JavaScript, set it in the page/worker before running queries.
- In webR, consumers can set it from R with `webr::eval_js()`; they do not need to hand-edit the host HTML/JS as long as they can run that call before the relevant HTTP reads.

Plain JavaScript example:

```js
Module.duckhtsWasmHttpConfig = {
  headers: {
    Authorization: "Bearer <short-lived-token>",
    "X-Request-Source": "webr-local"
  },
  allowHosts: ["ftp.ebi.ac.uk", ".ebi.ac.uk"],
  enforceHostAllowlist: true,
  withCredentials: false,
  allowInsecureAuth: false
};
```

Inside webR, set the same config through `webr::eval_js()` because the code runs inside the webR worker:

```r
webr::eval_js("
  Module.duckhtsWasmHttpConfig = {
    headers: {
      Authorization: 'Bearer <short-lived-token>',
      'X-Request-Source': 'webr-local'
    },
    allowHosts: ['ftp.ebi.ac.uk', '.ebi.ac.uk'],
    enforceHostAllowlist: true,
    withCredentials: false,
    allowInsecureAuth: false
  };
")
```

Configuration fields:

- `headers`: named request headers to attach to matching hosts.
- `allowHosts`: hostname allowlist for header injection. Entries can be exact hosts (`"ftp.ebi.ac.uk"`) or suffix matches with a leading dot (`".ebi.ac.uk"`).
- `enforceHostAllowlist`: when `true`, block requests to hosts outside `allowHosts` instead of merely omitting configured headers.
- `withCredentials`: when `true`, send cookies/credentials with XHR requests.
- `allowInsecureAuth`: when `true`, allow `Authorization` headers on non-HTTPS URLs. The default is `false`.

Security behavior of this config:

- Headers are only attached when URL hostnames match `allowHosts`.
- `Authorization` is blocked for non-HTTPS URLs unless `allowInsecureAuth: true` is set.
- Cookies/credentials are only sent when `withCredentials: true` is set.
- The config can be updated or cleared between queries if different hosts need different policies.


## Quick Start 

The extension is loaded with `rduckhts_load(con, extension_path = NULL)`. The wrappers break down into:

- readers: `rduckhts_bcf()`, `rduckhts_bam()`, `rduckhts_fasta()`, `rduckhts_fastq()`, `rduckhts_gff()`, `rduckhts_gtf()`, `rduckhts_tabix()`, `rduckhts_bed()`
- multi-file readers: `rduckhts_bam_multi()`, `rduckhts_bcf_multi()`, `rduckhts_fastq_multi()`, `rduckhts_fasta_multi()`, `rduckhts_gff_multi()`, `rduckhts_gtf_multi()`, `rduckhts_tabix_multi()`, `rduckhts_bed_multi()`
- reference helpers: `rduckhts_fasta_index()`, `rduckhts_fasta_nuc()`
- compression/indexing: `rduckhts_bgzip()`, `rduckhts_bgunzip()`, `rduckhts_bam_index()`, `rduckhts_bcf_index()`, `rduckhts_tabix_index()`
- metadata helpers: `rduckhts_hts_header()`, `rduckhts_hts_index()`, `rduckhts_hts_index_spans()`, `rduckhts_hts_index_raw()`

Start with one reader, then materialize tables and compose the richer helpers around them.

```{r}
library(DBI)
library(duckdb)
library(Rduckhts)


fasta_path <- system.file("extdata", "ce.fa", package = "Rduckhts")
fastq_r1 <- system.file("extdata", "r1.fq", package = "Rduckhts")
fastq_r2 <- system.file("extdata", "r2.fq", package = "Rduckhts")
con <- dbConnect(duckdb::duckdb(config = list(allow_unsigned_extensions = "true")))
rduckhts_load(con)

rduckhts_fasta(con, "sequences", fasta_path, overwrite = TRUE)
rduckhts_fastq(con, "reads", fastq_r1, mate_path = fastq_r2, overwrite = TRUE)

dbGetQuery(con, "SELECT COUNT(*) AS n FROM sequences")
dbGetQuery(con, "SELECT COUNT(*) AS n FROM reads")

```

## Multi-file Reading

The `rduckhts_*_multi` family reads multiple files into a single DuckDB table
with a `filename` column, following the same `(con, table_name, ...)` convention
as the single-file wrappers:

```{r}
fq_files <- c(
  system.file("extdata", "r1.fq", package = "Rduckhts"),
  system.file("extdata", "r2.fq", package = "Rduckhts")
)
rduckhts_fastq_multi(con, "fq_multi", fq_files, overwrite = TRUE)
dbGetQuery(con, "SELECT filename, count(*) AS n FROM fq_multi GROUP BY ALL")
```

Per-file parameters are supported via a `.params` data.frame with a `file` column
and columns matching reader arguments. `NA` values fall back to the uniform
default:

```{r}
bam_path <- system.file("extdata", "range.bam", package = "Rduckhts")
bam_idx  <- system.file("extdata", "range.bam.bai", package = "Rduckhts")

params <- data.frame(
  file       = bam_path,
  region     = "CHROMOSOME_I:1-1000",
  index_path = bam_idx
)
rduckhts_bam_multi(con, "bam_multi", bam_path, .params = params,
                   overwrite = TRUE)
dbGetQuery(con, "SELECT count(*) AS n FROM bam_multi")
```

## Function Catalog

Use `rduckhts_functions()` inside R to inspect the generated extension catalog.

```{r function-catalog, echo=FALSE, results='asis'}
cat(paste(readLines("inst/function_catalog/functions.md"), collapse = "\n"))
```

## Common Workflows

### Region-aware variant and alignment queries

```{r}
bcf_path <- system.file("extdata", "vcf_file.bcf", package = "Rduckhts")
bcf_index_path <- system.file("extdata", "vcf_file.bcf.csi", package = "Rduckhts")
bam_path <- system.file("extdata", "range.bam", package = "Rduckhts")
bam_index_path <- system.file("extdata", "range.bam.bai", package = "Rduckhts")

rduckhts_bcf(
  con, "variants_idx", bcf_path,
  region = "1:3000150-3000151",
  index_path = bcf_index_path,
  overwrite = TRUE
)
dbGetQuery(con, "SELECT CHROM, POS, REF, ALT FROM variants_idx")

rduckhts_bam(
  con, "bam_idx_reads", bam_path,
  region = "CHROMOSOME_I:1-1000",
  index_path = bam_index_path,
  overwrite = TRUE
)
dbGetQuery(con, "SELECT QNAME, FLAG, POS, MAPQ FROM bam_idx_reads")
```

### Interval + reference helpers

```{r}
bed_path <- system.file("extdata", "targets.bed", package = "Rduckhts")
fai_path <- tempfile("duckhts_readme_", fileext = ".fai")
rduckhts_fasta_index(con, fasta_path, index_path = fai_path)

rduckhts_bed(con, "targets", bed_path, overwrite = TRUE)
dbGetQuery(con, "SELECT chrom, start, \"end\", name, block_count FROM targets")

rduckhts_fasta_nuc(con, fasta_path, bed_path = bed_path, index_path = fai_path)
rduckhts_fasta_nuc(con, fasta_path, bin_width = 10, region = "CHROMOSOME_I:1-20", index_path = fai_path)
unlink(fai_path)
```

### Liftover score-style rows

```{r}
lift_src <- tempfile("duckhts_liftover_src_", fileext = ".fa")
lift_dst <- tempfile("duckhts_liftover_dst_", fileext = ".fa")
lift_chain <- tempfile("duckhts_liftover_", fileext = ".chain")

writeLines(c(
  ">chrF",
  "ACGTACGTAA",
  ">chrR",
  "AACCGGTTAA"
), lift_src)
writeLines(c(
  ">chrLiftF",
  "ACGTACGTAA",
  ">chrLiftR",
  "TTAACCGGTT"
), lift_dst)
writeLines(c(
  "chain 100 chrF 10 + 0 10 chrLiftF 10 + 0 10 1",
  "10",
  "",
  "chain 100 chrR 10 + 0 10 chrLiftR 10 - 0 10 2",
  "10"
), lift_chain)

rduckhts_fasta_index(con, lift_src, index_path = paste0(lift_src, ".fai"))
rduckhts_fasta_index(con, lift_dst, index_path = paste0(lift_dst, ".fai"))

lifted <- rduckhts_liftover(
  con,
  query = paste(
    "SELECT * FROM (VALUES",
    "('chrF', 2, 'C', 'T'),",
    "('chrR', 2, 'A', 'G'),",
    "('chrF', 11, 'A', 'T')",
    ") AS t(chrom, pos, ref, alt)"
  ),
  chain_path = lift_chain,
  dst_fasta_ref = lift_dst,
  ref_col = "ref",
  alt_col = "alt",
  src_fasta_ref = lift_src
)

lifted[, c(
  "src_chrom", "src_pos", "dest_chrom", "dest_pos",
  "dest_ref", "dest_alt", "mapped", "reverse_complemented",
  "reject_reason", "note"
)]

unlink(c(lift_src, paste0(lift_src, ".fai"), lift_dst, paste0(lift_dst, ".fai"), lift_chain))
```

### Munge score-style rows

```{r}
munge_fasta <- tempfile("duckhts_munge_", fileext = ".fa")
writeLines(c(
  ">chrF",
  "ACGTACGTAA"
), munge_fasta)
rduckhts_fasta_index(con, munge_fasta, index_path = paste0(munge_fasta, ".fai"))

munge_out <- rduckhts_munge(
  con,
  query = paste(
    "SELECT * FROM (VALUES",
    "('rs1', 2, 'chrF', 'A', 'C', 0.01, 1.10, 0.20, 0.98, 0.10, 0.01, 1000),",
    "('rs2', 2, 'chrF', 'C', 'A', 0.02, 0.90, -0.20, 0.98, 0.90, 0.01, 1000)",
    ") AS t(SNP, BP, CHR, A1, A2, P, OR_VALUE, BETA, INFO, FRQ, SE, N)"
  ),
  fasta_ref = munge_fasta,
  column_map = c(
    SNP = "SNP", BP = "BP", CHR = "CHR", A1 = "A1", A2 = "A2",
    P = "P", OR = "OR_VALUE", BETA = "BETA", INFO = "INFO", FRQ = "FRQ", SE = "SE", N = "N"
  )
)

munge_out[, c("chrom", "pos", "id", "ref", "alt", "alleles_swapped", "filter", "af", "es", "ns")]

unlink(c(munge_fasta, paste0(munge_fasta, ".fai")))
```

### Polygenic risk scoring

`rduckhts_score()` computes per-sample polygenic risk scores (PRS) from a genotype VCF/BCF and a GWAS summary statistics file, wrapping `bcftools_score`.

```{r}
vcf_path    <- system.file("extdata", "score_input.vcf",        package = "Rduckhts")
dosage_path <- system.file("extdata", "score_dosage.vcf",       package = "Rduckhts")
sumf_path   <- system.file("extdata", "score_summary.tsv",      package = "Rduckhts")
gwas_path   <- system.file("extdata", "score_gwas_summary.vcf", package = "Rduckhts")

# Hard-call (GT) PRS with PLINK-format summary statistics
# S1: 0×0.5 + 1×(−0.2) + 2×1.0 = 1.8
# S2: 1×0.5 + 2×(−0.2) + 0×1.0 = 0.1
gt_prs <- rduckhts_score(con, vcf_path, sumf_path, use = "GT", columns = "PLINK")
gt_prs[, c("SAMPLE", "score_summary")]

# Dosage-based PRS (DS field) for imputed genotypes
# S1: 0.1×0.5 + 0.8×(−0.2) + 1.8×1.0 = 1.69
# S2: 1.0×0.5 + 1.9×(−0.2) + 0.2×1.0 = 0.32
ds_prs <- rduckhts_score(con, dosage_path, sumf_path, use = "DS", columns = "PLINK")
ds_prs[, c("SAMPLE", "score_summary")]

# GWAS-VCF multi-PRS: each FORMAT/ES sample column becomes a separate PRS track
gwas_prs <- rduckhts_score(con, vcf_path, gwas_path, use = "GT")
gwas_prs[, c("SAMPLE", "PRS_A", "PRS_B")]
```

### Compression + tabix round-trips

```{r}
bed_src <- system.file("extdata", "targets.bed", package = "Rduckhts")
bam_src <- system.file("extdata", "range.bam", package = "Rduckhts")
bcf_src <- system.file("extdata", "vcf_file.bcf", package = "Rduckhts")

tmp_bed <- tempfile("duckhts_targets_", fileext = ".bed")
tmp_bgz <- paste0(tmp_bed, ".gz")
tmp_tbi <- paste0(tmp_bgz, ".tbi")
tmp_roundtrip <- tempfile("duckhts_targets_roundtrip_", fileext = ".bed")
tmp_bai <- tempfile("duckhts_range_", fileext = ".bam.bai")
tmp_csi <- tempfile("duckhts_variants_", fileext = ".bcf.csi")
file.copy(bed_src, tmp_bed, overwrite = TRUE)

bgzip_meta <- rduckhts_bgzip(
  con, tmp_bed,
  output_path = tmp_bgz,
  threads = 1,
  keep = TRUE,
  overwrite = TRUE
)
bgzip_meta[, c("success", "output_path", "bytes_out")]

bgunzip_meta <- rduckhts_bgunzip(
  con, tmp_bgz,
  output_path = tmp_roundtrip,
  threads = 1,
  keep = TRUE,
  overwrite = TRUE
)
bgunzip_meta[, c("success", "output_path", "bytes_out")]

bam_index_meta <- rduckhts_bam_index(
  con, bam_src,
  index_path = tmp_bai,
  threads = 1
)
bam_index_meta

bcf_index_meta <- rduckhts_bcf_index(
  con, bcf_src,
  index_path = tmp_csi,
  threads = 1
)
bcf_index_meta

tabix_meta <- rduckhts_tabix_index(
  con, tmp_bgz,
  preset = "bed",
  index_path = tmp_tbi,
  threads = 1
)
tabix_meta

rduckhts_bed(con, "targets_idx", tmp_bgz, region = "CHROMOSOME_I:1-20", index_path = tmp_tbi, overwrite = TRUE)
dbGetQuery(con, "SELECT * FROM targets_idx")

unlink(c(tmp_bed, tmp_bgz, tmp_tbi, tmp_roundtrip, tmp_bai, tmp_csi))
```

## Sequence UDFs

The extension also exposes sequence utility UDFs directly in DuckDB SQL, including 4-bit IUPAC DNA encode/decode helpers. These can be applied to `SEQUENCE` columns from FASTA and FASTQ scans.

```{r}
dbGetQuery(
  con,
  "SELECT
     NAME,
     seq_hash_2bit(substr(SEQUENCE, 1, 12)) AS hash_2bit_prefix,
     seq_encode_4bit(substr(SEQUENCE, 1, 16)) AS codes,
     seq_decode_4bit(seq_encode_4bit(substr(SEQUENCE, 1, 16))) AS roundtrip
   FROM sequences
   LIMIT 2"
)

dbGetQuery(
  con,
  "SELECT
     NAME,
     MATE,
     seq_encode_4bit(substr(SEQUENCE, 1, 12)) AS codes,
     seq_decode_4bit(seq_encode_4bit(substr(SEQUENCE, 1, 12))) AS roundtrip
   FROM reads
   LIMIT 2"
)
```

### FASTA region queries

`read_fasta` supports indexed region queries via `rduckhts_fasta(..., region = ...)`.

```{r}
fai_path <- tempfile("duckhts_readme_", fileext = ".fai")
fai_info <- rduckhts_fasta_index(con, fasta_path, index_path = fai_path)
fai_info

rduckhts_fasta(
  con, "fasta_region", fasta_path,
  region = "CHROMOSOME_I:1-25",
  overwrite = TRUE
)
dbGetQuery(con, "SELECT NAME, length(SEQUENCE) AS n FROM fasta_region")
unlink(fai_path)
```



##  Examples

### Region Queries

Region queries can use implicit sidecar indexes or an explicit `index_path`
for custom index names/locations.

```{r}
bcf_path <- system.file("extdata", "vcf_file.bcf", package = "Rduckhts")
bcf_index_path <- system.file("extdata", "vcf_file.bcf.csi", package = "Rduckhts")
rduckhts_bcf(con, "variants", bcf_path, overwrite = TRUE)
variants <- dbGetQuery(con, "SELECT * FROM variants LIMIT 5")
variants

rduckhts_bcf(
  con, "variants_idx", bcf_path,
  region = "1:3000150-3000151",
  index_path = bcf_index_path,
  overwrite = TRUE
)
dbGetQuery(con, "SELECT count(*) AS n FROM variants_idx")

# Span-oriented index view from the same file
index_spans_preview <- rduckhts_hts_index_spans(con, bcf_path, index_path = bcf_index_path)
head(index_spans_preview[, c("seqname", "tid", "index_type", "chunk_beg_vo", "chunk_end_vo")], 5)
```

### Remote VCF on S3

S3 files can be query when `htslib` is built with plugins enable. This is not the case on RTools

```{r, eval=TRUE}
# Enable htslib plugins for remote access (S3/GCS/HTTP)
setup_hts_env()

# Example S3 URL (1000 Genomes cohort VCF)
s3_base <- "s3://1000genomes-dragen-v3.7.6/data/cohorts/"
s3_path <- "gvcf-genotyper-dragen-3.7.6/hg19/3202-samples-cohort/"
s3_vcf_file <- "3202_samples_cohort_gg_chr22.vcf.gz"
s3_vcf_uri <- paste0(s3_base, s3_path, s3_vcf_file)

rduckhts_bcf(con, "s3_variants", s3_vcf_uri, region = "chr22:16050000-16050500", overwrite = TRUE)
dbGetQuery(con, "SELECT CHROM, COUNT(*) AS n FROM s3_variants GROUP BY CHROM")
```

### FASTQ files

Three modes for fastq files, single, paired and interleaved

```{r}
r1 <- system.file("extdata", "r1.fq", package = "Rduckhts")
r2 <- system.file("extdata", "r2.fq", package = "Rduckhts")
interleaved <- system.file("extdata", "interleaved.fq", package = "Rduckhts")
rduckhts_fastq(con, "paired_reads", r1, mate_path = r2, overwrite = TRUE)
rduckhts_fastq(con, "interleaved_reads", interleaved, interleaved = TRUE, overwrite = TRUE)
pairs <- dbGetQuery(con, "SELECT * FROM paired_reads WHERE MATE = 1 LIMIT 5")
pairs
```

### FASTQ quality decoding and per-position histograms

FASTQ quality handling has two separate knobs:

- `input_quality_encoding` controls how incoming FASTQ ASCII is decoded. The default is modern `phred33`; use `phred64`, `solexa64`, or `auto` only for legacy data.
- `quality_representation` controls how qualities are returned to DuckDB: canonical Phred+33 text (`"string"`) or numeric Phred arrays (`"phred"`).

The flow is:

1. Decode FASTQ ASCII using `input_quality_encoding`.
2. Normalize to numeric Phred qualities.
3. Return either numeric arrays or canonical Phred+33 text.

`BAM`/`CRAM` reads skip the text-decoding step because qualities are already stored numerically.

```{r}
legacy_fastq <- system.file("extdata", "legacy_phred64.fq", package = "Rduckhts")

rduckhts_detect_quality_encoding(con, legacy_fastq)

quality_hist <- dbGetQuery(
  con,
  sprintf(
    "WITH q AS (
       SELECT NAME, QUALITY
       FROM read_fastq('%s', quality_representation := 'phred')
     ),
     expanded AS (
       SELECT
         NAME,
         generate_subscripts(QUALITY, 1) AS pos,
         unnest(QUALITY) AS q
       FROM q
     )
     SELECT pos, q AS phred, count(*) AS n_reads
     FROM expanded
     GROUP BY pos, phred
     ORDER BY pos, phred
     LIMIT 12",
    fastq_r1
  )
)
quality_hist
```

### GFF files

These can be open with or with attributes maps

```{r}
gff_path <- system.file("extdata", "gff_file.gff.gz", package = "Rduckhts")
rduckhts_gff(con, "genes", gff_path, attributes_map = TRUE, overwrite = TRUE)
gene_annotations <- dbGetQuery(con, "SELECT seqname, start, \"end\" FROM genes WHERE feature = 'gene' LIMIT 5")
gene_annotations
```

### BAM/CRAM

When built with htslib codec, `CRAM` can be opened in addition to `BAM` files.
`index_path` can also be passed for region scans with non-standard index names.

```{r}
cram_path <- system.file("extdata", "range.cram", package = "Rduckhts")
ref_path <- system.file("extdata", "ce.fa", package = "Rduckhts")
bam_path <- system.file("extdata", "range.bam", package = "Rduckhts")
bam_index_path <- system.file("extdata", "range.bam.bai", package = "Rduckhts")

rduckhts_bam(con, "cram_reads", cram_path, reference = ref_path, overwrite = TRUE)
cram_reads <- dbGetQuery(con, "SELECT QNAME, FLAG, POS, MAPQ FROM cram_reads LIMIT 5")
cram_reads

rduckhts_bam(
  con, "bam_idx_reads", bam_path,
  region = "CHROMOSOME_I:1-1000",
  index_path = bam_index_path,
  overwrite = TRUE
)
dbGetQuery(con, "SELECT count(*) AS n FROM bam_idx_reads")
```

### SAMtags + auxiliary tags

Standard SAMtags can be exposed as typed columns, and any remaining tags are
available via `AUXILIARY_TAGS`:

```{r}
aux_path <- system.file("extdata", "aux_tags.sam.gz", package = "Rduckhts")
rduckhts_bam(con, "aux_reads", aux_path, standard_tags = TRUE, auxiliary_tags = TRUE, overwrite = TRUE)
dbGetQuery(con, "SELECT RG, NM, map_extract(AUXILIARY_TAGS, 'XZ') AS XZ FROM aux_reads LIMIT 1")
```

### Tabix headers + types

Use `header = TRUE` to use the first non-meta row as column names, and
`auto_detect = TRUE` / `column_types` to control column typing:

```{r}
tabix_header <- system.file("extdata", "header_tabix.tsv.gz", package = "Rduckhts")
tabix_meta <- system.file("extdata", "meta_tabix.tsv.gz", package = "Rduckhts")

rduckhts_tabix(con, "header_tabix", tabix_header, header = TRUE, overwrite = TRUE)
dbGetQuery(con, "SELECT chrom, pos FROM header_tabix LIMIT 2")

rduckhts_tabix(con, "typed_tabix", tabix_meta, auto_detect = TRUE, overwrite = TRUE)
dbGetQuery(con, "SELECT typeof(column1) AS column1_type FROM typed_tabix LIMIT 1")

rduckhts_tabix(con, "typed_tabix_explicit", tabix_header,
               header = TRUE,
               column_types = c("VARCHAR", "BIGINT", "VARCHAR"),
               overwrite = TRUE)
dbGetQuery(con, "SELECT pos + 1 AS pos_plus_one FROM typed_tabix_explicit LIMIT 1")
```

### HTS header and index metadata

Use metadata helpers to inspect parsed headers, raw header lines, index summaries,
span-oriented index views, and raw index blobs.

```{r}
header_meta <- rduckhts_hts_header(con, bcf_path)
head(header_meta[, c("record_type", "id", "number", "value_type")], 5)

header_raw <- rduckhts_hts_header(con, bcf_path, mode = "raw")
head(header_raw[, c("idx", "raw")], 5)

index_meta <- rduckhts_hts_index(con, bcf_path, index_path = bcf_index_path)
head(index_meta[, c("seqname", "mapped", "unmapped", "index_type")], 5)

index_spans <- rduckhts_hts_index_spans(con, bcf_path, index_path = bcf_index_path)
head(index_spans[, c("seqname", "tid", "index_type", "chunk_beg_vo", "chunk_end_vo")], 5)

index_raw <- rduckhts_hts_index_raw(con, bcf_path, index_path = bcf_index_path)
head(index_raw, 1)
```

### Remote GTEx tabix example

GTEx eQTL matrices on EBI are tabix-indexed. In browser wasm/webR, this still depends on CORS policy on both the data object and index object.

```{r, eval=TRUE}
gtex_url <- "https://ftp.ebi.ac.uk/pub/databases/spot/eQTL/imported/GTEx_V8/ge/Brain_Cerebellar_Hemisphere.tsv.gz" 
rduckhts_tabix(con, "gtex_eqtl", gtex_url, region = "1:11868-14409",
                  header = TRUE, auto_detect = TRUE, overwrite = TRUE)
dbGetQuery(con, "SELECT * FROM gtex_eqtl LIMIT 5")

```

```{r}
dbDisconnect(con, shutdown = TRUE)
```

## References

- DuckDB: https://duckdb.org/
- DuckDB Extension API: https://duckdb.org/docs/extensions/overview
- DuckDB extension template (C): https://github.com/duckdb/extension-template-c
- htslib: https://github.com/samtools/htslib
- RBCFTools: https://github.com/RGenomicsETL/RBCFTools

## License

GPL-3.
