---
title: "Automatic BAM full-file scan: small-fixture cost"
output:
  github_document:
    html_preview: false
---

This compares automatic BAM scans at two named reader revisions,
using the shared `benchmark_simd_bam_gc.R` driver on its committed nanopore
fixture, with and without projecting `FILE_OFFSET` via an `IS NOT NULL` filter.
It also times full materialization with standard/AUX tags and text or numeric
CIGAR/SEQ/QUAL, covering formatter scratch and DuckDB list writes.
It exercises many empty contig claims and full `SEQ` decoding, but this
186-record fixture is dominated by setup cost: it is not a WGS throughput benchmark.
There are no unplaced reads in this performance input, so both builds do equal work.
The separate SQL/R regression corpus tests retained no-coordinate records,
physical duplicates, BAI/CSI, CRAM, and output chunks.

The nearest identical R/DBI workload is this report's
[previous rendering](https://github.com/RGenomicsETL/duckhts/blob/c9cfce7f69aeef97a492578f17a905900ac542ee/benchmarks/benchmark_bam_tail_scan.md):
source `a1509946e9f7d425ce9af755b3dad4d41f03dba3`, 186 records / 249,110 bases,
1 or 4 DuckDB workers and 2 HTSlib workers per handle. Its candidate medians for
full text materialization were 3.431 ms and 3.508 ms respectively; numeric
materialization, 3.886 ms and 4.000 ms. The comparison below retains the same
pre-fix baseline binary at `e87b9f3ee5e27068964e4e57e107bacd7b5def3e` and measures
both binaries with the same R driver after correcting AUX scratch amplification.
Older Python-host SEQ/GC timings are not directly comparable to this R/DBI host.

The older larger-input report is
[`benchmark_simd_bam_gc.md` at the baseline revision](https://github.com/RGenomicsETL/duckhts/blob/e87b9f3ee5e27068964e4e57e107bacd7b5def3e/benchmarks/benchmark_simd_bam_gc.md), whose historical run used
3,245,905 HG00106 exome reads / 327,836,405 bases and one DuckDB worker. That input
is not staged here; its timings are not a comparable regression baseline. The
same-run comparison below uses the current driver's committed default input.
Remote I/O, large-input throughput, and CRAM performance remain unmeasured.

## Reproduction and identity

Build each named source revision separately. Set `BAM_SCAN_BASELINE_EXTENSION` to
the baseline binary, `BAM_SCAN_BASELINE_REVISION` to its revision, then render this
report from the repository root. `DUCKHTS_EXTENSION` selects the candidate binary.
The R driver uses `callr`, `duckdb`, `DBI`, and `digest`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r measurement, echo=FALSE}
source("scripts/benchmark_simd_bam_gc.R")
baseline <- Sys.getenv("BAM_SCAN_BASELINE_EXTENSION")
baseline_revision <- Sys.getenv("BAM_SCAN_BASELINE_REVISION")
candidate <- Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension")
stopifnot(nzchar(baseline_revision), file.exists(baseline), file.exists(candidate))
revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE)
source_tree <- system2("git", c("rev-parse", "HEAD:src"), stdout = TRUE)
input <- "test/data/nanopore.bam"
identity <- data.frame(
  property = c("baseline revision", "candidate revision", "candidate src tree",
               "baseline binary MD5", "candidate binary MD5", "input", "input MD5",
               "input bytes", "HTSlib", "host", "R", "DuckDB R package", "CPU affinity", "repetitions"),
  value = c(baseline_revision, revision, source_tree, unname(tools::md5sum(baseline)),
            unname(tools::md5sum(candidate)), input, unname(tools::md5sum(input)),
            file.info(input)$size, readLines("third_party/htslib/VERSION", n = 1),
            Sys.info()[["sysname"]], R.version.string, as.character(packageVersion("duckdb")),
            system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE),
            "9 per build/thread setting after warm-up; fresh R process per setting")
)
knitr::kable(identity)
results <- list()
for (threads in c(1L, 4L)) {
  for (build in c("baseline", "candidate")) {
    ext <- if (build == "baseline") baseline else candidate
    measurements <- benchmark_simd_bam_gc(
      ext, input, iterations = 9L, threads = threads, backends = "auto",
      modes = c("bam_scan", "bam_scan_offset", "bam_materialize_text", "bam_materialize_numeric")
    )
    for (measured in measurements) {
      stopifnot(!measured$skipped, measured$reads == 186L)
      results[[length(results) + 1L]] <- data.frame(
        build = build, workload = measured$benchmark,
        duckdb_workers = threads, htslib_workers_per_handle = 2L,
        input_records = 186L, decoded_sequence_rows = measured$reads,
        input_bases = measured$total_bases,
        sql_output_rows = if (startsWith(measured$benchmark, "bam_materialize_")) 186L else 1L,
        row_multiset_sha256 = if (is.null(measured$row_multiset_sha256)) NA_character_ else measured$row_multiset_sha256,
        gc_sum = measured$gc_fraction_sum, backend = measured$kernel_backend,
        median_seconds = measured$median_sec, min_seconds = measured$min_sec,
        max_seconds = max(unlist(measured$timings_sec))
      )
    }
  }
}
results <- do.call(rbind, results)
stopifnot(length(unique(results$input_bases)) == 1L,
          max(results$gc_sum) - min(results$gc_sum) < 1e-9)
for (workload in c("bam_materialize_text", "bam_materialize_numeric")) {
  stopifnot(length(unique(results$row_multiset_sha256[results$workload == workload])) == 1L)
}
knitr::kable(results[names(results) != "row_multiset_sha256"], digits = 6)
knitr::kable(unique(results[!is.na(results$row_multiset_sha256), c("workload", "row_multiset_sha256")]))
```

Each SEQ/GC query scans 186 BAM records, materializes their sequences into DuckDB vectors,
and returns one aggregate row containing the read count, total bases, and sum of
GC fractions. All nine repetitions validate those values; equality here is not
a substitute for the independent SAM-field and row-multiset regression tests.
The `bam_scan_offset` query additionally materializes and filters `FILE_OFFSET`;
all 186 BGZF BAM records must still reach the aggregate. Exact offset values are
tested separately against physical BGZF/BAM record ends in SQL and installed R.
The full-materialization queries time `CREATE OR REPLACE TEMP TABLE AS SELECT *`
with standard/AUX tags enabled. Their output denominator is 186 materialized table
rows, not the one-row CTAS completion response. Both representations include
249,110 sequence and quality elements; numeric mode also writes packed CIGAR lists.
Outside each timed call, the driver checks the aggregate and a SHA-256 digest of
the complete typed row multiset (including every projected tag), sorted by SQL
`ORDER BY ALL` and serialized by R. Each digest
agrees across revisions and thread settings; text and numeric schemas intentionally
differ. This is a small-input latency measurement, not a materialized-byte or RSS
benchmark. The adversarial formatter/fault fixtures are separate correctness tests.
No claim of general performance parity or speedup follows from these small timings.
Elapsed wall time uses R's `Sys.time()` and includes the DBI call/result overhead;
connection creation, warm-up, and post-materialization verification are not timed.
