---
title: "DuckHTS Score Benchmark"
output: github_document
---

<!-- benchmark_score.md is generated from benchmark_score.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This benchmark validates and times DuckHTS `bcftools_score()` against the
`bcftools +score` plugin bundled by
[`RGenomicsETL/RBCFTools`](https://github.com/RGenomicsETL/RBCFTools).
It covers both a single TSV/PLINK summary and multi-PRS TSV scoring where
several summary files are scored in one genotype scan.

# Run

`make bench-score`

Useful overrides:

- `SCORE_BENCH_ROWS`: synthetic VCF variants, default `100000`
- `SCORE_BENCH_SAMPLES`: genotype samples, default `10`
- `SCORE_BENCH_PRS`: TSV summary files for the multi-PRS case, default `4`
- `SCORE_BENCH_RUNS`: timed repeats, default `3`
- `BCFTOOLS_BIN`: optional override for the bcftools executable
- `SCORE_PLUGIN_DIR`: optional override for the bcftools plugin directory

```{r setup, include = FALSE}
library(DBI)
library(duckdb)

if (!requireNamespace("RBCFTools", quietly = TRUE)) {
  stop("RBCFTools is required for bcftools +score benchmarking", call. = FALSE)
}

RBCFTools::setup_hts_env()

bcftools_bin <- Sys.getenv("BCFTOOLS_BIN", unset = RBCFTools::bcftools_path())
score_plugin_dir <- Sys.getenv("SCORE_PLUGIN_DIR", unset = RBCFTools::bcftools_plugins_dir())
if (!file.exists(bcftools_bin)) stop("bcftools binary not found: ", bcftools_bin, call. = FALSE)
if (!file.exists(file.path(score_plugin_dir, "score.so"))) {
  stop("bcftools score plugin not found under: ", score_plugin_dir, call. = FALSE)
}

Sys.setenv(BCFTOOLS_PLUGINS = score_plugin_dir)

bench_rows <- as.integer(Sys.getenv("SCORE_BENCH_ROWS", unset = "100000"))
bench_runs <- as.integer(Sys.getenv("SCORE_BENCH_RUNS", unset = "3"))
n_samples <- as.integer(Sys.getenv("SCORE_BENCH_SAMPLES", unset = "10"))
n_prs <- as.integer(Sys.getenv("SCORE_BENCH_PRS", unset = "4"))

if (is.na(bench_rows) || bench_rows < 1L) stop("SCORE_BENCH_ROWS must be positive")
if (is.na(bench_runs) || bench_runs < 1L) stop("SCORE_BENCH_RUNS must be positive")
if (is.na(n_samples) || n_samples < 1L) stop("SCORE_BENCH_SAMPLES must be positive")
if (is.na(n_prs) || n_prs < 1L) stop("SCORE_BENCH_PRS must be positive")

drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
con <- dbConnect(drv, dbdir = ":memory:")

ext_path <- normalizePath("build/release/duckhts.duckdb_extension", mustWork = TRUE)
DBI::dbExecute(con, sprintf("LOAD '%s';", gsub("\\\\", "/", ext_path)))
DBI::dbExecute(con, "PRAGMA threads=4;")
```

```{r helpers, include = FALSE}
shell_quote <- function(x) shQuote(x, type = "sh")

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

sql_summary_arg <- function(paths) {
  if (length(paths) == 1L) {
    sql_string(paths)
  } else {
    sprintf("[%s]", paste(sql_string(paths), collapse = ", "))
  }
}

run_cmd <- function(cmd) {
  status <- system(cmd, ignore.stdout = FALSE, ignore.stderr = FALSE)
  if (!identical(status, 0L)) stop("command failed: ", cmd, call. = FALSE)
}

write_score_fixture <- function(dir_path, n_variants, n_samples, n_prs) {
  dir.create(dir_path, recursive = TRUE, showWarnings = FALSE)

  vcf_path <- file.path(dir_path, "score_input.vcf")
  vcf_gz <- paste0(vcf_path, ".gz")
  summary_paths <- file.path(dir_path, sprintf("score_summary_%02d.tsv", seq_len(n_prs)))

  sample_names <- sprintf("SAMPLE_%04d", seq_len(n_samples))
  header_lines <- c(
    "##fileformat=VCFv4.2",
    "##contig=<ID=1,length=250000000>",
    "##FORMAT=<ID=GT,Number=1,Type=String,Description=\"Genotype\">",
    paste0("#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO\tFORMAT\t",
           paste(sample_names, collapse = "\t"))
  )

  alleles <- list(c("A", "T"), c("G", "C"), c("C", "A"), c("T", "G"))

  vcf_lines <- vapply(seq_len(n_variants), function(i) {
    al <- alleles[[(i - 1L) %% length(alleles) + 1L]]
    pos <- i * 10L
    rsid <- sprintf("rs%d", i)
    gts <- sample(c("0/0", "0/1", "1/1"), n_samples, replace = TRUE,
                  prob = c(0.5, 0.35, 0.15))
    paste0("1\t", pos, "\t", rsid, "\t", al[1], "\t", al[2],
           "\t.\tPASS\t.\tGT\t", paste(gts, collapse = "\t"))
  }, character(1))
  writeLines(c(header_lines, vcf_lines), vcf_path)

  for (prs_i in seq_len(n_prs)) {
    set.seed(1000L + prs_i)
    summary_lines <- vapply(seq_len(n_variants), function(i) {
      al <- alleles[[(i - 1L) %% length(alleles) + 1L]]
      pos <- i * 10L
      rsid <- sprintf("rs%d", i)
      beta <- round(rnorm(1, mean = 0, sd = 0.5), 4)
      p <- runif(1, 1e-8, 1)
      paste0("1\t", pos, "\t", rsid, "\t", al[2], "\t", beta, "\t", p)
    }, character(1))
    writeLines(c("CHR\tBP\tSNP\tA1\tBETA\tP", summary_lines), summary_paths[[prs_i]])
  }

  run_cmd(sprintf(
    "%s view -Oz -o %s %s",
    shell_quote(bcftools_bin), shell_quote(vcf_gz), shell_quote(vcf_path)
  ))
  run_cmd(sprintf("%s index -t %s", shell_quote(bcftools_bin), shell_quote(vcf_gz)))

  list(
    vcf_path = vcf_path,
    vcf_gz = vcf_gz,
    summary_paths = summary_paths,
    n_variants = n_variants,
    n_samples = n_samples,
    n_prs = n_prs
  )
}

duckdb_score_sql <- function(vcf_path, summary_paths) {
  paste0(
    "SELECT * FROM bcftools_score(",
    sql_string(vcf_path), ", ",
    sql_summary_arg(summary_paths), ", ",
    "use := 'GT', columns := 'PLINK') ",
    "ORDER BY SAMPLE"
  )
}

run_duckdb_bench <- function(vcf_path, summary_paths, n = 3L) {
  sql <- duckdb_score_sql(vcf_path, summary_paths)
  times <- numeric(n)
  rows <- NA_integer_
  DBI::dbExecute(con, "PRAGMA threads=1")
  on.exit(DBI::dbExecute(con, "PRAGMA threads=4"), add = TRUE)

  for (i in seq_len(n)) {
    gc()
    t0 <- proc.time()[["elapsed"]]
    out <- DBI::dbGetQuery(con, sql)
    times[i] <- proc.time()[["elapsed"]] - t0
    rows <- nrow(out)
  }

  data.frame(
    engine = "duckhts",
    runs = n,
    median_sec = stats::median(times),
    min_sec = min(times),
    max_sec = max(times),
    output_rows = rows,
    stringsAsFactors = FALSE
  )
}

run_duckdb_output <- function(vcf_path, summary_paths) {
  DBI::dbExecute(con, "PRAGMA threads=1")
  on.exit(DBI::dbExecute(con, "PRAGMA threads=4"), add = TRUE)
  DBI::dbGetQuery(con, duckdb_score_sql(vcf_path, summary_paths))
}

run_bcftools_once <- function(vcf_gz, summary_paths, output_path) {
  summary_args <- paste(shell_quote(summary_paths), collapse = " ")
  cmd <- sprintf(
    "BCFTOOLS_PLUGINS=%s %s +score --use GT -c PLINK -o %s %s %s",
    shell_quote(score_plugin_dir),
    shell_quote(bcftools_bin),
    shell_quote(output_path),
    shell_quote(vcf_gz),
    summary_args
  )
  gc()
  t0 <- proc.time()[["elapsed"]]
  status <- system(cmd, ignore.stdout = TRUE, ignore.stderr = FALSE)
  elapsed <- proc.time()[["elapsed"]] - t0
  if (!identical(status, 0L)) stop("bcftools +score failed", call. = FALSE)

  list(
    elapsed = elapsed,
    output = output_path,
    data = utils::read.delim(output_path, stringsAsFactors = FALSE, check.names = FALSE)
  )
}

compare_outputs <- function(duck_out, bcf_out, rel_tol = 1e-2) {
  common_cols <- intersect(names(duck_out), names(bcf_out))
  score_cols <- setdiff(common_cols, "SAMPLE")

  merged <- merge(duck_out[, common_cols, drop = FALSE],
                  bcf_out[, common_cols, drop = FALSE],
                  by = "SAMPLE", all = TRUE, suffixes = c("_duck", "_bcf"))

  n_match <- 0L
  n_mismatch <- 0L
  n_only_duck <- sum(is.na(match(duck_out$SAMPLE, bcf_out$SAMPLE)))
  n_only_bcf <- sum(is.na(match(bcf_out$SAMPLE, duck_out$SAMPLE)))
  max_abs_diff <- 0
  max_rel_diff <- 0

  for (col in score_cols) {
    duck_col <- paste0(col, "_duck")
    bcf_col <- paste0(col, "_bcf")
    if (duck_col %in% names(merged) && bcf_col %in% names(merged)) {
      d <- as.numeric(merged[[duck_col]])
      b <- as.numeric(merged[[bcf_col]])
      abs_diff <- abs(d - b)
      denom <- pmax(abs(d), abs(b), 1.0)
      rel_diff <- abs_diff / denom
      matches <- !is.na(rel_diff) & rel_diff <= rel_tol
      n_match <- n_match + sum(matches, na.rm = TRUE)
      n_mismatch <- n_mismatch + sum(!matches, na.rm = TRUE)
      max_abs_diff <- max(max_abs_diff, max(abs_diff, na.rm = TRUE))
      max_rel_diff <- max(max_rel_diff, max(rel_diff, na.rm = TRUE))
    }
  }

  data.frame(
    duck_samples = nrow(duck_out),
    bcf_samples = nrow(bcf_out),
    score_columns = length(score_cols),
    cell_matches = n_match,
    cell_mismatches = n_mismatch,
    only_duck_samples = n_only_duck,
    only_bcftools_samples = n_only_bcf,
    max_abs_diff = max_abs_diff,
    max_rel_diff = max_rel_diff,
    outputs_match = (n_mismatch == 0L && n_only_duck == 0L && n_only_bcf == 0L),
    stringsAsFactors = FALSE
  )
}

run_benchmark_case <- function(fixture, summary_paths, n = 3L) {
  duck <- run_duckdb_bench(fixture$vcf_path, summary_paths, n = n)
  duck_out <- run_duckdb_output(fixture$vcf_path, summary_paths)

  bcftools_times <- numeric(n)
  bcftools_out <- NULL
  output_rows <- NA_integer_
  for (i in seq_len(n)) {
    out_tsv <- tempfile("bcftools_score_", fileext = ".tsv")
    run <- run_bcftools_once(fixture$vcf_gz, summary_paths, out_tsv)
    bcftools_times[i] <- run$elapsed
    output_rows <- nrow(run$data)
    if (i == 1L) bcftools_out <- run$data
    unlink(out_tsv, force = TRUE)
  }

  list(
    bench = rbind(
      duck,
      data.frame(
        engine = "bcftools_RBCFTools",
        runs = n,
        median_sec = stats::median(bcftools_times),
        min_sec = min(bcftools_times),
        max_sec = max(bcftools_times),
        output_rows = output_rows,
        stringsAsFactors = FALSE
      )
    ),
    compare = compare_outputs(duck_out, bcftools_out)
  )
}
```

## Tools

```{r tools, echo = FALSE}
data.frame(
  tool = c("bcftools", "bcftools_plugins"),
  path = c(bcftools_bin, score_plugin_dir),
  stringsAsFactors = FALSE
)
```

## Settings

```{r settings, echo = FALSE}
data.frame(
  variants = bench_rows,
  samples = n_samples,
  multi_prs_files = n_prs,
  runs = bench_runs,
  stringsAsFactors = FALSE
)
```

## Synthetic Single-PRS Case

```{r synthetic-single, echo = FALSE}
set.seed(42L)
fixture_dir <- file.path(tempdir(), "duckhts_score_benchmark")
fixture <- write_score_fixture(fixture_dir, n_variants = bench_rows,
                               n_samples = n_samples, n_prs = n_prs)
single_case <- run_benchmark_case(fixture, fixture$summary_paths[1], n = bench_runs)
single_case$bench
```

```{r synthetic-single-compare, echo = FALSE}
single_case$compare
```

## Synthetic Multi-PRS TSV Case

```{r synthetic-multi, echo = FALSE}
multi_case <- run_benchmark_case(fixture, fixture$summary_paths, n = bench_runs)
multi_case$bench
```

```{r synthetic-multi-compare, echo = FALSE}
multi_case$compare
```
