---
title: "Automatic BAM full-file scan: small-fixture cost"
output:
  github_document:
    html_preview: false
---

This compares automatic BAM scans at two named reader revisions,
using the existing `benchmark_simd_bam_gc.py` workload on its committed nanopore
fixture, with and without projecting `FILE_OFFSET` via an `IS NOT NULL` filter.
It exercises many empty contig claims and full `SEQ` decoding, but this
186-record fixture is dominated by setup cost: it is not a WGS throughput benchmark.
There are no unplaced reads in this performance input, so both builds do equal work.
The separate SQL/R regression corpus tests retained no-coordinate records,
physical duplicates, BAI/CSI, CRAM, and output chunks.

The nearest identical checked-in workload is this report's previous rendering
at `8e743aceb4109b5894ff23cf8f5c8e1b9a1607a1`: source `05fcdfdfcb32eb94a987d946e65d377040644c1a`,
186 records / 249,110 bases, 1 or 4 DuckDB workers and 2 HTSlib workers per handle.
Its candidate medians without `FILE_OFFSET` were 1.561 ms and 2.239 ms respectively.
The unchanged baseline binary below contains that same reader source. This is
the first rendered measurement requiring `FILE_OFFSET`; its same-run baseline
uses the pre-fix binary, not an invented historical offset workload.

The older larger-input report is
[`benchmark_simd_bam_gc.md`](benchmark_simd_bam_gc.md), whose historical run used
3,245,905 HG00106 exome reads / 327,836,405 bases and one DuckDB worker. That input
is not staged here; its timings are not a comparable regression baseline. The
same-run comparison below uses the current driver's committed default input.
Remote I/O, large-input throughput, and CRAM performance remain unmeasured.

## Reproduction and identity

Build each named source revision separately. Set `BAM_SCAN_BASELINE_EXTENSION` to
the baseline binary, `BAM_SCAN_BASELINE_REVISION` to its revision, then render this
report from the repository root. `DUCKHTS_EXTENSION` selects the candidate binary;
`PYTHON` selects the Python environment with DuckDB installed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r measurement, echo=FALSE}
library(jsonlite)
baseline <- Sys.getenv("BAM_SCAN_BASELINE_EXTENSION")
baseline_revision <- Sys.getenv("BAM_SCAN_BASELINE_REVISION")
candidate <- Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension")
python <- Sys.getenv("PYTHON", "configure/venv/bin/python3")
stopifnot(nzchar(baseline_revision), file.exists(baseline), file.exists(candidate))
revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE)
source_tree <- system2("git", c("rev-parse", "HEAD:src"), stdout = TRUE)
input <- "test/data/nanopore.bam"
identity <- data.frame(
  property = c("baseline revision", "candidate revision", "candidate src tree",
               "baseline binary MD5", "candidate binary MD5", "input", "input MD5",
               "input bytes", "HTSlib", "host", "CPU affinity", "repetitions"),
  value = c(baseline_revision, revision, source_tree, unname(tools::md5sum(baseline)),
            unname(tools::md5sum(candidate)), input, unname(tools::md5sum(input)),
            file.info(input)$size, readLines("third_party/htslib/VERSION", n = 1),
            Sys.info()[["sysname"]],
            system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE),
            "9 per build/thread setting after warm-up; fresh Python process per setting")
)
knitr::kable(identity)
results <- list()
for (threads in c(1L, 4L)) {
  for (build in c("baseline", "candidate")) {
    output <- tempfile(fileext = ".json")
    ext <- if (build == "baseline") baseline else candidate
    args <- c("scripts/benchmark_simd_bam_gc.py", "--extension", ext,
              "--bam", input, "--max-reads", "0", "--iterations", "9",
              "--threads", threads, "--backends", "auto", "--modes", "bam_scan,bam_scan_offset",
              "--json-out", output)
    log <- system2(python, shQuote(args), stdout = TRUE, stderr = TRUE)
    stopifnot(is.null(attr(log, "status")) || attr(log, "status") == 0L)
    for (measured in fromJSON(output, simplifyVector = FALSE)) {
      stopifnot(!measured$skipped, measured$reads == 186L)
      results[[length(results) + 1L]] <- data.frame(
        build = build, workload = measured$benchmark,
        duckdb_workers = threads, htslib_workers_per_handle = 2L,
        input_records = 186L, decoded_sequence_rows = measured$reads,
        input_bases = measured$total_bases, sql_output_rows = 1L,
        gc_sum = measured$gc_fraction_sum, backend = measured$kernel_backend,
        median_seconds = measured$median_sec, min_seconds = measured$min_sec,
        max_seconds = max(unlist(measured$timings_sec))
      )
    }
    unlink(output)
  }
}
results <- do.call(rbind, results)
stopifnot(length(unique(results$input_bases)) == 1L,
          max(results$gc_sum) - min(results$gc_sum) < 1e-9)
knitr::kable(results, digits = 6)
```

Each query scans 186 BAM records, materializes their sequences into DuckDB vectors,
and returns one aggregate row containing the read count, total bases, and sum of
GC fractions. All nine repetitions validate those values; equality here is not
a substitute for the independent SAM-field and row-multiset regression tests.
The `bam_scan_offset` query additionally materializes and filters `FILE_OFFSET`;
all 186 BGZF BAM records must still reach the aggregate. Exact offset values are
tested separately against physical BGZF/BAM record ends in SQL and installed R.
No claim of general performance parity or speedup follows from these small timings.
