---
title: "Ancestry panel building at scale"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
root <- normalizePath("..", mustWork = TRUE)
runs <- utils::read.delim(file.path(root, "benchmarks", "ancestry_panel_scaling.tsv"),
                          colClasses = c(panel_sha256 = "character"))
mib <- function(bytes) bytes / 1048576
# Budget, declared before measuring: fixed overhead plus three times the live state the
# builder needs, which is its per-locus eligibility state. Each reference locus keeps a
# group key and constant aggregate state on the frequency side (chromosome and position
# key 20 B; four counts 32 B; min/max of both alleles 64 B; group mask 8 B; validity
# flag 1 B) and on the loading side (key 20 B; count 8 B; PC mask 16 B; flag 1 B):
# 170 B per locus. Spill is not allowed.
state_bytes_per_locus <- 170
overhead <- subset(runs, mode == "overhead")
overhead_mib <- stats::median(overhead$peak_rss_kib) / 1024
runs <- subset(runs, mode != "overhead")
# Loci, groups and PCs depend only on the scale; fill them for runs that ran out of memory.
shape <- unique(subset(runs, status == "ok")[c("quarters", "loci", "groups", "pcs")])
stopifnot(!anyDuplicated(shape$quarters))
runs[c("loci", "groups", "pcs")] <- shape[match(runs$quarters, shape$quarters), c("loci", "groups", "pcs")]
runs$long_reference_rows <- runs$loci * runs$groups
runs$long_loading_rows <- runs$loci * runs$pcs
# Decoded required columns as DuckDB vectors: VARCHAR 16 B, UINTEGER/INTEGER 4 B, DOUBLE 8 B.
runs$decoded_input_mib <- mib(runs$long_reference_rows * (16 + 4 + 16 + 16 + 16 + 8) +
                                runs$long_loading_rows * (16 + 4 + 16 + 16 + 4 + 8))
runs$budget_mib <- overhead_mib + 3 * mib(runs$loci * state_bytes_per_locus)
runs$peak_rss_mib <- runs$peak_rss_kib / 1024
runs$peak_buffer_mib <- mib(runs$peak_buffer_bytes)
runs$spill_mib <- mib(runs$peak_temp_bytes)
keys <- c("implementation", "threads", "mode", "quarters")
summary <- do.call(rbind, lapply(split(runs, runs[keys], drop = TRUE), function(d) {
  ok <- d[d$status == "ok", ]
  data.frame(
    d[1, c(keys, "loci", "long_reference_rows")],
    decoded_input_mib = round(d$decoded_input_mib[1]),
    completed = sprintf("%d/%d", nrow(ok), nrow(d)),
    sites = if (nrow(ok)) ok$sites[1] else NA, contigs = if (nrow(ok)) ok$contigs[1] else NA,
    median_s = if (nrow(ok)) stats::median(ok$seconds) else NA,
    min_s = if (nrow(ok)) min(ok$seconds) else NA, max_s = if (nrow(ok)) max(ok$seconds) else NA,
    peak_rss_mib = if (nrow(ok)) round(max(ok$peak_rss_mib)) else NA,
    peak_buffer_mib = if (nrow(ok)) round(max(ok$peak_buffer_mib)) else NA,
    spill_mib = if (nrow(ok)) max(ok$spill_mib) else NA, budget_mib = round(d$budget_mib[1]),
    hashes = length(unique(ok$panel_sha256)),
    panel_sha256 = if (nrow(ok)) ok$panel_sha256[1] else NA)
}))
summary <- summary[order(summary$threads, summary$mode, summary$quarters,
                         match(summary$implementation, c("develop", "distinct", "final"))), ]
rownames(summary) <- NULL
final <- subset(summary, implementation == "final")
distinct <- subset(summary, implementation == "distinct")
same <- merge(final, subset(distinct, completed == "3/3"), by = c("threads", "mode", "quarters"))
failed <- subset(summary, completed != "3/3")
exponent <- function(d, y) unname(stats::coef(stats::lm(log(d[[y]]) ~ log(d$quarters)))[2])
exponents <- do.call(rbind, lapply(split(final, final[c("threads", "mode")], drop = TRUE),
  function(d) data.frame(threads = d$threads[1], mode = d$mode[1],
                         time_exponent = round(exponent(d, "median_s"), 2),
                         rss_exponent = round(exponent(d, "peak_rss_mib"), 2))))
rownames(exponents) <- NULL
stopifnot(
  nrow(overhead) == 9L, all(table(runs$implementation, runs$threads, runs$mode, runs$quarters) == 3L),
  all(runs$status %in% c("ok", "out_of_memory")), all(final$completed == "3/3"),
  all(runs$sites[runs$status == "ok"] == 17000L),
  all(summary$hashes[summary$completed != "0/3"] == 1L), nrow(same) > 0L,
  all(same$panel_sha256.x == same$panel_sha256.y),
  all(final$contigs == 22L), all(subset(summary, implementation == "develop")$contigs %in% c(1L, NA)),
  all(final$peak_rss_mib <= final$budget_mib), all(final$spill_mib == 0),
  all(final$min_s[final$threads == 1L & final$quarters == 1L] >= 5),
  all(exponents$time_exponent <= 1.25))
```

`rduckhts_ancestry_panel()` reads a long reference (loci × groups frequency rows and
loci × PCs loading rows), keeps the eligible loci, and caps the panel at `max_sites`.
This report measures three builders on the same inputs:

- **develop** keeps the first `max_sites` eligible loci in genomic order and checks
  eligibility with distinct aggregates (one allele pair, every group and PC once).
- **distinct** allocates the cap per contig (one site each, the rest in proportion to
  each contig's eligible sites minus one, spread along the contig) with the same
  distinct-aggregate eligibility checks.
- **final** keeps the per-contig allocation and checks eligibility with constant
  per-locus state: `min = max` for each allele, and a row count equal to the group (or
  PC) count whose ordinal bits cover the full mask, so each group and PC occurs exactly
  once.

## Method

`benchmark_ancestry_panel_run.R` stages bigsnpr's reference through the benchmark
registry (`ancestry_reference_parquet`: 5,816,590 loci, 21 groups, 16 PCs) and keeps 1,
2 or 4 quarters of its loci by hash, the one growth dimension (groups and PCs are fixed
by the reference). Each run expands them into the long relations the builder reads,
builds a committed 17,000-site panel through the public function, and records the
panel hash. `spaced` uses the default 5 kb spacing; `candidates` offers every locus as
a candidate, the largest input the allocation windows see. Every run is a fresh process
with `memory_limit = '`r runs$memory_limit[1]`'` and its own empty temporary directory, at
one and four threads, three repetitions interleaved across builders. The reference
Parquet file is read from a warm page cache. DuckDB peak buffer memory and spill are
the maxima of the per-statement profiles over every statement the builder runs; peak
RSS is the process high-water mark. Other jobs shared the machine; builders were
interleaved so they saw the same conditions.

The memory budget is fixed overhead
(`r round(overhead_mib)` MiB: packages, connection and extension load, measured by
runs that build no panel) plus three times the eligibility live state, 170 bytes per
locus of group key and constant aggregate state; spill must be zero. Decoded input is
the required long columns as DuckDB vectors.

## Results

```{r results}
knitr::kable(summary[c(keys, "loci", "long_reference_rows", "decoded_input_mib", "completed", "median_s",
                       "min_s", "max_s", "peak_rss_mib", "peak_buffer_mib", "spill_mib",
                       "budget_mib", "contigs")])
```

```{r failures, results = "asis"}
if (nrow(failed)) {
  cat("These runs stopped with DuckDB's out-of-memory error at the",
      runs$memory_limit[1], "limit:",
      paste0(paste(sprintf("%s at %d thread%s, %s, %d quarters", failed$implementation,
                           failed$threads, ifelse(failed$threads == 1, "", "s"), failed$mode,
                           failed$quarters), collapse = "; "), ".\n"))
}
```

Scaling exponents of the final builder over 1, 2 and 4 quarters:

```{r exponents}
knitr::kable(exponents)
```

## Reading

```{r reading-values}
pick <- function(d, threads, column) d[[column]][d$threads == threads & d$mode == "spaced" & d$quarters == 4]
old_spill <- max(subset(summary, implementation != "final" & quarters == 4 & threads == 1)$spill_mib)
```

The final builder produces the same panel hash as the distinct builder wherever the
distinct builder completed, so the constant-state checks select exactly the same loci.
At the full reference with one thread it takes
`r sprintf("%.1f", pick(final, 1, "median_s"))` s and
`r format(pick(final, 1, "peak_rss_mib"), big.mark = ",")` MiB peak RSS, against
`r sprintf("%.1f", pick(distinct, 1, "median_s"))` s and
`r format(pick(distinct, 1, "peak_rss_mib"), big.mark = ",")` MiB with distinct
aggregates, which hold a set of alleles, groups or PCs per locus across the whole
reference: they exceed the budget from the first quarter, spill up to
`r format(round(old_spill), big.mark = ",")` MiB at full scale with one thread, and run
out of memory at four threads. Every final run is within its budget with no spill; with
four threads the full reference takes `r sprintf("%.1f", pick(final, 4, "median_s"))` s.
Time and memory grow no faster than the reference.

The per-contig allocation itself is cheap: the develop and distinct builders differ only
in allocation and take similar time and memory. Every develop panel comes from
chromosome 1; every final panel spans all 22 autosomes.
