---
title: "GenBank reader benchmark"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

This report measures `read_genbank(...)` and `genbank_to_fasta(...)` on a real
annotation: the NCBI RefSeq record for *Escherichia coli* K-12 MG1655
(`NC_000913.3`), a 4,641,652 bp circular genome carrying 9,306 features across 8
feature kinds. Fetching the record is staging work and is outside every timed
interval.

There is no earlier checked-in DuckHTS GenBank benchmark. The nearest reader
report is
[`benchmark_gffbase_conformance.md`](benchmark_gffbase_conformance.md), which
measures `read_gff(...)`. It is the closest analogue because `read_genbank`
emits `read_gff`'s exact column shape, but the two are not numerically
comparable: that workload reads tabix-style tabular rows, while this parses a
GenBank flat file line by line through hFILE.

`read_genbank` declares no parallel scan and parses the whole record in `init`
before the first chunk, so the one- and four-thread conditions exist to show
whether threading adds overhead rather than to show it scales. Memory therefore
tracks the annotation size rather than the projection; that limit is visible in
the input denominators below but is not itself measured here.

## Reproduction

Stage the record from NCBI. `gbwithparts` returns the full FEATURES table and
the ORIGIN sequence, which is what both functions read.

```sh
mkdir -p /tmp/ncbi-genbank-bench
curl -fsSL -o /tmp/ncbi-genbank-bench/NC_000913.3.gb \
  "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=nuccore&id=NC_000913.3&rettype=gbwithparts&retmode=text"
```

Then render this report from a tracked-clean revision:

```sh
Rscript -e 'rmarkdown::render("benchmarks/benchmark_genbank_reader.Rmd")'
```

The benchmark pins the one-thread condition to logical CPU 12 and the
four-thread condition to logical CPUs 12--15 on the recorded host. Input
staging, extension loading, connection setup, warm-up, result verification, and
removal of the previous FASTA output are excluded from timing.

```{r benchmark, echo=FALSE}
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
  library(Rduckhts)
})

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

command_output <- function(command, args = character()) {
  paste(system2(command, args, stdout = TRUE, stderr = TRUE), collapse = " ")
}

input_path <- Sys.getenv(
  "DUCKHTS_GENBANK_BENCHMARK_INPUT",
  "/tmp/ncbi-genbank-bench/NC_000913.3.gb"
)
if (!file.exists(input_path)) {
  stop("Stage DUCKHTS_GENBANK_BENCHMARK_INPUT using the commands above")
}
input_path <- normalizePath(input_path, mustWork = TRUE)

source_revision <- command_output("git", c("rev-parse", "--short=12", "HEAD"))
tracked_dirty <- length(system2(
  "git", c("status", "--porcelain", "--untracked-files=no"), stdout = TRUE
)) > 0L
if (tracked_dirty) {
  stop("Render the benchmark from a tracked-clean source revision")
}

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
invisible(rduckhts_load(con))

# One row back to R, so the timing covers parsing and decoding rather than
# transport. The fingerprint is order independent, so a reordered scan that
# still produced the right bag of rows would pass, and a dropped or duplicated
# feature would not.
aggregate_sql <- function(attributes_map = FALSE) {
  # The scan only fills PROJECTED columns, so selecting attributes_map is what makes
  # the MAP get built at all. Aggregating its cardinality keeps one row coming back
  # while forcing every entry to be materialized.
  map_term <- if (attributes_map) {
    "sum(cardinality(attributes_map))::DOUBLE AS map_entries"
  } else {
    "0::DOUBLE AS map_entries"
  }
  sprintf(
    paste0(
      "SELECT count(*)::UBIGINT AS rows, ",
      "hex(bit_xor(hash(seqname, feature, start, \"end\", strand, frame, attributes))) ",
      "AS fingerprint, sum(\"end\" - start + 1)::DOUBLE AS span_sum, %s ",
      "FROM read_genbank(%s, attributes_map := %s)"
    ),
    map_term, sql_string(input_path), if (attributes_map) "TRUE" else "FALSE"
  )
}

plain_expected <- dbGetQuery(con, aggregate_sql(FALSE))
map_expected <- dbGetQuery(con, aggregate_sql(TRUE))
stopifnot(as.double(plain_expected$rows[[1L]]) > 9000)
# The MAP projection must not change which features are emitted.
stopifnot(identical(plain_expected$fingerprint[[1L]], map_expected$fingerprint[[1L]]))
# The MAP workload must actually have built entries, or it is measuring a pruned column.
stopifnot(map_expected$map_entries[[1L]] > plain_expected$rows[[1L]])

same_answer <- function(observed, expected) {
  identical(as.character(observed$rows[[1L]]), as.character(expected$rows[[1L]])) &&
    identical(observed$fingerprint[[1L]], expected$fingerprint[[1L]]) &&
    isTRUE(all.equal(
      observed$span_sum[[1L]], expected$span_sum[[1L]],
      tolerance = 1e-12
    )) &&
    isTRUE(all.equal(
      observed$map_entries[[1L]], expected$map_entries[[1L]],
      tolerance = 1e-12
    ))
}

set_affinity <- function(cpus) {
  output <- system2(
    "taskset", c("-pc", cpus, Sys.getpid()),
    stdout = TRUE, stderr = TRUE
  )
  if (!is.null(attr(output, "status"))) stop(paste(output, collapse = " "))
  invisible(output)
}

measure_aggregate <- function(workload, attributes_map, expected, threads, cpus,
                              repetitions = 9L) {
  set_affinity(cpus)
  invisible(dbExecute(con, sprintf("SET threads = %d", threads)))
  sql <- aggregate_sql(attributes_map)
  for (i in seq_len(2L)) stopifnot(same_answer(dbGetQuery(con, sql), expected))
  seconds <- numeric(repetitions)
  for (i in seq_len(repetitions)) {
    started <- proc.time()[["elapsed"]]
    observed <- dbGetQuery(con, sql)
    seconds[[i]] <- proc.time()[["elapsed"]] - started
    stopifnot(same_answer(observed, expected))
  }
  data.frame(
    workload = workload,
    threads = threads,
    affinity = cpus,
    repetitions = repetitions,
    output_rows = as.double(expected$rows[[1L]]),
    output_bytes = NA_real_,
    min_seconds = min(seconds),
    median_seconds = median(seconds),
    max_seconds = max(seconds),
    median_rows_per_second = as.double(expected$rows[[1L]]) / median(seconds),
    fingerprint = expected$fingerprint[[1L]],
    span_sum = expected$span_sum[[1L]],
    stringsAsFactors = FALSE
  )
}

# The write path reads ORIGIN rather than FEATURES, so it exercises the other
# half of the parser. Each pass is verified by reading the FASTA back.
measure_fasta <- function(threads, cpus, repetitions = 5L) {
  set_affinity(cpus)
  invisible(dbExecute(con, sprintf("SET threads = %d", threads)))
  output_path <- tempfile(sprintf("genbank-t%d-", threads), fileext = ".fa")
  convert_sql <- sprintf(
    "SELECT records_written FROM genbank_to_fasta(%s, output_path := %s, overwrite := TRUE)",
    sql_string(input_path), sql_string(output_path)
  )
  verify_sql <- sprintf(
    "SELECT count(*)::UBIGINT AS rows, sum(length(SEQUENCE))::DOUBLE AS bases FROM read_fasta(%s)",
    sql_string(output_path)
  )
  seconds <- numeric(repetitions)
  output_bytes <- numeric(repetitions)
  records <- NA_real_
  for (i in seq_len(repetitions)) {
    unlink(output_path)
    started <- proc.time()[["elapsed"]]
    written <- dbGetQuery(con, convert_sql)
    seconds[[i]] <- proc.time()[["elapsed"]] - started
    output_bytes[[i]] <- file.info(output_path)$size
    records <- as.double(written$records_written[[1L]])
    verified <- dbGetQuery(con, verify_sql)
    stopifnot(as.double(verified$rows[[1L]]) == records)
    stopifnot(as.double(verified$bases[[1L]]) == 4641652)
  }
  unlink(output_path)
  data.frame(
    workload = "ORIGIN to FASTA",
    threads = threads,
    affinity = cpus,
    repetitions = repetitions,
    output_rows = records,
    output_bytes = median(output_bytes),
    min_seconds = min(seconds),
    median_seconds = median(seconds),
    max_seconds = max(seconds),
    median_rows_per_second = 4641652 / median(seconds),
    fingerprint = NA_character_,
    span_sum = NA_real_,
    stringsAsFactors = FALSE
  )
}

results <- rbind(
  measure_aggregate("features + aggregate", FALSE, plain_expected, 1L, "12"),
  measure_aggregate("features + aggregate", FALSE, plain_expected, 4L, "12-15"),
  measure_aggregate("features + attribute MAP", TRUE, map_expected, 1L, "12"),
  measure_aggregate("features + attribute MAP", TRUE, map_expected, 4L, "12-15"),
  measure_fasta(1L, "12"),
  measure_fasta(4L, "12-15")
)

results$source_revision <- source_revision
results$run_date <- as.character(Sys.Date())
results$input_path <- input_path
results$input_bytes <- file.info(input_path)$size
results$input_span_bases <- 4641652
results$duckdb_version <- dbGetQuery(con, "SELECT version() AS v")$v[[1L]]
results$htslib_version <- rduckhts_htslib_version(con)

dir.create("benchmarks/data", showWarnings = FALSE, recursive = TRUE)
write.csv(
  results,
  "benchmarks/data/genbank_reader.csv",
  row.names = FALSE,
  na = ""
)

cpu_model <- sub(
  "^model name[[:space:]]*:[[:space:]]*", "",
  grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[[1L]]
)
metadata <- data.frame(
  property = c(
    "source revision", "run date", "input source", "input accession",
    "GenBank bytes", "genome bases", "emitted features", "distinct feature kinds",
    "feature fingerprint", "DuckDB version", "htslib version", "CPU"
  ),
  value = c(
    source_revision,
    as.character(Sys.Date()),
    "NCBI RefSeq efetch (gbwithparts)",
    "NC_000913.3",
    format(file.info(input_path)$size, big.mark = ",", scientific = FALSE),
    format(4641652, big.mark = ",", scientific = FALSE),
    format(as.double(plain_expected$rows[[1L]]), big.mark = ",", scientific = FALSE),
    as.character(dbGetQuery(
      con,
      sprintf("SELECT count(DISTINCT feature) AS k FROM read_genbank(%s)",
              sql_string(input_path))
    )$k[[1L]]),
    plain_expected$fingerprint[[1L]],
    results$duckdb_version[[1L]],
    results$htslib_version[[1L]],
    cpu_model
  ),
  stringsAsFactors = FALSE
)

knitr::kable(metadata, col.names = c("Property", "Recorded value"))
invisible(dbDisconnect(con, shutdown = TRUE))
```

## Results

The aggregate workloads parse the FEATURES table and return one row to R, so
the timing covers parsing and decoding rather than transport. The FASTA
workload reads ORIGIN instead and writes every base as a real file. Every timed
pass is verified: the aggregates by exact row count and an order-independent
full-row fingerprint, with the summed span checked to a relative tolerance of
`1e-12`, and the FASTA by reading it back and asserting both the record count
and 4,641,652 bases.

```{r results, echo=FALSE}
display <- results[c(
  "workload", "threads", "affinity", "repetitions", "output_rows",
  "output_bytes", "min_seconds", "median_seconds", "max_seconds",
  "median_rows_per_second"
)]
display$output_rows <- format(
  display$output_rows, big.mark = ",", scientific = FALSE
)
display$output_bytes <- ifelse(
  is.na(display$output_bytes), "--",
  format(display$output_bytes, big.mark = ",", scientific = FALSE)
)
for (column in c("min_seconds", "median_seconds", "max_seconds")) {
  display[[column]] <- sprintf("%.3f", display[[column]])
}
display$median_rows_per_second <- format(
  round(display$median_rows_per_second), big.mark = ",", scientific = FALSE
)
names(display) <- c(
  "Workload", "Threads", "CPU affinity", "Runs", "Output rows",
  "FASTA bytes", "Minimum seconds", "Median seconds", "Maximum seconds",
  "Median rows/s"
)
knitr::kable(display, align = c("l", rep("r", 9)))
```

The features and ORIGIN rows count different things: the aggregate workloads
emit one row per feature, while the FASTA rate is expressed in bases written,
since the converter produces a single record for this genome. Adding
`attributes_map := TRUE` builds a `MAP<VARCHAR,VARCHAR>` per feature on top of
the raw attribute string and is the more expensive projection, which is why
both are recorded.

The measurement does not include the NCBI fetch, extension loading, or any
downstream join against the emitted intervals. It also does not bound memory:
the reader materializes every feature before the first chunk, so a much larger
annotation is a different question than this report answers.
