---
title: "Indexed multi-region reader benchmark"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

This report measures the htslib 1.24 native multi-region iterator used by
`read_bcf(...)`. The same generated records are read from indexed BCF, which
uses `bcf_itr_regarray(...)`, and indexed VCF text, which uses
`tbx_itr_regarray(...)`. The requested intervals overlap by design; each query
must emit every matching VCF record once.

There is no earlier checked-in multi-region reader benchmark with the same
workload. The full-file `read_bcf(...)` measurements in
[`benchmark_variantkey_conformance.md`](benchmark_variantkey_conformance.md)
are the nearest reader baseline, but they are not directly comparable because
they do not create indexed region iterators.

## Reproduction

Build and install the current `Rduckhts` tarball, then render on one allowed
CPU. For example:

```bash
taskset -c 2 Rscript -e "rmarkdown::render('benchmarks/benchmark_multi_region_readers.Rmd', output_format = 'github_document')"
```

The fixture is generated locally and deterministically. Fixture creation,
BGZF compression, indexing, connection setup, and warm-up are outside the
timed interval.

```{r benchmark, echo=FALSE}
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
  library(Rduckhts)
})

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

command_output <- function(command, args = character()) {
  paste(system2(command, args, stdout = TRUE, stderr = TRUE), collapse = " ")
}

git_revision <- command_output("git", c("rev-parse", "--short=12", "HEAD"))
dirty <- length(system2(
  "git", c("status", "--porcelain", "--untracked-files=no"), stdout = TRUE
)) > 0L
source_state <- if (dirty) paste0(git_revision, " + report-source changes") else git_revision

row_count <- as.integer(Sys.getenv("DUCKHTS_MULTI_REGION_ROWS", "2000000"))
region_count <- 64L
region_width <- 50000L
region_step <- 25000L
region_starts <- 1L + (seq_len(region_count) - 1L) * region_step
region_ends <- region_starts + region_width - 1L
stopifnot(max(region_ends) <= row_count)
regions <- sprintf("1:%d-%d", region_starts, region_ends)
region_spec <- paste(regions, collapse = ",")

fixture_dir <- tempfile("duckhts-multi-region-")
invisible(dir.create(fixture_dir))
vcf_path <- file.path(fixture_dir, "variants.vcf")
vcf_gz_path <- paste0(vcf_path, ".gz")
bcf_path <- file.path(fixture_dir, "variants.bcf")

out <- file(vcf_path, open = "wt")
writeLines(c(
  "##fileformat=VCFv4.3",
  sprintf("##contig=<ID=1,length=%d>", row_count),
  "#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO"
), out)
chunk_size <- 100000L
for (first in seq.int(1L, row_count, by = chunk_size)) {
  last <- min(row_count, first + chunk_size - 1L)
  pos <- seq.int(first, last)
  writeLines(sprintf("1\t%d\t.\tA\tC\t.\tPASS\t.", pos), out)
}
close(out)

status <- system2("bgzip", c("-@", "1", "-c", vcf_path), stdout = vcf_gz_path)
stopifnot(status == 0L)
stopifnot(system2("tabix", c("-f", "-p", "vcf", vcf_gz_path)) == 0L)
stopifnot(system2("bcftools", c("view", "--no-version", "-Ob", "-o", bcf_path, vcf_gz_path)) == 0L)
stopifnot(system2("bcftools", c("index", "-f", bcf_path)) == 0L)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
invisible(rduckhts_load(con))
invisible(dbExecute(con, "PRAGMA threads=1"))

query_for <- function(path) {
  sprintf(
    paste0(
      "SELECT count(*) AS rows, sum(POS)::HUGEINT AS position_sum ",
      "FROM read_bcf(%s, region := %s, decompression_threads := 0)"
    ),
    sql_string(path), sql_string(region_spec)
  )
}

expected_rows <- as.double(max(region_ends) - min(region_starts) + 1L)
expected_sum <- expected_rows * (min(region_starts) + max(region_ends)) / 2
raw_region_hits <- as.double(region_count) * region_width

measure <- function(format, path, repetitions = 9L) {
  sql <- query_for(path)
  for (i in seq_len(2L)) DBI::dbGetQuery(con, sql)
  seconds <- numeric(repetitions)
  answer <- NULL
  for (i in seq_len(repetitions)) {
    elapsed <- system.time(answer <- DBI::dbGetQuery(con, sql))[["elapsed"]]
    seconds[[i]] <- elapsed
  }
  stopifnot(as.double(answer$rows[[1]]) == expected_rows)
  stopifnot(as.double(answer$position_sum[[1]]) == expected_sum)
  data.frame(
    format = format,
    repetitions = repetitions,
    unique_output_rows = expected_rows,
    min_seconds = min(seconds),
    median_seconds = stats::median(seconds),
    max_seconds = max(seconds),
    median_rows_per_second = expected_rows / stats::median(seconds),
    stringsAsFactors = FALSE
  )
}

results <- rbind(
  measure("BCF + CSI (bcf_itr_regarray)", bcf_path),
  measure("VCF.gz + TBI (tbx_itr_regarray)", vcf_gz_path)
)

cpu_model <- sub(
  "^model name[[:space:]]*:[[:space:]]*", "",
  grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[[1]]
)
affinity <- command_output("taskset", c("-pc", Sys.getpid()))

metadata <- data.frame(
  item = c(
    "source revision", "run date", "DuckDB version", "htslib version",
    "CPU", "process affinity", "DuckDB threads", "htslib decompression threads",
    "input records per format", "requested intervals", "interval width",
    "interval step", "raw interval-record hits", "unique output records",
    "overlap work factor", "uncompressed VCF bytes", "VCF.gz bytes", "BCF bytes",
    "bgzip version", "tabix version", "bcftools version"
  ),
  value = c(
    source_state,
    as.character(Sys.Date()),
    as.character(DBI::dbGetQuery(con, "SELECT version() AS version")$version[[1]]),
    trimws(readLines("third_party/htslib/VERSION", n = 1L)),
    cpu_model,
    affinity,
    "1",
    "0",
    format(row_count, big.mark = ",", scientific = FALSE),
    as.character(region_count),
    format(region_width, big.mark = ",", scientific = FALSE),
    format(region_step, big.mark = ",", scientific = FALSE),
    format(raw_region_hits, big.mark = ",", scientific = FALSE),
    format(expected_rows, big.mark = ",", scientific = FALSE),
    sprintf("%.3f", raw_region_hits / expected_rows),
    format(file.info(vcf_path)$size, big.mark = ",", scientific = FALSE),
    format(file.info(vcf_gz_path)$size, big.mark = ",", scientific = FALSE),
    format(file.info(bcf_path)$size, big.mark = ",", scientific = FALSE),
    command_output("bgzip", "--version"),
    command_output("tabix", "--version"),
    command_output("bcftools", "--version")
  ),
  stringsAsFactors = FALSE
)

knitr::kable(metadata, col.names = c("Property", "Recorded value"))
invisible(dbDisconnect(con, shutdown = TRUE))
invisible(unlink(fixture_dir, recursive = TRUE))
```

## Results

The denominator is unique VCF records returned after htslib merges the 64
overlapping intervals. `raw interval-record hits` is what independent
per-region scans would encounter before duplicate removal. Both formats are
validated against the exact expected row count and position checksum on every
timed run.

```{r results, echo=FALSE}
display_results <- results
display_results$unique_output_rows <- format(
  display_results$unique_output_rows, big.mark = ",", scientific = FALSE
)
for (column in c("min_seconds", "median_seconds", "max_seconds")) {
  display_results[[column]] <- sprintf("%.3f", display_results[[column]])
}
display_results$median_rows_per_second <- format(
  round(display_results$median_rows_per_second), big.mark = ",", scientific = FALSE
)
names(display_results) <- c(
  "Format and native iterator", "Runs", "Unique rows", "Minimum seconds",
  "Median seconds", "Maximum seconds", "Median unique rows/s"
)
knitr::kable(display_results, align = c("l", "r", "r", "r", "r", "r", "r"))
```

This is a warm-page-cache, one-core iterator/decode measurement. It does not
measure cold storage, remote object-store latency, rich INFO/FORMAT decoding,
or DuckDB materialization of every returned column. It specifically protects
the indexed overlapping-region path changed by the htslib 1.24 integration.
