---
title: "Base mismatch count scaling"
output:
  github_document:
    html_preview: false
    pandoc_args: ["--wrap=preserve", "--columns=400"]
---

This report gives the 1×/2×/4× scaling evidence `STYLE.md` asks for, for
`duckhts_bam_mismatch_counts()`. The function reads alignments once and counts
their aligned bases against the reference by mate, cycle, base quality and
substitution. Its state is meant to be fixed, so the memory verdict is the main
one. No earlier report measures this function; this is its first baseline.

## Workload

The workload is real and public: chr1, the largest contig, of the
1000 Genomes 30× CRAM of HG00403, staged as a CRAM slice (registry artifact
`bam_mismatch_hg00403_chr1_cram`), with every chr1 record of the phased
1000 Genomes panel as the mask (`bam_mismatch_chr1_mask_bcf`) and the
registered GRCh38 FASTA as the reference (`liftover_grch38_fasta`).
`benchmarks/bam_mismatch_stage.R` stages both derived files and checks them
against their registered identities. This report checks them again before it
measures, and it checks the reference against its registered size and MD5.
Those checks read all three files in full, so every measured run reads from a
warm page cache.

There are two growth dimensions, the aligned bases and the mask records, and
three series.

- The `mask` series doubles both together. Its 1× and 2× regions end at the
  25% and 50% quantiles of the alignment start positions of the slice, and its
  4× region is the whole contig. The mask records under a region grow with it.
- The `no mask` series reads the same regions without a mask, so the aligned
  bases grow alone. The difference between the two series is the whole cost
  of the mask at each size.
- The `mask records` series doubles the mask records alone. Its alignments
  are fixed: one two-base region every 2 MiB of the contig. The function reads
  the mask one reference window at a time, and a window starts at the first
  kept alignment that the last window does not hold and is 1 MiB long. A
  region 2 MiB after the last one is therefore outside the last window, and
  each region opens one window of its own. The render derives these windows
  from the first kept alignment of each region, checks that no alignment of a
  region leaves its window or reaches the next one, and counts the mask
  records that the windows hold; that count, not the size of the mask file,
  is the denominator of the series. The windows cover about half of the
  contig. The masks are the same panel records thinned to the positions
  divisible by 4 (`bam_mismatch_chr1_mask_quarter_bcf`), by 2
  (`bam_mismatch_chr1_mask_half_bcf`) and not at all, so 1×, 2× and 4×
  records; a run without a mask gives the cost of the alignments alone.

The cuts and the input denominators of each region are computed at render
time. That is untimed staging, as is the one-time staging of the files.

Every observation is one fresh DuckDB CLI process under `/usr/bin/time`. The
JSON profiler gives the query latency and DuckDB's peak buffer and temporary
bytes, and `/usr/bin/time` gives the peak RSS. The host is shared, so the
render command pins the processes to eight cores and each measured process
runs at scheduling priority −15 (`nice`), which makes other load on those
cores yield to it; the host load at the start is recorded, and the spread of
the repetitions shows what interference remained. `max_temp_directory_size` is
zero, so a spill fails the render; DuckDB's memory limit is left at its
default, which the report records. Each cell has three repetitions at one
thread and at four threads, run in an order that alternates sizes. The
function has one worker, so the thread setting should not change it, and the
four-thread cells show whether that holds for time and memory. Each query
reduces the rows of the function to their number, the counted bases and the
mismatches, so every row is produced and consumed. The R wrapper only builds
this SQL call; it is not timed here.

## Budget, declared before measuring

The overhead is the median peak RSS of three empty runs that start the CLI,
load the extension and scan nothing.

The ceiling is overhead + 3 × the live state. It is the same at every size,
and it is a gate: the render stops if any observation exceeds it or spills.
The live state is:

- the counters: 3 mates × 1,001 cycles × 95 base qualities × 16 substitutions
  × 8 bytes = 36,516,480 bytes;
- one reference window of 1 MiB and its mask of 1 MiB. An alignment that
  spans more reference than that grows both to its span, up to the 64 MiB
  limit; the render checks from the CIGARs that no alignment of this file
  spans more than 1 MiB, so the window stays at 1 MiB;
- htslib's state for the three files: the CRAM container in work with its
  decoded alignments, the reference of that container's span, the indexes and
  the file buffers. 16 MiB is allowed for it. The function sets no decoder
  threads, and for sorted input without them htslib's `cram_get_ref()` loads
  the reference of one slice's span, not the contig.

That is 52.8 MiB, so the ceiling is overhead + 158.5 MiB.

The render also stops if a time exponent of the `mask` or `no mask` series is
above 1.25 at either doubling, or if their 1× run on one thread takes less
than 5 s, which is the least that `STYLE.md` accepts for a timing verdict.
The `mask records` series reads a few million records in a few seconds, so
its 1× run is expected under 5 s: it gets the memory verdict in full and its
times are reported without a verdict, as `STYLE.md` provides.

``` sh
taskset -c 8-15 Rscript -e 'rmarkdown::render("benchmarks/benchmark_bam_mismatch_counts.Rmd")'
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r benchmark, echo=FALSE, message=FALSE}
stopifnot(requireNamespace("duckhtsbench", quietly = TRUE),
          requireNamespace("jsonlite", quietly = TRUE),
          requireNamespace("knitr", quietly = TRUE))
Sys.setenv(DUCKHTSBENCH_REGISTRY = normalizePath(
  "r/duckhtsbench/inst/benchmark_registry.tsv", mustWork = TRUE))
extension <- normalizePath(Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"),
                           mustWork = TRUE)
duckdb_cli <- Sys.getenv("DUCKDB_CLI", Sys.which("duckdb"))
stopifnot(nzchar(duckdb_cli), file.exists("/usr/bin/time"))
duckdb_runtime <- system2(duckdb_cli, "--version", stdout = TRUE)[[1L]]
revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE)
src_tree <- system2("git", c("rev-parse", "HEAD:src"), stdout = TRUE)
if (system2("git", c("diff", "--quiet", "HEAD", "--", "src"), stdout = FALSE, stderr = FALSE) != 0L) {
  stop("Build source must match the recorded HEAD")
}
host_load <- sub(".*load average: ", "", system2("uptime", stdout = TRUE))
quote_literal <- function(value) paste0("'", gsub("'", "''", value, fixed = TRUE), "'")

source("benchmarks/bam_mismatch_stage.R")
slice_id <- "bam_mismatch_hg00403_chr1_cram"
mask_id <- "bam_mismatch_chr1_mask_bcf"
slice_identity <- bam_mismatch_check_staged(slice_id, extension)
mask_identity <- bam_mismatch_check_staged(mask_id, extension)
cram <- duckhtsbench::duckhts_bench_artifact_path(slice_id)
mask <- duckhtsbench::duckhts_bench_artifact_path(mask_id)
# The thinned masks of the mask-records series, with the full mask as 4x.
mask_sizes <- data.frame(scale = c(0L, 1L, 2L, 4L), stringsAsFactors = FALSE,
                         id = c(NA, "bam_mismatch_chr1_mask_quarter_bcf", "bam_mismatch_chr1_mask_half_bcf", mask_id))
mask_sizes$path <- ifelse(is.na(mask_sizes$id), NA, vapply(mask_sizes$id, function(id) {
  if (is.na(id)) NA_character_ else duckhtsbench::duckhts_bench_artifact_path(id)
}, ""))
mask_sizes$records <- vapply(mask_sizes$id, function(id) {
  if (is.na(id)) 0 else bam_mismatch_check_staged(id, extension)$records
}, numeric(1L))
reference <- duckhtsbench::duckhts_bench_artifact_path("liftover_grch38_fasta")
duckhtsbench::duckhts_bench_validate_identity("liftover_grch38_fasta", reference)
# The slice is one whole contig; its length comes from the reference index.
contig <- bam_mismatch_registry_row(slice_id)$fields[["region"]]
reference_index <- utils::read.delim(paste0(reference, ".fai"), header = FALSE)
contig_length <- reference_index[[2L]][reference_index[[1L]] == contig]
stopifnot(length(contig_length) == 1L, identical(bam_mismatch_registry_row(mask_id)$fields[["contig"]], contig))
# The CRAM index lists each slice's start and reference span. htslib's decoder
# holds the reference of the slice in hand, so the widest span bounds that part
# of the live state.
slices <- utils::read.delim(gzfile(paste0(cram, ".crai")), header = FALSE,
                            col.names = c("sequence", "start", "span", "container", "slice", "size"))
widest_slice <- slices[which.max(slices$span), ]
widest_slice_mib <- widest_slice$span / 2^20

# One untimed query in its own CLI process; its rows come back as numbers.
query_untimed <- function(sql) {
  output <- system2(duckdb_cli, c("-unsigned", "-csv"),
                    input = c(sprintf("LOAD %s;", quote_literal(extension)), sql), stdout = TRUE)
  utils::read.csv(text = output, colClasses = "numeric")
}
# The window stays at its 1 MiB floor only while no alignment spans more.
long_spans <- query_untimed(sprintf(paste(
  "SELECT count(*) AS alignments, coalesce(max(cigar_reference_length(CIGAR)), 0) AS widest_span",
  "FROM read_bam(%s, reference := %s) WHERE cigar_reference_length(CIGAR) > %d;"),
  quote_literal(cram), quote_literal(reference), 2^20))
stopifnot(long_spans$alignments == 0)

# Run one SQL body in a fresh CLI process with the given threads and spill
# forbidden; the last statement's single row is returned with latency and memory.
run_fresh <- function(query, threads) {
  scratch <- tempfile("bam-mismatch-scaling-")
  dir.create(scratch)
  on.exit(unlink(scratch, recursive = TRUE), add = TRUE)
  sql_path <- file.path(scratch, "query.sql")
  result_path <- file.path(scratch, "result.csv")
  profile_path <- file.path(scratch, "profile.json")
  time_path <- file.path(scratch, "time.tsv")
  writeLines(c(
    sprintf("SET threads = %d;", threads),
    "SET max_temp_directory_size = '0KiB';",
    sprintf("LOAD %s;", quote_literal(extension)),
    "PRAGMA enable_profiling = 'json';",
    sprintf("PRAGMA profiling_output = %s;", quote_literal(profile_path)),
    sprintf("COPY (%s) TO %s (HEADER);", query, quote_literal(result_path)),
    "PRAGMA disable_profiling;"
  ), sql_path)
  status <- system2("/usr/bin/time", c("-f", shQuote("%M\t%e"), "-o", shQuote(time_path),
                                       "nice", "-n", "-15", shQuote(duckdb_cli), "-unsigned", "-batch", "-bail"),
                    stdin = sql_path, stdout = FALSE, stderr = FALSE)
  if (status != 0L) stop("DuckDB CLI failed (status ", status, ")")
  profile <- jsonlite::fromJSON(profile_path, simplifyVector = TRUE)
  timing <- utils::read.delim(time_path, header = FALSE, col.names = c("max_rss_kib", "process_seconds"))
  data.frame(utils::read.csv(result_path, colClasses = "character"),
             seconds = as.numeric(profile$latency), max_rss_kib = timing$max_rss_kib[[1L]],
             peak_buffer_bytes = as.numeric(profile$system_peak_buffer_memory),
             peak_temp_bytes = as.numeric(profile$system_peak_temp_dir_size))
}

# Untimed staging: the region ends at which a quarter and a half of the
# alignments of the slice have started, and what each region holds.
cuts <- query_untimed(sprintf(
  "SELECT quantile_disc(POS, 0.25)::BIGINT AS quarter, quantile_disc(POS, 0.5)::BIGINT AS half FROM read_bam(%s, reference := %s);",
  quote_literal(cram), quote_literal(reference)))
stopifnot(nrow(cuts) == 1L, cuts$quarter < cuts$half, cuts$half < contig_length)
inputs <- data.frame(scale = c(1L, 2L, 4L), region_end = c(cuts$quarter, cuts$half, contig_length))
inputs$region <- sprintf("%s:1-%s", contig, format(inputs$region_end, scientific = FALSE, trim = TRUE))
region_inputs <- do.call(rbind, lapply(inputs$region, function(region) {
  alignments <- query_untimed(sprintf(
    "SELECT count(*) AS alignments, sum(len(SEQ))::BIGINT AS stored_bases FROM read_bam(%s, reference := %s, region := %s);",
    quote_literal(cram), quote_literal(reference), quote_literal(region)))
  records <- query_untimed(sprintf("SELECT count(*) AS mask_records FROM read_bcf(%s, region := %s);",
                                   quote_literal(mask), quote_literal(region)))
  cbind(alignments, records)
}))
inputs <- cbind(inputs, region_inputs)
# htslib hands the function each alignment as a BAM record: four bits for a
# base and eight for its quality.
inputs$decoded_base_mib <- inputs$stored_bases * 1.5 / 2^20
memory_limit <- system2(duckdb_cli, c("-list", "-noheader"),
                        input = "SELECT current_setting('memory_limit');", stdout = TRUE)[[1L]]

count_query <- function(region, mask_path = NULL) sprintf(paste(
  "SELECT count(*) AS output_rows, sum(bases)::BIGINT AS bases,",
  "sum(bases) FILTER (WHERE reference_base != read_base)::BIGINT AS mismatches",
  "FROM duckhts_bam_mismatch_counts(%s, %s, region := %s%s)"),
  quote_literal(cram), quote_literal(reference), quote_literal(region),
  if (is.null(mask_path)) "" else paste0(", mask := ", quote_literal(mask_path)))
# The fixed alignments of the mask-records series: two bases every 2 MiB,
# which is twice the reference window of the function.
window_bases <- 2^20
targets <- seq(window_bases, contig_length - 2, by = 2 * window_bases)
sparse_region <- paste(sprintf("%s:%d-%d", contig, targets, targets + 1), collapse = ",")
# The windows the function fetches for these regions. An alignment is kept by
# the default filters of the function; the first kept alignment of a region
# starts the region's window. Coordinates are zero-based and half-open.
kept <- query_untimed(sprintf(paste(
  "SELECT POS - 1 AS beg0, POS - 1 + cigar_reference_length(CIGAR) AS end0",
  "FROM read_bam(%s, reference := %s, region := %s)",
  "WHERE (FLAG & 3844) = 0 AND MAPQ >= 20 AND CIGAR != '*' AND len(SEQ) > 0;"),
  quote_literal(cram), quote_literal(reference), quote_literal(sparse_region)))
kept$target <- findInterval(kept$beg0, targets - window_bases)
windows <- data.frame(beg0 = tapply(kept$beg0, kept$target, min), last_end0 = tapply(kept$end0, kept$target, max))
windows$end0 <- pmin(windows$beg0 + window_bases, contig_length)
# One window for each region holds only if no alignment of a region leaves its
# window and the next region starts past it.
stopifnot(nrow(windows) > 1L, all(windows$last_end0 <= windows$end0),
          all(windows$beg0[-1L] >= windows$end0[-nrow(windows)]))
# The function asks the mask for the window with the indel flank on each side.
default_flank <- 5
window_values <- paste(sprintf("(%.0f, %.0f)", pmax(windows$beg0 - default_flank, 0), windows$end0 + default_flank),
                       collapse = ", ")
mask_sizes$window_records <- vapply(mask_sizes$path, function(path) {
  if (is.na(path)) return(0)
  query_untimed(sprintf(paste(
    "SELECT count(*) AS records FROM read_bcf(%s) m JOIN (VALUES %s) w(beg0, end0)",
    "ON m.POS - 1 < w.end0 AND m.POS - 1 + len(m.REF) > w.beg0;"), quote_literal(path), window_values))$records
}, numeric(1L))

empty_runs <- vapply(seq_len(3L), function(run) run_fresh("SELECT 1 AS output_rows", 1L)$max_rss_kib, numeric(1L))
overhead_mib <- median(empty_runs) / 1024
live_state_bytes <- 3 * 1001 * 95 * 16 * 8 + 2 * 2^20 + 16 * 2^20
budget_mib <- overhead_mib + 3 * live_state_bytes / 2^20

observations <- list()
for (repetition in 1:3) {
  order <- if (repetition %% 2L == 1L) 1:3 else 3:1
  for (threads in c(1L, 4L)) {
    for (masked in c(TRUE, FALSE)) {
      for (k in order) {
        observed <- run_fresh(count_query(inputs$region[[k]], if (masked) mask else NULL), threads)
        observations[[length(observations) + 1L]] <- data.frame(
          series = if (masked) "mask" else "no mask", scale = inputs$scale[[k]], region = inputs$region[[k]],
          threads = threads, repetition = repetition, observed, budget_mib = budget_mib,
          stringsAsFactors = FALSE)
      }
    }
    for (k in if (repetition %% 2L == 1L) seq_len(nrow(mask_sizes)) else rev(seq_len(nrow(mask_sizes)))) {
      mask_path <- if (is.na(mask_sizes$path[[k]])) NULL else mask_sizes$path[[k]]
      observed <- run_fresh(count_query(sparse_region, mask_path), threads)
      observations[[length(observations) + 1L]] <- data.frame(
        series = "mask records", scale = mask_sizes$scale[[k]], region = "two bases every 2 MiB",
        threads = threads, repetition = repetition, observed, budget_mib = budget_mib,
        stringsAsFactors = FALSE)
    }
  }
}
runs <- do.call(rbind, observations)
for (column in c("output_rows", "bases", "mismatches")) runs[[column]] <- as.numeric(runs[[column]])
stopifnot(all(runs$peak_temp_bytes == 0), all(runs$bases > 0))
write.csv(runs, "benchmarks/benchmark_bam_mismatch_counts_runs.csv", row.names = FALSE)
over_budget <- runs[runs$max_rss_kib / 1024 > runs$budget_mib, ]
if (nrow(over_budget)) {
  print(over_budget[c("series", "scale", "threads", "max_rss_kib", "budget_mib")])
  stop("observations exceed the declared memory budget", call. = FALSE)
}

cells <- split(runs, list(runs$scale, runs$threads, runs$series), drop = TRUE)
summary <- do.call(rbind, lapply(cells, function(cell) {
  # Every repetition of a cell must count the same bases.
  stopifnot(length(unique(cell$bases)) == 1L, length(unique(cell$mismatches)) == 1L,
            length(unique(cell$output_rows)) == 1L)
  data.frame(series = cell$series[[1L]], scale = cell$scale[[1L]], threads = cell$threads[[1L]],
             bases = cell$bases[[1L]], mismatches = cell$mismatches[[1L]], output_rows = cell$output_rows[[1L]],
             median_seconds = median(cell$seconds), min_seconds = min(cell$seconds),
             max_seconds = max(cell$seconds),
             median_rss_mib = round(median(cell$max_rss_kib) / 1024, 1),
             max_rss_mib = round(max(cell$max_rss_kib) / 1024, 1),
             max_buffer_mib = round(max(cell$peak_buffer_bytes) / 2^20, 1),
             budget_mib = round(cell$budget_mib[[1L]], 1), stringsAsFactors = FALSE)
}))
summary <- summary[order(summary$series, summary$threads, summary$scale), ]
rownames(summary) <- NULL
# The three numbers of a query must be the same at both thread settings.
counted <- c("series", "scale", "bases", "mismatches", "output_rows")
stopifnot(isTRUE(all.equal(summary[summary$threads == 1L, counted], summary[summary$threads == 4L, counted],
                           check.attributes = FALSE)))

# The growth of the two alignment series is measured against the bases
# counted, which the quantile cuts double only approximately.
alignment_series <- summary[summary$series != "mask records", ]
rownames(alignment_series) <- NULL
exponents <- do.call(rbind, lapply(split(alignment_series, list(alignment_series$threads, alignment_series$series),
                                         drop = TRUE), function(s) {
  s <- s[order(s$scale), ]
  growth <- function(low, high) {
    log(s$median_seconds[[high]] / s$median_seconds[[low]]) / log(s$bases[[high]] / s$bases[[low]])
  }
  data.frame(series = s$series[[1L]], threads = s$threads[[1L]], one_x_seconds = s$median_seconds[[1L]],
             bases_ratio_2x = s$bases[[2L]] / s$bases[[1L]], bases_ratio_4x = s$bases[[3L]] / s$bases[[2L]],
             time_exponent_2x = growth(1L, 2L), time_exponent_4x = growth(2L, 3L),
             rss_ratio_2x = s$median_rss_mib[[2L]] / s$median_rss_mib[[1L]],
             rss_ratio_4x = s$median_rss_mib[[3L]] / s$median_rss_mib[[1L]], stringsAsFactors = FALSE)
}))
rownames(exponents) <- NULL
# The mask-records series: its growth is measured against the records, for
# the whole run and for the mask's own cost, the run without a mask taken off.
mask_series <- summary[summary$series == "mask records", ]
mask_series$mask_records <- mask_sizes$window_records[match(mask_series$scale, mask_sizes$scale)]
mask_series <- mask_series[order(mask_series$threads, mask_series$scale), ]
rownames(mask_series) <- NULL
mask_exponents <- do.call(rbind, lapply(split(mask_series, mask_series$threads), function(s) {
  base_seconds <- s$median_seconds[s$scale == 0L]
  s <- s[s$scale > 0L, ]
  mask_seconds <- s$median_seconds - base_seconds
  growth <- function(seconds, low, high) {
    log(seconds[[high]] / seconds[[low]]) / log(s$mask_records[[high]] / s$mask_records[[low]])
  }
  data.frame(threads = s$threads[[1L]], no_mask_seconds = base_seconds, one_x_seconds = s$median_seconds[[1L]],
             records_ratio_2x = s$mask_records[[2L]] / s$mask_records[[1L]],
             records_ratio_4x = s$mask_records[[3L]] / s$mask_records[[2L]],
             time_exponent_2x = growth(s$median_seconds, 1L, 2L), time_exponent_4x = growth(s$median_seconds, 2L, 3L),
             mask_cost_exponent_2x = growth(mask_seconds, 1L, 2L), mask_cost_exponent_4x = growth(mask_seconds, 2L, 3L),
             rss_ratio_4x = s$median_rss_mib[[3L]] / s$median_rss_mib[[1L]], stringsAsFactors = FALSE)
}))
rownames(mask_exponents) <- NULL
mask_full <- mask_series[mask_series$threads == 1L & mask_series$scale == 4L, ]
mask_none <- mask_series[mask_series$threads == 1L & mask_series$scale == 0L, ]
records_per_second <- mask_full$mask_records / (mask_full$median_seconds - mask_none$median_seconds)
one_thread <- alignment_series[alignment_series$threads == 1L, ]
masked_one <- one_thread[one_thread$series == "mask", ]
unmasked_one <- one_thread[one_thread$series == "no mask", ]
stopifnot(identical(masked_one$scale, unmasked_one$scale), identical(masked_one$scale, inputs$scale),
          all(one_thread$median_seconds[one_thread$scale == 1L] >= 5),
          all(exponents$time_exponent_2x <= 1.25), all(exponents$time_exponent_4x <= 1.25))
mask_cost <- data.frame(
  scale = masked_one$scale, mask_records = inputs$mask_records,
  seconds_with_mask = masked_one$median_seconds, seconds_without_mask = unmasked_one$median_seconds,
  mask_seconds = masked_one$median_seconds - unmasked_one$median_seconds,
  bases_left_out = unmasked_one$bases - masked_one$bases,
  rss_mib_with_mask = masked_one$median_rss_mib, rss_mib_without_mask = unmasked_one$median_rss_mib)
largest <- masked_one[masked_one$scale == 4L, ]
megabases_per_second <- largest$bases / largest$median_seconds / 1e6

# Untimed: the counts of the whole contig by base quality, with the mask.
by_quality <- query_untimed(sprintf(paste(
  "SELECT base_quality, sum(bases)::BIGINT AS bases,",
  "coalesce(sum(bases) FILTER (WHERE reference_base != read_base), 0)::BIGINT AS mismatches",
  "FROM duckhts_bam_mismatch_counts(%s, %s, region := %s, mask := %s) GROUP BY base_quality ORDER BY base_quality;"),
  quote_literal(cram), quote_literal(reference), quote_literal(contig), quote_literal(mask)))
stopifnot(sum(by_quality$bases) == largest$bases, sum(by_quality$mismatches) == largest$mismatches)
by_quality$mismatch_rate <- by_quality$mismatches / by_quality$bases
by_quality$share_of_bases <- by_quality$bases / sum(by_quality$bases)
by_quality$share_of_mismatches <- by_quality$mismatches / sum(by_quality$mismatches)
```

Source revision: `r revision`; `src` tree `r src_tree`. DuckDB runtime: `r duckdb_runtime` (CLI). Extension SHA-256: `r strsplit(system2("sha256sum", extension, stdout = TRUE), " ", fixed = TRUE)[[1L]][[1L]]`. Host load at start: `r host_load`. DuckDB memory limit: `r memory_limit` (the default); temporary directory limit: 0.

Staged inputs: the slice has `r format(slice_identity$alignments, big.mark = ",")` alignments (position sum `r format(slice_identity$position_sum, scientific = FALSE)`, `r format(slice_identity$base_sum, scientific = FALSE, big.mark = ",")` stored bases) in `r format(nrow(slices), big.mark = ",")` CRAM slices, of which the widest starts at `r format(widest_slice$start, big.mark = ",")` and spans `r format(widest_slice$span, big.mark = ",")` reference bases; the mask has `r format(mask_identity$records, big.mark = ",")` records, its half `r format(mask_sizes$records[mask_sizes$scale == 2L], big.mark = ",")` and its quarter `r format(mask_sizes$records[mask_sizes$scale == 1L], big.mark = ",")`.

Overhead: `r round(overhead_mib, 1)` MiB. Ceiling: `r round(budget_mib, 1)` MiB.

## Inputs of each region

`alignments` and `stored_bases` are what the region selects before the
function's own filters; `decoded_base_mib` is their bases and qualities as
htslib hands them over, at 1.5 bytes for each stored base. `mask_records` are
the mask records that the region selects.

```{r inputs, echo=FALSE}
knitr::kable(inputs[c("scale", "region", "alignments", "stored_bases", "decoded_base_mib", "mask_records")],
             digits = 1, format.args = list(big.mark = ","))
```

## Measurements

`bases`, `mismatches` and `output_rows` are the result of the query. Times are
the query latency of three repetitions. The two alignment series:

```{r summary, echo=FALSE}
knitr::kable(alignment_series, digits = 3, format.args = list(big.mark = ","))
```

Growth of the median time against the bases counted (an exponent of 1 is linear), and peak-RSS ratios to 1×:

```{r exponents, echo=FALSE}
knitr::kable(exponents, digits = 3)
```

The cost of the mask on one thread, as the difference of the two series:

```{r mask-cost, echo=FALSE}
knitr::kable(mask_cost, digits = 3, format.args = list(big.mark = ","))
```

The mask-records series, at fixed alignments. Its `r nrow(windows)` windows
hold `r format(sum(windows$end0 - windows$beg0), big.mark = ",")` reference
bases, and `mask_records` are the records of each mask that those windows
hold, out of `r format(mask_identity$records, big.mark = ",")` in the full
mask file. Scale 0 is the same regions without a mask:

```{r mask-series, echo=FALSE}
knitr::kable(mask_series[c("scale", "threads", "mask_records", "bases", "mismatches", "output_rows", "median_seconds",
                           "min_seconds", "max_seconds", "median_rss_mib", "max_rss_mib", "budget_mib")],
             digits = 3, format.args = list(big.mark = ","))
```

Its growth against the mask records, for the whole run and for the mask's own cost (the run without a mask taken off), without a timing verdict:

```{r mask-exponents, echo=FALSE}
knitr::kable(mask_exponents, digits = 3)
```

## Findings

Every observation is within the ceiling and none spills. The 1× runs of the
two alignment series on one thread take at least 5 s and none of their time
exponents is above 1.25. The render enforces all of this. It also stops
unless the number of rows, the counted bases and the mismatches of a region
are the same in every repetition and at both thread settings. This report
does not check the counts themselves; the SQL tests do, against counts
derived by hand and against the same counts made independently in SQL.

Mask records alone: reading the `r format(mask_full$mask_records, big.mark = ",")`
records that the `r nrow(windows)` windows hold in the full mask costs
`r round(mask_full$median_seconds - mask_none$median_seconds, 2)` s
on one thread (`r round(records_per_second / 1e6, 2)` million records a
second) on top of the `r round(mask_none$median_seconds, 2)` s of the fixed
alignments, and peak RSS is `r mask_none$median_rss_mib` MiB without a mask
and `r mask_full$median_rss_mib` MiB with the full mask: the mask window is
1 MiB whatever the mask holds. The whole-run time exponents at one thread are
`r round(mask_exponents$time_exponent_2x[mask_exponents$threads == 1L], 2)` and
`r round(mask_exponents$time_exponent_4x[mask_exponents$threads == 1L], 2)`, and
the mask's own cost grows with exponents
`r round(mask_exponents$mask_cost_exponent_2x[mask_exponents$threads == 1L], 2)` and
`r round(mask_exponents$mask_cost_exponent_4x[mask_exponents$threads == 1L], 2)`.
These times carry no verdict, because the 1× run takes
`r round(mask_exponents$one_x_seconds[mask_exponents$threads == 1L], 2)` s.

With the mask on one thread, the whole contig takes
`r round(largest$median_seconds, 1)` s for `r format(largest$bases, big.mark = ",")`
counted bases, which is `r round(megabases_per_second)` million counted bases
a second, at `r largest$median_rss_mib` MiB peak RSS. The 1× region takes
`r round(masked_one$median_seconds[[1L]], 1)` s at `r masked_one$median_rss_mib[[1L]]` MiB.
Peak RSS at 4× is `r round(exponents$rss_ratio_4x[exponents$threads == 1L & exponents$series == "mask"], 2)`
times the 1× value, and the largest peak RSS of any run is
`r max(summary$max_rss_mib)` MiB against the ceiling of `r round(budget_mib, 1)` MiB.

The difference between the sizes in peak RSS is htslib's, not the function's.
The decoder holds the reference of the CRAM slice in hand, and the widest
slice of this file spans `r format(widest_slice$span, big.mark = ",")` bases
(`r round(widest_slice_mib, 1)` MiB of reference) from position
`r format(widest_slice$start, big.mark = ",")`, over the sparse region of the
contig. Only the whole-contig runs read that slice. That one span is above the
16 MiB the budget allows for all of htslib's state, so the allowance was too
small for this file; the ceiling, which is what gates the render, held with a
wide margin. The step depends on how the file's slices are laid out, not on
the number of alignments.

The operating scale this report supports is what it measures: one contig of
one 30× short-read sample, read from CRAM, with a mask of
`r format(mask_identity$records, big.mark = ",")` records. It does not measure
a whole genome in one call, BAM input, unsorted input or long reads.

## What the counts say

The whole contig with the mask, by base quality, from one untimed run:

```{r quality, echo=FALSE}
knitr::kable(by_quality, digits = c(0, 0, 0, 6, 4, 4), format.args = list(big.mark = ","))
```

The mask leaves out every position of a panel record, and the sample is one of
the panel's samples. So these mismatches are errors of the read, the library
or the alignment, together with the variants of the sample that the panel
does not hold.
