---
title: "DuckHTS cgranges Streaming-Provider Benchmark"
output:
  github_document:
    html_preview: true
---

<!-- benchmark_cgranges.md is generated from benchmark_cgranges.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This benchmark measures the cgranges path that matters for provider streaming:

- build one session-scoped cgranges target index with
  `duckhts_cgranges_from_query(...)` / `duckhts_cgranges_index(...)`
- stream query intervals from `read_bed(...)`
- filter query rows with the vectorized scalar predicate
  `duckhts_cgranges_has_overlap(...)`
- annotate query rows with vectorized scalar counts via
  `duckhts_cgranges_count_overlaps(...)`
- expand streaming provider rows to one row per hit with
  `duckhts_cgranges_overlaps_list(...)` plus `UNNEST(...)`
- compare overlap-existence against `bedtk flt`
- compare overlap-existence, overlap counts, and one-row-per-hit expansion
  against `bedtools intersect -u`, `bedtools intersect -c`, and
  `bedtools intersect -wa -wb` when `bedtools` is installed

The old benchmark shape generated thousands of `UNION ALL` calls to
`duckhts_cgranges_overlaps(...)`, or issued one SQL statement per probe. That
measured bind/SQL dispatch overhead rather than the desired streaming provider
path. The current scripts avoid that pattern.

# Run

Synthetic, deterministic default used for the rendered table:

```sh
python3 scripts/cgranges_benchmark.py \
  --extension build/release/duckhts.duckdb_extension \
  --bedtk .sync/bedtk/bedtk \
  --bedtools bedtools \
  --subjects 50000 \
  --queries 5000 \
  --passes 3 \
  --out-dir .tmp/cgranges_benchmark
```

Real DuckBedQC BED files can be benchmarked with:

```sh
python3 scripts/cgranges_benchmark_real.py \
  --extension build/release/duckhts.duckdb_extension \
  --passes 1 \
  --out-dir .tmp/cgranges_benchmark_real
```

For a shell/CLI-only smoke path that still avoids generated bulk-query SQL:

```sh
scripts/cgranges_benchmark_cli.sh
```

```{r setup, include = FALSE}
library(knitr)

repo_root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", unset = getwd()), mustWork = TRUE)
out_dir <- file.path(repo_root, ".tmp", "cgranges_benchmark")
summary_csv <- file.path(out_dir, "summary.csv")
metadata_json <- file.path(out_dir, "metadata.json")

subjects <- as.integer(Sys.getenv("CGRANGES_BENCH_SUBJECTS", unset = "50000"))
queries <- as.integer(Sys.getenv("CGRANGES_BENCH_QUERIES", unset = "5000"))
passes <- as.integer(Sys.getenv("CGRANGES_BENCH_PASSES", unset = "3"))
force <- identical(Sys.getenv("CGRANGES_BENCH_FORCE", unset = "0"), "1")

expected_cols <- c(
  "tool", "variant", "subject_intervals", "query_intervals", "passes",
  "build_index_sec", "query_total_sec", "query_pass_1_sec", "total_elapsed_sec",
  "peak_rss_mb", "matched_query_intervals", "total_hits"
)
needs_run <- force || !file.exists(summary_csv) || !file.exists(metadata_json)
if (!needs_run && file.exists(summary_csv)) {
  existing_header <- names(utils::read.csv(summary_csv, nrows = 0, check.names = FALSE))
  needs_run <- !all(expected_cols %in% existing_header)
}

if (needs_run) {
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  cmd <- c(
    file.path(repo_root, "scripts", "cgranges_benchmark.py"),
    "--extension", file.path(repo_root, "build", "release", "duckhts.duckdb_extension"),
    "--bedtk", file.path(repo_root, ".sync", "bedtk", "bedtk"),
    "--bedtools", Sys.getenv("BEDTOOLS", unset = "bedtools"),
    "--subjects", as.character(subjects),
    "--queries", as.character(queries),
    "--passes", as.character(passes),
    "--out-dir", out_dir
  )
  out <- system2("python3", cmd, stdout = TRUE, stderr = TRUE)
  status <- attr(out, "status")
  if (!is.null(status) && status != 0) {
    stop(paste(out, collapse = "\n"))
  }
}

bench_raw <- utils::read.csv(summary_csv, check.names = FALSE, na.strings = c("", "NA", "None"))
bench <- bench_raw
bench$time_per_query_ms <- (bench$query_total_sec / (bench$query_intervals * bench$passes)) * 1000
num_cols <- c("build_index_sec", "query_total_sec", "query_pass_1_sec", "total_elapsed_sec", "peak_rss_mb", "time_per_query_ms")
for (col in intersect(num_cols, names(bench))) bench[[col]] <- round(bench[[col]], if (col == "time_per_query_ms") 4 else 3)
bench$peak_rss_mb <- round(bench$peak_rss_mb, 1)

filter_rows <- bench_raw[bench_raw$variant %in% c("scalar_filter", "flt", "intersect_u"), ]
filter_agree <- length(unique(filter_rows$matched_query_intervals)) == 1
count_rows <- bench_raw[bench_raw$variant %in% c("scalar_count", "scalar_expand", "intersect_c", "intersect_wa_wb") & !is.na(bench_raw$total_hits), ]
count_agree <- nrow(count_rows) <= 1 || length(unique(count_rows$total_hits)) == 1
```

# Configuration

```{r, echo=FALSE}
knitr::kable(
  data.frame(
    parameter = c("dataset", "subjects", "queries", "passes", "bedtk", "bedtools"),
    value = c(
      "synthetic deterministic BED4",
      subjects,
      queries,
      passes,
      if (file.exists(file.path(repo_root, ".sync", "bedtk", "bedtk"))) "available" else "missing",
      if (nzchar(Sys.which(Sys.getenv("BEDTOOLS", unset = "bedtools")))) "available" else "missing"
    )
  )
)
```

# Results

```{r, echo=FALSE}
knitr::kable(bench)
```

# Semantic checks

```{r, echo=FALSE}
knitr::kable(
  data.frame(
    check = c(
      "matched query intervals agree for DuckHTS scalar_filter, bedtk flt, and bedtools -u when present",
      "overlap counts agree for DuckHTS scalar_count, scalar_expand, bedtools -c, and bedtools -wa -wb when present"
    ),
    result = c(filter_agree, count_agree)
  )
)
```

# Notes

- `duckhts:scalar_filter` is the row-preserving provider-streaming predicate
  path.
- `duckhts:scalar_count` computes one cgranges overlap count per streamed query
  row and is comparable to `bedtools intersect -c`.
- `duckhts:scalar_expand` uses `duckhts_cgranges_overlaps_list(...)` plus
  `UNNEST(...)` to emit one row per hit while preserving streamed provider rows;
  it is comparable to `bedtools intersect -wa -wb`.
- `bedtk flt` and `bedtools intersect -u` are overlap-existence comparators;
  they do not report total hit counts.
- DuckHTS reports explicit target-index build time because the SQL API exposes
  index construction as a session-scoped operation. The external tools build any
  internal structures inside their command runtime.
- The rendered table is intentionally a modest synthetic run. Use
  `scripts/cgranges_benchmark_real.py` for the larger DuckBedQC WGS/exome BED
  workload, and optional `--limit-subjects` / `--limit-queries` for quick smoke
  runs.
