---
title: "BCF appender contig-parallel benchmark"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

Historical experiment: `read_bcf_appender` is removed from the public extension.
The recorded measurements are retained, not relabeled as canonical-reader
performance. Reproduction requires an explicitly selected pre-retirement binary.

This report measures `read_bcf_appender(...)` after making `region_threads`
partition work by primary contig. Each contig keeps its complete interval set
in one htslib 1.24 multi-region iterator, so repeated and overlapping intervals
produce the same target schema and row multiset at every thread count.

The previous parallel implementation scanned each requested interval
independently, emitted duplicate rows, and added `duckhts_region_idx`. Its row
denominator and write schema therefore differ from this workload. There is no
valid before/after throughput comparison. The nearest checked-in measurement
is [`benchmark_multi_region_readers.md`](benchmark_multi_region_readers.md),
which measures native iterator decoding without appender materialization.

## Reproduction

Build an extension revision containing the experiment, then render with the process
allowed on all CPUs named below. The script pins the serial run to one CPU and
the four-job run to four CPUs.

```bash
DUCKHTS_APPENDER_EXTENSION=/path/to/old/duckhts.duckdb_extension \
DUCKHTS_APPENDER_REVISION=408fbcb5bc7c \
DUCKHTS_BENCH_SERIAL_CPU=16 \
DUCKHTS_BENCH_PARALLEL_CPUS=16-19 \
taskset -c 16-19 Rscript -e "rmarkdown::render('benchmarks/benchmark_bcf_appender_contigs.Rmd', output_format = 'github_document')"
```

The fixture contains four contigs with consecutive SNVs. Each contig receives
eight 50,000-base intervals at a 25,000-base step. Fixture creation, BCF
conversion, indexing, connection setup, and warm-up are outside the timed
interval. The timed `seconds` value covers target-table replacement, contig-job
planning, indexed reads, narrow-row materialization through DuckDB appenders,
and commit.

```{r benchmark, echo=FALSE}
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
})

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

command_output <- function(command, args = character()) {
  paste(system2(command, args, stdout = TRUE, stderr = TRUE), collapse = " ")
}

set_affinity <- function(cpus) {
  output <- system2(
    "taskset", c("-pc", cpus, Sys.getpid()), stdout = TRUE, stderr = TRUE
  )
  status <- attr(output, "status")
  stopifnot(is.null(status) || identical(status, 0L))
  cpus
}

extension <- Sys.getenv("DUCKHTS_APPENDER_EXTENSION")
source_state <- Sys.getenv("DUCKHTS_APPENDER_REVISION")
stopifnot(nzchar(extension), file.exists(extension), nzchar(source_state))

serial_cpu <- Sys.getenv("DUCKHTS_BENCH_SERIAL_CPU", "16")
parallel_cpus <- Sys.getenv("DUCKHTS_BENCH_PARALLEL_CPUS", "16-19")
contig_count <- 4L
rows_per_contig <- as.integer(Sys.getenv("DUCKHTS_APPENDER_ROWS_PER_CONTIG", "250000"))
intervals_per_contig <- 8L
interval_width <- 50000L
interval_step <- 25000L
starts <- 1L + (seq_len(intervals_per_contig) - 1L) * interval_step
ends <- starts + interval_width - 1L
stopifnot(max(ends) <= rows_per_contig)

regions <- unlist(lapply(seq_len(contig_count), function(contig) {
  sprintf("%d:%d-%d", contig, starts, ends)
}), use.names = FALSE)
region_spec <- paste(regions, collapse = ",")
unique_rows_per_contig <- as.double(max(ends) - min(starts) + 1L)
expected_rows <- unique_rows_per_contig * contig_count
expected_position_sum <- contig_count *
  unique_rows_per_contig * (min(starts) + max(ends)) / 2
raw_interval_hits <- as.double(contig_count) * intervals_per_contig * interval_width

fixture_dir <- tempfile("duckhts-appender-contigs-")
invisible(dir.create(fixture_dir))
vcf_path <- file.path(fixture_dir, "variants.vcf")
bcf_path <- file.path(fixture_dir, "variants.bcf")

out <- file(vcf_path, open = "wt")
writeLines(c(
  "##fileformat=VCFv4.3",
  sprintf("##contig=<ID=%d,length=%d>", seq_len(contig_count), rows_per_contig),
  "#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO"
), out)
chunk_size <- 100000L
for (contig in seq_len(contig_count)) {
  for (first in seq.int(1L, rows_per_contig, by = chunk_size)) {
    last <- min(rows_per_contig, first + chunk_size - 1L)
    pos <- seq.int(first, last)
    writeLines(sprintf("%d\t%d\t.\tA\tC\t.\tPASS\t.", contig, pos), out)
  }
}
close(out)

stopifnot(system2(
  "bcftools", c("view", "--no-version", "-Ob", "-o", bcf_path, vcf_path)
) == 0L)
stopifnot(system2("bcftools", c("index", "-f", bcf_path)) == 0L)

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
invisible(dbExecute(con, sprintf("LOAD %s", sql_string(normalizePath(extension)))))
stopifnot(dbGetQuery(con, paste(
  "SELECT count(*) AS n FROM duckdb_functions()",
  "WHERE function_name = 'read_bcf_appender'"))$n[[1]] == 1)
invisible(dbExecute(con, "PRAGMA threads=1"))

appender_sql <- function(target, region_threads) {
  sprintf(
    paste0(
      "SELECT rows_written, seconds FROM read_bcf_appender(",
      "%s, %s, region := %s, tidy_format := false, overwrite := true, ",
      "include_file_offset := true, region_threads := %d, ",
      "decompression_threads := 0)"
    ),
    sql_string(bcf_path), sql_string(target), sql_string(region_spec), region_threads
  )
}

validate_target <- function(target) {
  answer <- DBI::dbGetQuery(
    con,
    sprintf(
      paste0(
        "SELECT count(*) AS rows, sum(POS)::HUGEINT AS position_sum, ",
        "count(DISTINCT FILE_OFFSET) AS offsets FROM %s"
      ),
      target
    )
  )
  stopifnot(as.double(answer$rows[[1]]) == expected_rows)
  stopifnot(as.double(answer$position_sum[[1]]) == expected_position_sum)
  stopifnot(as.double(answer$offsets[[1]]) == expected_rows)
}

measure <- function(mode, target, region_threads, cpus, repetitions = 15L) {
  affinity <- set_affinity(cpus)
  sql <- appender_sql(target, region_threads)
  DBI::dbGetQuery(con, sql)
  validate_target(target)
  seconds <- numeric(repetitions)
  rows <- numeric(repetitions)
  for (i in seq_len(repetitions)) {
    answer <- DBI::dbGetQuery(con, sql)
    seconds[[i]] <- answer$seconds[[1]]
    rows[[i]] <- answer$rows_written[[1]]
  }
  validate_target(target)
  stopifnot(all(rows == expected_rows))
  data.frame(
    mode = mode,
    region_threads = region_threads,
    affinity = affinity,
    repetitions = repetitions,
    unique_output_rows = expected_rows,
    median_seconds = stats::median(seconds),
    median_rows_per_second = expected_rows / stats::median(seconds),
    stringsAsFactors = FALSE
  )
}

serial <- measure("one complete multi-region iterator", "appender_serial", 1L, serial_cpu)
parallel <- measure("one iterator per primary contig", "appender_parallel", 4L, parallel_cpus)
results <- rbind(serial, parallel)

delta <- DBI::dbGetQuery(
  con,
  paste0(
    "SELECT count(*) AS n FROM (",
    "(SELECT * FROM appender_serial EXCEPT ALL SELECT * FROM appender_parallel) ",
    "UNION ALL ",
    "(SELECT * FROM appender_parallel EXCEPT ALL SELECT * FROM appender_serial))"
  )
)$n[[1]]
stopifnot(delta == 0)

cpu_model <- sub(
  "^model name[[:space:]]*:[[:space:]]*", "",
  grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[[1]]
)
bcftools_version <- system2("bcftools", "--version", stdout = TRUE, stderr = TRUE)
metadata <- data.frame(
  item = c(
    "source revision", "run date", "DuckDB version", "htslib version", "CPU",
    "serial CPU", "parallel CPUs", "DuckDB threads", "htslib decompression threads per job",
    "input BCF records", "primary contigs", "requested intervals", "interval width",
    "interval step", "raw interval-record hits", "unique output records",
    "overlap work factor", "uncompressed VCF bytes", "BCF bytes", "bcftools version"
  ),
  value = c(
    source_state,
    as.character(Sys.Date()),
    as.character(DBI::dbGetQuery(con, "SELECT version() AS version")$version[[1]]),
    trimws(readLines("third_party/htslib/VERSION", n = 1L)),
    cpu_model,
    serial_cpu,
    parallel_cpus,
    "1",
    "0",
    format(as.double(contig_count) * rows_per_contig, big.mark = ",", scientific = FALSE),
    as.character(contig_count),
    as.character(length(regions)),
    format(interval_width, big.mark = ",", scientific = FALSE),
    format(interval_step, big.mark = ",", scientific = FALSE),
    format(raw_interval_hits, big.mark = ",", scientific = FALSE),
    format(expected_rows, big.mark = ",", scientific = FALSE),
    sprintf("%.3f", raw_interval_hits / expected_rows),
    format(file.info(vcf_path)$size, big.mark = ",", scientific = FALSE),
    format(file.info(bcf_path)$size, big.mark = ",", scientific = FALSE),
    paste(utils::head(bcftools_version, 2L), collapse = "; ")
  ),
  stringsAsFactors = FALSE
)

knitr::kable(metadata, col.names = c("Property", "Recorded value"))
```

## Results

The denominator is the 900,000 unique target rows after htslib merges the
overlapping intervals. Both runs materialize `FILE_OFFSET`, validate the exact
row count, position checksum, and distinct record offsets, and finish with a
bidirectional `EXCEPT ALL` comparison of zero rows.

```{r results, echo=FALSE}
display_results <- results
display_results$unique_output_rows <- format(
  display_results$unique_output_rows, big.mark = ",", scientific = FALSE
)
display_results$median_seconds <- sprintf("%.3f", display_results$median_seconds)
display_results$median_rows_per_second <- format(
  round(display_results$median_rows_per_second), big.mark = ",", scientific = FALSE
)
names(display_results) <- c(
  "Scheduling", "region_threads", "Recorded affinity", "Runs", "Unique rows",
  "Median seconds", "Median unique rows/s"
)
knitr::kable(display_results, align = c("l", "r", "l", "r", "r", "r", "r"))
```

This is a warm-page-cache local-BCF materialization measurement. It does not
measure cold storage, remote object-store latency, rich INFO/FORMAT decoding,
or full `read_bcf(...)` output. The parallel result measures four independent
contig iterators and four DuckDB appenders; a request confined to one contig
correctly remains a one-iterator workload.

```{r cleanup, include=FALSE}
invisible(dbDisconnect(con, shutdown = TRUE))
invisible(unlink(fixture_dir, recursive = TRUE))
```
