---
title: "DuckHTS Normalization Benchmark"
output: github_document
---

<!-- benchmark_norm.md is generated from benchmark_norm.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This benchmark:

- times DuckHTS `duckhts_bcftools_norm(...)` against installed `bcftools norm`
- compares normalized outputs after grouping identical normalized alleles
- covers both site-preserving normalization and `split_multiallelic := TRUE`

# Run

`make bench-norm`

Useful overrides:

- `NORM_BENCH_ROWS`: synthetic VCF records, default `100000`
- `NORM_BENCH_RUNS`: timed repeats, default `3`
- `BCFTOOLS_BIN`: optional override for the bcftools executable
- `NORM_REAL_VCF`, `NORM_REAL_REGION`, `NORM_REAL_FASTA`: optional real-callset conformance case
- `NORM_REAL_SPLIT`: `1` to run the optional real-callset case in split mode too
- `NORM_REAL_LABEL`: optional label for the real-callset section

For a shell-first comparison, see `bash scripts/norm_conformance.sh ...`.

```{r setup, include = FALSE}
library(DBI)
library(duckdb)

if (!requireNamespace("RBCFTools", quietly = TRUE)) {
  stop("RBCFTools is required for bcftools norm benchmarking", call. = FALSE)
}

RBCFTools::setup_hts_env()

bcftools_bin <- Sys.getenv("BCFTOOLS_BIN", unset = RBCFTools::bcftools_path())
if (!file.exists(bcftools_bin)) stop("bcftools binary not found: ", bcftools_bin, call. = FALSE)

bench_rows <- as.integer(Sys.getenv("NORM_BENCH_ROWS", unset = "100000"))
bench_runs <- as.integer(Sys.getenv("NORM_BENCH_RUNS", unset = "3"))
real_input_path <- Sys.getenv("NORM_REAL_VCF", unset = "")
real_region <- Sys.getenv("NORM_REAL_REGION", unset = "")
real_fasta <- Sys.getenv("NORM_REAL_FASTA", unset = "")
real_split <- identical(Sys.getenv("NORM_REAL_SPLIT", unset = "0"), "1")
real_label <- Sys.getenv("NORM_REAL_LABEL", unset = "real_callset")

if (is.na(bench_rows) || bench_rows < 1L) stop("NORM_BENCH_ROWS must be positive")
if (is.na(bench_runs) || bench_runs < 1L) stop("NORM_BENCH_RUNS must be positive")

drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
con <- dbConnect(drv, dbdir = ":memory:")

ext_path <- normalizePath("build/release/duckhts.duckdb_extension", mustWork = TRUE)
DBI::dbExecute(con, sprintf("LOAD '%s';", gsub("\\\\", "/", ext_path)))
DBI::dbExecute(con, "PRAGMA threads=4;")
```

```{r helpers, include = FALSE}
shell_quote <- function(x) shQuote(x, type = "sh")

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

run_cmd <- function(cmd) {
  status <- system(cmd, ignore.stdout = FALSE, ignore.stderr = FALSE)
  if (!identical(status, 0L)) stop("command failed: ", cmd, call. = FALSE)
}

run_cmd_capture <- function(cmd) {
  out <- system(cmd, intern = TRUE, ignore.stderr = FALSE)
  status <- attr(out, "status")
  if (!is.null(status) && !identical(status, 0L)) stop("command failed: ", cmd, call. = FALSE)
  out
}

flatten_duck_alt_sql <- paste0(
  "NULLIF(CASE ",
  "WHEN typeof(alt_normed) = 'VARCHAR[]' ",
  "THEN replace(replace(replace(alt_normed::VARCHAR, '[', ''), ']', ''), ', ', ',') ",
  "ELSE alt_normed::VARCHAR END, '')"
)

write_norm_fixture <- function(dir_path, n_rows) {
  dir.create(dir_path, recursive = TRUE, showWarnings = FALSE)

  fasta_path <- file.path(dir_path, "norm_ref.fa")
  vcf_path <- file.path(dir_path, "norm_input.vcf")
  vcf_gz <- paste0(vcf_path, ".gz")

  writeLines(c(
    ">chrS",
    "GTTTTCT"
  ), fasta_path)

  header_lines <- c(
    "##fileformat=VCFv4.2",
    "##contig=<ID=chrS,length=7>",
    "#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO"
  )

  variants <- list(
    c("T", "TT"),
    c("TT", "T"),
    c("T", "TTT"),
    c("T", "TT,TTT"),
    c("T", ".")
  )

  write_chunked <- function(path, header, n, make_line, chunk_size = 50000L) {
    con_out <- file(path, open = "wt")
    on.exit(close(con_out), add = TRUE)
    writeLines(header, con_out)
    for (start in seq.int(1L, n, by = chunk_size)) {
      end <- min(n, start + chunk_size - 1L)
      idx <- seq.int(start, end)
      writeLines(vapply(idx, make_line, character(1)), con_out)
    }
  }

  write_chunked(vcf_path, header_lines, n_rows, function(i) {
    allele <- variants[[(i - 1L) %% length(variants) + 1L]]
    paste0("chrS\t2\tbench_", i, "\t", allele[[1]], "\t", allele[[2]], "\t.\tPASS\t.")
  })

  DBI::dbGetQuery(
    con,
    sprintf(
      "SELECT * FROM fasta_index(%s, index_path := %s)",
      sql_string(fasta_path),
      sql_string(paste0(fasta_path, ".fai"))
    )
  )

  run_cmd(sprintf(
    "%s view -Oz -o %s %s",
    shell_quote(bcftools_bin),
    shell_quote(vcf_gz),
    shell_quote(vcf_path)
  ))
  run_cmd(sprintf("%s index -f -t %s", shell_quote(bcftools_bin), shell_quote(vcf_gz)))

  list(
    input_path = vcf_gz,
    fasta_path = fasta_path,
    n_rows = n_rows
  )
}

prepare_real_fixture <- function(input_path, fasta_path, region = "") {
  if (!file.exists(fasta_path)) stop("FASTA not found: ", fasta_path, call. = FALSE)

  out_dir <- file.path(tempdir(), "duckhts_norm_real")
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  input_local <- file.path(out_dir, "real_input.vcf.gz")

  region_arg <- if (nzchar(region)) sprintf("-r %s ", shell_quote(region)) else ""
  run_cmd(sprintf(
    "%s view %s-Oz -o %s %s",
    shell_quote(bcftools_bin),
    region_arg,
    shell_quote(input_local),
    shell_quote(input_path)
  ))
  run_cmd(sprintf("%s index -f -t %s", shell_quote(bcftools_bin), shell_quote(input_local)))

  list(
    input_path = input_local,
    fasta_path = normalizePath(fasta_path, mustWork = TRUE),
    region = region
  )
}

prepare_duckdb_input_view <- function(input_path) {
  DBI::dbExecute(con, "DROP VIEW IF EXISTS bench_input")
  DBI::dbExecute(
    con,
    paste0(
      "CREATE TEMP VIEW bench_input AS ",
      "SELECT CHROM, POS, REF, ALT FROM read_bcf(", sql_string(input_path), ")"
    )
  )
}

duckdb_norm_group_sql <- function(fasta_path, split = FALSE) {
  paste0(
    "SELECT CHROM AS chrom, pos_normed, end_pos_normed, ref_normed, ",
    flatten_duck_alt_sql, " AS alt_normed, COUNT(*) AS duck_n ",
    "FROM duckhts_bcftools_norm(",
    sql_string("bench_input"), ", ",
    sql_string(fasta_path), ", ",
    "chrom_col := 'CHROM', pos_col := 'POS', ref_col := 'REF', alt_col := 'ALT', ",
    "split_multiallelic := ", if (split) "true" else "false", 
    ") ",
    "WHERE coalesce(norm_status, '') IN ('Normalized', 'Unchanged', 'RefOnly') ",
    "GROUP BY ALL ORDER BY chrom, pos_normed, end_pos_normed, ref_normed, alt_normed"
  )
}

duckdb_norm_status_sql <- function(fasta_path, split = FALSE) {
  paste0(
    "SELECT norm_status, COUNT(*) AS n ",
    "FROM duckhts_bcftools_norm(",
    sql_string("bench_input"), ", ",
    sql_string(fasta_path), ", ",
    "chrom_col := 'CHROM', pos_col := 'POS', ref_col := 'REF', alt_col := 'ALT', ",
    "split_multiallelic := ", if (split) "true" else "false", 
    ") GROUP BY ALL ORDER BY norm_status"
  )
}

run_duckdb_bench <- function(input_path, fasta_path, split = FALSE, n = 3L) {
  prepare_duckdb_input_view(input_path)
  sql <- duckdb_norm_group_sql(fasta_path, split = split)
  status_sql <- duckdb_norm_status_sql(fasta_path, split = split)
  times <- numeric(n)
  rows <- NA_integer_

  DBI::dbExecute(con, "PRAGMA threads=1")
  on.exit(DBI::dbExecute(con, "PRAGMA threads=4"), add = TRUE)

  for (i in seq_len(n)) {
    gc()
    t0 <- proc.time()[["elapsed"]]
    out <- DBI::dbGetQuery(con, sql)
    times[i] <- proc.time()[["elapsed"]] - t0
    rows <- sum(out$duck_n)
  }

  list(
    bench = data.frame(
      engine = "duckhts",
      mode = if (split) "split_multiallelic" else "site_preserving",
      runs = n,
      median_sec = stats::median(times),
      min_sec = min(times),
      max_sec = max(times),
      output_rows = rows,
      stringsAsFactors = FALSE
    ),
    grouped = out,
    status = DBI::dbGetQuery(con, status_sql)
  )
}

bcftools_output_table <- function(output_path) {
  header <- run_cmd_capture(sprintf("%s view -h %s", shell_quote(bcftools_bin), shell_quote(output_path)))
  has_end <- any(grepl("^##INFO=<ID=END,", header))

  if (has_end) {
    lines <- run_cmd_capture(sprintf(
      "%s query -f '%%CHROM\\t%%POS\\t%%INFO/END\\t%%REF\\t%%ALT\\n' %s",
      shell_quote(bcftools_bin),
      shell_quote(output_path)
    ))
    if (!length(lines)) {
      return(data.frame(chrom = character(), pos_normed = integer(), end_pos_normed = integer(), ref_normed = character(), alt_normed = character(), bcf_n = integer(), stringsAsFactors = FALSE))
    }
    raw <- utils::read.delim(text = lines, header = FALSE, sep = "\t", stringsAsFactors = FALSE)
    names(raw) <- c("chrom", "pos_normed", "end_pos_normed", "ref_normed", "alt_normed")
    raw$end_pos_normed[raw$end_pos_normed == "." | raw$end_pos_normed == ""] <- NA_character_
    raw$end_pos_normed <- as.integer(ifelse(is.na(raw$end_pos_normed), raw$pos_normed + nchar(raw$ref_normed) - 1L, raw$end_pos_normed))
  } else {
    lines <- run_cmd_capture(sprintf(
      "%s query -f '%%CHROM\\t%%POS\\t%%REF\\t%%ALT\\n' %s",
      shell_quote(bcftools_bin),
      shell_quote(output_path)
    ))
    if (!length(lines)) {
      return(data.frame(chrom = character(), pos_normed = integer(), end_pos_normed = integer(), ref_normed = character(), alt_normed = character(), bcf_n = integer(), stringsAsFactors = FALSE))
    }
    raw <- utils::read.delim(text = lines, header = FALSE, sep = "\t", stringsAsFactors = FALSE)
    names(raw) <- c("chrom", "pos_normed", "ref_normed", "alt_normed")
    raw$end_pos_normed <- raw$pos_normed + nchar(raw$ref_normed) - 1L
  }

  raw$alt_normed[raw$alt_normed == "."] <- NA_character_
  key <- paste(raw$chrom, raw$pos_normed, raw$end_pos_normed, raw$ref_normed, ifelse(is.na(raw$alt_normed), "<NULL>", raw$alt_normed), sep = "\r")
  counts <- sort(table(key), decreasing = FALSE)
  parts <- strsplit(names(counts), "\r", fixed = TRUE)
  out <- data.frame(
    chrom = vapply(parts, `[`, character(1), 1L),
    pos_normed = as.integer(vapply(parts, `[`, character(1), 2L)),
    end_pos_normed = as.integer(vapply(parts, `[`, character(1), 3L)),
    ref_normed = vapply(parts, `[`, character(1), 4L),
    alt_normed = vapply(parts, `[`, character(1), 5L),
    bcf_n = as.integer(counts),
    stringsAsFactors = FALSE
  )
  out$alt_normed[out$alt_normed == "<NULL>"] <- NA_character_
  out[order(out$chrom, out$pos_normed, out$end_pos_normed, out$ref_normed, out$alt_normed), , drop = FALSE]
}

run_bcftools_bench <- function(input_path, fasta_path, split = FALSE, n = 3L) {
  times <- numeric(n)
  out <- NULL

  for (i in seq_len(n)) {
    output_path <- file.path(tempdir(), sprintf("duckhts_norm_bench_%s_%02d.vcf.gz", if (split) "split" else "site", i))
    args <- c("norm")
    if (split) args <- c(args, "-m", "-any")
    args <- c(args, "-f", fasta_path, "-Oz", "-o", output_path, input_path)

    gc()
    t0 <- proc.time()[["elapsed"]]
    status <- system2(bcftools_bin, args = args, stdout = TRUE, stderr = TRUE)
    times[i] <- proc.time()[["elapsed"]] - t0
    exit_code <- attr(status, "status")
    if (!is.null(exit_code) && !identical(exit_code, 0L)) {
      stop("bcftools norm failed", call. = FALSE)
    }

    out <- bcftools_output_table(output_path)
    unlink(c(output_path, paste0(output_path, ".tbi"), paste0(output_path, ".csi")))
  }

  list(
    bench = data.frame(
      engine = "bcftools_norm",
      mode = if (split) "split_multiallelic" else "site_preserving",
      runs = n,
      median_sec = stats::median(times),
      min_sec = min(times),
      max_sec = max(times),
      output_rows = sum(out$bcf_n),
      stringsAsFactors = FALSE
    ),
    grouped = out
  )
}

compare_outputs <- function(duck_out, bcf_out) {
  duck_cmp <- duck_out
  bcf_cmp <- bcf_out
  duck_cmp$alt_key <- ifelse(is.na(duck_cmp$alt_normed), "<NULL>", duck_cmp$alt_normed)
  bcf_cmp$alt_key <- ifelse(is.na(bcf_cmp$alt_normed), "<NULL>", bcf_cmp$alt_normed)

  merged <- merge(
    duck_cmp[, c("chrom", "pos_normed", "end_pos_normed", "ref_normed", "alt_key", "duck_n")],
    bcf_cmp[, c("chrom", "pos_normed", "end_pos_normed", "ref_normed", "alt_key", "bcf_n")],
    by = c("chrom", "pos_normed", "end_pos_normed", "ref_normed", "alt_key"),
    all = TRUE
  )
  merged$alt_normed <- merged$alt_key
  merged$alt_normed[merged$alt_normed == "<NULL>"] <- NA_character_
  merged$status <- ifelse(
    is.na(merged$duck_n),
    "only_bcftools",
    ifelse(is.na(merged$bcf_n), "only_duckhts", ifelse(merged$duck_n != merged$bcf_n, "diff_counts", "match"))
  )
  merged <- merged[order(merged$chrom, merged$pos_normed, merged$end_pos_normed, merged$ref_normed, merged$alt_key), , drop = FALSE]

  list(
    summary = data.frame(
      status = names(sort(table(merged$status))),
      n = as.integer(sort(table(merged$status))),
      stringsAsFactors = FALSE
    ),
    mismatches = merged[merged$status != "match", c("chrom", "pos_normed", "end_pos_normed", "ref_normed", "alt_normed", "duck_n", "bcf_n", "status"), drop = FALSE]
  )
}

run_case <- function(input_path, fasta_path, split = FALSE, n = 3L) {
  duck <- run_duckdb_bench(input_path, fasta_path, split = split, n = n)
  bcf <- run_bcftools_bench(input_path, fasta_path, split = split, n = n)
  cmp <- compare_outputs(duck$grouped, bcf$grouped)
  list(
    bench = rbind(duck$bench, bcf$bench),
    duck_status = duck$status,
    compare_summary = cmp$summary,
    mismatches = head(cmp$mismatches, 10L)
  )
}
```

# Synthetic fixture

```{r synthetic-run}
fixture <- write_norm_fixture(file.path(tempdir(), "duckhts_norm_bench_fixture"), bench_rows)
site_case <- run_case(fixture$input_path, fixture$fasta_path, split = FALSE, n = bench_runs)
split_case <- run_case(fixture$input_path, fixture$fasta_path, split = TRUE, n = bench_runs)

synthetic_bench <- rbind(site_case$bench, split_case$bench)
knitr::kable(synthetic_bench, digits = 3)
```

Synthetic fixture mix:

- normalized insertions/deletions
- a multiallelic `TT,TTT` row
- a ref-only `ALT='.'` row that exercises split-mode row preservation

## Synthetic conformance

### Site-preserving mode

```{r synthetic-site-conformance}
knitr::kable(site_case$compare_summary)
knitr::kable(site_case$duck_status)
knitr::kable(site_case$mismatches)
```

### Split mode

```{r synthetic-split-conformance}
knitr::kable(split_case$compare_summary)
knitr::kable(split_case$duck_status)
knitr::kable(split_case$mismatches)
```

# Optional real-callset case

```{r real-case, results='asis'}
if (!nzchar(real_input_path) || !nzchar(real_fasta)) {
  cat("Real-callset benchmark skipped. Set `NORM_REAL_VCF` and `NORM_REAL_FASTA` to enable it.\n")
} else {
  real_fixture <- prepare_real_fixture(real_input_path, real_fasta, real_region)
  real_site <- run_case(real_fixture$input_path, real_fixture$fasta_path, split = FALSE, n = bench_runs)
  cat("## ", real_label, " (site-preserving)\n\n", sep = "")
  print(knitr::kable(real_site$bench, digits = 3))
  cat("\n")
  print(knitr::kable(real_site$compare_summary))
  cat("\n")
  print(knitr::kable(real_site$duck_status))
  cat("\n")
  print(knitr::kable(real_site$mismatches))
  cat("\n")

  if (real_split) {
    real_split_case <- run_case(real_fixture$input_path, real_fixture$fasta_path, split = TRUE, n = bench_runs)
    cat("## ", real_label, " (split_multiallelic)\n\n", sep = "")
    print(knitr::kable(real_split_case$bench, digits = 3))
    cat("\n")
    print(knitr::kable(real_split_case$compare_summary))
    cat("\n")
    print(knitr::kable(real_split_case$duck_status))
    cat("\n")
    print(knitr::kable(real_split_case$mismatches))
    cat("\n")
  }
}
```
