---
title: "DuckHTS SIMD BAM GC-content Benchmark"
output: github_document
---

<!-- benchmark_simd_bam_gc.md is generated from benchmark_simd_bam_gc.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This report measures `seq_gc_content(SEQ)` on real BAM read sequences.  It is
intended to complement the synthetic `seq_gc_content(...)` benchmark: the input
strings come from `read_bam(...)`, so read lengths, lowercase/uppercase handling,
`N` bases, DuckDB vector flow, and BAM decoding overhead are closer to a real HTS
workload.

Two workloads are reported for each SIMD backend request:

- `bam_scan`: scan the BAM and compute `seq_gc_content(SEQ)` in the same query.
  This is the end-to-end path and includes BAM decoding, decompression, DuckDB
  materialization, and GC-content computation.
- `materialized_seq`: materialize BAM `SEQ` strings once, then time only the
  real BAM sequence strings flowing through `seq_gc_content(...)`.  This better
  isolates the SIMD byte-scanning kernel on realistic sequence lengths.

The default input is the committed `test/data/nanopore.bam` fixture. Set
`SIMD_BAM_GC_BAM` to benchmark a separately staged BAM under the DuckHTS cache.

# Run

```sh
make bench-simd-bam-gc
```

Useful overrides:

- `SIMD_BAM_GC_BAM`: BAM path; default is the local HG00106 exome BAM if present
- `SIMD_BAM_GC_MAX_READS`: read limit before aggregation, default `0` for all reads
- `SIMD_BAM_GC_RUNS`: timed repeats per backend/workload, default `5`
- `SIMD_BAM_GC_THREADS`: DuckDB threads, default `1` to isolate SIMD effects
- `SIMD_BAM_GC_BACKENDS`: comma-separated backend requests, default `scalar,auto,avx2,avx512`
- `SIMD_BAM_GC_MODES`: comma-separated workloads, default `bam_scan,materialized_seq`
- `SIMD_BAM_GC_FORCE=1`: rerun even if cached JSON exists
- `DUCKHTS_EXTENSION`: extension path override

```{r setup, include = FALSE}
library(knitr)
library(jsonlite)

repo_root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", unset = getwd()), mustWork = TRUE)
duckhts_cache_dir <- Sys.getenv(
  "DUCKHTS_CACHE_DIR",
  unset = file.path(path.expand("~"), ".cache", "duckhts")
)
out_dir <- Sys.getenv("SIMD_BAM_GC_OUT_DIR", unset = file.path(duckhts_cache_dir, "benchmarks", "simd-bam-gc"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
json_out <- file.path(out_dir, "benchmark_simd_bam_gc.json")
meta_out <- file.path(out_dir, "benchmark_simd_bam_gc_meta.json")

extension <- Sys.getenv(
  "DUCKHTS_EXTENSION",
  unset = file.path(repo_root, "build", "release", "duckhts.duckdb_extension")
)
extension <- normalizePath(extension, mustWork = TRUE)

default_bam <- file.path(repo_root, "test", "data", "nanopore.bam")
bam <- Sys.getenv("SIMD_BAM_GC_BAM", unset = default_bam)
bam <- normalizePath(bam, mustWork = TRUE)

max_reads <- as.integer(Sys.getenv("SIMD_BAM_GC_MAX_READS", unset = "0"))
runs <- as.integer(Sys.getenv("SIMD_BAM_GC_RUNS", unset = "5"))
threads <- as.integer(Sys.getenv("SIMD_BAM_GC_THREADS", unset = "1"))
backends <- Sys.getenv("SIMD_BAM_GC_BACKENDS", unset = "scalar,auto,avx2,avx512")
modes <- Sys.getenv("SIMD_BAM_GC_MODES", unset = "bam_scan,materialized_seq")
force <- identical(Sys.getenv("SIMD_BAM_GC_FORCE", unset = "0"), "1")
python <- Sys.getenv("PYTHON", unset = "python3")

if (is.na(max_reads) || max_reads < 0L) stop("SIMD_BAM_GC_MAX_READS must be >= 0", call. = FALSE)
if (is.na(runs) || runs < 1L) stop("SIMD_BAM_GC_RUNS must be positive", call. = FALSE)
if (is.na(threads) || threads < 1L) stop("SIMD_BAM_GC_THREADS must be positive", call. = FALSE)

requested_meta <- list(
  schema = 1L,
  bam = bam,
  max_reads = max_reads,
  runs = runs,
  threads = threads,
  backends = backends,
  modes = modes,
  extension = extension
)
needs_run <- force || !file.exists(json_out) || !file.exists(meta_out)
if (!needs_run) {
  old_meta <- tryCatch(jsonlite::fromJSON(meta_out, simplifyVector = FALSE), error = function(e) list())
  needs_run <- !identical(old_meta, requested_meta)
}

if (needs_run) {
  cmd <- c(
    file.path(repo_root, "scripts", "benchmark_simd_bam_gc.py"),
    "--extension", extension,
    "--bam", bam,
    "--max-reads", as.character(max_reads),
    "--iterations", as.character(runs),
    "--threads", as.character(threads),
    "--backends", backends,
    "--modes", modes,
    "--json-out", json_out
  )
  out <- system2(python, cmd, stdout = TRUE, stderr = TRUE)
  status <- attr(out, "status")
  if (!is.null(status) && status != 0) {
    stop(paste(out, collapse = "\n"), call. = FALSE)
  }
  writeLines(jsonlite::toJSON(requested_meta, auto_unbox = TRUE, pretty = TRUE), meta_out)
}

raw_results <- jsonlite::fromJSON(json_out, simplifyVector = FALSE)
null_chr <- function(x) if (is.null(x)) NA_character_ else as.character(x)
null_num <- function(x) if (is.null(x)) NA_real_ else as.numeric(x)
null_int <- function(x) if (is.null(x)) NA_integer_ else as.integer(x)
null_lgl <- function(x) if (is.null(x)) FALSE else isTRUE(x)

results <- do.call(rbind, lapply(raw_results, function(x) {
  data.frame(
    benchmark = null_chr(x$benchmark),
    backend_request = null_chr(x$backend_request),
    requested_backend = null_chr(x$requested_backend),
    selected_backend = null_chr(x$selected_backend),
    kernel_backend = null_chr(x$kernel_backend),
    kernel_capability = null_chr(x$kernel_capability),
    kernel_scalar_fallback = if (is.null(x$kernel_scalar_fallback)) NA else isTRUE(x$kernel_scalar_fallback),
    available = if (is.null(x$available)) NA else isTRUE(x$available),
    skipped = null_lgl(x$skipped),
    skip_reason = null_chr(x$skip_reason),
    reads = null_int(x$reads),
    total_bases = null_num(x$total_bases),
    iterations = null_int(x$iterations),
    load_sec = null_num(x$load_sec),
    gc_fraction_sum = null_num(x$gc_fraction_sum),
    median_sec = null_num(x$median_sec),
    min_sec = null_num(x$min_sec),
    reads_per_sec_median = null_num(x$reads_per_sec_median),
    mbases_per_sec_median = null_num(x$mbases_per_sec_median),
    stringsAsFactors = FALSE
  )
}))

results$speedup_vs_scalar <- NA_real_
for (bench in unique(results$benchmark)) {
  idx <- which(!results$skipped & results$benchmark == bench & results$backend_request == "scalar")[1]
  if (length(idx) && !is.na(idx)) {
    scalar_rate <- results$mbases_per_sec_median[idx]
    if (is.finite(scalar_rate) && scalar_rate > 0) {
      rows <- which(results$benchmark == bench)
      results$speedup_vs_scalar[rows] <- results$mbases_per_sec_median[rows] / scalar_rate
    }
  }
}
results$status <- ifelse(results$skipped, "skipped", "ok")

path_display <- function(path) {
  if (startsWith(path, paste0(repo_root, .Platform$file.sep))) {
    sub(paste0("^", repo_root, .Platform$file.sep), "", path)
  } else {
    path
  }
}

configuration <- data.frame(
  setting = c("bam", "max_reads", "iterations", "threads", "backend_requests", "workloads", "extension"),
  value = c(
    path_display(bam),
    if (max_reads == 0L) "all" else format(max_reads, big.mark = ","),
    as.character(runs),
    as.character(threads),
    backends,
    modes,
    path_display(extension)
  ),
  stringsAsFactors = FALSE
)
```

# Configuration

```{r configuration, echo=FALSE}
knitr::kable(configuration)
```

# Results

```{r results, echo=FALSE}
display <- results[, c(
  "benchmark",
  "backend_request",
  "requested_backend",
  "selected_backend",
  "kernel_backend",
  "status",
  "reads",
  "total_bases",
  "load_sec",
  "median_sec",
  "min_sec",
  "mbases_per_sec_median",
  "speedup_vs_scalar",
  "skip_reason"
)]
knitr::kable(display, digits = 3)
```

# Interpretation

```{r interpretation, echo=FALSE, results='asis'}
measured <- results[!results$skipped, , drop = FALSE]
if (!nrow(measured)) {
  cat("No backend requests were available for measurement.\n")
} else {
  for (bench in unique(measured$benchmark)) {
    subset <- measured[measured$benchmark == bench, , drop = FALSE]
    best <- subset[which.max(subset$mbases_per_sec_median), , drop = FALSE]
    cat(sprintf(
      "For `%s`, the fastest measured `seq_base_counts` kernel backend was `%s` selected from request `%s`, at %.1f Mbases/s",
      bench, best$kernel_backend, best$backend_request, best$mbases_per_sec_median
    ))
    if (is.finite(best$speedup_vs_scalar)) {
      cat(sprintf(" (%.2fx scalar).", best$speedup_vs_scalar))
    } else {
      cat(".")
    }
    cat("\n\n")
  }
  cat("`bam_scan` is the practical end-to-end view. `materialized_seq` removes repeated BAM I/O and shows the SIMD kernel headroom on realistic read strings.\n")
}
```

# Raw timing vectors

```{r raw-timings, echo=FALSE}
timing_rows <- do.call(rbind, lapply(raw_results, function(x) {
  if (isTRUE(x$skipped) || !length(x$timings_sec)) return(NULL)
  data.frame(
    benchmark = x$benchmark,
    backend_request = x$backend_request,
    selected_backend = x$selected_backend,
    kernel_backend = x$kernel_backend,
    iteration = seq_along(x$timings_sec),
    seconds = unlist(x$timings_sec),
    stringsAsFactors = FALSE
  )
}))
if (is.null(timing_rows) || !nrow(timing_rows)) {
  cat("No timing rows were recorded.\n")
} else {
  knitr::kable(timing_rows, digits = 6)
}
```
