---
title: "Indexed BCF/VCF scan initialization: small-fixture cost"
output:
  github_document:
    html_preview: false
---

This paired measurement exercises automatic full-file scan planning and complete
materialization on committed BCF/VCF fixtures staged through duckhtsbench's registry.
The registry pins encoded file bytes; staging is a network-free verified copy,
not fixture reconstruction. The original two-record inputs have
one occupied contig and 256 empty header contigs. A second five-record corpus has
two occupied contigs, two samples, equal-position distinct alleles, and a physical
duplicate; its VCF header declares all, some, or none of the contigs. It uses
non-colocated indexes. Timings measure setup and small-result materialization,
not cohort throughput.

Both builds plan VCF work using the Tabix dictionary. The one-occupied-contig VCF
therefore selects a single full-file stream instead of 257 indexed contig tasks;
the five-record VCFs use two contig tasks irrespective of header completeness.
BCF retains its header dictionary. This is a deliberate scheduling change, not a
promise that the requested DuckDB thread limit equals the number of active workers.
The candidate retains the parsed index at bind and removes worker index reloads;
the baseline reloads an independent index in each indexed scan worker. Data and
indexes remain unchanged during this timing workload.

The nearest larger full-column report is
[`benchmark_reader_allocations.md`](benchmark_reader_allocations.md):
4,048,342 GIAB HG002 records in wide/tidy form, one sequential scan handle and
zero HTSlib decompression workers. It does not exercise automatic parallel
index reloads and is not an identical-workload baseline. The previous identical
two/five-record workload is retained at
[`40c2b52:benchmark_bcf_scan_init.md`](https://github.com/RGenomicsETL/duckhts/blob/40c2b52793740064956709e44b825553efe76cd3/benchmarks/benchmark_bcf_scan_init.md).
This report measures the pre-fix binary alongside the candidate on the same host.

## Reproduction and identity

Install duckhtsbench from this checkout. Set `BCF_SCAN_BASELINE_EXTENSION` and `BCF_SCAN_BASELINE_REVISION` to a separately
built baseline. `DUCKHTS_EXTENSION` selects the candidate. From the repository root:

```sh
taskset -c 0,2,4,6 Rscript -e 'rmarkdown::render("benchmarks/benchmark_bcf_scan_init.Rmd")'
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r measurement}
baseline <- Sys.getenv("BCF_SCAN_BASELINE_EXTENSION")
baseline_revision <- Sys.getenv("BCF_SCAN_BASELINE_REVISION")
candidate <- Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension")
stopifnot(nzchar(baseline_revision), file.exists(baseline), file.exists(candidate))
stopifnot(system2("git", c("diff", "--quiet", "HEAD", "--", "src")) == 0L)
identity <- data.frame(
  property = c("baseline revision", "candidate revision", "candidate src tree",
               "baseline binary MD5", "candidate binary MD5", "R", "DuckDB R package",
               "HTSlib", "CPU affinity", "repetitions"),
  value = c(baseline_revision, system2("git", c("rev-parse", "HEAD"), stdout = TRUE),
            system2("git", c("rev-parse", "HEAD:src"), stdout = TRUE),
            unname(tools::md5sum(baseline)), unname(tools::md5sum(candidate)),
            R.version.string, as.character(packageVersion("duckdb")),
            readLines("third_party/htslib/VERSION", n = 1),
            system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE),
            "9 timed calls after warm-up; fresh R process per build/input/thread setting")
)
knitr::kable(identity)
paths <- duckhtsbench::duckhts_bench_stage_bcf_scan(".")
artifact <- function(ids) unname(paths[paste0("bcf_scan_", ids)])
old_inputs <- artifact(c("old_bcf", "old_vcfgz"))
vcf_inputs <- artifact(c("full", "partial", "none"))
cases <- data.frame(
  input = c(old_inputs, vcf_inputs),
  index = artifact(c("old_csi", "old_tbi", "full_tbi", "partial_tbi", "none_tbi")),
  explicit_index = c(FALSE, FALSE, TRUE, TRUE, TRUE),
  oracle = c(rep(old_inputs[1], 2), rep(artifact("oracle"), 3)),
  format = c("BCF/CSI", "VCF.gz/TBI", rep("VCF.gz/TBI", 3)),
  header = c("257 contigs", "257 contigs", "full", "partial", "none"),
  header_contigs = c(257L, 257L, 3L, 2L, 0L),
  input_records = c(2L, 2L, 5L, 5L, 5L)
)
files <- unique(c(cases$input, cases$index, cases$oracle))
stopifnot(all(file.exists(files)),
          sum(grepl("^##contig=", readLines(artifact("old_vcf")))) == 257L)
knitr::kable(data.frame(file = basename(files), bytes = file.info(files)$size,
                        md5 = unname(tools::md5sum(files))))
results <- list()
oracles <- list()
for (i in seq_len(nrow(cases))) for (threads in c(1L, 4L)) for (build in c("baseline", "candidate")) {
  case <- cases[i, ]
  measured <- callr::r(function(extension, input, index, explicit_index, oracle, threads, build, input_records) {
    con <- DBI::dbConnect(duckdb::duckdb(config = list(allow_unsigned_extensions = "true")))
    on.exit(DBI::dbDisconnect(con, shutdown = TRUE))
    DBI::dbExecute(con, paste("LOAD", DBI::dbQuoteString(con, extension)))
    DBI::dbExecute(con, sprintf("SET threads=%d", threads))
    reference <- DBI::dbGetQuery(con, sprintf(
      "SELECT * FROM read_bcf(%s,scan_mode:='sequential',decompression_threads:=0) ORDER BY ALL",
      DBI::dbQuoteString(con, oracle)))
    stopifnot(nrow(reference) == input_records)
    index_option <- if (explicit_index) paste0(",index_path:=", DBI::dbQuoteString(con, index)) else ""
    sql <- sprintf("CREATE OR REPLACE TEMP TABLE scanned AS SELECT * FROM read_bcf(%s%s,decompression_threads:=0)",
                   DBI::dbQuoteString(con, input), index_option)
    DBI::dbExecute(con, sql)
    expected <- DBI::dbGetQuery(con, "SELECT * FROM scanned ORDER BY ALL")
    exact <- identical(expected, reference)
    stopifnot(exact)
    timings <- numeric(9)
    for (i in seq_along(timings)) {
      start <- Sys.time()
      DBI::dbExecute(con, sql)
      timings[i] <- as.numeric(difftime(Sys.time(), start, units = "secs"))
      stopifnot(identical(DBI::dbGetQuery(con, "SELECT * FROM scanned ORDER BY ALL"), expected))
    }
    list(rows = expected, reference = reference, exact = exact, timings = timings)
  }, args = list(extension = normalizePath(if (build == "baseline") baseline else candidate),
                 input = normalizePath(case$input), index = normalizePath(case$index),
                 explicit_index = case$explicit_index,
                 oracle = normalizePath(case$oracle), threads = threads, build = build,
                 input_records = case$input_records))
  if (is.null(oracles[[case$oracle]])) oracles[[case$oracle]] <- measured$reference
  stopifnot(identical(oracles[[case$oracle]], measured$reference))
  stopifnot(measured$exact) # Never compare an incomplete materialization.
  results[[length(results) + 1L]] <- data.frame(
    build, format = case$format, header = case$header,
    duckdb_threads = threads, htslib_workers_per_handle = 0L,
    header_contigs = case$header_contigs, input_records = case$input_records,
    materialized_rows = nrow(measured$rows), materialized_columns = ncol(measured$rows),
    exact_rows = measured$exact,
    median_ms = if (measured$exact) 1000 * median(measured$timings) else NA_real_,
    min_ms = if (measured$exact) 1000 * min(measured$timings) else NA_real_,
    max_ms = if (measured$exact) 1000 * max(measured$timings) else NA_real_
  )
}
knitr::kable(do.call(rbind, results), digits = 3)
```

Each timed `CREATE OR REPLACE TEMP TABLE AS SELECT *` materializes all columns,
not merely a count. The output denominator is two or five complete rows on both
builds, including partial-header scans fixed by the preceding change.
Elapsed `Sys.time()` includes the DBI call;
connection creation, warm-up, and complete typed-row verification are outside the
timer. Every candidate result and every valid baseline agrees exactly with the
complete BCF sequential rows, including GT, ALT and physical duplicate multiplicity,
without asserting scan arrival order. The BCF oracle also agrees across builds.

This does not measure the latency of an index failure, remote I/O, large-cohort
throughput, RSS, or maximum worker concurrency. Thread settings are DuckDB's
requested limits. Small timings can vary with scheduling; no general speedup or
no-regression claim follows from them.
