---
title: "Ancestry relational projection: synthetic scaling and public-reference parity"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(DBI)
library(Rduckhts)
library(duckhtsbench)
repo <- normalizePath("..")
source_revision <- system2("git", c("-C", shQuote(repo), "rev-parse", "HEAD"), stdout = TRUE)
registry <- file.path(repo, "r/duckhtsbench/inst/benchmark_registry.tsv")
cache <- tempfile("duckhts-ancestry-benchmark-")
dir.create(cache)
Sys.setenv(DUCKHTSBENCH_REGISTRY = registry, DUCKHTS_CACHE_DIR = cache)
correction_file <- duckhts_bench_stage_repository_fixtures(repo, "ancestry-projection")[["ancestry_correction"]]
con <- rduckhts_connect()
dbExecute(con, "SET threads=4")
dbExecute(con, paste(
  "CREATE TABLE bench_ancestry_ref AS SELECT 'chr1' AS chromosome,",
  "(100000+i)::INTEGER AS position, 'A' AS allele_a, 'G' AS allele_b,",
  "g.group_id, CASE WHEN g.group_id='A' THEN 0.25 + 0.16*sin(i*0.017)",
  "ELSE 0.68 + 0.12*cos(i*0.019) END AS frequency",
  "FROM range(17000) t(i), (VALUES ('A'),('B')) g(group_id)"
))
dbExecute(con, paste(
  "CREATE TABLE bench_ancestry_pc AS SELECT DISTINCT chromosome, position, allele_a, allele_b,",
  "pc, CASE WHEN pc=1 THEN sin(position*0.031) ELSE cos(position*0.027) END AS loading",
  "FROM bench_ancestry_ref, (VALUES (1),(2)) t(pc)"
))
dbExecute(con, sprintf(
  "CREATE TABLE bench_ancestry_correction AS SELECT pc::INTEGER AS pc, coefficient::DOUBLE AS coefficient FROM read_csv_auto(%s, delim='\\t', header=true) WHERE pc<=2",
  as.character(dbQuoteString(con, correction_file))
))
dbExecute(con, paste(
  "CREATE TABLE bench_ancestry_input AS SELECT sample_id, chromosome, position, allele_a, allele_b,",
  "sum(CASE WHEN group_id='A' THEN a_share ELSE 1-a_share END * frequency)",
  "+ 0.002*sin(position * sample_number) AS frequency",
  "FROM bench_ancestry_ref, (VALUES ('S1',0.7,1),('S2',0.4,2),('S3',0.1,3))",
  "s(sample_id,a_share,sample_number)",
  "GROUP BY sample_id, sample_number, chromosome, position, allele_a, allele_b"
))
peak_rss_mb <- function() {
  status <- grep("^VmHWM:", readLines("/proc/self/status"), value = TRUE)
  as.numeric(strsplit(trimws(status), "[[:space:]]+")[[1L]][[2L]]) / 1024
}
baseline_rss_mb <- peak_rss_mb()
run_sites <- function(n, repeat_id) {
  dbExecute(con, sprintf("CREATE OR REPLACE TEMP VIEW bench_ancestry_slice AS SELECT * FROM bench_ancestry_input WHERE position < %d", 100000 + n))
  start <- proc.time()[["elapsed"]]
  out <- rduckhts_ancestry_proportions(
    con, "bench_ancestry_slice", "bench_ancestry_ref", "bench_ancestry_pc",
    "bench_ancestry_correction", min_cor = 0.4
  )
  elapsed <- proc.time()[["elapsed"]] - start
  stopifnot(nrow(out) == 6L, all(out$used_variants == n))
  data.frame(samples = 3L, variants_per_sample = n, input_rows = 3L*n,
             output_group_rows = nrow(out), groups = 2L, pcs = 2L,
             threads = 4L, repeat_id = repeat_id, elapsed_seconds = elapsed,
             process_peak_rss_mb = peak_rss_mb(),
             failures = paste(unique(out$status[out$status != "ok"]), collapse = ";"))
}
invisible(run_sites(1000L, 0L))
results <- do.call(rbind, lapply(c(1000L, 5000L, 17000L), function(n) {
  do.call(rbind, lapply(seq_len(3L), function(rep) run_sites(n, rep)))
}))
```

Revision: `r source_revision`. Workload: deterministic biallelic synthetic
reference frequencies and PC loadings; three independently perturbed samples;
4 DuckDB threads; two reference groups; two PCs; `min_cor = 0.4`. Each variant
is present in both groups and PCs, and every input row matches. This report
measures query and materialization together. The public-reference workload is
measured separately below.

The correction vector is staged from the committed, checksummed bigsnpr vignette
product. `VmHWM` is cumulative process high-water RSS (MiB), including package
loading and all preceding cases; the baseline before the first run was
`r round(baseline_rss_mb, 1)` MiB. It is not per-query allocated memory.

```{r results}
knitr::kable(results, digits = 3)
```

Failed gates are retained in the `failures` column for each repeat. The nearest
identical synthetic workload is the earlier rendered
`benchmarks/benchmark_ancestry.md` at repository revision `8ab38b99` (source
revision `56aad9cf`): its three 17,000-site repeats had median 0.078 seconds,
versus `r round(median(results$elapsed_seconds[results$variants_per_sample == 17000L]), 3)`
seconds here. These are separate R processes with uncontrolled background load;
this difference cannot establish a regression or equivalence. There is no
ancestry-projection baseline on `develop`; the synthetic rows are not evidence
about real-reference loading or BAM/CRAM depth sensitivity.

## Public reference and GRCh37 chr22 genotypes

Two checksum-verified Figshare CSV products each contain 5,816,590 rows:
21 reference-frequency groups and 16 PC loadings. A staged, key-sorted Parquet
join holds the four site-key columns, 16 DOUBLE loadings and 21 DOUBLE
frequencies. Its SHA-256 is
`8f0b5f7d3cd98fafd1b0598c63e743d07697d7536526c89948883d5aff791489`;
the staging receipt binds this digest to both registered CSV digests and DuckDB
1.5.5. The one-time conversion from checksum-verified CSVs took 322.26 seconds
and peaked at 15,269 MiB RSS; it is not included in the query measurements.
The epilepsy summary input has 4,880,492 records. A separate public phase-3 chr22 VCF supplies 11 individuals
from AFR (3), EUR (2), EAS (2), SAS (2) and AMR (2). Each workload uses four
DuckDB threads and the bigsnpr vignette's 16 correction coefficients. These
are complete-process peak RSS measurements, including package loading,
CSV-based bigsnpr oracles and preceding operations in that process. The SQL
time includes scanning the staged Parquet view, matching sites and aggregation,
but excludes loading the oracle CSVs.

```{r real-reference}
real_epilepsy <- callr::r(function(repo) {
  setwd(repo)
  Sys.unsetenv("DUCKHTS_CACHE_DIR")
  source("test/scripts/ancestry_bigsnpr_real.R", local = TRUE)
  rss <- grep("^VmHWM:", readLines("/proc/self/status"), value = TRUE)
  list(reference_load_seconds = load_seconds, query_seconds = elapsed,
       peak_rss_mib = as.numeric(strsplit(trimws(rss), "[[:space:]]+")[[1L]][[2L]]) / 1024,
       summary_input = nrow(sumstats), upstream_matched = nrow(matched),
       eligible_chr22 = eligible, matched = nrow(input),
       group_rows = nrow(parity), groups = length(shared),
       cor_pred = result$cor_pred[[1L]], oracle_cor_pred = attr(shared, "cor_pred"),
       rounding_bound = correlation_bound, group_parity = parity)
}, args = list(repo = repo))
real_epilepsy_full <- callr::r(function(repo) {
  setwd(repo)
  Sys.unsetenv("DUCKHTS_CACHE_DIR")
  Sys.setenv(ANCESTRY_EPILEPSY_CHR22_ONLY = "false")
  source("test/scripts/ancestry_bigsnpr_real.R", local = TRUE)
  rss <- grep("^VmHWM:", readLines("/proc/self/status"), value = TRUE)
  list(reference_load_seconds = load_seconds, query_seconds = elapsed,
       peak_rss_mib = as.numeric(strsplit(trimws(rss), "[[:space:]]+")[[1L]][[2L]]) / 1024,
       summary_input = nrow(sumstats), upstream_matched = nrow(matched),
       eligible = eligible, matched = nrow(input), group_rows = nrow(parity),
       groups = length(shared), cor_pred = result$cor_pred[[1L]],
       oracle_cor_pred = attr(shared, "cor_pred"),
       rounding_bound = correlation_bound, group_parity = parity)
}, args = list(repo = repo))
real_individuals <- callr::r(function(repo) {
  setwd(repo)
  Sys.unsetenv("DUCKHTS_CACHE_DIR")
  source("test/scripts/ancestry_1000g_real.R", local = TRUE)
  rss <- grep("^VmHWM:", readLines("/proc/self/status"), value = TRUE)
  list(reference_load_seconds = load_seconds, query_seconds = elapsed,
       peak_rss_mib = as.numeric(strsplit(trimws(rss), "[[:space:]]+")[[1L]][[2L]]) / 1024,
       vcf_snvs = nrow(vcf), reference_overlap = nrow(joined),
       samples = length(samples), sites_per_sample = nrow(site),
       input_rows = nrow(input), group_rows = nrow(parity), groups = ncol(ref) - 5L,
       summary = as.data.frame(summary))
}, args = list(repo = repo))
real_timings <- data.frame(
  workload = c("epilepsy (chr22)", "epilepsy (genome-wide)",
               "1000G phase-3 chr22 genotypes"),
  samples = c(1L, 1L, real_individuals$samples),
  variants_per_sample = c(real_epilepsy$matched, real_epilepsy_full$matched,
                          real_individuals$sites_per_sample),
  input_rows = c(real_epilepsy$matched, real_epilepsy_full$matched,
                 real_individuals$input_rows),
  output_group_rows = c(real_epilepsy$group_rows, real_epilepsy_full$group_rows,
                        real_individuals$group_rows),
  groups = c(real_epilepsy$groups, real_epilepsy_full$groups,
             real_individuals$groups),
  threads = 4L,
  reference_rows = 5816590L,
  reference_load_seconds = c(real_epilepsy$reference_load_seconds,
                             real_epilepsy_full$reference_load_seconds,
                             real_individuals$reference_load_seconds),
  sql_seconds = c(real_epilepsy$query_seconds, real_epilepsy_full$query_seconds,
                  real_individuals$query_seconds),
  process_peak_rss_mib = c(real_epilepsy$peak_rss_mib,
                           real_epilepsy_full$peak_rss_mib,
                           real_individuals$peak_rss_mib))
knitr::kable(real_timings, digits = 2)
```

Epilepsy matching audit: bigsnpr matched `r real_epilepsy_full$upstream_matched` /
`r real_epilepsy_full$summary_input` records genome-wide (6 ambiguous SNPs
excluded, 1,576,097 reversed, 0 strand-flipped). DuckHTS matched
`r real_epilepsy_full$matched` / `r real_epilepsy_full$matched` pre-matched
inputs, after matching against the real reference product. The independently
selected, position-unique, unambiguous chr22 subset is
`r real_epilepsy$matched` / `r real_epilepsy$eligible_chr22` eligible chr22
matches, with `r real_epilepsy$matched` / `r real_epilepsy$matched` matched
by DuckHTS. Both genome-wide solvers consume the same allele-aligned sites.

```{r epilepsy-parity}
knitr::kable(real_epilepsy_full$group_parity[, c("group_id", "proportion", "oracle",
  "absolute_difference")], digits = 7)
```

Genome-wide `cor_pred`: DuckHTS
`r sprintf("%.12f", real_epilepsy_full$cor_pred)`, bigsnpr
`r sprintf("%.12f", real_epilepsy_full$oracle_cor_pred)`. The conservative
seven-decimal QP rounding correlation bound is
`r sprintf("%.9f", real_epilepsy_full$rounding_bound)`; the 0.4 gate is
farther away. On chr22, both return `r real_epilepsy$cor_pred` at seven
decimals with group difference
`r max(real_epilepsy$group_parity$absolute_difference)`; its rounding bound is
`r sprintf("%.9f", real_epilepsy$rounding_bound)`. All 21 genome-wide
coefficients agree at seven decimals; the script checks rounded equality.

The individual input is restricted from `r real_individuals$vcf_snvs` unique
biallelic SNVs to `r real_individuals$reference_overlap` same-assembly,
allele-agreeing, unambiguous reference overlaps before spaced site selection.
Both solvers see the same per-sample genotypes. bigsnpr's
imperfect-match warning (`cor_pred < 0.99`) is retained; passing the
0.4 gate does not imply that the frequencies match perfectly. The 11
individuals' 231 coefficients agree with bigsnpr at seven decimals.

```{r individual-parity}
knitr::kable(real_individuals$summary, digits = 7)
```

## Keyed Parquet profile and input scaling

The pre-conversion genome-wide DuckDB JSON profile, on revision
`0a5fd898c2bb8449259e01dd7aff161bb4dc3bc8`, measured 30.27 s and
40,232 MiB peak buffer memory. Its complete R process peaked at 38,783 MiB
RSS while retaining the CSV-based bigsnpr oracle. The 53,491,760-row
materialized `basis` and 1,123,326,960-row `x` join dominate cardinality;
`pred_sites` separately groups 3,343,235 variants after a long-reference
join. The old `list()` calls assemble only the small post-aggregation solver
vectors; there is no per-variant list construction. The profile also shows
hash joins and grouping over the long reference, but DuckDB does not
attribute its single reported peak buffer value to individual operators.
CSV parsing occurs in R before this SQL profile, not in a DuckDB scan
operator. Sorting is limited to the input window partition and final group
ordering.

The keyed query instead scans projected Parquet columns and aggregates 336
`sum(PC * group_frequency)` values, 16 corrected `sum(PC * aligned_frequency)`
values and group correlations per sample. Its only `list_value()` expressions
assemble the 336-element solver vector and 21 group correlations **after**
aggregation. It retains narrow site classification and aligned-frequency
rows; the 5,816,590-site reference hash/group and joins are the dominant
build-side state. The four-thread full-epilepsy query took 3.39 s with
3,806 MiB peak DuckDB buffer, zero temp spill and 8,898 MiB process peak
RSS including the CSV oracle. Standalone Parquet-only query process RSS is
reported below; it does not retain CSV reference tables or a bigsnpr oracle.
Buffer peaks describe DuckDB's accounting, not resident bytes, and cannot
be added to RSS.

`benchmark_ancestry_reference_load.R` scans every numeric column in each
source CSV and the staged Parquet product with DuckDB; neither scan
materializes 5.8 million rows in R. These timings differ from the CSV
`fread()` oracle-loading times above and have their own denominator.

```{r reference-load-comparison}
reference_load <- utils::read.delim("ancestry_reference_load.tsv")
stopifnot(nrow(reference_load) == 2L,
          identical(reference_load$threads, c(1L, 4L)),
          all(reference_load$reference_rows == 5816590L),
          all(reference_load$numeric_columns == 37L))
knitr::kable(reference_load, digits = 2)
```

`benchmark_ancestry_memory.R` uses fresh R processes and the same staged
Parquet reference for each row. Epilepsy 1×/2×/4× are nested prefixes of
835,809/1,671,618/3,343,235 physical input variants (one sample).
Genotype 11/22/44 is the same 20,000 physical chr22 sites per individual
with copied sample IDs; the extra sample IDs are a scaling probe, not new
independent genotypes. Every input row participates, and each group produces
one output row. Peak RSS includes DuckDB, the input Parquet reader and R
package loading; peak buffer is DuckDB's query-session maximum. Query time
includes reading Parquet, site joins, aggregation, the solver and quality
correlations, but not one-time conversion or source-CSV checksum validation.
The full epilepsy and 11-individual coefficient tables also agree at seven
decimals between one and four threads (21 and 231 keyed rows, respectively).

```{r parquet-scaling}
scaling <- utils::read.delim("ancestry_memory_scaling.tsv")
stopifnot(nrow(scaling) == 12L,
          all(scaling$threads %in% c(1L, 4L)),
          all(scaling$scale %in% c(1L, 2L, 4L)),
          all(scaling$input_rows == scaling$used_variants),
          all(scaling$output_rows == scaling$samples * 21L))
knitr::kable(scaling, digits = 2)
```

## Indexed 30x CRAM sensitivity

These retained measurements were produced at
`0a5fd898c2bb8449259e01dd7aff161bb4dc3bc8` with the existing long-form
BAM ancestry path, not remeasured for the keyed Parquet projection.
Three public 30x CRAMs (NA18507/AFR, HG00188/EUR, HG00403/EAS) were
read via their indexes at chr22:15,335,303–50,800,284 (GRCh38). The
registered GRCh37-to-GRCh38 chain maps 19,981/20,000 selected GRCh37
loci onto chr22; 430 mapped loci require reference swaps. The 19,551
unswapped, unique SNVs provide a spaced 17,000-site panel. No whole
CRAM was downloaded. Each 25% read subsample uses samtools seed 42;
median depth and allele-read totals are measured after downsampling.
Each mode sees the same `min_depth = 7` criterion and the same VCF site
subset. Each query emits 21 group rows with four DuckDB threads and
two htslib workers. Peak RSS is the R process high-water mark (MiB)
after reference loading and regional CRAM staging; it excludes samtools
child-process RSS and is not incremental SQL memory. The `seconds`
column is the ancestry BAM query alone, excluding CRAM staging and
depth audit.
The native and quarter-depth columns retain every status, including
failed gates and their `cor_pred` values.

```{r cram-sensitivity}
cram_sensitivity <- utils::read.delim("ancestry_30x_sensitivity.tsv",
  stringsAsFactors = FALSE)
stopifnot(nrow(cram_sensitivity) == 36L,
  all(cram_sensitivity$sites %in% c(1000L, 5000L, 17000L)),
  length(unique(cram_sensitivity$sample_id)) == 3L,
  all(c("allele_fraction", "called_genotype") %in% cram_sensitivity$mode),
  nrow(unique(cram_sensitivity[, c("sites", "panel_sha256")])) == 3L,
  all(with(cram_sensitivity, gate_distance[is.finite(gate_distance)] >
           rounding_bound[is.finite(gate_distance)])))
knitr::kable(unique(cram_sensitivity[, c("sites", "panel_sha256")]),
             row.names = FALSE)
knitr::kable(cram_sensitivity[, c("sample_id", "sites", "depth", "mode",
  "vcf_used", "bam_used", "depth_eligible", "median_depth", "allele_reads",
  "vcf_cor_pred", "cor_pred", "status", "max_group_difference",
  "rounding_bound", "gate_distance", "seconds", "process_peak_rss_mib")], digits = 4)
```

`r sum(cram_sensitivity$status != "ok")` / `r nrow(cram_sensitivity)`
correlation gates failed. The minimum distance from the 0.4 gate among
the measurable cases is `r min(cram_sensitivity$gate_distance, na.rm = TRUE)`;
the largest
coefficient-rounding correlation bound across the three site counts is
`r max(cram_sensitivity$rounding_bound)`.
Called-genotype frequency excludes sites whose balance rule makes no call;
its `bam_used` denominator is therefore smaller than the depth-eligible
count. Maximum group differences compare the same-site VCF result with
each BAM/CRAM mode; they are sensitivity, not an equality assertion.

```{r cleanup, include=FALSE}
dbDisconnect(con, shutdown = TRUE)
unlink(cache, recursive = TRUE)
```
