---
title: "Somalier find-sites: 1000 Genomes 30x phased chromosome 22"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
root <- normalizePath("..", mustWork = TRUE)
read_result <- function(name) utils::read.delim(
  file.path(root, "benchmarks", name), check.names = FALSE
)
observations <- read_result("somalier_find_sites_chr22_observations.tsv")
gates <- read_result("somalier_find_sites_chr22_gates.tsv")
timing <- read_result("somalier_find_sites_chr22_timing.tsv")
classes <- read_result("somalier_find_sites_chr22_disagreement_summary.tsv")
disagreements <- read_result("somalier_find_sites_chr22_disagreements.tsv")
xobservations <- read_result("somalier_find_sites_chrx_observations.tsv")
xgates <- read_result("somalier_find_sites_chrx_gates.tsv")
xtiming <- read_result("somalier_find_sites_chrx_timing.tsv")
xclasses <- read_result("somalier_find_sites_chrx_disagreement_summary.tsv")
xdisagreements <- read_result("somalier_find_sites_chrx_disagreements.tsv")
xpar <- read_result("somalier_find_sites_chrx_par_records.tsv")
```

This is a whole-chromosome differential of DuckHTS and pinned Somalier
[v0.3.4](https://github.com/brentp/somalier/tree/ff58fdade8f4f8293d904f10e0a4a13f1fac808d)
(binary SHA-256 `18717c205a9c4b65d479f1d2cf069a30b047a5378edf544d403d4081f06a3a78`).
The unmodified [1000 Genomes high-coverage 30x phased chr22 VCF](https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20201028_3202_phased/CCDG_14151_B01_GRM_WGS_2020-08-05_chr22.filtered.shapeit2-duohmm-phased.vcf.gz)
is staged through `duckhtsbench` (SHA-256
`70ff24cfd01eab24c6f4c6e825f55a2b3475ff595f6e12e179c4ef06b9cc4df6`,
519,930,289 bytes, GRCh38). Its declared INFO/AF and INFO/AN are used without
recomputing from genotypes. Somalier skips sample genotypes; DuckHTS reads the
physical VCF in sequential mode with `samples := ''`. The cache VCF is not committed.

- Source revision: `r observations$revision`. Workload: `--min-AN 6000`,
  `--snp-dist 10000`, minimum AF 0.15, AF target 0.48, no interval or
  gnotate exclusion. One run per tool, warm filesystem cache, GNU `/usr/bin/time -v`.
- Denominators: `r format(observations$source_records, big.mark = ",")` input
  records; `r format(observations$upstream_candidates, big.mark = ",")` Somalier
  candidates and `r format(observations$duckhts_candidates, big.mark = ",")`
  DuckHTS candidates before nearby-variant exclusion;
  `r format(observations$max_per_chromosome_candidates, big.mark = ",")`
  chr22 spacing inputs. The native spacing kernel's per-list limit is
  1,000,000 candidates; this chromosome's largest list is below it.
- Selected: `r observations$upstream_sites` Somalier sites and
  `r observations$duckhts_sites` DuckHTS sites;
  `r observations$common_sites` shared four-field keys
  `(region, position, source_ref, source_alt)`. Input-order and lexical
  tie policies differed on `r observations$lexical_tie_dependent_sites`
  selected sites on this chromosome.
- Threads: Somalier uses two reader/decompression threads and one writer
  thread (pinned source); DuckHTS uses one DuckDB thread and a sequential
  reader. The timed DuckHTS process also verifies its cached input and
  starts R/DuckDB; Somalier timing covers the CLI only. These single-process
  measurements are not matched-concurrency speedups.

```{r gates}
knitr::kable(gates, caption = "Whole-chromosome stage counts. Indels and snps are exclusion-neighbor inventories, not sequential pass counts. NA denotes stages the pinned Somalier executable does not expose; its own log reports the gated candidate count.")
```

```{r timing}
knitr::kable(timing, caption = paste(
  "Wall seconds and peak RSS KiB for each timed process, including its own",
  "initialization and I/O. Single runs are not a variance estimate."))
```

```{r disagreement-summary}
knitr::kable(classes, caption = "Keyed disagreement classes (all records counted).")
```

The retained `somalier_find_sites_chr22_disagreements.tsv` contains the first
100 disagreement keys and origins; `r nrow(disagreements)` rows are present.
An empty table means no disagreements, not an absent comparison. The
`somalier_find_sites_chr22_observations.tsv` and gate table retain the full
comparison denominators. The previous
[`benchmark_somalier_find_sites.md`](benchmark_somalier_find_sites.md)
uses two small fixture slices and is not a comparable whole-chromosome baseline.

## Chromosome X, including physical PAR records

The unmodified [1000 Genomes 30x phased chrX VCF](https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20201028_3202_phased/CCDG_14151_B01_GRM_WGS_2020-08-05_chrX.filtered.eagle2-phased.v2.vcf.gz)
(SHA-256 `f42c786cba67ff74c30e9bd37254b5a605a5957cfe3a5dc54e7dd41d9c9d8edc`,
2,800,398,832 bytes) contains `r format(xpar$par1_records, big.mark = ",")`
records in PAR1 (before POS 2,781,480) and
`r format(xpar$par2_records, big.mark = ",")` records in PAR2 (after POS
154,931,045). The physical first and last positions are
`r xpar$first_position` and `r xpar$last_position`.
Somalier's pinned X-range rule admits only zero-based positions
2,781,479..154,931,044; the physical PAR records exercise that filter.
Somalier's log says “0 autosomal variants” for this X-only input; the
actual compressed output VCF, not that autosomal-only log counter, is used
for the selected-site denominator and keyed comparison.

- Source revision: `r xobservations$revision`.
- Denominators: `r format(xobservations$source_records, big.mark = ",")`
  input records; `r format(xobservations$upstream_candidates, big.mark = ",")`
  Somalier candidates and `r format(xobservations$duckhts_candidates, big.mark = ",")`
  DuckHTS gated candidates;
  `r format(xobservations$max_per_chromosome_candidates, big.mark = ",")`
  spacing inputs (below the 1,000,000 limit).
- Selected: `r xobservations$upstream_sites` Somalier and
  `r xobservations$duckhts_sites` DuckHTS sites;
  `r xobservations$common_sites` matching four-field keys.
  Input and lexical tie policies differed on
  `r xobservations$lexical_tie_dependent_sites` selected sites.
  `r xobservations$duckhts_only` DuckHTS-only and
  `r xobservations$upstream_only` Somalier-only keys.

```{r x-gates}
knitr::kable(xgates, caption = "chrX stage counts; NA denotes stages not reported by the pinned executable.")
```

```{r x-timing}
knitr::kable(xtiming, caption = "chrX whole-process wall seconds and peak RSS KiB; thread and measurement caveats above also apply.")
```

```{r x-disagreement-summary}
knitr::kable(xclasses, caption = sprintf(
  "chrX keyed disagreement classes. First 100 rows in somalier_find_sites_chrx_disagreements.tsv; retained: %s rows.",
  nrow(xdisagreements)))
```

Reproduce from the repository root with installed `Rduckhts` and
`duckhtsbench`, the pinned binary staged in `duckhtsbench`, and GNU time:

```bash
export DUCKHTSBENCH_REGISTRY="$PWD/r/duckhtsbench/inst/benchmark_registry.tsv"
Rscript test/scripts/somalier_find_sites_chromosome.R upstream "$PWD" chr22
time_file="$(Rscript -e 'cat(file.path(dirname(duckhtsbench::duckhts_bench_artifact_path("somalier_find_sites_chr22_source")), "chr22.duckhts.time"))')"
/usr/bin/time -v -o "$time_file" \
  Rscript test/scripts/somalier_find_sites_chromosome.R select "$PWD" chr22
Rscript test/scripts/somalier_find_sites_chromosome.R compare "$PWD" chr22
Rscript test/scripts/somalier_find_sites_chromosome.R upstream "$PWD" chrX
x_time_file="$(Rscript -e 'cat(file.path(dirname(duckhtsbench::duckhts_bench_artifact_path("somalier_find_sites_chrx_source")), "chrX.duckhts.time"))')"
/usr/bin/time -v -o "$x_time_file" \
  Rscript test/scripts/somalier_find_sites_chromosome.R select "$PWD" chrX
Rscript test/scripts/somalier_find_sites_chromosome.R compare "$PWD" chrX
cd benchmarks && Rscript -e "rmarkdown::render('benchmark_somalier_find_sites_chr22.Rmd')"
```
