---
title: "DuckHTS VariantKey Join and Overlap Benchmark"
output: github_document
---

<!-- benchmark_variantkey_join_overlap.md is generated from benchmark_variantkey_join_overlap.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This report focuses on the workloads that matter for a future DuckVEP-style
annotation layer:

1. exact variant joins with stored `VariantKey`
2. safe mixed joins where hashed/nonreversible rows refine on exact alleles
3. exact span joins with stored `RegionKey`
4. interval overlap through RegionKey-ordered candidate ranges, DuckDB range
   joins, and `duckhts_cgranges_*`

This is intentionally separate from `benchmark_variantkey_conformance.Rmd`.
That report measures encoding and `%VKX` parity. This one measures join and
overlap throughput.

## Run

```sh
make bench-variantkey-join
```

Useful overrides:

- `VARIANTKEY_JOIN_ROWS`: synthetic exact-join rows, default `1000000`
- `VARIANTKEY_INTERVAL_ROWS`: synthetic interval rows, default `250000`
- `VARIANTKEY_JOIN_RUNS`: timed repeats, default `3`
- `VARIANTKEY_JOIN_THREADS_LIST`: DuckDB thread grid, default `1,2,4`
- `DUCKHTS_EXTENSION`: optional extension path override

```{r setup, include = FALSE}
library(DBI)
library(duckdb)

ext_path <- Sys.getenv(
  "DUCKHTS_EXTENSION",
  unset = normalizePath("build/release/duckhts.duckdb_extension", mustWork = TRUE)
)
join_rows <- as.integer(Sys.getenv("VARIANTKEY_JOIN_ROWS", unset = "1000000"))
interval_rows <- as.integer(Sys.getenv("VARIANTKEY_INTERVAL_ROWS", unset = "250000"))
bench_runs <- as.integer(Sys.getenv("VARIANTKEY_JOIN_RUNS", unset = "3"))
thread_grid_env <- Sys.getenv("VARIANTKEY_JOIN_THREADS_LIST", unset = "1,2,4")

parse_thread_grid <- function(x) {
  vals <- trimws(strsplit(x, ",", fixed = TRUE)[[1]])
  vals <- vals[nzchar(vals)]
  out <- unique(as.integer(vals))
  if (!length(out) || any(is.na(out)) || any(out < 1L)) {
    stop("VARIANTKEY_JOIN_THREADS_LIST must be a comma-separated list of positive integers", call. = FALSE)
  }
  out
}

thread_grid <- parse_thread_grid(thread_grid_env)

if (is.na(join_rows) || join_rows < 1L) stop("VARIANTKEY_JOIN_ROWS must be positive", call. = FALSE)
if (is.na(interval_rows) || interval_rows < 1L) stop("VARIANTKEY_INTERVAL_ROWS must be positive", call. = FALSE)
if (is.na(bench_runs) || bench_runs < 1L) stop("VARIANTKEY_JOIN_RUNS must be positive", call. = FALSE)

benchmark_settings <- data.frame(
  setting = c(
    "variant_join_rows",
    "interval_rows",
    "benchmark_runs",
    "duckdb_threads_grid",
    "mixed_hash_fraction"
  ),
  value = c(
    as.character(join_rows),
    as.character(interval_rows),
    as.character(bench_runs),
    paste(thread_grid, collapse = ","),
    "20%"
  ),
  stringsAsFactors = FALSE
)

source_revision <- trimws(system2("git", c("rev-parse", "HEAD"), stdout = TRUE))
cpu_model <- sub(
  "^model name\\s*:\\s*",
  "",
  grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[1L]
)
cpu_affinity <- trimws(system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE))

drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
con <- dbConnect(drv, dbdir = ":memory:")
DBI::dbExecute(con, sprintf("LOAD '%s';", gsub("\\\\", "/", ext_path)))
```

```{r helpers, include = FALSE}
rows_per_second <- function(rows, seconds) {
  rows / seconds
}

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

set_duck_threads <- function(n) {
  DBI::dbExecute(con, sprintf("PRAGMA threads=%d;", as.integer(n)))
}

measure_runs <- function(fun, runs = bench_runs) {
  times <- numeric(runs)
  value <- NULL
  for (i in seq_len(runs)) {
    gc()
    times[i] <- system.time({
      value <- fun()
    })[["elapsed"]]
  }
  list(
    value = value,
    times = times,
    median_seconds = unname(stats::median(times))
  )
}

measure_thread_grid <- function(operation, rows, fun, runs = bench_runs) {
  do.call(
    rbind,
    lapply(thread_grid, function(th) {
      set_duck_threads(th)
      m <- measure_runs(fun, runs = runs)
      result_rows <- as.numeric(m$value[[1L]][1L])
      data.frame(
        operation = operation,
        threads = th,
        input_rows = rows,
        result_rows = result_rows,
        median_seconds = m$median_seconds,
        input_rows_per_second = rows_per_second(rows, m$median_seconds),
        result_rows_per_second = rows_per_second(result_rows, m$median_seconds),
        stringsAsFactors = FALSE
      )
    })
  )
}

sql_path <- function(path) {
  sql_string(normalizePath(path, mustWork = FALSE))
}

unlink_generated <- function(path) {
  if (file.exists(path) && unlink(path) != 0L) {
    stop("failed to replace generated benchmark file: ", path, call. = FALSE)
  }
}

copy_timed <- function(
  query,
  path,
  options = "FORMAT PARQUET, COMPRESSION ZSTD"
) {
  unlink_generated(path)
  elapsed <- system.time(
    DBI::dbExecute(
      con,
      sprintf(
        "COPY (%s) TO %s (%s)",
        query,
        sql_path(path),
        options
      )
    )
  )[["elapsed"]]
  data.frame(
    path = path,
    rows = as.numeric(DBI::dbGetQuery(
      con,
      sprintf("SELECT count(*) AS n FROM read_parquet(%s)", sql_path(path))
    )$n[1L]),
    bytes = file.info(path)$size,
    seconds = unname(elapsed),
    stringsAsFactors = FALSE
  )
}

parse_time_elapsed <- function(x) {
  pieces <- as.numeric(strsplit(x, ":", fixed = TRUE)[[1L]])
  if (length(pieces) == 2L) return(pieces[1L] * 60 + pieces[2L])
  if (length(pieces) == 3L) return(pieces[1L] * 3600 + pieces[2L] * 60 + pieces[3L])
  stop("unexpected /usr/bin/time elapsed value: ", x, call. = FALSE)
}

isolated_duckdb_measure <- function(
  operation,
  sql,
  threads,
  cpus,
  repeats = 1L,
  cleanup_paths = character()
) {
  time_bin <- "/usr/bin/time"
  duckdb_cli <- Sys.which("duckdb")
  taskset_bin <- Sys.which("taskset")
  if (!file.exists(time_bin) || !nzchar(duckdb_cli) || !nzchar(taskset_bin)) {
    stop("isolated RSS measurements require /usr/bin/time, taskset, and duckdb", call. = FALSE)
  }

  rss_kb <- elapsed <- numeric(repeats)
  for (i in seq_len(repeats)) {
    unlink(cleanup_paths)
    sql_file <- tempfile(fileext = ".sql")
    time_file <- tempfile(fileext = ".time")
    stdout_file <- tempfile(fileext = ".out")
    stderr_file <- tempfile(fileext = ".err")
    on.exit(unlink(c(sql_file, time_file, stdout_file, stderr_file)), add = TRUE)
    writeLines(
      c(
        sprintf("LOAD %s;", sql_path(ext_path)),
        sprintf("PRAGMA threads=%d;", as.integer(threads)),
        sql
      ),
      sql_file
    )
    status <- system2(
      time_bin,
      c("-v", "-o", time_file, taskset_bin, "-c", cpus, duckdb_cli, "-unsigned", ":memory:"),
      stdin = sql_file,
      stdout = stdout_file,
      stderr = stderr_file
    )
    if (!identical(status, 0L)) {
      stop(
        "isolated DuckDB operation failed: ", operation, "\n",
        paste(readLines(stderr_file, warn = FALSE), collapse = "\n"),
        call. = FALSE
      )
    }
    timing <- readLines(time_file, warn = FALSE)
    rss_line <- grep("Maximum resident set size", timing, value = TRUE)
    elapsed_line <- grep("Elapsed \\(wall clock\\) time", timing, value = TRUE)
    if (length(rss_line) != 1L || length(elapsed_line) != 1L) {
      stop("failed to parse /usr/bin/time output for ", operation, call. = FALSE)
    }
    rss_kb[i] <- as.numeric(sub(".*: ", "", rss_line))
    elapsed[i] <- parse_time_elapsed(sub(".*: ", "", elapsed_line))
  }

  data.frame(
    operation = operation,
    threads = as.integer(threads),
    repeats = as.integer(repeats),
    peak_rss_mib = max(rss_kb) / 1024,
    median_peak_rss_mib = stats::median(rss_kb) / 1024,
    median_elapsed_seconds = stats::median(elapsed),
    stringsAsFactors = FALSE
  )
}
```

```{r synthetic-data-setup, include = FALSE}
DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TEMP TABLE vk_ann_reversible AS",
      "SELECT '1' AS chrom,",
      "       i::BIGINT AS pos,",
      "       'A' AS ref,",
      "       'C' AS alt,",
      "       variantkey('1', i, 'A', 'C') AS vk,",
      "       i::BIGINT AS payload",
      "FROM range(1, %d) t(i)"
    ),
    join_rows + 1L
  )
)
DBI::dbExecute(
  con,
  "CREATE OR REPLACE TEMP TABLE vk_probe_reversible AS SELECT * FROM vk_ann_reversible"
)

DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TEMP TABLE vk_ann_mixed AS",
      "WITH src AS (",
      "  SELECT '1' AS chrom,",
      "         i::BIGINT AS pos,",
      "         CASE WHEN i %% 5 = 0 THEN 'AAAAAAAAAAAA' ELSE 'A' END AS ref,",
      "         CASE WHEN i %% 5 = 0 THEN 'T' ELSE 'C' END AS alt,",
      "         i::BIGINT AS payload",
      "  FROM range(1, %d) t(i)",
      "), keyed AS (",
      "  SELECT src.*, variantkey(chrom, pos, ref, alt) AS vk",
      "  FROM src",
      ")",
      "SELECT keyed.*, (mod(extract_variantkey_refalt(vk), 2) = 1) AS is_hash",
      "FROM keyed"
    ),
    join_rows + 1L
  )
)
DBI::dbExecute(
  con,
  "CREATE OR REPLACE TEMP TABLE vk_probe_mixed AS SELECT * FROM vk_ann_mixed"
)

DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TEMP TABLE rk_ann AS",
      "WITH src AS (",
      "  SELECT '1' AS chrom,",
      "         ((i - 1) * 10)::BIGINT AS start,",
      "         (((i - 1) * 10) + 20)::BIGINT AS \"end\",",
      "         i::BIGINT AS label",
      "  FROM range(1, %d) t(i)",
      ")",
      "SELECT src.*, regionkey(chrom, start, \"end\") AS rk FROM src"
    ),
    interval_rows + 1L
  )
)
DBI::dbExecute(con, "CREATE OR REPLACE TEMP TABLE rk_probe_eq AS SELECT * FROM rk_ann")

DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TABLE ov_subject AS",
      "WITH src AS (",
      "  SELECT '1' AS chrom,",
      "         ((i - 1) * 20)::BIGINT AS start,",
      "         (((i - 1) * 20) + 30)::BIGINT AS \"end\",",
      "         i::BIGINT AS label",
      "  FROM range(1, %d) t(i)",
      "), keyed AS (",
      "  SELECT src.*, regionkey(chrom, start, \"end\") AS rk",
      "  FROM src",
      ")",
      "SELECT keyed.*,",
      "       rk >> 31 AS rk_chrom_start,",
      "       regionkey(chrom, \"end\", \"end\") >> 31 AS rk_chrom_end",
      "FROM keyed",
      "ORDER BY rk"
    ),
    interval_rows + 1L
  )
)
DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TABLE ov_probe AS",
      "WITH src AS (",
      "  SELECT '1' AS chrom,",
      "         (((i - 1) * 20) + 10)::BIGINT AS start,",
      "         (((i - 1) * 20) + 15)::BIGINT AS \"end\",",
      "         i::BIGINT AS probe_id",
      "  FROM range(1, %d) t(i)",
      ")",
      "SELECT src.*,",
      "       regionkey(chrom, start, \"end\") AS rk,",
      "       regionkey(chrom, start, start) >> 31 AS rk_chrom_start,",
      "       regionkey(chrom, \"end\", \"end\") >> 31 AS rk_chrom_end",
      "FROM src",
      "ORDER BY rk"
    ),
    interval_rows + 1L
  )
)

idx_name <- "vk_join_overlap_idx"
try(DBI::dbGetQuery(con, sprintf("SELECT duckhts_cgranges_destroy('%s') AS ok", idx_name)), silent = TRUE)
DBI::dbGetQuery(
  con,
  paste0(
    "SELECT duckhts_cgranges_from_query(",
    sql_string(idx_name), ", ",
    sql_string('SELECT chrom, start, "end", label FROM ov_subject'), ", ",
    sql_string("chrom"), ", ",
    sql_string("start"), ", ",
    sql_string("end"), ", ",
    sql_string("label"),
    ") AS ok"
  )
)
DBI::dbGetQuery(con, sprintf("SELECT duckhts_cgranges_index('%s') AS ok", idx_name))
```

## Benchmark settings

```{r benchmark-settings}
benchmark_settings
```

```{r benchmark-environment}
benchmark_environment <- data.frame(
  field = c("source_revision", "cpu", "process_affinity", "duckdb_version"),
  value = c(
    source_revision,
    cpu_model,
    cpu_affinity,
    DBI::dbGetQuery(con, "SELECT version() AS v")$v[1L]
  ),
  stringsAsFactors = FALSE
)
benchmark_environment

```

## 1. Exact reversible VariantKey joins

This section benchmarks exact self-joins over normalized reversible variants.

```{r reversible-joins}
reversible_multicol <- measure_thread_grid(
  "reversible_multicolumn_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_reversible p",
        "JOIN vk_ann_reversible a",
        "  ON p.chrom = a.chrom AND p.pos = a.pos AND p.ref = a.ref AND p.alt = a.alt"
      )
    )
  }
)

reversible_stored_vk <- measure_thread_grid(
  "reversible_stored_vk_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      "SELECT count(*) AS n FROM vk_probe_reversible p JOIN vk_ann_reversible a USING (vk)"
    )
  }
)

reversible_on_the_fly_vk <- measure_thread_grid(
  "reversible_on_the_fly_vk_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_reversible p",
        "JOIN vk_ann_reversible a",
        "  ON variantkey(p.chrom, p.pos, p.ref, p.alt) = a.vk"
      )
    )
  }
)

reversible_join_results <- rbind(
  reversible_multicol,
  reversible_stored_vk,
  reversible_on_the_fly_vk
)
reversible_join_results
```

## 2. Mixed joins with hashed-row refinement

This section benchmarks the safe join policy for mixed data:

- join on stored `vk`
- refine only when the probe row is hashed / nonreversible

```{r mixed-joins}
mixed_multicol <- measure_thread_grid(
  "mixed_multicolumn_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_mixed p",
        "JOIN vk_ann_mixed a",
        "  ON p.chrom = a.chrom AND p.pos = a.pos AND p.ref = a.ref AND p.alt = a.alt"
      )
    )
  }
)

mixed_stored_vk <- measure_thread_grid(
  "mixed_stored_vk_join_only",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      "SELECT count(*) AS n FROM vk_probe_mixed p JOIN vk_ann_mixed a USING (vk)"
    )
  }
)

mixed_safe_refine <- measure_thread_grid(
  "mixed_stored_vk_join_plus_hash_refine",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_mixed p",
        "JOIN vk_ann_mixed a",
        "  ON p.vk = a.vk",
        " AND (NOT p.is_hash OR (p.chrom = a.chrom AND p.pos = a.pos AND p.ref = a.ref AND p.alt = a.alt))"
      )
    )
  }
)

mixed_join_results <- rbind(
  mixed_multicol,
  mixed_stored_vk,
  mixed_safe_refine
)
mixed_join_results
```

## 3. Exact RegionKey span joins

Exact equality is only one RegionKey operation. RegionKey also sorts by
chromosome and start coordinate, can be binary-searched by that prefix, and
provides exact overlap predicates. The next section measures those properties
without claiming that a bare 64-bit key is an augmented interval tree.

```{r regionkey-joins}
region_multicol <- measure_thread_grid(
  "region_multicolumn_join",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM rk_probe_eq p",
        "JOIN rk_ann a",
        "  ON p.chrom = a.chrom AND p.start = a.start AND p.\"end\" = a.\"end\""
      )
    )
  }
)

region_rk <- measure_thread_grid(
  "region_stored_rk_join",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      "SELECT count(*) AS n FROM rk_probe_eq p JOIN rk_ann a USING (rk)"
    )
  }
)

region_join_results <- rbind(region_multicol, region_rk)
region_join_results
```

## 4. Interval overlap implementations

The subject intervals have a declared maximum span of 30 bases. That lets the
RegionKey query derive a correct lower start bound, range-search the sortable
chromosome/start prefix, and apply `are_overlapping_regionkeys(...)` to the
candidate rows. The maximum span is part of the index receipt: omitting it
would miss long intervals that start before the query.

The same probes and subjects are also evaluated through two vanilla DuckDB
forms and the immutable cgranges index. Because this isolated distribution has
one chromosome, the direct start/end query needs only the two inequalities and
plans as DuckDB's `IE_JOIN`. Packing chromosome and position into the
RegionKey-compatible 33-bit coordinate generalizes that two-range plan across
chromosomes. All pair-producing methods must return the same overlap count.

```{r duckdb-regionkey-overlap}
duckdb_interval_overlap <- measure_thread_grid(
  "duckdb_single_contig_interval_iejoin_pairs",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM ov_probe p",
        "JOIN ov_subject s",
        "  ON p.start < s.\"end\"",
        " AND p.\"end\" > s.start"
      )
    )
  }
)

duckdb_packed_interval_overlap <- measure_thread_grid(
  "duckdb_packed_coordinate_iejoin_pairs",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM ov_probe p",
        "JOIN ov_subject s",
        "  ON p.rk_chrom_start < s.rk_chrom_end",
        " AND p.rk_chrom_end > s.rk_chrom_start"
      )
    )
  }
)

regionkey_bounded_overlap <- measure_thread_grid(
  "regionkey_bounded_start_range_pairs",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM ov_probe p",
        "JOIN ov_subject s",
        "  ON s.rk_chrom_start >=",
        "     (regionkey(p.chrom, greatest(0, p.start - 30 + 1), greatest(0, p.start - 30 + 1)) >> 31)",
        " AND s.rk_chrom_start < (regionkey(p.chrom, p.\"end\", p.\"end\") >> 31)",
        " AND are_overlapping_regionkeys(p.rk, s.rk)"
      )
    )
  }
)
```

```{r cgranges-overlap}
cgranges_has_overlap <- measure_thread_grid(
  "cgranges_has_overlap_filter",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n",
          "FROM ov_probe p",
          "WHERE duckhts_cgranges_has_overlap('%s', p.chrom, p.start, p.\"end\")"
        ),
        idx_name
      )
    )
  }
)

cgranges_count_overlap <- measure_thread_grid(
  "cgranges_count_overlaps_sum",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT sum(duckhts_cgranges_count_overlaps('%s', p.chrom, p.start, p.\"end\"))::BIGINT AS n",
          "FROM ov_probe p"
        ),
        idx_name
      )
    )
  }
)

cgranges_overlap_results <- rbind(cgranges_has_overlap, cgranges_count_overlap)
cgranges_overlap_results
```

```{r overlap-correctness}
overlap_pair_results <- rbind(
  duckdb_interval_overlap,
  duckdb_packed_interval_overlap,
  regionkey_bounded_overlap,
  cgranges_count_overlap
)

stopifnot(length(unique(overlap_pair_results$result_rows)) == 1L)
overlap_pair_results
```

## External provider composition

Real supplementary-provider joins are intentionally outside this benchmark.
The former report depended on an undeclared local mixture of ClinVar, GIAB,
Ensembl, gnomAD, REVEL, AlphaMissense, and ClinvArbitration files. A result
cannot be portable merely because paths are environment variables. This
workload returns only after one explicit cache staging command declares every
source release, access requirement, transformation, cached output, and
consumer variable.

## Notes

- The exact-join sections are the relevant comparison class for future
  `echtvar`-style annotation throughput work.
- The `mixed_stored_vk_join_plus_hash_refine` query is the safe SQL pattern for
  nonreversible VariantKey rows: equal keys still refine on exact `REF/ALT`.
- VariantKey does not normalize alleles. A real provider comparison must name
  one normalized biallelic representation before timing a join.
- RegionKey's chromosome/start ordering supports bounded range discovery; a
  general nested-interval workload needs a complete overlap index such as
  cgranges or an explicit maximum-end structure.
- DuckDB's interval join is the zero-registry SQL baseline. The RegionKey range
  query and cgranges have different preparation and repeated-query trade-offs.
  A staged real workload, not the key format alone, decides between them.
