---
title: "DuckVEP SQL transcript presentation"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
runs <- read.csv(file.path(root, "benchmarks/data/duckvep_projection.csv"),
  stringsAsFactors = FALSE, check.names = FALSE)
stopifnot(all(runs$build_binding != "diagnostic_unbound"))
conformance <- read.csv(file.path(root, "benchmarks/data/duckvep_projection_conformance.csv"),
  stringsAsFactors = FALSE, check.names = FALSE)
stopifnot(all(conformance$build_binding != "diagnostic_unbound"),
  all(conformance$mismatches == 0L), all(conformance$actual_pairs == conformance$expected_pairs))
```

This measures the complete 23-column result of
`duckvep_transcript_projection(events, annotations, transcripts)` as a DuckDB
temporary table. It is the SQL reference, not a fused native implementation.
The literal alleles and consequence rows are prepared before timing. The timer
includes the macro's validation, joins, coordinate projection, codon/peptide
calculation and complete output materialization. One warm-up precedes the
reported measured passes; preparation and full-output verification are excluded
from those timings.

The raw FASTA, index and minimal GFF are pinned in `r/duckhtsbench` under the
`duckvep-projection` workload. `projection_fixtures.R` derives a forward model,
a reverse model, and a three-exon model with two-base CDS phase padding from
those inputs. The unchanged independent-event generator adds 1,000 seed-173
random alleles to its targeted witnesses. Only record IDs are relabelled so
repeated source alleles retain separate event keys. Each set is repeated eight
times with distinct keys. These tiny warm models are a reproducible adapter
workload, not a genome-scale annotation or file-ingestion benchmark.

Each model/worker-count pair runs in a fresh R process. Peak RSS is the Linux
process high-water mark, including model preparation and verification; it is
not isolated DuckDB workspace usage. Output bytes are the sum of the UTF-8
lengths of the complete rows serialized as JSON, not DuckDB physical storage
size. The SHA-256 covers the schema and every typed row sorted by all columns,
including duplicates and NULLs. It must match across passes and worker counts.

## Measured materialization

```{r timings}
table <- runs[c("source_revision", "case", "threads", "input_alleles",
  "expanded_transcript_rows", "output_rows", "json_utf8_bytes", "iterations",
  "median_seconds", "min_seconds", "max_seconds", "peak_rss_bytes")]
table$source_revision <- substr(table$source_revision, 1L, 12L)
knitr::kable(table)
```

## Identity and complete output

```{r identity}
knitr::kable(unique(runs[c("source_revision", "extension_sha256", "build_binding",
  "machine", "cpu", "duckdb_version")]))
knitr::kable(unique(runs[c("case", "input_vcf_sha256", "model_gff_sha256",
  "reference_sha256", "output_sha256")]))
```

The first `f0184155b0fa` runs establish this complete-contract baseline. Later
rows retain that measurement rather than replace it. The following comparison
pairs the first and last recorded runs only when input hash, model, output hash,
copy count, worker count, DuckDB version and machine agree.

```{r comparison}
keys <- c("case", "threads", "input_vcf_sha256", "model_gff_sha256", "reference_sha256",
  "output_sha256", "copies", "input_alleles", "duckdb_version", "machine", "cpu")
groups <- split(runs, do.call(paste, c(runs[keys], sep = "|")))
comparison <- lapply(groups, function(rows) {
  if (nrow(rows) < 2L) return(NULL)
  first <- rows[1L, ]
  last <- rows[nrow(rows), ]
  data.frame(case = last$case, threads = last$threads,
    baseline_revision = substr(first$source_revision, 1L, 12L),
    current_revision = substr(last$source_revision, 1L, 12L),
    baseline_seconds = first$median_seconds, current_seconds = last$median_seconds,
    current_over_baseline = last$median_seconds / first$median_seconds)
})
knitr::kable(do.call(rbind, comparison), row.names = FALSE)
```

The [sorted-stream throughput report](duckvep_throughput.md) instead measures
consequence/HGVS output and is not an identical comparator. No fused
implementation is measured here. A production-density model and comparison
with a future fused implementation remain unmeasured; small timing differences
on this warm adapter workload are not proof of a speedup or absence of regression.

## Pinned VEP field differential

The separate differential checks all 21 presentation fields, plus the complete
event/transcript key set, against VEP 116 JSON. Six models add a partial-CDS end,
a noncoding first exon and a noncoding transcript to the three timed models.
Each seed runs the unchanged generator's targeted cases plus 1,000 random cases;
the 128-base differing-allele campaign supplements the default 10-base campaign
to exercise longer exon/intron spans. Every physical input record is retained,
including duplicate alleles. Missing or duplicate keys, changed fields, and
changes to/from NULL must all trip deliberate verifier controls. Actual and
expected tables, original and relabelled VCFs, GFFs and oracle JSON have separate
hashes in the checked-in CSV; detailed artifacts remain in the run directory.

```{r conformance}
knitr::kable(conformance[c("source_revision", "case", "seed", "max_random_length",
  "input_alleles", "expected_pairs", "actual_pairs", "mismatches", "detected_controls")])
```

The oracle sources are Ensembl VEP
`57ea5c52340acc1f156267f810ad162e26597082` and Variation
`2fb834b987ede3824e200197a838ce11e91aeb4b`, with the exact package set in
`test/duckvep/upstream/receipts/vep116_2026-07-22.conda-explicit.txt`.
GFF input does not import the full core transcript-quality and translation-edit
attributes. Positive CDS quality flags and reference peptide edits are therefore
tested as SQL model-attribute properties, not represented as positive GFF oracle
coverage. Symbolic structural and phased events are outside this function's
literal-allele contract; existing consequence/HGVS/phased acceptance histories
and generator settings are unchanged.

Reproduce from a clean source revision with an extension build receipt:

```{r reproduce, eval=FALSE, echo=TRUE}
source("scripts/benchmark_duckvep_projection.R")
benchmark_duckvep_projection(
  normalizePath("."), "build/release/duckhts.duckdb_extension",
  "benchmarks/data/duckvep_projection.csv",
  extension_receipt = Sys.getenv("DUCKHTS_EXTENSION_RECEIPT")
)
rmarkdown::render("benchmarks/benchmark_duckvep_projection.Rmd",
  knit_root_dir = normalizePath("."))
```

For each reported seed and maximum length, record conformance separately:

```bash
make test-duckvep-projection ARGS="--seed 173 --random-cases 1000 --max-random-length 10 --extension-receipt $DUCKHTS_EXTENSION_RECEIPT --summary-output benchmarks/data/duckvep_projection_conformance.csv"
```

`VEP_PREFIX` names the environment matching the checked-in package lock.
