---
title: "DuckHTS VariantKey Benchmark and Conformance"
output: github_document
---

<!-- benchmark_variantkey_conformance.md is generated from benchmark_variantkey_conformance.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark and conformance

This report separates three different workloads so the full conformance harness
is not mistaken for raw VariantKey encode throughput:

1. pure in-memory VariantKey encoding
2. real `read_bcf(...)` scan + VariantKey encoding
3. full DuckHTS-vs-bcftools `%VKX` conformance

DuckHTS uses the vendored official VariantKey C API and follows the bcftools
`%VKX` / `+add-variantkey` convention of accepting 1-based VCF `POS` at the SQL
boundary while encoding the upstream 0-based field internally.

VariantKey / RegionKey are based on Nicola Asuni's work:
<https://doi.org/10.1101/473744>.

## Run

```sh
make bench-variantkey
```

Useful overrides:

- `BCFTOOLS_BIN`: optional bcftools executable override
- `VARIANTKEY_REAL_CASES`: optional comma-separated local VCF/BCF paths
- `VARIANTKEY_OUT_DIR`: optional directory for emitted TSVs and summaries
- `DUCKHTS_EXTENSION`: optional extension path override
- `VARIANTKEY_BENCH_ROWS`: synthetic benchmark rows, default `4000000`
- `VARIANTKEY_BENCH_RUNS`: timed repeats for the benchmark sections, default `3`
- `VARIANTKEY_BENCH_CASE`: optional real VCF/BCF path for the scan+encode section
- `VARIANTKEY_DUCK_THREADS_LIST`: DuckDB thread grid for the benchmark sections, default `1,2,4`
- `VARIANTKEY_CONFORMANCE_DUCK_THREADS`: DuckDB threads for the full conformance harness, default `max(VARIANTKEY_DUCK_THREADS_LIST)`
- `VARIANTKEY_DUCK_DECOMPRESSION_THREADS`: `read_bcf(...)` worker threads, default `0`

If `VARIANTKEY_REAL_CASES` is unset, this report looks for the locally
downloaded whole-WGS GIAB files already used in earlier DuckHTS conformance
work under `/root/giab_norm/`.

```{r setup, include = FALSE}
library(utils)
library(DBI)
library(duckdb)

if (!requireNamespace("RBCFTools", quietly = TRUE)) {
  stop("RBCFTools is required for VariantKey conformance rendering", call. = FALSE)
}

RBCFTools::setup_hts_env()

bcftools_bin <- Sys.getenv("BCFTOOLS_BIN", unset = RBCFTools::bcftools_path())
if (!file.exists(bcftools_bin)) stop("bcftools binary not found: ", bcftools_bin, call. = FALSE)

ext_path <- Sys.getenv(
  "DUCKHTS_EXTENSION",
  unset = normalizePath("build/release/duckhts.duckdb_extension", mustWork = TRUE)
)
out_root <- Sys.getenv("VARIANTKEY_OUT_DIR", unset = file.path(tempdir(), "duckhts_variantkey_real"))
dir.create(out_root, recursive = TRUE, showWarnings = FALSE)

bench_rows <- as.integer(Sys.getenv("VARIANTKEY_BENCH_ROWS", unset = "4000000"))
bench_runs <- as.integer(Sys.getenv("VARIANTKEY_BENCH_RUNS", unset = "3"))
duck_decompression_threads <- as.integer(Sys.getenv("VARIANTKEY_DUCK_DECOMPRESSION_THREADS", unset = "0"))
bench_case_env <- Sys.getenv("VARIANTKEY_BENCH_CASE", unset = "")
thread_grid_env <- Sys.getenv("VARIANTKEY_DUCK_THREADS_LIST", unset = "1,2,4")

parse_thread_grid <- function(x) {
  vals <- trimws(strsplit(x, ",", fixed = TRUE)[[1]])
  vals <- vals[nzchar(vals)]
  out <- unique(as.integer(vals))
  if (!length(out) || any(is.na(out)) || any(out < 1L)) {
    stop("VARIANTKEY_DUCK_THREADS_LIST must be a comma-separated list of positive integers", call. = FALSE)
  }
  out
}

thread_grid <- parse_thread_grid(thread_grid_env)
conformance_threads <- as.integer(Sys.getenv("VARIANTKEY_CONFORMANCE_DUCK_THREADS", unset = as.character(max(thread_grid))))
if (is.na(conformance_threads) || conformance_threads < 1L) {
  stop("VARIANTKEY_CONFORMANCE_DUCK_THREADS must be positive", call. = FALSE)
}

if (is.na(bench_rows) || bench_rows < 1L) stop("VARIANTKEY_BENCH_ROWS must be positive", call. = FALSE)
if (is.na(bench_runs) || bench_runs < 1L) stop("VARIANTKEY_BENCH_RUNS must be positive", call. = FALSE)
if (is.na(duck_decompression_threads) || duck_decompression_threads < 0L) {
  stop("VARIANTKEY_DUCK_DECOMPRESSION_THREADS must be a non-negative integer", call. = FALSE)
}

case_env <- Sys.getenv("VARIANTKEY_REAL_CASES", unset = "")
default_cases <- data.frame(
  label = c("giab_hg001_grch37_full", "giab_hg002_grch37_full", "giab_hg006_grch37_full"),
  path = c(
    "/root/giab_norm/HG001_GRCh37_1_22_v4.2.1_benchmark.vcf.gz",
    "/root/giab_norm/HG002_GRCh37_1_22_v4.2.1_benchmark.vcf.gz",
    "/root/giab_norm/HG006_GRCh37_1_22_v4.2.1_benchmark.vcf.gz"
  ),
  stringsAsFactors = FALSE
)

if (nzchar(case_env)) {
  raw_paths <- trimws(strsplit(case_env, ",", fixed = TRUE)[[1]])
  raw_paths <- raw_paths[nzchar(raw_paths)]
  cases <- data.frame(
    label = sub("\\.(vcf|bcf)(\\.gz)?$", "", basename(raw_paths), ignore.case = TRUE),
    path = raw_paths,
    stringsAsFactors = FALSE
  )
} else {
  cases <- default_cases[file.exists(default_cases$path), , drop = FALSE]
}

if (!nrow(cases)) {
  stop(
    "No real GIAB inputs found. Set VARIANTKEY_REAL_CASES or place local files under /root/giab_norm/.",
    call. = FALSE
  )
}

bench_case_path <- if (nzchar(bench_case_env)) bench_case_env else cases$path[1]
if (!file.exists(bench_case_path)) stop("Benchmark case not found: ", bench_case_path, call. = FALSE)
bench_case_label <- cases$label[match(bench_case_path, cases$path)]
if (is.na(bench_case_label)) {
  bench_case_label <- sub("\\.(vcf|bcf)(\\.gz)?$", "", basename(bench_case_path), ignore.case = TRUE)
}

benchmark_settings <- data.frame(
  setting = c(
    "synthetic_rows",
    "benchmark_runs",
    "duckdb_threads_grid",
    "full_conformance_duckdb_threads",
    "read_bcf_decompression_threads",
    "scan_encode_case"
  ),
  value = c(
    as.character(bench_rows),
    as.character(bench_runs),
    paste(thread_grid, collapse = ","),
    as.character(conformance_threads),
    as.character(duck_decompression_threads),
    bench_case_label
  ),
  stringsAsFactors = FALSE
)

drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
con <- dbConnect(drv, dbdir = ":memory:")
DBI::dbExecute(con, sprintf("LOAD '%s';", gsub("\\\\", "/", ext_path)))
```

```{r helpers, include = FALSE}
rows_per_second <- function(rows, seconds) {
  rows / seconds
}

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

shell_quote <- function(x) shQuote(x, type = "sh")

set_duck_threads <- function(n) {
  DBI::dbExecute(con, sprintf("PRAGMA threads=%d;", as.integer(n)))
}

run_bash <- function(cmd) {
  stderr_file <- tempfile("duckhts_variantkey_bench_stderr_")
  script_file <- tempfile("duckhts_variantkey_bench_", fileext = ".sh")
  writeLines(c("#!/usr/bin/env bash", "set -euo pipefail", cmd), script_file)
  Sys.chmod(script_file, mode = "0700")
  on.exit(unlink(c(stderr_file, script_file)), add = TRUE)

  status <- system2("bash", script_file, stdout = FALSE, stderr = stderr_file)
  stderr <- if (file.exists(stderr_file)) readLines(stderr_file, warn = FALSE) else character()
  if (!identical(status, 0L)) {
    stop(paste(c("command failed:", cmd, stderr), collapse = "\n"), call. = FALSE)
  }
  invisible(NULL)
}

measure_runs <- function(fun, runs = bench_runs) {
  times <- numeric(runs)
  value <- NULL
  for (i in seq_len(runs)) {
    gc()
    times[i] <- system.time({
      value <- fun()
    })[["elapsed"]]
  }
  list(
    value = value,
    times = times,
    median_seconds = unname(stats::median(times))
  )
}

measure_thread_grid <- function(operation, fun) {
  do.call(
    rbind,
    lapply(thread_grid, function(th) {
      set_duck_threads(th)
      m <- measure_runs(fun)
      data.frame(
        threads = th,
        operation = operation,
        rows = m$value$n[1],
        median_seconds = m$median_seconds,
        rows_per_second = rows_per_second(m$value$n[1], m$median_seconds),
        stringsAsFactors = FALSE
      )
    })
  )
}

run_case <- function(label, path) {
  set_duck_threads(conformance_threads)

  case_out <- file.path(out_root, label)
  dir.create(case_out, recursive = TRUE, showWarnings = FALSE)

  bcftools_tsv <- file.path(case_out, sprintf("%s.bcftools.tsv", label))
  duckhts_tsv <- file.path(case_out, sprintf("%s.duckhts.tsv", label))
  compare_tsv <- file.path(case_out, sprintf("%s.compare.tsv", label))
  summary_tsv <- file.path(case_out, sprintf("%s.summary.tsv", label))

  elapsed <- system.time({
    writeLines("CHROM\tPOS\tREF\tALT\tVKX", bcftools_tsv)
    run_bash(sprintf(
      "%s query -f '%%CHROM\\t%%POS\\t%%REF\\t%%ALT{0}\\t%%VKX\\n' %s >> %s",
      shell_quote(bcftools_bin),
      shell_quote(path),
      shell_quote(bcftools_tsv)
    ))

    DBI::dbExecute(
      con,
      paste0(
        "COPY (",
        "SELECT CHROM, POS, REF, coalesce(ALT[1], '.') AS ALT, ",
        "variantkey_hex(variantkey(CHROM, POS, REF, coalesce(ALT[1], '.'))) AS VKX ",
        "FROM read_bcf(", sql_string(path), ", decompression_threads := ", duck_decompression_threads, ") ",
        "ORDER BY CHROM, POS, REF, ALT",
        ") TO ", sql_string(duckhts_tsv), " (FORMAT CSV, DELIMITER '\\t', HEADER TRUE)"
      )
    )

    DBI::dbExecute(
      con,
      paste0(
        "COPY (",
        "WITH bcftools_rows AS (",
        "  SELECT CHROM, POS, REF, ALT, VKX, COUNT(*) AS n ",
        "  FROM read_csv(", sql_string(bcftools_tsv), ", header := TRUE, delim := '\\t', ",
        "                columns := {'CHROM':'VARCHAR','POS':'BIGINT','REF':'VARCHAR','ALT':'VARCHAR','VKX':'VARCHAR'}) ",
        "  GROUP BY ALL",
        "), duckhts_rows AS (",
        "  SELECT CHROM, POS, REF, ALT, VKX, COUNT(*) AS n ",
        "  FROM read_csv(", sql_string(duckhts_tsv), ", header := TRUE, delim := '\\t', ",
        "                columns := {'CHROM':'VARCHAR','POS':'BIGINT','REF':'VARCHAR','ALT':'VARCHAR','VKX':'VARCHAR'}) ",
        "  GROUP BY ALL",
        ") ",
        "SELECT coalesce(b.CHROM, d.CHROM) AS CHROM, ",
        "       coalesce(b.POS, d.POS) AS POS, ",
        "       coalesce(b.REF, d.REF) AS REF, ",
        "       coalesce(b.ALT, d.ALT) AS ALT, ",
        "       coalesce(b.VKX, d.VKX) AS VKX, ",
        "       coalesce(b.n, 0) AS bcftools_n, ",
        "       coalesce(d.n, 0) AS duckhts_n ",
        "FROM bcftools_rows b FULL OUTER JOIN duckhts_rows d USING (CHROM, POS, REF, ALT, VKX) ",
        "WHERE coalesce(b.n, 0) <> coalesce(d.n, 0) ",
        "ORDER BY 1, 2, 3, 4, 5",
        ") TO ", sql_string(compare_tsv), " (FORMAT CSV, DELIMITER '\\t', HEADER TRUE)"
      )
    )

    summary <- DBI::dbGetQuery(
      con,
      paste0(
        "WITH bcftools_rows AS (",
        "  SELECT COUNT(*) AS n FROM read_csv(", sql_string(bcftools_tsv), ", header := TRUE, delim := '\\t', ",
        "    columns := {'CHROM':'VARCHAR','POS':'BIGINT','REF':'VARCHAR','ALT':'VARCHAR','VKX':'VARCHAR'})",
        "), duckhts_rows AS (",
        "  SELECT COUNT(*) AS n FROM read_csv(", sql_string(duckhts_tsv), ", header := TRUE, delim := '\\t', ",
        "    columns := {'CHROM':'VARCHAR','POS':'BIGINT','REF':'VARCHAR','ALT':'VARCHAR','VKX':'VARCHAR'})",
        "), mismatch_rows AS (",
        "  SELECT COUNT(*) AS n FROM read_csv(", sql_string(compare_tsv), ", header := TRUE, delim := '\\t', ",
        "    columns := {'CHROM':'VARCHAR','POS':'BIGINT','REF':'VARCHAR','ALT':'VARCHAR','VKX':'VARCHAR','bcftools_n':'BIGINT','duckhts_n':'BIGINT'})",
        ") ",
        "SELECT (SELECT n FROM bcftools_rows) AS bcftools_rows, ",
        "       (SELECT n FROM duckhts_rows) AS duckhts_rows, ",
        "       (SELECT n FROM mismatch_rows) AS mismatch_groups"
      )
    )
    utils::write.table(summary, summary_tsv, sep = "\t", quote = FALSE, row.names = FALSE)
  })[["elapsed"]]

  summary <- read.delim(summary_tsv, sep = "\t", check.names = FALSE)
  data.frame(
    label = label,
    path = path,
    duckdb_threads = conformance_threads,
    bcftools_rows = summary$bcftools_rows[1],
    duckhts_rows = summary$duckhts_rows[1],
    mismatch_groups = summary$mismatch_groups[1],
    elapsed_seconds = elapsed,
    rows_per_second = rows_per_second(summary$duckhts_rows[1], elapsed),
    stringsAsFactors = FALSE
  )
}
```

## Benchmark settings

```{r benchmark-settings}
benchmark_settings
```

## 1. Pure in-memory encode kernel

This section isolates the VariantKey UDF itself from HTS decoding, file I/O,
and conformance comparison work.

```{r pure-encode-kernel}
pure_kernel_results <- rbind(
  measure_thread_grid(
    "numeric_variantkey",
    function() {
      DBI::dbGetQuery(
        con,
        sprintf(
          "SELECT count(*) AS n, max(variantkey('1', i, 'A', 'C')) AS max_vk FROM range(1, %d) t(i)",
          bench_rows + 1L
        )
      )
    }
  ),
  measure_thread_grid(
    "variantkey_hex",
    function() {
      DBI::dbGetQuery(
        con,
        sprintf(
          paste(
            "SELECT count(*) AS n,",
            "max(variantkey_hex(variantkey('1', i, 'A', 'C'))) AS max_vkx",
            "FROM range(1, %d) t(i)"
          ),
          bench_rows + 1L
        )
      )
    }
  )
)

pure_kernel_results
```

## 2. Real BCF scan + encode

This section measures the cost of scanning a real indexed GIAB BCF/VCF through
`read_bcf(...)` and computing VariantKey values, without TSV materialization or
bcftools-side conformance comparison.

```{r real-scan-encode}
scan_case_sql <- sql_string(bench_case_path)
scan_results <- rbind(
  measure_thread_grid(
    "read_bcf_count",
    function() {
      DBI::dbGetQuery(
        con,
        sprintf(
          "SELECT count(*) AS n FROM read_bcf(%s, decompression_threads := %d)",
          scan_case_sql,
          duck_decompression_threads
        )
      )
    }
  ),
  measure_thread_grid(
    "read_bcf_plus_numeric_variantkey",
    function() {
      DBI::dbGetQuery(
        con,
        sprintf(
          paste(
            "SELECT count(*) AS n,",
            "max(variantkey(CHROM, POS, REF, coalesce(ALT[1], '.'))) AS max_vk",
            "FROM read_bcf(%s, decompression_threads := %d)"
          ),
          scan_case_sql,
          duck_decompression_threads
        )
      )
    }
  ),
  measure_thread_grid(
    "read_bcf_plus_variantkey_hex",
    function() {
      DBI::dbGetQuery(
        con,
        sprintf(
          paste(
            "SELECT count(*) AS n,",
            "max(variantkey_hex(variantkey(CHROM, POS, REF, coalesce(ALT[1], '.')))) AS max_vkx",
            "FROM read_bcf(%s, decompression_threads := %d)"
          ),
          scan_case_sql,
          duck_decompression_threads
        )
      )
    }
  )
)
scan_results <- cbind(case = bench_case_label, scan_results, stringsAsFactors = FALSE)
scan_results
```

## 3. Full DuckHTS vs bcftools conformance harness

This final section is the end-to-end parity workflow. It includes both sides of
the comparison and is therefore much slower than the pure encode kernel.

```{r run-real-cases}
results <- do.call(
  rbind,
  lapply(seq_len(nrow(cases)), function(i) run_case(cases$label[i], cases$path[i]))
)
results
```

All listed cases should have `mismatch_groups == 0`.

```{r stop-on-mismatch, include = FALSE}
if (any(results$mismatch_groups != 0)) {
  stop("VariantKey conformance mismatches detected; inspect VARIANTKEY_OUT_DIR artifacts.", call. = FALSE)
}
```

## Notes

- For paper-style raw VariantKey throughput comparisons, use the **pure
  in-memory encode kernel** section rather than the full conformance harness.
- The **real BCF scan + encode** section includes HTS decoding and DuckDB table
  function overhead, so it is the right comparison for DuckHTS-as-reader
  performance rather than bare header-only VariantKey microbenchmarks.
- The **full conformance harness** includes bcftools `%VKX` export, DuckHTS
  export, TSV materialization, and rowwise comparison, so it should be read as
  a parity-validation workflow, not as a bare encoding benchmark.
- This framework compares against bcftools `%VKX`, which is the same VariantKey
  hexadecimal output exposed by `+add-variantkey`.
- DuckHTS currently compares `ALT[1]` from `read_bcf(...)`, mirroring the
  biallelic `%VKX` field that bcftools derives from the first ALT allele on the
  current record.
- Large, ambiguous, and symbolic alleles are expected to fall back to the
  official hashed nonreversible VariantKey mode. That still preserves bcftools
  `%VKX` parity, but it does not encode `END`, `SVLEN`, mate breakend
  coordinates, or other SV metadata.
