---
title: "Indexed BCF/VCF scan initialization: small-fixture cost"
output:
  github_document:
    html_preview: false
---

This paired measurement exercises automatic contig scanning with a usable
worker index. Each committed fixture contains two records on one occupied
contig and 256 empty header contigs. It measures setup and materialization
latency, not cohort throughput. The separate SQL/R regressions corrupt or remove
a private index after preparing a query and require an error instead of zero rows.

The nearest larger full-column report is
[`benchmark_reader_allocations.md`](benchmark_reader_allocations.md):
4,048,342 GIAB HG002 records in wide/tidy form, one sequential scan handle and
zero HTSlib decompression workers. It does not exercise automatic parallel
index reloads and is not an identical-workload baseline. This first small-fixture
scan-init report measures the pre-fix binary alongside the candidate.

## Reproduction and identity

Set `BCF_SCAN_BASELINE_EXTENSION` and `BCF_SCAN_BASELINE_REVISION` to a separately
built baseline. `DUCKHTS_EXTENSION` selects the candidate. From the repository root:

```sh
taskset -c 0,2,4,6 Rscript -e 'rmarkdown::render("benchmarks/benchmark_bcf_scan_init.Rmd")'
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r measurement}
baseline <- Sys.getenv("BCF_SCAN_BASELINE_EXTENSION")
baseline_revision <- Sys.getenv("BCF_SCAN_BASELINE_REVISION")
candidate <- Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension")
stopifnot(nzchar(baseline_revision), file.exists(baseline), file.exists(candidate))
stopifnot(system2("git", c("diff", "--quiet", "HEAD", "--", "src")) == 0L)
identity <- data.frame(
  property = c("baseline revision", "candidate revision", "candidate src tree",
               "baseline binary MD5", "candidate binary MD5", "R", "DuckDB R package",
               "HTSlib", "CPU affinity", "repetitions"),
  value = c(baseline_revision, system2("git", c("rev-parse", "HEAD"), stdout = TRUE),
            system2("git", c("rev-parse", "HEAD:src"), stdout = TRUE),
            unname(tools::md5sum(baseline)), unname(tools::md5sum(candidate)),
            R.version.string, as.character(packageVersion("duckdb")),
            readLines("third_party/htslib/VERSION", n = 1),
            system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE),
            "9 timed calls after warm-up; fresh R process per build/input/thread setting")
)
knitr::kable(identity)
inputs <- paste0("test/data/parallel_empty_contigs.", c("bcf", "vcf.gz"))
indexes <- paste0(inputs, c(".csi", ".tbi"))
stopifnot(all(file.exists(indexes)),
          sum(grepl("^##contig=", readLines("test/data/parallel_empty_contigs.vcf"))) == 257L)
knitr::kable(data.frame(file = c(inputs, indexes), bytes = file.info(c(inputs, indexes))$size,
                        md5 = unname(tools::md5sum(c(inputs, indexes)))))
results <- list()
expected <- NULL
for (input in inputs) for (threads in c(1L, 4L)) for (build in c("baseline", "candidate")) {
  measured <- callr::r(function(extension, input, threads) {
    con <- DBI::dbConnect(duckdb::duckdb(config = list(allow_unsigned_extensions = "true")))
    on.exit(DBI::dbDisconnect(con, shutdown = TRUE))
    DBI::dbExecute(con, paste("LOAD", DBI::dbQuoteString(con, extension)))
    DBI::dbExecute(con, sprintf("SET threads=%d", threads))
    sql <- sprintf("CREATE OR REPLACE TEMP TABLE scanned AS SELECT * FROM read_bcf(%s, decompression_threads:=0)",
                   DBI::dbQuoteString(con, input))
    DBI::dbExecute(con, sql)
    expected <- DBI::dbGetQuery(con, "SELECT * FROM scanned ORDER BY ALL")
    stopifnot(nrow(expected) == 2L, identical(as.numeric(expected$POS), c(10, 40)))
    timings <- numeric(9)
    for (i in seq_along(timings)) {
      start <- Sys.time()
      DBI::dbExecute(con, sql)
      timings[i] <- as.numeric(difftime(Sys.time(), start, units = "secs"))
      stopifnot(identical(DBI::dbGetQuery(con, "SELECT * FROM scanned ORDER BY ALL"), expected))
    }
    list(rows = expected, timings = timings)
  }, args = list(extension = normalizePath(if (build == "baseline") baseline else candidate),
                 input = normalizePath(input), threads = threads))
  if (is.null(expected)) expected <- measured$rows
  stopifnot(identical(expected, measured$rows))
  results[[length(results) + 1L]] <- data.frame(
    build, format = if (endsWith(input, ".bcf")) "BCF/CSI" else "VCF.gz/TBI",
    duckdb_threads = threads, htslib_workers_per_handle = 0L,
    header_contigs = 257L, input_records = 2L, materialized_rows = nrow(measured$rows),
    materialized_columns = ncol(measured$rows), median_ms = 1000 * median(measured$timings),
    min_ms = 1000 * min(measured$timings), max_ms = 1000 * max(measured$timings)
  )
}
knitr::kable(do.call(rbind, results), digits = 3)
```

Each timed `CREATE OR REPLACE TEMP TABLE AS SELECT *` produces two complete
materialized rows, not merely a count. Elapsed `Sys.time()` includes the DBI call;
connection creation, warm-up, and complete typed-row verification are outside the
timer. All repetitions, formats, thread settings and builds agree exactly on those
rows, including GT and ALT, without asserting scan arrival order.

This does not measure the latency of an index failure, remote I/O, large-cohort
throughput, RSS, or maximum worker concurrency. Thread settings are DuckDB's
requested limits. Small timings can vary with scheduling; no general speedup or
no-regression claim follows from them.
