---
title: "Tabix GFF/GTF single-split reader benchmark"
output:
  github_document:
    html_preview: false
    pandoc_args: ["--wrap=preserve", "--columns=400"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

This report compares identical release builds of origin/develop and the single-split
reader on the same host. Each query returns one aggregate row but hashes every
requested column so DuckDB must materialize it. The two builds must agree on
row count, XOR fingerprint and hash sum for each workload. Warm-up, load, and staging are
outside the five timed repetitions. The input is GENCODE mouse vM25 BASIC GFF3
and GTF; the BED-like file is the first 100,000 GFF3 data records rearranged
into 16 columns (chromosome, zero-based start, end, feature, score, strand,
source, phase, attributes, then seven repeated source fields). The `gff_plain`
projection selects `seqname`, `start` and `end`; `gff_five` selects `seqname`,
`start`, `end`, `feature` and `strand`. `gff_all_nine` and `gtf_all_nine`
select all nine physical columns. The BED-like query selects all 16 fields.
This derived file tests generic column projection, not BED semantics or the
BED reader.

## Reproduction

The inputs are registered under workload `tabix-single-split` in
`r/duckhtsbench/inst/benchmark_registry.tsv`. Stage with network access only
in the explicit staging command; `--offline` validates both source checksums
and reconstructs the derived input from an already-cached GFF3:

``` sh
Rscript scripts/stage_tabix_split.R
Rscript scripts/stage_tabix_split.R --offline
```

Build origin/develop from a source archive in a separate directory with
`make configure && make release -j4` (copy the pinned extension-ci-tools
submodule into that directory before building).
Set `DUCKHTS_TABIX_BASELINE_EXTENSION` to its release binary and
`DUCKHTS_TABIX_BASELINE_REVISION` to its exact commit; the candidate defaults
to `build/release/duckhts.duckdb_extension`. Render with
`Rscript -e 'rmarkdown::render("benchmarks/benchmark_tabix_split.Rmd")'`.
The `default` condition leaves DuckDB's `threads` setting untouched. Both
conditions run on the same process affinity (reported below); the reader is
not a parallel table scan, so the thread conditions are not a scaling claim.

```{r benchmark, echo=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
})
Sys.setenv(DUCKHTSBENCH_REGISTRY = normalizePath(
  "r/duckhtsbench/inst/benchmark_registry.tsv", mustWork = TRUE
))
ids <- c("tabix_split_gff3", "tabix_split_gtf", "tabix_split_bed")
paths <- stats::setNames(vapply(ids, duckhtsbench::duckhts_bench_artifact_path,
                                character(1)), ids)
stopifnot(all(file.exists(paths)))
for (id in ids) duckhtsbench::duckhts_bench_validate_identity(id, paths[[id]])
paths <- stats::setNames(normalizePath(paths), ids)

baseline <- Sys.getenv("DUCKHTS_TABIX_BASELINE_EXTENSION")
baseline_revision <- Sys.getenv("DUCKHTS_TABIX_BASELINE_REVISION")
candidate <- Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension")
stopifnot(nzchar(baseline_revision), file.exists(baseline), file.exists(candidate))
binaries <- c(develop = normalizePath(baseline), branch = normalizePath(candidate))
source_revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE)
source_tree <- system2("git", c("rev-parse", "HEAD:src"), stdout = TRUE)
affinity <- system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE)
knitr::kable(data.frame(
  build = names(binaries), revision = c(baseline_revision, source_revision),
  binary = unname(binaries)
), caption = "Release binaries (built from the named revisions).")
cat("Host:", Sys.info()[["nodename"]], "; affinity:", affinity, "; candidate src tree:", source_tree, "\n")

quote_path <- function(path) paste0("'", gsub("'", "''", path, fixed = TRUE), "'")
queries <- c(
  gff_plain = sprintf(
    paste0('SELECT count(*) AS rows, hex(bit_xor(hash(seqname, start, "end"))) AS digest, ',
           'sum(hash(seqname, start, "end")) AS digest_sum ',
           'FROM read_gff(%s, scan_mode := \'sequential\')'),
    quote_path(paths[["tabix_split_gff3"]])
  ),
  gff_five = sprintf(
    paste0('SELECT count(*) AS rows, hex(bit_xor(hash(seqname, start, "end", feature, strand))) AS digest, ',
           'sum(hash(seqname, start, "end", feature, strand)) AS digest_sum ',
           'FROM read_gff(%s, scan_mode := \'sequential\')'),
    quote_path(paths[["tabix_split_gff3"]])
  ),
  gff_all_nine = sprintf(
    paste0('SELECT count(*) AS rows, hex(bit_xor(hash(seqname, source, feature, start, "end", score, strand, frame, attributes))) AS digest, ',
           'sum(hash(seqname, source, feature, start, "end", score, strand, frame, attributes)) AS digest_sum ',
           'FROM read_gff(%s, scan_mode := \'sequential\')'),
    quote_path(paths[["tabix_split_gff3"]])
  ),
  gtf_all_nine = sprintf(
    paste0('SELECT count(*) AS rows, hex(bit_xor(hash(seqname, source, feature, start, "end", score, strand, frame, attributes))) AS digest, ',
           'sum(hash(seqname, source, feature, start, "end", score, strand, frame, attributes)) AS digest_sum ',
           'FROM read_gtf(%s, scan_mode := \'sequential\')'),
    quote_path(paths[["tabix_split_gtf"]])
  ),
  tabix_bed16 = sprintf(
    paste0('SELECT count(*) AS rows, hex(bit_xor(hash(',
           paste(sprintf('column%d', 0:15), collapse = ', '),
           '))) AS digest, sum(hash(',
           paste(sprintf('column%d', 0:15), collapse = ', '),
           ')) AS digest_sum FROM read_tabix(%s, scan_mode := \'sequential\')'),
    quote_path(paths[["tabix_split_bed"]])
  )
)
inputs <- c(gff_plain = "tabix_split_gff3", gff_five = "tabix_split_gff3",
            gff_all_nine = "tabix_split_gff3",
            gtf_all_nine = "tabix_split_gtf", tabix_bed16 = "tabix_split_bed")
measure <- function(extension, threads, sql) {
  config <- list(allow_unsigned_extensions = "true")
  if (identical(threads, "1")) config$threads <- "1"
  con <- dbConnect(duckdb(config = config))
  on.exit(dbDisconnect(con, shutdown = TRUE))
  dbExecute(con, paste("LOAD", quote_path(extension)))
  actual_threads <- dbGetQuery(con, "SELECT current_setting('threads') AS threads")$threads[[1L]]
  expected <- dbGetQuery(con, sql)
  elapsed <- numeric(5L)
  for (i in seq_along(elapsed)) {
    started <- proc.time()[["elapsed"]]
    observed <- dbGetQuery(con, sql)
    elapsed[[i]] <- proc.time()[["elapsed"]] - started
    stopifnot(identical(observed, expected))
  }
  list(result = expected, seconds = stats::median(elapsed), threads = actual_threads)
}
results <- list()
for (threads in c("1", "default")) {
  for (name in names(queries)) {
    reference <- measure(binaries[["develop"]], threads, queries[[name]])
    candidate_run <- measure(binaries[["branch"]], threads, queries[[name]])
    stopifnot(identical(reference$result, candidate_run$result))
    for (build in names(binaries)) {
      run <- if (build == "develop") reference else candidate_run
      results[[length(results) + 1L]] <- data.frame(
        workload = name, build = build, threads = threads,
        effective_threads = run$threads, input_bytes = file.info(paths[[inputs[[name]]]])$size,
        input_rows = format(run$result$rows[[1L]], scientific = FALSE, trim = TRUE),
        result_rows = nrow(run$result),
        digest = run$result$digest[[1L]], median_seconds = run$seconds
      )
    }
  }
}
result <- do.call(rbind, results)
knitr::kable(result[, c("workload", "build", "threads", "effective_threads",
                        "input_bytes", "input_rows", "result_rows", "median_seconds")],
             digits = 3, caption = "Five-run median wall time (seconds); output fingerprints and hash sums match across builds.")
```

The raw input checksum, byte count and derivation are pinned in the registry
and validated before measurement. `input_rows` is the count returned by each
scan; `result_rows` counts aggregate rows materialized by the query. Times
include scanning, decompression of GENCODE inputs, aggregation and DuckDB
execution but exclude opening the connection and loading the extension.
Fingerprint and hash-sum equality check output values, not ordering; no
output rows are filtered by the benchmark.
