---
title: "DuckHTS VariantKey Join and Overlap Benchmark"
output: github_document
---

<!-- benchmark_variantkey_join_overlap.md is generated from benchmark_variantkey_join_overlap.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This report focuses on the workloads that matter for a future DuckVEP-style
annotation layer:

1. exact variant joins with stored `VariantKey`
2. safe mixed joins where hashed/nonreversible rows refine on exact alleles
3. exact span joins with stored `RegionKey`
4. interval overlap through RegionKey-ordered candidate ranges, DuckDB range
   joins, and `duckhts_cgranges_*`

This is intentionally separate from `benchmark_variantkey_conformance.Rmd`.
That report measures encoding and `%VKX` parity. This one measures join and
overlap throughput.

## Run

```sh
make bench-variantkey-join
```

Useful overrides:

- `VARIANTKEY_JOIN_ROWS`: synthetic exact-join rows, default `1000000`
- `VARIANTKEY_INTERVAL_ROWS`: synthetic interval rows, default `250000`
- `VARIANTKEY_JOIN_RUNS`: timed repeats, default `3`
- `VARIANTKEY_JOIN_THREADS_LIST`: DuckDB thread grid, default `1,2,4`
- `DUCKHTS_EXTENSION`: optional extension path override
- `VARIANTKEY_CLINVAR_VCF`, `VARIANTKEY_GIAB_VCF`, and provider variables: cache-path overrides
- `VARIANTKEY_REAL_CACHE_DIR`: generated benchmark Parquet directory

Stage the provider workload explicitly before rendering:

```sh
make stage-variantkey-providers
```

The command stages declared public raw sources and exports provider relations
from the registered model cache entry. It fails closed if that model is absent:
the public Ensembl 116 model producer is tracked in
<https://github.com/RGenomicsETL/duckhts/issues/179>, rather than being
replaced by an unreceipted URL.

```{r setup, include = FALSE}
library(DBI)
library(duckdb)

ext_path <- Sys.getenv(
  "DUCKHTS_EXTENSION",
  unset = normalizePath("build/release/duckhts.duckdb_extension", mustWork = TRUE)
)

join_rows <- as.integer(Sys.getenv("VARIANTKEY_JOIN_ROWS", unset = "1000000"))
interval_rows <- as.integer(Sys.getenv("VARIANTKEY_INTERVAL_ROWS", unset = "250000"))
bench_runs <- as.integer(Sys.getenv("VARIANTKEY_JOIN_RUNS", unset = "3"))
thread_grid_env <- Sys.getenv("VARIANTKEY_JOIN_THREADS_LIST", unset = "1,2,4")
registry_source <- normalizePath("r/duckhtsbench/R/registry.R", mustWork = TRUE)
if (!nzchar(Sys.getenv("DUCKHTSBENCH_REGISTRY", unset = ""))) {
  Sys.setenv(DUCKHTSBENCH_REGISTRY = normalizePath("r/duckhtsbench/inst/benchmark_registry.tsv", mustWork = TRUE))
}
source(registry_source)
artifact_path <- duckhts_bench_artifact_path

clinvar_vcf <- Sys.getenv("VARIANTKEY_CLINVAR_VCF", unset = artifact_path("variantkey_clinvar_20260706"))
giab_vcf <- Sys.getenv("VARIANTKEY_GIAB_VCF", unset = artifact_path("variantkey_giab_hg002_v421"))
reference_fasta <- Sys.getenv("VARIANTKEY_REFERENCE_FASTA", unset = artifact_path("ensembl116_grch38_fasta_fa"))
regulatory_parquet <- Sys.getenv("VARIANTKEY_REGULATORY_PARQUET", unset = artifact_path("duckvep_ensembl116_regulatory_parquet"))
transcripts_parquet <- Sys.getenv("VARIANTKEY_TRANSCRIPTS_PARQUET", unset = artifact_path("duckvep_ensembl116_transcripts_parquet"))
gnomad_constraint <- Sys.getenv("VARIANTKEY_GNOMAD_CONSTRAINT", unset = artifact_path("gnomad_v211_constraint"))
real_cache_dir <- Sys.getenv("VARIANTKEY_REAL_CACHE_DIR", unset = dirname(regulatory_parquet))
revel_parquet <- Sys.getenv("VARIANTKEY_REVEL_PARQUET", unset = artifact_path("revel_v13_grch37"))
alphamissense_gz <- Sys.getenv("VARIANTKEY_ALPHAMISSENSE_GZ", unset = artifact_path("alphamissense_v2_hg38"))
clinvarbitration_tsv <- Sys.getenv("VARIANTKEY_CLINVARBITRATION_TSV", unset = artifact_path("clinvarbitration_202508"))

real_source_paths <- c(
  clinvar_vcf = clinvar_vcf,
  giab_vcf = giab_vcf,
  reference_fasta = reference_fasta,
  regulatory_parquet = regulatory_parquet,
  transcripts_parquet = transcripts_parquet,
  gnomad_constraint = gnomad_constraint,
  revel_v13_derived_parquet = revel_parquet,
  alphamissense_v2_hg38 = alphamissense_gz,
  clinvarbitration_zenodo_16792026 = clinvarbitration_tsv
)
missing_real_sources <- real_source_paths[!file.exists(real_source_paths)]
if (length(missing_real_sources)) {
  stop(
    "missing staged benchmark sources: ",
    paste(names(missing_real_sources), collapse = ", "),
    ". Run `make stage-variantkey-providers`; it reports any unavailable registered model.",
    call. = FALSE
  )
}
dir.create(real_cache_dir, recursive = TRUE, showWarnings = FALSE)

parse_thread_grid <- function(x) {
  vals <- trimws(strsplit(x, ",", fixed = TRUE)[[1]])
  vals <- vals[nzchar(vals)]
  out <- unique(as.integer(vals))
  if (!length(out) || any(is.na(out)) || any(out < 1L)) {
    stop("VARIANTKEY_JOIN_THREADS_LIST must be a comma-separated list of positive integers", call. = FALSE)
  }
  out
}

thread_grid <- parse_thread_grid(thread_grid_env)

if (is.na(join_rows) || join_rows < 1L) stop("VARIANTKEY_JOIN_ROWS must be positive", call. = FALSE)
if (is.na(interval_rows) || interval_rows < 1L) stop("VARIANTKEY_INTERVAL_ROWS must be positive", call. = FALSE)
if (is.na(bench_runs) || bench_runs < 1L) stop("VARIANTKEY_JOIN_RUNS must be positive", call. = FALSE)

benchmark_settings <- data.frame(
  setting = c(
    "variant_join_rows",
    "interval_rows",
    "benchmark_runs",
    "duckdb_threads_grid",
    "mixed_hash_fraction"
  ),
  value = c(
    as.character(join_rows),
    as.character(interval_rows),
    as.character(bench_runs),
    paste(thread_grid, collapse = ","),
    "20%"
  ),
  stringsAsFactors = FALSE
)

source_revision <- trimws(system2("git", c("rev-parse", "HEAD"), stdout = TRUE))
cpu_model <- sub(
  "^model name\\s*:\\s*",
  "",
  grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[1L]
)
cpu_affinity <- trimws(system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE))

drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
con <- dbConnect(drv, dbdir = ":memory:")
DBI::dbExecute(con, sprintf("LOAD '%s';", gsub("\\\\", "/", ext_path)))
```

```{r helpers, include = FALSE}
rows_per_second <- function(rows, seconds) {
  rows / seconds
}

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

set_duck_threads <- function(n) {
  DBI::dbExecute(con, sprintf("PRAGMA threads=%d;", as.integer(n)))
}

measure_runs <- function(fun, runs = bench_runs) {
  times <- numeric(runs)
  value <- NULL
  for (i in seq_len(runs)) {
    gc()
    times[i] <- system.time({
      value <- fun()
    })[["elapsed"]]
  }
  list(
    value = value,
    times = times,
    median_seconds = unname(stats::median(times))
  )
}

measure_thread_grid <- function(operation, rows, fun, runs = bench_runs) {
  do.call(
    rbind,
    lapply(thread_grid, function(th) {
      set_duck_threads(th)
      m <- measure_runs(fun, runs = runs)
      result_rows <- as.numeric(m$value[[1L]][1L])
      data.frame(
        operation = operation,
        threads = th,
        input_rows = rows,
        result_rows = result_rows,
        median_seconds = m$median_seconds,
        input_rows_per_second = rows_per_second(rows, m$median_seconds),
        result_rows_per_second = rows_per_second(result_rows, m$median_seconds),
        stringsAsFactors = FALSE
      )
    })
  )
}

sha256_file <- function(path) {
  strsplit(system2("sha256sum", path, stdout = TRUE), "\\s+")[[1L]][1L]
}

sql_path <- function(path) {
  sql_string(normalizePath(path, mustWork = FALSE))
}

unlink_generated <- function(path) {
  if (file.exists(path) && unlink(path) != 0L) {
    stop("failed to replace generated benchmark file: ", path, call. = FALSE)
  }
}

copy_timed <- function(
  query,
  path,
  options = "FORMAT PARQUET, COMPRESSION ZSTD"
) {
  unlink_generated(path)
  elapsed <- system.time(
    DBI::dbExecute(
      con,
      sprintf(
        "COPY (%s) TO %s (%s)",
        query,
        sql_path(path),
        options
      )
    )
  )[["elapsed"]]
  data.frame(
    path = path,
    rows = as.numeric(DBI::dbGetQuery(
      con,
      sprintf("SELECT count(*) AS n FROM read_parquet(%s)", sql_path(path))
    )$n[1L]),
    bytes = file.info(path)$size,
    seconds = unname(elapsed),
    stringsAsFactors = FALSE
  )
}

parse_time_elapsed <- function(x) {
  pieces <- as.numeric(strsplit(x, ":", fixed = TRUE)[[1L]])
  if (length(pieces) == 2L) return(pieces[1L] * 60 + pieces[2L])
  if (length(pieces) == 3L) return(pieces[1L] * 3600 + pieces[2L] * 60 + pieces[3L])
  stop("unexpected /usr/bin/time elapsed value: ", x, call. = FALSE)
}

isolated_duckdb_measure <- function(
  operation,
  sql,
  threads,
  cpus,
  repeats = 1L,
  cleanup_paths = character()
) {
  time_bin <- "/usr/bin/time"
  duckdb_cli <- Sys.which("duckdb")
  taskset_bin <- Sys.which("taskset")
  if (!file.exists(time_bin) || !nzchar(duckdb_cli) || !nzchar(taskset_bin)) {
    stop("isolated RSS measurements require /usr/bin/time, taskset, and duckdb", call. = FALSE)
  }

  rss_kb <- elapsed <- numeric(repeats)
  for (i in seq_len(repeats)) {
    unlink(cleanup_paths)
    sql_file <- tempfile(fileext = ".sql")
    time_file <- tempfile(fileext = ".time")
    stdout_file <- tempfile(fileext = ".out")
    stderr_file <- tempfile(fileext = ".err")
    on.exit(unlink(c(sql_file, time_file, stdout_file, stderr_file)), add = TRUE)
    writeLines(
      c(
        sprintf("LOAD %s;", sql_path(ext_path)),
        sprintf("PRAGMA threads=%d;", as.integer(threads)),
        sql
      ),
      sql_file
    )
    status <- system2(
      time_bin,
      c("-v", "-o", time_file, taskset_bin, "-c", cpus, duckdb_cli, "-unsigned", ":memory:"),
      stdin = sql_file,
      stdout = stdout_file,
      stderr = stderr_file
    )
    if (!identical(status, 0L)) {
      stop(
        "isolated DuckDB operation failed: ", operation, "\n",
        paste(readLines(stderr_file, warn = FALSE), collapse = "\n"),
        call. = FALSE
      )
    }
    timing <- readLines(time_file, warn = FALSE)
    rss_line <- grep("Maximum resident set size", timing, value = TRUE)
    elapsed_line <- grep("Elapsed \\(wall clock\\) time", timing, value = TRUE)
    if (length(rss_line) != 1L || length(elapsed_line) != 1L) {
      stop("failed to parse /usr/bin/time output for ", operation, call. = FALSE)
    }
    rss_kb[i] <- as.numeric(sub(".*: ", "", rss_line))
    elapsed[i] <- parse_time_elapsed(sub(".*: ", "", elapsed_line))
  }

  data.frame(
    operation = operation,
    threads = as.integer(threads),
    repeats = as.integer(repeats),
    peak_rss_mib = max(rss_kb) / 1024,
    median_peak_rss_mib = stats::median(rss_kb) / 1024,
    median_elapsed_seconds = stats::median(elapsed),
    stringsAsFactors = FALSE
  )
}
```

```{r synthetic-data-setup, include = FALSE}
DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TEMP TABLE vk_ann_reversible AS",
      "SELECT '1' AS chrom,",
      "       i::BIGINT AS pos,",
      "       'A' AS ref,",
      "       'C' AS alt,",
      "       variantkey('1', i, 'A', 'C') AS vk,",
      "       i::BIGINT AS payload",
      "FROM range(1, %d) t(i)"
    ),
    join_rows + 1L
  )
)
DBI::dbExecute(
  con,
  "CREATE OR REPLACE TEMP TABLE vk_probe_reversible AS SELECT * FROM vk_ann_reversible"
)

DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TEMP TABLE vk_ann_mixed AS",
      "WITH src AS (",
      "  SELECT '1' AS chrom,",
      "         i::BIGINT AS pos,",
      "         CASE WHEN i %% 5 = 0 THEN 'AAAAAAAAAAAA' ELSE 'A' END AS ref,",
      "         CASE WHEN i %% 5 = 0 THEN 'T' ELSE 'C' END AS alt,",
      "         i::BIGINT AS payload",
      "  FROM range(1, %d) t(i)",
      "), keyed AS (",
      "  SELECT src.*, variantkey(chrom, pos, ref, alt) AS vk",
      "  FROM src",
      ")",
      "SELECT keyed.*, (mod(extract_variantkey_refalt(vk), 2) = 1) AS is_hash",
      "FROM keyed"
    ),
    join_rows + 1L
  )
)
DBI::dbExecute(
  con,
  "CREATE OR REPLACE TEMP TABLE vk_probe_mixed AS SELECT * FROM vk_ann_mixed"
)

DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TEMP TABLE rk_ann AS",
      "WITH src AS (",
      "  SELECT '1' AS chrom,",
      "         ((i - 1) * 10)::BIGINT AS start,",
      "         (((i - 1) * 10) + 20)::BIGINT AS \"end\",",
      "         i::BIGINT AS label",
      "  FROM range(1, %d) t(i)",
      ")",
      "SELECT src.*, regionkey(chrom, start, \"end\") AS rk FROM src"
    ),
    interval_rows + 1L
  )
)
DBI::dbExecute(con, "CREATE OR REPLACE TEMP TABLE rk_probe_eq AS SELECT * FROM rk_ann")

DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TABLE ov_subject AS",
      "WITH src AS (",
      "  SELECT '1' AS chrom,",
      "         ((i - 1) * 20)::BIGINT AS start,",
      "         (((i - 1) * 20) + 30)::BIGINT AS \"end\",",
      "         i::BIGINT AS label",
      "  FROM range(1, %d) t(i)",
      "), keyed AS (",
      "  SELECT src.*, regionkey(chrom, start, \"end\") AS rk",
      "  FROM src",
      ")",
      "SELECT keyed.*,",
      "       rk >> 31 AS rk_chrom_start,",
      "       regionkey(chrom, \"end\", \"end\") >> 31 AS rk_chrom_end",
      "FROM keyed",
      "ORDER BY rk"
    ),
    interval_rows + 1L
  )
)
DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TABLE ov_probe AS",
      "WITH src AS (",
      "  SELECT '1' AS chrom,",
      "         (((i - 1) * 20) + 10)::BIGINT AS start,",
      "         (((i - 1) * 20) + 15)::BIGINT AS \"end\",",
      "         i::BIGINT AS probe_id",
      "  FROM range(1, %d) t(i)",
      ")",
      "SELECT src.*,",
      "       regionkey(chrom, start, \"end\") AS rk,",
      "       regionkey(chrom, start, start) >> 31 AS rk_chrom_start,",
      "       regionkey(chrom, \"end\", \"end\") >> 31 AS rk_chrom_end",
      "FROM src",
      "ORDER BY rk"
    ),
    interval_rows + 1L
  )
)

idx_name <- "vk_join_overlap_idx"
try(DBI::dbGetQuery(con, sprintf("SELECT duckhts_cgranges_destroy('%s') AS ok", idx_name)), silent = TRUE)
DBI::dbGetQuery(
  con,
  paste0(
    "SELECT duckhts_cgranges_from_query(",
    sql_string(idx_name), ", ",
    sql_string('SELECT chrom, start, "end", label FROM ov_subject'), ", ",
    sql_string("chrom"), ", ",
    sql_string("start"), ", ",
    sql_string("end"), ", ",
    sql_string("label"),
    ") AS ok"
  )
)
DBI::dbGetQuery(con, sprintf("SELECT duckhts_cgranges_index('%s') AS ok", idx_name))
```

## Benchmark settings

```{r benchmark-settings}
benchmark_settings
```

```{r benchmark-environment}
benchmark_environment <- data.frame(
  field = c("source_revision", "cpu", "process_affinity", "duckdb_version"),
  value = c(
    source_revision,
    cpu_model,
    cpu_affinity,
    DBI::dbGetQuery(con, "SELECT version() AS v")$v[1L]
  ),
  stringsAsFactors = FALSE
)
benchmark_environment

source_receipts <- data.frame(
  source = names(real_source_paths),
  path = unname(real_source_paths),
  bytes = unname(file.info(real_source_paths)$size),
  sha256 = vapply(real_source_paths, sha256_file, character(1L)),
  stringsAsFactors = FALSE
)
source_receipts
```

AlphaMissense is the official v2 GRCh38 genomic file from
[Zenodo record 8360242](https://zenodo.org/records/8360242); its published MD5
is `9fd167735f16a1b87da6eb3e4c25fcb5`. REVEL is the locally derived,
column-pruned Parquet form of the official
[REVEL v1.3 download](https://sites.google.com/site/revelgenomics/downloads).
The ClinvArbitration profile is the extracted TSV from
[Zenodo record 16792026](https://zenodo.org/records/16792026). That TSV does
not carry an assembly column; the ClinvArbitration/Talos release contract
declares GRCh38, and the provider receipt below supplies that semantic input
before keying. The fact does not become inferable from the coordinate columns.

## 1. Exact reversible VariantKey joins

This section benchmarks exact self-joins over normalized reversible variants.

```{r reversible-joins}
reversible_multicol <- measure_thread_grid(
  "reversible_multicolumn_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_reversible p",
        "JOIN vk_ann_reversible a",
        "  ON p.chrom = a.chrom AND p.pos = a.pos AND p.ref = a.ref AND p.alt = a.alt"
      )
    )
  }
)

reversible_stored_vk <- measure_thread_grid(
  "reversible_stored_vk_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      "SELECT count(*) AS n FROM vk_probe_reversible p JOIN vk_ann_reversible a USING (vk)"
    )
  }
)

reversible_on_the_fly_vk <- measure_thread_grid(
  "reversible_on_the_fly_vk_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_reversible p",
        "JOIN vk_ann_reversible a",
        "  ON variantkey(p.chrom, p.pos, p.ref, p.alt) = a.vk"
      )
    )
  }
)

reversible_join_results <- rbind(
  reversible_multicol,
  reversible_stored_vk,
  reversible_on_the_fly_vk
)
reversible_join_results
```

## 2. Mixed joins with hashed-row refinement

This section benchmarks the safe join policy for mixed data:

- join on stored `vk`
- refine only when the probe row is hashed / nonreversible

```{r mixed-joins}
mixed_multicol <- measure_thread_grid(
  "mixed_multicolumn_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_mixed p",
        "JOIN vk_ann_mixed a",
        "  ON p.chrom = a.chrom AND p.pos = a.pos AND p.ref = a.ref AND p.alt = a.alt"
      )
    )
  }
)

mixed_stored_vk <- measure_thread_grid(
  "mixed_stored_vk_join_only",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      "SELECT count(*) AS n FROM vk_probe_mixed p JOIN vk_ann_mixed a USING (vk)"
    )
  }
)

mixed_safe_refine <- measure_thread_grid(
  "mixed_stored_vk_join_plus_hash_refine",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_mixed p",
        "JOIN vk_ann_mixed a",
        "  ON p.vk = a.vk",
        " AND (NOT p.is_hash OR (p.chrom = a.chrom AND p.pos = a.pos AND p.ref = a.ref AND p.alt = a.alt))"
      )
    )
  }
)

mixed_join_results <- rbind(
  mixed_multicol,
  mixed_stored_vk,
  mixed_safe_refine
)
mixed_join_results
```

## 3. Exact RegionKey span joins

Exact equality is only one RegionKey operation. RegionKey also sorts by
chromosome and start coordinate, can be binary-searched by that prefix, and
provides exact overlap predicates. The next section measures those properties
without claiming that a bare 64-bit key is an augmented interval tree.

```{r regionkey-joins}
region_multicol <- measure_thread_grid(
  "region_multicolumn_join",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM rk_probe_eq p",
        "JOIN rk_ann a",
        "  ON p.chrom = a.chrom AND p.start = a.start AND p.\"end\" = a.\"end\""
      )
    )
  }
)

region_rk <- measure_thread_grid(
  "region_stored_rk_join",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      "SELECT count(*) AS n FROM rk_probe_eq p JOIN rk_ann a USING (rk)"
    )
  }
)

region_join_results <- rbind(region_multicol, region_rk)
region_join_results
```

## 4. Interval overlap implementations

The subject intervals have a declared maximum span of 30 bases. That lets the
RegionKey query derive a correct lower start bound, range-search the sortable
chromosome/start prefix, and apply `are_overlapping_regionkeys(...)` to the
candidate rows. The maximum span is part of the index receipt: omitting it
would miss long intervals that start before the query.

The same probes and subjects are also evaluated through two vanilla DuckDB
forms and the immutable cgranges index. Because this isolated distribution has
one chromosome, the direct start/end query needs only the two inequalities and
plans as DuckDB's `IE_JOIN`. Packing chromosome and position into the
RegionKey-compatible 33-bit coordinate generalizes that two-range plan across
chromosomes. All pair-producing methods must return the same overlap count.

```{r duckdb-regionkey-overlap}
duckdb_interval_overlap <- measure_thread_grid(
  "duckdb_single_contig_interval_iejoin_pairs",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM ov_probe p",
        "JOIN ov_subject s",
        "  ON p.start < s.\"end\"",
        " AND p.\"end\" > s.start"
      )
    )
  }
)

duckdb_packed_interval_overlap <- measure_thread_grid(
  "duckdb_packed_coordinate_iejoin_pairs",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM ov_probe p",
        "JOIN ov_subject s",
        "  ON p.rk_chrom_start < s.rk_chrom_end",
        " AND p.rk_chrom_end > s.rk_chrom_start"
      )
    )
  }
)

regionkey_bounded_overlap <- measure_thread_grid(
  "regionkey_bounded_start_range_pairs",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM ov_probe p",
        "JOIN ov_subject s",
        "  ON s.rk_chrom_start >=",
        "     (regionkey(p.chrom, greatest(0, p.start - 30 + 1)::BIGINT, greatest(0, p.start - 30 + 1)::BIGINT) >> 31)",
        " AND s.rk_chrom_start < (regionkey(p.chrom, p.\"end\"::BIGINT, p.\"end\"::BIGINT) >> 31)",
        " AND are_overlapping_regionkeys(p.rk, s.rk)"
      )
    )
  }
)
```

```{r cgranges-overlap}
cgranges_has_overlap <- measure_thread_grid(
  "cgranges_has_overlap_filter",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n",
          "FROM ov_probe p",
          "WHERE duckhts_cgranges_has_overlap('%s', p.chrom, p.start, p.\"end\")"
        ),
        idx_name
      )
    )
  }
)

cgranges_count_overlap <- measure_thread_grid(
  "cgranges_count_overlaps_sum",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT sum(duckhts_cgranges_count_overlaps('%s', p.chrom, p.start::BIGINT, p.\"end\"::BIGINT))::BIGINT AS n",
          "FROM ov_probe p"
        ),
        idx_name
      )
    )
  }
)

cgranges_overlap_results <- rbind(cgranges_has_overlap, cgranges_count_overlap)
cgranges_overlap_results
```

```{r overlap-correctness}
overlap_pair_results <- rbind(
  duckdb_interval_overlap,
  duckdb_packed_interval_overlap,
  regionkey_bounded_overlap,
  cgranges_count_overlap
)

stopifnot(length(unique(overlap_pair_results$result_rows)) == 1L)
overlap_pair_results
```

## 5. Real supplementary-annotation composition

The synthetic sections isolate operator costs. This section exercises the
storage and join architecture on real release data:

- 2026-07-06 ClinVar GRCh38 as an exact and positional provider;
- GIAB HG002 GRCh38 v4.2.1 as a whole-genome probe relation;
- Ensembl 116 regulatory features as the interval provider; and
- gnomAD v2.1.1 constraint as a gene provider joined to the full Ensembl 116
  transcript relation;
- REVEL v1.3's 77.97 million GRCh37 SNV scores; and
- the official AlphaMissense v2 GRCh38 genomic release's 71.70 million rows.

ClinvArbitration Zenodo record 16792026 is also exercised as a GRCh38 exact
provider. Its TSV has 3.65 million rows and omits an assembly column; GRCh38 is
therefore an explicit receipt fact, not a guessed property of `chr`-prefixed
coordinates.

VCF is decoded once into narrow, typed, ZSTD Parquet. VariantKey and RegionKey
are prepared columns, not recomputed for every downstream provider join. The
Parquet files are generated benchmark artifacts and are not committed.

```{r real-parquet-staging}
real_case_path <- file.path(real_cache_dir, "giab_hg002_grch38_keyed.parquet")
real_case_raw_path <- file.path(real_cache_dir, "giab_hg002_grch38_raw_keyed.parquet")
real_clinvar_path <- file.path(real_cache_dir, "clinvar_20260706_grch38_keyed.parquet")
real_regulatory_path <- file.path(real_cache_dir, "ensembl116_grch38_regulatory_keyed.parquet")
real_constraint_path <- file.path(real_cache_dir, "gnomad_v211_constraint_gene.parquet")
real_case_reversible_path <- file.path(real_cache_dir, "giab_hg002_reversible.parquet")
real_case_hashed_path <- file.path(real_cache_dir, "giab_hg002_hashed.parquet")
real_clinvar_reversible_path <- file.path(real_cache_dir, "clinvar_20260706_reversible.parquet")
real_clinvar_hashed_path <- file.path(real_cache_dir, "clinvar_20260706_hashed.parquet")
real_revel_path <- file.path(real_cache_dir, "revel_grch37_variantkey.parquet")
real_revel_probe_path <- file.path(real_cache_dir, "revel_probe_hash19.parquet")
real_alphamissense_path <- file.path(real_cache_dir, "alphamissense_hg38_variantkey.parquet")
real_clinvarbitration_path <- file.path(real_cache_dir, "clinvarbitration_16792026_grch38_keyed.parquet")
real_clinvarbitration_reversible_path <- file.path(
  real_cache_dir,
  "clinvarbitration_16792026_reversible.parquet"
)
real_clinvarbitration_hashed_path <- file.path(
  real_cache_dir,
  "clinvarbitration_16792026_hashed.parquet"
)
real_clinvar_reversible_rg32k_path <- file.path(
  real_cache_dir,
  "clinvar_20260706_reversible_rg32768.parquet"
)
real_clinvar_reversible_rg1m_path <- file.path(
  real_cache_dir,
  "clinvar_20260706_reversible_rg1048576.parquet"
)

canonical_contigs_sql <- paste(
  sql_string(c(as.character(1:22), "X", "Y", "MT")),
  collapse = ", "
)

giab_raw_stage_query <- sprintf(
  paste(
    "WITH alleles AS (",
    "  SELECT duckhts_contig_key(CHROM) AS chrom, POS::BIGINT AS pos,",
    "         upper(REF) AS ref, upper(allele.alt) AS alt",
    "  FROM read_bcf(%s, scan_mode := 'sequential')",
    "  CROSS JOIN UNNEST(ALT) AS allele(alt)",
    "), keyed AS (",
    "  SELECT alleles.*, variantkey(chrom, pos, ref, alt) AS vk",
    "  FROM alleles WHERE chrom IN (%s)",
    "    AND regexp_full_match(ref, '[ACGTN]+')",
    "    AND regexp_full_match(alt, '[ACGTN]+') AND ref <> alt",
    ")",
    "SELECT *, mod(extract_variantkey_refalt(vk), 2) = 1 AS is_hash",
    "FROM keyed WHERE vk IS NOT NULL ORDER BY vk"
  ),
  sql_path(giab_vcf),
  canonical_contigs_sql
)

giab_source_query <- sprintf(
  paste(
    "(SELECT row_number() OVER ()::UBIGINT AS record_index,",
    "        duckhts_contig_key(CHROM) AS chrom, POS::BIGINT AS pos,",
    "        REF AS ref, ALT AS alt",
    " FROM read_bcf(%s, scan_mode := 'sequential')) AS giab_records"
  ),
  sql_path(giab_vcf)
)

giab_stage_query <- sprintf(
  paste(
    "WITH normalized AS (",
    "  SELECT record_index, alt_index, chrom, pos_normed::BIGINT AS pos,",
    "         upper(ref_normed) AS ref, upper(alt_normed) AS alt",
    "  FROM duckhts_bcftools_norm(%s, %s, split_multiallelic := TRUE)",
    "  WHERE pos_normed IS NOT NULL AND chrom IN (%s)",
    "    AND regexp_full_match(ref_normed, '[ACGTNacgtn]+')",
    "    AND regexp_full_match(alt_normed, '[ACGTNacgtn]+')",
    "    AND upper(ref_normed) <> upper(alt_normed)",
    "), interpreted AS (",
    "  SELECT normalized.*, duckvep_allele_geometry(pos::UBIGINT, ref, alt) AS geometry",
    "  FROM normalized",
    "), keyed AS (",
    "  SELECT interpreted.*,",
    "         variantkey(chrom, pos, ref, alt) AS vk,",
    "         regionkey(chrom, geometry.feature_start0::BIGINT, geometry.feature_end0::BIGINT) AS rk",
    "  FROM interpreted",
    ")",
    "SELECT record_index, alt_index, chrom, pos, ref, alt, vk,",
    "       mod(extract_variantkey_refalt(vk), 2) = 1 AS is_hash,",
    "       (rk >> 31) AS rk_chrom_start,",
    "       (regionkey(chrom, geometry.feature_end0::BIGINT, geometry.feature_end0::BIGINT) >> 31) AS rk_chrom_end,",
    "       geometry.raw_start0, geometry.raw_end0,",
    "       geometry.feature_start0 AS start0, geometry.feature_end0 AS end0,",
    "       geometry.edit_start0, geometry.edit_end0, geometry.interbase, rk",
    "FROM keyed",
    "WHERE vk IS NOT NULL AND rk IS NOT NULL",
    "ORDER BY vk"
  ),
  sql_string(giab_source_query),
  sql_path(reference_fasta),
  canonical_contigs_sql
)

clinvar_source_query <- sprintf(
  paste(
    "(SELECT duckhts_contig_key(CHROM) AS chrom, POS::BIGINT AS pos,",
    "        REF AS ref, ALT AS alt, INFO_ALLELEID AS allele_id,",
    "        INFO_CLNSIG AS clinical_significance, INFO_GENEINFO AS gene_info",
    " FROM read_bcf(%s, scan_mode := 'sequential')) AS clinvar_records"
  ),
  sql_path(clinvar_vcf)
)

clinvar_stage_query <- sprintf(
  paste(
    "WITH normalized AS (",
    "  SELECT chrom, pos_normed::BIGINT AS pos, upper(ref_normed) AS ref,",
    "         upper(alt_normed) AS alt, allele_id, clinical_significance, gene_info",
    "  FROM duckhts_bcftools_norm(%s, %s, split_multiallelic := TRUE)",
    "  WHERE pos_normed IS NOT NULL AND chrom IN (%s)",
    "    AND regexp_full_match(ref_normed, '[ACGTNacgtn]+')",
    "    AND regexp_full_match(alt_normed, '[ACGTNacgtn]+')",
    "), keyed AS (",
    "  SELECT normalized.*, variantkey(chrom, pos, ref, alt) AS vk",
    "  FROM normalized",
    ")",
    "SELECT chrom, pos, ref, alt, vk,",
    "       mod(extract_variantkey_refalt(vk), 2) = 1 AS is_hash,",
    "       allele_id, clinical_significance, gene_info",
    "FROM keyed",
    "WHERE vk IS NOT NULL",
    "ORDER BY vk"
  ),
  sql_string(clinvar_source_query),
  sql_path(reference_fasta),
  canonical_contigs_sql
)

regulatory_stage_query <- sprintf(
  paste(
    "WITH source AS (",
    "  SELECT stable_id, duckhts_contig_key(chrom) AS chrom,",
    "         start - 1 AS start0, \"end\" AS end0,",
    "         feature_type, so_term, so_accession",
    "  FROM read_parquet(%s)",
    "), keyed AS (",
    "  SELECT source.*, regionkey(chrom, start0::BIGINT, end0::BIGINT) AS rk",
    "  FROM source",
    "  WHERE chrom IN (%s)",
    ")",
    "SELECT keyed.*, rk >> 31 AS rk_chrom_start,",
    "       regionkey(chrom, end0::BIGINT, end0::BIGINT) >> 31 AS rk_chrom_end",
    "FROM keyed",
    "WHERE rk IS NOT NULL",
    "ORDER BY rk"
  ),
  sql_path(regulatory_parquet),
  canonical_contigs_sql
)

constraint_stage_query <- sprintf(
  paste(
    "SELECT gene_id, gene AS gene_symbol, transcript,",
    "       try_cast(pLI AS DOUBLE) AS pLI,",
    "       try_cast(oe_lof_upper AS DOUBLE) AS oe_lof_upper",
    "FROM read_csv_auto(%s, delim = '\\t', header = TRUE, nullstr = 'NA')",
    "WHERE gene_id IS NOT NULL"
  ),
  sql_path(gnomad_constraint)
)

revel_stage_query <- sprintf(
  paste(
    "WITH keyed AS (",
    "  SELECT variantkey(chrom, pos, ref, alt) AS vk, revel",
    "  FROM read_parquet(%s)",
    ")",
    "SELECT vk, revel FROM keyed WHERE vk IS NOT NULL"
  ),
  sql_path(revel_parquet)
)

alphamissense_columns_sql <- paste0(
  "{'chrom':'VARCHAR','pos':'BIGINT','ref':'VARCHAR','alt':'VARCHAR',",
  "'genome':'VARCHAR','uniprot_id':'VARCHAR','transcript_id':'VARCHAR',",
  "'protein_variant':'VARCHAR','am_pathogenicity':'DOUBLE','am_class':'VARCHAR'}"
)
alphamissense_stage_query <- sprintf(
  paste(
    "WITH keyed AS (",
    "  SELECT variantkey(duckhts_contig_key(chrom), pos, ref, alt) AS vk,",
    "         regexp_replace(transcript_id, '\\.[0-9]+$', '') AS transcript_stable_id,",
    "         try_cast(regexp_extract(transcript_id, '\\.([0-9]+)$', 1) AS UINTEGER) AS transcript_version,",
    "         transcript_id, protein_variant, uniprot_id, am_pathogenicity, am_class",
    "  FROM read_csv(%s, delim = '\\t', skip = 4, header = FALSE, columns = %s)",
    ")",
    "SELECT vk, transcript_stable_id, transcript_version, transcript_id,",
    "       protein_variant, uniprot_id, am_pathogenicity, am_class",
    "FROM keyed WHERE vk IS NOT NULL",
    "ORDER BY vk, transcript_stable_id, transcript_version, protein_variant"
  ),
  sql_path(alphamissense_gz),
  alphamissense_columns_sql
)

clinvarbitration_stage_query <- sprintf(
  paste(
    "WITH source AS (",
    "  SELECT duckhts_contig_key(contig) AS chrom,",
    "         try_cast(position AS BIGINT) AS pos,",
    "         reference AS ref, alternate AS alt,",
    "         clinical_significance, try_cast(gold_stars AS UTINYINT) AS gold_stars,",
    "         try_cast(allele_id AS BIGINT) AS allele_id",
    "  FROM read_csv_auto(%s, delim = '\\t', header = TRUE, all_varchar = TRUE)",
    "), keyed AS (",
    "  SELECT source.*, variantkey(chrom, pos, ref, alt) AS vk",
    "  FROM source",
    ")",
    "SELECT keyed.*, mod(extract_variantkey_refalt(vk), 2) = 1 AS is_hash",
    "FROM keyed WHERE vk IS NOT NULL"
  ),
  sql_path(clinvarbitration_tsv)
)

staging_results <- rbind(
  cbind(source = "GIAB HG002 raw allele keys", copy_timed(giab_raw_stage_query, real_case_raw_path)),
  cbind(source = "GIAB HG002 normalized allele keys and geometry", copy_timed(giab_stage_query, real_case_path)),
  cbind(source = "ClinVar VCF to keyed Parquet", copy_timed(clinvar_stage_query, real_clinvar_path)),
  cbind(source = "Ensembl regulatory to keyed Parquet", copy_timed(regulatory_stage_query, real_regulatory_path)),
  cbind(source = "gnomAD constraint to gene Parquet", copy_timed(constraint_stage_query, real_constraint_path)),
  cbind(
    source = "REVEL v1.3 to VariantKey Parquet",
    copy_timed(revel_stage_query, real_revel_path, "FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 122880")
  ),
  cbind(
    source = "AlphaMissense v2 to VariantKey Parquet",
    copy_timed(
      alphamissense_stage_query,
      real_alphamissense_path,
      "FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 122880"
    )
  ),
  cbind(
    source = "ClinvArbitration 16792026 to keyed Parquet",
    copy_timed(
      clinvarbitration_stage_query,
      real_clinvarbitration_path,
      "FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 122880"
    )
  )
)
staging_results$rows_per_second <- staging_results$rows / staging_results$seconds
staging_results$output_mb_per_second <- staging_results$bytes / staging_results$seconds / 1e6
staging_results

clinvarbitration_profile <- data.frame(
  source = "ClinvArbitration Zenodo 16792026 TSV",
  rows = as.numeric(DBI::dbGetQuery(
    con,
    sprintf(
      paste(
        "SELECT count(*) AS n FROM read_csv_auto(%s, delim = '\\t', header = TRUE,",
        "all_varchar = TRUE, ignore_errors = TRUE)"
      ),
      sql_path(clinvarbitration_tsv)
    )
  )$n[1L]),
  bytes = file.info(clinvarbitration_tsv)$size,
  assembly_in_relation = FALSE,
  receipt_assembly = "GRCh38",
  stringsAsFactors = FALSE
)
clinvarbitration_profile

lane_staging_results <- rbind(
  cbind(
    source = "GIAB reversible lane",
    copy_timed(
      sprintf(
        "SELECT * EXCLUDE (is_hash) FROM read_parquet(%s) WHERE NOT is_hash ORDER BY vk",
        sql_path(real_case_path)
      ),
      real_case_reversible_path
    )
  ),
  cbind(
    source = "GIAB hashed lane",
    copy_timed(
      sprintf(
        "SELECT * EXCLUDE (is_hash) FROM read_parquet(%s) WHERE is_hash ORDER BY vk",
        sql_path(real_case_path)
      ),
      real_case_hashed_path
    )
  ),
  cbind(
    source = "ClinVar reversible numeric lane",
    copy_timed(
      sprintf(
        paste(
          "SELECT vk, allele_id, clinical_significance, gene_info",
          "FROM read_parquet(%s) WHERE NOT is_hash ORDER BY vk"
        ),
        sql_path(real_clinvar_path)
      ),
      real_clinvar_reversible_path
    )
  ),
  cbind(
    source = "ClinVar hashed refinement lane",
    copy_timed(
      sprintf(
        paste(
          "SELECT chrom, pos, ref, alt, vk, allele_id, clinical_significance, gene_info",
          "FROM read_parquet(%s) WHERE is_hash ORDER BY vk"
        ),
        sql_path(real_clinvar_path)
      ),
      real_clinvar_hashed_path
    )
  ),
  cbind(
    source = "ClinvArbitration reversible numeric lane",
    copy_timed(
      sprintf(
        paste(
          "SELECT vk, allele_id, clinical_significance, gold_stars",
          "FROM read_parquet(%s) WHERE NOT is_hash ORDER BY vk"
        ),
        sql_path(real_clinvarbitration_path)
      ),
      real_clinvarbitration_reversible_path
    )
  ),
  cbind(
    source = "ClinvArbitration hashed refinement lane",
    copy_timed(
      sprintf(
        paste(
          "SELECT chrom, pos, ref, alt, vk, allele_id, clinical_significance, gold_stars",
          "FROM read_parquet(%s) WHERE is_hash ORDER BY vk"
        ),
        sql_path(real_clinvarbitration_path)
      ),
      real_clinvarbitration_hashed_path
    )
  )
)
lane_staging_results$rows_per_second <- lane_staging_results$rows / lane_staging_results$seconds
lane_staging_results$output_mb_per_second <- lane_staging_results$bytes / lane_staging_results$seconds / 1e6
lane_staging_results

revel_probe_staging <- copy_timed(
  sprintf(
    paste(
      "SELECT vk FROM read_parquet(%s)",
      "WHERE hash(vk) %% 19 = 0 ORDER BY vk"
    ),
    sql_path(real_revel_path)
  ),
  real_revel_probe_path
)
revel_probe_staging

row_group_staging_results <- rbind(
  cbind(
    row_group_size = 32768L,
    copy_timed(
      sprintf("SELECT * FROM read_parquet(%s)", sql_path(real_clinvar_reversible_path)),
      real_clinvar_reversible_rg32k_path,
      "FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 32768"
    )
  ),
  cbind(
    row_group_size = 1048576L,
    copy_timed(
      sprintf("SELECT * FROM read_parquet(%s)", sql_path(real_clinvar_reversible_path)),
      real_clinvar_reversible_rg1m_path,
      "FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 1048576"
    )
  )
)

default_row_groups <- DBI::dbGetQuery(
  con,
  sprintf(
    "SELECT count(DISTINCT row_group_id)::BIGINT AS n FROM parquet_metadata(%s)",
    sql_path(real_clinvar_reversible_path)
  )
)$n[1L]
row_group_staging_results$row_groups <- vapply(
  row_group_staging_results$path,
  function(path) {
    as.numeric(DBI::dbGetQuery(
      con,
      sprintf(
        "SELECT count(DISTINCT row_group_id)::BIGINT AS n FROM parquet_metadata(%s)",
        sql_path(path)
      )
    )$n[1L])
  },
  numeric(1L)
)
row_group_layouts <- rbind(
  data.frame(
    row_group_size = 122880L,
    path = real_clinvar_reversible_path,
    rows = lane_staging_results$rows[
      lane_staging_results$source == "ClinVar reversible numeric lane"
    ],
    bytes = file.info(real_clinvar_reversible_path)$size,
    seconds = NA_real_,
    row_groups = as.numeric(default_row_groups),
    stringsAsFactors = FALSE
  ),
  row_group_staging_results[, c(
    "row_group_size", "path", "rows", "bytes", "seconds", "row_groups"
  )]
)
row_group_layouts
```

### Exact and positional ClinVar joins

The exact join uses the collision-safe policy: reversible rows join on the
64-bit VariantKey alone; hashed/nonreversible rows additionally refine on the
normalized allele tuple. The multicolumn query is the correctness oracle. The
positional query intentionally has a different, larger result cardinality.

```{r real-exact-positional}
real_case_rows <- staging_results$rows[
  staging_results$source == "GIAB HG002 normalized allele keys and geometry"
]

real_exact_multicol <- measure_thread_grid(
  "real_clinvar_exact_multicolumn_pairs",
  real_case_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n",
          "FROM read_parquet(%s) q",
          "JOIN read_parquet(%s) a",
          "  ON q.chrom = a.chrom AND q.pos = a.pos",
          " AND q.ref = a.ref AND q.alt = a.alt"
        ),
        sql_path(real_case_path),
        sql_path(real_clinvar_path)
      )
    )
  }
)

real_exact_variantkey <- measure_thread_grid(
  "real_clinvar_exact_variantkey_refined_pairs",
  real_case_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n",
          "FROM read_parquet(%s) q",
          "JOIN read_parquet(%s) a",
          "  ON q.vk = a.vk",
          " AND q.is_hash = a.is_hash",
          " AND (NOT q.is_hash",
          "      OR (q.chrom = a.chrom AND q.pos = a.pos AND q.ref = a.ref AND q.alt = a.alt))"
        ),
        sql_path(real_case_path),
        sql_path(real_clinvar_path)
      )
    )
  }
)

real_exact_variantkey_lanes <- measure_thread_grid(
  "real_clinvar_exact_variantkey_split_lanes_pairs",
  real_case_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste0(
          "SELECT sum(n)::BIGINT AS n FROM (",
          "SELECT count(*)::BIGINT AS n ",
          "FROM read_parquet(%s) q JOIN read_parquet(%s) a USING (vk) ",
          "UNION ALL ",
          "SELECT count(*)::BIGINT AS n ",
          "FROM read_parquet(%s) q JOIN read_parquet(%s) a ",
          "ON q.vk = a.vk AND q.chrom = a.chrom AND q.pos = a.pos ",
          "AND q.ref = a.ref AND q.alt = a.alt)"
        ),
        sql_path(real_case_reversible_path),
        sql_path(real_clinvar_reversible_path),
        sql_path(real_case_hashed_path),
        sql_path(real_clinvar_hashed_path)
      )
    )
  }
)

real_positional <- measure_thread_grid(
  "real_clinvar_positional_pairs",
  real_case_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n",
          "FROM read_parquet(%s) q",
          "JOIN read_parquet(%s) a",
          "  ON q.chrom = a.chrom AND q.pos = a.pos"
        ),
        sql_path(real_case_path),
        sql_path(real_clinvar_path)
      )
    )
  }
)

real_exact_results <- rbind(
  real_exact_multicol,
  real_exact_variantkey,
  real_exact_variantkey_lanes,
  real_positional
)
stopifnot(
  length(unique(rbind(
    real_exact_multicol,
    real_exact_variantkey,
    real_exact_variantkey_lanes
  )$result_rows)) == 1L
)
real_exact_results
```

The same join is also run once on the uploaded, unnormalized GIAB alleles. This
is a semantic control, not a competing serving layout: the difference counts
real ClinVar records that an exact raw-VCF key would miss.

```{r real-normalization-effect}
raw_exact_matches <- as.numeric(DBI::dbGetQuery(
  con,
  sprintf(
    paste(
      "SELECT count(*)::BIGINT AS n",
      "FROM read_parquet(%s) q JOIN read_parquet(%s) a",
      "  ON q.vk = a.vk",
      " AND q.is_hash = a.is_hash",
      " AND (NOT q.is_hash",
      "      OR (q.chrom = a.chrom AND q.pos = a.pos",
      "          AND q.ref = a.ref AND q.alt = a.alt))"
    ),
    sql_path(real_case_raw_path),
    sql_path(real_clinvar_path)
  )
)$n[1L])
normalized_exact_matches <- unique(real_exact_multicol$result_rows)
stopifnot(length(normalized_exact_matches) == 1L,
          normalized_exact_matches >= raw_exact_matches)
normalization_match_effect <- data.frame(
  representation = c("uploaded GIAB alleles", "normalized GIAB alleles"),
  exact_clinvar_records = c(raw_exact_matches, normalized_exact_matches),
  stringsAsFactors = FALSE
)
normalization_match_effect$recovered_vs_raw <-
  normalization_match_effect$exact_clinvar_records - raw_exact_matches
normalization_match_effect
```

The split-lane layout processes `r round(real_exact_variantkey_lanes$input_rows_per_second[real_exact_variantkey_lanes$threads == 1] / 1e6, 2)` million GIAB alleles/s on one thread and `r round(real_exact_variantkey_lanes$input_rows_per_second[real_exact_variantkey_lanes$threads == 4] / 1e6, 2)` million/s on four threads. It is faster than the multicolumn oracle because only the `r round(100 * lane_staging_results$rows[lane_staging_results$source == "GIAB hashed lane"] / real_case_rows, 2)`% GIAB hashed lane and `r round(100 * lane_staging_results$rows[lane_staging_results$source == "ClinVar hashed refinement lane"] / staging_results$rows[staging_results$source == "ClinVar VCF to keyed Parquet"], 2)`% ClinVar hashed lane carry allele strings into the refinement join.

The same collision-safe split is applied to the GRCh38 ClinvArbitration release.
Its classification and star columns are provider payload; DuckHTS does not
recompute the ClinvArbitration policy in this benchmark.

```{r real-clinvarbitration}
real_clinvarbitration_exact <- measure_thread_grid(
  "ClinvArbitration_16792026_exact_split_lanes",
  real_case_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste0(
          "SELECT sum(n)::BIGINT AS n FROM (",
          "SELECT count(*)::BIGINT AS n FROM read_parquet(%s) q ",
          "JOIN read_parquet(%s) a USING (vk) UNION ALL ",
          "SELECT count(*)::BIGINT AS n FROM read_parquet(%s) q ",
          "JOIN read_parquet(%s) a ON q.vk = a.vk AND q.chrom = a.chrom ",
          "AND q.pos = a.pos AND q.ref = a.ref AND q.alt = a.alt)"
        ),
        sql_path(real_case_reversible_path),
        sql_path(real_clinvarbitration_reversible_path),
        sql_path(real_case_hashed_path),
        sql_path(real_clinvarbitration_hashed_path)
      )
    )
  }
)
real_clinvarbitration_exact
```

### Dense pathogenicity-score providers

REVEL and AlphaMissense are intentionally measured as separate exact providers,
not folded into the consequence kernel. The AlphaMissense run uses the real
GRCh38 GIAB relation. The REVEL v1.3 source is GRCh37, so its throughput probe is
a deterministic `hash(vk) % 19 = 0` sample spanning the provider's complete
chromosome/key range. This prevents a leading-key sample from making DuckDB
dynamically prune most provider row groups. It is a storage and execution
measurement, not a biological cross-build comparison.

```{r real-dense-score-providers}
alpha_probe_rows <- lane_staging_results$rows[
  lane_staging_results$source == "GIAB reversible lane"
]
revel_provider_rows <- staging_results$rows[
  staging_results$source == "REVEL v1.3 to VariantKey Parquet"
]
alpha_provider_rows <- staging_results$rows[
  staging_results$source == "AlphaMissense v2 to VariantKey Parquet"
]

real_revel_join <- measure_thread_grid(
  "REVEL_v1.3_exact_score_join",
  revel_probe_staging$rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n, sum(a.revel) AS score_sum",
          "FROM read_parquet(%s) q",
          "JOIN read_parquet(%s) a USING (vk)"
        ),
        sql_path(real_revel_probe_path),
        sql_path(real_revel_path)
      )
    )
  }
)

real_alphamissense_join <- measure_thread_grid(
  "AlphaMissense_v2_exact_score_join",
  alpha_probe_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n, sum(a.am_pathogenicity) AS score_sum",
          "FROM read_parquet(%s) q",
          "JOIN read_parquet(%s) a USING (vk)"
        ),
        sql_path(real_case_reversible_path),
        sql_path(real_alphamissense_path)
      )
    )
  }
)

dense_score_results <- rbind(real_revel_join, real_alphamissense_join)
dense_score_results
```

The slower AlphaMissense path still processes `r round(real_alphamissense_join$input_rows_per_second[real_alphamissense_join$threads == 1] / 1e6, 2)` million GIAB alleles/s on one thread and `r round(real_alphamissense_join$input_rows_per_second[real_alphamissense_join$threads == 4] / 1e6, 2)` million/s on four. The provider has `r format(alpha_provider_rows, big.mark = ",", scientific = FALSE)` rows and its hot Parquet projection contains only VariantKey, score, and class. REVEL's hot projection has `r format(revel_provider_rows, big.mark = ",", scientific = FALSE)` rows.

### Multiple exact providers in one plan

The previous measurements isolate one provider at a time. This section tests
whether ordinary DuckDB composition remains efficient when a case relation
needs several supplementary sources together. The real three-provider query
keeps the case side to one reversible `UBIGINT` key and projects one numeric
payload from each provider. The equivalent separate-query total is the sum of
three medians over the same probe, sources, thread count, and process.

Provider key scope remains a semantic contract. ClinVar is unique on this hot
key, while the prepared ClinvArbitration and AlphaMissense files contain a
small number of repeated VariantKeys. The raw join therefore measures provider
records and may return more rows than the probe. A production provider must
either declare the finer key that makes those rows distinct or publish an
explicit one-row-per-variant aggregation.

```{r real-multiple-exact-providers}
provider_key_cardinality <- DBI::dbGetQuery(
  con,
  sprintf(
    paste0(
      "SELECT 'ClinVar' AS provider, count(*)::BIGINT AS rows, ",
      "count(DISTINCT vk)::BIGINT AS distinct_keys FROM read_parquet(%s) UNION ALL ",
      "SELECT 'ClinvArbitration', count(*)::BIGINT, count(DISTINCT vk)::BIGINT ",
      "FROM read_parquet(%s) UNION ALL ",
      "SELECT 'AlphaMissense', count(*)::BIGINT, count(DISTINCT vk)::BIGINT ",
      "FROM read_parquet(%s)"
    ),
    sql_path(real_clinvar_reversible_path),
    sql_path(real_clinvarbitration_reversible_path),
    sql_path(real_alphamissense_path)
  )
)
provider_key_cardinality$duplicate_key_rows <-
  provider_key_cardinality$rows - provider_key_cardinality$distinct_keys
provider_key_cardinality

real_multi_provider_sql <- sprintf(
  paste(
    "SELECT count(*)::BIGINT AS n,",
    "count(cv.allele_id)::BIGINT AS clinvar_hits,",
    "count(ca.allele_id)::BIGINT AS clinvarbitration_hits,",
    "count(am.am_pathogenicity)::BIGINT AS alphamissense_hits,",
    "sum(am.am_pathogenicity) AS alphamissense_score_sum",
    "FROM read_parquet(%s) q",
    "LEFT JOIN read_parquet(%s) cv USING (vk)",
    "LEFT JOIN read_parquet(%s) ca USING (vk)",
    "LEFT JOIN read_parquet(%s) am USING (vk)"
  ),
  sql_path(real_case_reversible_path),
  sql_path(real_clinvar_reversible_path),
  sql_path(real_clinvarbitration_reversible_path),
  sql_path(real_alphamissense_path)
)

real_multi_provider <- measure_thread_grid(
  "GIAB_three_real_exact_providers_one_plan",
  alpha_probe_rows,
  function() DBI::dbGetQuery(con, real_multi_provider_sql)
)

measure_one_exact_provider <- function(operation, provider_path, payload) {
  measure_thread_grid(
    operation,
    alpha_probe_rows,
    function() {
      DBI::dbGetQuery(
        con,
        sprintf(
          paste(
            "SELECT count(*)::BIGINT AS n, count(p.%s)::BIGINT AS hits",
            "FROM read_parquet(%s) q",
            "LEFT JOIN read_parquet(%s) p USING (vk)"
          ),
          payload,
          sql_path(real_case_reversible_path),
          sql_path(provider_path)
        )
      )
    }
  )
}

real_multi_provider_separate <- rbind(
  measure_one_exact_provider(
    "GIAB_ClinVar_reversible_left_join",
    real_clinvar_reversible_path,
    "allele_id"
  ),
  measure_one_exact_provider(
    "GIAB_ClinvArbitration_reversible_left_join",
    real_clinvarbitration_reversible_path,
    "allele_id"
  ),
  measure_one_exact_provider(
    "GIAB_AlphaMissense_left_join",
    real_alphamissense_path,
    "am_pathogenicity"
  )
)

real_multi_provider_comparison <- aggregate(
  median_seconds ~ threads,
  data = real_multi_provider_separate,
  FUN = sum
)
names(real_multi_provider_comparison)[2L] <- "three_separate_seconds"
real_multi_provider_comparison <- merge(
  real_multi_provider_comparison,
  real_multi_provider[, c("threads", "median_seconds")],
  by = "threads",
  sort = FALSE
)
names(real_multi_provider_comparison)[3L] <- "one_plan_seconds"
real_multi_provider_comparison$one_plan_input_rows_per_second <-
  alpha_probe_rows / real_multi_provider_comparison$one_plan_seconds
real_multi_provider_comparison$speedup_vs_three_separate <-
  real_multi_provider_comparison$three_separate_seconds /
    real_multi_provider_comparison$one_plan_seconds

real_multi_provider
real_multi_provider_separate
real_multi_provider_comparison

multi_provider_plan <- DBI::dbGetQuery(
  con,
  paste("EXPLAIN", real_multi_provider_sql)
)
multi_provider_plan_text <- paste(multi_provider_plan[[2L]], collapse = "\n")
multi_provider_plan_summary <- data.frame(
  hash_join_operators = lengths(regmatches(
    multi_provider_plan_text,
    gregexpr("HASH_JOIN", multi_provider_plan_text, fixed = TRUE)
  )),
  parquet_scans = lengths(regmatches(
    multi_provider_plan_text,
    gregexpr("READ_PARQUET", multi_provider_plan_text, fixed = TRUE)
  )),
  dynamic_filter_sites = lengths(regmatches(
    multi_provider_plan_text,
    gregexpr("Dynamic Filters", multi_provider_plan_text, fixed = TRUE)
  )),
  order_operators = lengths(regmatches(
    multi_provider_plan_text,
    gregexpr("ORDER_BY", multi_provider_plan_text, fixed = TRUE)
  )),
  stringsAsFactors = FALSE
)
multi_provider_plan_summary
```

The real combined plan processes `r round(real_multi_provider_comparison$one_plan_input_rows_per_second[real_multi_provider_comparison$threads == 1] / 1e6, 2)` million GIAB alleles/s on one thread and `r round(real_multi_provider_comparison$one_plan_input_rows_per_second[real_multi_provider_comparison$threads == 4] / 1e6, 2)` million/s on four. It is `r round(real_multi_provider_comparison$speedup_vs_three_separate[real_multi_provider_comparison$threads == 1], 2)` times faster than issuing the three joins separately on one thread because the case relation is scanned once and DuckDB can plan the equality joins together.

The following 1/4/12-provider test isolates join-count overhead. It reuses the
unique real ClinVar numeric lane under independent aliases, projects one value
per alias, and forces every payload to contribute to a checksum. This is a
planner and memory stress over real Parquet pages, not a biological assertion
that twelve sources have ClinVar's density or schema.

```{r repeated-provider-join-count}
clinvar_key_check <- provider_key_cardinality[
  provider_key_cardinality$provider == "ClinVar",
]
stopifnot(clinvar_key_check$duplicate_key_rows == 0)

repeated_provider_sql <- function(provider_count) {
  aliases <- sprintf("p%02d", seq_len(provider_count))
  joins <- paste(
    sprintf(
      "LEFT JOIN read_parquet(%s) %s USING (vk)",
      sql_path(real_clinvar_reversible_path),
      aliases
    ),
    collapse = "\n"
  )
  checksum <- paste(
    sprintf("coalesce(%s.allele_id, 0)::HUGEINT", aliases),
    collapse = " + "
  )
  sprintf(
    paste(
      "SELECT count(*)::BIGINT AS n, sum(%s) AS payload_checksum",
      "FROM read_parquet(%s) q",
      "%s"
    ),
    checksum,
    sql_path(real_case_reversible_path),
    joins
  )
}

repeated_provider_counts <- c(1L, 4L, 12L)
repeated_provider_join_results <- do.call(
  rbind,
  lapply(repeated_provider_counts, function(provider_count) {
    result <- measure_thread_grid(
      sprintf("GIAB_%02d_unique_exact_provider_joins", provider_count),
      alpha_probe_rows,
      function() DBI::dbGetQuery(con, repeated_provider_sql(provider_count))
    )
    result$provider_count <- provider_count
    result
  })
)
stopifnot(all(repeated_provider_join_results$result_rows == alpha_probe_rows))
repeated_provider_join_results
```

### Row-group zone-map pruning for a local delta

Whole-genome joins necessarily visit the whole provider. Incremental
annotation is different: a new local batch can carry an explicit VariantKey
range, allowing Parquet row-group statistics to reject unrelated provider
pages. This real 5 Mb chromosome-17 slice compares three row-group sizes while
holding rows, ordering, compression, query, and result cardinality fixed.

```{r real-row-group-pruning}
delta_chrom <- "17"
delta_pos_min <- 40000000L
delta_pos_max <- 45000000L
delta_rows <- as.numeric(DBI::dbGetQuery(
  con,
  sprintf(
    paste(
      "SELECT count(*) AS n FROM read_parquet(%s)",
      "WHERE chrom = %s AND pos BETWEEN %d AND %d"
    ),
    sql_path(real_case_reversible_path),
    sql_string(delta_chrom),
    delta_pos_min,
    delta_pos_max
  )
)$n[1L])

row_group_layouts$candidate_row_groups <- vapply(
  row_group_layouts$path,
  function(path) {
    as.numeric(DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "WITH bounds AS (SELECT variantkey_range(%s, %d, %d) AS b)",
          "SELECT count(DISTINCT row_group_id)::BIGINT AS n",
          "FROM parquet_metadata(%s), bounds",
          "WHERE path_in_schema = 'vk'",
          "  AND try_cast(stats_max_value AS UBIGINT) >= b.min",
          "  AND try_cast(stats_min_value AS UBIGINT) <= b.max"
        ),
        sql_string(delta_chrom),
        delta_pos_min,
        delta_pos_max,
        sql_path(path)
      )
    )$n[1L])
  },
  numeric(1L)
)
row_group_layouts

row_group_query_results <- do.call(
  rbind,
  lapply(seq_len(nrow(row_group_layouts)), function(i) {
    provider_path <- row_group_layouts$path[i]
    measure_thread_grid(
      sprintf("clinvar_local_delta_rg%d", row_group_layouts$row_group_size[i]),
      delta_rows,
      function() {
        DBI::dbGetQuery(
          con,
          sprintf(
            paste(
              "SELECT count(*) AS n",
              "FROM read_parquet(%s) q",
              "JOIN (",
              "  SELECT vk FROM read_parquet(%s)",
              "  WHERE vk BETWEEN",
              "        (variantkey_range(%s, %d, %d)).min AND",
              "        (variantkey_range(%s, %d, %d)).max",
              ") a USING (vk)",
              "WHERE q.chrom = %s AND q.pos BETWEEN %d AND %d"
            ),
            sql_path(real_case_reversible_path),
            sql_path(provider_path),
            sql_string(delta_chrom),
            delta_pos_min,
            delta_pos_max,
            sql_string(delta_chrom),
            delta_pos_min,
            delta_pos_max,
            sql_string(delta_chrom),
            delta_pos_min,
            delta_pos_max
          )
        )
      },
      runs = max(bench_runs, 10L)
    )
  })
)
stopifnot(length(unique(row_group_query_results$result_rows)) == 1L)
row_group_query_results
```

The default 122,880-row layout admits `r row_group_layouts$candidate_row_groups[row_group_layouts$row_group_size == 122880]` of `r row_group_layouts$row_groups[row_group_layouts$row_group_size == 122880]` provider row groups and answers the 6,023-allele local delta in `r row_group_query_results$median_seconds[row_group_query_results$operation == "clinvar_local_delta_rg122880" & row_group_query_results$threads == 1]` seconds on one thread. The 1,048,576-row layout also admits one row group, but that group contains much more unrelated data and is slower. Smaller row groups do not help this 5 Mb query enough to offset their metadata and scheduling overhead.

### Regulatory interval joins

Ensembl coordinates are converted from one-based inclusive to zero-based
half-open before RegionKey construction. The maximum source span is read from
the staged relation and used as the proven RegionKey look-back bound. A
per-contig vanilla SQL query exposes typed start/end inequalities directly to
DuckDB's interval join, while the packed-coordinate form exposes the same
problem in one numeric domain.

```{r real-regulatory-index-setup}
real_regulatory_stats <- DBI::dbGetQuery(
  con,
  sprintf(
    paste(
      "SELECT count(*) AS rows, max(end0 - start0)::BIGINT AS max_span,",
      "       sum(end0 - start0)::UBIGINT AS summed_span",
      "FROM read_parquet(%s)"
    ),
    sql_path(real_regulatory_path)
  )
)
real_regulatory_stats

real_idx_name <- "variantkey_real_regulatory_idx"
try(
  DBI::dbGetQuery(con, sprintf("SELECT duckhts_cgranges_destroy('%s') AS ok", real_idx_name)),
  silent = TRUE
)
real_index_query <- sprintf(
  "SELECT chrom, start0, end0, stable_id FROM read_parquet(%s)",
  sql_path(real_regulatory_path)
)
real_index_seconds <- system.time({
  DBI::dbGetQuery(
    con,
    paste0(
      "SELECT duckhts_cgranges_from_query(",
      sql_string(real_idx_name), ", ",
      sql_string(real_index_query), ", ",
      sql_string("chrom"), ", ", sql_string("start0"), ", ",
      sql_string("end0"), ", ", sql_string("stable_id"),
      ") AS ok"
    )
  )
  DBI::dbGetQuery(con, sprintf("SELECT duckhts_cgranges_index('%s') AS ok", real_idx_name))
})[["elapsed"]]

data.frame(
  operation = "cgranges_build_and_finalize",
  intervals = real_regulatory_stats$rows,
  seconds = unname(real_index_seconds),
  intervals_per_second = real_regulatory_stats$rows / real_index_seconds
)
```

```{r real-regulatory-overlap}
real_contigs <- DBI::dbGetQuery(
  con,
  sprintf("SELECT DISTINCT chrom FROM read_parquet(%s) ORDER BY chrom", sql_path(real_regulatory_path))
)$chrom

real_typed_contig_queries <- vapply(
  real_contigs,
  function(chrom) {
    sprintf(
      paste(
        "SELECT count(*)::BIGINT AS n",
        "FROM read_parquet(%s) q",
        "JOIN read_parquet(%s) s",
        "  ON q.start0 < s.end0 AND q.end0 > s.start0",
        "WHERE q.chrom = %s AND s.chrom = %s"
      ),
      sql_path(real_case_path),
      sql_path(real_regulatory_path),
      sql_string(chrom),
      sql_string(chrom)
    )
  },
  character(1L)
)
real_typed_interval_sql <- paste0(
  "SELECT sum(n)::BIGINT AS n FROM (",
  paste(real_typed_contig_queries, collapse = " UNION ALL "),
  ")"
)

real_typed_interval_plan <- DBI::dbGetQuery(
  con,
  paste("EXPLAIN", real_typed_contig_queries[[1L]])
)$explain_value
real_typed_iejoin_count <- sum(lengths(regmatches(
  real_typed_interval_plan,
  gregexpr("IE_JOIN", real_typed_interval_plan, fixed = TRUE)
)))
stopifnot(real_typed_iejoin_count == 1L)

data.frame(
  physical_operator = "IE_JOIN",
  representative_contig = real_contigs[[1L]],
  operators_in_representative_plan = real_typed_iejoin_count,
  contig_shards_in_full_query = length(real_contigs)
)

real_typed_interval <- measure_thread_grid(
  "real_regulatory_typed_per_contig_iejoin_pairs",
  real_case_rows,
  function() DBI::dbGetQuery(con, real_typed_interval_sql)
)

real_packed_interval <- measure_thread_grid(
  "real_regulatory_packed_coordinate_iejoin_pairs",
  real_case_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n",
          "FROM read_parquet(%s) q",
          "JOIN read_parquet(%s) s",
          "  ON q.rk_chrom_start < s.rk_chrom_end",
          " AND q.rk_chrom_end > s.rk_chrom_start"
        ),
        sql_path(real_case_path),
        sql_path(real_regulatory_path)
      )
    )
  }
)

real_max_span <- as.numeric(real_regulatory_stats$max_span[1L])
real_regionkey_interval <- measure_thread_grid(
  "real_regulatory_regionkey_bounded_start_pairs",
  real_case_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n",
          "FROM read_parquet(%s) q",
          "JOIN read_parquet(%s) s",
          "  ON s.rk_chrom_start >=",
          "     (regionkey(q.chrom, greatest(0, q.start0 - %d + 1)::BIGINT, greatest(0, q.start0 - %d + 1)::BIGINT) >> 31)",
          " AND s.rk_chrom_start < (regionkey(q.chrom, q.end0::BIGINT, q.end0::BIGINT) >> 31)",
          " AND are_overlapping_regionkeys(q.rk, s.rk)"
        ),
        sql_path(real_case_path),
        sql_path(real_regulatory_path),
        real_max_span,
        real_max_span
      )
    )
  }
)

real_cgranges_interval <- measure_thread_grid(
  "real_regulatory_cgranges_count_pairs",
  real_case_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT sum(duckhts_cgranges_count_overlaps('%s', chrom, start0::BIGINT, end0::BIGINT))::BIGINT AS n",
          "FROM read_parquet(%s)"
        ),
        real_idx_name,
        sql_path(real_case_path)
      )
    )
  }
)

real_interval_results <- rbind(
  real_typed_interval,
  real_packed_interval,
  real_regionkey_interval,
  real_cgranges_interval
)
stopifnot(length(unique(real_interval_results$result_rows)) == 1L)
real_interval_results

real_interval_speedups <- merge(
  real_interval_results[, c("operation", "threads", "input_rows_per_second")],
  real_typed_interval[, c("threads", "input_rows_per_second")],
  by = "threads",
  suffixes = c("", "_typed_baseline")
)
real_interval_speedups$speedup_vs_typed_iejoin <-
  real_interval_speedups$input_rows_per_second /
  real_interval_speedups$input_rows_per_second_typed_baseline
real_interval_speedups
```

The physical-plan assertion above matters. DuckDB's
[IEJoin description](https://duckdb.org/2022/05/27/iejoin) explains that a
double-inequality join sorts both relations on the first condition, constructs
merged and permutation arrays for the second condition, and uses a bitmap to
identify matching pairs. DuckDB parallelizes sorted blocks and can spool join
blocks to disk. The measured 665/712 MiB is therefore the resident cost of real
parallel `IE_JOIN` planning/execution over these Parquet relations, not a nested
loop mislabeled as an interval join. The per-contig `UNION ALL` is deliberate:
it makes contig equality a static shard filter and leaves exactly the two
coordinate inequalities for each `IE_JOIN` operator.

For this real source, cgranges is `r round(real_interval_speedups$speedup_vs_typed_iejoin[real_interval_speedups$operation == "real_regulatory_cgranges_count_pairs" & real_interval_speedups$threads == 1], 2)` times the typed DuckDB interval join on one thread and `r round(real_interval_speedups$speedup_vs_typed_iejoin[real_interval_speedups$operation == "real_regulatory_cgranges_count_pairs" & real_interval_speedups$threads == 4], 2)` times on four threads. The globally bounded RegionKey range is correct but loses because the single 48,907-base source interval widens candidate discovery for every GIAB allele.

### Gene annotations at gene cardinality

Gene annotations are not a multi-million-row allele problem. They should be
joined after affected transcript rows have been reduced to distinct genes.
This comparison uses the same full Ensembl model both ways to expose the
avoidable transcript multiplicity.

```{r real-gene-join}
gene_transcript_join <- measure_thread_grid(
  "gnomad_constraint_join_per_transcript",
  644292,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n",
          "FROM read_parquet(%s) t",
          "JOIN read_parquet(%s) g USING (gene_id)"
        ),
        sql_path(transcripts_parquet),
        sql_path(real_constraint_path)
      )
    )
  }
)

gene_distinct_join <- measure_thread_grid(
  "gnomad_constraint_join_after_gene_dedup",
  644292,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n",
          "FROM (SELECT DISTINCT gene_id FROM read_parquet(%s)) t",
          "JOIN read_parquet(%s) g USING (gene_id)"
        ),
        sql_path(transcripts_parquet),
        sql_path(real_constraint_path)
      )
    )
  }
)

rbind(gene_transcript_join, gene_distinct_join)
```

### Materialized provider output

Counting is useful for operator isolation, but the clinical pipeline must keep
the matching annotation rows. These four-thread runs write the exact ClinVar
and regulatory results to ZSTD Parquet and report both row and byte output.

```{r real-materialized-output}
set_duck_threads(4L)
real_exact_output_path <- file.path(real_cache_dir, "giab_clinvar_exact_matches.parquet")
real_regulatory_output_path <- file.path(real_cache_dir, "giab_regulatory_overlaps.parquet")

real_exact_output_query <- sprintf(
  paste0(
    "SELECT q.chrom, q.pos, q.ref, q.alt, ",
    "a.allele_id, a.clinical_significance, a.gene_info ",
    "FROM read_parquet(%s) q JOIN read_parquet(%s) a USING (vk) ",
    "UNION ALL ",
    "SELECT q.chrom, q.pos, q.ref, q.alt, ",
    "a.allele_id, a.clinical_significance, a.gene_info ",
    "FROM read_parquet(%s) q JOIN read_parquet(%s) a ",
    "ON q.vk = a.vk AND q.chrom = a.chrom AND q.pos = a.pos ",
    "AND q.ref = a.ref AND q.alt = a.alt"
  ),
  sql_path(real_case_reversible_path),
  sql_path(real_clinvar_reversible_path),
  sql_path(real_case_hashed_path),
  sql_path(real_clinvar_hashed_path)
)

real_regulatory_output_query <- sprintf(
  paste(
    "SELECT q.chrom, q.pos, q.ref, q.alt,",
    "       s.stable_id, s.feature_type, s.so_term, s.so_accession",
    "FROM read_parquet(%s) q",
    "JOIN read_parquet(%s) s",
    "  ON q.rk_chrom_start < s.rk_chrom_end",
    " AND q.rk_chrom_end > s.rk_chrom_start"
  ),
  sql_path(real_case_path),
  sql_path(real_regulatory_path)
)

materialized_results <- rbind(
  cbind(output = "GIAB exact ClinVar annotations", copy_timed(real_exact_output_query, real_exact_output_path)),
  cbind(output = "GIAB Ensembl regulatory overlaps", copy_timed(real_regulatory_output_query, real_regulatory_output_path))
)
materialized_results$rows_per_second <- materialized_results$rows / materialized_results$seconds
materialized_results$input_rows <- real_case_rows
materialized_results$input_rows_per_second <- real_case_rows / materialized_results$seconds
materialized_results$output_mb_per_second <- materialized_results$bytes / materialized_results$seconds / 1e6
stopifnot(materialized_results$rows[1L] == real_exact_variantkey_lanes$result_rows[1L])
stopifnot(materialized_results$rows[2L] == real_packed_interval$result_rows[1L])
materialized_results
```

## 6. Isolated peak memory and population-scale projection

The preceding timings run in one long-lived R/DuckDB process. Peak memory is
measured separately because cumulative allocator state would make an in-process
RSS sample meaningless. Each row below is a fresh DuckDB CLI process wrapped by
`/usr/bin/time -v`, with the same CPU pinning and an in-memory database. Reported
RSS includes the DuckDB process, decompression buffers, operator state, cgranges
construction transients, and the loaded extension; it excludes the operating
system page cache.

```{r isolated-memory}
rss_cpu_1 <- Sys.getenv("VARIANTKEY_RSS_CPU_1", unset = "0")
rss_cpu_4 <- Sys.getenv("VARIANTKEY_RSS_CPU_4", unset = "0,2,4,6")

exact_lanes_sql <- sprintf(
  paste0(
    "SELECT sum(n)::BIGINT AS n FROM (",
    "SELECT count(*)::BIGINT AS n FROM read_parquet(%s) q ",
    "JOIN read_parquet(%s) a USING (vk) UNION ALL ",
    "SELECT count(*)::BIGINT AS n FROM read_parquet(%s) q ",
    "JOIN read_parquet(%s) a ON q.vk = a.vk AND q.chrom = a.chrom ",
    "AND q.pos = a.pos AND q.ref = a.ref AND q.alt = a.alt)"
  ),
  sql_path(real_case_reversible_path),
  sql_path(real_clinvar_reversible_path),
  sql_path(real_case_hashed_path),
  sql_path(real_clinvar_hashed_path)
)
clinvarbitration_lanes_sql <- sprintf(
  paste0(
    "SELECT sum(n)::BIGINT AS n FROM (",
    "SELECT count(*)::BIGINT AS n FROM read_parquet(%s) q ",
    "JOIN read_parquet(%s) a USING (vk) UNION ALL ",
    "SELECT count(*)::BIGINT AS n FROM read_parquet(%s) q ",
    "JOIN read_parquet(%s) a ON q.vk = a.vk AND q.chrom = a.chrom ",
    "AND q.pos = a.pos AND q.ref = a.ref AND q.alt = a.alt)"
  ),
  sql_path(real_case_reversible_path),
  sql_path(real_clinvarbitration_reversible_path),
  sql_path(real_case_hashed_path),
  sql_path(real_clinvarbitration_hashed_path)
)

revel_join_sql <- sprintf(
  paste(
    "SELECT count(*) AS n, sum(a.revel) AS score_sum",
    "FROM read_parquet(%s) q JOIN read_parquet(%s) a USING (vk)"
  ),
  sql_path(real_revel_probe_path),
  sql_path(real_revel_path)
)
alpha_join_sql <- sprintf(
  paste(
    "SELECT count(*) AS n, sum(a.am_pathogenicity) AS score_sum",
    "FROM read_parquet(%s) q JOIN read_parquet(%s) a USING (vk)"
  ),
  sql_path(real_case_reversible_path),
  sql_path(real_alphamissense_path)
)

alpha_chr1_rows <- as.numeric(DBI::dbGetQuery(
  con,
  sprintf(
    "SELECT count(*) AS n FROM read_parquet(%s) WHERE extract_variantkey_chrom(vk) = 1",
    sql_path(real_alphamissense_path)
  )
)$n[1L])
alpha_chr1_join_sql <- sprintf(
  paste(
    "SELECT count(*) AS n, sum(a.am_pathogenicity) AS score_sum",
    "FROM (SELECT vk FROM read_parquet(%s)",
    "      WHERE extract_variantkey_chrom(vk) = 1) q",
    "JOIN (SELECT vk, am_pathogenicity FROM read_parquet(%s)",
    "      WHERE extract_variantkey_chrom(vk) = 1) a USING (vk)"
  ),
  sql_path(real_alphamissense_path),
  sql_path(real_alphamissense_path)
)

real_cgranges_sql <- paste0(
  "SELECT duckhts_cgranges_from_query('rss_reg', ",
  sql_string(real_index_query), ", 'chrom', 'start0', 'end0', 'stable_id'); ",
  "SELECT duckhts_cgranges_index('rss_reg'); ",
  sprintf(
    paste(
      "SELECT sum(duckhts_cgranges_count_overlaps('rss_reg', chrom, start0::BIGINT, end0::BIGINT))",
      "FROM read_parquet(%s)"
    ),
    sql_path(real_case_path)
  )
)

cgranges_synthetic_sql <- function(n, label_kind = "bigint") {
  label_sql <- switch(
    label_kind,
    none = "",
    bigint = ", i::BIGINT AS label",
    varchar = ", 'feature_' || i::VARCHAR AS label",
    stop("unknown cgranges label kind", call. = FALSE)
  )
  label_arg <- if (label_kind == "none") "" else ", 'label'"
  query <- sprintf(
    "SELECT '1' AS chrom, i * 10 AS start0, i * 10 + 5 AS end0%s FROM range(%d) t(i)",
    label_sql,
    as.numeric(n)
  )
  paste0(
    "SELECT duckhts_cgranges_from_query('rss_syn', ", sql_string(query),
    ", 'chrom', 'start0', 'end0'", label_arg, "); ",
    "SELECT duckhts_cgranges_index('rss_syn')"
  )
}

rss_alpha_stage_path <- file.path(real_cache_dir, "rss_alphamissense_stage.parquet")
rss_revel_stage_path <- file.path(real_cache_dir, "rss_revel_stage.parquet")
alpha_stage_sql <- sprintf(
  "COPY (%s) TO %s (FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 122880)",
  alphamissense_stage_query,
  sql_path(rss_alpha_stage_path)
)
revel_stage_sql <- sprintf(
  "COPY (%s) TO %s (FORMAT PARQUET, COMPRESSION ZSTD, ROW_GROUP_SIZE 122880)",
  revel_stage_query,
  sql_path(rss_revel_stage_path)
)

isolated_memory <- rbind(
  isolated_duckdb_measure("baseline", "SELECT 1", 1L, rss_cpu_1),
  isolated_duckdb_measure("baseline", "SELECT 1", 4L, rss_cpu_4),
  isolated_duckdb_measure("ClinVar split exact join", exact_lanes_sql, 1L, rss_cpu_1),
  isolated_duckdb_measure("ClinVar split exact join", exact_lanes_sql, 4L, rss_cpu_4),
  isolated_duckdb_measure(
    "ClinvArbitration split exact join",
    clinvarbitration_lanes_sql,
    1L,
    rss_cpu_1
  ),
  isolated_duckdb_measure(
    "ClinvArbitration split exact join",
    clinvarbitration_lanes_sql,
    4L,
    rss_cpu_4
  ),
  isolated_duckdb_measure("REVEL 77.97M exact join", revel_join_sql, 1L, rss_cpu_1),
  isolated_duckdb_measure("REVEL 77.97M exact join", revel_join_sql, 4L, rss_cpu_4),
  isolated_duckdb_measure("AlphaMissense 71.70M exact join", alpha_join_sql, 1L, rss_cpu_1),
  isolated_duckdb_measure("AlphaMissense 71.70M exact join", alpha_join_sql, 4L, rss_cpu_4),
  isolated_duckdb_measure(
    "GIAB three real exact providers one plan",
    real_multi_provider_sql,
    1L,
    rss_cpu_1
  ),
  isolated_duckdb_measure(
    "GIAB three real exact providers one plan",
    real_multi_provider_sql,
    4L,
    rss_cpu_4
  ),
  isolated_duckdb_measure(
    "GIAB twelve unique exact provider joins",
    repeated_provider_sql(12L),
    1L,
    rss_cpu_1
  ),
  isolated_duckdb_measure(
    "GIAB twelve unique exact provider joins",
    repeated_provider_sql(12L),
    4L,
    rss_cpu_4
  ),
  isolated_duckdb_measure("AlphaMissense chromosome 1 shard self-join", alpha_chr1_join_sql, 1L, rss_cpu_1),
  isolated_duckdb_measure("AlphaMissense chromosome 1 shard self-join", alpha_chr1_join_sql, 4L, rss_cpu_4),
  isolated_duckdb_measure("regulatory cgranges build and full GIAB probe", real_cgranges_sql, 1L, rss_cpu_1),
  isolated_duckdb_measure("regulatory DuckDB per-contig IEJoin", real_typed_interval_sql, 1L, rss_cpu_1),
  isolated_duckdb_measure("regulatory DuckDB per-contig IEJoin", real_typed_interval_sql, 4L, rss_cpu_4),
  isolated_duckdb_measure("cgranges 1M no labels", cgranges_synthetic_sql(1000000, "none"), 1L, rss_cpu_1),
  isolated_duckdb_measure("cgranges 1M BIGINT labels", cgranges_synthetic_sql(1000000, "bigint"), 1L, rss_cpu_1),
  isolated_duckdb_measure("cgranges 1M VARCHAR labels", cgranges_synthetic_sql(1000000, "varchar"), 1L, rss_cpu_1),
  isolated_duckdb_measure("cgranges 10M BIGINT labels", cgranges_synthetic_sql(10000000, "bigint"), 1L, rss_cpu_1),
  isolated_duckdb_measure(
    "AlphaMissense provider preparation",
    alpha_stage_sql,
    4L,
    rss_cpu_4,
    cleanup_paths = rss_alpha_stage_path
  ),
  isolated_duckdb_measure(
    "REVEL provider preparation",
    revel_stage_sql,
    4L,
    rss_cpu_4,
    cleanup_paths = rss_revel_stage_path
  )
)
unlink(c(rss_alpha_stage_path, rss_revel_stage_path))

baseline_rss <- isolated_memory[isolated_memory$operation == "baseline", c("threads", "peak_rss_mib")]
names(baseline_rss)[2L] <- "baseline_rss_mib"
isolated_memory <- merge(isolated_memory, baseline_rss, by = "threads", all.x = TRUE, sort = FALSE)
isolated_memory$incremental_peak_rss_mib <- pmax(
  0,
  isolated_memory$peak_rss_mib - isolated_memory$baseline_rss_mib
)
isolated_memory
```

The real 643,528-row regulatory cgranges index uses string labels because an
expanding annotation needs a stable ordinal or identifier. Its peak includes
index construction and the complete GIAB count probe. The synthetic rows expose
the current API's payload cost: besides cgranges' 16-byte core interval, DuckHTS
retains return coordinates, label validity, and currently one owned chromosome
string per interval. A count-only compact constructor that interns chromosome
names could reduce this, but it is an optimization—not permission to put dbSNP
or TOPMed into an interval index.

```{r cgranges-memory-density}
cgranges_memory_density <- data.frame(
  operation = c(
    "cgranges 1M no labels",
    "cgranges 1M BIGINT labels",
    "cgranges 1M VARCHAR labels",
    "cgranges 10M BIGINT labels"
  ),
  intervals = c(1e6, 1e6, 1e6, 1e7),
  stringsAsFactors = FALSE
)
cgranges_memory_density <- merge(
  cgranges_memory_density,
  isolated_memory[, c("operation", "peak_rss_mib", "incremental_peak_rss_mib")],
  by = "operation",
  sort = FALSE
)
cgranges_memory_density$incremental_peak_bytes_per_interval <-
  cgranges_memory_density$incremental_peak_rss_mib * 1024^2 / cgranges_memory_density$intervals
cgranges_memory_density
```

### Storage and logical time at catalog scale

The provider hot files below intentionally exclude source presentation columns.
They are serving projections, not replacements for the source receipts. Exact
providers may share one costed plan when the measured RSS is acceptable, as in
the real three-provider result above. At larger catalog counts the execution
macro may instead schedule bounded provider groups; twelve independent groups
then have approximately the sum of their elapsed times but the maximum of their
group RSS values. Loading twelve indexes concurrently is neither required nor
the default design.

```{r provider-scale}
provider_storage <- data.frame(
  provider = c(
    "ClinVar reversible exact lane",
    "ClinvArbitration exact lanes",
    "REVEL v1.3",
    "AlphaMissense v2"
  ),
  rows = c(
    lane_staging_results$rows[lane_staging_results$source == "ClinVar reversible numeric lane"],
    sum(lane_staging_results$rows[
      lane_staging_results$source %in% c(
        "ClinvArbitration reversible numeric lane",
        "ClinvArbitration hashed refinement lane"
      )
    ]),
    revel_provider_rows,
    alpha_provider_rows
  ),
  bytes = c(
    lane_staging_results$bytes[lane_staging_results$source == "ClinVar reversible numeric lane"],
    sum(lane_staging_results$bytes[
      lane_staging_results$source %in% c(
        "ClinvArbitration reversible numeric lane",
        "ClinvArbitration hashed refinement lane"
      )
    ]),
    staging_results$bytes[staging_results$source == "REVEL v1.3 to VariantKey Parquet"],
    staging_results$bytes[staging_results$source == "AlphaMissense v2 to VariantKey Parquet"]
  ),
  stringsAsFactors = FALSE
)
provider_storage$compressed_bytes_per_row <- provider_storage$bytes / provider_storage$rows
provider_storage

catalog_scenarios <- data.frame(
  catalog = c("TOPMed Freeze 8 BRAVO", "dbSNP Build 157 live RefSNP"),
  rows = c(705486649, 1172689405),
  stringsAsFactors = FALSE
)
catalog_scenarios$narrow_hot_gb_low <- catalog_scenarios$rows *
  min(provider_storage$compressed_bytes_per_row) / 1e9
catalog_scenarios$narrow_hot_gb_high <- catalog_scenarios$rows *
  max(provider_storage$compressed_bytes_per_row) / 1e9

dense_rates <- dense_score_results[dense_score_results$threads %in% c(1L, 4L), ]
logical_time <- do.call(
  rbind,
  lapply(seq_len(nrow(catalog_scenarios)), function(i) {
    do.call(
      rbind,
      lapply(c(1L, 4L), function(th) {
        rates <- dense_rates$input_rows_per_second[dense_rates$threads == th]
        data.frame(
          catalog = catalog_scenarios$catalog[i],
          threads = th,
          catalog_rows = catalog_scenarios$rows[i],
          twelve_dense_providers_minutes_fast = catalog_scenarios$rows[i] / max(rates) * 12 / 60,
          twelve_dense_providers_minutes_slow = catalog_scenarios$rows[i] / min(rates) * 12 / 60,
          stringsAsFactors = FALSE
        )
      })
    )
  })
)
catalog_scenarios
logical_time
```

These are deliberately labelled logical-time projections: they linearly apply
the measured REVEL-to-AlphaMissense throughput envelope to twelve equally dense
providers. They do not claim that all real sources are that large, equally
compressible, or equally expensive. For a 705-million-row or 1.17-billion-row
query, the execution macro must compile chromosome or coordinate-tile tasks so
one bounded group of provider shards and query keys supplies join state while
the large provider relations stream.
The measured largest AlphaMissense chromosome has `r format(alpha_chr1_rows, big.mark = ",", scientific = FALSE)` source rows; its isolated self-join is the concrete memory denominator for that scheduling model. Four chromosome workers may share immutable Parquet pages, but each owns its join state, so concurrency must be bounded by measured RSS rather than by the number of available source files.

## Notes

- The exact-join sections are the relevant comparison class for future
  `echtvar`-style annotation throughput work.
- The `mixed_stored_vk_join_plus_hash_refine` query represents the safe SQL
  pattern for hashed/nonreversible VariantKey rows when exact `REF/ALT`
  verification is required.
- VariantKey does not normalize alleles. Exact provider joins require the same
  normalized biallelic representation on both sides; the real ClinVar result
  counts exact stored representations, not merely equivalent biological edits.
- The reversible exact lane joins only `UBIGINT vk`; chromosome and allele
  strings are absent from that provider file. A general non-human store should
  use a release-scoped numeric contig ordinal and retain the ordinal-to-name
  map in the dataset manifest, Parquet key/value metadata, or a companion
  relation. Official VariantKey/RegionKey chromosome codes remain limited to
  canonical human contigs.
- Multiple equality joins are valid composition, but provider count alone is
  not the controlling cost. The provider receipt must declare key scope and
  uniqueness, the plan must keep the case side narrow, and peak join state must
  remain within the measured memory budget. Provider-specific rows should be
  reduced or nested before transcript expansion rather than multiplied across
  every transcript consequence.
- RegionKey is more than exact equality: its chromosome/start ordering supports
  binary search, ordered merge, and bounded start-range discovery. Arbitrary
  nested intervals still require a proven look-back bound, an augmented
  max-end index, or a complete overlap index such as cgranges.
- DuckDB's native interval join is the zero-registry SQL baseline. The
  RegionKey range query and cgranges are alternatives with different storage,
  preparation, and repeated-query trade-offs; the measured report, not the
  key format alone, decides which path is preferable.
- The Ensembl regulatory source has one 48,907-base interval, so the globally
  bounded RegionKey query admits a wide candidate window for every probe. The
  result is correct but slower than both cgranges and DuckDB's interval join;
  span-class partitioning or an augmented max-end index would be required to
  improve that RegionKey search without dropping long features.
- Sorted VariantKey row groups already provide useful min/max pruning. Physical
  chromosome/tile partitions become worthwhile when a distributed delta can
  be compiled into static touched-prefix scans; a plain hash join over sparse
  keys does not by itself guarantee file or row-group pruning.
- For richer overlap comparisons against external tools such as `bedtk` or
  `bedtools`, see the existing `benchmarks/benchmark_cgranges.Rmd` report.
- The real-source section is the 1.5.0 SQL-composition proof. It measures
  actual provider and case relations rather than claiming that the existence
  of scalar keys alone constitutes a supplementary-annotation system.
