---
title: "GenBank named-qualifier projections"
output:
  github_document:
    html_preview: false
    pandoc_args: ["--wrap=preserve", "--columns=400"]
---

This report measures `read_genbank(..., attributes := [...])` against the two
ways a query could obtain the same qualifiers before it existed: the
`attributes_map` MAP and the raw `attributes` string. The input is NCBI RefSeq
release 237's plasmid division (workload `genbank-plasmid`): its first
`plasmid.N.genomic.gbff.gz` part alone, its first two parts joined, and its
first four parts joined, each part pinned by NCBI's published MD5 and each
joined file by its own SHA-256 and record count. Joining consecutive release
parts grows the record count while every other dimension stays what the
release supplies; the `records` column in every table is the LOCUS count of
the input. A record whose only feature is `source` emits no row, so row counts
are below record counts. Stage all four parts without network access at render
time:

``` sh
Rscript -e 'duckhtsbench::duckhts_bench_stage_genbank_plasmid()'
```

Every workload is a DuckDB table macro whose parameters and output columns
carry a `p_` prefix, so no name can resolve to a reader column, and the input
path is bound at execution rather than pasted into SQL. Each workload returns
one row: a row count and a scalar metric over the stated columns. Before any
timing, the render asserts row-level parity: on every input, every named column
equals `attributes_map[key]` on every row, or the render stops. The
`named_three` and `map_three` workloads then compute the same metric over the
same three keys through the two representations, and the render stops if they
disagree, so the timed queries are known to return the same answer. Five timed
repetitions follow one verified warm-up run per workload and input; the
one-part run exceeds the five-second floor `STYLE.md` sets for a timing
verdict. Memory is measured the way `benchmark_genbank_memory.Rmd` does: one
fresh DuckDB CLI process per observation under `/usr/bin/time`, with the JSON
profiler's peak buffer-manager and temporary-directory bytes, for the three-key
MAP and named-column workloads at every scale, three observations each, with
`max_temp_directory_size` set to zero so any spill is an error. The budget is
the reader's contract, peak memory proportional to the largest record, not to
records consumed: the ceiling is three times the one-part fresh-process peak
RSS, applied to every observation. The one-thread condition is pinned to the
highest logical CPU in the rendering process's affinity mask, so wrap the
render in `taskset -c` to choose it; the table records the affinity.
`read_genbank` declares one worker, so a multi-thread condition would time the
same scan and is not recorded.

``` sh
taskset -c 8-11 Rscript -e 'rmarkdown::render("benchmarks/benchmark_genbank_named_attributes.Rmd")'
```

---
title: "GenBank named-qualifier projections"
output:
  github_document:
    html_preview: false
    pandoc_args: ["--wrap=preserve", "--columns=400"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r benchmark, echo=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
})
stopifnot(requireNamespace("duckhtsbench", quietly = TRUE))
stopifnot(requireNamespace("knitr", quietly = TRUE))

Sys.setenv(DUCKHTSBENCH_REGISTRY = normalizePath(
  "r/duckhtsbench/inst/benchmark_registry.tsv", mustWork = TRUE
))
staged <- duckhtsbench::duckhts_bench_stage_genbank_plasmid(fetch = FALSE)
plan <- duckhtsbench::duckhts_bench_stage_plan("genbank-plasmid")
artifact_ids <- c("genbank_plasmid_records_1", "genbank_plasmid_records_2", "genbank_plasmid_records_4")
registered_records <- vapply(artifact_ids, function(id) {
  fields <- duckhtsbench:::duckhts_bench_identity_fields(plan$supplier_identity[[match(id, plan$id)]])
  as.integer(fields[["records"]])
}, integer(1L))
inputs <- data.frame(
  parts = c(1L, 2L, 4L),
  artifact = artifact_ids,
  path = normalizePath(unname(unlist(staged[c("records_1", "records_2", "records_4")])), mustWork = TRUE),
  records = unname(registered_records),
  stringsAsFactors = FALSE
)
for (i in seq_len(nrow(inputs))) {
  duckhtsbench::duckhts_bench_validate_identity(inputs$artifact[[i]], inputs$path[[i]])
}
extension <- normalizePath(Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"),
                           mustWork = TRUE)
revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE)
# The rendered outputs are committed after rendering, which changes the commit
# hash but not the measured source. The src tree hash is what a reviewer
# compares against the PR head's `git rev-parse HEAD:src`.
src_tree <- system2("git", c("rev-parse", "HEAD:src"), stdout = TRUE)
if (system2("git", c("diff", "--quiet", "HEAD", "--", "src"),
            stdout = FALSE, stderr = FALSE) != 0L) {
  stop("Build source must match the recorded HEAD")
}
host_load <- sub(".*load average: ", "", system2("uptime", stdout = TRUE))

set_affinity <- function(cpus) {
  output <- system2("taskset", c("-pc", cpus, Sys.getpid()), stdout = TRUE, stderr = TRUE)
  if (!is.null(attr(output, "status"))) stop(paste(output, collapse = " "))
  invisible(output)
}
affinity_cpus <- function() {
  line <- system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE)
  spec <- sub("^.*: *", "", line[[length(line)]])
  unlist(lapply(strsplit(spec, ",", fixed = TRUE)[[1L]], function(part) {
    bounds <- as.integer(strsplit(part, "-", fixed = TRUE)[[1L]])
    if (length(bounds) == 1L) bounds else seq.int(bounds[[1L]], bounds[[2L]])
  }))
}
host_cpus <- affinity_cpus()
cpu_one <- as.character(host_cpus[[length(host_cpus)]])

quote_literal <- function(value) paste0("'", gsub("'", "''", value, fixed = TRUE), "'")
driver <- duckdb::duckdb(shared_home = FALSE, config = list(allow_unsigned_extensions = "true"))
con <- DBI::dbConnect(driver)
invisible(DBI::dbExecute(con, paste("LOAD", quote_literal(extension))))
# The engine the queries run on, read from the connection rather than from the
# R package index, so the render does not depend on where duckdb is installed.
duckdb_runtime <- DBI::dbGetQuery(con, "SELECT version() AS p_version")$p_version[[1L]]
invisible(DBI::dbExecute(con, "SET threads = 1"))

# One table macro per workload. Parameters and outputs are p_-prefixed so they
# cannot resolve to a reader column; the path is bound at execution.
keys <- "['gene', 'product', 'locus_tag']"
macros <- c(
  plain_columns = "SELECT count(*) AS p_n, sum(start + \"end\") AS p_metric FROM read_genbank(p_path)",
  named_none_projected = sprintf(
    "SELECT count(*) AS p_n, sum(start + \"end\") AS p_metric FROM read_genbank(p_path, attributes := %s)", keys),
  attributes_string = "SELECT count(*) AS p_n, sum(length(attributes)) AS p_metric FROM read_genbank(p_path)",
  map_gene = "SELECT count(*) AS p_n, count(attributes_map['gene']) AS p_metric FROM read_genbank(p_path, attributes_map := true)",
  map_three = paste(
    "SELECT count(*) AS p_n, sum(length(coalesce(attributes_map['gene'], ''))",
    "+ length(coalesce(attributes_map['product'], '')) + length(coalesce(attributes_map['locus_tag'], ''))) AS p_metric",
    "FROM read_genbank(p_path, attributes_map := true)"),
  named_three = sprintf(paste(
    "SELECT count(*) AS p_n, sum(length(coalesce(gene, '')) + length(coalesce(product, ''))",
    "+ length(coalesce(locus_tag, ''))) AS p_metric FROM read_genbank(p_path, attributes := %s)"), keys),
  named_one_projected = sprintf(
    "SELECT count(*) AS p_n, sum(length(coalesce(gene, ''))) AS p_metric FROM read_genbank(p_path, attributes := %s)", keys)
)
for (name in names(macros)) {
  invisible(DBI::dbExecute(con, sprintf("CREATE MACRO bench_%s(p_path) AS TABLE %s", name, macros[[name]])))
}
run_workload <- function(name, path) {
  DBI::dbGetQuery(con, sprintf("SELECT p_n, p_metric FROM bench_%s(?)", name), params = list(path))
}
same_answer <- function(observed, expected) {
  identical(as.character(observed$p_n[[1L]]), as.character(expected$p_n[[1L]])) &&
    identical(as.character(observed$p_metric[[1L]]), as.character(expected$p_metric[[1L]]))
}

# Row-level parity on every input: each named column must equal the MAP entry
# on every row. p_-prefixed outputs; the path is bound.
invisible(DBI::dbExecute(con, paste(
  "CREATE MACRO bench_parity(p_path) AS TABLE",
  "SELECT count(*) AS p_rows, count(*) FILTER (WHERE",
  "gene IS DISTINCT FROM attributes_map['gene'] OR product IS DISTINCT FROM attributes_map['product']",
  "OR locus_tag IS DISTINCT FROM attributes_map['locus_tag']) AS p_mismatches",
  sprintf("FROM read_genbank(p_path, attributes_map := true, attributes := %s)", keys)
)))
parity <- do.call(rbind, lapply(seq_len(nrow(inputs)), function(i) {
  observed <- DBI::dbGetQuery(con, "SELECT p_rows, p_mismatches FROM bench_parity(?)",
                              params = list(inputs$path[[i]]))
  stopifnot(as.numeric(observed$p_rows[[1L]]) > 0,
            identical(as.numeric(observed$p_mismatches[[1L]]), 0))
  data.frame(parts = inputs$parts[[i]], records = inputs$records[[i]], rows = observed$p_rows[[1L]],
             mismatches = observed$p_mismatches[[1L]], stringsAsFactors = FALSE)
}))
inputs$rows <- as.numeric(parity$rows)

set_affinity(cpu_one)
rows <- list()
measured_runs <- list()
repetitions <- 5L
for (i in seq_len(nrow(inputs))) {
  path <- inputs$path[[i]]
  expected_rows <- inputs$rows[[i]]
  for (name in names(macros)) {
    expected <- run_workload(name, path)
    stopifnot(identical(as.numeric(expected$p_n[[1L]]), expected_rows))
    seconds <- numeric(repetitions)
    for (run in seq_len(repetitions)) {
      started <- proc.time()[["elapsed"]]
      observed <- run_workload(name, path)
      seconds[[run]] <- proc.time()[["elapsed"]] - started
      stopifnot(same_answer(observed, expected))
    }
    measured_runs[[length(measured_runs) + 1L]] <- data.frame(
      parts = inputs$parts[[i]], records = inputs$records[[i]], workload = name, run = seq_len(repetitions), seconds = seconds,
      input_rows = expected_rows, output_rows = 1L, metric = as.character(expected$p_metric[[1L]]),
      stringsAsFactors = FALSE
    )
    rows[[length(rows) + 1L]] <- data.frame(
      parts = inputs$parts[[i]], records = inputs$records[[i]], workload = name, affinity = cpu_one, input_rows = expected_rows,
      output_rows = 1L, metric = as.character(expected$p_metric[[1L]]),
      min_seconds = min(seconds), median_seconds = median(seconds), max_seconds = max(seconds),
      repetitions = repetitions, stringsAsFactors = FALSE
    )
  }
}
results <- do.call(rbind, rows)
# The named columns and the MAP must agree on the same three-key metric at
# every scale, or the report is timing two different answers.
for (parts in inputs$parts) {
  pair <- results[results$parts == parts & results$workload %in% c("map_three", "named_three"), "metric"]
  stopifnot(length(unique(pair)) == 1L)
}
write.csv(do.call(rbind, measured_runs),
          "benchmarks/benchmark_genbank_named_attributes_runs.csv", row.names = FALSE)
DBI::dbDisconnect(con, shutdown = TRUE)

# Memory: one fresh DuckDB CLI process per observation, as benchmark_genbank_memory.Rmd
# measures it. Spill is forbidden; the profiler reports peak buffer and temp bytes.
duckdb_cli <- Sys.which("duckdb")
stopifnot(nzchar(duckdb_cli), file.exists("/usr/bin/time"), requireNamespace("jsonlite", quietly = TRUE))
memory_sql <- c(
  map_three = macros[["map_three"]],
  named_three = macros[["named_three"]]
)
profile_memory <- function(workload, path, expected_rows, expected_metric, repetition) {
  scratch <- tempfile(sprintf("genbank-named-memory-%s-%d-", workload, repetition))
  dir.create(scratch)
  on.exit(unlink(scratch, recursive = TRUE), add = TRUE)
  sql_path <- file.path(scratch, "query.sql")
  result_path <- file.path(scratch, "result.csv")
  profile_path <- file.path(scratch, "profile.json")
  time_path <- file.path(scratch, "time.tsv")
  # The path enters as a SQL variable set once and read through getvariable(),
  # so the macro body never sees it as text. COPY cannot wrap EXECUTE, which is
  # why this is not a prepared statement.
  writeLines(c(
    "SET threads = 1;",
    "SET max_temp_directory_size = '0KiB';",
    sprintf("LOAD %s;", quote_literal(extension)),
    sprintf("SET VARIABLE p_path = %s;", quote_literal(path)),
    sprintf("CREATE MACRO bench_%s(p_path) AS TABLE %s;", workload, memory_sql[[workload]]),
    "PRAGMA enable_profiling = 'json';",
    sprintf("PRAGMA profiling_output = %s;", quote_literal(profile_path)),
    sprintf("COPY (SELECT p_n, p_metric FROM bench_%s(getvariable('p_path'))) TO %s (HEADER);",
            workload, quote_literal(result_path)),
    "PRAGMA disable_profiling;"
  ), sql_path)
  status <- system2("/usr/bin/time",
                    c("-f", shQuote("%M\t%e"), "-o", shQuote(time_path),
                      shQuote(duckdb_cli), "-unsigned", "-batch"),
                    stdin = sql_path, stdout = FALSE, stderr = FALSE)
  stopifnot(status == 0L)
  observed <- utils::read.csv(result_path, check.names = FALSE, colClasses = "character")
  stopifnot(identical(as.numeric(observed$p_n[[1L]]), expected_rows),
            identical(observed$p_metric[[1L]], expected_metric))
  profile <- jsonlite::fromJSON(profile_path, simplifyVector = TRUE)
  timing <- utils::read.delim(time_path, header = FALSE, col.names = c("max_rss_kib", "elapsed_seconds"))
  data.frame(workload = workload, repetition = repetition,
             max_rss_kib = timing$max_rss_kib[[1L]],
             duckdb_peak_buffer_bytes = as.numeric(profile$system_peak_buffer_memory),
             duckdb_peak_temp_bytes = as.numeric(profile$system_peak_temp_dir_size),
             stringsAsFactors = FALSE)
}
memory_rows <- list()
for (i in seq_len(nrow(inputs))) {
  expected_rows <- inputs$rows[[i]]
  for (workload in names(memory_sql)) {
    expected_metric <- results$metric[results$parts == inputs$parts[[i]] & results$workload == workload]
    observations <- do.call(rbind, lapply(seq_len(3L), function(repetition) {
      profile_memory(workload, inputs$path[[i]], expected_rows, expected_metric, repetition)
    }))
    memory_rows[[length(memory_rows) + 1L]] <- data.frame(
      parts = inputs$parts[[i]], records = inputs$records[[i]], workload = workload, input_rows = expected_rows,
      median_max_rss_mib = round(median(observations$max_rss_kib) / 1024, 1),
      max_max_rss_mib = round(max(observations$max_rss_kib) / 1024, 1),
      duckdb_peak_buffer_mib = round(max(observations$duckdb_peak_buffer_bytes) / 2^20, 2),
      duckdb_peak_temp_bytes = max(observations$duckdb_peak_temp_bytes),
      observations = nrow(observations), stringsAsFactors = FALSE
    )
  }
}
memory <- do.call(rbind, memory_rows)
stopifnot(all(memory$duckdb_peak_temp_bytes == 0))
# Budget: the reader streams one record at a time, so peak memory must not grow
# with record count. Ceiling: three times the one-part named-column peak RSS.
rss_ceiling_mib <- 3 * memory$median_max_rss_mib[memory$parts == 1L & memory$workload == "named_three"]
stopifnot(all(memory$max_max_rss_mib <= rss_ceiling_mib))
write.csv(memory, "benchmarks/benchmark_genbank_named_attributes_memory.csv", row.names = FALSE)

# Scaling of the named-column path: two- and four-part medians over the one-part median,
# beside the record-count ratios, since release parts are not equal in records.
scaling <- do.call(rbind, lapply(c("named_three", "map_three", "plain_columns"), function(name) {
  medians <- results$median_seconds[results$workload == name]
  data.frame(workload = name, seconds_1_part = medians[[1L]],
             ratio_2_parts = medians[[2L]] / medians[[1L]], ratio_4_parts = medians[[3L]] / medians[[1L]],
             records_ratio_2_parts = inputs$records[[2L]] / inputs$records[[1L]],
             records_ratio_4_parts = inputs$records[[3L]] / inputs$records[[1L]],
             stringsAsFactors = FALSE)
}))
```

Source revision: `r revision`; measured `src` tree: `r src_tree` (compare with `git rev-parse HEAD:src` at the reviewed head, since committing this render moves the commit hash but not the tree). DuckDB runtime: `r duckdb_runtime`.
Extension SHA-256: `r strsplit(system2("sha256sum", extension, stdout = TRUE), " ", fixed = TRUE)[[1L]][[1L]]`.
Host load before the run: `r host_load`. One thread, affinity `r cpu_one`, `r repetitions` timed repetitions per cell.

```{r results, echo=FALSE}
knitr::kable(results, digits = 3)
```

Row-level parity of the three named columns against `attributes_map`, asserted before timing:

```{r parity, echo=FALSE}
knitr::kable(parity)
```

Growth with record count: the ratio of the two- and four-part medians to the one-part median, beside the ratio of their record counts (release parts are not equal in size, so linear growth means the time ratio tracks the record ratio):

```{r scaling, echo=FALSE}
knitr::kable(scaling, digits = 3)
```

Memory, one fresh DuckDB CLI process per observation, spill forbidden. The peak-RSS ceiling is `r rss_ceiling_mib` MiB (three times the one-part named-column median); the render stops if any observation exceeds it or spills:

```{r memory, echo=FALSE}
knitr::kable(memory)
```

`named_none_projected` requests three keys at bind but reads only the
coordinate columns; its scan resolves no qualifiers. `named_one_projected`
requests three keys and scans only `gene`. `map_three` and `named_three` are
the same metric through the two representations. Comparisons across workloads
reflect different aggregation and expression costs, not isolated lookup cycles.
The nearest recorded GenBank measurements are in
[`benchmark_genbank_reader.md`](benchmark_genbank_reader.md), which times the
plain-column and MAP scans of a single E. coli record without named columns
and is not directly comparable because its input and workloads differ.
