---
title: "SQL entry-point scaling audit: measured subset"
output: github_document
---

<!-- benchmark_sql_scaling_audit.md is rendered from this file. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
read_measurements <- function(path) {
  utils::read.delim(path, stringsAsFactors = FALSE)
}
somalier <- read_measurements("benchmarks/sql_scaling_audit_measurements.tsv")
norm <- read_measurements("benchmarks/sql_scaling_audit_norm.tsv")
fixed <- read_measurements("benchmarks/sql_scaling_audit_fixed_pairs.tsv")
measured <- rbind(somalier, norm)
stopifnot(all(table(interaction(measured$case, measured$threads)) == 3L),
          all(table(interaction(fixed$case, fixed$threads)) == 3L))
labels <- c(panel_hash = "duckhts_somalier_panel_sha256",
            frequency_hash = "duckhts_somalier_frequency_sha256",
            sketches = "duckhts_somalier_prepare_sketches",
            charr = "duckhts_somalier_charr",
            matched = "duckhts_somalier_matched_contamination",
            norm = "duckhts_bcftools_norm")
summary_rows <- lapply(split(measured, interaction(measured$case, measured$threads)),
                       function(rows) {
  rows <- rows[order(rows$size), ]
  data.frame(entry_point = labels[[rows$case[[1L]]]],
             threads = rows$threads[[1L]],
             input_1x_2x_4x = if (rows$case[[1L]] %in%
                                  c("panel_hash", "frequency_hash")) {
               "4,250 / 8,500 / 17,000 sites"
             } else if (rows$case[[1L]] == "norm") {
               "10k / 20k / 40k VCF records"
             } else {
               "8 / 16 / 32 samples (17k sites each)"
             },
             time_s = paste(sprintf("%.3f", rows$seconds), collapse = " / "),
             peak_rss_mib = paste(sprintf("%.0f", rows$peak_rss_kib / 1024),
                                  collapse = " / "),
             peak_buffer_mib = paste(sprintf("%.0f", rows$peak_buffer_mib),
                                     collapse = " / "),
             exponent = sprintf("%.2f", log(rows$seconds[[3L]] /
                                              rows$seconds[[1L]]) / log(4)),
             verdict = "No superlinear signal at these sizes")
})
summary_table <- do.call(rbind, summary_rows)
summary_table <- summary_table[order(match(summary_table$entry_point, labels),
                                     summary_table$threads), ]
row.names(summary_table) <- NULL
```

## Workload and measurement contract

Source revision: `ecf3d4a6a48dc81872ffb65e6a2a6e90346384e6` (no extension or SQL changes). DuckDB R 1.5.5; one or four DuckDB threads; one cold DuckDB connection per timed query. Linux `VmHWM` is process high-water resident memory in KiB, converted to MiB; DuckDB JSON profiling supplies `system_peak_buffer_memory`, converted to MiB. RSS includes process startup and input-table preparation; DuckDB buffer memory is the profiled query's peak. Wall time covers only the query execution, including output materialization to a temporary table; input staging and reading into the GIAB input table are excluded. Each cell is one run, not a confidence interval. This is an exploratory scaling check, not a performance regression claim.

The Somalier input is the `somalier-synthetic` registry workload: 17,000 autosomal sites and `samples × 17,000` count-evidence records, with one directional pair per additional sample. For the hashes, the dense panel/frequency prefix is 4,250, 8,500 or 17,000 sites; other Somalier calls use the whole panel. The public normalization input is the registry's GIAB NIST v4.2.1 HG002 GRCh38 phased VCF (`geno_giab_phased_source`) with the registered GRCh38 FASTA (`liftover_grch38_fasta`), both checksum-validated. Prefixes of 10,000, 20,000 and 40,000 physical VCF records are materialized before timing. Output denominators are one digest for each hash, `size` sketches or charr rows, `size - 1` matched pairs, and `size` normalized rows. The normalization probe uses the default site-preserving mode; split-multiallelic and reference-block cases are not covered.

Reproduce from the repository root after `make release`:

```sh
DUCKHTS_CACHE_DIR=/tmp/duckhts-sql-scaling-cache Rscript benchmarks/benchmark_sql_scaling_audit_run.R build/release/duckhts.duckdb_extension
DUCKHTS_CACHE_DIR=/tmp/duckhts-sql-scaling-cache Rscript benchmarks/benchmark_sql_scaling_audit_run.R build/release/duckhts.duckdb_extension --fixed-pairs
Rscript benchmarks/benchmark_sql_scaling_audit_run.R build/release/duckhts.duckdb_extension --norm-run
Rscript -e 'rmarkdown::render("benchmarks/benchmark_sql_scaling_audit.Rmd", knit_root_dir = getwd(), quiet = TRUE)'
```

The default cache must contain the registered public VCF and FASTA for the `--norm-run` command; see the `genotype-phase-set` and `liftover-reference-bundle` registry workloads for staging. Raw per-process observations are in `sql_scaling_audit_measurements.tsv`, `sql_scaling_audit_fixed_pairs.tsv`, and `sql_scaling_audit_norm.tsv`.

## Measured scaling

The exponent is `log(time at 4× / time at 1×) / log(4)`; startup and planner cost dominate the fastest queries. Time, RSS and buffer columns give 1× / 2× / 4× values. A verdict applies only to the displayed input range and workload.

```{r table, results='asis'}
knitr::kable(summary_table, row.names = FALSE)
```

## Candidate-set control

For `duckhts_somalier_matched_contamination`, a second staged run grows the evidence from 8 to 32 samples (136k / 272k / 544k input records) while holding the candidate set at three selected samples and two output pairs. Peak RSS (MiB) at one thread is `r paste(sprintf('%.0f', fixed$peak_rss_kib[fixed$threads == 1L] / 1024), collapse = ' / ')`; at four threads it is `r paste(sprintf('%.0f', fixed$peak_rss_kib[fixed$threads == 4L] / 1024), collapse = ' / ')`. DuckDB peak buffer memory (MiB) at one thread is `r paste(sprintf('%.0f', fixed$peak_buffer_mib[fixed$threads == 1L]), collapse = ' / ')`; at four threads it is `r paste(sprintf('%.0f', fixed$peak_buffer_mib[fixed$threads == 4L]), collapse = ' / ')`. All six runs produce two pairs. This probe does not show memory tracking unselected evidence. Matched contamination's full-candidate run is close to linear in candidate pairs at these sizes; its ordered `list()` aggregates hold per-pair site vectors, so larger candidate or panel sweeps remain necessary.

## Open audit work (ranked by likely exposure, not confirmed defect)

1. **Somalier VCF/BAM extraction, `duckhts_somalier_vcf_counts`, `duckhts_somalier_import_sites` and the R extraction wrappers:** no public multi-size input was exercised here. The VCF-count SQL has materialized validation and a per-record alternate/AD path; real indexed VCF and BAM workloads need 1×/2×/4× records, thread sweeps and operator profiles.
2. **`duckhts_somalier_verify_sketches`, R sketch wrappers, and matched contamination on a real multi-sample cohort:** synthetic count evidence cannot resolve transport/decompression or high-panel-cardinality costs. Hold selected pairs fixed and then grow candidates on a publicly staged cohort.
3. **`duckdb_munge`, `duckdb_munge_metal`, `duckdb_liftover` and their R builders:** their public input and reference workloads need independent staging, output-count controls, two thread settings and complete plan profiling. No scaling verdict is assigned.
4. **BAM/BCF/GFF/tabix `*_convert_parquet_sql` execution:** these macros produce COPY statements; scaling must include executing the generated statements, including partition/cardinality and disk output denominators. SQL-string generation alone would not test the writer.
5. **DuckVEP annotate, transcript projection and Ensembl macros:** lower priority, not measured; high-cardinality transcript cohorts and repeated exon/reference joins need a separate audit.

No query in the measured subset exceeded a 1.2 time-scaling exponent. Accordingly there is no flagged plan to diagnose, semantic rewrite, or before/after differential in this report. This is not an audit verdict on the unmeasured entry points. A source search of `src/` and `r/Rduckhts/R/` found no `row_number() OVER ()` without ordering; the `row_number()` call in `duckhts_parquet_metadata_args` orders by metadata priority. No order-dependent SQL was changed.
