---
title: "Selected FORMAT materialization in record-major genotype calls"
output:
  github_document:
    html_preview: false
---

This measures `read_geno` on every record of the registered GIAB HG002 GRCh38
NIST v4.2.1 benchmark VCF.gz. The single sample has declared GT, PS, AD, DP and
GQ fields. GT/PS, +AD and +AD/DP/GQ have separate measurements; each selection
also runs with `calls` unprojected. No record, allele or sample is filtered.

Stage the input explicitly with
`duckhtsbench::duckhts_bench_fetch("variantkey_giab_hg002_v421")`.
Rendering is network-free and resolves the input through the artifact registry.
The output is a single-sample workload, not a large-cohort or remote-I/O result.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
knitr::opts_knit$set(root.dir=normalizePath(".."))
```

```{r measurement, include=FALSE}
library(DBI)
source("r/duckhtsbench/R/registry.R")
source("r/duckhtsbench/R/stage.R")
source("benchmarks/genotype_format_run.R")
Sys.setenv(DUCKHTSBENCH_REGISTRY=normalizePath("r/duckhtsbench/inst/benchmark_registry.tsv"))
artifact <- "variantkey_giab_hg002_v421"
input <- duckhts_bench_artifact_path(artifact)
duckhts_bench_validate_identity(artifact, input)
extension <- normalizePath(Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"), mustWork=TRUE)
stopifnot(system2("git", c("diff", "--quiet", "HEAD", "--", "src")) == 0L)
revision <- system2("git", c("rev-parse", "HEAD"), stdout=TRUE)
source_tree <- system2("git", c("rev-parse", "HEAD:src"), stdout=TRUE)
hts <- function(args) {
  text <- system2("bcftools", shQuote(args), stdout=TRUE)
  stopifnot(is.null(attr(text, "status")))
  text
}
header <- hts(c("view", "-h", input))
stopifnot(identical(hts(c("query", "-l", input)), "HG002"),
          "##fileformat=VCFv4.2" %in% header,
          all(vapply(c("GT","PS","AD","DP","GQ"), function(tag)
            any(startsWith(header, paste0("##FORMAT=<ID=",tag,","))), logical(1))))
oracle <- genotype_format_oracle(input)
cases <- expand.grid(selection=c("GT_PS","GT_PS_AD","GT_PS_AD_DP_GQ"),
                      calls_projected=c(TRUE,FALSE), stringsAsFactors=FALSE)
selections <- list(GT_PS=character(), GT_PS_AD="AD", GT_PS_AD_DP_GQ=c("AD","DP","GQ"))
directory <- tempfile("genotype-format-report-")
dir.create(directory)
results <- list()
snapshots <- setNames(character(nrow(cases)), paste(cases$selection, cases$calls_projected))
repetitions <- as.integer(Sys.getenv("GENO_FORMAT_RUNS", "3"))
stopifnot(repetitions >= 1L)
for (repetition in seq_len(repetitions)) {
  order <- if (repetition %% 2L) seq_len(nrow(cases)) else rev(seq_len(nrow(cases)))
  for (i in order) {
    case <- cases[i,]
    fields <- selections[[case$selection]]
    snapshot <- if (repetition == 1L) file.path(directory, paste0(i,".parquet")) else ""
    if (nzchar(snapshot)) snapshots[i] <- snapshot
    observed <- callr::r(genotype_format_run, args=list(extension=extension, input=input,
      fields=fields, calls_projected=case$calls_projected, snapshot=snapshot))
    expected <- oracle
    if (!case$calls_projected) expected[names(expected)!="records"] <- 0
    if (!"AD" %in% fields) expected[c("ad_slots","ad_values")] <- 0
    if (!"DP" %in% fields) expected["dp_values"] <- 0
    if (!"GQ" %in% fields) expected["gq_values"] <- 0
    stopifnot(identical(as.numeric(observed[1,names(expected)]), as.numeric(expected)))
    results[[length(results)+1L]] <- cbind(repetition,case,observed)
  }
}
results <- do.call(rbind,results)
for (rows in split(results, interaction(results$selection,results$calls_projected)))
  stopifnot(length(unique(rows$checksum)) == 1L)

# Complete duplicate-preserving comparisons are outside every measurement.
con <- dbConnect(duckdb::duckdb(config=list(allow_unsigned_extensions="true")))
dbExecute(con,paste("LOAD",dbQuoteString(con,extension)))
dbExecute(con,"SET threads=1")
dbExecute(con,"SET memory_limit='8GB'")
table <- function(path) paste0("read_parquet(",dbQuoteString(con,path),")")
site <- function(path) paste("SELECT record_index,CHROM,POS,ID,REF,ALT FROM",table(path))
base <- function(path) paste("SELECT record_index,CHROM,POS,ID,REF,ALT,",
  "c.sample_index,c.alleles,c.phase_before,c.phase_set FROM",table(path),"g,unnest(calls) u(c)")
comparisons <- list()
for (i in seq_along(snapshots)) {
  differences <- genotype_format_difference(con,site(snapshots[1]),site(snapshots[i]))
  stopifnot(differences == 0)
  comparisons[[length(comparisons)+1L]] <- data.frame(comparison=paste("sites",names(snapshots)[i]),differences)
  if (cases$calls_projected[i]) {
    differences <- genotype_format_difference(con,base(snapshots[1]),base(snapshots[i]))
    stopifnot(differences == 0)
    comparisons[[length(comparisons)+1L]] <- data.frame(comparison=paste("GT/PS",names(snapshots)[i]),differences)
  }
}
dbExecute(con,sprintf(paste("CREATE TEMP TABLE canonical AS SELECT CHROM,POS,ID,REF,ALT,",
  "FORMAT_AD AS AD,FORMAT_DP AS DP,FORMAT_GQ AS GQ FROM read_bcf(%s,",
  "tidy_format := true,scan_mode := 'sequential',decode_error_policy := 'error')"),dbQuoteString(con,input)))
for (i in which(cases$calls_projected & cases$selection != "GT_PS")) {
  fields <- selections[[cases$selection[i]]]
  left <- paste("SELECT CHROM,POS,ID,REF,ALT,",paste0("c.format.",fields," AS ",fields,collapse=","),
                "FROM",table(snapshots[i]),"g,unnest(calls) u(c)")
  right <- paste("SELECT CHROM,POS,ID,REF,ALT,",paste(fields,collapse=","),"FROM canonical")
  differences <- genotype_format_difference(con,left,right)
  stopifnot(differences == 0)
  comparisons[[length(comparisons)+1L]] <- data.frame(comparison=paste("read_bcf",cases$selection[i]),differences)
}
# The same comparator must reject dropped records, altered values and missing-slot compaction.
controls <- c(
  genotype_format_difference(con,"SELECT 1 AS id, [10,NULL,5] AS AD","SELECT 1 AS id, [10,5] AS AD"),
  genotype_format_difference(con,"SELECT 1 AS id UNION ALL SELECT 1","SELECT 1 AS id"),
  genotype_format_difference(con,"SELECT 15 AS DP","SELECT 14 AS DP"),
  genotype_format_difference(con,"SELECT NULL::INTEGER AS GQ","SELECT 0 AS GQ"))
stopifnot(all(controls > 0))
comparisons <- do.call(rbind,comparisons)
dbDisconnect(con,shutdown=TRUE)
unlink(directory,recursive=TRUE)
```

## Source and denominators

```{r identity}
cat("Source revision:",revision,"\nSource tree:",source_tree,"\n")
cat("Extension SHA-256:",system2("sha256sum",shQuote(extension),stdout=TRUE),"\n")
cat("Input artifact:",artifact,"; bytes:",file.info(input)$size,"; observed MD5:",unname(tools::md5sum(input)),"\n")
cat("bcftools:",hts("--version")[1],"\nR:",R.version.string,"; DuckDB:",as.character(packageVersion("duckdb")),"\n")
cat(system2("uname","-srmo",stdout=TRUE),"\n")
cat(system2("taskset",c("-pc",Sys.getpid()),stdout=TRUE),"\n")
cat(system2("sh",c("-c",shQuote("lscpu | sed -n '/Model name:/p'")),stdout=TRUE),"\n")
cat("DuckDB threads: 1; scan handles: 1; HTSlib decompression workers: 0; repetitions:",repetitions,"\n")
knitr::kable(data.frame(denominator=names(oracle),count=as.numeric(oracle)))
```

Input denominators come from complete `bcftools view -H` records in bounded R
batches. Every physical duplicate counts. AD slots include missing list elements;
AD values count only nonmissing elements. DP/GQ values and PS values count nonmissing
scalars. An absent AD list has no child slots. An emitted scalar field occupies one
slot per call even when NULL. No PS throughput claim is made when its observed count
is zero.

## Materialization cost

Each timer includes bind, sequential read/decompression, decoding, vector writing
and `CREATE TABLE AS` in a fresh R/DuckDB process with an 8 GiB memory limit.
Calls-unprojected cases still materialize all site columns and scan-local ordinals;
they are not index-only counts. Downstream counts, checksums, checkpoints, snapshots
and comparisons are outside timing. Peak RSS is the process high-water mark immediately
after CTAS, not first-party workspace size. Database bytes are measured after checkpoint.
Case order reverses on alternate repetitions; input data is warm in the OS cache.

```{r timings}
summary <- aggregate(cbind(elapsed,cpu,peak_rss_kib,database_bytes) ~ selection+calls_projected,results,median)
counts <- unique(results[c("selection","calls_projected","records","calls","gt_slots","extra_slots",
                           "ad_values","dp_values","gq_values")])
summary <- merge(summary,counts,by=c("selection","calls_projected"),sort=FALSE)
knitr::kable(summary,digits=3)
knitr::kable(results[c("repetition","selection","calls_projected","elapsed","cpu","peak_rss_kib")],digits=3)
```

## Complete comparisons and scope

All six layouts preserve site fields and scan-local record ordinals. Projected
layouts preserve the complete default GT/PS call state. Selected AD/DP/GQ values
are compared with projected `read_bcf` across every record using `EXCEPT ALL`
in both directions. Per-pass complete-row fingerprints must also agree within
each layout. Four corruption controls test that missing-slot compaction, dropped
duplicates, changed depth and NULL-to-zero substitution are rejected.

```{r comparisons}
knitr::kable(comparisons)
cat("Rejected corruption controls:",length(controls),"\n")
```

This is the selected-FORMAT baseline, not a pre-change/post-change speedup claim.
The nearest prior [GIAB full-reader report](benchmark_bcf_shared_scan.md) reads
all declared columns with `read_bcf`; it does not measure this nested schema or
projection set. The [matched HPRC report](benchmark_genotypes_format_shared.md)
separately exercises the existing GT-only interfaces against
[their recorded baseline](benchmark_genotypes.md). These workloads do not establish
universal fastest-reader performance, multi-sample FORMAT scaling, remote-I/O
throughput or sealed allocation.

## Matched GT-only cohort comparison

The following comparison requires identical registered input hashes and all four
input-denominator rows in the rendered HPRC reports. Both use one scan thread,
no HTSlib decompression workers and three fresh-process passes on CPU 2. Values
are medians from those reports; the percentage is descriptive, not a statistical
significance test or a general no-regression claim. The reports retain complete
within-revision typed comparisons, not cross-revision output snapshots.

```{r cohort-comparison}
cohort_baseline <- "benchmarks/benchmark_genotypes.md"
cohort_current <- "benchmarks/benchmark_genotypes_format_shared.md"
for (path in c(cohort_baseline,cohort_current)) {
  cat(basename(path),":",grep("Source revision:",readLines(path),value=TRUE),"\n")
}
knitr::kable(genotype_hprc_comparison(cohort_baseline,cohort_current),digits=3)
```
