---
title: "Ancestry projection: GRCh38 reference product and 30x CRAM acceptance"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(DBI)
library(Rduckhts)
library(duckhtsbench)
repo <- normalizePath("..")
source_revision <- system2("git", c("-C", shQuote(repo), "rev-parse", "HEAD"), stdout = TRUE)
Sys.setenv(DUCKHTSBENCH_REGISTRY = file.path(repo, "r/duckhtsbench/inst/benchmark_registry.tsv"))

# Declared before any result was computed. The acceptance criterion is a
# per-sample maximum absolute difference between the 21 group proportions of the
# CRAM result (GRCh38 product) and of the phased-genotype result (GRCh37
# product) at the same panel loci, for both CRAM frequency methods. A sample
# also fails when either side fails the correlation gate or when the two
# largest-proportion groups differ.
tolerance <- 0.05
min_cor <- 0.4
min_depth <- 7L

peak_rss_mib <- function() {
  status <- grep("^VmHWM:", readLines("/proc/self/status"), value = TRUE)
  as.numeric(strsplit(trimws(status), "[[:space:]]+")[[1L]][[2L]]) / 1024
}
timed <- function(expr) {
  start <- proc.time()[["elapsed"]]
  value <- force(expr)
  list(value = value, seconds = proc.time()[["elapsed"]] - start)
}
```

```{r stage}
staged_product <- timed(duckhts_bench_stage_ancestry_grch38_parquet())
product <- staged_product$value
product_receipt <- utils::read.delim(paste0(product, ".sources.tsv"), colClasses = "character")
receipt_value <- function(field) product_receipt$value[product_receipt$field == field]
source_reference <- duckhts_bench_stage_ancestry_parquet()
staged_acceptance <- timed(duckhts_bench_stage_ancestry_grch38_acceptance())
genotypes_path <- staged_acceptance$value
acceptance_receipt <- utils::read.delim(paste0(genotypes_path, ".sources.tsv"),
                                        colClasses = "character")
acceptance_value <- function(field) acceptance_receipt$value[acceptance_receipt$field == field]
samples <- strsplit(acceptance_value("samples"), ",", fixed = TRUE)[[1L]]
stopifnot(identical(receipt_value("output_sha256"),
                    acceptance_value("ancestry_reference_grch38_output_sha256")))
correction_file <- duckhts_bench_stage_repository_fixtures(repo, "ancestry-projection")[["ancestry_correction"]]
destination_fasta <- duckhts_bench_fetch("liftover_grch38_fasta")
sample_panel <- utils::read.table(duckhts_bench_fetch("ancestry_1000g_panel"), header = FALSE,
  col.names = c("sample_id", "population", "superpopulation", "sex"))
```

The GRCh38 ancestry reference is derived from the keyed GRCh37 product by
`duckdb_liftover` and checked here against the GRCh37 phased genotypes of the
same 1000 Genomes individuals. Revision: `r source_revision`.

## GRCh38 reference product

`duckhts_bench_stage_ancestry_grch38_parquet()` lifts each of the 5,816,590
loci of the keyed GRCh37 Parquet with the registered Ensembl GRCh37-to-GRCh38
chain, the registered `human_g1k_v37` source FASTA and the registered Broad
`Homo_sapiens_assembly38` destination FASTA. The coded allele of a locus is
`allele_b`, the allele that the 21 group frequencies and 16 PC loadings
describe. Liftover reports its destination spelling through
`reverse_complemented`; the derived key is `allele_a` = the other allele and
`allele_b` = the coded allele, and every frequency and loading is copied
unchanged. This is the reversal convention of `rduckhts_ancestry_proportions()`:
a reversed input meets an unchanged reference row. Rewriting a swapped locus as
`1 - f` with negated loadings is not equivalent once the per-PC correction
coefficients apply (the constant shift of `-sum(U)` in each PC is multiplied by
the coefficient on one side only), so it is not done. The output is sorted by
`(chromosome, position, allele_a, allele_b)` with the same columns, types and
row groups as the GRCh37 product, so the panel builder and the proportion
wrapper consume it unchanged. A liftover map Parquet retains the source locus of
every output locus.

```{r denominators}
denominator_fields <- c("input_loci", "mapped", "rejected",
  grep("^rejected_", product_receipt$field, value = TRUE), "reverse_complemented",
  "swapped", "duplicate_destination_dropped", "output_loci")
denominators <- data.frame(
  quantity = sub("^rejected_", "  rejected: ", denominator_fields),
  loci = as.integer(vapply(denominator_fields, receipt_value, "")))
knitr::kable(denominators, format.args = list(big.mark = ","))
```

`mapped` counts loci for which liftover reported a destination, before the
later classification. Every input locus is in exactly one of rejected,
duplicate-destination dropped or output
(`r paste(receipt_value("rejected"), receipt_value("duplicate_destination_dropped"), receipt_value("output_loci"), sep = " + ")`
= `r receipt_value("input_loci")`). `swapped` and `reverse_complemented` count
output loci. Non-single-base source loci are rejected because ancestry panels
select biallelic SNVs and an indel has no coded allele that survives
left-alignment. The receipt binds these counts to the source Parquet, chain,
both FASTAs, the output and the DuckDB `r receipt_value("duckdb_version")`,
htslib `r receipt_value("duckhts_htslib_version")` and Rduckhts
`r receipt_value("rduckhts_version")` versions:

```{r receipt}
knitr::kable(product_receipt[product_receipt$field %in% c("source_parquet_sha256",
  "chain_sha256", "source_fasta_sha256", "destination_fasta_sha256", "output_sha256",
  "liftover_map_sha256"), ], row.names = FALSE)
```

## Acceptance comparison

```{r inputs}
con <- rduckhts_connect()
dbExecute(con, "SET threads=4")
quote_str <- function(x) as.character(dbQuoteString(con, x))
dbExecute(con, paste("CREATE VIEW ref37 AS SELECT * FROM read_parquet(",
                     quote_str(source_reference), ")"))
dbExecute(con, paste("CREATE VIEW ref38 AS SELECT * FROM read_parquet(",
                     quote_str(product), ")"))
dbExecute(con, sprintf(
  "CREATE TABLE ancestry_correction AS SELECT pc::INTEGER AS pc, coefficient::DOUBLE AS coefficient FROM read_csv_auto(%s, delim='\\t', header=true)",
  quote_str(correction_file)))
dbExecute(con, sprintf("CREATE TABLE phase3 AS SELECT * FROM read_parquet(%s)", quote_str(genotypes_path)))
dbExecute(con, paste(
  "CREATE TABLE ancestry_panel AS SELECT 'GRCh38' AS assembly,",
  "(row_number() OVER (ORDER BY site_index) - 1)::UBIGINT AS site_index,",
  "chromosome AS region, position::UBIGINT AS position,",
  "least(allele_a, allele_b) AS allele_a, greatest(allele_a, allele_b) AS allele_b",
  "FROM (SELECT DISTINCT site_index, chromosome, position, allele_a, allele_b",
  "FROM phase3) ORDER BY site_index"))
panel_sha256 <- dbGetQuery(con,
  "SELECT duckhts_somalier_panel_sha256('ancestry_panel') AS s")$s[[1L]]
stopifnot(identical(panel_sha256, acceptance_value("panel_sha256")))
panel_sites <- dbGetQuery(con, "SELECT count(*) AS n FROM ancestry_panel")$n
```

The panel is the 1,000-site genome-wide output of `rduckhts_ancestry_panel()`
on the GRCh38 product (5,000 bp windows, GRCh38 assembly); its SHA-256 is
`r panel_sha256`, verified equal to the staging receipt. Each panel site's
GRCh37 source locus comes from the liftover map. The GRCh37 side reads the
public phase-3 VCFs (`ALL.chrN.phase3_shapeit2_mvncall_integrated_v5b`) by
indexed `read_bcf(region := ...)` at only those loci, for
`r paste(samples, collapse = ", ")`; keeping REF/ALT that match the source
alleles in either orientation left `r acceptance_value("genotype_rows")`
sample-locus rows of `r panel_sites * length(samples)` possible. The GRCh38 side
is the registered 30x CRAM of each individual, cut by
`samtools view -M -L` to the reads that overlap a panel site and indexed
(`<sample>.panel-sites.cram`), because the whole-genome CRAMs are 14 to 16 GB.
Both stagings have a receipt with remote byte counts, ETags and output SHA-256
values.

```{r compare}
dose <- paste(
  "CASE WHEN gt ~ '^[01][|/][01]$' THEN CASE WHEN ref = source_allele_a THEN",
  "length(gt) - length(replace(gt, '1', '')) ELSE 2 - (length(gt) - length(replace(gt, '1', ''))) END END")
dbExecute(con, sprintf(paste(
  "CREATE TABLE genotype37 AS SELECT sample_id, source_chromosome AS chromosome,",
  "source_position AS position, source_allele_a AS allele_a, source_allele_b AS allele_b,",
  "(%s) / 2.0 AS frequency FROM phase3"), dose))
groups <- setdiff(dbGetQuery(con, "DESCRIBE SELECT * FROM ref38")$column_name,
                  c("chromosome", "position", "allele_a", "allele_b", paste0("PC", 1:16)))
results <- list()
proportions <- list()
for (sample in samples) {
  dbExecute(con, "DROP VIEW IF EXISTS genotype_input")
  dbExecute(con, sprintf(paste(
    "CREATE VIEW genotype_input AS SELECT * FROM genotype37 WHERE sample_id = %s",
    "AND frequency IS NOT NULL"), quote_str(sample)))
  called <- dbGetQuery(con, sprintf(paste(
    "SELECT count(*) AS n FROM phase3 WHERE sample_id = %s"), quote_str(sample)))$n
  genotype <- timed(rduckhts_ancestry_proportions(con, "genotype_input", "ref37",
    "ancestry_correction", min_cor = min_cor))
  genotype_result <- genotype$value
  cram <- file.path(dirname(genotypes_path), paste0(sample, ".panel-sites.cram"))
  for (method in c("allele_fraction", "called_genotype")) {
    counted <- timed(rduckhts_ancestry_bam(con, cram, sample, destination_fasta,
      "ancestry_panel", "ref38", "ancestry_correction", frequency_method = method,
      min_depth = min_depth, min_cor = min_cor, worker_count = 2L))
    result <- counted$value
    both_ok <- identical(unique(genotype_result$status), "ok") &&
      identical(unique(result$status), "ok")
    matched <- match(genotype_result$group_id, result$group_id)
    difference <- if (both_ok) max(abs(result$proportion[matched] -
                                       genotype_result$proportion)) else NA_real_
    top <- function(x) sort(x$group_id[order(-x$proportion)][1:2])
    top_agree <- both_ok && identical(top(result), top(genotype_result))
    results[[length(results) + 1L]] <- data.frame(
      sample_id = sample,
      superpopulation = sample_panel$superpopulation[match(sample, sample_panel$sample_id)],
      method = method, panel_sites = panel_sites, phase3_called = called,
      genotype_matched = genotype_result$used_variants[[1L]],
      cram_matched = result$used_variants[[1L]],
      genotype_cor = genotype_result$cor_pred[[1L]], cram_cor = result$cor_pred[[1L]],
      genotype_gate = unique(genotype_result$status), cram_gate = unique(result$status),
      max_abs_difference = difference, top2_agree = top_agree,
      within_tolerance = both_ok && difference <= tolerance && top_agree,
      cram_seconds = counted$seconds, genotype_seconds = genotype$seconds,
      process_peak_rss_mib = peak_rss_mib())
    proportions[[length(proportions) + 1L]] <- data.frame(
      sample_id = sample, method = method, group_id = genotype_result$group_id,
      genotype_grch37 = genotype_result$proportion,
      cram_grch38 = result$proportion[matched])
  }
}
results <- do.call(rbind, results)
proportions <- do.call(rbind, proportions)
```

Declared tolerance (fixed in this report's source before any result): the
maximum absolute difference over the 21 groups is at most `r tolerance` for each
CRAM frequency method (`allele_fraction`, `called_genotype`), both sides pass the
correlation gate (`cor_pred` at least `r min_cor`), and the two largest groups agree. `min_depth` is
`r min_depth`. Per sample, matched sites are the panel loci that survive
allele matching; gate failures are retained as `status` and would make the
difference undefined.

```{r acceptance}
shown <- results[, c("sample_id", "superpopulation", "method", "panel_sites",
  "genotype_matched", "cram_matched", "genotype_cor", "cram_cor", "genotype_gate",
  "cram_gate", "max_abs_difference", "top2_agree", "within_tolerance")]
knitr::kable(shown, digits = 4)
```

Gate failures: `r sum(results$genotype_gate != "ok")` genotype and
`r sum(results$cram_gate != "ok")` CRAM of `r nrow(results)` comparisons.
Samples within tolerance: `r sum(results$within_tolerance)` of `r nrow(results)`.
Largest difference: `r round(max(results$max_abs_difference, na.rm = TRUE), 4)`
against the `r tolerance` tolerance.

Proportions, group by group (GRCh37 phased genotypes against the GRCh37 product
versus the GRCh38 CRAM against the GRCh38 product):

```{r proportions}
wide <- reshape(proportions, idvar = c("group_id", "sample_id"), timevar = "method",
                direction = "wide")
wide <- wide[order(wide$sample_id, -wide$genotype_grch37.allele_fraction), ]
names(wide) <- sub("cram_grch38\\.", "cram_", sub("genotype_grch37\\.[a-z_]+", "genotype", names(wide)))
wide <- wide[, c("sample_id", "group_id", "genotype", "cram_allele_fraction", "cram_called_genotype")]
knitr::kable(wide, digits = 4, row.names = FALSE)
```

## Timing and memory

```{r timing}
timing <- data.frame(
  step = c("GRCh38 product staging call (cache hit: hashes of the sources)",
           "acceptance staging call (cache hit)"),
  seconds = c(staged_product$seconds, staged_acceptance$seconds))
knitr::kable(timing, digits = 1)
knitr::kable(results[, c("sample_id", "method", "genotype_seconds", "cram_seconds",
                         "process_peak_rss_mib")], digits = 2)
```

Scale notes. Dimensions: 1,000 panel sites, 3 samples, 21 groups, 16 PCs,
5,012,826 reference loci. Each call reads the panel-site CRAM once, then joins at
most 1,000 counted sites to the 5.0 million-locus Parquet reference, so the
matched relation and the aggregates hold at most 1,000 x 37 numeric values plus
21 x 16 solver terms per sample. Every measured call is under the 5 s floor of
`STYLE.md` on one input size with one fresh process, so this report gives
timing and peak RSS (cumulative `VmHWM` of the render process, including package
loading and the preceding calls) without a scaling or speed claim. It is a
correctness benchmark: it does not vary sites, samples or threads, and it does not
replace the 1x/2x/4x measurements of `benchmark_ancestry.md`. The one-time
staging (product derivation, remote fetches) is measured outside these tables.

Restrictions. The panel has 1,000 genome-wide sites, not the 17,000 of the
default panel, because cutting each 30x CRAM to the panel sites by indexed remote
access costs one region fetch per site. Sites without a phase-3 record whose
alleles match the source alleles, or with a missing genotype or depth below
`r min_depth`, drop out of the corresponding side; the matched counts above are
the audit of that loss. Phased 1000 Genomes genotypes and the 30x CRAMs come from
different sequencing datasets of the same individuals, so agreement includes
genotype-call and library differences, not only the GRCh37-to-GRCh38 derivation.
