---
title: "DuckHTS BCF record-cache ownership"
output: github_document
---

<!-- benchmark_bcf_record_cache.md is generated from this Rmd. -->

This paired workload materializes **every declared column** of the registered
GIAB HG002 GRCh38 NIST v4.2.1 benchmark VCF with `read_bcf`, in wide and tidy
form. The file has one sample, so input records
and output rows have the same denominator. This exercises typed INFO, GT, and
FORMAT decoding across many output chunks, not multi-sample expansion or CSQ.

Stage `variantkey_giab_hg002_v421` explicitly with the `duckhtsbench`
`duckhts_bench_fetch()` helper before running; rendering never downloads data.
Then build both revisions with the same toolchain and run from the repo root:

```sh
BCF_CACHE_BASELINE_EXTENSION=/path/to/baseline/duckhts.duckdb_extension \
BCF_CACHE_BASELINE_REV=<commit> \
Rscript -e 'rmarkdown::render("benchmarks/benchmark_bcf_record_cache.Rmd", knit_root_dir=getwd())'
```

The earlier `benchmark_bcf_record_cache.md` records the experiment at its named
source revisions. Preserve that report when measuring retirement: render this
Rmd with `output_file = "benchmark_bcf_reader_retirement.md"`. Set
`BCF_CACHE_READERS=read_bcf,read_bcf_v2` only to reproduce the historical comparison
with two old binaries that both expose the experimental reader.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(DBI)
library(duckdb)
root <- normalizePath(".", mustWork = TRUE)
Sys.setenv(DUCKHTSBENCH_REGISTRY = file.path(root, "r/duckhtsbench/inst/benchmark_registry.tsv"))
source("r/duckhtsbench/R/registry.R")
source("r/duckhtsbench/R/stage.R")
input <- duckhts_bench_artifact_path("variantkey_giab_hg002_v421")
duckhts_bench_validate_identity("variantkey_giab_hg002_v421", input)
extensions <- c(baseline = Sys.getenv("BCF_CACHE_BASELINE_EXTENSION"),
  candidate = Sys.getenv("BCF_CACHE_CANDIDATE_EXTENSION", "build/release/duckhts.duckdb_extension"))
extensions <- vapply(extensions, normalizePath, character(1), mustWork = TRUE)
revisions <- c(baseline = Sys.getenv("BCF_CACHE_BASELINE_REV"),
  candidate = system2("git", c("rev-parse", "HEAD"), stdout = TRUE))
stopifnot(nzchar(revisions["baseline"]))
if (system2("git", c("diff", "--quiet", "HEAD", "--", "src")) != 0L)
  stop("Commit candidate extension sources before revision-labelled measurements")
source_trees <- vapply(revisions, function(revision)
  system2("git", c("rev-parse", paste0(revision, ":src")), stdout = TRUE), character(1))
runs <- as.integer(Sys.getenv("BCF_CACHE_RUNS", "3"))
readers <- strsplit(Sys.getenv("BCF_CACHE_READERS", "read_bcf"), ",", fixed = TRUE)[[1L]]
stopifnot(length(readers) > 0L, !anyDuplicated(readers),
  all(readers %in% c("read_bcf", "read_bcf_v2")))
stopifnot(runs > 0L, nzchar(Sys.which("bcftools")))
# This independent HTSlib executable supplies the input denominator, not either
# reader's optimized COUNT path. No VCF-derived fixture or private cache is made.
count_command <- paste("bcftools query -f", shQuote("%CHROM\n"), shQuote(input), "| wc -l")
input_rows <- as.numeric(system2("bash", c("-o", "pipefail", "-c", shQuote(count_command)), stdout = TRUE))
stopifnot(length(input_rows) == 1L, !is.na(input_rows), input_rows > 0)
samples <- system2("bcftools", c("query", "-l", shQuote(input)), stdout = TRUE)
stopifnot(identical(samples, "HG002"))
directory <- tempfile("duckhts_bcf_cache_bench_")
dir.create(directory)
timings <- comparisons <- list()
con <- dbConnect(duckdb())
dbExecute(con, "SET threads=1")
dbExecute(con, "SET memory_limit='16GB'")
for (mode in c("wide", "tidy")) {
  snapshots <- character()
  for (reader in readers) for (run in seq_len(runs)) {
    # Reverse revision order on alternate repetitions to expose order effects.
    order <- if (run %% 2L) names(extensions) else rev(names(extensions))
    for (revision in order) {
      key <- paste(mode, reader, revision, sep = "_")
      snapshot <- if (run == 1L) file.path(directory, paste0(key, ".parquet")) else ""
      if (nzchar(snapshot)) snapshots[key] <- snapshot
      output <- file.path(directory, "timing.tsv")
      args <- c("benchmarks/bcf_record_cache_run.R", shQuote(extensions[revision]),
        shQuote(input), reader, mode, input_rows, shQuote(snapshot), shQuote(output))
      log <- system2(file.path(R.home("bin"), "Rscript"), args, stdout = TRUE, stderr = TRUE)
      status <- attr(log, "status")
      if (!is.null(status) && status != 0L) stop(paste(log, collapse = "\n"))
      timings[[length(timings) + 1L]] <- data.frame(mode, reader, revision, run,
        read.delim(output, colClasses = c(checksum = "character")))
    }
  }
  # Compare typed row multisets, including multiplicity, without imposing file
  # arrival order. First-run snapshots from every case must match exactly.
  base <- sprintf("read_parquet(%s)", dbQuoteString(con, snapshots[1]))
  for (i in seq_along(snapshots)[-1]) {
    other <- sprintf("read_parquet(%s)", dbQuoteString(con, snapshots[i]))
    sql <- sprintf(paste("SELECT count(*) AS n FROM ((SELECT * FROM %s EXCEPT ALL SELECT * FROM %s)",
      "UNION ALL (SELECT * FROM %s EXCEPT ALL SELECT * FROM %s))"), base, other, other, base)
    difference <- as.numeric(dbGetQuery(con, sql)$n)
    stopifnot(difference == 0)
    comparisons[[length(comparisons) + 1L]] <- data.frame(mode,
      reference = names(snapshots)[1], compared = names(snapshots)[i], different_rows = difference)
  }
  unlink(unname(snapshots))
}
dbDisconnect(con, shutdown = TRUE)
timings <- do.call(rbind, timings)
comparisons <- do.call(rbind, comparisons)
for (mode in unique(timings$mode)) {
  stopifnot(length(unique(timings$checksum[timings$mode == mode])) == 1L)
}
```

## Revisions, input, and environment

```{r identity}
print(data.frame(revision = names(revisions), commit = unname(revisions),
  src_tree = unname(source_trees), extension_md5 = unname(tools::md5sum(extensions))), row.names = FALSE)
cat("Artifact: variantkey_giab_hg002_v421; registered bytes:", file.info(input)$size, "\n")
cat("Observed input MD5:", unname(tools::md5sum(input)), "\n")
cat("Input records:", input_rows, "; sample: HG002; output rows per scan:", input_rows, "\n")
cat("DuckDB threads: 1; sequential scan handles: 1; HTSlib decompression workers: 0\n")
cat("Measured repetitions per reader/mode/revision:", runs, "\n")
cat("R:", R.version.string, "; DuckDB:", as.character(packageVersion("duckdb")), "\n")
cat(system2("uname", "-srmo", stdout = TRUE), "\n")
cat(system2("clang", "--version", stdout = TRUE)[1], "\n")
cat(system2("bcftools", "--version", stdout = TRUE)[1:2], sep = "\n")
cat("\n", system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE), "\n")
cat(system2("sh", c("-c", shQuote("lscpu | sed -n '/Model name:/p'")), stdout = TRUE), "\n")
```

## Full typed materialization

Elapsed and user + system CPU times include binding, initialization, VCF/BGZF
decoding, typed vectors, and `CREATE TABLE AS` materialization. Every repetition
uses a fresh R/DuckDB process. Startup, staging, independent counting, checksum
validation, Parquet snapshot writing, and exact comparison are outside timing.
The OS page cache is warm. RSS is the whole-process high-water mark immediately
after materialization, not an allocation count or cache-only measurement.

```{r timing}
summary <- aggregate(cbind(elapsed, cpu, peak_rss_kib) ~ mode + reader + revision,
  timings, median)
summary$input_records <- input_rows
summary$output_rows <- input_rows
summary$rows_per_second <- input_rows / summary$elapsed
print(summary, row.names = FALSE, digits = 6)
print(timings[c("mode", "reader", "revision", "run", "rows", "columns", "elapsed", "cpu")],
  row.names = FALSE, digits = 6)
```

## Output agreement and scope

```{r agreement}
print(comparisons, row.names = FALSE)
print(unique(timings[c("mode", "rows", "columns", "checksum")]), row.names = FALSE)
```

Each first-run snapshot is compared against the pre-change `read_bcf` output
with `EXCEPT ALL` in both directions. Every repetition also checks the independent
record count and a sum of hashes covering every materialized field. Hash agreement
alone is not an exact oracle; the separate first-run comparisons check complete
typed row multisets. The registry pins the supplier byte count; the input MD5
above is observed at measurement time, not a supplier-authenticated checksum.

The earlier [appender contig report](benchmark_bcf_appender_contigs.md) exercises
a narrow scalar appender with different data and scheduling. It is not an
identical-workload baseline; the paired pre-change build here supplies that baseline.
This report does not measure indexed or remote scans, appender pipelining, many-sample
throughput, CSQ parsing throughput, or bounded/static allocation. It establishes
neither universal performance parity nor a fastest-reader claim.

The independent SQL/R regression fixtures separately exercise two-sample CSQ
replication, all 2,053 sample rows of a record crossing an output chunk, missing
annotations and FORMAT values, growing list/string buffers, and changing ploidy.
The pre-change legacy reader returns incorrect CSQ rows; that incorrect output
is not used as a valid throughput baseline here. Reader names, scan defaults,
sample-selection policies, and output ordering guarantees remain unchanged.
