---
title: "DuckHTS VariantKey Join and Overlap Benchmark"
output: github_document
---

<!-- benchmark_variantkey_join_overlap.md is generated from benchmark_variantkey_join_overlap.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This report focuses on the workloads that matter for a future DuckVEP-style
annotation layer:

1. exact variant joins with stored `VariantKey`
2. safe mixed joins where hashed/nonreversible rows refine on exact alleles
3. exact span joins with stored `RegionKey`
4. true interval overlap queries through `duckhts_cgranges_*`

This is intentionally separate from `benchmark_variantkey_conformance.Rmd`.
That report measures encoding and `%VKX` parity. This one measures join and
overlap throughput.

## Run

```sh
make bench-variantkey-join
```

Useful overrides:

- `VARIANTKEY_JOIN_ROWS`: synthetic exact-join rows, default `1000000`
- `VARIANTKEY_INTERVAL_ROWS`: synthetic interval rows, default `250000`
- `VARIANTKEY_JOIN_RUNS`: timed repeats, default `3`
- `VARIANTKEY_JOIN_THREADS_LIST`: DuckDB thread grid, default `1,2,4`
- `DUCKHTS_EXTENSION`: optional extension path override

```{r setup, include = FALSE}
library(DBI)
library(duckdb)

ext_path <- Sys.getenv(
  "DUCKHTS_EXTENSION",
  unset = normalizePath("build/release/duckhts.duckdb_extension", mustWork = TRUE)
)

join_rows <- as.integer(Sys.getenv("VARIANTKEY_JOIN_ROWS", unset = "1000000"))
interval_rows <- as.integer(Sys.getenv("VARIANTKEY_INTERVAL_ROWS", unset = "250000"))
bench_runs <- as.integer(Sys.getenv("VARIANTKEY_JOIN_RUNS", unset = "3"))
thread_grid_env <- Sys.getenv("VARIANTKEY_JOIN_THREADS_LIST", unset = "1,2,4")

parse_thread_grid <- function(x) {
  vals <- trimws(strsplit(x, ",", fixed = TRUE)[[1]])
  vals <- vals[nzchar(vals)]
  out <- unique(as.integer(vals))
  if (!length(out) || any(is.na(out)) || any(out < 1L)) {
    stop("VARIANTKEY_JOIN_THREADS_LIST must be a comma-separated list of positive integers", call. = FALSE)
  }
  out
}

thread_grid <- parse_thread_grid(thread_grid_env)

if (is.na(join_rows) || join_rows < 1L) stop("VARIANTKEY_JOIN_ROWS must be positive", call. = FALSE)
if (is.na(interval_rows) || interval_rows < 1L) stop("VARIANTKEY_INTERVAL_ROWS must be positive", call. = FALSE)
if (is.na(bench_runs) || bench_runs < 1L) stop("VARIANTKEY_JOIN_RUNS must be positive", call. = FALSE)

benchmark_settings <- data.frame(
  setting = c(
    "variant_join_rows",
    "interval_rows",
    "benchmark_runs",
    "duckdb_threads_grid",
    "mixed_hash_fraction"
  ),
  value = c(
    as.character(join_rows),
    as.character(interval_rows),
    as.character(bench_runs),
    paste(thread_grid, collapse = ","),
    "20%"
  ),
  stringsAsFactors = FALSE
)

drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
con <- dbConnect(drv, dbdir = ":memory:")
DBI::dbExecute(con, sprintf("LOAD '%s';", gsub("\\\\", "/", ext_path)))
```

```{r helpers, include = FALSE}
rows_per_second <- function(rows, seconds) {
  rows / seconds
}

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

set_duck_threads <- function(n) {
  DBI::dbExecute(con, sprintf("PRAGMA threads=%d;", as.integer(n)))
}

measure_runs <- function(fun, runs = bench_runs) {
  times <- numeric(runs)
  value <- NULL
  for (i in seq_len(runs)) {
    gc()
    times[i] <- system.time({
      value <- fun()
    })[["elapsed"]]
  }
  list(
    value = value,
    times = times,
    median_seconds = unname(stats::median(times))
  )
}

measure_thread_grid <- function(operation, rows, fun) {
  do.call(
    rbind,
    lapply(thread_grid, function(th) {
      set_duck_threads(th)
      m <- measure_runs(fun)
      data.frame(
        operation = operation,
        threads = th,
        rows = rows,
        median_seconds = m$median_seconds,
        rows_per_second = rows_per_second(rows, m$median_seconds),
        stringsAsFactors = FALSE
      )
    })
  )
}
```

```{r synthetic-data-setup, include = FALSE}
DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TEMP TABLE vk_ann_reversible AS",
      "SELECT '1' AS chrom,",
      "       i::BIGINT AS pos,",
      "       'A' AS ref,",
      "       'C' AS alt,",
      "       variantkey('1', i, 'A', 'C') AS vk,",
      "       i::BIGINT AS payload",
      "FROM range(1, %d) t(i)"
    ),
    join_rows + 1L
  )
)
DBI::dbExecute(
  con,
  "CREATE OR REPLACE TEMP TABLE vk_probe_reversible AS SELECT * FROM vk_ann_reversible"
)

DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TEMP TABLE vk_ann_mixed AS",
      "WITH src AS (",
      "  SELECT '1' AS chrom,",
      "         i::BIGINT AS pos,",
      "         CASE WHEN i %% 5 = 0 THEN 'AAAAAAAAAAAA' ELSE 'A' END AS ref,",
      "         CASE WHEN i %% 5 = 0 THEN 'T' ELSE 'C' END AS alt,",
      "         i::BIGINT AS payload",
      "  FROM range(1, %d) t(i)",
      "), keyed AS (",
      "  SELECT src.*, variantkey(chrom, pos, ref, alt) AS vk",
      "  FROM src",
      ")",
      "SELECT keyed.*, (mod(extract_variantkey_refalt(vk), 2) = 1) AS is_hash",
      "FROM keyed"
    ),
    join_rows + 1L
  )
)
DBI::dbExecute(
  con,
  "CREATE OR REPLACE TEMP TABLE vk_probe_mixed AS SELECT * FROM vk_ann_mixed"
)

DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TEMP TABLE rk_ann AS",
      "WITH src AS (",
      "  SELECT '1' AS chrom,",
      "         ((i - 1) * 10)::BIGINT AS start,",
      "         (((i - 1) * 10) + 20)::BIGINT AS \"end\",",
      "         i::BIGINT AS label",
      "  FROM range(1, %d) t(i)",
      ")",
      "SELECT src.*, regionkey(chrom, start, \"end\") AS rk FROM src"
    ),
    interval_rows + 1L
  )
)
DBI::dbExecute(con, "CREATE OR REPLACE TEMP TABLE rk_probe_eq AS SELECT * FROM rk_ann")

DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TABLE ov_subject AS",
      "SELECT '1' AS chrom,",
      "       ((i - 1) * 20)::BIGINT AS start,",
      "       (((i - 1) * 20) + 30)::BIGINT AS \"end\",",
      "       i::BIGINT AS label",
      "FROM range(1, %d) t(i)"
    ),
    interval_rows + 1L
  )
)
DBI::dbExecute(
  con,
  sprintf(
    paste(
      "CREATE OR REPLACE TABLE ov_probe AS",
      "SELECT '1' AS chrom,",
      "       (((i - 1) * 20) + 10)::BIGINT AS start,",
      "       (((i - 1) * 20) + 15)::BIGINT AS \"end\",",
      "       i::BIGINT AS probe_id",
      "FROM range(1, %d) t(i)"
    ),
    interval_rows + 1L
  )
)

idx_name <- "vk_join_overlap_idx"
try(DBI::dbGetQuery(con, sprintf("SELECT duckhts_cgranges_destroy('%s') AS ok", idx_name)), silent = TRUE)
DBI::dbGetQuery(
  con,
  paste0(
    "SELECT duckhts_cgranges_from_query(",
    sql_string(idx_name), ", ",
    sql_string('SELECT chrom, start, "end", label FROM ov_subject'), ", ",
    sql_string("chrom"), ", ",
    sql_string("start"), ", ",
    sql_string("end"), ", ",
    sql_string("label"),
    ") AS ok"
  )
)
DBI::dbGetQuery(con, sprintf("SELECT duckhts_cgranges_index('%s') AS ok", idx_name))
```

## Benchmark settings

```{r benchmark-settings}
benchmark_settings
```

## 1. Exact reversible VariantKey joins

This section benchmarks exact self-joins over normalized reversible variants.

```{r reversible-joins}
reversible_multicol <- measure_thread_grid(
  "reversible_multicolumn_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_reversible p",
        "JOIN vk_ann_reversible a",
        "  ON p.chrom = a.chrom AND p.pos = a.pos AND p.ref = a.ref AND p.alt = a.alt"
      )
    )
  }
)

reversible_stored_vk <- measure_thread_grid(
  "reversible_stored_vk_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      "SELECT count(*) AS n FROM vk_probe_reversible p JOIN vk_ann_reversible a USING (vk)"
    )
  }
)

reversible_on_the_fly_vk <- measure_thread_grid(
  "reversible_on_the_fly_vk_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_reversible p",
        "JOIN vk_ann_reversible a",
        "  ON variantkey(p.chrom, p.pos, p.ref, p.alt) = a.vk"
      )
    )
  }
)

reversible_join_results <- rbind(
  reversible_multicol,
  reversible_stored_vk,
  reversible_on_the_fly_vk
)
reversible_join_results
```

## 2. Mixed joins with hashed-row refinement

This section benchmarks the safe join policy for mixed data:

- join on stored `vk`
- refine only when the probe row is hashed / nonreversible

```{r mixed-joins}
mixed_multicol <- measure_thread_grid(
  "mixed_multicolumn_join",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_mixed p",
        "JOIN vk_ann_mixed a",
        "  ON p.chrom = a.chrom AND p.pos = a.pos AND p.ref = a.ref AND p.alt = a.alt"
      )
    )
  }
)

mixed_stored_vk <- measure_thread_grid(
  "mixed_stored_vk_join_only",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      "SELECT count(*) AS n FROM vk_probe_mixed p JOIN vk_ann_mixed a USING (vk)"
    )
  }
)

mixed_safe_refine <- measure_thread_grid(
  "mixed_stored_vk_join_plus_hash_refine",
  join_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM vk_probe_mixed p",
        "JOIN vk_ann_mixed a",
        "  ON p.vk = a.vk",
        " AND (NOT p.is_hash OR (p.chrom = a.chrom AND p.pos = a.pos AND p.ref = a.ref AND p.alt = a.alt))"
      )
    )
  }
)

mixed_join_results <- rbind(
  mixed_multicol,
  mixed_stored_vk,
  mixed_safe_refine
)
mixed_join_results
```

## 3. Exact RegionKey span joins

`RegionKey` is useful for exact stored span equality. It is not the general
interval-overlap engine; that comes next.

```{r regionkey-joins}
region_multicol <- measure_thread_grid(
  "region_multicolumn_join",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      paste(
        "SELECT count(*) AS n",
        "FROM rk_probe_eq p",
        "JOIN rk_ann a",
        "  ON p.chrom = a.chrom AND p.start = a.start AND p.\"end\" = a.\"end\""
      )
    )
  }
)

region_rk <- measure_thread_grid(
  "region_stored_rk_join",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      "SELECT count(*) AS n FROM rk_probe_eq p JOIN rk_ann a USING (rk)"
    )
  }
)

region_join_results <- rbind(region_multicol, region_rk)
region_join_results
```

## 4. cgranges overlap probes

This is the actual overlap path of interest for transcript/gene/regulatory
annotation candidate generation.

```{r cgranges-overlap}
cgranges_has_overlap <- measure_thread_grid(
  "cgranges_has_overlap_filter",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT count(*) AS n",
          "FROM ov_probe p",
          "WHERE duckhts_cgranges_has_overlap('%s', p.chrom, p.start, p.\"end\")"
        ),
        idx_name
      )
    )
  }
)

cgranges_count_overlap <- measure_thread_grid(
  "cgranges_count_overlaps_sum",
  interval_rows,
  function() {
    DBI::dbGetQuery(
      con,
      sprintf(
        paste(
          "SELECT sum(duckhts_cgranges_count_overlaps('%s', p.chrom, p.start, p.\"end\"))::BIGINT AS n",
          "FROM ov_probe p"
        ),
        idx_name
      )
    )
  }
)

cgranges_overlap_results <- rbind(cgranges_has_overlap, cgranges_count_overlap)
cgranges_overlap_results
```

## Notes

- The exact-join sections are the relevant comparison class for future
  `echtvar`-style annotation throughput work.
- The `mixed_stored_vk_join_plus_hash_refine` query represents the safe SQL
  pattern for hashed/nonreversible VariantKey rows when exact `REF/ALT`
  verification is required.
- `RegionKey` equality is useful for exact stored spans, but real overlap work
  should go through `duckhts_cgranges_*` or later overlap kernels rather than
  a plain `rk = rk` join.
- For richer overlap comparisons against external tools such as `bedtk` or
  `bedtools`, see the existing `benchmarks/benchmark_cgranges.Rmd` report.
