---
title: "DuckHTS VariantKey Conformance"
output: github_document
---

<!-- benchmark_variantkey_conformance.md is generated from benchmark_variantkey_conformance.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Conformance

This report compares DuckHTS VariantKey output against installed `bcftools`
`%VKX` formatting on real whole-genome GIAB VCF files.

DuckHTS uses the vendored official VariantKey C API and follows the bcftools
`%VKX` / `+add-variantkey` convention of accepting 1-based VCF `POS` at the SQL
boundary while encoding the upstream 0-based field internally.

VariantKey / RegionKey are based on Nicola Asuni's work:
<https://doi.org/10.1101/473744>.

## Run

```sh
make bench-variantkey
```

Useful overrides:

- `BCFTOOLS_BIN`: optional bcftools executable override
- `VARIANTKEY_REAL_CASES`: optional comma-separated local VCF/BCF paths
- `VARIANTKEY_OUT_DIR`: optional directory for emitted TSVs and summaries
- `DUCKHTS_EXTENSION`: optional extension path override

If `VARIANTKEY_REAL_CASES` is unset, this report looks for the locally
downloaded whole-WGS GIAB files already used in earlier DuckHTS conformance
work under `/root/giab_norm/`.

```{r setup, include = FALSE}
library(utils)

if (!requireNamespace("RBCFTools", quietly = TRUE)) {
  stop("RBCFTools is required for VariantKey conformance rendering", call. = FALSE)
}

RBCFTools::setup_hts_env()

bcftools_bin <- Sys.getenv("BCFTOOLS_BIN", unset = RBCFTools::bcftools_path())
if (!file.exists(bcftools_bin)) stop("bcftools binary not found: ", bcftools_bin, call. = FALSE)

ext_path <- Sys.getenv(
  "DUCKHTS_EXTENSION",
  unset = normalizePath("build/release/duckhts.duckdb_extension", mustWork = TRUE)
)
script_path <- normalizePath("scripts/variantkey_conformance.sh", mustWork = TRUE)
out_root <- Sys.getenv("VARIANTKEY_OUT_DIR", unset = file.path(tempdir(), "duckhts_variantkey_real"))
dir.create(out_root, recursive = TRUE, showWarnings = FALSE)

case_env <- Sys.getenv("VARIANTKEY_REAL_CASES", unset = "")
default_cases <- data.frame(
  label = c("giab_hg001_grch37_full", "giab_hg002_grch37_full", "giab_hg006_grch37_full"),
  path = c(
    "/root/giab_norm/HG001_GRCh37_1_22_v4.2.1_benchmark.vcf.gz",
    "/root/giab_norm/HG002_GRCh37_1_22_v4.2.1_benchmark.vcf.gz",
    "/root/giab_norm/HG006_GRCh37_1_22_v4.2.1_benchmark.vcf.gz"
  ),
  stringsAsFactors = FALSE
)

if (nzchar(case_env)) {
  raw_paths <- trimws(strsplit(case_env, ",", fixed = TRUE)[[1]])
  raw_paths <- raw_paths[nzchar(raw_paths)]
  cases <- data.frame(
    label = sub("\\.(vcf|bcf)(\\.gz)?$", "", basename(raw_paths), ignore.case = TRUE),
    path = raw_paths,
    stringsAsFactors = FALSE
  )
} else {
  cases <- default_cases[file.exists(default_cases$path), , drop = FALSE]
}

if (!nrow(cases)) {
  stop(
    "No real GIAB inputs found. Set VARIANTKEY_REAL_CASES or place local files under /root/giab_norm/.",
    call. = FALSE
  )
}

run_case <- function(label, path) {
  case_out <- file.path(out_root, label)
  dir.create(case_out, recursive = TRUE, showWarnings = FALSE)

  env <- c(
    sprintf("VARIANTKEY_VCF=%s", path),
    sprintf("VARIANTKEY_LABEL=%s", label),
    sprintf("VARIANTKEY_OUT_DIR=%s", case_out),
    sprintf("VARIANTKEY_BCFTOOLS_BIN=%s", bcftools_bin),
    sprintf("DUCKHTS_EXTENSION=%s", ext_path)
  )

  elapsed <- system.time({
    status <- system2("bash", script_path, env = env, stdout = TRUE, stderr = TRUE)
    attr(status, "status") <- NULL
  })[["elapsed"]]

  summary_path <- file.path(case_out, sprintf("%s.summary.tsv", label))
  compare_path <- file.path(case_out, sprintf("%s.compare.tsv", label))
  summary <- read.delim(summary_path, sep = "\t", check.names = FALSE)
  mismatches <- read.delim(compare_path, sep = "\t", check.names = FALSE)

  data.frame(
    label = label,
    path = path,
    bcftools_rows = summary$bcftools_rows[1],
    duckhts_rows = summary$duckhts_rows[1],
    mismatch_groups = summary$mismatch_groups[1],
    elapsed_seconds = elapsed,
    stringsAsFactors = FALSE
  )
}
```

```{r run-real-cases}
results <- do.call(
  rbind,
  lapply(seq_len(nrow(cases)), function(i) run_case(cases$label[i], cases$path[i]))
)
results
```

All listed cases should have `mismatch_groups == 0`.

```{r stop-on-mismatch, include = FALSE}
if (any(results$mismatch_groups != 0)) {
  stop("VariantKey conformance mismatches detected; inspect VARIANTKEY_OUT_DIR artifacts.", call. = FALSE)
}
```

## Notes

- This framework compares against bcftools `%VKX`, which is the same VariantKey
  hexadecimal output exposed by `+add-variantkey`.
- DuckHTS currently compares `ALT[1]` from `read_bcf(...)`, mirroring the
  biallelic `%VKX` field that bcftools derives from the first ALT allele on the
  current record.
- Large, ambiguous, and symbolic alleles are expected to fall back to the
  official hashed nonreversible VariantKey mode. That still preserves bcftools
  `%VKX` parity, but it does not encode `END`, `SVLEN`, mate breakend
  coordinates, or other SV metadata.
