---
title: "DuckHTS SIMD seq_gc_content Benchmark"
output: github_document
---

<!-- benchmark_simd_seq_gc.md is generated from benchmark_simd_seq_gc.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This report measures the first DuckHTS runtime-SIMD dispatch path:
`seq_gc_content(...)` uses the shared byte-oriented base-count vtable with a
scalar fallback and optional platform-specific SIMD implementations such as
AVX2. The benchmark requests backends explicitly through
`duckhts_simd_set_backend(...)`; unavailable requests are reported as skipped so
the same report remains valid on x86, ARM, and wasm builds.

The goal is to decide whether a SIMD path is worth keeping before expanding the
same dispatch scaffold to quality-score scans, BAM packed-sequence GC counts,
depth reductions, and variant normalization helpers.

# Run

```sh
make bench-simd-seq-gc
```

Useful overrides:

- `SIMD_SEQ_GC_ROWS`: number of rows, default `200000`
- `SIMD_SEQ_GC_LEN`: sequence length per row, default `512`
- `SIMD_SEQ_GC_RUNS`: timed repeats per backend, default `7`
- `SIMD_SEQ_GC_BACKENDS`: comma-separated backend requests, default `scalar,auto,avx2,avx512`
- `SIMD_SEQ_GC_PATTERN`: repeated sequence pattern, default `ACGTNNacgtnn`
- `SIMD_SEQ_GC_FORCE=1`: rerun even if cached JSON exists
- `DUCKHTS_EXTENSION`: extension path override

```{r setup, include = FALSE}
library(knitr)
library(jsonlite)

repo_root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", unset = getwd()), mustWork = TRUE)
out_dir <- file.path(repo_root, ".tmp", "simd_seq_gc")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
json_out <- file.path(out_dir, "benchmark_simd_seq_gc.json")

extension <- Sys.getenv(
  "DUCKHTS_EXTENSION",
  unset = file.path(repo_root, "build", "release", "duckhts.duckdb_extension")
)
extension <- normalizePath(extension, mustWork = TRUE)

rows <- as.integer(Sys.getenv("SIMD_SEQ_GC_ROWS", unset = "200000"))
seq_len <- as.integer(Sys.getenv("SIMD_SEQ_GC_LEN", unset = "512"))
runs <- as.integer(Sys.getenv("SIMD_SEQ_GC_RUNS", unset = "7"))
backends <- Sys.getenv("SIMD_SEQ_GC_BACKENDS", unset = "scalar,auto,avx2,avx512")
pattern <- Sys.getenv("SIMD_SEQ_GC_PATTERN", unset = "ACGTNNacgtnn")
force <- identical(Sys.getenv("SIMD_SEQ_GC_FORCE", unset = "0"), "1")
python <- Sys.getenv("PYTHON", unset = "python3")

if (is.na(rows) || rows < 1L) stop("SIMD_SEQ_GC_ROWS must be positive", call. = FALSE)
if (is.na(seq_len) || seq_len < 1L) stop("SIMD_SEQ_GC_LEN must be positive", call. = FALSE)
if (is.na(runs) || runs < 1L) stop("SIMD_SEQ_GC_RUNS must be positive", call. = FALSE)
if (!nzchar(pattern)) stop("SIMD_SEQ_GC_PATTERN must be non-empty", call. = FALSE)

meta_out <- file.path(out_dir, "benchmark_simd_seq_gc_meta.json")
requested_meta <- list(
  schema = 2L,
  rows = rows,
  seq_len = seq_len,
  runs = runs,
  backends = backends,
  pattern = pattern,
  extension = extension
)
needs_run <- force || !file.exists(json_out) || !file.exists(meta_out)
if (!needs_run) {
  old_meta <- tryCatch(jsonlite::fromJSON(meta_out, simplifyVector = FALSE), error = function(e) list())
  needs_run <- !identical(old_meta, requested_meta)
}

if (needs_run) {
  cmd <- c(
    file.path(repo_root, "scripts", "benchmark_simd_seq_gc.py"),
    "--extension", extension,
    "--rows", as.character(rows),
    "--seq-len", as.character(seq_len),
    "--iterations", as.character(runs),
    "--pattern", pattern,
    "--backends", backends,
    "--json-out", json_out
  )
  out <- system2(python, cmd, stdout = TRUE, stderr = TRUE)
  status <- attr(out, "status")
  if (!is.null(status) && status != 0) {
    stop(paste(out, collapse = "\n"), call. = FALSE)
  }
  writeLines(jsonlite::toJSON(requested_meta, auto_unbox = TRUE, pretty = TRUE), meta_out)
}

raw_results <- jsonlite::fromJSON(json_out, simplifyVector = FALSE)
null_chr <- function(x) if (is.null(x)) NA_character_ else as.character(x)
null_num <- function(x) if (is.null(x)) NA_real_ else as.numeric(x)
null_lgl <- function(x) if (is.null(x)) FALSE else isTRUE(x)

results <- do.call(rbind, lapply(raw_results, function(x) {
  data.frame(
    backend_request = null_chr(x$backend_request),
    requested_backend = null_chr(x$requested_backend),
    selected_backend = null_chr(x$selected_backend),
    available = if (is.null(x$available)) NA else isTRUE(x$available),
    skipped = null_lgl(x$skipped),
    skip_reason = null_chr(x$skip_reason),
    rows = as.integer(x$rows),
    seq_len = as.integer(x$seq_len),
    total_bases = as.numeric(x$total_bases),
    iterations = as.integer(x$iterations),
    median_sec = null_num(x$median_sec),
    min_sec = null_num(x$min_sec),
    mbases_per_sec_median = null_num(x$mbases_per_sec_median),
    stringsAsFactors = FALSE
  )
}))

scalar_idx <- which(!results$skipped & results$backend_request == "scalar")[1]
scalar_rate <- if (length(scalar_idx) && !is.na(scalar_idx)) results$mbases_per_sec_median[scalar_idx] else NA_real_
results$speedup_vs_scalar <- if (is.finite(scalar_rate) && scalar_rate > 0) {
  results$mbases_per_sec_median / scalar_rate
} else {
  NA_real_
}
results$status <- ifelse(results$skipped, "skipped", "ok")

extension_display <- if (startsWith(extension, paste0(repo_root, .Platform$file.sep))) {
  sub(paste0("^", repo_root, .Platform$file.sep), "", extension)
} else {
  extension
}

configuration <- data.frame(
  setting = c("rows", "sequence_length", "total_bases", "iterations", "backend_requests", "pattern", "extension"),
  value = c(
    format(rows, big.mark = ","),
    format(seq_len, big.mark = ","),
    format(rows * seq_len, big.mark = ","),
    as.character(runs),
    backends,
    pattern,
    extension_display
  ),
  stringsAsFactors = FALSE
)
```

# Configuration

```{r configuration, echo=FALSE}
knitr::kable(configuration)
```

# Results

```{r results, echo=FALSE}
display <- results[, c(
  "backend_request",
  "requested_backend",
  "selected_backend",
  "status",
  "median_sec",
  "min_sec",
  "mbases_per_sec_median",
  "speedup_vs_scalar",
  "skip_reason"
)]
knitr::kable(display, digits = 3)
```

# Interpretation

```{r interpretation, echo=FALSE, results='asis'}
measured <- results[!results$skipped, , drop = FALSE]
if (!nrow(measured)) {
  cat("No backend requests were available for measurement.\n")
} else {
  best <- measured[which.max(measured$mbases_per_sec_median), , drop = FALSE]
  cat(sprintf(
    "The fastest measured backend was `%s` selected from request `%s`, at %.1f Mbases/s.\n\n",
    best$selected_backend, best$backend_request, best$mbases_per_sec_median
  ))
  if (is.finite(best$speedup_vs_scalar)) {
    cat(sprintf(
      "That is %.2fx the scalar baseline for this synthetic workload.\n\n",
      best$speedup_vs_scalar
    ))
  }
  cat("The benchmark is intentionally synthetic: it isolates the byte-scanning kernel and does not include FASTA/VCF/BAM I/O. Real end-to-end wins will be smaller when decompression, SQL materialization, and joins dominate.\n")
}

```

# Raw timing vectors

```{r raw-timings, echo=FALSE}
timing_rows <- do.call(rbind, lapply(raw_results, function(x) {
  if (isTRUE(x$skipped) || !length(x$timings_sec)) return(NULL)
  data.frame(
    backend_request = x$backend_request,
    selected_backend = x$selected_backend,
    iteration = seq_along(x$timings_sec),
    seconds = unlist(x$timings_sec),
    stringsAsFactors = FALSE
  )
}))
if (is.null(timing_rows) || !nrow(timing_rows)) {
  cat("No timing rows were recorded.\n")
} else {
  knitr::kable(timing_rows, digits = 6)
}
```
