---
title: "Indexed multi-region reader benchmark"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

This compares the canonical `read_bcf` indexed region path before and after
shared region-list validation and retirement of the appender experiment.
HTSlib 1.24 still owns interval union and record visitation: BCF uses
`bcf_itr_regarray`, and indexed VCF text uses `tbx_itr_regarray`.

The same-run baseline and candidate read identical cached files, with equal
schema, record and aggregate-output denominators. The nearest identical logical
workload was recorded at `2abd25645223` on July 18, 2026: 0.086 s BCF and
0.284 s VCF.gz medians, one DuckDB thread and no HTSlib workers. That historical
run used different compression tools; it is context, not an environment-matched
regression comparison. The old rendered report remains in git history.

## Reproduction and identity

Build the two named revisions separately. Set `DUCKHTS_REGION_BASELINE_EXTENSION`
and `DUCKHTS_REGION_BASELINE_REVISION` to the pre-fix binary and commit.
`DUCKHTS_EXTENSION` selects the candidate (default: release build).
Render on one allowed CPU, for example:

```bash
taskset -c 2 Rscript -e "rmarkdown::render('benchmarks/benchmark_multi_region_readers.Rmd')"
```

The network-free `duckhts_bench_stage_multi_region()` stage resolves the
synthetic VCF and its derived BCF/BGZF/index artifacts through the benchmark
registry. It regenerates the exact consecutive-SNV workload before timing;
creation, compression, indexing, connection setup and warm-up are excluded.
Each build uses a separate in-memory database connection, with baseline first.

```{r benchmark, echo=FALSE}
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
})
Sys.setenv(DUCKHTSBENCH_REGISTRY = "r/duckhtsbench/inst/benchmark_registry.tsv")
for (name in c("registry", "stage", "multi_region")) {
  source(file.path("r/duckhtsbench/R", paste0(name, ".R")))
}
paths <- duckhts_bench_stage_multi_region()
row_count <- as.integer(duckhts_bench_identity_fields(
  duckhts_bench_stage_plan("multi-region-readers")$supplier_identity[[1]]
)[["records"]])
region_count <- 64L
region_width <- 50000L
region_step <- 25000L
starts <- 1L + (seq_len(region_count) - 1L) * region_step
ends <- starts + region_width - 1L
stopifnot(max(ends) <= row_count)
region_spec <- paste(sprintf("1:%d-%d", starts, ends), collapse = ",")
expected_rows <- as.double(max(ends) - min(starts) + 1L)
expected_sum <- expected_rows * (min(starts) + max(ends)) / 2
sql_string <- function(x) sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
command_output <- function(command, args = character()) {
  paste(system2(command, args, stdout = TRUE, stderr = TRUE), collapse = " ")
}
baseline <- Sys.getenv("DUCKHTS_REGION_BASELINE_EXTENSION")
baseline_revision <- Sys.getenv("DUCKHTS_REGION_BASELINE_REVISION")
candidate <- Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension")
stopifnot(nzchar(baseline_revision), file.exists(baseline), file.exists(candidate))
stopifnot(length(system2("git", c("diff", "--name-only", "HEAD", "--", "src"), stdout = TRUE)) == 0L)
revision <- command_output("git", c("rev-parse", "HEAD"))
source_tree <- command_output("git", c("rev-parse", "HEAD:src"))

measure_build <- function(build, extension) {
  con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
  on.exit(dbDisconnect(con, shutdown = TRUE))
  dbExecute(con, sprintf("LOAD %s", sql_string(normalizePath(extension))))
  dbExecute(con, "SET threads=1")
  measure <- function(format, path) {
    sql <- sprintf(paste(
      "SELECT count(*) AS rows, sum(POS)::HUGEINT AS position_sum",
      "FROM read_bcf(%s, region := %s, decompression_threads := 0)"),
      sql_string(path), sql_string(region_spec))
    check <- function(answer) {
      stopifnot(as.double(answer$rows[[1]]) == expected_rows,
                as.double(answer$position_sum[[1]]) == expected_sum)
    }
    for (i in seq_len(2L)) check(dbGetQuery(con, sql))
    seconds <- numeric(9L)
    for (i in seq_along(seconds)) {
      seconds[[i]] <- system.time(answer <- dbGetQuery(con, sql))[["elapsed"]]
      check(answer)
    }
    data.frame(build = build, format = format, runs = length(seconds),
               reader_rows = expected_rows, sql_output_rows = 1L,
               min_seconds = min(seconds), median_seconds = median(seconds),
               max_seconds = max(seconds))
  }
  rbind(measure("BCF + CSI", paths[["bcf"]]),
        measure("VCF.gz + TBI", paths[["vcf_gz"]]))
}
results <- rbind(measure_build("baseline", baseline), measure_build("candidate", candidate))
metadata <- data.frame(
  property = c("baseline revision", "candidate revision", "candidate src tree",
               "baseline binary MD5", "candidate binary MD5", "run date",
               "DuckDB R package", "HTSlib", "CPU", "CPU affinity",
               "DuckDB threads", "HTSlib workers per handle", "input records per format",
               "requested intervals", "interval width", "interval step",
               "raw interval-record hits", "unique reader rows", "SQL aggregate rows",
               "bgzip version", "tabix version", "bcftools version"),
  value = c(baseline_revision, revision, source_tree,
            unname(tools::md5sum(baseline)), unname(tools::md5sum(candidate)),
            as.character(Sys.Date()), as.character(packageVersion("duckdb")),
            readLines("third_party/htslib/VERSION", n = 1),
            sub("^model name[[:space:]]*:[[:space:]]*", "",
                grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[[1]]),
            command_output("taskset", c("-pc", Sys.getpid())),
            1, 0, row_count, region_count, region_width, region_step,
            region_count * region_width, expected_rows, 1,
            command_output("bgzip", "--version"), command_output("tabix", "--version"),
            command_output("bcftools", "--version"))
)
knitr::kable(metadata)
knitr::kable(data.frame(artifact = names(paths), bytes = file.info(paths)$size,
                       md5 = unname(tools::md5sum(paths))), row.names = FALSE)
```

## Results

The reader-row denominator is the 1,625,000 stored records returned after
HTSlib unions 64 overlapping intervals, from 2,000,000 input records per format.
Independent interval scans would encounter 3,200,000 record hits. Each SQL query
returns one aggregate row. All nine timed runs validate both count and position
checksum; separate SQL/R tests cover long REF/END spans, missing targets and
physical duplicate records.

```{r results, echo=FALSE}
knitr::kable(results, digits = 3, row.names = FALSE)
comparison <- merge(results[results$build == "baseline", c("format", "median_seconds")],
                    results[results$build == "candidate", c("format", "median_seconds")],
                    by = "format", suffixes = c("_baseline", "_candidate"))
comparison$percent_change <- 100 * (comparison$median_seconds_candidate /
                                     comparison$median_seconds_baseline - 1)
knitr::kable(comparison, digits = 3, row.names = FALSE)
```

This is a warm-page-cache, one-core iterator/decode measurement, including
region parsing and initialization. It does not measure remote seeks, cold
storage, rich INFO/FORMAT decoding, every-column materialization, BAM/tabix/FASTA
reader performance, or staged pipelining. The removed appender cannot be
benchmarked at this revision; its historical narrow-row timings in
[benchmark_bcf_appender_contigs.md](benchmark_bcf_appender_contigs.md) are not
canonical-reader throughput measurements.
