---
title: "GenBank named-attribute projections"
output:
  github_document:
    html_preview: false
---

`read_genbank(attributes := [...])` adds one VARCHAR column per qualifier key.
The C API table function learns which columns a query projects, never which MAP
keys, so a query that needs one qualifier used to build the whole
`attributes_map` for every feature. This report compares that MAP projection
with named columns.

The input is the registered NCBI RefSeq record for *Escherichia coli* K-12
MG1655 (`genbank_ecoli_k12_gbff`, staged by
`duckhtsbench::duckhts_bench_stage_genbank()`), 11,450,954 bytes yielding 9,306
feature rows. Each measurement is one query in a fresh R process with one DuckDB
thread: the extension is loaded, the query runs once, and the process reports
elapsed seconds and its peak resident set (`VmHWM`). Three fresh processes per
workload run in an interleaved order; the table reports medians, and the
individual runs are in `benchmark_genbank_named_attributes_runs.csv`. Every
query returns one aggregate row, so transport is not measured. Named columns
carry the same values as the map lookups (`n` and `metric` agree between paired
workloads, and the SQL and R tests compare every column with the lookup).

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r benchmark, include=FALSE}
stopifnot(requireNamespace("duckhtsbench", quietly = TRUE))
stopifnot(requireNamespace("knitr", quietly = TRUE))
input <- normalizePath(
  duckhtsbench::duckhts_bench_stage_genbank(fetch = FALSE)[["gbff"]],
  mustWork = TRUE
)
extension <- normalizePath(
  Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"),
  mustWork = TRUE
)
worker <- normalizePath("benchmarks/benchmark_genbank_named_attributes_worker.R",
                        mustWork = TRUE)
revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE)
if (system2("git", c("diff", "--quiet", "HEAD", "--", "src"),
            stdout = FALSE, stderr = FALSE) != 0L) {
  stop("Build source must match the recorded HEAD")
}
quote_sql <- function(value) paste0("'", gsub("'", "''", value, fixed = TRUE), "'")
source_sql <- function(options = "") {
  paste0("read_genbank(", quote_sql(input), options, ")")
}
plain <- source_sql()
map <- source_sql(", attributes_map := true")
named <- source_sql(", attributes := ['product', 'locus_tag', 'gene']")
queries <- c(
  plain_columns = paste0("SELECT count(*) n, sum(start + \"end\") metric FROM ", plain),
  attributes_string = paste0("SELECT count(*) n, sum(length(attributes)) metric FROM ", plain),
  map_one = paste0("SELECT count(*) n, count(attributes_map['product']) metric FROM ", map),
  named_one = paste0("SELECT count(*) n, count(product) metric FROM ", named),
  map_three = paste0("SELECT count(*) n, count(attributes_map['product']) + count(attributes_map['locus_tag']) + count(attributes_map['gene']) metric FROM ", map),
  named_three = paste0("SELECT count(*) n, count(product) + count(locus_tag) + count(gene) metric FROM ", named),
  named_none_projected = paste0("SELECT count(*) n, sum(start + \"end\") metric FROM ", named)
)
repeats <- 3L
runs <- list()
for (run in seq_len(repeats)) {
  for (kind in names(queries)) {
    line <- system2("Rscript", c(shQuote(worker), shQuote(extension), shQuote(queries[[kind]])),
                    stdout = TRUE)
    field <- strsplit(line, "\t", fixed = TRUE)[[1L]]
    runs[[length(runs) + 1L]] <- data.frame(
      workload = kind, run = run, seconds = as.numeric(field[[1L]]),
      rss_before_query_mib = as.numeric(field[[2L]]),
      peak_rss_mib = as.numeric(field[[3L]]),
      input_rows = as.numeric(field[[4L]]), output_rows = 1L, metric = field[[5L]]
    )
  }
}
runs <- do.call(rbind, runs)
write.csv(runs, "benchmarks/benchmark_genbank_named_attributes_runs.csv", row.names = FALSE)
summary <- do.call(rbind, lapply(names(queries), function(kind) {
  rows <- runs[runs$workload == kind, ]
  stopifnot(length(unique(rows$metric)) == 1L, length(unique(rows$input_rows)) == 1L)
  data.frame(
    workload = kind, input_rows = rows$input_rows[[1L]], output_rows = 1L,
    metric = rows$metric[[1L]], median_seconds = median(rows$seconds),
    median_peak_rss_mib = median(rows$peak_rss_mib),
    input_mib_per_second = 11450954 / 1048576 / median(rows$seconds), runs = nrow(rows)
  )
}))
stopifnot(summary$metric[summary$workload == "map_one"] == summary$metric[summary$workload == "named_one"])
stopifnot(summary$metric[summary$workload == "map_three"] == summary$metric[summary$workload == "named_three"])
```

Source revision: `r revision`. Extension SHA-256: `r strsplit(system2("sha256sum", extension, stdout = TRUE), " ", fixed = TRUE)[[1L]][[1L]]`.
Input SHA-256: `r strsplit(system2("sha256sum", input, stdout = TRUE), " ", fixed = TRUE)[[1L]][[1L]]`.
DuckDB runtime: `r as.character(packageVersion("duckdb"))`. Host: `r trimws(sub("^Model name:", "", grep("^Model name:", system2("lscpu", stdout = TRUE), value = TRUE)[[1L]]))`; `uptime` before rendering: `r trimws(system2("uptime", stdout = TRUE))`.

```{r results, echo=FALSE}
knitr::kable(summary, digits = 3)
```

```{r ratios, echo=FALSE}
med <- function(kind, column) summary[[column]][summary$workload == kind]
ratio <- data.frame(
  keys = c(1L, 3L),
  map_seconds = c(med("map_one", "median_seconds"), med("map_three", "median_seconds")),
  named_seconds = c(med("named_one", "median_seconds"), med("named_three", "median_seconds")),
  map_peak_rss_mib = c(med("map_one", "median_peak_rss_mib"), med("map_three", "median_peak_rss_mib")),
  named_peak_rss_mib = c(med("named_one", "median_peak_rss_mib"), med("named_three", "median_peak_rss_mib"))
)
ratio$speedup <- ratio$map_seconds / ratio$named_seconds
knitr::kable(ratio, digits = 3)
```

## Reading the result

This is a single 1x input. Its runtime is below the 5 s floor that `STYLE.md`
(Scale) sets for timing claims, so the report makes no claim about how time
scales with input size and has no 2x/4x series. It reports the measured seconds,
peak resident memory and input throughput for this file only. Memory is the
process peak, dominated by DuckDB and R baseline plus the parser's one-record
buffers; `benchmark_genbank_memory.md` is the report for the memory budget across
input sizes.

`named_none_projected` requests three keys at bind but projects none, so the scan
computes no qualifier values. `map_one` builds the entire MAP for each feature to
read a single key, which is the cost named columns avoid; `named_one` and
`named_three` compute values only for matching keys. The paired workloads must
return identical metrics, and the render stops if they differ. Workloads with
different aggregation expressions are not isolated parsing costs. No earlier
recorded GenBank named-attribute workload exists; the nearest report for this
change is `benchmark_gff_named_attributes.md`, which measures a different
format on a different input and is not comparable.
