---
title: "Somalier find-sites: public 1000 Genomes slice"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
root <- normalizePath("..", mustWork = TRUE)
read_result <- function(name) utils::read.delim(
  file.path(root, "benchmarks", name), check.names = FALSE
)
observations <- read_result("somalier_find_sites_observations.tsv")
gates <- read_result("somalier_find_sites_gates.tsv")
gnomad <- read_result("somalier_find_sites_gnomad_gates.tsv")
disagreements <- read_result("somalier_find_sites_disagreements.tsv")
peak_kb <- as.numeric(readLines(file.path(root, "benchmarks",
  "somalier_find_sites_peak_kb.txt")))
```

This report measures `rduckhts_somalier_find_sites()` on a retained, unaltered
three-sample 1000 Genomes chromosome-1 VCF slice distributed by RBCFTools at
[commit 9adeaf4](https://github.com/RGenomicsETL/RBCFTools/blob/9adeaf4cfcc3bff40efca6237749fefb53391678/inst/extdata/1000G_3samples.vcf.gz).
The 1,586-byte input has SHA-256
`d5b129eb4f7e431e0e8d9b73229fe3e304ed14863d2cb26c055f631eabb34110`.
The differential also uses the unaltered 9-record, 32,182-byte
[gnomAD VCF fixture](https://github.com/pysam-developers/pysam/blob/79293854c45889bedef09491ee7d20de3dbc76a6/tests/cbcf_data/gnomad.vcf)
(SHA-256 `afb6133e19dca997769ab5c163389a42a28c241243898e513b3c823b2371a73a`).
The binary oracle is Somalier v0.3.4 at
[`ff58fda`](https://github.com/brentp/somalier/tree/ff58fdade8f4f8293d904f10e0a4a13f1fac808d)
(SHA-256 `18717c205a9c4b65d479f1d2cf069a30b047a5378edf544d403d4081f06a3a78`).
Inputs are staged through the artifact registry without changing physical records.

- Source revision: `r observations$revision[[1]]`.
- Workload: `--min-AN 6 --snp-dist 100`, default AF 0.15, target AF 0.48,
  one DuckDB thread; five repeats of the caller-connection R wrapper after
  staging and connection setup.
- Denominators: `r observations$source_records` source rows,
  `r observations$duckhts_candidates` gated candidates, and
  `r observations$duckhts_sites` selected rows. Somalier has
  `r observations$upstream_candidates` candidates and
  `r observations$upstream_sites` final sites.
- Wall time: median `r observations$median_seconds` s, minimum
  `r observations$min_seconds` s. Process peak resident memory:
  `r format(peak_kb / 1024, digits = 4)` MiB (GNU time, includes R,
  DuckDB and staging; **not** isolated query memory).
- Differential on `(region, position, source_ref, source_alt)`:
  `r nrow(disagreements)` disagreements, all retained in
  `somalier_find_sites_disagreements.tsv`. The second public slice has
  `r gnomad$records[gnomad$gate == 'src']` source rows,
  `r gnomad$records[gnomad$gate == 'gated']` candidates, and
  `r gnomad$records[gnomad$gate == 'selected']` selected rows, equal to
  the pinned binary at `--min-AN 100 --snp-dist 100`.

```{r gates}
knitr::kable(gates, caption = "Source records and candidate gates. Indels and snps are exclusion-neighbor inventories, not sequential pass counts.")
```

The small public slice verifies the pipeline but does not predict memory or time
on a full population cohort. The nearest existing rendered Somalier workload,
`benchmark_somalier_site_extraction.md`, measures different inputs and kernels;
no comparable full-cohort selection baseline is available.

Reproduce locally from the repository root with installed `Rduckhts` and
`duckhtsbench`, using the registry-staged pinned executable:

```bash
/usr/bin/time -f '%M' -o benchmarks/somalier_find_sites_peak_kb.txt \
  Rscript test/scripts/somalier_find_sites_differential.R "$PWD"
cd benchmarks && Rscript -e "rmarkdown::render('benchmark_somalier_find_sites.Rmd')"
```
