---
title: "ROH from read counts: real-data validation on chr20"
output:
  github_document:
    html_preview: false
    pandoc_args: ["--wrap=preserve", "--columns=400"]
---

`duckhts_roh_counts` finds runs of homozygosity from allele read counts, with a
per-read error and a contamination fraction. Its SQL tests compare it with the
same model computed in SQL and with a synthetic titration. This report adds two
checks on real data, as section 2 of
<https://github.com/RGenomicsETL/duckhts/issues/329> asks:

- **(a) Agreement.** The read-count path against the VCF genotype path, on the
  same samples, sites and allele frequencies.
- **(b) Titration.** Real read counts of two samples mixed at a known fraction,
  decoded without and with the contamination term.

The scope is small on purpose: chr20, three samples, no timing. It makes no
performance claim.

## Inputs

All inputs are `r/duckhtsbench` registry artifacts.

- Read counts of NA18507, HG00403 and HG00188 at the 108,757 chr20 evaluation
  sites of the ROH ancestry evaluation: `roh_counts_chr20_na18507`,
  `roh_counts_chr20_hg00403` and `roh_counts_chr20_hg00188`. They come from the
  1000 Genomes 30× CRAMs (`benchmarks/roh_counts_stage.R`). Each row has the
  site's `af`, the source VCF's INFO/AF of the ALT allele, clamped to
  [0.001, 0.999].
- Genotypes of the same three samples at the same sites:
  `roh_counts_validation_chr20_genotypes`. It is derived from the phased chr20
  VCF of the 3,202 samples (`roh_ancestry_chr20_source`) and the site list
  (`roh_ancestry_chr20_af_sites`) by
  `benchmarks/roh_counts_validation/stage_genotypes.R`: three samples, FORMAT/GT
  only, exactly one record per site. Staging checks the SHA-256 of the cached
  source and the record digest of the output.

Before any decode, the run compares each cached input with the identity its
registry row records: the sample, row count and canonical counts digest of
each read-count file, and the sample order, record count and record digest of
the genotype BCF (`benchmarks/roh_counts_validation/inputs.R`). A file that
differs stops the render.

The read counts and the genotypes come from the same 30× sequencing of the
1000 Genomes samples. The VCF path is an independent decode path, with joint calling and phasing behind its
genotypes. It is not an independent measurement of the samples.

## Design, declared before any run was decoded

The values below are in `benchmarks/roh_counts_validation/declaration.R`. That
file and this section were committed before the first decode.

**Denominator.** FROH is the bases in runs divided by the span from the first
to the last evaluation site on chr20, inclusive. A run's bases are
`end - start + 1`.

**Run classes.** Every metric is given for all runs and for long runs, those of
at least 1,000,000 bases. Each decode is filtered by its own run lengths.

**Metrics**, per sample and run class, for a test decode against a reference:

- FROH of each decode and the difference, test minus reference;
- the base-level overlap as a Jaccard index: bases in the runs of both decodes
  divided by bases in the runs of either. It is undefined when neither decode
  has a run in the class.

**Tolerance.**

- The absolute FROH difference is at most 0.01.
- The Jaccard index is at least 0.9 for long runs and at least 0.7 for all runs.
- A class in which neither decode has a run agrees. A class in which only one
  decode has a run has a Jaccard index of 0 and fails.

Justification: both paths use the same sites, frequencies and hidden Markov
model, and differ only in the genotype evidence (a called genotype at phred 30,
or about 30 reads at the default `seq_error`). A long run rests on hundreds of
sites and should not depend on that difference. A short run rests on a few
sites, so its ends can move.

**(a) Agreement.** The reference is `duckhts_roh_af_table(..., gt_error := 30)`
on the genotype BCF, with the `af` of the counts as the frequency relation. The
test is `duckhts_roh_counts` on the counts with the default `seq_error` (0.001)
and `contamination := 0`. A sample passes a run class when it meets the
tolerance.

**(b) Titration.** For each ordered pair (receiver, contaminant) of the three
samples and each α in 1%, 2%, 5% and 10%, the mixed count of each allele at
each site is

    Binomial(receiver count, 1 − α) + Binomial(contaminant count, α).

This thins real counts and adds them. It is not a mixture of reads in a CRAM:
no read is realigned, filtered or counted again, and the two samples' counts
are drawn independently at each site. The mean depth stays close to that of
one sample.

Cell *k* of the grid, in the order receiver, contaminant, α, draws with
`set.seed(20261004 + k)` and R's default generators. The draws are, in order:
receiver REF, contaminant REF, receiver ALT, contaminant ALT.

Each mixture is decoded twice: with `contamination := 0` and with
`contamination := α`. The reference is the decode of the receiver's unmixed
counts from (a). The same metrics and tolerance apply. The declared expectation
is for the decode with `contamination := α`; the decode without the term is the
control that shows what the term changes.

The contaminant is one individual, and `duckhts_roh_counts` models a
contaminant read by the population frequency `af`. That approximation is part
of what (b) measures.

All observations are kept: every run of every decode is in
`benchmarks/results/roh-counts-validation/`.

``` sh
Rscript benchmarks/roh_counts_validation/stage.R
taskset -c 4-7 Rscript -e 'rmarkdown::render("benchmarks/benchmark_roh_counts_validation.Rmd")'
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r run, include=FALSE}
Sys.setenv(DUCKHTSBENCH_REGISTRY = normalizePath(
  "r/duckhtsbench/inst/benchmark_registry.tsv", mustWork = TRUE))
source("benchmarks/roh_counts_validation/run.R")
results <- roh_validation_run(
  extension = Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"),
  output_dir = "benchmarks/results/roh-counts-validation", threads = 2L)
declaration <- roh_validation_declaration
verdict <- function(pass) ifelse(pass, "pass", "fail")
class_label <- c(all = "all runs", long = "runs ≥ 1 Mb")
```

## Run

```{r metadata, echo=FALSE}
knitr::kable(results$metadata, col.names = c("Item", "Value"))
```

## (a) Read counts against VCF genotypes

`reference` is the VCF genotype path and `test` is the read-count path.

```{r agreement, echo=FALSE}
agreement <- results$agreement
knitr::kable(data.frame(
  Sample = agreement$sample,
  `Run class` = class_label[agreement$run_class],
  `VCF runs` = agreement$reference_runs,
  `Count runs` = agreement$test_runs,
  `VCF FROH` = sprintf("%.4f", agreement$reference_froh),
  `Count FROH` = sprintf("%.4f", agreement$test_froh),
  `FROH difference` = sprintf("%+.4f", agreement$froh_difference),
  `Shared bases` = format(agreement$shared_bases, big.mark = ","),
  `Bases in either` = format(agreement$either_bases, big.mark = ","),
  Jaccard = ifelse(is.na(agreement$jaccard), "none", sprintf("%.3f", agreement$jaccard)),
  FROH = verdict(agreement$froh_pass),
  Overlap = verdict(agreement$jaccard_pass),
  Verdict = verdict(agreement$pass),
  check.names = FALSE), align = "llrrrrrrrrlll")
```

`r sum(agreement$pass)` of `r nrow(agreement)` sample and run-class rows meet
the declared tolerance. "none" means that neither decode has a run in the
class.

## What (a) compares, written after its result

The declared comparison fails, and the reason is in what the two error
parameters mean. This section was written after the result above. It changes
no tolerance and no verdict.

**`gt_error` is an error of the site.** With genotype evidence, a called
genotype has likelihood close to 1 and a genotype one allele away has
10^(−`gt_error`/10). At `gt_error := 30`, one heterozygous call can count
against a run by a factor of 1,000 and no more, whatever caused it: a calling
error, a mapping artefact or a wrong site. The run can continue through it.

**`seq_error` is an error of one read.** With read counts, the reads of a site
are independent and their probabilities multiply. A site with 16 reads of each
allele has likelihood `seq_error`^16 under either homozygous genotype. No
value of `seq_error` makes that site compatible with a run, and nothing in the
read model says that a whole site can be wrong.

So the read-count decode is the genotype decode with no tolerated genotype
error. At `gt_error := 30` the genotype decode keeps runs through isolated
heterozygous sites, and the read-count decode ends them there. The two paths
then differ by design, not by a defect of either, and the declared tolerance
compared them at settings that do not correspond.

```{r diagnosis, include=FALSE}
source("benchmarks/roh_counts_validation/diagnose.R")
diagnosis <- roh_diagnosis_run(
  extension = Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"),
  output_dir = "benchmarks/results/roh-counts-validation", threads = 2L)
comparison_table <- function(rows, parameter) {
  knitr::kable(data.frame(
    parameter = format(rows[[parameter]], scientific = FALSE, drop0trailing = TRUE),
    Sample = rows$sample,
    `VCF FROH` = sprintf("%.4f", rows$reference_froh),
    `Count FROH` = sprintf("%.4f", rows$test_froh),
    `FROH difference` = sprintf("%+.4f", rows$froh_difference),
    Jaccard = sprintf("%.3f", rows$jaccard),
    check.names = FALSE), col.names = c(parameter, "Sample", "VCF FROH", "Count FROH",
                                         "FROH difference", "Jaccard"),
    align = "rlrrrr", row.names = FALSE)
}
```

**The genotype decode approaches the read-count decode as `gt_error` rises.**
The read-count decode is fixed at the default `seq_error`; all runs.

```{r diagnosis-gt-error, echo=FALSE}
comparison_table(diagnosis$by_gt_error, "gt_error")
```

**`seq_error` hardly moves the read-count decode.** The genotype decode is
fixed at `gt_error := 30`; all runs.

```{r diagnosis-seq-error, echo=FALSE}
comparison_table(diagnosis$by_seq_error, "seq_error")
```

**The genotypes and the read counts agree at the sites.** Sites of the three
samples, by called genotype and by the share of reads of the rarer allele:

```{r diagnosis-sites, echo=FALSE}
site_classes <- diagnosis$site_classes
knitr::kable(stats::reshape(site_classes, idvar = "read_class", timevar = "genotype",
                            direction = "wide"),
             row.names = FALSE, format.args = list(big.mark = ","))
```

**The lost parts of genotype runs hold the heterozygous calls.** Sites inside
the runs of the genotype decode at `gt_error := 30`, split by whether the
read-count decode keeps the site in a run:

```{r diagnosis-inside, echo=FALSE}
knitr::kable(diagnosis$inside_runs, row.names = FALSE, format.args = list(big.mark = ","))
```

The observations are in `diagnosis_*.csv` under
`benchmarks/results/roh-counts-validation/`.

What this leaves open:

- Which decode is closer to the truth is not measured here. A heterozygous
  call with balanced reads inside a long run can be a real heterozygous site
  that ends the run, or reads from a paralogous sequence at a site that is
  homozygous.
- The read model has no term for a site whose reads do not reflect the
  sample's genotype. Such a term would be to read counts what `gt_error` is to
  genotypes. It is not implemented.
- A like-for-like check of the read model is the genotype-likelihood path
  (FORMAT/PL) from the same reads. The staged genotypes carry FORMAT/GT only.

## (b) Count-level contamination titration

Each row summarises the six ordered pairs at one α and one decode. The FROH
difference and the Jaccard index are against the receiver's unmixed decode;
they are shown as the median and the range over the pairs.

```{r titration-summary, echo=FALSE}
titration <- results$titration
titration$contamination_term <- factor(titration$contamination_term, levels = c("none", "alpha"))
range_text <- function(values, pattern) {
  if (all(is.na(values))) return("none")
  sprintf(paste0(pattern, " (", pattern, " to ", pattern, ")"),
          stats::median(values, na.rm = TRUE), min(values, na.rm = TRUE),
          max(values, na.rm = TRUE))
}
groups <- split(titration, list(titration$contamination_term, titration$alpha,
                                titration$run_class), drop = TRUE)
summary_rows <- do.call(rbind, lapply(groups, function(group) {
  data.frame(
    `Run class` = class_label[[group$run_class[[1L]]]],
    alpha = sprintf("%.0f%%", 100 * group$alpha[[1L]]),
    `Contamination term` = if (group$contamination_term[[1L]] == "none") "0" else "α",
    `FROH difference` = range_text(group$froh_difference, "%+.4f"),
    Jaccard = range_text(group$jaccard, "%.3f"),
    `Pairs without runs` = sum(is.na(group$jaccard)),
    `Pairs passing` = sprintf("%d of %d", sum(group$pass), nrow(group)),
    check.names = FALSE)
}))
names(summary_rows)[[2L]] <- "α"
knitr::kable(summary_rows, row.names = FALSE, align = "lllllrl")
```

```{r titration-verdict, echo=FALSE}
with_term <- titration[titration$contamination_term == "alpha", ]
without_term <- titration[titration$contamination_term == "none", ]
```

With `contamination := α`, `r sum(with_term$pass)` of `r nrow(with_term)` pair,
α and run-class rows meet the declared tolerance. With `contamination := 0`,
`r sum(without_term$pass)` of `r nrow(without_term)` do.

The realised contaminant read fraction of the mixtures, over all sites, is
within `r sprintf("%.4f", max(abs(titration$realised_alpha - titration$alpha)))`
of α in every cell, and the mean depth of a mixture is between
`r sprintf("%.1f", min(titration$mean_depth))` and
`r sprintf("%.1f", max(titration$mean_depth))` reads per site.

### Every pair

```{r titration-pairs, echo=FALSE}
pairs <- titration[order(titration$run_class, titration$cell, titration$contamination_term), ]
knitr::kable(data.frame(
  `Run class` = class_label[pairs$run_class],
  Receiver = pairs$receiver,
  Contaminant = pairs$contaminant,
  alpha = sprintf("%.0f%%", 100 * pairs$alpha),
  `Contamination term` = ifelse(pairs$contamination_term == "none", "0", "α"),
  `Reference runs` = pairs$reference_runs,
  Runs = pairs$test_runs,
  `Reference FROH` = sprintf("%.4f", pairs$reference_froh),
  FROH = sprintf("%.4f", pairs$test_froh),
  `FROH difference` = sprintf("%+.4f", pairs$froh_difference),
  Jaccard = ifelse(is.na(pairs$jaccard), "none", sprintf("%.3f", pairs$jaccard)),
  Verdict = verdict(pairs$pass),
  check.names = FALSE), row.names = FALSE, align = "lllllrrrrrrl")
```

## Limitations

- One chromosome and three samples. The rows above are observations, not an
  estimate of a rate.
- The two paths of (a) share their sequencing data, sites and frequencies.
  They do not share an error model: see "What (a) compares". Their agreement
  at a large `gt_error` shows that the two paths decode the same evidence
  alike. It does not show that either set of runs is true.
- (b) thins and adds counts. It does not reproduce what a mixture of reads
  would change upstream: alignment, duplicate marking, base and mapping quality
  filters, or the correlation of reads within a fragment.
- The contamination fraction is supplied to the decode. Nothing here estimates
  it.
- The reference of (b) is the receiver's own decode, so (b) measures how much
  a mixture moves the runs, not whether the runs are true.
