---
title: "Ancestry relational projection: synthetic site scaling"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(DBI)
library(Rduckhts)
library(duckhtsbench)
repo <- normalizePath("..")
source_revision <- system2("git", c("-C", shQuote(repo), "rev-parse", "HEAD"), stdout = TRUE)
registry <- file.path(repo, "r/duckhtsbench/inst/benchmark_registry.tsv")
cache <- tempfile("duckhts-ancestry-benchmark-")
dir.create(cache)
Sys.setenv(DUCKHTSBENCH_REGISTRY = registry, DUCKHTS_CACHE_DIR = cache)
correction_file <- duckhts_bench_stage_repository_fixtures(repo, "ancestry-projection")[["ancestry_correction"]]
con <- rduckhts_connect()
dbExecute(con, "SET threads=4")
dbExecute(con, paste(
  "CREATE TABLE bench_ancestry_ref AS SELECT 'chr1' AS chromosome,",
  "(100000+i)::INTEGER AS position, 'A' AS allele_a, 'G' AS allele_b,",
  "g.group_id, CASE WHEN g.group_id='A' THEN 0.25 + 0.16*sin(i*0.017)",
  "ELSE 0.68 + 0.12*cos(i*0.019) END AS frequency",
  "FROM range(17000) t(i), (VALUES ('A'),('B')) g(group_id)"
))
dbExecute(con, paste(
  "CREATE TABLE bench_ancestry_pc AS SELECT DISTINCT chromosome, position, allele_a, allele_b,",
  "pc, CASE WHEN pc=1 THEN sin(position*0.031) ELSE cos(position*0.027) END AS loading",
  "FROM bench_ancestry_ref, (VALUES (1),(2)) t(pc)"
))
dbExecute(con, sprintf(
  "CREATE TABLE bench_ancestry_correction AS SELECT pc::INTEGER AS pc, coefficient::DOUBLE AS coefficient FROM read_csv_auto(%s, delim='\\t', header=true) WHERE pc<=2",
  as.character(dbQuoteString(con, correction_file))
))
dbExecute(con, paste(
  "CREATE TABLE bench_ancestry_input AS SELECT sample_id, chromosome, position, allele_a, allele_b,",
  "sum(CASE WHEN group_id='A' THEN a_share ELSE 1-a_share END * frequency)",
  "+ 0.002*sin(position * sample_number) AS frequency",
  "FROM bench_ancestry_ref, (VALUES ('S1',0.7,1),('S2',0.4,2),('S3',0.1,3))",
  "s(sample_id,a_share,sample_number)",
  "GROUP BY sample_id, sample_number, chromosome, position, allele_a, allele_b"
))
peak_rss_mb <- function() {
  status <- grep("^VmHWM:", readLines("/proc/self/status"), value = TRUE)
  as.numeric(strsplit(trimws(status), "[[:space:]]+")[[1L]][[2L]]) / 1024
}
baseline_rss_mb <- peak_rss_mb()
run_sites <- function(n, repeat_id) {
  dbExecute(con, sprintf("CREATE OR REPLACE TEMP VIEW bench_ancestry_slice AS SELECT * FROM bench_ancestry_input WHERE position < %d", 100000 + n))
  start <- proc.time()[["elapsed"]]
  out <- rduckhts_ancestry_proportions(
    con, "bench_ancestry_slice", "bench_ancestry_ref", "bench_ancestry_pc",
    "bench_ancestry_correction", min_cor = 0.4
  )
  elapsed <- proc.time()[["elapsed"]] - start
  stopifnot(nrow(out) == 6L, all(out$used_variants == n))
  data.frame(samples = 3L, variants_per_sample = n, input_rows = 3L*n,
             output_group_rows = nrow(out), groups = 2L, pcs = 2L,
             threads = 4L, repeat_id = repeat_id, elapsed_seconds = elapsed,
             process_peak_rss_mb = peak_rss_mb(),
             failures = paste(unique(out$status[out$status != "ok"]), collapse = ";"))
}
invisible(run_sites(1000L, 0L))
results <- do.call(rbind, lapply(c(1000L, 5000L, 17000L), function(n) {
  do.call(rbind, lapply(seq_len(3L), function(rep) run_sites(n, rep)))
}))
```

Revision: `r source_revision`. Workload: deterministic biallelic synthetic
reference frequencies and PC loadings; three independently perturbed samples;
4 DuckDB threads; two reference groups; two PCs; `min_cor = 0.4`. Each variant
is present in both groups and PCs, and every input row matches. This report
measures query and materialization together, not the 850 MB Figshare products.

The correction vector is staged from the committed, checksummed bigsnpr vignette
product. `VmHWM` is cumulative process high-water RSS (MiB), including package
loading and all preceding cases; the baseline before the first run was
`r round(baseline_rss_mb, 1)` MiB. It is not per-query allocated memory.

```{r results}
knitr::kable(results, digits = 3)
```

Failed gates are retained in the `failures` column for each repeat. There is no
identical previously rendered ancestry workload for a longitudinal comparison.

```{r cleanup, include=FALSE}
dbDisconnect(con, shutdown = TRUE)
unlink(cache, recursive = TRUE)
```
