---
title: "Indexed BCF/VCF scan initialization: small-fixture cost"
output:
  github_document:
    html_preview: false
---

This paired measurement exercises automatic full-file scan planning, complete and
projected materialization, metadata counts and overlapping indexed regions on
committed BCF/VCF fixtures staged through duckhtsbench's registry.
The registry pins encoded file bytes; staging is a network-free verified copy,
not fixture reconstruction. The original two-record inputs have
one occupied contig and 256 empty header contigs. A second five-record corpus has
two occupied contigs, two samples, equal-position distinct alleles, and a physical
duplicate; its VCF header declares all, some, or none of the contigs. It uses
non-colocated indexes. Timings measure setup and small-result materialization,
not cohort throughput.

Both builds plan VCF work using the Tabix dictionary. The one-occupied-contig VCF
therefore selects a single full-file stream instead of 257 indexed contig tasks;
the five-record VCFs use two contig tasks irrespective of header completeness.
BCF retains its header dictionary. The requested DuckDB thread limit is not a
promise about the number of active workers. Both builds retain parsed indexes at
bind. The candidate shares its file/header/iterator and decode-validation code
with a host-neutral scanner. Data and indexes remain unchanged during timing.

The nearest larger full-column report is
[`benchmark_reader_allocations.md`](benchmark_reader_allocations.md):
4,048,342 GIAB HG002 records in wide/tidy form, one sequential scan handle and
zero HTSlib decompression workers. It does not exercise automatic parallel
index reloads and is not an identical-workload baseline. The previous identical
two/five-record full-materialization workload is retained at
[`4929162:benchmark_bcf_scan_init.md`](https://github.com/RGenomicsETL/duckhts/blob/4929162aa103fe703c848a9b87bfc062cd53aa15/benchmarks/benchmark_bcf_scan_init.md).
Count-only, projected and explicit-region cases below are new paired workloads;
they do not have a previous identical recorded baseline.
This report measures the pre-fix binary alongside the candidate on the same host.

## Reproduction and identity

Install duckhtsbench from this checkout. Set `BCF_SCAN_BASELINE_EXTENSION` and `BCF_SCAN_BASELINE_REVISION` to a separately
built baseline. `DUCKHTS_EXTENSION` selects the candidate. From the repository root:

```sh
taskset -c 0,2,4,6 Rscript -e 'rmarkdown::render("benchmarks/benchmark_bcf_scan_init.Rmd")'
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r measurement}
baseline <- Sys.getenv("BCF_SCAN_BASELINE_EXTENSION")
baseline_revision <- Sys.getenv("BCF_SCAN_BASELINE_REVISION")
candidate <- Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension")
stopifnot(nzchar(baseline_revision), file.exists(baseline), file.exists(candidate))
stopifnot(system2("git", c("diff", "--quiet", "HEAD", "--", "src")) == 0L)
identity <- data.frame(
  property = c("baseline revision", "candidate revision", "candidate src tree",
               "baseline binary MD5", "candidate binary MD5", "R", "DuckDB R package",
               "HTSlib", "CPU affinity", "repetitions"),
  value = c(baseline_revision, system2("git", c("rev-parse", "HEAD"), stdout = TRUE),
            system2("git", c("rev-parse", "HEAD:src"), stdout = TRUE),
            unname(tools::md5sum(baseline)), unname(tools::md5sum(candidate)),
            R.version.string, as.character(packageVersion("duckdb")),
            readLines("third_party/htslib/VERSION", n = 1),
            system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE),
            "9 timed calls after warm-up; fresh R process per build/input/thread/workload setting")
)
knitr::kable(identity)
paths <- duckhtsbench::duckhts_bench_stage_bcf_scan(".")
artifact <- function(ids) unname(paths[paste0("bcf_scan_", ids)])
old_inputs <- artifact(c("old_bcf", "old_vcfgz"))
vcf_inputs <- artifact(c("full", "partial", "none"))
cases <- data.frame(
  input = c(old_inputs, vcf_inputs),
  index = artifact(c("old_csi", "old_tbi", "full_tbi", "partial_tbi", "none_tbi")),
  explicit_index = c(FALSE, FALSE, TRUE, TRUE, TRUE),
  oracle = c(rep(old_inputs[1], 2), rep(artifact("oracle"), 3)),
  format = c("BCF/CSI", "VCF.gz/TBI", rep("VCF.gz/TBI", 3)),
  header = c("257 contigs", "257 contigs", "full", "partial", "none"),
  header_contigs = c(257L, 257L, 3L, 2L, 0L),
  input_records = c(2L, 2L, 5L, 5L, 5L)
)
files <- unique(c(cases$input, cases$index, cases$oracle))
stopifnot(all(file.exists(files)),
          sum(grepl("^##contig=", readLines(artifact("old_vcf")))) == 257L)
knitr::kable(data.frame(file = basename(files), bytes = file.info(files)$size,
                        md5 = unname(tools::md5sum(files))))
results <- list()
oracles <- list()
for (i in seq_len(nrow(cases))) for (threads in c(1L, 4L))
  for (workload in c("full", "count", "projected", "regions")) for (build in c("baseline", "candidate")) {
  case <- cases[i, ]
  measured <- callr::r(function(extension, input, index, explicit_index, oracle, threads, input_records, workload) {
    con <- DBI::dbConnect(duckdb::duckdb(config = list(allow_unsigned_extensions = "true")))
    on.exit(DBI::dbDisconnect(con, shutdown = TRUE))
    DBI::dbExecute(con, paste("LOAD", DBI::dbQuoteString(con, extension)))
    DBI::dbExecute(con, sprintf("SET threads=%d", threads))
    reference <- DBI::dbGetQuery(con, sprintf(
      "SELECT * FROM read_bcf(%s,scan_mode:='sequential',decompression_threads:=0) ORDER BY ALL",
      DBI::dbQuoteString(con, oracle)))
    stopifnot(nrow(reference) == input_records)
    physical_reference <- reference
    selected_records <- input_records
    index_option <- if (explicit_index) paste0(",index_path:=", DBI::dbQuoteString(con, index)) else ""
    projection <- "*"
    region_option <- ""
    if (workload == "count") {
      projection <- "count(*) AS n"
      reference <- DBI::dbGetQuery(con, sprintf("SELECT %d::BIGINT AS n", input_records))
    } else if (workload == "projected") {
      projection <- "CHROM,POS,REF,ALT"
      reference <- reference[c("CHROM", "POS", "REF", "ALT")]
    } else if (workload == "regions") {
      region_option <- ",region:='chr1:1-20,chr1:10-30,chr1:1-20'"
      reference <- reference[reference$CHROM == "chr1" & reference$POS <= 30, ]
      rownames(reference) <- NULL
      selected_records <- nrow(reference)
      stopifnot(selected_records == 1L)
    }
    sql <- sprintf("CREATE OR REPLACE TEMP TABLE scanned AS SELECT %s FROM read_bcf(%s%s%s,decompression_threads:=0)",
                   projection, DBI::dbQuoteString(con, input), index_option, region_option)
    DBI::dbExecute(con, sql)
    expected <- DBI::dbGetQuery(con, "SELECT * FROM scanned ORDER BY ALL")
    exact <- identical(expected, reference)
    if (!exact) stop(basename(input), "/", workload, ": ",
                    paste(all.equal(expected, reference), collapse = "; "))
    timings <- numeric(9)
    for (i in seq_along(timings)) {
      start <- Sys.time()
      DBI::dbExecute(con, sql)
      timings[i] <- as.numeric(difftime(Sys.time(), start, units = "secs"))
      stopifnot(identical(DBI::dbGetQuery(con, "SELECT * FROM scanned ORDER BY ALL"), expected))
    }
    list(rows = expected, reference = physical_reference, selected_records = selected_records,
         exact = exact, timings = timings)
  }, args = list(extension = normalizePath(if (build == "baseline") baseline else candidate),
                 input = normalizePath(case$input), index = normalizePath(case$index),
                 explicit_index = case$explicit_index,
                 oracle = normalizePath(case$oracle), threads = threads,
                 input_records = case$input_records, workload = workload))
  if (is.null(oracles[[case$oracle]])) oracles[[case$oracle]] <- measured$reference
  stopifnot(identical(oracles[[case$oracle]], measured$reference))
  stopifnot(measured$exact) # Never compare an incomplete materialization.
  results[[length(results) + 1L]] <- data.frame(
    build, workload, format = case$format, header = case$header,
    duckdb_threads = threads, htslib_workers_per_handle = 0L,
    header_contigs = case$header_contigs, input_records = case$input_records,
    selected_records = measured$selected_records,
    materialized_rows = nrow(measured$rows), materialized_columns = ncol(measured$rows),
    exact_rows = measured$exact,
    median_ms = if (measured$exact) 1000 * median(measured$timings) else NA_real_,
    min_ms = if (measured$exact) 1000 * min(measured$timings) else NA_real_,
    max_ms = if (measured$exact) 1000 * max(measured$timings) else NA_real_
  )
}
knitr::kable(do.call(rbind, results), digits = 3)
```

Each timed `CREATE OR REPLACE TEMP TABLE AS SELECT` materializes the requested
result. Full/projected cases return two or five rows, regions return one selected
record, and count returns one aggregate row representing two or five records.
Count may read only index statistics; it is not a physical decode throughput test.
Elapsed `Sys.time()` includes the DBI call;
connection creation, warm-up, and complete typed-row verification are outside the
timer. Every candidate result and every valid baseline agrees exactly with the
complete BCF sequential rows, including GT, ALT and physical duplicate multiplicity,
without asserting scan arrival order. The BCF oracle also agrees across builds.

There is no public sample-selection option on canonical `read_bcf`; retired
experimental selectors are not reinstated to produce a timing. Native selected
sample tests are correctness evidence, not a selected-cohort benchmark.

This does not measure the latency of an index failure, remote I/O, large-cohort
throughput, RSS, or maximum worker concurrency. Thread settings are DuckDB's
requested limits. Small timings can vary with scheduling; no general speedup or
no-regression claim follows from them.
