---
title: "Somalier X/Y sex evidence at scale"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
root <- normalizePath("..", mustWork = TRUE)
runs <- utils::read.delim(file.path(root, "benchmarks", "somalier_sex_scaling.tsv"),
                          colClasses = c(git_sha = "character", extension_sha256 = "character"))
mib <- function(bytes) bytes / 1048576
# Budget, declared before measuring. The peak-RSS ceiling of a cell is the fixed
# overhead (the median VmHWM of the empty-run processes at the same thread count:
# packages, connection and extension load, attach, no query) plus three times the
# live state the function needs plus the DuckDB buffer pool, which is capped by
# memory_limit and holds clean, evictable pages of the scanned count relation.
# Live state is one aggregate group per sample and the panel-sized build side:
#   per sample: group key 16 B, 12 aggregate numbers 96 B, hash-table slot and
#               coverage sums 88 B                              = 200 B per sample;
#   panel:      20,000 rows x (assembly 16 B, ordinal 8 B, region 16 B, position 8 B,
#               alleles 32 B, kind and hashes 20 B)             = 100 B per site;
#   per thread: one 2,048-row vector chunk of the eight scanned columns and the
#               join probe state, at most 1 MiB.
# Spill must be zero. DuckDB's peak buffer memory is reported but not gated: the
# first measured matrix showed the profile metric reaching about twice memory_limit
# for a file-backed relation, so a gate of "buffer <= memory_limit" declared for that
# run was withdrawn and the process-level peak RSS ceiling above is the gate.
memory_limit_mib <- as.numeric(sub("MB$", "", runs$memory_limit[1]))
panel_sites <- 20000
state_bytes <- function(samples, threads) samples * 200 + panel_sites * 100 + threads * 1048576
overhead <- subset(runs, mode == "overhead")
overhead_mib <- tapply(overhead$peak_rss_kib / 1024, overhead$threads, stats::median)
runs <- subset(runs, mode == "run")
runs$peak_rss_mib <- runs$peak_rss_kib / 1024
runs$peak_buffer_mib <- mib(runs$peak_buffer_bytes)
runs$spill_mib <- mib(runs$peak_temp_bytes)
runs$budget_mib <- overhead_mib[as.character(runs$threads)] +
  3 * mib(state_bytes(runs$samples, runs$threads)) + memory_limit_mib
keys <- c("threads", "samples")
summary <- do.call(rbind, lapply(split(runs, runs[keys], drop = TRUE), function(d) {
  data.frame(d[1, keys], input_rows = format(d$input_rows[1], big.mark = ",", scientific = FALSE),
             output_rows = d$output_rows[1],
             median_s = stats::median(d$seconds), min_s = min(d$seconds), max_s = max(d$seconds),
             peak_rss_mib = round(max(d$peak_rss_mib)),
             peak_buffer_mib = round(max(d$peak_buffer_mib)),
             spill_mib = max(d$spill_mib), budget_mib = round(d$budget_mib[1]),
             overhead_mib = round(overhead_mib[as.character(d$threads[1])]))
}))
rownames(summary) <- NULL
one_thread <- subset(summary, threads == 1L)
exponent <- function(d) unname(stats::coef(stats::lm(log(d$median_s) ~ log(d$samples)))[2])
exponents <- do.call(rbind, lapply(split(summary, summary$threads), function(d)
  data.frame(threads = d$threads[1], time_exponent = round(exponent(d), 2))))
rownames(exponents) <- NULL
stopifnot(
  nrow(overhead) == 6L, all(table(runs$samples, runs$threads) == 3L),
  length(unique(runs$git_sha)) == 1L, length(unique(runs$extension_sha256)) == 1L,
  all(runs$input_rows == runs$samples * panel_sites), all(runs$output_rows == runs$samples),
  all(runs$peak_rss_mib <= runs$budget_mib), all(runs$spill_mib == 0),
  all(exponents$time_exponent <= 1.25))
```

`duckhts_somalier_sex()` reduces a sample-by-panel count relation (one row per sample
and panel site) to one row per sample: the mean autosomal depth over sites that pass
the relatedness gate, the X and Y depth over usable sites, and the X genotype counts.
It is a single scan and hash aggregate with one group per sample, so its live state is
samples × a few numbers plus the panel-sized join side; the count relation is
streamed.

## Inputs

No registered public cohort has X/Y sites in a Somalier panel, so this workload is
synthetic and generated by `benchmark_somalier_sex_run.R --data`: a 20,000-site panel
(17,000 autosomal on 22 contigs, 2,500 chrX, 500 chrY) and counts for 500, 1,000 and
2,000 samples (1×, 2×, 4×; 10, 20 and 40 million rows), with depths 8–57 and a mix of
homozygous and heterozygous balances. Growth is along samples only; the panel is fixed.
Because the data are synthetic the report certifies memory and scaling, not the
biology; agreement with Somalier v0.3.4 is tested separately by
`test/scripts/somalier_sex_differential.R`.

## Method

`benchmark_somalier_sex_run.R` runs each cell as fresh processes, three repetitions
interleaved across cells, at one and four threads, each with
`memory_limit = '`r runs$memory_limit[1]`'` and an empty temporary directory. Each
worker attaches the count database read-only, times the public SQL macro including
fetching the result into R (which is consumed: every sample must have a call and a
status), and records DuckDB's peak buffer memory and temporary-directory size from the
query profile and the process's peak RSS (VmHWM). The database file is read from a warm
page cache. Every row records the git revision of the tree and the SHA-256 of the
bundled extension the process loaded; the report requires one of each.

```{r identity}
knitr::kable(data.frame(git_sha = runs$git_sha[1], extension_sha256 = runs$extension_sha256[1]))
```

## Results

```{r results}
knitr::kable(summary, digits = 2)
```

`overhead_mib` is the median peak RSS of the empty-run processes and `budget_mib` is
the ceiling declared in the report source: overhead + 3 × live state +
`memory_limit`. Every cell stays within it and no cell spills. DuckDB's peak buffer
memory, the maximum of the query profile, is reported without a gate: it reaches about
twice `memory_limit` for the attached file-backed relation, and the first measured
matrix led to withdrawing a `buffer <= memory_limit` gate that had been declared for it.

```{r exponents}
knitr::kable(exponents)
```

The 1× single-thread median is `r round(one_thread$median_s[one_thread$samples == 500], 2)` s,
under the 5 s floor of `STYLE.md`, so this report gives no timing verdict at that
scale; time is shown at 2× and 4× and the fitted exponent above (expected linear work,
limit 1.25) is a scaling check, not a claim about a production cohort. Memory does not
depend on that floor.

## Limits

- Synthetic counts, no registered public X/Y cohort.
- Panel size is fixed at 20,000 sites; only samples grow.
- The buffer-pool allowance in the ceiling covers evictable pages of the scanned
  count relation; it makes the ceiling loose relative to the live state. What the
  gate certifies is that the process stays within overhead + live state + that bound
  with no spill, not that the live state alone is that small.
