---
title: "DuckHTS vs GFFBase GFF3 Conformance Benchmark"
output:
  github_document:
    html_preview: true
---

<!-- benchmark_gffbase_conformance.md is generated from benchmark_gffbase_conformance.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This benchmark compares DuckHTS `read_gff(...)` with
[GFFBase](https://github.com/Kuanhao-Chao/gffbase) across parser,
conformance, scan and feature-database workloads:

1. **GFFBase audit:** record whether GFFBase is using its Rust parser or Python
   fallback, check Rust/Python parser parity when both are available, and verify
   basic GFF3/GTF dialect detection. We do not treat marketing claims as proof.
2. **GFF3 conformance:** compare DuckHTS default permissive `read_gff(...)` and
   `read_gff(..., strict := true)` against observed GFFBase strict-parser
   behavior. The cases are adapted from GFFBase's NCBI/GFF3 compliance tests at
   commit `78714cf30a9d799eab544e00a79a4da9754987ca`.
3. **Direct scan/parser throughput:** DuckHTS scans the same GFF3 files through
   `read_gff(...)`; GFFBase parses them with `parse_gff(strict = TRUE)`.
4. **Feature-database core:** SQL over `read_gff(...)` materializes features,
   parent edges and descendants for comparison with `gffbase.create_db`.

DuckHTS supports GFF3 through `read_gff(...)`, GTF through `read_gtf(...)`, and
plain tabix-style GFF/GTF-like rows through `read_tabix(...)`. The new
`strict := true` option is specifically GFF3 structural validation for
`read_gff(...)`; it is not a full feature-database hierarchy validator.

# Run

Stage GFFBase into the DuckHTS cache, then render this report:

```sh
make stage-gffbase
make bench-gffbase
```

Override defaults with `GFFBASE_BENCH_ROWS`, `GFFBASE_BENCH_PASSES`,
`GFFBASE_BENCH_FORCE=1`, `GFFBASE_BENCH_INCLUDE_CREATE_DB=1`, or
`GFFBASE_HUMAN_GFF=/path/to/gencode.gff3.gz[,/path/to/refseq.gff3.gz]` for
real human-scale files.

```{r setup, include = FALSE}
library(knitr)
library(jsonlite)

repo_root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", unset = getwd()), mustWork = TRUE)
if (!nzchar(Sys.getenv("DUCKHTSBENCH_REGISTRY", unset = ""))) {
  Sys.setenv(DUCKHTSBENCH_REGISTRY = normalizePath("r/duckhtsbench/inst/benchmark_registry.tsv", mustWork = TRUE))
}
source(normalizePath("r/duckhtsbench/R/registry.R", mustWork = TRUE))
duckhts_cache_dir <- duckhts_bench_cache_dir()
gffbase_site <- Sys.getenv("GFFBASE_SITE", unset = duckhts_bench_artifact_path("gffbase_010"))
if (!dir.exists(gffbase_site)) {
  stop("GFFBase is absent from the DuckHTS cache. Run `make stage-gffbase`.", call. = FALSE)
}
old_pythonpath <- Sys.getenv("PYTHONPATH", unset = "")
Sys.setenv(PYTHONPATH = paste(c(gffbase_site, old_pythonpath[nzchar(old_pythonpath)]), collapse = .Platform$path.sep))
out_dir <- Sys.getenv("GFFBASE_BENCH_OUT_DIR", unset = file.path(duckhts_cache_dir, "benchmarks", "gffbase", "results"))
audit_csv <- file.path(out_dir, "gffbase_audit.csv")
conformance_csv <- file.path(out_dir, "conformance.csv")
performance_csv <- file.path(out_dir, "performance.csv")
metadata_json <- file.path(out_dir, "metadata.json")

rows <- as.integer(Sys.getenv("GFFBASE_BENCH_ROWS", unset = "200000"))
passes <- as.integer(Sys.getenv("GFFBASE_BENCH_PASSES", unset = "3"))
warmup <- as.integer(Sys.getenv("GFFBASE_BENCH_WARMUP", unset = "1"))
force <- identical(Sys.getenv("GFFBASE_BENCH_FORCE", unset = "0"), "1")
include_create_db <- identical(Sys.getenv("GFFBASE_BENCH_INCLUDE_CREATE_DB", unset = "0"), "1")
human_gff <- strsplit(Sys.getenv("GFFBASE_HUMAN_GFF", unset = ""), ",", fixed = TRUE)[[1]]
human_gff <- human_gff[nzchar(human_gff)]
human_gff <- normalizePath(human_gff, mustWork = FALSE)
extension_path <- normalizePath(file.path(repo_root, "build", "release", "duckhts.duckdb_extension"), mustWork = TRUE)
current_identity <- list(
  duckhts_extension = extension_path,
  duckhts_extension_md5 = unname(tools::md5sum(extension_path)),
  gffbase_site = normalizePath(gffbase_site, mustWork = TRUE),
  rows = rows, passes = passes, warmup = warmup,
  include_create_db = include_create_db, human_gff = human_gff
)

needs_run <- force || !file.exists(audit_csv) || !file.exists(conformance_csv) || !file.exists(performance_csv) || !file.exists(metadata_json)
if (!needs_run && file.exists(metadata_json)) {
  old_meta <- tryCatch(jsonlite::fromJSON(metadata_json), error = function(error) list())
  identity_matches <- identical(as.character(old_meta$duckhts_extension), current_identity$duckhts_extension) &&
    identical(as.character(old_meta$duckhts_extension_md5), current_identity$duckhts_extension_md5) &&
    startsWith(as.character(old_meta$gffbase_site), current_identity$gffbase_site) &&
    identical(as.integer(old_meta$rows), current_identity$rows) &&
    identical(as.integer(old_meta$passes), current_identity$passes) &&
    identical(as.integer(old_meta$warmup), current_identity$warmup) &&
    identical(as.logical(old_meta$include_create_db), current_identity$include_create_db) &&
    identical(as.character(old_meta$human_gff), current_identity$human_gff)
  needs_run <- !identity_matches
}

if (needs_run) {
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  cmd <- c(
    file.path(repo_root, "scripts", "gffbase_conformance_benchmark.py"),
    "--extension", extension_path,
    "--rows", as.character(rows),
    "--passes", as.character(passes),
    "--warmup", as.character(warmup),
    "--out-dir", out_dir
  )
  for (p in human_gff) cmd <- c(cmd, "--human-gff", p)
  if (include_create_db) cmd <- c(cmd, "--include-create-db")

  pythonpath <- Sys.getenv("PYTHONPATH", unset = "")
  out <- system2("python3", cmd, stdout = TRUE, stderr = TRUE, env = c(sprintf("PYTHONPATH=%s", pythonpath)))
  status <- attr(out, "status")
  if (!is.null(status) && status != 0) {
    stop(paste(out, collapse = "\n"))
  }
}

audit <- utils::read.csv(audit_csv, check.names = FALSE, stringsAsFactors = FALSE)
conf <- utils::read.csv(conformance_csv, check.names = FALSE, stringsAsFactors = FALSE)
conf$gffbase_expectation_ok <- tolower(as.character(conf$gffbase_expectation_ok)) == "true"
perf <- utils::read.csv(performance_csv, check.names = FALSE, stringsAsFactors = FALSE)
meta <- jsonlite::fromJSON(metadata_json)
```

# Configuration

```{r configuration, echo=FALSE}
server_specs <- meta$server_specs
spec <- function(name, default = "") {
  if (is.null(server_specs) || is.null(server_specs[[name]])) default else as.character(server_specs[[name]])
}
duckhts_rev <- substr(meta$duckhts_git_rev, 1, 12)
if (!is.null(meta$duckhts_git_dirty) && isTRUE(meta$duckhts_git_dirty)) {
  duckhts_rev <- paste0(duckhts_rev, "+dirty")
}
knitr::kable(
  data.frame(
    parameter = c(
      "DuckHTS git rev",
      "DuckHTS extension",
      "GFFBase version",
      "GFFBase native parser",
      "GFFBase upstream commit",
      "DuckDB Python",
      "server hostname",
      "server OS",
      "kernel / machine",
      "CPU model",
      "CPU logical / affinity",
      "memory",
      "DuckDB threads",
      "synthetic rows",
      "timed passes"
    ),
    value = c(
      duckhts_rev,
      file.path("build", "release", "duckhts.duckdb_extension"),
      meta$gffbase_version,
      as.character(meta$gffbase_native_available),
      substr(meta$gffbase_upstream_commit, 1, 12),
      meta$duckdb_python_version,
      spec("hostname"),
      spec("os_pretty_name", spec("platform")),
      paste(spec("release"), spec("machine"), sep = " / "),
      spec("cpu_model"),
      paste(spec("cpu_logical"), spec("cpu_affinity"), sep = " / "),
      paste0(spec("mem_total_gib"), " GiB"),
      meta$duckdb_threads,
      format(meta$rows, big.mark = ","),
      meta$passes
    )
  )
)
```

# GFFBase audit

```{r audit, echo=FALSE}
knitr::kable(audit, row.names = FALSE)
```

# GFF3 conformance summary

```{r conformance-summary, echo=FALSE}
summary_tbl <- data.frame(
  metric = c(
    "cases",
    "GFFBase matched expected strict behavior",
    "DuckHTS default matched GFFBase strict behavior",
    "DuckHTS default gaps vs GFFBase strict behavior",
    "DuckHTS strict matched GFFBase strict behavior",
    "DuckHTS strict gaps vs GFFBase strict behavior"
  ),
  value = c(
    nrow(conf),
    sum(conf$gffbase_expectation_ok),
    sum(conf$duckhts_default_vs_gffbase == "match"),
    sum(conf$duckhts_default_vs_gffbase == "gap"),
    sum(conf$duckhts_strict_vs_gffbase == "match"),
    sum(conf$duckhts_strict_vs_gffbase == "gap")
  )
)
knitr::kable(summary_tbl)
```

## Remaining DuckHTS strict gaps

```{r conformance-gaps, echo=FALSE}
gaps <- conf[conf$duckhts_strict_vs_gffbase == "gap", c(
  "case", "focus", "expected_gffbase_kind", "duckhts_strict_detail"
)]
if (nrow(gaps) == 0) {
  gaps <- data.frame(note = "No remaining gaps in the local strict-conformance cases when DuckHTS uses attributes_list / attributes_pairs.")
}
knitr::kable(gaps, row.names = FALSE)
```

# Direct scan/parser throughput

```{r performance, echo=FALSE}
perf_show <- perf
num_cols <- c("min_sec", "median_sec", "mean_sec", "max_sec", "rows_per_sec", "mb_per_sec")
for (col in intersect(num_cols, names(perf_show))) {
  perf_show[[col]] <- round(perf_show[[col]], if (col %in% c("rows_per_sec", "mb_per_sec")) 1 else 4)
}
perf_show$vs_gffbase_parse <- NA_real_
if ("dataset" %in% names(perf)) {
  for (ds in unique(perf$dataset)) {
    idx <- perf$dataset == ds
    base <- perf$median_sec[idx & perf$tool == "GFFBase" & grepl("parse_gff", perf$variant)][1]
    perf_show$vs_gffbase_parse[idx] <- round(base / perf$median_sec[idx], 2)
  }
} else {
  base <- perf$median_sec[perf$tool == "GFFBase" & grepl("parse_gff", perf$variant)][1]
  perf_show$vs_gffbase_parse <- round(base / perf$median_sec, 2)
}
perf_show$rows <- format(perf_show$rows, big.mark = ",", scientific = FALSE)
show_cols <- c("dataset", "tool", "variant", "rows", "passes", "median_sec", "rows_per_sec", "mb_per_sec", "vs_gffbase_parse")
show_cols <- intersect(show_cols, names(perf_show))
knitr::kable(perf_show[, show_cols], row.names = FALSE)
```

# Interpretation

- GFFBase should be read as a comparison implementation, not an unquestioned
  oracle. The audit table exposes whether `auto` used Rust or Python and whether
  the two parsers agree on the local strict cases.
- DuckHTS default reading remains permissive for existing ingestion workflows.
  `strict := true` now rejects the structural GFF3 failures covered here.
- `attributes_list := true` and `attributes_pairs := true` close the local
  strict-conformance attribute gaps by representing repeated/comma-split GFF3
  values and URL-decoded values. The older `attributes_map := true` remains a
  backward-compatible scalar convenience column and intentionally cannot express
  duplicate keys or multi-valued attributes losslessly.
- Requesting `attributes_map`, `attributes_list`, and `attributes_pairs` together
  materializes three nested representations independently. That is useful as a
  stress test and compatibility escape hatch, but callers should normally choose
  the one representation their query needs.
- The timing table compares direct scans/parsing, not GFFBase's full persistent
  feature database API. Set `GFFBASE_HUMAN_GFF` to real GENCODE/RefSeq/MANE files
  for human-scale rows, and enable `GFFBASE_BENCH_INCLUDE_CREATE_DB=1` when the
  workload of interest is database materialization plus hierarchy indexing.

# Exercise the thesis: a feature database from `read_gff`

DuckHTS observes GFF3 records; SQL builds the persistent, position-sorted
feature table (including ordered, repeated attributes), parent edges and
transitive closure. This section uses the registered MANE and GENCODE inputs
and GFFBase v0.2.1; the conformance and parser sections above use v0.1.0.

```{r featuredb-sql, echo=FALSE, results='asis'}
cat("```sql\n", paste(readLines(file.path(repo_root, "scripts/gffbase_featuredb.sql")),
                      collapse = "\n"), "\n```\n", sep = "")
```

`gffbase.create_db` uses `force=True`, `merge_strategy="create_unique"`,
`force_gff=True`, `pragmas={"threads": threads}` and `validation=None`.
Validation modes are not run. SQL assigns suffixes to repeated physical IDs
in source order to match `create_unique`; both sides retain every input row.

```{r featuredb-data, include=FALSE}
featuredb_dir <- file.path(repo_root, "benchmarks/data")
parity <- read.csv(file.path(featuredb_dir, "gffbase_featuredb_parity.csv"),
                   stringsAsFactors = FALSE)
times <- read.csv(file.path(featuredb_dir, "gffbase_featuredb_timings.csv"),
                  stringsAsFactors = FALSE)
stopifnot(nrow(parity) > 0, all(parity$parity == "PASS"),
          nrow(times) == 2 * nrow(parity),
          !anyDuplicated(parity[c("dataset", "pass_number")]),
          !anyDuplicated(times[c("dataset", "pass_number", "engine")]))
stopifnot(setequal(unique(times$engine), c("DuckHTS", "gffbase")))
joined <- merge(times, parity, by = c("dataset", "pass_number"),
                suffixes = c("_measured", "_parity"))
stopifnot(nrow(joined) == nrow(times))
for (column in c("features", "anchors", "windows", "descendants",
                 "descendants_sha256", "region_hits", "region_sha256")) {
  stopifnot(all(joined[[paste0(column, "_measured")]] ==
                joined[[paste0(column, "_parity")]]))
}
for (column in c("anchors_sha256", "windows_sha256")) {
  stopifnot(all(vapply(split(times[[column]],
                             interaction(times$dataset, times$pass_number, drop = TRUE)),
                       function(values) length(unique(values)) == 1L, logical(1))))
}
```

## Parity before timing

Seeded genes and windows are identical across engines; sorted descendant rows
compare anchor, descendant ID, chromosome, source, type, coordinates, score,
strand, frame, source order and depth (missing numeric score is rendered as
`'.'` on the SQL side). Region hits compare sorted feature IDs for every
window, not just the total count.

```{r featuredb-parity, echo=FALSE}
parity_show <- parity[!duplicated(parity$dataset),
                      c("dataset", "features", "anchors", "windows", "descendants",
                        "region_hits", "parity")]
parity_show$passes <- as.integer(table(parity$dataset)[parity_show$dataset])
knitr::kable(parity_show, row.names = FALSE)
```

## Materialization and queries

Wall time and peak resident memory include process startup, build, and both
queries; build time covers database creation only. Each pass starts with a new
process and database. File sizes follow a closed checkpoint. The GENCODE
GFFBase process runs in a `systemd-run` scope with memory cap
`r unique(times$gffbase_memory_cap[grepl("^gencode", times$dataset)])`.
GENCODE has a single pass, so it supplies no run-to-run variance estimate.

```{r featuredb-timing, echo=FALSE}
timing_show <- do.call(rbind, lapply(split(times, list(times$dataset, times$engine),
                                            drop = TRUE), function(group) {
  data.frame(dataset = group$dataset[[1]], engine = group$engine[[1]],
             passes = nrow(group),
             build_s = round(stats::median(group$build_s), 2),
             wall_s = round(stats::median(group$wall_s), 2),
             peak_rss_GiB = round(stats::median(group$peak_rss_kb) / 1024^2, 2),
             database_GiB = round(stats::median(group$db_bytes) / 1024^3, 2),
             descendants_s = round(stats::median(group$descendant_s), 2),
             windows_s = round(stats::median(group$region_s), 2))
}))
timing_show <- timing_show[order(match(timing_show$dataset, unique(parity$dataset)),
                                 match(timing_show$engine, c("DuckHTS", "gffbase"))), ]
knitr::kable(timing_show, row.names = FALSE)
```

```{r featuredb-provenance, echo=FALSE}
input_provenance <- unique(times[c("dataset", "input_sha256", "gffbase_memory_cap")])
knitr::kable(input_provenance, row.names = FALSE)

run_provenance <- times[1, c("revision", "extension_sha256", "host", "os", "cpu",
                             "threads", "duckdb_version", "gffbase_version",
                             "duckhts_memory_limit", "validation")]
knitr::kable(data.frame(parameter = names(run_provenance),
                        value = unname(unlist(run_provenance, use.names = FALSE))),
              row.names = FALSE)
```

This is the storage and query core, not GFFBase's validation modes, GTF
synthesis, gffutils API or R-tree. The recorded DuckHTS batched descendant
query is faster: its one SQL join handles all anchors; GFFBase also batches
through its dedicated `children_batched` API. This result does not describe
per-gene SQL lookups, where GFFBase's indexed batched API can avoid repeated
query overhead. DuckDB's default memory limit is 80% of RAM; these DuckHTS
runs set a limit explicitly. GENCODE DuckHTS peak RSS exceeds both that limit
and GFFBase's measured RSS despite a smaller database. The DuckDB limit does
not cap all process allocations.

To restage inputs and the pinned wheel, repeat the measurements and render:

```sh
make bench-gffbase-featuredb
```

`GFFBASE_FEATUREDB_PYTHON` selects a Python with DuckDB and PyArrow installed;
`DUCKHTS_CACHE_DIR` selects the registry cache. Set
`GFFBASE_FEATUREDB_MANE_PASSES` and `GFFBASE_FEATUREDB_GENCODE_PASSES` to
increase repetitions. Python dependencies are installed from the verified wheel
without network access during the measurement; staging may download inputs.
