---
title: "DuckVEP SQL transcript presentation"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
runs <- read.csv(file.path(root, "benchmarks/data/duckvep_projection.csv"),
  stringsAsFactors = FALSE, check.names = FALSE)
stopifnot(all(runs$build_binding != "diagnostic_unbound"))
```

This measures the complete 23-column result of
`duckvep_transcript_projection(events, annotations, transcripts)` as a DuckDB
temporary table. It is the SQL reference, not a fused native implementation.
The literal alleles and consequence rows are prepared before timing. The timer
includes the macro's validation, joins, coordinate projection, codon/peptide
calculation and complete output materialization. One warm-up precedes the
reported measured passes; preparation and full-output verification are excluded
from those timings.

The raw FASTA, index and minimal GFF are pinned in `r/duckhtsbench` under the
`duckvep-projection` workload. `projection_fixtures.R` derives a forward model,
a reverse model, and a three-exon model with two-base CDS phase padding from
those inputs. The unchanged independent-event generator adds 1,000 seed-173
random alleles to its targeted witnesses. Only record IDs are relabelled so
repeated source alleles retain separate event keys. Each set is repeated eight
times with distinct keys. These tiny warm models are a reproducible adapter
workload, not a genome-scale annotation or file-ingestion benchmark.

Each model/worker-count pair runs in a fresh R process. Peak RSS is the Linux
process high-water mark, including model preparation and verification; it is
not isolated DuckDB workspace usage. Output bytes are the sum of the UTF-8
lengths of the complete rows serialized as JSON, not DuckDB physical storage
size. The SHA-256 covers the schema and every typed row sorted by all columns,
including duplicates and NULLs. It must match across passes and worker counts.

## Measured materialization

```{r timings}
table <- runs[c("source_revision", "case", "threads", "input_alleles",
  "expanded_transcript_rows", "output_rows", "json_utf8_bytes", "iterations",
  "median_seconds", "min_seconds", "max_seconds", "peak_rss_bytes")]
table$source_revision <- substr(table$source_revision, 1L, 12L)
knitr::kable(table)
```

## Identity and complete output

```{r identity}
knitr::kable(unique(runs[c("source_revision", "extension_sha256", "build_binding",
  "machine", "cpu", "duckdb_version")]))
knitr::kable(unique(runs[c("case", "input_vcf_sha256", "model_gff_sha256",
  "reference_sha256", "output_sha256")]))
```

The nearest existing [sorted-stream throughput report](duckvep_throughput.md)
measures consequence/HGVS output, not this complete presentation contract.
There is no identical earlier workload and no fused implementation measured
here, so these results establish a baseline rather than prove a speedup or
absence of regression. A production-density model and comparison with a future
fused implementation remain unmeasured.

Reproduce from a clean source revision with an extension build receipt:

```{r reproduce, eval=FALSE, echo=TRUE}
source("scripts/benchmark_duckvep_projection.R")
benchmark_duckvep_projection(
  normalizePath("."), "build/release/duckhts.duckdb_extension",
  "benchmarks/data/duckvep_projection.csv",
  extension_receipt = Sys.getenv("DUCKHTS_EXTENSION_RECEIPT")
)
rmarkdown::render("benchmarks/benchmark_duckvep_projection.Rmd",
  knit_root_dir = normalizePath("."))
```
