---
title: "DuckHTS Liftover Benchmark"
output: github_document
---

<!-- benchmark_liftover.md is generated from benchmark_liftover.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This benchmark:

- times DuckHTS liftover against `bcftools +liftover`
- compares normalized lifted outputs

# Run

`make bench-lift`

Useful overrides:

- `LIFTOVER_BENCH_ROWS`: synthetic VCF records, default `1000000`
- `LIFTOVER_BENCH_RUNS`: timed repeats, default `3`
- `BCFTOOLS_BIN`: optional override for the bcftools executable
- `BCFTOOLS_PLUGIN_DIR`: optional override for the bcftools plugin directory
- `LIFTOVER_REAL_VCF`, `LIFTOVER_REAL_PROVENANCE`, `LIFTOVER_REAL_REGION`, `LIFTOVER_REAL_CHAIN`, `LIFTOVER_REAL_SRC_FASTA`, `LIFTOVER_REAL_DST_FASTA`: cached real-callset inputs; use `make stage-giab-v4.2.1` and `make stage-liftover-references` to populate their defaults

For a shell-first real-callset comparison, see
`bash scripts/liftover_conformance.sh ...`.

For curated GIAB benchmark slices, see `scripts/conformance_case_table.tsv`
and run them in batch with `bash scripts/liftover_conformance_batch.sh`.

```{r setup, include = FALSE}
library(DBI)
library(duckdb)

if (!requireNamespace("RBCFTools", quietly = TRUE)) {
  stop("RBCFTools is required for bcftools +liftover benchmarking", call. = FALSE)
}

RBCFTools::setup_hts_env()

bcftools_bin <- Sys.getenv("BCFTOOLS_BIN", unset = RBCFTools::bcftools_path())
bcftools_plugin_dir <- Sys.getenv(
  "BCFTOOLS_PLUGIN_DIR",
  unset = Sys.getenv("SCORE_PLUGIN_DIR", unset = RBCFTools::bcftools_plugins_dir())
)
if (!file.exists(bcftools_bin)) stop("bcftools binary not found: ", bcftools_bin, call. = FALSE)
if (!file.exists(file.path(bcftools_plugin_dir, "liftover.so"))) {
  stop("bcftools liftover plugin not found under: ", bcftools_plugin_dir, call. = FALSE)
}
Sys.setenv(BCFTOOLS_PLUGINS = bcftools_plugin_dir)

bench_rows <- as.integer(Sys.getenv("LIFTOVER_BENCH_ROWS", unset = "1000000"))
bench_runs <- as.integer(Sys.getenv("LIFTOVER_BENCH_RUNS", unset = "3"))
duckhts_cache_dir <- Sys.getenv(
  "DUCKHTS_CACHE_DIR",
  unset = file.path(path.expand("~"), ".cache", "duckhts")
)
liftover_reference_dir <- file.path(duckhts_cache_dir, "references", "liftover", "grch37-to-grch38")
giab_cache_dir <- file.path(duckhts_cache_dir, "datasets", "giab", "nist-v4.2.1", "grch37")
real_input_path <- Sys.getenv(
  "LIFTOVER_REAL_VCF",
  unset = file.path(giab_cache_dir, "HG001_GRCh37_1_22_v4.2.1_benchmark.vcf.gz")
)
real_provenance <- Sys.getenv("LIFTOVER_REAL_PROVENANCE", unset = paste0(real_input_path, ".provenance.tsv"))
real_region <- Sys.getenv("LIFTOVER_REAL_REGION", unset = "")
real_chain_path <- Sys.getenv("LIFTOVER_REAL_CHAIN", unset = file.path(liftover_reference_dir, "GRCh37_to_GRCh38.chain.gz"))
real_src_fasta <- Sys.getenv("LIFTOVER_REAL_SRC_FASTA", unset = file.path(liftover_reference_dir, "human_g1k_v37.fasta"))
real_dst_fasta <- Sys.getenv("LIFTOVER_REAL_DST_FASTA", unset = file.path(liftover_reference_dir, "GCA_000001405.15_GRCh38_no_alt_analysis_set.fna"))

if (is.na(bench_rows) || bench_rows < 1L) stop("LIFTOVER_BENCH_ROWS must be positive")
if (is.na(bench_runs) || bench_runs < 1L) stop("LIFTOVER_BENCH_RUNS must be positive")

drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
con <- dbConnect(drv, dbdir = ":memory:")

ext_path <- normalizePath("build/release/duckhts.duckdb_extension", mustWork = TRUE)
DBI::dbExecute(con, sprintf("LOAD '%s';", gsub("\\\\", "/", ext_path)))
DBI::dbExecute(con, "PRAGMA threads=4;")
```

```{r helpers, include = FALSE}
shell_quote <- function(x) shQuote(x, type = "sh")

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

run_cmd <- function(cmd) {
  status <- system(cmd, ignore.stdout = FALSE, ignore.stderr = FALSE)
  if (!identical(status, 0L)) stop("command failed: ", cmd, call. = FALSE)
}

normalize_warning <- function(x) {
  out <- as.character(x)
  out[is.na(out) | out == "" | out == "."] <- "PASS"
  out
}

write_large_fixture <- function(dir_path, n_rows) {
  dir.create(dir_path, recursive = TRUE, showWarnings = FALSE)

  src_fa <- file.path(dir_path, "liftover_src.fa")
  dst_fa <- file.path(dir_path, "liftover_dst.fa")
  chain_path <- file.path(dir_path, "liftover.chain")
  vcf_path <- file.path(dir_path, "liftover_input.vcf")
  vcf_gz <- paste0(vcf_path, ".gz")

  writeLines(c(
    ">chrF",
    "ACGTACGTAA",
    ">chrR",
    "AACCGGTTAA"
  ), src_fa)
  writeLines(c(
    ">chrLiftF",
    "ACGTACGTAA",
    ">chrLiftR",
    "TTAACCGGTT"
  ), dst_fa)
  writeLines(c(
    "chain 100 chrF 10 + 0 10 chrLiftF 10 + 0 10 1",
    "10",
    "",
    "chain 100 chrR 10 + 0 10 chrLiftR 10 - 0 10 2",
    "10"
  ), chain_path)

  n_f <- n_rows %/% 2L + n_rows %% 2L
  n_r <- n_rows %/% 2L
  chroms <- c(rep.int("chrF", n_f), rep.int("chrR", n_r))
  refs <- c(rep.int("C", n_f), rep.int("A", n_r))
  alts <- c(rep.int("T", n_f), rep.int("G", n_r))

  vcf_lines <- sprintf(
    "%s\t2\tbench_%d\t%s\t%s\t.\t.\t.",
    chroms,
    seq_len(n_rows),
    refs,
    alts
  )
  writeLines(c(
    "##fileformat=VCFv4.2",
    "##contig=<ID=chrF,length=10>",
    "##contig=<ID=chrR,length=10>",
    "#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO",
    vcf_lines
  ), vcf_path)

  DBI::dbGetQuery(
    con,
    sprintf(
      "SELECT * FROM fasta_index(%s, index_path := %s)",
      sql_string(src_fa),
      sql_string(paste0(src_fa, ".fai"))
    )
  )
  DBI::dbGetQuery(
    con,
    sprintf(
      "SELECT * FROM fasta_index(%s, index_path := %s)",
      sql_string(dst_fa),
      sql_string(paste0(dst_fa, ".fai"))
    )
  )
  DBI::dbGetQuery(
    con,
    sprintf(
      "SELECT * FROM bgzip(%s, output_path := %s, keep := true, overwrite := true)",
      sql_string(vcf_path),
      sql_string(vcf_gz)
    )
  )
  DBI::dbGetQuery(
    con,
    sprintf(
      "SELECT * FROM bcf_index(%s, index_path := %s, threads := 1)",
      sql_string(vcf_gz),
      sql_string(paste0(vcf_gz, ".tbi"))
    )
  )

  list(
    input_path = vcf_gz,
    chain_path = chain_path,
    src_fasta = src_fa,
    dst_fasta = dst_fa,
    n_rows = n_rows
  )
}

prepare_real_fixture <- function(input_path, chain_path, src_fasta, dst_fasta, region = "") {
  if (!file.exists(chain_path)) stop("Chain file not found: ", chain_path, call. = FALSE)
  if (!file.exists(src_fasta)) stop("Source FASTA not found: ", src_fasta, call. = FALSE)
  if (!file.exists(dst_fasta)) stop("Destination FASTA not found: ", dst_fasta, call. = FALSE)

  out_dir <- file.path(tempdir(), "duckhts_liftover_real")
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  input_local <- file.path(out_dir, "real_input.vcf.gz")

  region_arg <- if (nzchar(region)) sprintf("-r %s ", shell_quote(region)) else ""
  run_cmd(sprintf(
    "%s view %s-Oz -o %s %s",
    shell_quote(bcftools_bin),
    region_arg,
    shell_quote(input_local),
    shell_quote(input_path)
  ))
  run_cmd(sprintf("%s index -f -t %s", shell_quote(bcftools_bin), shell_quote(input_local)))

  list(
    input_path = input_local,
    chain_path = chain_path,
    src_fasta = src_fasta,
    dst_fasta = dst_fasta,
    region = region
  )
}

prepare_duckdb_input_table <- function(input_path) {
  DBI::dbExecute(con, "DROP TABLE IF EXISTS bench_input")
  DBI::dbExecute(
    con,
    paste0(
      "CREATE TEMP TABLE bench_input AS ",
      "SELECT CHROM AS chrom, POS AS pos, REF AS ref, array_to_string(ALT, ',') AS alt ",
      "FROM read_bcf(", sql_string(input_path), ")"
    )
  )
}

duckdb_liftover_sql <- function(chain_path, src_fasta, dst_fasta) {
  paste0(
    "SELECT dest_chrom, dest_pos, dest_ref, dest_alt, COUNT(*) AS n ",
    "FROM duckdb_liftover(",
    "'bench_input', 'chrom', 'pos', ref_col := 'ref', alt_col := 'alt', ",
    "chain_path := ", sql_string(chain_path), ", ",
    "dst_fasta_ref := ", sql_string(dst_fasta), ", ",
    "src_fasta_ref := ", sql_string(src_fasta), ", ",
    "max_snp_gap := 1, max_indel_inc := 250",
    ") lifted ",
    "WHERE mapped ",
    "GROUP BY ALL ",
    "ORDER BY dest_chrom, dest_pos, dest_ref, dest_alt"
  )
}

run_duckdb_bench <- function(input_path, chain_path, src_fasta, dst_fasta, n = 3L) {
  prepare_duckdb_input_table(input_path)
  sql <- duckdb_liftover_sql(chain_path, src_fasta, dst_fasta)
  times <- numeric(n)
  rows <- NA_integer_

  for (i in seq_len(n)) {
    gc()
    t0 <- proc.time()[["elapsed"]]
    out <- DBI::dbGetQuery(con, sql)
    times[i] <- proc.time()[["elapsed"]] - t0
    rows <- sum(out$n)
  }

  data.frame(
    engine = "duckhts",
    runs = n,
    median_sec = stats::median(times),
    min_sec = min(times),
    max_sec = max(times),
    output_rows = rows,
    stringsAsFactors = FALSE
  )
}

run_duckdb_output <- function(input_path, chain_path, src_fasta, dst_fasta) {
  prepare_duckdb_input_table(input_path)
  DBI::dbGetQuery(con, duckdb_liftover_sql(chain_path, src_fasta, dst_fasta))
}

run_bcftools_once <- function(input_path, chain_path, src_fasta, dst_fasta, output_path) {
  cmd <- sprintf(
    "BCFTOOLS_PLUGINS=%s %s +liftover --no-version -Ob -o %s %s -- -s %s -f %s -c %s",
    shell_quote(bcftools_plugin_dir),
    shell_quote(bcftools_bin),
    shell_quote(output_path),
    shell_quote(input_path),
    shell_quote(src_fasta),
    shell_quote(dst_fasta),
    shell_quote(chain_path)
  )
  gc()
  t0 <- proc.time()[["elapsed"]]
  status <- system(cmd, ignore.stdout = TRUE, ignore.stderr = FALSE)
  elapsed <- proc.time()[["elapsed"]] - t0
  if (!identical(status, 0L)) stop("bcftools +liftover failed", call. = FALSE)
  list(
    elapsed = elapsed,
    output = output_path,
    normalized = DBI::dbGetQuery(con, bcftools_output_sql(output_path))
  )
}

bcftools_output_sql <- function(output_path) {
  paste0(
    "SELECT CHROM AS dest_chrom, POS AS dest_pos, REF AS dest_ref, array_to_string(ALT, ',') AS dest_alt, ",
    "COUNT(*) AS bcftools_n ",
    "FROM read_bcf(", sql_string(output_path), ") ",
    "GROUP BY ALL ",
    "ORDER BY dest_chrom, dest_pos, dest_ref, dest_alt"
  )
}

run_benchmark_case <- function(input_path, chain_path, src_fasta, dst_fasta, n = 3L) {
  duck <- run_duckdb_bench(input_path, chain_path, src_fasta, dst_fasta, n = n)
  duck_out <- run_duckdb_output(input_path, chain_path, src_fasta, dst_fasta)

  bcftools_times <- numeric(n)
  bcftools_out <- NULL
  output_rows <- NA_integer_
  for (i in seq_len(n)) {
    out_bcf <- tempfile("bcftools_liftover_", fileext = ".bcf")
    run <- run_bcftools_once(input_path, chain_path, src_fasta, dst_fasta, out_bcf)
    bcftools_times[i] <- run$elapsed
    output_rows <- sum(run$normalized$bcftools_n)
    if (i == 1L) {
      bcftools_out <- run$normalized
    }
    unlink(c(out_bcf, paste0(out_bcf, ".csi")), force = TRUE)
  }

  list(
    bench = rbind(
      duck,
      data.frame(
        engine = "bcftools_RBCFTools",
        runs = n,
        median_sec = stats::median(bcftools_times),
        min_sec = min(bcftools_times),
        max_sec = max(bcftools_times),
        output_rows = output_rows,
        stringsAsFactors = FALSE
      )
    ),
    compare = compare_outputs(duck_out, bcftools_out),
    compare_examples = compare_examples(duck_out, bcftools_out)
  )
}

compare_outputs <- function(duck_out, bcf_out) {
  names(duck_out)[names(duck_out) == "n"] <- "duck_n"
  merged <- merge(
    duck_out,
    bcf_out,
    by = c("dest_chrom", "dest_pos", "dest_ref", "dest_alt"),
    all = TRUE
  )
  merged$duck_n[is.na(merged$duck_n)] <- 0L
  merged$bcftools_n[is.na(merged$bcftools_n)] <- 0L

  data.frame(
    duck_rows = sum(merged$duck_n),
    bcftools_rows = sum(merged$bcftools_n),
    only_duck_groups = sum(merged$duck_n > 0L & merged$bcftools_n == 0L),
    only_bcftools_groups = sum(merged$bcftools_n > 0L & merged$duck_n == 0L),
    mismatched_count_groups = sum(
      merged$duck_n > 0L &
        merged$bcftools_n > 0L &
        merged$duck_n != merged$bcftools_n
    ),
    outputs_match = all(merged$duck_n == merged$bcftools_n),
    stringsAsFactors = FALSE
  )
}

compare_examples <- function(duck_out, bcf_out, n = 20L) {
  names(duck_out)[names(duck_out) == "n"] <- "duck_n"
  merged <- merge(
    duck_out,
    bcf_out,
    by = c("dest_chrom", "dest_pos", "dest_ref", "dest_alt"),
    all = TRUE
  )
  merged$duck_n[is.na(merged$duck_n)] <- 0L
  merged$bcftools_n[is.na(merged$bcftools_n)] <- 0L
  head(merged[merged$duck_n != merged$bcftools_n, , drop = FALSE], n)
}
```

## Tools

```{r tools, echo = FALSE}
data.frame(
  tool = c("bcftools", "bcftools_plugins"),
  path = c(bcftools_bin, bcftools_plugin_dir),
  stringsAsFactors = FALSE
)
```

## Settings

```{r settings, echo = FALSE}
data.frame(
  synthetic_rows = bench_rows,
  runs = bench_runs,
  real_region = if (nzchar(real_region)) real_region else NA_character_,
  stringsAsFactors = FALSE
)
```

## Synthetic Stress Case

```{r synthetic, echo = FALSE}
fixture_dir <- file.path(tempdir(), "duckhts_liftover_benchmark")
fixture <- write_large_fixture(fixture_dir, n_rows = bench_rows)
bench_case <- run_benchmark_case(
  fixture$input_path,
  fixture$chain_path,
  fixture$src_fasta,
  fixture$dst_fasta,
  n = bench_runs
)

bench_case$bench
```

```{r synthetic-compare, echo = FALSE}
bench_case$compare
```

```{r real-case-setup, include = FALSE}
required_real_paths <- c(
  input_vcf = real_input_path,
  provenance = real_provenance,
  chain = real_chain_path,
  src_fasta = real_src_fasta,
  dst_fasta = real_dst_fasta
)
real_enabled <- all(nzchar(required_real_paths)) && all(file.exists(required_real_paths))
if (real_enabled) {
  real_fixture <- prepare_real_fixture(
    real_input_path,
    real_chain_path,
    real_src_fasta,
    real_dst_fasta,
    region = real_region
  )
  real_case <- run_benchmark_case(
    real_fixture$input_path,
    real_fixture$chain_path,
    real_fixture$src_fasta,
    real_fixture$dst_fasta,
    n = bench_runs
  )
} else {
  real_case <- NULL
}
```

## Real-Callset Case

```{r real-case-config, echo = FALSE}
if (real_enabled) {
  data.frame(
    input_vcf = real_input_path,
    provenance = real_provenance,
    region = if (nzchar(real_region)) real_region else "full_input",
    chain = real_chain_path,
    src_fasta = real_src_fasta,
    dst_fasta = real_dst_fasta,
    stringsAsFactors = FALSE
  )
} else {
  data.frame(status = "skipped; set all LIFTOVER_REAL_* paths", stringsAsFactors = FALSE)
}
```

```{r real-case-bench, echo = FALSE}
if (real_enabled) real_case$bench
```

```{r real-case-compare, echo = FALSE}
if (real_enabled) real_case$compare
```

```{r real-case-examples, echo = FALSE}
if (real_enabled && nrow(real_case$compare_examples)) real_case$compare_examples
```
