---
title: "Record-major genotypes and carrier expansion on an HPRC cohort"
output:
  github_document:
    html_preview: false
---

The registered HPRC v2.0 GRCh38 cohort region `chr22:20000000-21000000`
retains every source sample and allele. This is not the consequence-only,
split-allele corpus. Stage explicitly with
`duckhtsbench::duckhts_bench_stage_genotypes()` before rendering; rendering is
network-free. BCF and VCF.gz contain the same complete records, verified by
the staging test and full HTSlib text comparison.

This compares `read_geno()` with the canonical `read_bcf(..., tidy_format:=true)`
GT projection. The experiment named `read_bcf_v2` in the original request was
deleted; it is not reintroduced as a benchmark dependency. Full cohorts, the
first eight header samples, sparse non-reference calls, and downstream typed
carrier expansion have separate denominators. The input has GT but no PS;
these timings do not measure PS decoding, which SQL/R/native fixtures test.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
knitr::opts_knit$set(root.dir=normalizePath(".."))
library(DBI)
library(duckhtsbench)
source("benchmarks/genotype_reader_run.R")
extension <- normalizePath(Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"), mustWork=TRUE)
stopifnot(system2("git", c("diff", "--quiet", "HEAD", "--", "src")) == 0L)
revision <- system2("git", c("rev-parse", "HEAD"), stdout=TRUE)
src_tree <- system2("git", c("rev-parse", "HEAD:src"), stdout=TRUE)
paths <- c(VCF = duckhts_bench_artifact_path("geno_hprc_vcfgz"), BCF = duckhts_bench_artifact_path("geno_hprc_bcf"))
identities <- list()
for (i in seq_along(paths)) {
  id <- c("geno_hprc_vcfgz", "geno_hprc_bcf")[[i]]
  duckhts_bench_validate_identity(id, paths[[i]])
  receipt <- read.delim(paste0(paths[[i]], ".provenance.tsv"))
  values <- setNames(receipt$value, receipt$field)
  stopifnot(identical(unname(tools::md5sum(paths[[i]])), values[["observed_md5"]]))
  identities[[i]] <- data.frame(format=names(paths)[i], artifact=id,
                                bytes=file.info(paths[[i]])$size, md5=values[["observed_md5"]])
}
identities <- do.call(rbind, identities)
bcftools <- function(args) {
  result <- system2("bcftools", shQuote(args), stdout=TRUE)
  stopifnot(is.null(attr(result,"status")))
  result
}
samples <- bcftools(c("query","-l",paths[[1]]))
header <- bcftools(c("view","-h",paths[[1]]))
stopifnot("##fileformat=VCFv4.2" %in% header,
          sum(startsWith(header,"##FORMAT=")) == 1L,
          any(startsWith(header,"##FORMAT=<ID=GT,")), length(samples) >= 8L)
gt_lines <- bcftools(c("query","-f","[%GT\t]\n",paths[[1]]))
input_records <- length(gt_lines)
# Independent HTSlib GT strings supply all workload denominators. No optimized
# reader count path, dropped records or changed oracle is used for a rate.
denominators <- lapply(c("full","selected","sparse","carriers"), function(workload) {
  selected <- if (workload == "selected") seq_len(8L) else seq_along(samples)
  sums <- vapply(strsplit(gt_lines,"\t",fixed=TRUE), function(gt) {
    stopifnot(length(gt) == length(samples))
    gt <- gt[selected]
    slots <- nchar(gsub("[^|/]","",gt)) + 1L
    nonref <- grepl("(^|[|/])[1-9][0-9]*($|[|/])",gt)
    alt <- lengths(regmatches(gt, gregexpr("(^|[|/])[1-9][0-9]*",gt)))
    c(calls=length(gt), slots=sum(slots), nonref=sum(nonref),
      sparse_slots=sum(slots[nonref]), alt_slots=sum(alt))
  }, numeric(5))
  total <- rowSums(sums)
  data.frame(workload, input_records, selected_samples=length(selected),
             selected_calls=total[["calls"]], selected_slots=total[["slots"]],
             non_reference_calls=total[["nonref"]], sparse_slots=total[["sparse_slots"]],
             alt_slots=total[["alt_slots"]])
})
denominators <- do.call(rbind,denominators)
rm(gt_lines)
repetitions <- as.integer(Sys.getenv("GENO_BENCH_RUNS","3"))
stopifnot(repetitions > 0L)
directory <- tempfile("duckhts-genotype-report-")
dir.create(directory)
results <- comparisons <- list()
con <- dbConnect(duckdb::duckdb())
dbExecute(con,"SET threads=1")
dbExecute(con,"SET memory_limit='8GB'")
for (format in names(paths)) for (workload in denominators$workload) {
  expected <- denominators[denominators$workload==workload,]
  snapshots <- c()
  for (repetition in seq_len(repetitions)) {
    readers <- if (repetition %% 2L) c("read_bcf","read_geno") else c("read_geno","read_bcf")
    for (reader in readers) {
      snapshot <- if (repetition==1L) file.path(directory,paste(format,workload,reader,"parquet",sep=".")) else ""
      if (nzchar(snapshot)) snapshots[reader] <- snapshot
      observed <- callr::r(genotype_reader_run, args=list(extension=extension, input=paths[[format]],
        reader=reader, workload=workload,
        selector=if(workload=="selected") paste(samples[seq_len(8L)],collapse=",") else "",
        sample_names=samples, snapshot=snapshot, threads=1L))
      if (workload=="carriers") stopifnot(observed$output_rows==expected$alt_slots)
      else {
        stopifnot(observed$output_calls==if(workload=="sparse") expected$non_reference_calls else expected$selected_calls,
                  observed$output_slots==if(workload=="sparse") expected$sparse_slots else expected$selected_slots,
                  observed$output_rows==if(reader=="read_geno") input_records else observed$output_calls)
      }
      results[[length(results)+1L]] <- cbind(format, repetition, observed)
    }
  }
  reference <- paste0("read_parquet(",dbQuoteString(con,snapshots[["read_bcf"]]),")")
  candidate <- paste0("read_parquet(",dbQuoteString(con,snapshots[["read_geno"]]),")")
  sql <- sprintf(paste("SELECT count(*) AS n FROM ((SELECT * FROM %s EXCEPT ALL SELECT * FROM %s)",
    "UNION ALL (SELECT * FROM %s EXCEPT ALL SELECT * FROM %s))"),reference,candidate,candidate,reference)
  difference <- as.numeric(dbGetQuery(con,sql)$n)
  stopifnot(difference==0)
  comparisons[[length(comparisons)+1L]] <- data.frame(format,workload,different_typed_rows=difference)
  unlink(snapshots)
}
dbDisconnect(con,shutdown=TRUE)
unlink(directory,recursive=TRUE)
results <- do.call(rbind,results)
comparisons <- do.call(rbind,comparisons)
for (key in split(results,interaction(results$format,results$reader,results$workload)))
  stopifnot(length(unique(key$checksum))==1L)
```

## Source and workload identity

```{r identity}
cat("Source revision:",revision,"\nSource tree:",src_tree,"\n")
cat("Extension MD5:",unname(tools::md5sum(extension)),"\n")
cat("R:",R.version.string,"; DuckDB:",as.character(packageVersion("duckdb")),"\n")
cat(system2("uname","-srmo",stdout=TRUE),"\n")
cat(system2("taskset",c("-pc",Sys.getpid()),stdout=TRUE),"\n")
cat(system2("sh",c("-c",shQuote("lscpu | sed -n '/Model name:/p'")),stdout=TRUE),"\n")
cat("DuckDB threads: 1; scan handles: 1; HTSlib decompression workers: 0; repetitions:",repetitions,"\n")
knitr::kable(identities)
knitr::kable(denominators)
```

## Materialization and carrier expansion

Timing includes binding, decoding, selected output vectors and `CREATE TABLE AS`.
Carrier timing also includes allele-slot expansion and ALT selection. Every
measurement uses a fresh R/DuckDB process; source data is warm in the OS cache.
Startup, independent counting, checksums, checkpoint, snapshots and exact output
comparison are outside timing. RSS is the process high-water mark immediately
after materialization. Materialized database bytes are measured after checkpoint
and include the tiny sample catalog and database overhead; they are physical
compressed storage, not vector-memory bytes. The fixed DuckDB memory limit is
8 GiB; these are not sealed/static-allocation measurements.

```{r timing}
summary <- aggregate(cbind(elapsed,peak_rss_kib,materialized_database_bytes) ~ format+reader+workload,results,median)
summary <- merge(summary,denominators,by="workload",sort=FALSE)
summary$records_per_second <- summary$input_records/summary$elapsed
summary$calls_per_second <- summary$selected_calls/summary$elapsed
summary$allele_slots_per_second <- summary$selected_slots/summary$elapsed
summary$input_bytes_per_second <- identities$bytes[match(summary$format,identities$format)]/summary$elapsed
knitr::kable(summary[c("format","reader","workload","elapsed","records_per_second","calls_per_second",
                      "allele_slots_per_second","input_bytes_per_second","peak_rss_kib","materialized_database_bytes")],digits=3)
knitr::kable(results[c("format","reader","workload","repetition","output_rows","output_calls","output_slots","elapsed","cpu")],digits=3)
```

## Complete output comparison

First-run results are normalized outside the timer to typed variant identity,
original-header sample index, every allele slot, every phase flag and phase set.
For the VCF 4.2 source, the comparison reconstructs the leading phase flag from
HTSlib's string representation; VCF 4.4 explicit-prefix semantics are tested in
the committed SQL/R witnesses, not inferred from this dataset. Every field and
duplicate contributes to `EXCEPT ALL` in both directions. Carrier comparisons
retain variant identity, sample, allele slot/index, selected ALT, phase and PS.
Each repetition separately checks the complete materialized-row checksum.

```{r agreement}
knitr::kable(comparisons)
```

The nearest earlier full-materialization report is
[the shared BCF scanner report](benchmark_bcf_shared_scan.md): a single-sample
GIAB workload, not an identical cohort baseline. This report compares the two
current interfaces, not pre-change and post-change revisions. The separately
rendered paired BCF regression workload measures changes to the existing reader.
These results do not establish universal fastest-reader performance, remote-I/O
throughput, multi-worker scaling, PS throughput, or phased annotation correctness.
