---
title: "DuckHTS cgranges Benchmark"
output: github_document
---

<!-- benchmark_cgranges.md is generated from benchmark_cgranges.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This benchmark measures:

- `duckhts_cgranges_from_query(...)` + `duckhts_cgranges_index(...)`
- repeated bound overlap probes through `duckhts_cgranges_overlaps(...)`
- upstream `bedtk flt` as the overlap-existence comparator
- wall time and peak RSS for both engines

It uses a deterministic synthetic interval set so the benchmark is stable and
rerunnable.

# Run

```sh
python3 scripts/cgranges_benchmark.py \
  --extension build/release/duckhts.duckdb_extension \
  --bedtk .sync/bedtk/bedtk \
  --subjects 50000 \
  --queries 5000 \
  --passes 3 \
  --out-dir .tmp/cgranges_benchmark
```

```{r setup, include = FALSE}
library(readr)
library(knitr)

repo_root <- normalizePath(file.path(dirname(knitr::current_input()), ".."), mustWork = TRUE)
out_dir <- file.path(repo_root, ".tmp", "cgranges_benchmark")
summary_csv <- file.path(out_dir, "summary.csv")

subjects <- as.integer(Sys.getenv("CGRANGES_BENCH_SUBJECTS", unset = "50000"))
queries <- as.integer(Sys.getenv("CGRANGES_BENCH_QUERIES", unset = "5000"))
passes <- as.integer(Sys.getenv("CGRANGES_BENCH_PASSES", unset = "3"))

if (!file.exists(summary_csv)) {
  cmd <- c(
    file.path(repo_root, "scripts", "cgranges_benchmark.py"),
    "--extension", file.path(repo_root, "build", "release", "duckhts.duckdb_extension"),
    "--bedtk", file.path(repo_root, ".sync", "bedtk", "bedtk"),
    "--subjects", as.character(subjects),
    "--queries", as.character(queries),
    "--passes", as.character(passes),
    "--out-dir", out_dir
  )
  system2("python3", cmd, stdout = TRUE, stderr = TRUE)
}

bench <- readr::read_csv(summary_csv, show_col_types = FALSE)
bench$time_per_query_ms <- (bench$query_total_sec / (bench$queries * bench$passes)) * 1000
bench$peak_rss_mb <- round(bench$peak_rss_mb, 1)
bench$build_index_sec <- round(bench$build_index_sec, 3)
bench$query_total_sec <- round(bench$query_total_sec, 3)
bench$query_pass_1_sec <- round(bench$query_pass_1_sec, 3)
bench$total_elapsed_sec <- round(bench$total_elapsed_sec, 3)
bench$time_per_query_ms <- round(bench$time_per_query_ms, 4)
```

# Configuration

```{r, echo=FALSE}
knitr::kable(
  data.frame(
    parameter = c("subjects", "queries", "passes"),
    value = c(subjects, queries, passes)
  )
)
```

# Results

```{r, echo=FALSE}
knitr::kable(bench)
```

# Notes

- `bedtk` reports overlap-existence for the whole query BED in one pass.
- DuckHTS builds an immutable registry index once, then reuses it across probe passes.
- `matched_query_intervals` should agree across both engines.
- `build_index_sec` is only meaningful for DuckHTS because the public API exposes an explicit build/finalize step.
