---
title: "Ancestry-tuned allele frequencies for ROH"
output: github_document
---

## Evaluation protocol

This report evaluates whether ancestry-tuned allele frequencies reduce unsupported long runs of homozygosity (ROH) in admixed 1000 Genomes trio children without changing non-admixed controls. The acceptance criterion is declared before calculating or comparing any arm:

- In admixed children, ancestry-tuned AF must yield fewer unsupported ROH of at least 1 Mb than both pooled and single-population AF.
- Its fraction of frequency-free truth-run length covered must be no more than 5% relatively below either comparator.
- In controls, both unsupported-ROH count and truth coverage must remain within 5% relative of the single-population arm. For a zero comparator, equality is required.

No amendment to the truth threshold has been made. A truth segment is a merged run of 100 kb windows with at most one heterozygous call per window, and a retained run covers at least 1 Mb of actual bases. An ROH call is truth-supported when at least 90% of its length overlaps these truth intervals. Truth calls use all biallelic SNVs in each full source VCF, not just ancestry-reference sites. The window distribution and separation check must be reviewed before any ROH-arm comparison; if inadequate, the threshold may only be amended here before computing arm results. The primary length threshold is 1 Mb, with 2 Mb and 5 Mb also reported. FROH is total called ROH length divided by callable autosomal length.

## Data identity and processing status

The phased high-coverage VCF source is `https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20201028_3202_phased/CCDG_14151_B01_GRM_WGS_2020-08-05_chr{N}.filtered.shapeit2-duohmm-phased.vcf.gz`. The pedigree source is `https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/20130606_g1k_3202_samples_ped_population.txt`. The ancestry reference is registry artifact `ancestry_reference_grch38_parquet` (GRCh38, 5,012,826 sites).

| Input | Identity / SHA-256 | Status |
|:--|:--|:--|
| chr20 phased VCF | Source SHA-256 `c2624c726c6cb27e288fe5908621125740379a3d523299dd07c13abf97645e4b`; remote ETag `2fbb7ed2-5b2d2762b9fc0`; Last-Modified `Thu, 29 Oct 2020 17:17:59 GMT` | Local source checksum verified; 3,202-sample source used to make the 377-child BCF |
| chr20 children BCF | 849,143 records; SHA-256 `a53f83cebd7de6fb4661c14ecd93e5346022ff453392095a085e0b7cda116541` | All 377 eligible children present; biallelic SNVs with INFO/AF, INFO/AF_AFR/AMR/EAS/EUR retained |
| GRCh38 ancestry reference | SHA-256 `4982810aa69b9c71924e9a02dcd454ccedccc34bd8bcf71506ce9752589847fa` | Chr20 wide rows converted to long form; 108,757 non-palindromic, matched dosage loci per child |
| Pedigree | SHA-256 `4e164b717433fc9bfc16e29b20e5b4659b4216185b3275322d421fe3d323c132` | Downloaded; expected child counts matched in all ten populations; all 377 found in the chr20 VCF header |

The planned trio counts are ACB 20, ASW 13, CLM 35, MXL 32, PEL 35 and PUR 35 (admixed), and YRI 56, ESN 43, CEU 57 and CHS 51 (controls). Single-population AF uses AFR for ACB, ASW, YRI and ESN; AMR for CLM, MXL, PEL and PUR; EUR for CEU; and EAS for CHS.

Chr20 has the complete chromosome-specific truth, q and arm evaluation. The 22-autosome truth-window table and retained intervals are computed; chr20 windows were reused and the 102 retained chr20 intervals match the prior result exactly. Genome-wide q now contains 7,917 estimates (377 children × 21 groups), all with `status=ok`; each child contributed 4,883,749 matched loci and the `cor_pred` range is 0.643847–0.719384. Genome-wide arm comparisons remain pending. The population child counts match the declared expectations: ACB 20, ASW 13, CLM 35, MXL 32, PEL 35, PUR 35, YRI 56, ESN 43, CEU 57 and CHS 51. The chr20 source VCF was copied from the supplied local file after its checksum was verified; its remote metadata is recorded above. bcftools reported `1.23.1-70-g6dbd8fef`; the source header had 3,202 samples and the children BCF has 377. The DuckHTS extension was rebuilt from revision `86f53052` for the arm measurements. The remote chr1 input was streamed through bcftools; its ETag, Last-Modified value, record count and output SHA-256 are recorded in [its receipt](results/roh-ancestry/chr1/input_receipt.tsv).

## Autosome BCF staging

Every children BCF has 377 samples. The per-chromosome source URL, Last-Modified value, ETag, record count and output SHA-256 are listed in [the staging summary](results/roh-ancestry/autosome_staging_summary.csv); raw source VCFs were streamed and not cached.

```{r, echo=FALSE}
staging <- utils::read.csv("results/roh-ancestry/autosome_staging_summary.csv")
knitr::kable(staging[c("chromosome", "source_last_modified", "source_etag",
  "source_samples", "child_samples", "child_records", "output_sha256")],
  row.names = FALSE)
```

## Truth-window separation

The complete chr20 truth-window distribution contains 243,165 child-windows (645 windows per child, including the terminal partial window). Of these, 6,572 have zero heterozygotes and 2,840 have one; 9,412 (3.87%) meet the declared `<=1` criterion, while 233,753 (96.13%) have at least two. The binned distribution is [here](results/roh-ancestry/chr20/truth_window_bins.csv), and the exact per-count distribution is [here](results/roh-ancestry/chr20/truth_window_distribution.csv). This separates the low-heterozygosity tail from the remaining windows sufficiently to retain the preregistered cutoff; no amendment was made. The retained truth is 102 runs of at least 1 Mb across 92 children; the run counts and truth-base denominators by population are [here](results/roh-ancestry/chr20/truth_runs_by_population.csv).

The binned distribution is summarized below; the exact per-count table is [here](results/roh-ancestry/chr20/truth_window_distribution.csv).

```{r, echo=FALSE}
knitr::kable(utils::read.csv("results/roh-ancestry/chr20/truth_window_bins.csv"))
```

### Full-autosome truth windows

The full truth set contains all child BCF biallelic SNVs across autosomes 1–22. Window sizes are 100 kb; terminal windows use their actual GRCh38 primary-assembly length from the Ensembl 116 FASTA index. A retained truth run must cover at least 1,000,000 actual bases. The chr20 window Parquet was reused; its 102 intervals exactly match the chromosome-specific evaluation.

Across 377 children there are 10,842,520 child-windows: 745,963 (6.88%) have at most one heterozygote and 10,096,557 (93.12%) have at least two. There are 7,965 retained 1 Mb-or-longer intervals across all 377 children. The [exact per-count window distribution](results/roh-ancestry/autosome_truth_window_distribution.csv), [retained intervals](results/roh-ancestry/autosome_truth_intervals.csv), [per-child truth metrics](results/roh-ancestry/autosome_truth_by_child.csv) and [population summary](results/roh-ancestry/autosome_truth_by_population.csv) are available as result tables.

```{r, echo=FALSE}
autosome_truth_by_population <- utils::read.csv(
  "results/roh-ancestry/autosome_truth_by_population.csv")
knitr::kable(autosome_truth_by_population, digits = 4, row.names = FALSE)
```

The pooled and single-population comparisons began only after the chr20 truth distribution was reviewed. The full-autosome separation check also retains the preregistered cutoff.

## Ancestry proportions

Proportions were estimated from chr20 dosage input only, using the GRCh38 wide reference and the bigsnpr 1.12.21 correction. Each child contributed 108,757 matched, non-palindromic loci. Every child has `status=ok`; the `cor_pred` range and top three groups are reported per child in [this table](results/roh-ancestry/chr20/q_by_child.csv). This is a chr20-only estimate, not the final genome-wide q.

The three ROH arms used the same 108,757 sites for every child. The AF and ancestry reference site keys have zero entries in either direction of their symmetric difference; the exact comparison is [here](results/roh-ancestry/chr20/site_set_equivalence.csv). The `INFO_AF*` values are array-valued Number=A fields; because the input BCF is biallelic, the AF relation takes its sole value and clamps it to `[1e-3, 0.999]`. The ancestry arm uses the same clamp. Per-child site counts are [here](results/roh-ancestry/chr20/site_counts_by_child.csv).

```{r, echo=FALSE}
knitr::kable(utils::read.csv("results/roh-ancestry/chr20/q_by_child.csv"), digits = 4)
```

### Full-autosome q estimates

Genome-wide q uses all 22 child BCFs matched to GRCh38 non-palindromic biallelic reference sites and the bigsnpr 1.12.21 correction, with `min_cor = 0.4`. Eight-child batches were used under a 12 GB DuckDB memory limit. All 377 children have estimates for 21 groups, each child's proportions sum to one, and every estimate has `status=ok`. The full per-child, per-group estimates and allele-harmonization counts are available in [the q table](results/roh-ancestry/autosome_q_by_child.csv) and [per-child site diagnostics](results/roh-ancestry/autosome_q_site_counts_by_child.csv).

```{r, echo=FALSE}
q_quality <- utils::read.csv("results/roh-ancestry/autosome_q_quality_by_population.csv")
q_by_population <- utils::read.csv("results/roh-ancestry/autosome_q_summary_by_population.csv")
knitr::kable(q_quality, digits = 4, row.names = FALSE)
knitr::kable(q_by_population, digits = 4, row.names = FALSE)
```

## ROH results

All count and length summaries are per-child quantities and are aggregated by population only after retaining each child's observations. The chr20-only FROH column is reported as a chromosome-length fraction; it is not the declared genome-wide FROH denominator.

```{r, echo=FALSE}
summary <- utils::read.csv("results/roh-ancestry/chr20/population_arm_summary.csv", check.names = FALSE)
population_order <- c("ACB", "ASW", "CLM", "MXL", "PEL", "PUR", "YRI", "ESN", "CEU", "CHS")
arm_order <- c("pooled", "single_population", "ancestry_tuned")
summary$Population <- factor(summary$Population, levels = population_order)
summary$Arm <- factor(summary$Arm, levels = arm_order)
summary <- summary[order(summary$Population, summary$Arm), ]
summary$Population <- as.character(summary$Population)
summary$Arm <- c(pooled = "Pooled", single_population = "Single population",
                 ancestry_tuned = "Ancestry tuned")[as.character(summary$Arm)]
knitr::kable(summary, row.names = FALSE)
```

## Criterion

The chr20 criterion **`r if (all(utils::read.csv("results/roh-ancestry/chr20/criterion_by_population.csv")$criterion_pass)) "held" else "did not hold"`**. The comparison for each population is shown below.

```{r, echo=FALSE}
criterion <- utils::read.csv("results/roh-ancestry/chr20/criterion_by_population.csv")
knitr::kable(criterion, digits = 4, row.names = FALSE)
```

## Runtime and memory

Each value is from a fresh R process using four DuckDB threads. The ancestry arm used 16-child batches after the 64-child batch exceeded the 12 GB DuckDB memory limit; other arms used 64-child batches. Per-replicate runtime and peak resident memory are reported below.

```{r, echo=FALSE}
runtime <- utils::read.csv("results/roh-ancestry/chr20/runtime_replicates.csv")
knitr::kable(runtime, digits = 2, row.names = FALSE)
```
