---
title: "Ancestry projection: public-input scaling and reference parity"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(DBI)
library(Rduckhts)
library(duckhtsbench)
repo <- normalizePath("..")
source_revision <- system2("git", c("-C", shQuote(repo), "rev-parse", "HEAD"), stdout = TRUE)
registry <- file.path(repo, "r/duckhtsbench/inst/benchmark_registry.tsv")
cache <- tempfile("duckhts-ancestry-benchmark-")
dir.create(cache)
Sys.setenv(DUCKHTSBENCH_REGISTRY = registry, DUCKHTS_CACHE_DIR = cache)
correction_file <- duckhts_bench_stage_repository_fixtures(repo, "ancestry-projection")[["ancestry_correction"]]
con <- rduckhts_connect()
dbExecute(con, "SET threads=4")
dbExecute(con, paste(
  "CREATE TABLE bench_ancestry_ref AS SELECT 'chr1' AS chromosome,",
  "(100000+i)::INTEGER AS position, 'A' AS allele_a, 'G' AS allele_b,",
  "g.group_id, CASE WHEN g.group_id='A' THEN 0.25 + 0.16*sin(i*0.017)",
  "ELSE 0.68 + 0.12*cos(i*0.019) END AS frequency",
  "FROM range(17000) t(i), (VALUES ('A'),('B')) g(group_id)"
))
dbExecute(con, paste(
  "CREATE TABLE bench_ancestry_pc AS SELECT DISTINCT chromosome, position, allele_a, allele_b,",
  "pc, CASE WHEN pc=1 THEN sin(position*0.031) ELSE cos(position*0.027) END AS loading",
  "FROM bench_ancestry_ref, (VALUES (1),(2)) t(pc)"
))
dbExecute(con, sprintf(
  "CREATE TABLE bench_ancestry_correction AS SELECT pc::INTEGER AS pc, coefficient::DOUBLE AS coefficient FROM read_csv_auto(%s, delim='\\t', header=true) WHERE pc<=2",
  as.character(dbQuoteString(con, correction_file))
))
dbExecute(con, paste(
  "CREATE TABLE bench_ancestry_input AS SELECT sample_id, chromosome, position, allele_a, allele_b,",
  "sum(CASE WHEN group_id='A' THEN a_share ELSE 1-a_share END * frequency)",
  "+ 0.002*sin(position * sample_number) AS frequency",
  "FROM bench_ancestry_ref, (VALUES ('S1',0.7,1),('S2',0.4,2),('S3',0.1,3))",
  "s(sample_id,a_share,sample_number)",
  "GROUP BY sample_id, sample_number, chromosome, position, allele_a, allele_b"
))
peak_rss_mb <- function() {
  status <- grep("^VmHWM:", readLines("/proc/self/status"), value = TRUE)
  as.numeric(strsplit(trimws(status), "[[:space:]]+")[[1L]][[2L]]) / 1024
}
baseline_rss_mb <- peak_rss_mb()
run_sites <- function(n, repeat_id) {
  dbExecute(con, sprintf("CREATE OR REPLACE TEMP VIEW bench_ancestry_slice AS SELECT * FROM bench_ancestry_input WHERE position < %d", 100000 + n))
  start <- proc.time()[["elapsed"]]
  out <- rduckhts_ancestry_proportions(
    con, "bench_ancestry_slice", "bench_ancestry_ref", "bench_ancestry_pc",
    "bench_ancestry_correction", min_cor = 0.4
  )
  elapsed <- proc.time()[["elapsed"]] - start
  stopifnot(nrow(out) == 6L, all(out$used_variants == n))
  data.frame(samples = 3L, variants_per_sample = n, input_rows = 3L*n,
             output_group_rows = nrow(out), groups = 2L, pcs = 2L,
             threads = 4L, repeat_id = repeat_id, elapsed_seconds = elapsed,
             process_peak_rss_mb = peak_rss_mb(),
             failures = paste(unique(out$status[out$status != "ok"]), collapse = ";"))
}
invisible(run_sites(1000L, 0L))
results <- do.call(rbind, lapply(c(1000L, 5000L, 17000L), function(n) {
  do.call(rbind, lapply(seq_len(3L), function(rep) run_sites(n, rep)))
}))
```

Revision: `r source_revision`. Workload: deterministic biallelic synthetic
reference frequencies and PC loadings; three independently perturbed samples;
4 DuckDB threads; two reference groups; two PCs; `min_cor = 0.4`. Each variant
is present in both groups and PCs, and every input row matches. This report
measures query and materialization together. The public-reference workload is
measured separately below.

The correction vector is staged from the committed, checksummed bigsnpr vignette
product. `VmHWM` is cumulative process high-water RSS (MiB), including package
loading and all preceding cases; the baseline before the first run was
`r round(baseline_rss_mb, 1)` MiB. It is not per-query allocated memory.

```{r results}
knitr::kable(results, digits = 3)
```

Failed gates are retained in the `failures` column for each repeat. The nearest
identical synthetic workload is the earlier rendered
`benchmarks/benchmark_ancestry.md` at repository revision `8ab38b99` (source
revision `56aad9cf`): its three 17,000-site repeats had median 0.078 seconds,
versus `r round(median(results$elapsed_seconds[results$variants_per_sample == 17000L]), 3)`
seconds here. These are separate R processes with uncontrolled background load;
this difference cannot establish a regression or equivalence. There is no
ancestry-projection baseline on `develop`; the synthetic rows are not evidence
about real-reference loading or BAM/CRAM depth sensitivity.

## Public reference and GRCh37 chr22 genotypes

Two checksum-verified Figshare CSV products each contain 5,816,590 rows:
21 reference-frequency groups and 16 PC loadings. A staged, key-sorted Parquet
join holds the four site-key columns, 16 DOUBLE loadings and 21 DOUBLE
frequencies. Its SHA-256 is
`a0bf01ae16db605ff0add59e5ebb747c35454406900c11fd8596f23c167fe02b`;
the staging receipt binds this digest to both registered CSV digests and DuckDB
1.5.5. The one-time conversion is excluded from query times. Converting each
compressed CSV to narrow typed Parquet before the keyed sort keeps the CSV
scanner and sort from holding wide decompressed rows simultaneously; reading
just the compressed header also avoids a full R-side CSV scan.

```{r staging-comparison}
staging <- utils::read.delim("ancestry_stage_comparison.tsv")
stopifnot(nrow(staging) == 3L, all(staging$reference_rows == 5816590L),
          all(staging$pc_columns == 16L), all(staging$group_columns == 21L))
staging$budget_mib <- 2560
staging$within_budget <- staging$peak_rss_mib <= staging$budget_mib
stopifnot(staging$within_budget[staging$implementation == "keyed_narrow_parquet"],
          staging$seconds[staging$implementation == "keyed_narrow_parquet"] <= 30)
knitr::kable(staging, digits = 2)
```

All three staging rows use the same two public CSV products and four threads.
The independently run Astra prototype is an external comparator, not a DuckHTS
revision. Source conversion and digest validation are included in staging RSS;
these figures are not query memory. The epilepsy summary input has 4,880,492
records. A separate public phase-3 chr22 VCF supplies 11 individuals
from AFR (3), EUR (2), EAS (2), SAS (2) and AMR (2). Each workload uses four
DuckDB threads and the bigsnpr vignette's 16 correction coefficients. These
are complete-process peak RSS measurements, including package loading,
CSV-based bigsnpr oracles and preceding operations in that process. The SQL
time includes scanning the staged Parquet view, matching sites and aggregation,
but excludes loading the oracle CSVs.

```{r real-reference}
real_epilepsy <- callr::r(function(repo) {
  setwd(repo)
  Sys.unsetenv("DUCKHTS_CACHE_DIR")
  source("test/scripts/ancestry_bigsnpr_real.R", local = TRUE)
  rss <- grep("^VmHWM:", readLines("/proc/self/status"), value = TRUE)
  list(reference_load_seconds = load_seconds, query_seconds = elapsed,
       peak_rss_mib = as.numeric(strsplit(trimws(rss), "[[:space:]]+")[[1L]][[2L]]) / 1024,
       summary_input = nrow(sumstats), upstream_matched = nrow(matched),
       eligible_chr22 = eligible, matched = nrow(input),
       group_rows = nrow(parity), groups = length(shared),
       cor_pred = result$cor_pred[[1L]], oracle_cor_pred = attr(shared, "cor_pred"),
       rounding_bound = correlation_bound, group_parity = parity)
}, args = list(repo = repo))
real_epilepsy_full <- callr::r(function(repo) {
  setwd(repo)
  Sys.unsetenv("DUCKHTS_CACHE_DIR")
  Sys.setenv(ANCESTRY_EPILEPSY_CHR22_ONLY = "false")
  source("test/scripts/ancestry_bigsnpr_real.R", local = TRUE)
  rss <- grep("^VmHWM:", readLines("/proc/self/status"), value = TRUE)
  list(reference_load_seconds = load_seconds, query_seconds = elapsed,
       peak_rss_mib = as.numeric(strsplit(trimws(rss), "[[:space:]]+")[[1L]][[2L]]) / 1024,
       summary_input = nrow(sumstats), upstream_matched = nrow(matched),
       eligible = eligible, matched = nrow(input), group_rows = nrow(parity),
       groups = length(shared), cor_pred = result$cor_pred[[1L]],
       oracle_cor_pred = attr(shared, "cor_pred"),
       rounding_bound = correlation_bound, group_parity = parity)
}, args = list(repo = repo))
real_individuals <- callr::r(function(repo) {
  setwd(repo)
  Sys.unsetenv("DUCKHTS_CACHE_DIR")
  source("test/scripts/ancestry_1000g_real.R", local = TRUE)
  rss <- grep("^VmHWM:", readLines("/proc/self/status"), value = TRUE)
  list(reference_load_seconds = load_seconds, query_seconds = elapsed,
       peak_rss_mib = as.numeric(strsplit(trimws(rss), "[[:space:]]+")[[1L]][[2L]]) / 1024,
       vcf_snvs = nrow(vcf), reference_overlap = nrow(joined),
       samples = length(samples), sites_per_sample = nrow(site),
       input_rows = nrow(input), group_rows = nrow(parity), groups = ncol(ref) - 5L,
       summary = as.data.frame(summary))
}, args = list(repo = repo))
real_timings <- data.frame(
  workload = c("epilepsy (chr22)", "epilepsy (genome-wide)",
               "1000G phase-3 chr22 genotypes"),
  samples = c(1L, 1L, real_individuals$samples),
  variants_per_sample = c(real_epilepsy$matched, real_epilepsy_full$matched,
                          real_individuals$sites_per_sample),
  input_rows = c(real_epilepsy$matched, real_epilepsy_full$matched,
                 real_individuals$input_rows),
  output_group_rows = c(real_epilepsy$group_rows, real_epilepsy_full$group_rows,
                        real_individuals$group_rows),
  groups = c(real_epilepsy$groups, real_epilepsy_full$groups,
             real_individuals$groups),
  threads = 4L,
  reference_rows = 5816590L,
  reference_load_seconds = c(real_epilepsy$reference_load_seconds,
                             real_epilepsy_full$reference_load_seconds,
                             real_individuals$reference_load_seconds),
  sql_seconds = c(real_epilepsy$query_seconds, real_epilepsy_full$query_seconds,
                  real_individuals$query_seconds),
  process_peak_rss_mib = c(real_epilepsy$peak_rss_mib,
                           real_epilepsy_full$peak_rss_mib,
                           real_individuals$peak_rss_mib))
knitr::kable(real_timings, digits = 2)
```

Epilepsy matching audit: bigsnpr matched `r real_epilepsy_full$upstream_matched` /
`r real_epilepsy_full$summary_input` records genome-wide (6 ambiguous SNPs
excluded, 1,576,097 reversed, 0 strand-flipped). DuckHTS matched
`r real_epilepsy_full$matched` / `r real_epilepsy_full$matched` pre-matched
inputs, after matching against the real reference product. The independently
selected, position-unique, unambiguous chr22 subset is
`r real_epilepsy$matched` / `r real_epilepsy$eligible_chr22` eligible chr22
matches, with `r real_epilepsy$matched` / `r real_epilepsy$matched` matched
by DuckHTS. Both genome-wide solvers consume the same allele-aligned sites.

```{r epilepsy-parity}
knitr::kable(real_epilepsy_full$group_parity[, c("group_id", "proportion", "oracle",
  "absolute_difference")], digits = 7)
```

Genome-wide `cor_pred`: DuckHTS
`r sprintf("%.12f", real_epilepsy_full$cor_pred)`, bigsnpr
`r sprintf("%.12f", real_epilepsy_full$oracle_cor_pred)`. The conservative
seven-decimal QP rounding correlation bound is
`r sprintf("%.9f", real_epilepsy_full$rounding_bound)`; the 0.4 gate is
farther away. On chr22, both return `r real_epilepsy$cor_pred` at seven
decimals with group difference
`r max(real_epilepsy$group_parity$absolute_difference)`; its rounding bound is
`r sprintf("%.9f", real_epilepsy$rounding_bound)`. All 21 genome-wide
coefficients agree at seven decimals; the script checks rounded equality.

The individual input is restricted from `r real_individuals$vcf_snvs` unique
biallelic SNVs to `r real_individuals$reference_overlap` same-assembly,
allele-agreeing, unambiguous reference overlaps before spaced site selection.
Both solvers see the same per-sample genotypes. bigsnpr's
imperfect-match warning (`cor_pred < 0.99`) is retained; passing the
0.4 gate does not imply that the frequencies match perfectly. The 11
individuals' 231 coefficients agree with bigsnpr at seven decimals.

```{r individual-parity}
knitr::kable(real_individuals$summary, digits = 7)
```

## Keyed Parquet profile and input scaling

The pre-conversion genome-wide DuckDB JSON profile, on revision
`0a5fd898c2bb8449259e01dd7aff161bb4dc3bc8`, measured 30.27 s and
40,232 MiB peak buffer memory. Its complete R process peaked at 38,783 MiB
RSS while retaining the CSV-based bigsnpr oracle. The 53,491,760-row
materialized `basis` and 1,123,326,960-row `x` join dominate cardinality;
`pred_sites` separately groups 3,343,235 variants after a long-reference
join. The old `list()` calls assemble only the small post-aggregation solver
vectors; there is no per-variant list construction. The profile also shows
hash joins and grouping over the long reference, but DuckDB does not
attribute its single reported peak buffer value to individual operators.
CSV parsing occurs in R before this SQL profile, not in a DuckDB scan
operator. Sorting is limited to the input window partition and final group
ordering.

The keyed query scans projected Parquet columns and aggregates 336
`sum(PC * group_frequency)` values, 16 corrected `sum(PC * aligned_frequency)`
values and group correlations per sample. The 336-element solver vector and
21 group correlations are assembled **after** aggregation. The earlier keyed
plan grouped all 5,816,590 reference loci into a site hash and retained
3,343,235 classified rows while the correlation branch rebuilt aligned
input. Its four-thread fresh-process run took 3.43 s, 1,872 MiB RSS and
3,175 MiB DuckDB peak buffer. The old JSON profile attributes 0.80 s to
the reference grouping and has two large join branches; profile peak memory
is query-wide, not operator-attributed.

Astra's external plan converts CSVs to narrow Parquet, then builds its solver
hash from the 3,343,235 matched inputs, not the 5,816,590 reference sites.
Its solver-only process ran 2.39–2.43 s with 587 MiB RSS and zero spill under
`memory_limit='1GB'`; classification/preparation ran separately. DuckHTS
keeps classification and its complete audit in the measured wrapper call:
it writes a temporary 3,343,235-row, approximately 22.6 MiB aligned Parquet relation,
then materializes only per-sample moments, solver output and correlation.
The single-sample reductions use ungrouped aggregates. For multiple samples,
the 1..N sample index from the small audit table is the group key: a string
`GROUP BY sample_id` led DuckDB to estimate 440,000 groups for 22 samples
and spill 951 MiB with one thread, while the indexed 44-sample query has
zero spill. The temporary Parquet file is removed before returning; its
bytes are explicit bounded intermediate I/O, not a DuckDB spill.

The hash-build input is 95.65 MiB of decoded key/frequency payload for the
full epilepsy run (3,343,235 rows, measured by
`benchmark_ancestry_query_stages.R`). With a fixed 512 MiB R/DuckDB overhead
and 3× this payload, the process-RSS budget is 799 MiB, further capped at
1 GiB under `memory_limit='1GB'`. Its measured 4-thread RSS below is within
both limits. Every stage's JSON profile reported zero temp spill; the largest
query-session DuckDB buffer estimate was 1,218 MiB. DuckDB buffer accounting
is not resident process RSS and must not be added to it. Genotype 44-sample
build payload is 27.69 MiB (880,000 rows, including the four-byte
sample index used for aggregation), for a 595 MiB budget; the
one-thread process measured 476 MiB RSS with zero spill.

`benchmark_ancestry_reference_load.R` scans every numeric column in each
source CSV and the staged Parquet product with DuckDB; neither scan
materializes 5.8 million rows in R. These timings differ from the CSV
`fread()` oracle-loading times above and have their own denominator.

```{r reference-load-comparison}
reference_load <- utils::read.delim("ancestry_reference_load.tsv")
stopifnot(nrow(reference_load) == 2L,
          identical(reference_load$threads, c(1L, 4L)),
          all(reference_load$reference_rows == 5816590L),
          all(reference_load$numeric_columns == 37L))
knitr::kable(reference_load, digits = 2)
```

`benchmark_ancestry_scaling.R` runs three independent R processes per point
with one or four DuckDB threads and the same registry-staged
`ancestry_reference_parquet`. Files are reused from the local cache across
processes; the OS page cache is not cleared between repetitions. Input Parquet is emitted by
`test/scripts/ancestry_bigsnpr_real.R` and `test/scripts/ancestry_1000g_real.R`
from registry-staged `ancestry_epilepsy`, `ancestry_1000g_chr22` and the
1000G sample panel; the scaling driver accepts those output paths as arguments.
Epilepsy 1×/2×/4× contains nested prefixes of 835,809/1,671,618/3,343,235 physical
input variants (one sample). The sample series keeps 20,000 chr22 sites per
person while selecting 11/22/44 distinct phase-3 VCF individuals. The joint
series selects 5,000/10,000/20,000 sites for 11/22/44 individuals: both axes
grow, with 55,000/220,000/880,000 input rows. No sample IDs or genotypes
are cloned. Every input row participates; each group produces one output row.
Query time covers the public wrapper, including reference validation, aligned
file I/O, solver, quality gates and result materialisation, but excludes
one-time input staging, source checksums and bigsnpr oracle runs. Peak RSS
includes R and DuckDB; DuckDB buffer peaks are not process RSS. The aligned
Parquet size is reported separately from DuckDB temporary spill. The
single-thread 1× floor determines whether a workload has a timing verdict;
sub-five-second series are memory profiles only. Each repetition has a 1 GB
DuckDB memory limit and zero DuckDB spill allowance. The declared process-RSS
ceilings are 512 MiB fixed overhead plus three times the full-scale decoded
required-column payload: 798.95 MiB for epilepsy (95.65 MiB decoded), and
595.07 MiB for both sample and joint cases (27.69 MiB decoded), capped at
1 GiB. These ceilings apply to every point, including the joint series. The keyed
reference scans 5,816,590 loci with 16 DOUBLE PC and 21 DOUBLE group columns
(296 decoded numeric bytes per row plus four keys). The admitted aligned
relation holds at most the physical input row count: 3,343,235 dense
summary-statistic sites or 880,000 joint genotype rows, not
sites × PCs × groups. The largest per-sample aggregate consumes 3,343,235
sites and emits 16 × (21 + 1) numeric moments; each fitted sample returns
21 rows. The aligned file is written once and scanned for moments and
prediction; its reported size is logical file I/O, not DuckDB spill or
measured physical disk traffic.
An optional single-run TEMP-table probe retained identical proportions and
zero spill, but its full-epilepsy peak was approximately 865 MiB, above the
798.95 MiB ceiling. It is not a repeated comparable workload; the measured
path here keeps the aligned Parquet file. The 21 full-epilepsy
coefficients and 231 individual coefficients agree with bigsnpr at seven decimals.

```{r parquet-scaling}
scaling_before <- utils::read.delim("ancestry_memory_scaling_before.tsv")
prior_scaling <- utils::read.delim("ancestry_memory_scaling.tsv")
repetitions <- utils::read.delim("ancestry_memory_repetitions.tsv")
stopifnot(nrow(repetitions) == 54L, nrow(scaling_before) == 12L,
          nrow(prior_scaling) == 12L,
          all(repetitions$threads %in% c(1L, 4L)),
          all(repetitions$scale %in% c(1L, 2L, 4L)),
          all(repetitions$input_rows == repetitions$used_variants),
          all(repetitions$output_rows == repetitions$samples * 21L),
          all(repetitions$peak_temp_mib == 0),
          all(repetitions$peak_rss_mib <= repetitions$budget_mib))
points <- split(repetitions, interaction(repetitions$workload,
                                        repetitions$threads, repetitions$scale,
                                        drop = TRUE))
stopifnot(length(points) == 18L,
          all(vapply(points, function(point)
            identical(sort(point$repetition), 1:3), logical(1L))))
scaling <- do.call(rbind, lapply(points, function(point) data.frame(
  workload = point$workload[1L], threads = point$threads[1L],
  scale = point$scale[1L], samples = point$samples[1L],
  input_rows = point$input_rows[1L], output_rows = point$output_rows[1L],
  decoded_mib = point$decoded_required_column_mib[1L],
  median_s = median(point$seconds), iqr_s = IQR(point$seconds),
  min_s = min(point$seconds), max_s = max(point$seconds),
  max_rss_mib = max(point$peak_rss_mib),
  max_buffer_mib = max(point$peak_buffer_mib),
  aligned_file_mib = max(point$aligned_file_mib),
  spill_mib = max(point$peak_temp_mib), budget_mib = point$budget_mib[1L])))
scaling <- scaling[order(scaling$workload, scaling$threads, scaling$scale), ]
scaling$exponent <- NA_real_
for (workload in unique(scaling$workload)) {
  for (threads in c(1L, 4L)) {
    rows <- which(scaling$workload == workload & scaling$threads == threads)
    scaling$exponent[rows[-1L]] <- log(scaling$median_s[rows[-1L]] /
                                     scaling$median_s[rows[-length(rows)]]) /
      log(scaling$input_rows[rows[-1L]] /
          scaling$input_rows[rows[-length(rows)]])
  }
}
meets_floor <- tapply(scaling$median_s[scaling$threads == 1L & scaling$scale == 1L],
                scaling$workload[scaling$threads == 1L & scaling$scale == 1L],
                function(seconds) seconds >= 5)
scaling$timing_verdict <- ifelse(meets_floor[scaling$workload], "timed", "memory only")
stopifnot(all(scaling$max_rss_mib <= scaling$budget_mib),
          all(scaling$spill_mib == 0))
knitr::kable(scaling_before, digits = 2, caption =
  "Keyed-query baseline at 29f3e695; genotype 22/44 repeated sample IDs")
knitr::kable(prior_scaling, digits = 2, caption =
  "Single-run bounded-engine baseline at 58fb3264; reference admission contract differs")
knitr::kable(scaling, digits = 2, caption =
  "Three fresh processes per point; spread is IQR and range; exponent is log median-time ratio / log input-row ratio to preceding scale")
comparison <- merge(
  scaling[scaling$threads == 4L & scaling$scale == 4L &
          scaling$workload %in% c("epilepsy", "genotypes"),
          c("workload", "input_rows", "median_s", "max_rss_mib")],
  prior_scaling[prior_scaling$threads == 4L & prior_scaling$scale == 4L,
                c("workload", "input_rows", "seconds", "peak_rss_mib")],
  by = c("workload", "input_rows"))
stopifnot(nrow(comparison) == 2L)
comparison$extra_seconds <- comparison$median_s - comparison$seconds
comparison$extra_rss_mib <- comparison$max_rss_mib - comparison$peak_rss_mib
knitr::kable(comparison, digits = 2,
  caption = "Nearest identical input baseline (single run, 58fb3264) versus repeated checked-reference wrapper; absolute differences are not a no-regression claim")
```

Full-reference duplicate and finite-column admission scans account for additional
fixed work across all input scales, especially the smaller sample runs. The
comparison uses the same staged input denominators but different validation
contracts; the latency/RSS trade-off requires explicit review before merging.

## Indexed 30x CRAM sensitivity

These retained measurements were produced at
`0a5fd898c2bb8449259e01dd7aff161bb4dc3bc8` with the existing long-form
BAM ancestry path, not remeasured for the keyed Parquet projection.
Three public 30x CRAMs (NA18507/AFR, HG00188/EUR, HG00403/EAS) were
read via their indexes at chr22:15,335,303–50,800,284 (GRCh38). The
registered GRCh37-to-GRCh38 chain maps 19,981/20,000 selected GRCh37
loci onto chr22; 430 mapped loci require reference swaps. The 19,551
unswapped, unique SNVs provide a spaced 17,000-site panel. No whole
CRAM was downloaded. Each 25% read subsample uses samtools seed 42;
median depth and allele-read totals are measured after downsampling.
Each mode sees the same `min_depth = 7` criterion and the same VCF site
subset. Each query emits 21 group rows with four DuckDB threads and
two htslib workers. Peak RSS is the R process high-water mark (MiB)
after reference loading and regional CRAM staging; it excludes samtools
child-process RSS and is not incremental SQL memory. The `seconds`
column is the ancestry BAM query alone, excluding CRAM staging and
depth audit.
The native and quarter-depth columns retain every status, including
failed gates and their `cor_pred` values.

```{r cram-sensitivity}
cram_sensitivity <- utils::read.delim("ancestry_30x_sensitivity.tsv",
  stringsAsFactors = FALSE)
stopifnot(nrow(cram_sensitivity) == 36L,
  all(cram_sensitivity$sites %in% c(1000L, 5000L, 17000L)),
  length(unique(cram_sensitivity$sample_id)) == 3L,
  all(c("allele_fraction", "called_genotype") %in% cram_sensitivity$mode),
  nrow(unique(cram_sensitivity[, c("sites", "panel_sha256")])) == 3L,
  all(with(cram_sensitivity, gate_distance[is.finite(gate_distance)] >
           rounding_bound[is.finite(gate_distance)])))
knitr::kable(unique(cram_sensitivity[, c("sites", "panel_sha256")]),
             row.names = FALSE)
knitr::kable(cram_sensitivity[, c("sample_id", "sites", "depth", "mode",
  "vcf_used", "bam_used", "depth_eligible", "median_depth", "allele_reads",
  "vcf_cor_pred", "cor_pred", "status", "max_group_difference",
  "rounding_bound", "gate_distance", "seconds", "process_peak_rss_mib")], digits = 4)
```

`r sum(cram_sensitivity$status != "ok")` / `r nrow(cram_sensitivity)`
correlation gates failed. The minimum distance from the 0.4 gate among
the measurable cases is `r min(cram_sensitivity$gate_distance, na.rm = TRUE)`;
the largest
coefficient-rounding correlation bound across the three site counts is
`r max(cram_sensitivity$rounding_bound)`.
Called-genotype frequency excludes sites whose balance rule makes no call;
its `bam_used` denominator is therefore smaller than the depth-eligible
count. Maximum group differences compare the same-site VCF result with
each BAM/CRAM mode; they are sensitivity, not an equality assertion.

```{r cleanup, include=FALSE}
dbDisconnect(con, shutdown = TRUE)
unlink(cache, recursive = TRUE)
```
