---
title: "DuckHTS worker-local reference cache"
output: github_document
---

<!-- benchmark_reference_cache.md is generated from benchmark_reference_cache.Rmd. -->

This paired benchmark exercises `bcftools_norm_row` and `bcftools_munge_row`
against the full, uncompressed Broad hg38 v0 reference from the `duckhtsbench`
registry (`liftover_grch38_fasta`). The registered 3,249,912,778-byte size and
MD5 are validated before use. Input rows are synthetic SNVs at the first
1,000,000 A/C/G/T positions starting at chr22:20,000,000; each ALT is C for
reference A, otherwise A. Each row requests one reference base and must return
one unchanged, correctly oriented row. Every timed run checks positions,
alleles, status/filter, row count, and the input-ID sum, not just throughput.

Run from the repo root after building both revisions with the same toolchain:

```sh
REFERENCE_CACHE_BASELINE_EXTENSION=/path/to/baseline/duckhts.duckdb_extension \
REFERENCE_CACHE_BASELINE_REV=<commit> \
Rscript -e 'rmarkdown::render("benchmarks/benchmark_reference_cache.Rmd", knit_root_dir=getwd())'
```

The candidate defaults to `build/release/duckhts.duckdb_extension`. Stage the
registered reference explicitly with `duckhts_bench_fetch("liftover_grch38_fasta")`
before rendering. No benchmark download or alternate cache layout is embedded
here. Linux is required for peak RSS and the optional htslib interposition probe.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(".", mustWork = TRUE)
Sys.setenv(DUCKHTSBENCH_REGISTRY = file.path(root, "r/duckhtsbench/inst/benchmark_registry.tsv"))
source("r/duckhtsbench/R/registry.R")
source("r/duckhtsbench/R/stage.R")
reference <- duckhts_bench_artifact_path("liftover_grch38_fasta")
stopifnot(file.exists(reference))
duckhts_bench_validate_identity("liftover_grch38_fasta", reference)
extensions <- c(
  baseline = Sys.getenv("REFERENCE_CACHE_BASELINE_EXTENSION"),
  candidate = Sys.getenv("REFERENCE_CACHE_CANDIDATE_EXTENSION", "build/release/duckhts.duckdb_extension"))
extensions <- vapply(extensions, normalizePath, character(1), mustWork = TRUE)
revisions <- c(baseline = Sys.getenv("REFERENCE_CACHE_BASELINE_REV"),
  candidate = system2("git", c("rev-parse", "HEAD"), stdout = TRUE))
stopifnot(nzchar(revisions["baseline"]))
if (system2("git", c("diff", "--quiet", "HEAD", "--", "src")) != 0L)
  stop("Commit candidate extension sources before recording revision-labelled measurements")
rows <- as.integer(Sys.getenv("REFERENCE_CACHE_ROWS", "1000000"))
runs <- as.integer(Sys.getenv("REFERENCE_CACHE_RUNS", "3"))
stopifnot(rows > 0L, rows <= 1000000L, runs > 0L)
directory <- tempfile("duckhts_reference_bench_")
dir.create(directory)
probe <- file.path(directory, "metrics.so")
compile <- c("-std=c11", "-O2", "-Wall", "-Wextra", "-Werror", "-fPIC", "-shared",
  "-Ithird_party/htslib", "benchmarks/reference_cache_metrics.c", "-ldl", "-pthread", "-o", shQuote(probe))
stopifnot(system2(Sys.getenv("CC", "cc"), compile) == 0L)
run_case <- function(revision, operator, threads, measured) {
  output <- file.path(directory, "result.tsv")
  env <- c(paste0("REFERENCE_CACHE_EXTENSION=", shQuote(extensions[revision])))
  if (measured) env <- c(env, paste0("REFERENCE_CACHE_METRICS=", shQuote(probe)),
    paste0("LD_PRELOAD=", shQuote(probe)))
  args <- c("benchmarks/reference_cache_run.R", operator, threads, shQuote(reference),
    rows, if (measured) 1L else runs, shQuote(output))
  log <- system2(file.path(R.home("bin"), "Rscript"), args, env = env, stdout = TRUE, stderr = TRUE)
  status <- attr(log, "status")
  if (!is.null(status) && status != 0L) stop(paste(log, collapse = "\n"))
  data.frame(revision = revision, operator = operator, threads = threads,
    read.delim(output, check.names = FALSE))
}
# Alternate baseline/candidate within each workload; never time with interposition.
timings <- counters <- list()
for (operator in c("norm", "munge")) for (threads in c(1L, 4L)) {
  for (revision in names(extensions)) {
    key <- paste(operator, threads, revision)
    timings[[key]] <- run_case(revision, operator, threads, FALSE)
    counters[[key]] <- run_case(revision, operator, threads, TRUE)
  }
}
timings <- do.call(rbind, timings)
counters <- do.call(rbind, counters)
for (operator in c("norm", "munge"))
  counters <- rbind(counters, run_case("candidate", operator, 8L, TRUE))
row.names(timings) <- row.names(counters) <- NULL
```

## Revisions and environment

```{r identity}
print(data.frame(revision = names(revisions), commit = unname(revisions),
  extension_md5 = unname(tools::md5sum(extensions))), row.names = FALSE)
cat(sprintf("R: %s\nDuckDB: %s\n", R.version.string, as.character(packageVersion("duckdb"))))
cat(system2("uname", "-srmo", stdout = TRUE), "\n")
cat(system2("sh", c("-c", shQuote("lscpu | sed -n '/Model name:/p'")), stdout = TRUE), "\n")
cat("Rows per query:", rows, "; measured repetitions:", runs, "\n")
cat(system2("clang", "--version", stdout = TRUE)[1], "\n")
cat(system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE), "\n")
```

## Timings

Each function/thread/revision combination uses a fresh R process; repetitions
within that process reuse worker caches when DuckDB retains workers. Table
construction, FASTA staging, indexing, and R startup are excluded from elapsed
and CPU time. CPU is user + system process CPU during the query. Peak RSS is
the **whole-process high-water mark**, including setup, not cache-only memory.
The operating-system page cache is warm; this is not a cold-storage benchmark.

```{r timing}
summary <- aggregate(cbind(elapsed, cpu, peak_rss_kib) ~ revision + operator + threads,
  timings, median)
names(summary)[4:6] <- c("median_elapsed_s", "median_cpu_s", "median_peak_rss_kib")
print(summary, row.names = FALSE, digits = 6)
print(timings[c("revision", "operator", "threads", "run", "elapsed", "cpu")], row.names = FALSE)
```

## Reference work, counted separately

These are separate untimed runs with a benchmark-only interposition library;
counter overhead is absent from the timing table. All rows take one reference
lookup. Thus a cache hit is `input rows - htslib fetches`. Fetched bases include
read-ahead. One FASTA key is used per worker, so there are no capacity evictions;
handle destruction can occur when DuckDB tears down workers. The independent
native test, rather than this locality workload, exercises more than eight
references and verifies destruction, retained-window limits, and exact bytes.
The candidate also runs at eight requested workers on the same full plain
FASTA as a regression stress case; handle-load counts expose actual workers.

```{r counters}
print(counters[c("revision", "operator", "threads", "rows", "output_rows", "correct_rows",
  "fetches", "fetched_bases", "cache_hits", "handle_loads", "handle_destroys")], row.names = FALSE)
```

## Scope

The nearest earlier reports are [normalization](benchmark_norm.md) and
[munging](benchmark_munge.md). They use different workloads, so their timings
are not an identical-input baseline for this change. The paired baseline above
rebuilds the pre-change code with the candidate toolchain and dependencies.
This measures clustered SNVs on a real plain FASTA, not arbitrary indels,
random genome-wide access, remote transport, BGZF throughput, or BAM generation.
No result here establishes a general "no regression" or "fastest" claim.

The first post-transport-policy run is retained in
[benchmark_reference_cache.md](benchmark_reference_cache.md). It recorded slower
one-worker munging and faster four-worker munging. A fixed-CPU-placement repeat,
when present as `benchmark_reference_cache_affinity.md`, is a separate experiment,
not a replacement for that result. Its process affinity is recorded above;
all subprocesses inherit it. Neither experiment establishes the cause of a
difference by itself.

The shared cache retains at most eight handles and eight 65,536-base sequence
windows **per thread**, plus terminators and metadata. HTSlib index/transport
memory and caller-owned fetch results are separate; remote BGZF tuning can
retain additional per-handle blocks. Tuning remains caller-specific and is part
of cache identity; this patch does not add public I/O options. This is not a
sealed-allocation claim.

Independent cache tests use seed 171, twelve mixed-case/IUPAC references
(alternating plain and BGZF), eight concurrent workers, and 16,000 random
requests checked against retained source bytes. They also test missing-index
creation, exact versus clipped contig ends, explicit FAI/GZI and tuning keys, cache
eviction, caller-copy isolation, requests at and above 65,536 bases, and TLS
destruction. ASan/UBSan and `make test-reference-cache-tsan CC=clang` pass.
The GCC TSAN runtime cannot initialize on this host (unexpected memory mapping);
Clang TSAN passes both the standard probe and a directly HTSlib-linked build.
HTSlib itself is not sanitizer-instrumented in these probes; they do not prove
the absence of every dependency race. SQL/R checks additionally exercise mixed
normalization/munging calls, one million rows, input reversal, and 1/4/8 workers.
