---
title: "Reader performance across initialization compatibility changes"
output: github_document
params:
  baseline_extension: !r Sys.getenv("DUCKHTS_BASELINE_EXTENSION")
  candidate_extension: !r Sys.getenv("DUCKHTS_EXTENSION")
  baseline_revision: "183ae5fc5f464e68c697b990406a26a5d8138b51"
  candidate_revision: "e602cd66796c9f1c17dce8f27cbf3fce5f779363"
  repetitions: 7
  measure: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r measure, include=FALSE, eval=params$measure}
library(DBI)
library(duckdb)
library(duckhtsbench)
repetitions <- as.integer(params$repetitions)
stopifnot(repetitions >= 3L)
extensions <- setNames(
  normalizePath(c(params$baseline_extension, params$candidate_extension), mustWork = TRUE),
  c("baseline", "candidate")
)
inputs <- data.frame(
  workload = c("BAM", "FASTQ.gz", "FASTA", "VCF.gz", "BCF"),
  artifact = c("ont_ecoli_k12_bam", "ont_ecoli_k12_reads_fastq_gz",
               "ont_ecoli_k12_reference_fna", "multi_region_vcf_gz", "multi_region_bcf"),
  reader = c("read_bam", "read_fastq", "read_fasta", "read_bcf", "read_bcf"),
  expected_rows = c(27377, 25950, 1, 2000000, 2000000),
  options = c(
    ", scan_mode := 'sequential', decompression_threads := 0",
    "", "",
    ", scan_mode := 'sequential', decompression_threads := 0",
    ", scan_mode := 'sequential', decompression_threads := 0"
  )
)
paths <- vapply(inputs$artifact, duckhts_bench_artifact_path, character(1))
stopifnot(all(file.exists(paths)))
inputs$bytes <- file.info(paths)$size
inputs$sha256 <- vapply(paths, digest::digest, character(1), algo = "sha256", file = TRUE)
metadata <- data.frame(
  build = names(extensions),
  source_revision = c(params$baseline_revision, params$candidate_revision),
  extension_sha256 = vapply(extensions, digest::digest, character(1), algo = "sha256", file = TRUE),
  duckdb = as.character(packageVersion("duckdb")),
  R = R.version.string,
  threads = 1L,
  measured_repetitions = repetitions,
  run_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  cpu_affinity = sub("^Cpus_allowed_list:\\s+", "",
                    grep("^Cpus_allowed_list:", readLines("/proc/self/status"), value = TRUE)),
  host_system = paste(Sys.info()[c("sysname", "release", "machine")], collapse = " / "),
  cpu_model = sub("^model name\\s*:\\s*", "",
                  grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[[1]])
)
scratch <- tempfile("init-reader-benchmark-")
dir.create(scratch)

# One extension per fresh process. Timing includes all-column CTAS, not checksums
# or Parquet output. No COUNT-only projection can replace decoding the records.
measure <- function(extension, input, reader, options, expected_rows, snapshot) {
  library(DBI)
  library(duckdb)
  con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true"),
                         shared_home = FALSE))
  on.exit(dbDisconnect(con, shutdown = TRUE))
  load_time <- system.time(dbExecute(con, paste("LOAD", dbQuoteString(con, extension))))
  dbExecute(con, "SET threads=1")
  dbExecute(con, "SET memory_limit='8GB'")
  query <- sprintf("CREATE TABLE records AS SELECT * FROM %s(%s%s)",
                   reader, dbQuoteString(con, input), options)
  elapsed <- system.time(dbExecute(con, query))
  counts <- dbGetQuery(con, paste(
    "SELECT count(*) AS rows, sum(hash(r)::HUGEINT)::VARCHAR AS checksum",
    "FROM records r"
  ))
  stopifnot(as.numeric(counts$rows) == expected_rows)
  schema <- dbGetQuery(con, "DESCRIBE records")
  if (nzchar(snapshot)) {
    dbExecute(con, sprintf("COPY records TO %s (FORMAT PARQUET)",
                          dbQuoteString(con, snapshot)))
  }
  data.frame(rows = as.numeric(counts$rows), columns = nrow(schema),
             checksum = counts$checksum,
             schema_sha256 = digest::digest(schema, algo = "sha256"),
             elapsed = unname(elapsed["elapsed"]),
             load_elapsed = unname(load_time["elapsed"]))
}

results <- list()
snapshots <- matrix("", nrow(inputs), length(extensions),
                    dimnames = list(inputs$workload, names(extensions)))
for (i in seq_len(nrow(inputs))) {
  for (iteration in seq_len(repetitions)) {
    # Alternating order limits systematic drift between the two builds.
    order <- if (iteration %% 2L) names(extensions) else rev(names(extensions))
    for (build in order) {
      snapshot <- if (iteration == 1L) file.path(scratch, paste0(i, "-", build, ".parquet")) else ""
      if (nzchar(snapshot)) snapshots[i, build] <- snapshot
      result <- callr::r(measure, list(extensions[[build]], paths[[i]], inputs$reader[[i]],
                                      inputs$options[[i]], inputs$expected_rows[[i]], snapshot))
      results[[length(results) + 1L]] <- cbind(
        data.frame(workload = inputs$workload[[i]], build, iteration), result
      )
    }
  }
}
results <- do.call(rbind, results)

# Compare complete first-run multisets in both directions, retaining duplicates.
con <- dbConnect(duckdb(shared_home = FALSE))
multiset_difference <- function(con, a, b) {
  query <- sprintf(paste(
    "SELECT count(*) AS n FROM ((SELECT * FROM %s EXCEPT ALL SELECT * FROM %s)",
    "UNION ALL (SELECT * FROM %s EXCEPT ALL SELECT * FROM %s)) differences"
  ), a, b, b, a)
  as.numeric(dbGetQuery(con, query)$n)
}
proofs <- list()
for (i in seq_len(nrow(inputs))) {
  a <- paste0("read_parquet(", dbQuoteString(con, snapshots[i, "baseline"]), ")")
  b <- paste0("read_parquet(", dbQuoteString(con, snapshots[i, "candidate"]), ")")
  mismatches <- multiset_difference(con, a, b)
  group <- results[results$workload == inputs$workload[[i]], ]
  stopifnot(mismatches == 0, length(unique(group$checksum)) == 1L,
            length(unique(group$schema_sha256)) == 1L)
  proofs[[i]] <- data.frame(workload = inputs$workload[[i]], mismatches,
                            rows = group$rows[[1]], columns = group$columns[[1]],
                            checksum = group$checksum[[1]])
}
# The same multiset comparison must detect a missing duplicate record.
stopifnot(multiset_difference(con, "(VALUES (1),(1)) a(x)", "(VALUES (1)) b(x)") == 1,
          multiset_difference(con, "(VALUES (1)) a(x)", "(VALUES (1),(1)) b(x)") == 1,
          multiset_difference(con, "(VALUES (NULL),(1)) a(x)", "(VALUES (1),(1)) b(x)") == 2)
dbDisconnect(con, shutdown = TRUE)
unlink(scratch, recursive = TRUE)
proofs <- do.call(rbind, proofs)
write.csv(results, "data/init_readers_timings.csv", row.names = FALSE)
write.csv(inputs, "data/init_readers_inputs.csv", row.names = FALSE)
write.csv(metadata, "data/init_readers_metadata.csv", row.names = FALSE)
write.csv(proofs, "data/init_readers_proofs.csv", row.names = FALSE)
```

```{r tables, include=FALSE}
results <- read.csv("data/init_readers_timings.csv", colClasses = c(checksum = "character"))
inputs <- read.csv("data/init_readers_inputs.csv")
metadata <- read.csv("data/init_readers_metadata.csv")
proofs <- read.csv("data/init_readers_proofs.csv", colClasses = "character")
repetitions <- unique(metadata$measured_repetitions)
stopifnot(length(repetitions) == 1L)
summary <- aggregate(elapsed ~ workload + build, results, median)
comparison <- merge(summary[summary$build == "baseline", c("workload", "elapsed")],
                    summary[summary$build == "candidate", c("workload", "elapsed")],
                    by = "workload", suffixes = c("_baseline", "_candidate"))
comparison$ratio <- comparison$elapsed_candidate / comparison$elapsed_baseline
```

## Scope and method

This is a matched check of five reader/format combinations on one host, with
**one DuckDB thread**, automatic SIMD dispatch, and no htslib decompression
workers for BAM/VCF/BCF. Each observation starts a fresh R process and database.
All reader columns are decoded and materialized by `CREATE TABLE AS SELECT *`;
checksums and comparison snapshots are outside the timed interval. Extension
load time is recorded separately. There are `r repetitions` measured
observations per build/workload, in alternating build order; no cold-cache or
peak-memory claim is made.

Inputs resolve through `duckhtsbench`: the ONT E. coli BAM, its FASTQ reads and
reference, and the registered two-million-record synthetic VCF/BCF. Expected
counts come from the retained ONT staging evidence, reference/read registry
metadata, and synthetic generator contract. They are not inferred from candidate
output. Every repetition must match its expected record count, complete schema,
and all-field checksum. First-run outputs additionally undergo exact multiset
comparison in both directions, including duplicate rows.

This workload does not measure DuckVEP, CRAM, indexed interval selection, or
reader scaling across thread counts. The nearest broader reader baselines are
[BCF record caching](benchmark_bcf_record_cache.md),
[multi-region readers](benchmark_multi_region_readers.md), and
[FASTQ reading](benchmark_fastq_reader.md); their inputs/projections differ, so
the matched baseline in this report is the performance comparator.

## Artifact identity

```{r metadata}
knitr::kable(metadata[setdiff(names(metadata), c("host_system", "cpu_model"))])
```

```{r host}
cat(metadata$host_system[[1]], "\n", metadata$cpu_model[[1]], "\n", sep = "")
```

## Inputs and output denominators

```{r inputs}
knitr::kable(inputs[c("workload", "artifact", "expected_rows", "bytes", "sha256")])
knitr::kable(proofs)
```

## Matched reader timing

Median seconds and candidate/baseline ratios. The single-record FASTA has
millisecond-scale variation and limited timing resolution; its 14–15 ms range
is not evidence of a general reader speed difference.

```{r timing}
knitr::kable(comparison, digits = 3)
```

```{r distribution}
distribution <- aggregate(elapsed ~ workload + build, results,
                          function(x) c(min = min(x), median = median(x), max = max(x)))
knitr::kable(do.call(data.frame, distribution), digits = 3)
```

```{r load}
knitr::kable(aggregate(load_elapsed ~ build, results, median), digits = 3)
```

Raw timings, source/binary identities, input hashes and equality proofs are kept
in `data/init_readers_{timings,metadata,inputs,proofs}.csv`. The default render
measures and validates the workload; `measure: false` renders those recorded
CSVs without executing readers or creating new measurements.
