---
title: "Count error fit scaling"
output:
  github_document:
    html_preview: false
    pandoc_args: ["--wrap=preserve", "--columns=400"]
---

This report gives the 1×/2×/4× scaling evidence `STYLE.md` asks for, for
`duckhts_count_error_fit()`, and its first measurement on real counts. The
function fits read error, contamination and five related numbers of each
sample from allele read counts at sites of known frequency. DuckDB builds a
histogram of (block, frequency bin, depth, alt count) cells; the native code
packs a sample's cells into one bounded buffer and fits the model on it. So
the work has two parts: the histogram, which grows with the rows, and the
fit, which grows with the cells and not with the rows. No earlier report
measures this function; this is its first baseline.

## Workloads

- **Sites.** One synthetic sample of 2, 4 and 8 million sites, depth 20 to
  40, Hardy-Weinberg genotypes at frequencies 0.15 to 0.85 with a second
  genome at 2%, read error 1e-3. The positions are 100 bases apart, so the
  sample spans 200, 400 and 800 Mb and has 20, 40 and 80 blocks of 10 Mb.
- **Samples.** 8, 16 and 32 synthetic samples of 250,000 sites each, the same
  model with the second genome at 2%, each sample spanning 25 Mb and three
  blocks. The buffers of the samples share one byte budget.
- **Real.** The three staged chr20 count tables of 1000 Genomes samples
  (`roh_counts_chr20_na18507`, `roh_counts_chr20_hg00403`,
  `roh_counts_chr20_hg00188`: `duckhts_somalier_bam_counts` at the chr20
  sites of the ancestry panel, about 32× each), checked against their
  registered identities before anything is measured. One chromosome of one
  sample fits in under the 5 s floor that `STYLE.md` sets for a timing
  verdict, so this workload is measured at its one size, three samples, and
  carries no scaling verdict: it records what the function costs on real
  counts and what it finds in them. A genome is 45 times one chromosome and
  is not measured here.

The synthetic inputs are generated untimed at render time, with every draw a
hash of its coordinates, and the number of histogram cells of each input is
counted untimed: it is the input denominator of the fit.

Every observation is one fresh DuckDB CLI process under `/usr/bin/time`,
pinned to eight cores by the render command and run at scheduling priority
−15, with three repetitions at one thread and at four threads in an order
that alternates sizes. The JSON profiler gives the query latency and DuckDB's
peak buffer and temporary bytes; `/usr/bin/time` gives the peak RSS.
`max_temp_directory_size` is zero, so a spill fails the render; the memory
limit is DuckDB's default, which the report records. Each query consumes
every row of the fit. One fit runs on one thread; the fits of a query's
samples run in the threads that hold their rows, so the four-thread cells
measure both the histogram and that spread of the samples over threads.

## Budget, declared before measuring

The overhead is the median peak RSS of three empty runs that start the CLI,
load the extension and scan nothing.

The ceiling is overhead + 3 × the live state of the largest input of each
workload, and it is a gate on the peak RSS: the render stops if any
observation's peak RSS exceeds it or if any observation spills. DuckDB's
peak buffer count is its own accounting of what its buffer manager held at
most; it is reported beside the RSS and not gated. The live state of one
query is, summed over its samples:

- the packed cells, 16 bytes each, held natively while the histogram is
  built and then by DuckDB as one BLOB per sample (two copies at the
  hand-over), plus one copy and the model's pair tables in the fit: 16 × 3
  bytes per cell, plus 72 bytes per distinct (depth, alt) pair;
- DuckDB's hash aggregate that builds the cells from the rows, 160 bytes
  per cell, and the window that numbers the blocks, 64 bytes per cell, both
  of which DuckDB manages and could spill. The rows themselves stream from
  Parquet into the aggregate and are not held.

The cells are counted from the input before measuring, so the budget is a
number for each workload, written below with the measurements.

The render also stops if a time exponent of a synthetic workload is above
1.25 at either doubling, or if its 1× run takes less than 5 s on one thread,
which is the least that `STYLE.md` accepts for a timing verdict.

``` sh
taskset -c 8-15 Rscript -e 'rmarkdown::render("benchmarks/benchmark_count_error_fit.Rmd")'
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r benchmark, echo=FALSE, message=FALSE}
stopifnot(requireNamespace("duckhtsbench", quietly = TRUE),
          requireNamespace("jsonlite", quietly = TRUE),
          requireNamespace("knitr", quietly = TRUE))
Sys.setenv(DUCKHTSBENCH_REGISTRY = normalizePath(
  "r/duckhtsbench/inst/benchmark_registry.tsv", mustWork = TRUE))
extension <- normalizePath(Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension"),
                           mustWork = TRUE)
duckdb_cli <- Sys.getenv("DUCKDB_CLI", Sys.which("duckdb"))
stopifnot(nzchar(duckdb_cli), file.exists("/usr/bin/time"))
duckdb_runtime <- system2(duckdb_cli, "--version", stdout = TRUE)[[1L]]
revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE)
src_tree <- system2("git", c("rev-parse", "HEAD:src"), stdout = TRUE)
if (system2("git", c("diff", "--quiet", "HEAD", "--", "src"), stdout = FALSE, stderr = FALSE) != 0L) {
  stop("Build source must match the recorded HEAD")
}
host_load <- sub(".*load average: ", "", system2("uptime", stdout = TRUE))
quote_literal <- function(value) paste0("'", gsub("'", "''", value, fixed = TRUE), "'")

# One untimed query in its own CLI process; its rows come back as numbers.
query_untimed <- function(sql) {
  output <- system2(duckdb_cli, c("-unsigned", "-csv"),
                    input = c(sprintf("LOAD %s;", quote_literal(extension)), sql), stdout = TRUE)
  utils::read.csv(text = output, colClasses = "numeric")
}
memory_limit <- system2(duckdb_cli, c("-list", "-noheader"),
                        input = "SELECT current_setting('memory_limit');", stdout = TRUE)[[1L]]

# Run one SQL body in a fresh CLI process with the given threads and spill
# forbidden, after an untimed setup statement; the query's single row is
# returned with latency and memory.
run_fresh <- function(query, threads, setup = NULL) {
  scratch <- tempfile("count-error-scaling-")
  dir.create(scratch)
  on.exit(unlink(scratch, recursive = TRUE), add = TRUE)
  sql_path <- file.path(scratch, "query.sql")
  result_path <- file.path(scratch, "result.csv")
  profile_path <- file.path(scratch, "profile.json")
  time_path <- file.path(scratch, "time.tsv")
  writeLines(c(
    sprintf("SET threads = %d;", threads),
    "SET max_temp_directory_size = '0KiB';",
    sprintf("LOAD %s;", quote_literal(extension)),
    setup,
    "PRAGMA enable_profiling = 'json';",
    sprintf("PRAGMA profiling_output = %s;", quote_literal(profile_path)),
    sprintf("COPY (%s) TO %s (HEADER);", query, quote_literal(result_path)),
    "PRAGMA disable_profiling;"
  ), sql_path)
  status <- system2("/usr/bin/time", c("-f", shQuote("%M\t%e"), "-o", shQuote(time_path),
                                       "nice", "-n", "-15", shQuote(duckdb_cli), "-unsigned", "-batch", "-bail"),
                    stdin = sql_path, stdout = FALSE, stderr = FALSE)
  if (status != 0L) stop("DuckDB CLI failed (status ", status, ")")
  profile <- jsonlite::fromJSON(profile_path, simplifyVector = TRUE)
  timing <- utils::read.delim(time_path, header = FALSE, col.names = c("max_rss_kib", "process_seconds"))
  data.frame(utils::read.csv(result_path, colClasses = "character"),
             seconds = as.numeric(profile$latency), max_rss_kib = timing$max_rss_kib[[1L]],
             peak_buffer_bytes = as.numeric(profile$system_peak_buffer_memory),
             peak_temp_bytes = as.numeric(profile$system_peak_temp_dir_size))
}

# Synthetic counts, staged untimed. Every draw is a hash of its coordinates.
staging <- tempfile("count-error-scaling-counts-")
dir.create(staging)
inputs <- data.frame(
  workload = c(rep("sites", 3L), rep("samples", 3L)),
  scale = rep(c(1L, 2L, 4L), 2L),
  samples = c(1L, 1L, 1L, 8L, 16L, 32L),
  sites = c(2000000L, 4000000L, 8000000L, 250000L, 250000L, 250000L),
  stringsAsFactors = FALSE)
inputs$rows <- inputs$samples * inputs$sites
inputs$path <- file.path(staging, sprintf("counts_%s_%d.parquet", inputs$workload, inputs$scale))
for (k in seq_len(nrow(inputs))) {
  generate <- sprintf(paste(
    "COPY (SELECT 'S' || s AS sample_id, '1' AS chrom, i * 100 + 1 AS pos, af, depth - alt AS ref_count, alt AS alt_count",
    "FROM (SELECT s, i, af, depth, (SELECT count(*) FROM range(depth) r(j) WHERE (hash(7, s, i, j) %% 1000003) / 1000003.0 <",
    "0.98 * [1e-3, 0.5, 1 - 1e-3][g + 1] + 0.02 * [1e-3, 0.5, 1 - 1e-3][h + 1]) AS alt",
    "FROM (SELECT s, i, af, 20 + (hash(8, s, i, 0) %% 21)::INTEGER AS depth,",
    "((hash(3, s, i, 0) %% 1000003) / 1000003.0 < af)::INTEGER + ((hash(4, s, i, 0) %% 1000003) / 1000003.0 < af)::INTEGER AS g,",
    "((hash(5, s, i, 0) %% 1000003) / 1000003.0 < af)::INTEGER + ((hash(6, s, i, 0) %% 1000003) / 1000003.0 < af)::INTEGER AS h",
    "FROM (SELECT s, i, 0.15 + 0.7 * ((hash(1, i, 0, 0) %% 1000003) / 1000003.0) AS af",
    "FROM range(0, %d) AS a(s), range(0, %d) AS t(i))))) TO %s (FORMAT parquet);"),
    inputs$samples[[k]], inputs$sites[[k]], quote_literal(inputs$path[[k]]))
  stopifnot(system2(duckdb_cli, "-batch", input = generate, stdout = FALSE) == 0L)
}
inputs$source_sql <- sprintf("read_parquet(%s)", vapply(inputs$path, quote_literal, character(1L)))

# Real counts: registered artifacts, checked against their registered identity.
source("benchmarks/roh_counts_stage.R")
real_ids <- c("roh_counts_chr20_na18507", "roh_counts_chr20_hg00403", "roh_counts_chr20_hg00188")
real_paths <- vapply(real_ids, duckhtsbench::duckhts_bench_artifact_path, character(1L))
identity_con <- DBI::dbConnect(duckdb::duckdb(shared_home = FALSE))
real_rows <- 0
for (k in seq_along(real_ids)) {
  registry <- duckhtsbench::duckhts_bench_registry()
  registered <- duckhtsbench:::duckhts_bench_identity_fields(registry$supplier_identity[match(real_ids[[k]], registry$id)])
  observed_identity <- roh_counts_identity(identity_con, real_paths[[k]])
  stopifnot(observed_identity$rows == as.numeric(registered[["rows"]]),
            identical(observed_identity$counts_sha256, registered[["counts_sha256"]]))
  real_rows <- real_rows + as.numeric(registered[["rows"]])
}
DBI::dbDisconnect(identity_con, shutdown = TRUE)
real_inputs <- data.frame(workload = "real", scale = 1L, samples = 3L, sites = NA, rows = NA,
                          path = NA, stringsAsFactors = FALSE)
real_inputs$source_sql <- sprintf("read_parquet([%s])",
  paste(vapply(real_paths, quote_literal, character(1L)), collapse = ", "))
inputs <- rbind(inputs, real_inputs)

# The denominators of each input, counted untimed: rows the fit uses, the
# histogram cells and the distinct (depth, alt) pairs, with the default
# options of the macro.
denominators <- do.call(rbind, lapply(inputs$source_sql, function(source) query_untimed(sprintf(paste(
  "WITH r AS (SELECT sample_id, chrom, pos // 10000000 AS block_index, floor(af * 64) AS bin,",
  "ref_count + alt_count AS depth, alt_count AS alt FROM %s",
  "WHERE ref_count IS NOT NULL AND alt_count IS NOT NULL AND af > 0 AND af < 1",
  "AND ref_count + alt_count BETWEEN 1 AND 1000)",
  "SELECT count(*) AS rows_used, (SELECT count(*) FROM (SELECT DISTINCT sample_id, chrom, block_index, bin, depth, alt FROM r)) AS cells,",
  "(SELECT count(*) FROM (SELECT DISTINCT sample_id, depth, alt FROM r)) AS pairs,",
  "count(DISTINCT sample_id) AS samples_used FROM r;"), source))))
inputs <- cbind(inputs, denominators)
inputs$rows[is.na(inputs$rows)] <- inputs$rows_used[is.na(inputs$rows)]
inputs$sites[is.na(inputs$sites)] <- inputs$rows_used[is.na(inputs$sites)] / inputs$samples_used[is.na(inputs$sites)]
inputs$live_state_bytes <- inputs$cells * (16 * 3 + 160 + 64) + inputs$pairs * 72

# The macro reads a named table; each run gives it a view of its Parquet input.
fit_query <- paste(
  "SELECT count(*) AS fits, count(*) FILTER (WHERE status = 'ok') AS ok,",
  "round(avg(contamination), 6) AS mean_contamination, sum(sites)::BIGINT AS sites",
  "FROM duckhts_count_error_fit('counts')")
fit_setup <- function(source) sprintf("CREATE TEMP VIEW counts AS SELECT * FROM %s;", source)

empty_runs <- vapply(seq_len(3L), function(run) run_fresh("SELECT 1 AS fits", 1L)$max_rss_kib, numeric(1L))
overhead_mib <- median(empty_runs) / 1024
budgets <- aggregate(live_state_bytes ~ workload, inputs, max)
budgets$budget_mib <- overhead_mib + 3 * budgets$live_state_bytes / 2^20
inputs$budget_mib <- budgets$budget_mib[match(inputs$workload, budgets$workload)]

observations <- list()
for (repetition in 1:3) {
  order <- if (repetition %% 2L == 1L) seq_len(nrow(inputs)) else rev(seq_len(nrow(inputs)))
  for (threads in c(1L, 4L)) {
    for (k in order) {
      observed <- run_fresh(fit_query, threads, fit_setup(inputs$source_sql[[k]]))
      observations[[length(observations) + 1L]] <- data.frame(
        workload = inputs$workload[[k]], scale = inputs$scale[[k]], samples = inputs$samples[[k]],
        rows = inputs$rows[[k]], cells = inputs$cells[[k]], threads = threads, repetition = repetition,
        observed, budget_mib = inputs$budget_mib[[k]], stringsAsFactors = FALSE)
    }
  }
}
runs <- do.call(rbind, observations)
for (column in c("fits", "ok", "mean_contamination", "sites")) runs[[column]] <- as.numeric(runs[[column]])
stopifnot(all(runs$peak_temp_bytes == 0), all(runs$fits == runs$samples), all(runs$ok == runs$fits))
write.csv(runs, "benchmarks/benchmark_count_error_fit_runs.csv", row.names = FALSE)
over_budget <- runs[runs$max_rss_kib / 1024 > runs$budget_mib, ]
if (nrow(over_budget)) {
  print(over_budget[c("workload", "scale", "threads", "max_rss_kib", "budget_mib")])
  stop("observations exceed the declared memory budget", call. = FALSE)
}

cells <- split(runs, list(runs$scale, runs$threads, runs$workload), drop = TRUE)
summary <- do.call(rbind, lapply(cells, function(cell) {
  stopifnot(length(unique(cell$mean_contamination)) == 1L, length(unique(cell$sites)) == 1L)
  data.frame(workload = cell$workload[[1L]], scale = cell$scale[[1L]], threads = cell$threads[[1L]],
             samples = cell$samples[[1L]], rows = cell$rows[[1L]], cells = cell$cells[[1L]],
             mean_contamination = cell$mean_contamination[[1L]],
             median_seconds = median(cell$seconds), min_seconds = min(cell$seconds), max_seconds = max(cell$seconds),
             median_rss_mib = round(median(cell$max_rss_kib) / 1024, 1),
             max_rss_mib = round(max(cell$max_rss_kib) / 1024, 1),
             max_buffer_mib = round(max(cell$peak_buffer_bytes) / 2^20, 1),
             budget_mib = round(cell$budget_mib[[1L]], 1), stringsAsFactors = FALSE)
}))
summary <- summary[order(summary$workload, summary$threads, summary$scale), ]
rownames(summary) <- NULL

synthetic <- summary[summary$workload != "real", ]
exponents <- do.call(rbind, lapply(split(synthetic, list(synthetic$threads, synthetic$workload), drop = TRUE), function(s) {
  s <- s[order(s$scale), ]
  dimension <- if (s$workload[[1L]] == "samples") s$samples else s$rows
  growth <- function(low, high) log(s$median_seconds[[high]] / s$median_seconds[[low]]) / log(dimension[[high]] / dimension[[low]])
  data.frame(workload = s$workload[[1L]], threads = s$threads[[1L]], one_x_seconds = s$median_seconds[[1L]],
             time_exponent_2x = growth(1L, 2L), time_exponent_4x = growth(2L, 3L),
             rss_ratio_2x = s$median_rss_mib[[2L]] / s$median_rss_mib[[1L]],
             rss_ratio_4x = s$median_rss_mib[[3L]] / s$median_rss_mib[[1L]], stringsAsFactors = FALSE)
}))
rownames(exponents) <- NULL
one_thread <- synthetic[synthetic$threads == 1L, ]
stopifnot(all(one_thread$median_seconds[one_thread$scale == 1L] >= 5),
          all(exponents$time_exponent_2x <= 1.25), all(exponents$time_exponent_4x <= 1.25))

# Untimed: the fit of the three real samples, what the function is for.
real_fit <- utils::read.csv(text = system2(duckdb_cli, c("-unsigned", "-csv"), input = c(
  sprintf("LOAD %s;", quote_literal(extension)), fit_setup(real_inputs$source_sql[[1L]]), paste(
  "SELECT sample, sites, round(mean_depth, 1) AS depth, blocks, seq_error, contamination, contamination_sd,",
  "allele_balance, spread_hom, spread_het, artefact_weight, contamination_relative,",
  "round(log_likelihood_relative - log_likelihood, 1) AS relative_minus_unrelated, status",
  "FROM duckhts_count_error_fit('counts') ORDER BY sample;")), stdout = TRUE))

# Untimed: a count-level mixture of two real samples. Each read of the
# receiver is kept with chance 1 - share and each read of the other sample
# is added with chance share, every draw a hash of (position, read); the
# receiver's frequencies are the panel's.
mixture_sql <- c(
  sprintf("LOAD %s;", quote_literal(extension)),
  sprintf("CREATE TEMP TABLE receiver AS SELECT * FROM read_parquet(%s);", quote_literal(real_paths[["roh_counts_chr20_hg00403"]])),
  sprintf("CREATE TEMP TABLE other AS SELECT * FROM read_parquet(%s);", quote_literal(real_paths[["roh_counts_chr20_na18507"]])),
  "CREATE MACRO thin(n, keep, salt, pos) AS (SELECT count(*) FROM range(n) t(i) WHERE (hash(pos, salt, i) % 1000000) / 1000000.0 < keep);",
  paste("CREATE TEMP VIEW counts AS SELECT share AS sample_id, a.chrom, a.pos, a.af,",
        "thin(a.ref_count, 1 - share, 1, a.pos) + thin(b.ref_count, share, 2, a.pos) AS ref_count,",
        "thin(a.alt_count, 1 - share, 3, a.pos) + thin(b.alt_count, share, 4, a.pos) AS alt_count",
        "FROM receiver a JOIN other b USING (chrom, pos), (VALUES (0.0), (0.02), (0.05), (0.20)) v(share)",
        "WHERE a.ref_count IS NOT NULL AND a.alt_count IS NOT NULL AND b.ref_count IS NOT NULL AND b.alt_count IS NOT NULL;"),
  paste("SELECT sample AS share, sites, round(mean_depth, 1) AS depth, seq_error, contamination, contamination_sd,",
        "homozygosity_excess, allele_balance, artefact_weight, contamination_relative,",
        "round(log_likelihood_relative - log_likelihood, 1) AS relative_minus_unrelated, status",
        "FROM duckhts_count_error_fit('counts') ORDER BY CAST(sample AS DOUBLE);"))
mixture_fit <- utils::read.csv(text = system2(duckdb_cli, c("-unsigned", "-csv"), input = mixture_sql, stdout = TRUE))
unlink(staging, recursive = TRUE)
```

Source revision: `r revision`; `src` tree `r src_tree`. DuckDB runtime: `r duckdb_runtime` (CLI). Extension SHA-256: `r strsplit(system2("sha256sum", extension, stdout = TRUE), " ", fixed = TRUE)[[1L]][[1L]]`. Host load at start: `r host_load`. DuckDB memory limit: `r memory_limit` (the default); temporary directory limit: 0.

Overhead: `r round(overhead_mib, 1)` MiB. Ceilings: `r paste(sprintf("%s %s MiB", budgets$workload, round(budgets$budget_mib, 1)), collapse = ", ")`.

## Inputs

`rows_used` are the rows the fit keeps, `cells` the histogram cells they make and `pairs` the distinct (depth, alt count) pairs, all with the default options.

```{r inputs, echo=FALSE}
knitr::kable(inputs[c("workload", "scale", "samples", "sites", "rows_used", "cells", "pairs", "budget_mib")],
             digits = 1, format.args = list(big.mark = ","))
```

## Measurements

`mean_contamination` is the mean fitted share over the samples of the query, the same in every repetition. Times are the query latency of three repetitions.

```{r summary, echo=FALSE}
knitr::kable(summary, digits = 3, format.args = list(big.mark = ","))
```

Growth of the median time of the synthetic workloads against the rows (sites) or the samples (an exponent of 1 is linear), and peak-RSS ratios to 1×:

```{r exponents, echo=FALSE}
knitr::kable(exponents, digits = 3)
```

## Findings

Every observation is within its ceiling and none spills; every synthetic 1× run on one thread takes at least 5 s and no time exponent is above 1.25. The render enforces all of this, and that every fit of every run reports `ok` and the same estimates in every repetition.

```{r findings, echo=FALSE, results="asis"}
sites <- summary[summary$workload == "sites" & summary$threads == 1L, ]
samples <- summary[summary$workload == "samples" & summary$threads == 1L, ]
samples4 <- summary[summary$workload == "samples" & summary$threads == 4L, ]
real <- summary[summary$workload == "real" & summary$threads == 1L, ]
cat(sprintf(paste(
  "One sample of %s sites takes %.1f s on one thread (%.1f s at %s sites) at %s MiB peak RSS;",
  "32 samples of 250,000 sites take %.1f s on one thread and %.1f s on four, at %s MiB and %s MiB.",
  "Four threads do not speed up one sample's fit, and spread the fits of 32 samples over the threads.",
  "The three real chr20 samples take %.1f s on one thread at %s MiB.\n"),
  format(sites$rows[[1L]], big.mark = ",", scientific = FALSE), sites$median_seconds[[1L]], sites$median_seconds[[3L]],
  format(sites$rows[[3L]], big.mark = ",", scientific = FALSE), sites$max_rss_mib[[3L]],
  samples$median_seconds[[3L]], samples4$median_seconds[[3L]], samples$max_rss_mib[[3L]], samples4$max_rss_mib[[3L]],
  real$median_seconds[[1L]], real$max_rss_mib[[1L]]))
```

## The real samples

The fit of the three staged chr20 samples with the default options, untimed. These are single 30× samples with no second genome known, so `contamination` is the share the model attributes to one; `seq_error` is the chance that a read shows the other panel allele at a homozygous site, with site artefacts that `artefact_weight` does not absorb.

```{r real, echo=FALSE}
knitr::kable(real_fit, digits = 6)
```

## A second genome of another population

The same function on a count-level mixture of two of those samples, untimed: HG00403 (CHS) receives reads of NA18507 (YRI) at a share of 0, 0.02, 0.05 and 0.20, each read of the receiver kept with chance 1 − share and each read of the other sample added with chance share, with the receiver's panel frequencies. The fitted `contamination` is the number to compare with `share`. The other sample's genotypes do not follow the receiver's frequencies, which the unrelated model takes them to, so the fit attributes more reads to it than the share: this is the bias the catalog entry of `duckhts_count_error_fit()` names. The sites common to both tables are used, so the share-0 row is the receiver alone on those sites and differs slightly from its row above.

```{r mixture, echo=FALSE}
knitr::kable(mixture_fit, digits = 6)
```
