---
title: "Ancestry panel building: per-contig allocation at scale"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
root <- normalizePath("..", mustWork = TRUE)
runs <- utils::read.delim(file.path(root, "benchmarks", "ancestry_panel_scaling.tsv"))
runs$peak_rss_gib <- runs$peak_rss_kb / 1024^2
keys <- c("implementation", "mode", "quarters")
medians <- aggregate(cbind(seconds, peak_rss_gib, sites, contigs) ~ implementation + mode + quarters,
                     data = runs, FUN = stats::median)
old <- subset(medians, implementation == "old")
new <- subset(medians, implementation == "new")
paired <- merge(old, new, by = c("mode", "quarters"), suffixes = c("_old", "_new"))
paired$time_ratio <- paired$seconds_new / paired$seconds_old
paired$rss_delta_mib <- 1024 * (paired$peak_rss_gib_new - paired$peak_rss_gib_old)
exponents <- do.call(rbind, lapply(split(new, new$mode), function(d) data.frame(
  mode = d$mode[1],
  time_exponent = unname(stats::coef(stats::lm(log(seconds) ~ log(quarters), d))[2]),
  memory_exponent = unname(stats::coef(stats::lm(log(peak_rss_gib) ~ log(quarters), d))[2]))))
stopifnot(nrow(runs) == 36L, all(table(runs$implementation, runs$mode, runs$quarters) == 3L),
          all(runs$sites == 17000L),
          all(paired$time_ratio <= 1.10), all(paired$rss_delta_mib <= 64),
          all(runs$contigs[runs$implementation == "new"] == 22L),
          all(exponents$time_exponent <= 1.25))
```

`rduckhts_ancestry_panel()` caps a panel at `max_sites`. The previous builder kept
the first `max_sites` candidates in genomic order, so a 17,000-site panel over
bigsnpr's reference came entirely from chromosome 1. The current builder gives each
contig one site, shares the rest in proportion to each contig's eligible sites
minus that one, and spreads each contig's sites evenly along it. That adds
per-contig row numbers and counts over every candidate, where the old builder
kept a bounded top-N. This report measures what that costs.

## Method

`benchmark_ancestry_panel_run.R` stages bigsnpr's reference and loadings
(`ancestry_reference_parquet`, 5,816,590 loci, 21 groups, 16 PCs) and keeps 1, 2 or 4
quarters of its loci by hash. Each run expands them into the long frequency and
loading relations the builder reads, builds a committed 17,000-site panel, and
counts it. Every run is a fresh single-thread process; peak RSS comes from GNU time.
There are three repetitions of each implementation, mode and scale, in interleaved
order. `spaced` uses the default 5 kb window spacing; `candidates` offers every
locus as a candidate, the largest input the allocation windows can see. `old` is
the builder at the commit before this change and `new` is the per-contig builder.

## Results

Medians of three runs:

```{r results}
out <- medians[order(medians$mode, medians$quarters, medians$implementation != "old"),
               c(keys, "sites", "contigs", "seconds", "peak_rss_gib")]
out$peak_rss_gib <- round(out$peak_rss_gib, 2)
rownames(out) <- NULL
knitr::kable(out)
```

The new builder against the old one at each scale:

```{r paired}
cmp <- paired[order(paired$mode, paired$quarters),
              c("mode", "quarters", "time_ratio", "rss_delta_mib")]
cmp$time_ratio <- round(cmp$time_ratio, 3)
cmp$rss_delta_mib <- round(cmp$rss_delta_mib, 1)
rownames(cmp) <- NULL
knitr::kable(cmp)
```

Scaling exponents of the new builder over 1, 2 and 4 quarters:

```{r exponents}
exponents$time_exponent <- round(exponents$time_exponent, 2)
exponents$memory_exponent <- round(exponents$memory_exponent, 2)
rownames(exponents) <- NULL
knitr::kable(exponents)
```

## Reading

The allocation adds at most `r sprintf("%.1f", 100 * (max(paired$time_ratio) - 1))`%
to the median time and `r sprintf("%.0f", max(paired$rss_delta_mib))` MiB to peak
memory at any scale, and every new panel spans all 22 autosomes. Time grows
linearly with the reference in both builders.

Peak memory also grows linearly in both builders, to about
`r sprintf("%.1f", max(runs$peak_rss_gib))` GiB at full scale. That growth
predates this change: the builder groups the whole long reference (loci × groups)
to find eligible sites before any capping happens. Reducing it belongs to the
builder's site-eligibility stage and is tracked in
[#293](https://github.com/RGenomicsETL/duckhts/issues/293).
