---
title: "Real-input SQL scaling audit: measured paths and open coverage"
output: github_document
---

<!-- benchmark_sql_scaling_audit.md is rendered from this file. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
real <- utils::read.delim("benchmarks/sql_scaling_audit_real.tsv")
wrappers <- utils::read.delim("benchmarks/sql_scaling_audit_wrappers.tsv")
before <- utils::read.delim("benchmarks/sql_scaling_audit_vcf_before.tsv")
vcf_counts_runs <- utils::read.delim(
  "benchmarks/sql_scaling_vcf_counts_runs.tsv", stringsAsFactors = FALSE,
  check.names = FALSE, na.strings = "NA")
parity <- utils::read.delim("benchmarks/sql_scaling_vcf_keyed_differential.tsv")
import_parity <- utils::read.delim("benchmarks/sql_scaling_import_keyed_differential.tsv")
import_repeats <- utils::read.delim("benchmarks/sql_scaling_import_repetitions.tsv")
stopifnot(nrow(import_parity) == 1L,
          import_parity$rows_before == 562061L,
          import_parity$rows_after == 562061L,
          all(import_parity[c("duplicate_keys_before", "duplicate_keys_after",
                              "keyed_disagreements", "lost", "gained")] == 0L),
          nrow(import_repeats) == 18L,
          all(table(interaction(import_repeats$multiplier,
                                import_repeats$threads)) == 3L),
          all(import_repeats$input_records == import_repeats$result_rows),
          isTRUE(all.equal(import_repeats$rss_limit_mib,
                           256 + 3 * 192 * import_repeats$input_records / 1048576)),
          all(import_repeats$rss_verdict == ifelse(
            import_repeats$peak_rss_kib / 1024 <= import_repeats$rss_limit_mib,
            "within_budget", "over_budget")),
          all(import_repeats$spill_verdict == ifelse(
            import_repeats$peak_spill_mib <= import_repeats$spill_limit_mib,
            "within_budget", "over_budget")))
stopifnot(nrow(parity) == 1L, parity$rows_before == 2000L,
          parity$rows_after == 2000L,
          all(parity[c("duplicate_keys_before", "duplicate_keys_after",
                       "lost", "gained")] == 0L))
vcf_count_cells <- c(
  "panel_1x_fixed_file_4x", "panel_2x_fixed_file_4x",
  "panel_4x_fixed_file_4x", "file_1x_fixed_panel_1x",
  "file_2x_fixed_panel_1x", "file_4x_fixed_panel_1x",
  "joint_file_2x_panel_2x", "duplicate_error_1m")
stopifnot(nrow(vcf_counts_runs) == 96L,
          all(table(interaction(vcf_counts_runs$implementation,
                                vcf_counts_runs$cell,
                                vcf_counts_runs$threads)) == 3L),
          setequal(unique(vcf_counts_runs$cell), vcf_count_cells),
          setequal(unique(vcf_counts_runs$threads), c(1L, 4L)),
          all(vcf_counts_runs$memory_limit == "256MB"),
          all(vcf_counts_runs$max_temp_directory_size == "0B"),
          all(nzchar(vcf_counts_runs$private_temp_directory)),
          all(vcf_counts_runs$input_records > 0L),
          all(vcf_counts_runs$panel_sites > 0L))
format_metric <- function(values, digits = 1L) {
  values <- values[is.finite(values)]
  if (!length(values)) return("NA")
  sprintf(paste0("%.", digits, "f [ %.", digits, "f, %.", digits, "f ]"),
          median(values), min(values), max(values))
}
format_values <- function(values) {
  paste(sort(unique(ifelse(is.na(values), "NA", as.character(values)))),
        collapse = ", ")
}
summarize_vcf_counts <- function(implementation) {
  data <- vcf_counts_runs[vcf_counts_runs$implementation == implementation, ]
  keys <- unique(data[c("cell", "threads")])
  keys <- keys[order(match(keys$cell, vcf_count_cells), keys$threads), ]
  rows <- lapply(seq_len(nrow(keys)), function(index) {
    group <- data[data$cell == keys$cell[[index]] &
                    data$threads == keys$threads[[index]], ]
    passed <- sum(group$resource_verdict == "within_budget")
    verdict <- if (passed == 3L) "PASS" else {
      paste0("FAIL/INCOMPLETE (", passed, "/3 within budget)")
    }
    data.frame(
      cell = keys$cell[[index]], threads = keys$threads[[index]],
      input_records = format_values(group$input_records),
      panel_sites = format_values(group$panel_sites),
      output_rows = format_values(group$output_rows),
      execution = format_values(group$execution_status),
      seconds_median_min_max = format_metric(group$seconds, 3L),
      rss_mib_median_min_max = format_metric(group$peak_rss_mib),
      buffer_mib_median_min_max = format_metric(group$peak_buffer_mib),
      spill_mib_median_min_max = format_metric(group$peak_spill_mib),
      budget_verdict = verdict,
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}
vcf_counts_prechange <- summarize_vcf_counts("prechange")
vcf_counts_branch <- summarize_vcf_counts("branch")
vcf_counts_file_doublings <- local({
  data <- vcf_counts_runs[vcf_counts_runs$axis == "file", ]
  rows <- list()
  index <- 0L
  for (implementation in c("prechange", "branch")) {
    for (threads in c(1L, 4L)) {
      for (pair in list(c(1L, 2L), c(2L, 4L))) {
        from_cell <- paste0("file_", pair[[1L]], "x_fixed_panel_1x")
        to_cell <- paste0("file_", pair[[2L]], "x_fixed_panel_1x")
        from <- data[data$implementation == implementation &
                       data$threads == threads & data$cell == from_cell, ]
        to <- data[data$implementation == implementation &
                     data$threads == threads & data$cell == to_cell, ]
        delta_rss <- median(to$peak_rss_mib, na.rm = TRUE) -
          median(from$peak_rss_mib, na.rm = TRUE)
        delta_buffer <- median(to$peak_buffer_mib, na.rm = TRUE) -
          median(from$peak_buffer_mib, na.rm = TRUE)
        measured <- is.finite(delta_rss) && is.finite(delta_buffer)
        verdict <- if (!measured) "FAIL/INCOMPLETE (unmeasured)" else if (
          delta_rss <= 16 && delta_buffer <= 16
        ) "PASS" else "FAIL"
        index <- index + 1L
        rows[[index]] <- data.frame(
          implementation = implementation, threads = threads,
          doubling = paste0(pair[[1L]], "x to ", pair[[2L]], "x"),
          median_rss_delta_mib = delta_rss,
          median_buffer_delta_mib = delta_buffer,
          verdict = verdict, stringsAsFactors = FALSE)
      }
    }
  }
  do.call(rbind, rows)
})
vcf_counts_output_identity <- local({
  rows <- list()
  index <- 0L
  for (cell in vcf_count_cells) {
    for (threads in c(1L, 4L)) {
      group <- vcf_counts_runs[vcf_counts_runs$cell == cell &
                                 vcf_counts_runs$threads == threads, ]
      before_hash <- unique(na.omit(group$result_sha256[
        group$implementation == "prechange"]))
      branch_hash <- unique(na.omit(group$result_sha256[
        group$implementation == "branch"]))
      status <- if (!length(before_hash) || !length(branch_hash)) {
        "comparison unavailable"
      } else if (length(before_hash) == 1L && length(branch_hash) == 1L &&
                 identical(before_hash, branch_hash)) {
        "equal serialized full results"
      } else {
        "result hash disagreement"
      }
      index <- index + 1L
      rows[[index]] <- data.frame(cell = cell, threads = threads,
                                  output_identity = status,
                                  stringsAsFactors = FALSE)
    }
  }
  do.call(rbind, rows)
})
vcf_counts_build_identity <- unique(vcf_counts_runs[c(
  "implementation", "source_git_sha", "source_dirty_diff_sha256",
  "extension_sha256", "macro_text_sha256", "duckdb_version",
  "duckdb_source_id", "R_version", "DBI_version", "duckdb_R_version",
  "Rduckhts_installed_version", "Rduckhts_source_version",
  "duckhtsbench_source_version", "htslib_version")])
vcf_counts_input_identity <- unique(vcf_counts_runs[c(
  "input_id", "input_sha256", "input_bytes", "input_records",
  "panel_source_id", "panel_source_sha256", "panel_source_bytes",
  "panel_sites", "panel_sha256")])
stopifnot(all(table(interaction(real$case, real$threads)) == 3L),
          all(table(interaction(wrappers$case, wrappers$threads)) == 3L),
          all(real$multiplier %in% c(1L, 2L, 4L)),
          all(wrappers$multiplier %in% c(1L, 2L, 4L)),
          all(real$seconds > 0), all(real$result_rows > 0),
          all(wrappers$seconds > 0), all(wrappers$result_rows > 0))
summarize <- function(data) {
  rows <- lapply(split(data, interaction(data$case, data$threads)), function(group) {
    group <- group[order(group$multiplier), ]
    data.frame(entry_point = group$case[[1L]], threads = group$threads[[1L]],
               input_records = paste(format(group$input_records, big.mark = ",",
                                            scientific = FALSE, trim = TRUE),
                                     collapse = " / "),
               output_rows = paste(format(group$result_rows, big.mark = ",",
                                          scientific = FALSE, trim = TRUE),
                                   collapse = " / "),
               seconds = paste(sprintf("%.3f", group$seconds), collapse = " / "),
               peak_rss_mib = paste(sprintf("%.0f", group$peak_rss_kib / 1024),
                                    collapse = " / "),
               peak_buffer_mib = paste(sprintf("%.0f", group$peak_buffer_mib),
                                       collapse = " / "),
               exponent = sprintf("%.2f", log(group$seconds[[3L]] /
                                                  group$seconds[[1L]]) /
                                            log(group$input_records[[3L]] /
                                                  group$input_records[[1L]])))
  })
  result <- do.call(rbind, rows)
  result[order(match(result$entry_point, unique(data$case)), result$threads), ]
}
qualified <- summarize(rbind(real[real$case %in% c("munge", "munge_metal"), ],
                             wrappers))
probes <- summarize(real[!real$case %in% c("munge", "munge_metal"), ])
vcf_after <- real[real$case == "vcf_counts", ]
vcf_pair <- merge(before, vcf_after, by = c("case", "multiplier", "threads",
                                             "input_records"), suffixes = c("_before", "_after"))
stopifnot(nrow(vcf_pair) == 6L,
          all(vcf_pair$result_rows_before == vcf_pair$result_rows_after))
```

## Measurement contract

Source: `a0e7949dae4e7958d87cb9954f3cfa8fce59943a` (candidate); `0b9f32a8f572c4b01f2346f00086602ece473c67` (legacy VCF-count baseline only). DuckDB R 1.5.5; Linux; these previously recorded broad-run cells use one fresh DuckDB process and one observation per case/size/thread. The capped VCF-count matrix below uses three fresh processes per cell. `seconds` measures the query, including source reading and temporary-table materialization, but not source VCF subset derivation, panel preparation, connection, or COPY used for keyed differential. RSS is process `VmHWM` (includes preparation); DuckDB buffer is the query JSON profile's `system_peak_buffer_memory`. Numbers are MiB. These measurements are not cross-machine performance comparisons.

Source identity is checked against `r/duckhtsbench/inst/benchmark_registry.tsv`. `stage_sql_scaling_audit.R` derives real chromosome prefixes of the GIAB HG002 NIST v4.2.1 GRCh38 phased VCF (319,349 / 646,810 / 1,236,049 records), and unique biallelic SNV windows of the 1000 Genomes phase-3 GRCh37 chr22 VCF (124,658 / 274,567 / 562,061 records). VCF counts use 2,000 panel sites selected from each corresponding public VCF; import sites return one row per staged site. GRCh37 epilepsy METAL summary statistics supply genuine physical prefixes of 200,000 / 400,000 / 800,000 rows via a streaming `LIMIT`, with the registered GRCh37 FASTA supplied to both munge macros. Liftover uses 250,000 / 500,000 / 1,000,000 real phase-3 chr22 variants and the registered GRCh37→GRCh38 chain and FASTAs. Exponent is `log(t4/t1)/log(input4/input1)` on the displayed actual row counts. Source and output row counts are separate denominators.

Reproduce from repository root with `bcftools` and the registered inputs already in the cache:

```sh
Rscript benchmarks/stage_sql_scaling_audit.R
make release -j2
Rscript benchmarks/benchmark_sql_scaling_real_run.R build/release/duckhts.duckdb_extension
repo=$PWD
mkdir -p /tmp/duckhts-scaling-r-lib
(cd r/Rduckhts && Rscript bootstrap.R "$repo" && R_LIBS_USER=/tmp/duckhts-scaling-r-lib THREADS=4 make test)
R_LIBS_USER=/tmp/duckhts-scaling-r-lib Rscript benchmarks/benchmark_sql_scaling_real_run.R build/release/duckhts.duckdb_extension --wrappers
Rscript -e 'rmarkdown::render("benchmarks/benchmark_sql_scaling_audit.Rmd", knit_root_dir = getwd(), quiet = TRUE)'
```

Pass `--download` to the staging script to fetch missing direct-download inputs through the registry. The `0b9f32a8` VCF-count baseline is in `sql_scaling_audit_vcf_before.tsv`; `--wrappers` writes `sql_scaling_audit_wrappers.tsv`. The synthetic diagnostic inputs in `sql_scaling_audit_measurements.tsv`, `sql_scaling_audit_fixed_pairs.tsv`, and `sql_scaling_audit_norm.tsv` fall below the timing floor and are excluded from scaling conclusions.

## Earlier above-floor probes (one observation per cell)

Columns give 1× / 2× / 4× observations. The single-thread 1× workloads take 6–8 seconds. Both SQL macros and their R wrappers use reference-backed allele validation and emit all selected rows. The wrappers fetch the full result into R; their peak RSS includes the returned data frame, while the SQL timings materialize a DuckDB temporary table. The displayed exponents are exploratory: one observation per cell does not meet the three-repetition timing evidence requirement. These are distinct output-materialization workloads, not a wrapper-overhead comparison.

```{r qualified, results='asis'}
knitr::kable(qualified, row.names = FALSE)
```

## Real-input probes below the timing floor

These probes identify allocation behavior but their time exponents are **not scaling verdicts**: all one-thread 1× queries finish in less than two seconds. The liftover denominator is capped by the single staged chr22 VCF; the count and site-import prefixes do not reach a 5-second 1× while retaining a 4× prefix of the same public source.

```{r probes, results='asis'}
knitr::kable(probes, row.names = FALSE)
```

## VCF counts: a materialized scan worth removing

At 1,236,049 physical GIAB VCF records and 2,000 output panel rows, the legacy baseline JSON EXPLAIN ANALYZE shows `READ_GENO` (1,236,049 rows) under a materialized CTE feeding `HASH_JOIN`. Removing only `AS MATERIALIZED` from `__dht_source_records` lets the scan feed the join directly. At one thread, in that earlier uncapped one-observation run, before → after: **2.327 → 2.198 s**, **749 → 430 MiB peak RSS**, **1,367 → 612 MiB DuckDB buffer**. Across 319,349 / 646,810 / 1,236,049 records, before → after buffers (MiB) are **373 / 728 / 1,367 → 178 / 333 / 612**. Whole-result keyed differential on `(sample_id, site_index)` for the 4× input (full results from the 0b9f32a8 and candidate builds, exported separately to Parquet): 2,000 baseline and 2,000 candidate rows, zero duplicate keys; `EXCEPT ALL` in both directions returns zero rows. `check_sql_scaling_vcf_differential.R` reproduces the assertion from the two Parquet files; the checked-in receipt is `sql_scaling_vcf_keyed_differential.tsv`. This earlier allocation comparison is not a bounded-memory claim; the capped repeated matrix below applies the declared 256 MiB budgets.

## duckhts_somalier_vcf_counts: capped before/after matrix

This matrix uses the frozen pre-change extension and the extension built from the recorded branch revision. Every cell runs in three fresh serial processes at 1 and 4 DuckDB threads, and the query result is fetched with every output column consumed. The file-axis cells hold a 2,000-site panel fixed across the registered 1× / 2× / 4× GIAB inputs; the panel-axis cells hold the 4× file fixed and use nested 2,000 / 4,000 / 8,000-site panels. The joint-growth cell uses the 2× file and 4,000-site panel. The duplicate-error cell reads one million copies of a single registered chr1 biallelic SNP with its GT and AD calls and uses the matching one-site panel; the staging test exercises this deterministic transformation from a local VCF record without network access.

Before these measurements, the benchmark source declared **process RSS ≤256 MiB**, **query-wide DuckDB buffer ≤256 MiB**, **spill exactly 0 MiB**, and for each file doubling at fixed panel **both median RSS and median query buffer increase by no more than 16 MiB**. Each child sets `memory_limit='256MB'`, `max_temp_directory_size='0B'`, and a private temporary directory. The raw run file records the per-process settings, all input and panel checksums, output row counts, timings, memory observations, versions, build identities, and the frozen/candidate macro hashes. Any failed or unmeasured metric remains a budget failure/incomplete result. The fastest 1× run is below the 5-second timing floor, so no timing exponent or timing-scaling verdict is made.

### Frozen pre-change implementation

```{r vcf-counts-prechange, results='asis'}
knitr::kable(vcf_counts_prechange, row.names = FALSE)
```

### Branch implementation

```{r vcf-counts-branch, results='asis'}
knitr::kable(vcf_counts_branch, row.names = FALSE)
```

### File-doubling memory gates

The `1×→2×` and `2×→4×` verdicts require both median memory deltas to remain at or below 16 MiB. Failures and unavailable profiles remain visible.

```{r vcf-counts-file-doublings, results='asis'}
knitr::kable(vcf_counts_file_doublings, row.names = FALSE)
```

### Per-cell output and provenance checks

```{r vcf-counts-output-identity, results='asis'}
knitr::kable(vcf_counts_output_identity, row.names = FALSE)
```

The raw per-run records are `sql_scaling_vcf_counts_runs.tsv`. The following tables identify the exact builds and staged inputs used by those runs; the baseline source commit is `296d44a4d9c190c0551257b5e94740988332966e` and its empty dirty-diff hash identifies the clean commit tree.

```{r vcf-counts-build-identity, results='asis'}
knitr::kable(vcf_counts_build_identity, row.names = FALSE)
knitr::kable(vcf_counts_input_identity, row.names = FALSE)
```

## Site import: real-input materialization

The candidate build is the commit containing this report, based on `509190790a3ce9d51ec735b2313d0c46b0969a70`. The public phase-3 GRCh37 chr22 VCF supplies 124,658 / 274,567 / 562,061 biallelic SNVs; the importer emits exactly one row per site. Offline registry validation and VCF derivation (`stage_sql_scaling_audit.R`) are outside query timing. The full 562,061-row keyed differential on `(assembly, site_index)` is in `sql_scaling_import_keyed_differential.tsv`; zero duplicate keys, missing rows, or column disagreements. Its baseline and candidate Parquet exports were produced in separate fresh processes. The baseline physical plan materializes `__dht_source` (562,061 records), `__dht_autosomal`, and `__dht_oriented` before sorting; the candidate retains only the shared source, allowing validation and orientation to consume it without two further wide copies. `READ_BCF` runs once in either plan.

Before measurement, the per-process budget was set at **256 MiB runtime + 3 × (64 bytes decoded source + 128 bytes materialized output) per physical input row**; query spill limit **0 MiB**. The three fresh-process repetitions per size at each of 1 and 4 threads are in `sql_scaling_import_repetitions.tsv`. Each measurement materializes every output row in a temporary table. Peak RSS includes connection and extension overhead; buffer/spill are taken from the measured query's JSON profile, before any row-count probe. Staging and oracle work are reported separately. The 1× one-thread time is <5 s, so **no timing exponent or timing scaling verdict** is asserted. Memory stays under the predetermined RSS budget with zero spill at both thread counts.

```{r import-repetitions, results='asis'}
import_summary <- do.call(rbind, lapply(
  split(import_repeats, interaction(import_repeats$multiplier,
                                    import_repeats$threads)), function(group) {
    data.frame(input_rows = group$input_records[[1L]],
               output_rows = group$result_rows[[1L]],
               threads = group$threads[[1L]],
               seconds = paste(sprintf("%.3f", group$seconds), collapse = " / "),
               rss_mib = paste(sprintf("%.0f", group$peak_rss_kib / 1024),
                               collapse = " / "),
               rss_limit_mib = sprintf("%.0f", group$rss_limit_mib[[1L]]),
               buffer_mib = paste(sprintf("%.0f", group$peak_buffer_mib),
                                  collapse = " / "),
               spill_mib = paste(sprintf("%.0f", group$peak_spill_mib),
                                 collapse = " / "))
  }))
import_summary <- import_summary[order(import_summary$input_rows,
                                       import_summary$threads), ]
knitr::kable(import_summary, row.names = FALSE)
```

At one thread, baseline → candidate buffer for 1× / 2× / 4× is **222 / 469 / 942 → 149 / 309 / 614 MiB**. At 562,061 rows, the paired keyed-export runs show **687 → 527 MiB RSS**; both builds produced all 562,061 rows. The repeated candidate 4-thread 4× RSS is **537–539 MiB** against the **565 MiB** budget, with **0 MiB** spill. Reproduce query repetitions with `Rscript benchmarks/benchmark_sql_scaling_import_repetitions.R`; compare separately exported results with `Rscript benchmarks/check_sql_scaling_import_differential.R before.parquet after.parquet`. Timing excludes input staging, keyed exports, and result-count probes.

## Open findings, ranked by measured exposure

1. **VCF-count join:** 612 MiB of DuckDB buffer for 1.24 million GIAB records and 2,000 output rows. `READ_GENO` is estimated at one row; the left join therefore builds a hash table from the full VCF source. An exploratory panel `SEMI JOIN` was planned as `RIGHT_SEMI`, still building the source (658 MiB buffer in a single-process diagnostic), so it was not retained. A genuinely selective indexed input or keyed stream join needs a separate design and keyed differential; this diagnostic is not a timing verdict.
2. **Site import:** the output grows one-to-one with input; the candidate reduces its 4× one-thread buffer from 942 to 614 MiB, but the 1× query is still below the timing floor. A larger public source is required before a time-scaling verdict.
3. **Munge/Metal reference-backed modes:** the single-observation 200k / 400k / 800k real epilepsy probe approaches linear time, but does not meet the three-repetition timing gate. The 800k-row SQL output occupies about 779 MiB DuckDB buffer; COPY-to-disk output and any bounded streaming requirement are separate questions. The R wrapper 800k-row result reaches 645 MiB RSS (`r_munge`) and 710 MiB (`r_munge_metal`), consistent with returning a large R data frame.
4. **Liftover:** the entire staged chr22 GRCh37 source holds only ~1.1 million variants; the current 250k baseline is below the timing floor. A larger public source, allele/mapped-output verification, and EXPLAIN ANALYZE at that scale are required.
5. **Not measured with real multi-sample evidence:** BAM/CRAM extraction, sketch preparation/verification, Charr, matched contamination and R all-pairs relatedness. The phased 1000 Genomes chr22 VCFs have GT but no FORMAT/AD. Public indexed 30× CRAM regional access was probed on HG00188 chr22:16,000,000–16,020,000 (2,680 alignments; 1.75 s; 37 MiB RSS) using the registered GRCh38 reference. This establishes feasibility, not a checksum-validated 16/32/64-sample staged cohort or a Somalier performance measurement. All-pairs relatedness needs a separately capped quadratic output-size and memory contract.
6. **Not measured:** executed BAM/VCF/GFF `*_convert_parquet_sql` COPYs. These require physical multi-size public inputs, output denominator checks and independent disk-space limits.

The first-pass synthetic and 10k–40k-row rows have no extrapolatable scaling conclusion. This report does not claim completion of the unmeasured entry points.
