---
title: "Somalier find-sites: 1000 Genomes 30x phased chr22 and chrX"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
root <- normalizePath("..", mustWork = TRUE)
read_result <- function(name) utils::read.delim(
  file.path(root, "benchmarks", name), check.names = FALSE
)
observations <- read_result("somalier_find_sites_chr22_observations.tsv")
gates <- read_result("somalier_find_sites_chr22_gates.tsv")
timing <- read_result("somalier_find_sites_chr22_timing.tsv")
classes <- read_result("somalier_find_sites_chr22_disagreement_summary.tsv")
disagreements <- read_result("somalier_find_sites_chr22_disagreements.tsv")
xobservations <- read_result("somalier_find_sites_chrx_observations.tsv")
xgates <- read_result("somalier_find_sites_chrx_gates.tsv")
xtiming <- read_result("somalier_find_sites_chrx_timing.tsv")
xclasses <- read_result("somalier_find_sites_chrx_disagreement_summary.tsv")
xdisagreements <- read_result("somalier_find_sites_chrx_disagreements.tsv")
xpar <- read_result("somalier_find_sites_chrx_par_records.tsv")
profile <- read_result("somalier_find_sites_chrx_profile.tsv")
matched <- read_result("somalier_find_sites_matched_timing.tsv")
memory <- read_result("somalier_find_sites_memory.tsv")
baselines <- subset(memory, input == "empty")
memory_runs <- subset(memory, input == "VCF")
rownames(memory_runs) <- NULL
keys <- c("chromosome", "phase", "decompression_workers")
baseline_index <- match(do.call(paste, memory_runs[keys]),
                        do.call(paste, baselines[keys]))
memory_runs$baseline_rss_kib <- baselines$peak_rss_kib[baseline_index]
memory_runs$net_bytes_per_input_variant <- round(
  1024 * (memory_runs$peak_rss_kib - memory_runs$baseline_rss_kib) /
    memory_runs$input_records
)
memory_runs$net_bytes_per_candidate <- round(
  1024 * (memory_runs$peak_rss_kib - memory_runs$baseline_rss_kib) /
    memory_runs$gated_candidates
)
```

This is a whole-chromosome differential of DuckHTS and pinned Somalier
[v0.3.4](https://github.com/brentp/somalier/tree/ff58fdade8f4f8293d904f10e0a4a13f1fac808d)
(binary SHA-256 `18717c205a9c4b65d479f1d2cf069a30b047a5378edf544d403d4081f06a3a78`).
The unmodified [1000 Genomes high-coverage 30x phased chr22 VCF](https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20201028_3202_phased/CCDG_14151_B01_GRM_WGS_2020-08-05_chr22.filtered.shapeit2-duohmm-phased.vcf.gz)
is staged through `duckhtsbench` (SHA-256
`70ff24cfd01eab24c6f4c6e825f55a2b3475ff595f6e12e179c4ef06b9cc4df6`,
519,930,289 bytes, GRCh38). Its declared INFO/AF and INFO/AN are used without
recomputing from genotypes. Somalier skips sample genotypes; DuckHTS reads the
physical VCF in sequential mode with `samples := ''`. The cache VCF is not committed.

## Selection rules

| Rule | Somalier v0.3.4 and DuckHTS default | DuckHTS named alternative |
|:--|:--|:--|
| Human contigs | Exclude autosomal REF=C; X zero-based positions 2,781,479..154,931,044 | `compatibility_mode = "generic"` does not classify human sex contigs |
| AF rank ties | Retain input order on equal AF scores; a reversed-position two-site probe selects the first input record with pinned Somalier | `tie_order = "lexical"` orders by position and allele |
| Spacing | Greedy autosomal distance 10,000 by default; X/Y selections do not update spacing state | `sex_spacing = "enforced"` applies X distance 1,000 and Y distance 200 |
| Caps | Up to 65,535 autosomal, 10,001 X, and 5,001 Y selections after ranking | Named per-sex maximum arguments |

- Source revision: `r observations$revision`. Workload: `--min-AN 6000`,
  `--snp-dist 10000`, minimum AF 0.15, AF target 0.48, no interval or
  gnotate exclusion. One run per tool, warm filesystem cache, GNU `/usr/bin/time -v`.
- Denominators: `r format(observations$source_records, big.mark = ",")` input
  records; `r format(observations$upstream_candidates, big.mark = ",")` Somalier
  candidates and `r format(observations$duckhts_candidates, big.mark = ",")`
  DuckHTS candidates before nearby-variant exclusion;
  `r format(observations$max_per_chromosome_candidates, big.mark = ",")`
  chr22 spacing inputs. The native spacing kernel's per-list limit is
  1,000,000 candidates; this chromosome's largest list is below it.
- Selected: `r observations$upstream_sites` Somalier sites and
  `r observations$duckhts_sites` DuckHTS sites;
  `r observations$common_sites` shared four-field keys
  `(region, position, source_ref, source_alt)`. Input-order and lexical
  tie policies differed on `r observations$lexical_tie_dependent_sites`
  selected sites on this chromosome.
- Threads: Somalier uses two reader/decompression threads and one writer
  thread (pinned source); DuckHTS uses one DuckDB thread and a sequential
  reader. The timed DuckHTS process also verifies its cached input and
  starts R/DuckDB; Somalier timing covers the CLI only. These single-process
  measurements are not matched-concurrency speedups.

```{r gates}
knitr::kable(gates, caption = "Whole-chromosome stage counts. Indels and snps are exclusion-neighbor inventories, not sequential pass counts. NA denotes stages the pinned Somalier executable does not expose; its own log reports the gated candidate count.")
```

```{r timing}
knitr::kable(timing, caption = paste(
  "Wall seconds and peak RSS KiB for each timed process, including its own",
  "initialization and I/O. Single runs are not a variance estimate."))
```

```{r disagreement-summary}
knitr::kable(classes, caption = "Keyed disagreement classes (all records counted).")
```

The retained `somalier_find_sites_chr22_disagreements.tsv` contains the first
100 disagreement keys and origins; `r nrow(disagreements)` rows are present.
An empty table means no disagreements, not an absent comparison. The
`somalier_find_sites_chr22_observations.tsv` and gate table retain the full
comparison denominators. The previous
[`benchmark_somalier_find_sites.md`](benchmark_somalier_find_sites.md)
uses two small fixture slices and is not a comparable whole-chromosome baseline.

## Chromosome X, including physical PAR records

The unmodified [1000 Genomes 30x phased chrX VCF](https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20201028_3202_phased/CCDG_14151_B01_GRM_WGS_2020-08-05_chrX.filtered.eagle2-phased.v2.vcf.gz)
(SHA-256 `f42c786cba67ff74c30e9bd37254b5a605a5957cfe3a5dc54e7dd41d9c9d8edc`,
2,800,398,832 bytes) contains `r format(xpar$par1_records, big.mark = ",")`
records in PAR1 (before POS 2,781,480) and
`r format(xpar$par2_records, big.mark = ",")` records in PAR2 (after POS
154,931,045). The physical first and last positions are
`r format(xpar$first_position, big.mark = ",")` and
`r format(xpar$last_position, big.mark = ",")`.
Somalier's pinned X-range rule admits only zero-based positions
2,781,479..154,931,044. The `r format(xpar$par1_records + xpar$par2_records, big.mark = ",")`
PAR records account for the drop from source to eligible records.
Somalier's log says “0 autosomal variants” for this X-only input; the
actual compressed output VCF, not that autosomal-only log counter, is used
for the selected-site denominator and keyed comparison.

- Source revision: `r xobservations$revision`.
- Denominators: `r format(xobservations$source_records, big.mark = ",")`
  input records; `r format(xobservations$upstream_candidates, big.mark = ",")`
  Somalier candidates and `r format(xobservations$duckhts_candidates, big.mark = ",")`
  DuckHTS gated candidates;
  `r format(xobservations$max_per_chromosome_candidates, big.mark = ",")`
  spacing inputs (below the 1,000,000 limit).
- Selected: `r format(xobservations$upstream_sites, big.mark = ",")` Somalier and
  `r format(xobservations$duckhts_sites, big.mark = ",")` DuckHTS sites;
  `r format(xobservations$common_sites, big.mark = ",")` matching four-field keys.
  Input and lexical tie policies differed on
  `r xobservations$lexical_tie_dependent_sites` selected sites.
  `r xobservations$duckhts_only` DuckHTS-only and
  `r xobservations$upstream_only` Somalier-only keys.

```{r x-gates}
knitr::kable(xgates, caption = "chrX stage counts; NA denotes stages not reported by the pinned executable.")
```

```{r x-timing}
knitr::kable(xtiming, caption = "chrX whole-process wall seconds and peak RSS KiB; thread and measurement caveats above also apply.")
```

```{r x-disagreement-summary}
knitr::kable(xclasses, caption = sprintf(
  "chrX keyed disagreement classes. First 100 rows in somalier_find_sites_chrx_disagreements.tsv; retained: %s rows.",
  nrow(xdisagreements)))
```

## Operator profile and matched-thread selection

The before/after chrX profile uses source revisions `396f9599` and
`99d2f443` on the same 2,858,184 physical VCF records, with one DuckDB
thread, warm filesystem cache, no sample genotype decode, and the same
250,641 gated candidates and 234,367 spacing inputs. Each complete query
was profiled with DuckDB JSON profiling (query latency and operator timing);
`EXPLAIN ANALYZE` also checked the ordered-stream plan. The 14 named gate
counts were collected by one query with
shared materialized CTEs. Operator times are pipelined and **not additive**.
No gate count or keyed denominator was dropped to obtain these timings.

```{r operator-profile}
knitr::kable(profile, digits = 4, caption = paste(
  "chrX seconds. Named gate rows are FILTER operators; join/window rows",
  "isolate nearby-variant work; the mask projection bounds the spacing",
  "kernel and its projection rather than timing the kernel alone."))
```

A separate `count(*)` over `read_bcf` took 16.700 s; decoding and summing
only `INFO_AF[1]` and `INFO_AN` took 18.265 s. The selector's `READ_BCF`
operator took 19.237 s before and 19.330 s after. Its projected columns
are `CHROM`, `POS`, `REF`, `ALT`, `FILTER`, `INFO_AF`, `INFO_AN`,
`INFO_BaseQRankSum`, `INFO_ClippingRankSum`, `INFO_FS`, `INFO_MQ`,
`INFO_MQRankSum`, `INFO_QD`, and `INFO_ReadPosRankSum`; no genotype columns
or other INFO fields are materialized. These are separate full scans, not
additive decode components. The positional and QC predicates run after the
reader; this table function does not push them into VCF decoding. Sampling
with `perf` was unavailable (`perf_event_paranoid=4`).

For the matched-thread runs below, DuckHTS uses one SQL worker and either
zero or two htslib BGZF decompression workers; Somalier uses its pinned two
readers and one writer. DuckDB is set to one thread in both DuckHTS runs.
A separate warm-cache process profiles one selection per row; `process_seconds`
adds R startup, the connection, and result transfer to `selection_seconds`,
but excludes the input cache checksum. Somalier's CLI process includes its
own startup and VCF writer; these distinct scopes are reported, not treated
as a controlled throughput ratio. Values are single runs, not variance
estimates. The official one-thread `select` workflow *includes* cache
verification and took `r timing$wall_seconds[timing$tool == "DuckHTS"]` s
on chr22 and `r xtiming$wall_seconds[xtiming$tool == "DuckHTS"]` s on chrX.
Isolated warm-cache `duckhts_bench_fetch()` calls took 0.285 s and 1.440 s,
respectively; R and library startup added about 0.3 s and opening a DuckDB
connection about 0.04 s. All 2,402 chr22 and 10,001 chrX keys also agree
between one-worker and three-worker DuckHTS selections.

```{r matched-threads}
knitr::kable(matched, digits = 3, caption = paste(
  "Warm-cache whole-chromosome selection: input/output denominators,",
  "wall seconds and peak process RSS KiB. NA: Somalier has no separate",
  "in-process SQL selection timing."))
```

The ordered position passes reduce the chrX selection query from 219.640 s
to approximately 21 s at one thread. The initial ordered query (revision
`dd87028e0c33b9b0ed807e71a6aeeed7a3eec764`) still
materializes the full source and eligible relations. The filtered query
materializes only records contributing an indel inventory, a nearby-SNP
inventory, or an AF/AN-qualified site (code revision
`70ef384cc023350b7e10bfd01881ca9f26417498`). Indel and SNP inventory relations
hold only chromosome, position and span/count data. Both queries use the
same installed Rduckhts extension, source VCFs and pinned selection rules.
The 14 diagnostic gate counts share one source scan; that diagnostic query
retains the full source relation and is not included in this selection-only
memory comparison.

GNU time reports peak process RSS. DuckDB JSON profiling reports the query's
peak buffer-manager allocation, **not** total resident memory or a sample
synchronized to peak RSS. After the query, `duckdb_memory()` reports zero
buffer bytes for every row in this table. Each empty input keeps the reader
schema (`WHERE false`) and measures R, DuckDB, extension startup and query
planning in a fresh process. Net bytes per variant subtract the paired empty
process RSS, then divide by the physical input records; bytes per candidate
use the `gated` denominator (before the indel/SNP neighborhood passes).
The inventories contain more records than `gated`, so the latter is not a
complete allocation denominator. Times include JSON profiling, but exclude
cache checksum and initial input staging. Single runs are not a variance
estimate.

```{r memory}
knitr::kable(memory_runs[, c(
  "chromosome", "phase", "decompression_workers", "input_records",
  "gated_candidates", "selected_sites", "baseline_rss_kib", "peak_rss_kib",
  "net_bytes_per_input_variant", "net_bytes_per_candidate",
  "query_peak_buffer_bytes", "selection_seconds", "process_seconds"
)], caption = paste("Fresh-process selection and paired empty-input baseline.",
                   "RSS in KiB; DuckDB peak buffer allocations in bytes."))
```

At three active workers, chrX RSS falls from 1,225,804 to 560,992 KiB
and chr22 RSS from 453,552 to 212,312 KiB. The corresponding query times
are 13.420 vs 13.149 seconds on chrX and 5.068 vs 5.044 seconds on chr22.
The same physical chromosome inputs produce 10,001 and 2,402 selected keys;
one-worker and three-worker results are identical, with zero keyed
disagreements against pinned Somalier. Memory is still bounded below by R,
DuckDB and the extension (~134,000 KiB here); the retained candidate rows,
indel/SNP inventories, reader decode buffers and downstream sorts contribute
to the additional resident memory; these measurements do not isolate their
shares. Predicate pushdown into `read_bcf()` is not part of this selection
query.

Reproduce from the repository root with installed `Rduckhts` and
`duckhtsbench`, the pinned binary staged in `duckhtsbench`, and GNU time:

```bash
export DUCKHTSBENCH_REGISTRY="$PWD/r/duckhtsbench/inst/benchmark_registry.tsv"
Rscript test/scripts/somalier_find_sites_chromosome.R upstream "$PWD" chr22
time_file="$(Rscript -e 'cat(file.path(dirname(duckhtsbench::duckhts_bench_artifact_path("somalier_find_sites_chr22_source")), "chr22.duckhts.time"))')"
/usr/bin/time -v -o "$time_file" \
  Rscript test/scripts/somalier_find_sites_chromosome.R select "$PWD" chr22
Rscript test/scripts/somalier_find_sites_chromosome.R compare "$PWD" chr22
Rscript test/scripts/somalier_find_sites_chromosome.R upstream "$PWD" chrX
x_time_file="$(Rscript -e 'cat(file.path(dirname(duckhtsbench::duckhts_bench_artifact_path("somalier_find_sites_chrx_source")), "chrX.duckhts.time"))')"
/usr/bin/time -v -o "$x_time_file" \
  Rscript test/scripts/somalier_find_sites_chromosome.R select "$PWD" chrX
Rscript test/scripts/somalier_find_sites_chromosome.R compare "$PWD" chrX
# Compare both query revisions using the same installed Rduckhts extension.
old_query="$(mktemp)"
trap 'rm -f "$old_query"' EXIT
git show dd87028e:r/Rduckhts/R/somalier_find_sites.R > "$old_query"
for phase in before after; do
  if [ "$phase" = before ]; then
    export DUCKHTS_FIND_SITES_SOURCE="$old_query"
  else
    export DUCKHTS_FIND_SITES_SOURCE="$PWD/r/Rduckhts/R/somalier_find_sites.R"
  fi
  for chrom in chr22 chrX; do
    for decompression_workers in 0 2; do
      for input in empty input; do
        /usr/bin/time -f 'peak_rss_kib: %M process_seconds: %e' \
          Rscript scripts/benchmark_somalier_find_sites_profile.R \
            "$chrom" "$decompression_workers" "$input"
      done
    done
  done
done
cd benchmarks && Rscript -e "rmarkdown::render('benchmark_somalier_find_sites_chr22.Rmd')"
```
