---
title: "GENCODE GFF3 named-attribute projections"
output: github_document
---

The source is the public GENCODE mouse vM25 BASIC GFF3, staged by
`scripts/stage_tabix_split.R --offline` from the `tabix_split_gff3` registry
artifact (24,025,067 compressed bytes; 1,299,172 annotation rows). Each query
streams the whole file in sequential mode. Measured outputs are scalar
aggregates of the stated columns, not `count(*)` shortcuts. Five measured runs
follow one warm-up per workload and thread setting. Times include gzip decode,
row scan, aggregation and R/DBI query execution. Default threads are the
runtime's own default; the GFF reader uses one worker in either setting.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r benchmark, include=FALSE}
Sys.setenv(DUCKHTSBENCH_REGISTRY = normalizePath(
  "r/duckhtsbench/inst/benchmark_registry.tsv", mustWork = TRUE
))
stopifnot(requireNamespace("duckhtsbench", quietly = TRUE))
stopifnot(requireNamespace("duckdb", quietly = TRUE))
stopifnot(requireNamespace("DBI", quietly = TRUE))
stopifnot(requireNamespace("knitr", quietly = TRUE))
input <- duckhtsbench::duckhts_bench_artifact_path("tabix_split_gff3")
duckhtsbench::duckhts_bench_validate_identity("tabix_split_gff3", input)
input <- normalizePath(input, mustWork = TRUE)
extension <- normalizePath("build/release/duckhts.duckdb_extension", mustWork = TRUE)
quote_sql <- function(value) paste0("'", gsub("'", "''", value, fixed = TRUE), "'")
revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE)
if (system2("git", c("diff", "--quiet", "HEAD", "--", "src"),
            stdout = FALSE, stderr = FALSE) != 0L) {
  stop("Build source must match the recorded HEAD")
}
driver <- duckdb::duckdb(shared_home = FALSE, config = list(allow_unsigned_extensions = "true"))
con <- DBI::dbConnect(driver)
DBI::dbExecute(con, paste("LOAD", quote_sql(extension)))
default_threads <- DBI::dbGetQuery(con, "SELECT current_setting('threads')::INTEGER AS n")$n[[1L]]
source_sql <- paste0("read_gff(", quote_sql(input), ", scan_mode := 'sequential'")
plain <- paste0(source_sql, ")")
named <- paste0(source_sql, ", attributes := ['Parent', 'ID', 'gene_type'])")
map <- paste0(source_sql, ", attributes_map := true)")
queries <- c(
  plain_columns = paste0("SELECT count(*) n, sum(start + \"end\") metric FROM ", plain),
  named_none_projected = paste0("SELECT count(*) n, sum(start + \"end\") metric FROM ", named),
  attributes_string = paste0("SELECT count(*) n, sum(length(attributes)) metric FROM ", plain),
  map_parent = paste0("SELECT count(*) n, count(attributes_map['Parent']) metric FROM ", map),
  regexp_three = paste0("SELECT count(*) n, sum(length(regexp_extract(attributes, '(?:^|;)Parent=([^;]*)', 1)) + length(regexp_extract(attributes, '(?:^|;)ID=([^;]*)', 1)) + length(regexp_extract(attributes, '(?:^|;)gene_type=([^;]*)', 1))) metric FROM ", plain),
  named_three = paste0("SELECT count(*) n, sum(length(coalesce(Parent, '')) + length(coalesce(ID, '')) + length(coalesce(gene_type, ''))) metric FROM ", named),
  named_one_projected = paste0("SELECT count(*) n, sum(length(coalesce(Parent, ''))) metric FROM ", named)
)
rows <- list()
for (threads in c(1L, default_threads)) {
  DBI::dbExecute(con, paste0("SET threads = ", threads))
  for (kind in names(queries)) {
    sql <- queries[[kind]]
    invisible(DBI::dbGetQuery(con, sql))
    elapsed <- numeric(5L)
    result <- NULL
    for (run in seq_along(elapsed)) {
      start <- proc.time()[["elapsed"]]
      result <- DBI::dbGetQuery(con, sql)
      elapsed[[run]] <- proc.time()[["elapsed"]] - start
    }
    stopifnot(identical(as.numeric(result$n[[1L]]), 1299172))
    rows[[length(rows) + 1L]] <- data.frame(
      threads = if (threads == 1L) "1" else paste0("default (", threads, ")"),
      workload = kind, input_rows = result$n[[1L]],
      output_rows = 1L, metric = as.character(result$metric[[1L]]),
      median_seconds = median(elapsed), runs = length(elapsed)
    )
  }
}
results <- do.call(rbind, rows)
DBI::dbDisconnect(con, shutdown = TRUE)
```

Source revision: `r revision`. DuckDB runtime: `r as.character(packageVersion("duckdb"))`.
Extension SHA-256: `r strsplit(system2("sha256sum", extension, stdout = TRUE), " ", fixed = TRUE)[[1L]][[1L]]`.
Input SHA-256: `r strsplit(system2("sha256sum", input, stdout = TRUE), " ", fixed = TRUE)[[1L]][[1L]]`.

```{r results, echo=FALSE}
knitr::kable(results, digits = 3)
```

The one-key projection requests three keys at bind but scans only `Parent`.
`named_none_projected` requests three keys at bind but reads only the physical
coordinate columns; its scan skips attribute parsing entirely. Comparisons across workloads reflect different aggregation and expression costs,
not isolated parsing cycles. The local duckdb-wasm browser test could not load
this extension: `duckdb_get_list_size` has an incompatible wasm import type
with the staged browser runtime. Browser wasm timings are unavailable.
