---
title: "BigWig reader benchmark"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

This report measures `read_bigwig(...)` on a dense, real conservation track.
The input is a local 10 Mb slice of the UCSC GRCh38 phyloP 100-way BigWig.
Network transfer and slice construction are staging work and are outside every
timed interval.

There is no earlier checked-in DuckHTS BigWig benchmark. The indexed
multi-region BCF/VCF report in
[`benchmark_multi_region_readers.md`](benchmark_multi_region_readers.md) is the
nearest reader benchmark, but it exercises htslib variant iterators and is not
numerically comparable.

## Reproduction

Stage the real track with the UCSC utilities. Their interval flags are
zero-based and half-open, matching the stored BigWig coordinates.

```bash
mkdir -p /tmp/ucsc-bigwig-bench
curl -fsSL https://hgdownload.soe.ucsc.edu/admin/exe/linux.x86_64/bigWigToBedGraph \
  -o /tmp/ucsc-bigwig-bench/bigWigToBedGraph
curl -fsSL https://hgdownload.soe.ucsc.edu/admin/exe/linux.x86_64/bedGraphToBigWig \
  -o /tmp/ucsc-bigwig-bench/bedGraphToBigWig
chmod +x /tmp/ucsc-bigwig-bench/bigWigToBedGraph \
  /tmp/ucsc-bigwig-bench/bedGraphToBigWig

/tmp/ucsc-bigwig-bench/bigWigToBedGraph \
  -chrom=chr22 -start=20000000 -end=30000000 \
  https://hgdownload.soe.ucsc.edu/goldenPath/hg38/phyloP100way/hg38.phyloP100way.bw \
  /tmp/ucsc-bigwig-bench/chr22_20m_30m.bedGraph
printf 'chr22\t50818468\n' > /tmp/ucsc-bigwig-bench/hg38.chr22.sizes
/tmp/ucsc-bigwig-bench/bedGraphToBigWig \
  /tmp/ucsc-bigwig-bench/chr22_20m_30m.bedGraph \
  /tmp/ucsc-bigwig-bench/hg38.chr22.sizes \
  /tmp/ucsc-bigwig-bench/hg38_chr22_20m_30m_phyloP100way.bw
```

Build and install `Rduckhts` from the source revision under test, then render:

```bash
DUCKHTS_BIGWIG_BENCHMARK_INPUT=/tmp/ucsc-bigwig-bench/hg38_chr22_20m_30m_phyloP100way.bw \
  Rscript -e "rmarkdown::render('benchmarks/benchmark_bigwig_reader.Rmd', output_format='github_document')"
```

The benchmark pins the one-thread condition to CPU 12 and the four-thread
condition to physical CPUs 12--15 on the recorded host. Input construction,
extension loading, connection setup, warm-up, result verification, and removal
of the previous Parquet output are excluded from timing.

```{r benchmark, echo=FALSE}
suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
  library(Rduckhts)
})

sql_string <- function(x) {
  sprintf("'%s'", gsub("'", "''", x, fixed = TRUE))
}

command_output <- function(command, args = character()) {
  paste(system2(command, args, stdout = TRUE, stderr = TRUE), collapse = " ")
}

input_path <- Sys.getenv(
  "DUCKHTS_BIGWIG_BENCHMARK_INPUT",
  "/tmp/ucsc-bigwig-bench/hg38_chr22_20m_30m_phyloP100way.bw"
)
if (!file.exists(input_path)) {
  stop("Stage DUCKHTS_BIGWIG_BENCHMARK_INPUT using the commands above")
}
input_path <- normalizePath(input_path, mustWork = TRUE)

source_revision <- command_output("git", c("rev-parse", "--short=12", "HEAD"))
tracked_dirty <- length(system2(
  "git", c("status", "--porcelain", "--untracked-files=no"), stdout = TRUE
)) > 0L
if (tracked_dirty) {
  stop("Render the benchmark from a tracked-clean source revision")
}

con <- dbConnect(duckdb(config = list(allow_unsigned_extensions = "true")))
invisible(rduckhts_load(con))

aggregate_sql <- function(region = NULL, relation = NULL) {
  source <- if (is.null(relation)) {
    region_argument <- if (is.null(region)) "" else {
      paste0(", region := ", sql_string(region))
    }
    sprintf(
      "read_bigwig(%s%s, blocks_per_iteration := 64)",
      sql_string(input_path), region_argument
    )
  } else {
    relation
  }
  sprintf(
    paste0(
      "SELECT count(*)::UBIGINT AS rows, ",
      "hex(bit_xor(hash(CHROM, START0, END0, VALUE))) AS fingerprint, ",
      "sum(VALUE)::DOUBLE AS value_sum FROM %s"
    ),
    source
  )
}

full_expected <- dbGetQuery(con, aggregate_sql())
stopifnot(as.double(full_expected$rows[[1L]]) > 1000000)

# Twenty separated 250 kb ranges expose multiple indexed work items while
# selecting half of the 10 Mb slice.
region_starts <- seq.int(20000001L, 29750001L, by = 500000L)
regions <- sprintf("chr22:%d-%d", region_starts, region_starts + 249999L)
region_spec <- paste(regions, collapse = ",")
region_expected <- dbGetQuery(con, aggregate_sql(region_spec))

same_answer <- function(observed, expected) {
  identical(as.character(observed$rows[[1L]]), as.character(expected$rows[[1L]])) &&
    identical(observed$fingerprint[[1L]], expected$fingerprint[[1L]]) &&
    isTRUE(all.equal(
      observed$value_sum[[1L]], expected$value_sum[[1L]],
      tolerance = 1e-12
    ))
}

set_affinity <- function(cpus) {
  output <- system2(
    "taskset", c("-pc", cpus, Sys.getpid()),
    stdout = TRUE, stderr = TRUE
  )
  if (!is.null(attr(output, "status"))) stop(paste(output, collapse = " "))
  invisible(output)
}

measure_aggregate <- function(workload, region, expected, threads, cpus,
                              repetitions = 9L) {
  set_affinity(cpus)
  invisible(dbExecute(con, sprintf("SET threads = %d", threads)))
  sql <- aggregate_sql(region)
  for (i in seq_len(2L)) stopifnot(same_answer(dbGetQuery(con, sql), expected))
  seconds <- numeric(repetitions)
  for (i in seq_len(repetitions)) {
    started <- proc.time()[["elapsed"]]
    observed <- dbGetQuery(con, sql)
    seconds[[i]] <- proc.time()[["elapsed"]] - started
    stopifnot(same_answer(observed, expected))
  }
  data.frame(
    workload = workload,
    threads = threads,
    affinity = cpus,
    repetitions = repetitions,
    output_rows = as.double(expected$rows[[1L]]),
    output_bytes = NA_real_,
    min_seconds = min(seconds),
    median_seconds = median(seconds),
    max_seconds = max(seconds),
    median_rows_per_second = as.double(expected$rows[[1L]]) / median(seconds),
    fingerprint = expected$fingerprint[[1L]],
    value_sum = expected$value_sum[[1L]],
    stringsAsFactors = FALSE
  )
}

measure_parquet <- function(threads, cpus, repetitions = 5L) {
  set_affinity(cpus)
  invisible(dbExecute(con, sprintf("SET threads = %d", threads)))
  output_path <- tempfile(sprintf("bigwig-t%d-", threads), fileext = ".parquet")
  copy_sql <- sprintf(
    paste0(
      "COPY (SELECT * FROM read_bigwig(%s, blocks_per_iteration := 64)) ",
      "TO %s (FORMAT PARQUET, COMPRESSION ZSTD)"
    ),
    sql_string(input_path), sql_string(output_path)
  )
  seconds <- numeric(repetitions)
  output_bytes <- numeric(repetitions)
  for (i in seq_len(repetitions)) {
    unlink(output_path)
    started <- proc.time()[["elapsed"]]
    invisible(dbExecute(con, copy_sql))
    seconds[[i]] <- proc.time()[["elapsed"]] - started
    output_bytes[[i]] <- file.info(output_path)$size
    observed <- dbGetQuery(
      con,
      aggregate_sql(relation = sprintf("read_parquet(%s)", sql_string(output_path)))
    )
    stopifnot(same_answer(observed, full_expected))
  }
  unlink(output_path)
  data.frame(
    workload = "full scan + ZSTD Parquet",
    threads = threads,
    affinity = cpus,
    repetitions = repetitions,
    output_rows = as.double(full_expected$rows[[1L]]),
    output_bytes = median(output_bytes),
    min_seconds = min(seconds),
    median_seconds = median(seconds),
    max_seconds = max(seconds),
    median_rows_per_second = as.double(full_expected$rows[[1L]]) / median(seconds),
    fingerprint = full_expected$fingerprint[[1L]],
    value_sum = full_expected$value_sum[[1L]],
    stringsAsFactors = FALSE
  )
}

results <- rbind(
  measure_aggregate("full scan + aggregate", NULL, full_expected, 1L, "12"),
  measure_aggregate("full scan + aggregate", NULL, full_expected, 4L, "12-15"),
  measure_aggregate(
    "20 indexed ranges + aggregate", region_spec, region_expected, 1L, "12"
  ),
  measure_aggregate(
    "20 indexed ranges + aggregate", region_spec, region_expected, 4L, "12-15"
  ),
  measure_parquet(1L, "12"),
  measure_parquet(4L, "12-15")
)

results$source_revision <- source_revision
results$run_date <- as.character(Sys.Date())
results$input_path <- input_path
results$input_bytes <- file.info(input_path)$size
results$input_span_bases <- 10000000
results$duckdb_version <- dbGetQuery(con, "SELECT version() AS v")$v[[1L]]
results$htslib_version <- rduckhts_htslib_version(con)

dir.create("benchmarks/data", showWarnings = FALSE, recursive = TRUE)
write.csv(
  results,
  "benchmarks/data/bigwig_reader.csv",
  row.names = FALSE,
  na = ""
)

cpu_model <- sub(
  "^model name[[:space:]]*:[[:space:]]*", "",
  grep("^model name", readLines("/proc/cpuinfo"), value = TRUE)[[1L]]
)
metadata <- data.frame(
  property = c(
    "source revision", "run date", "input source", "input interval",
    "local BigWig bytes", "full-scan stored intervals", "full-scan value sum",
    "full-scan fingerprint", "indexed ranges", "indexed output intervals",
    "DuckDB version", "htslib version", "CPU"
  ),
  value = c(
    source_revision,
    as.character(Sys.Date()),
    "UCSC hg38.phyloP100way.bw",
    "chr22:[20,000,000, 30,000,000)",
    format(file.info(input_path)$size, big.mark = ",", scientific = FALSE),
    format(as.double(full_expected$rows[[1L]]), big.mark = ",", scientific = FALSE),
    sprintf("%.12g", full_expected$value_sum[[1L]]),
    full_expected$fingerprint[[1L]],
    as.character(length(regions)),
    format(as.double(region_expected$rows[[1L]]), big.mark = ",", scientific = FALSE),
    results$duckdb_version[[1L]],
    results$htslib_version[[1L]],
    cpu_model
  ),
  stringsAsFactors = FALSE
)

knitr::kable(metadata, col.names = c("Property", "Recorded value"))
invisible(dbDisconnect(con, shutdown = TRUE))
```

## Results

The aggregate workloads decode the reader output but return one row to R. The
Parquet workload decodes and writes every interval as a real ZSTD-compressed
four-column relation. Every timed pass is followed by exact row-count and
order-independent full-row fingerprint validation; floating-point sums are
also checked to a relative tolerance of `1e-12`.

```{r results, echo=FALSE}
display <- results[c(
  "workload", "threads", "affinity", "repetitions", "output_rows",
  "output_bytes", "min_seconds", "median_seconds", "max_seconds",
  "median_rows_per_second"
)]
display$output_rows <- format(
  display$output_rows, big.mark = ",", scientific = FALSE
)
display$output_bytes <- ifelse(
  is.na(display$output_bytes), "--",
  format(display$output_bytes, big.mark = ",", scientific = FALSE)
)
for (column in c("min_seconds", "median_seconds", "max_seconds")) {
  display[[column]] <- sprintf("%.3f", display[[column]])
}
display$median_rows_per_second <- format(
  round(display$median_rows_per_second), big.mark = ",", scientific = FALSE
)
names(display) <- c(
  "Workload", "Threads", "CPU affinity", "Runs", "Output intervals",
  "Parquet bytes", "Minimum seconds", "Median seconds", "Maximum seconds",
  "Median intervals/s"
)
knitr::kable(display, align = c("l", rep("r", 9)))
```

The full scan contains one nonempty contig, so it deliberately tests whether
threading overhead remains bounded when contig-level parallel work is absent.
The separated indexed ranges expose independent tasks. The measurement does
not include remote latency, BigWig construction, extension loading, or a
downstream conservation reduction over variant spans.
