---
title: "DuckHTS Mosdepth Benchmark"
output: github_document
---

<!-- Benchmarks_mosdepth.md is generated from Benchmarks_mosdepth.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

# Benchmark

This benchmark:

- times native `duckhts_mosdepth(...)` against upstream `mosdepth`
- uses the native benchmark harness in `scripts/mosdepth_benchmark.py`
- verifies output equality on the first run of each case
- exercises whole-genome `fast`, `default`, `fragment`, and filtered default mode

# Run

`make bench-mosdepth`

First stage the registered whole-genome CRAM/reference input and derived BAM:

```sh
make stage-riker-wgs
```

Override defaults with:

- `MOSDEPTH_BENCH_BAM`
- `MOSDEPTH_BENCH_RUNS`
- `MOSDEPTH_BENCH_THREADS`
- `MOSDEPTH_BENCH_WINDOW`
- `MOSDEPTH_BENCH_EXTENSION`

```{r setup, include = FALSE}
shell_quote <- function(x) shQuote(x, type = "sh")

if (!nzchar(Sys.getenv("DUCKHTSBENCH_REGISTRY", unset = ""))) {
  Sys.setenv(DUCKHTSBENCH_REGISTRY = normalizePath("r/duckhtsbench/inst/benchmark_registry.tsv", mustWork = TRUE))
}
source(normalizePath("r/duckhtsbench/R/registry.R", mustWork = TRUE))
bench_bam <- normalizePath(
  Sys.getenv(
    "MOSDEPTH_BENCH_BAM",
    unset = duckhts_bench_artifact_path("riker_hg00188_bam")
  ),
  mustWork = TRUE
)
bench_extension <- normalizePath(
  Sys.getenv(
    "MOSDEPTH_BENCH_EXTENSION",
    unset = "build/release/duckhts.duckdb_extension"
  ),
  mustWork = TRUE
)
bench_runs <- as.integer(Sys.getenv("MOSDEPTH_BENCH_RUNS", unset = "1"))
bench_threads <- as.integer(Sys.getenv("MOSDEPTH_BENCH_THREADS", unset = "4"))
bench_window <- as.integer(Sys.getenv("MOSDEPTH_BENCH_WINDOW", unset = "1000000"))

stopifnot(nzchar(Sys.which("python3")))
stopifnot(nzchar(Sys.which("mosdepth")))
stopifnot(nzchar(Sys.which("samtools")))
stopifnot(file.exists(paste0(bench_bam, ".bai")) || file.exists(sub("\\.bam$", ".crai", bench_bam)))
```

```{r helpers, include = FALSE}
extract_tool_seconds <- function(lines, tool) {
  idx <- grep(paste0("^", tool, "\\s+[0-9]"), lines)
  if (!length(idx)) {
    stop(sprintf("Could not find timing line for %s", tool), call. = FALSE)
  }
  fields <- strsplit(trimws(lines[idx[[1]]]), "\\s+")[[1]]
  as.numeric(sub("s$", "", fields[[2]]))
}

extract_ratio <- function(lines) {
  idx <- grep("^Ratio: DuckHTS is ", lines)
  if (!length(idx)) {
    stop("Could not find speed ratio", call. = FALSE)
  }
  sub("^Ratio: DuckHTS is ", "", trimws(lines[idx[[1]]]))
}

extract_verification <- function(lines) {
  idx <- grep("PASS \\(", lines)
  if (!length(idx)) {
    return(NA_character_)
  }
  trimws(lines[idx[[length(idx)]]])
}

run_case <- function(case) {
  cmd <- c(
    "scripts/mosdepth_benchmark.py",
    bench_bam,
    "--mode", case$mode,
    "--window-size", as.character(bench_window),
    "--no-per-base",
    "--threads", as.character(bench_threads),
    "--runs", as.character(bench_runs),
    "--extension", bench_extension,
    "--verify"
  )
  if (!is.null(case$extra_args)) {
    cmd <- c(cmd, case$extra_args)
  }

  out <- system2("python3", cmd, stdout = TRUE, stderr = TRUE)
  status <- attr(out, "status")
  if (!is.null(status) && status != 0L) {
    stop(paste(out, collapse = "\n"), call. = FALSE)
  }

  lines <- trimws(out)
  data.frame(
    case = case$case,
    mode = case$mode,
    filter = case$filter,
    mosdepth_sec = extract_tool_seconds(lines, "mosdepth"),
    duckhts_sec = extract_tool_seconds(lines, "duckhts_mosdepth"),
    speedup = extract_ratio(lines),
    verification = extract_verification(lines),
    command = paste(c("python3", vapply(cmd, shell_quote, character(1))), collapse = " "),
    raw_output = paste(out, collapse = "\n"),
    stringsAsFactors = FALSE
  )
}
```

```{r benchmark-cases, include = FALSE}
cases <- list(
  list(
    case = "whole_genome_fast",
    mode = "fast",
    filter = "default filters"
  ),
  list(
    case = "whole_genome_default",
    mode = "default",
    filter = "default filters"
  ),
  list(
    case = "whole_genome_fragment",
    mode = "fragment",
    filter = "default filters"
  ),
  list(
    case = "whole_genome_default_mapq60_include64",
    mode = "default",
    filter = "MAPQ >= 60, include_flag = 64",
    extra_args = c("--mapq", "60", "--include-flag", "64")
  )
)

benchmark_results <- do.call(
  rbind,
  lapply(cases, run_case)
)
```

# Inputs

```{r inputs, echo = FALSE}
data.frame(
  bam = bench_bam,
  extension = bench_extension,
  runs = bench_runs,
  threads = bench_threads,
  window_size = bench_window,
  stringsAsFactors = FALSE
)
```

# Whole-Genome Cases

```{r benchmark-summary, echo = FALSE}
benchmark_results[, c("case", "mode", "filter", "mosdepth_sec", "duckhts_sec", "speedup", "verification")]
```

# Commands

```{r commands, echo = FALSE}
benchmark_results[, c("case", "command")]
```

# Raw Output

```{r raw-output, results = 'asis', echo = FALSE}
for (i in seq_len(nrow(benchmark_results))) {
  cat("## ", benchmark_results$case[[i]], "\n\n", sep = "")
  cat("```text\n")
  cat(benchmark_results$raw_output[[i]])
  cat("\n```\n\n")
}
```
