---
title: "Automatic BAM full-file scan: small-fixture cost"
output:
  github_document:
    html_preview: false
---

This compares automatic BAM scans at two named reader revisions,
using the existing `benchmark_simd_bam_gc.py` workload on its committed nanopore
fixture. It exercises many empty contig claims and full `SEQ` decoding, but this
186-record fixture is dominated by setup cost: it is not a WGS throughput benchmark.
There are no unplaced reads in this performance input, so both builds do equal work.
The separate SQL/R regression corpus tests retained no-coordinate records,
physical duplicates, BAI/CSI, CRAM, and output chunks.

The nearest checked-in report is
[`benchmark_simd_bam_gc.md`](benchmark_simd_bam_gc.md), whose historical run used
3,245,905 HG00106 exome reads / 327,836,405 bases and one DuckDB worker. That input
is not staged here; its timings are not a comparable regression baseline. The
same-run comparison below uses the current driver's committed default input.
Remote I/O, large-input throughput, and CRAM performance remain unmeasured.

## Reproduction and identity

Build each named source revision separately. Set `BAM_SCAN_BASELINE_EXTENSION` to
the baseline binary, `BAM_SCAN_BASELINE_REVISION` to its revision, then render this
report from the repository root. `DUCKHTS_EXTENSION` selects the candidate binary;
`PYTHON` selects the Python environment with DuckDB installed.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r measurement, echo=FALSE}
library(jsonlite)
baseline <- Sys.getenv("BAM_SCAN_BASELINE_EXTENSION")
baseline_revision <- Sys.getenv("BAM_SCAN_BASELINE_REVISION")
candidate <- Sys.getenv("DUCKHTS_EXTENSION", "build/release/duckhts.duckdb_extension")
python <- Sys.getenv("PYTHON", "configure/venv/bin/python3")
stopifnot(nzchar(baseline_revision), file.exists(baseline), file.exists(candidate))
revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE)
source_tree <- system2("git", c("rev-parse", "HEAD:src"), stdout = TRUE)
input <- "test/data/nanopore.bam"
identity <- data.frame(
  property = c("baseline revision", "candidate revision", "candidate src tree",
               "baseline binary MD5", "candidate binary MD5", "input", "input MD5",
               "input bytes", "HTSlib", "host", "CPU affinity", "repetitions"),
  value = c(baseline_revision, revision, source_tree, unname(tools::md5sum(baseline)),
            unname(tools::md5sum(candidate)), input, unname(tools::md5sum(input)),
            file.info(input)$size, readLines("third_party/htslib/VERSION", n = 1),
            Sys.info()[["sysname"]],
            system2("taskset", c("-pc", Sys.getpid()), stdout = TRUE),
            "9 per build/thread setting after warm-up; fresh Python process per setting")
)
knitr::kable(identity)
results <- list()
for (threads in c(1L, 4L)) {
  for (build in c("baseline", "candidate")) {
    output <- tempfile(fileext = ".json")
    ext <- if (build == "baseline") baseline else candidate
    args <- c("scripts/benchmark_simd_bam_gc.py", "--extension", ext,
              "--bam", input, "--max-reads", "0", "--iterations", "9",
              "--threads", threads, "--backends", "auto", "--modes", "bam_scan",
              "--json-out", output)
    log <- system2(python, shQuote(args), stdout = TRUE, stderr = TRUE)
    stopifnot(is.null(attr(log, "status")) || attr(log, "status") == 0L)
    measured <- fromJSON(output, simplifyVector = FALSE)[[1]]
    stopifnot(!measured$skipped, measured$reads == 186L)
    results[[length(results) + 1L]] <- data.frame(
      build = build, duckdb_workers = threads, htslib_workers_per_handle = 2L,
      input_records = 186L, decoded_sequence_rows = measured$reads,
      input_bases = measured$total_bases, sql_output_rows = 1L,
      gc_sum = measured$gc_fraction_sum, backend = measured$kernel_backend,
      median_seconds = measured$median_sec, min_seconds = measured$min_sec,
      max_seconds = max(unlist(measured$timings_sec))
    )
    unlink(output)
  }
}
results <- do.call(rbind, results)
stopifnot(length(unique(results$input_bases)) == 1L,
          max(results$gc_sum) - min(results$gc_sum) < 1e-9)
knitr::kable(results, digits = 6)
```

Each query scans 186 BAM records, materializes their sequences into DuckDB vectors,
and returns one aggregate row containing the read count, total bases, and sum of
GC fractions. All nine repetitions validate those values; equality here is not
a substitute for the independent SAM-field and row-multiset regression tests.
No claim of general performance parity or speedup follows from these small timings.
