---
title: "DuckVEP sorted-stream throughput"
output: github_document
---

<!-- duckvep_throughput.md is generated from duckvep_throughput.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
history <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_throughput.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
history$revision <- substr(history$source_revision, 1, 8)
```

This benchmark drives `duckvep_annotate` through DuckDB's stable extension API
with nondecreasing sequence-region and position keys. It records vector adapter,
sweep reuse, coding classification, list materialization, `unnest`, and final
aggregation together.

The current checked-in workload has one transcript and two exons. It is a
repeatable hot-transcript floor, not a whole-genome throughput claim. Production
qualification also requires an imported Ensembl model and a sorted gnomAD-scale
variant stream spanning realistic transcript density.

## Recorded runs

```{r runs}
table <- history[, c(
  "run_date", "revision", "workload", "threads", "variants", "passes",
  "min_seconds", "median_seconds", "max_seconds", "variants_per_second",
  "ns_per_variant", "cpu"
)]
table$variants <- format(table$variants, big.mark = ",", scientific = FALSE)
table$variants_per_second <- round(table$variants_per_second)
table$ns_per_variant <- round(table$ns_per_variant, 1)
knitr::kable(table, row.names = FALSE)
```

Each pass consumes every generated input and checks output cardinality and a
consequence-string checksum. Rows with different workload, thread count, host,
or variant count are separate measurements and should not be compared as if
only the engine changed.
