---
title: "DuckVEP sorted-stream throughput"
output: github_document
---

<!-- duckvep_throughput.md is generated from duckvep_throughput.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
history <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_throughput.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
history$revision <- substr(history$source_revision, 1, 8)
```

This benchmark drives either the rich `duckvep_annotate` result or the numeric
`duckvep_annotate_compact` result through DuckDB's stable extension API with
nondecreasing sequence-region and position keys. It records candidate sweep,
consequence classification, list materialization, `unnest`, and final aggregation
together. Model loading and input staging are outside the timed pass.

The checked-in one-transcript workload is a repeatable adapter floor. The
production-density workloads use a pre-staged DuckDB database containing
`bench_regions`, `bench_transcripts`, `bench_exons`, and `bench_variants`. The
earlier chromosome-only staging model contains 25 sequence regions, 644,292
transcripts, and 5,078,384 exon memberships. The complete-primary-assembly
model contains 194 sequence regions, 646,577 transcripts, and 5,087,789 exon
memberships; 370,580 coding transcripts have prepared CDS sequence. Both use
the same deterministic 1-in-40 hash sample of 100,957 sorted GIAB sites. These
are distinct model workloads and their consequence checksums are not expected
to match. The external staging databases are not committed.
The production query restates `ORDER BY` in the timed input CTE, so its result
includes the cost of enforcing the kernel's nondecreasing-input contract even
though the staging table was already written in that order. The `79cfaaf2`
rows predate that review correction and relied on the single-thread physical
scan preserving insertion order; they remain only as historical measurements.

## Recorded runs

```{r runs}
table <- history[, c(
  "run_date", "revision", "workload", "output_mode", "threads", "variants",
  "transcripts", "exons", "annotated_rows", "passes", "min_seconds",
  "median_seconds", "max_seconds", "variants_per_second", "ns_per_variant",
  "cpu"
)]
table$variants <- format(table$variants, big.mark = ",", scientific = FALSE)
table$transcripts <- format(table$transcripts, big.mark = ",", scientific = FALSE)
table$exons <- format(table$exons, big.mark = ",", scientific = FALSE)
table$annotated_rows <- format(
  table$annotated_rows,
  big.mark = ",",
  scientific = FALSE
)
table$variants_per_second <- round(table$variants_per_second)
table$ns_per_variant <- round(table$ns_per_variant, 1)
knitr::kable(table, row.names = FALSE)
```

Each pass consumes every staged input and checks output cardinality plus either
the rendered consequence-byte total or the numeric consequence-mask sum. Rows
with different workload, output mode, thread count, host, or variant count are
separate measurements and should not be compared as if only the engine changed.

The production form is run with:

```sh
Rscript benchmarks/duckvep_throughput.R \
  --database /path/to/staged.duckdb \
  --variants 100957 --passes 9 --warmup 10000 --threads 1 \
  --output compact
```
