---
title: "DuckVEP conformance against Ensembl VEP 116"
output: github_document
---

<!-- duckvep_conformance.md is generated from duckvep_conformance.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
history_path <- file.path(
  root, "test", "duckvep", "conformance", "data", "conformance_history.csv"
)
history <- read.csv(
  history_path,
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(oracle_version = "character")
)
property_history <- read.csv(
  file.path(
    root, "test", "duckvep", "conformance", "data", "property_history.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character", seed = "character")
)
property_coverage_history <- read.csv(
  file.path(
    root,
    "test",
    "duckvep",
    "conformance",
    "data",
    "property_coverage_history.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character", seed = "character")
)
hgvs_history <- read.csv(
  file.path(
    root, "test", "duckvep", "conformance", "data", "hgvs_history.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character", oracle_version = "character")
)
release_parquet <- read.csv(
  file.path(
    root, "benchmarks", "data", "duckvep_release_parquet.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character", chromosome = "character")
)
model_receipts <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_model_receipts.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character", source_version = "character")
)
overall <- history[history$stratum_kind == "all", , drop = FALSE]
revision_ancestry <- suppressWarnings(system2(
  "git",
  c("-C", shQuote(root), "rev-list", "--topo-order", "HEAD"),
  stdout = TRUE,
  stderr = FALSE
))
corpus_labels <- c(
  final_dbsnp157_windows_seed29 = "GRCh38 dbSNP",
  final_giab_grch38_seed71 = "GRCh38 GIAB",
  final_clinvar_coding_seed113 = "GRCh38 ClinVar coding",
  final_clinvar_crosschrom_seed17 = "GRCh38 ClinVar cross-chromosome",
  final_grch37_cache_seed37 = "GRCh37",
  `plasmodium-falciparum-vep63-seed11663` = "P. falciparum",
  breakend_multichrom_seed31_isolated = "GRCh38 paired BND",
  regulation_giab_chr21_seed1 = "GRCh38 GIAB + core regulation",
  regulation_sv_chr21_seed17 = "GRCh38 exact SV + core regulation"
)
latest_runs <- do.call(rbind, lapply(names(corpus_labels), function(corpus) {
  rows <- overall[overall$corpus == corpus, , drop = FALSE]
  if (!nrow(rows)) return(NULL)
  tested_ancestors <- revision_ancestry[
    revision_ancestry %in% rows$source_revision
  ]
  revision <- if (length(tested_ancestors)) {
    tested_ancestors[[1L]]
  } else {
    rows$source_revision[[nrow(rows)]]
  }
  rows[rows$source_revision == revision, , drop = FALSE]
}))
latest_runs$corpus_label <- unname(corpus_labels[latest_runs$corpus])
latest_runs <- latest_runs[order(
  match(latest_runs$corpus, names(corpus_labels))
), , drop = FALSE]
state_overall <- overall[
  grepl("^state_exploration_seed_[0-9]+$", overall$corpus),
  ,
  drop = FALSE
]
state_revisions <- unique(state_overall$source_revision)
state_ancestry <- revision_ancestry[revision_ancestry %in% state_revisions]
latest_state_revision <- if (length(state_ancestry)) {
  state_ancestry[[1L]]
} else if (nrow(state_overall)) {
  state_overall$source_revision[[nrow(state_overall)]]
} else {
  ""
}
latest_state_runs <- state_overall[
  state_overall$source_revision == latest_state_revision,
  ,
  drop = FALSE
]
hgvs_runs <- unique(hgvs_history[, c(
  "source_revision", "corpus", "model", "oracle_version", "oracle_build"
), drop = FALSE])
latest_hgvs_run_ids <- NULL
if (nrow(hgvs_runs)) {
  groups <- interaction(hgvs_runs$corpus, hgvs_runs$model, drop = TRUE)
  latest_hgvs_run_ids <- do.call(rbind, lapply(
    split(hgvs_runs, groups),
    function(rows) {
      tested_ancestors <- revision_ancestry[
        revision_ancestry %in% rows$source_revision
      ]
      revision <- if (length(tested_ancestors)) {
        tested_ancestors[[1L]]
      } else {
        rows$source_revision[[nrow(rows)]]
      }
      rows[rows$source_revision == revision, , drop = FALSE]
    }
  ))
}
nmd_history <- history[
  history$stratum_kind == "nmd_prediction",
  ,
  drop = FALSE
]
nmd_run_keys <- c(
  "source_revision", "engine", "corpus", "model", "oracle_version",
  "oracle_build"
)
latest_nmd_run_ids <- NULL
if (nrow(nmd_history)) {
  nmd_runs <- unique(nmd_history[, nmd_run_keys, drop = FALSE])
  groups <- interaction(
    nmd_runs$engine,
    nmd_runs$corpus,
    nmd_runs$model,
    drop = TRUE
  )
  latest_nmd_run_ids <- do.call(rbind, lapply(
    split(nmd_runs, groups),
    function(rows) {
      tested_ancestors <- revision_ancestry[
        revision_ancestry %in% rows$source_revision
      ]
      revision <- if (length(tested_ancestors)) {
        tested_ancestors[[1L]]
      } else {
        rows$source_revision[[nrow(rows)]]
      }
      rows[rows$source_revision == revision, , drop = FALSE]
    }
  ))
}
upper95 <- function(k, n) {
  if (n == 0) NA_real_ else binom.test(k, n, conf.level = 0.95)$conf.int[[2L]]
}
percent <- function(x) {
  ifelse(is.na(x), "not estimable", sprintf("%.2f%%", 100 * x))
}
short_revision <- function(x) substr(x, 1, 8)
oracle_field <- function(x, field) {
  pattern <- paste0("(^|;)", field, "=([^;]+)")
  found <- regexec(pattern, x)
  values <- regmatches(x, found)
  vapply(values, function(value) {
    if (length(value) >= 3L) value[[3L]] else ""
  }, character(1L))
}
```

This report records exact consequence agreement with the real Ensembl VEP 116
executable at the object VEP annotates: a transcript, RegulatoryFeature, or
MotifFeature. It uses either VEP's declared indexed cache or a staged GFF oracle.
Unresolved DuckVEP rows remain in the denominator; they are not discarded as
unsupported cases. The CSV is append-only by source revision, corpus, and
resident model. Independent frozen distributions and seeds are kept separate so
a fix cannot improve its own hand-picked witnesses and hide a regression
elsewhere.

Official Ensembl variation release VCFs provide a second, precomputed oracle lane:
their CSQ projection can be compared in ordinary CI without starting Perl VEP. Pinned
release shards are fast regression evidence for known alleles; executable-VEP and
generated-corpus lanes remain necessary for novel allele states, option-dependent
semantics, phased combinations, and sampled structural geometry. Matching full models
belong in receipt-hashed external artifacts (for example a versioned Zenodo record), not
in git or the network-free extension build step.

## Declared conformance closure

The independent-event consequence engine is closed as a semantic implementation campaign
for the declared model and event surfaces: admitted Ensembl transcript, mature-miRNA,
RegulatoryFeature, and MotifFeature objects; independent literal small alleles; exact typed
DEL, DUP, tandem-DUP, INV, INS, and CNV events; structural tandem repeats (`STR`); paired
breakends; supported BioPerl codon tables and exceptional Ensembl peptide edits; and the
separately declared VEP NMD-plugin result. DEL/DUP/tandem-DUP/INV/INS/CNV and BND have
generated executable-VEP differentials. Structural `STR` has source-derived VEP-116
semantics plus fixed SQL/R and randomized C coverage; raw repeat reconstruction is a
separate input-preparation operation. The evidence spans GRCh38, GRCh37, and
*P. falciparum*, executable witnesses, indexed-cache corpora, generated state exploration,
sanitizer runs, and pure-C oracle properties.

“Closed” means future consequence changes are routine engineering behind these regression
gates. VEP 116 parses `CIPOS`/`CIEND` into inner/outer structural coordinates, but its
registered consequence predicates use nominal `POS`/`END`; DuckVEP therefore annotates
that nominal span while the surrounding relation preserves the uncertainty metadata.
The checked-in 12-record GRCh38 confidence witness records this directly: nominal and
`IMPRECISE;CIPOS;CIEND` forms of CNV, DEL, DUP, tandem DUP, INV, and INS produced 466/466
exact transcript pairs, and both engines had equal nominal/imprecise consequence
multisets for all six event-kind pairs.
VEP can also expand a bounded `<CNV:TR>` from `RN`, `RUS`, and `RUC` or `RB` into a literal
allele before consequence calculation. Implementing that lossless expansion and mapping
VEP's finite supported symbolic vocabulary into the typed event API are narrower input-
preparation tasks, not missing consequence predicates. VEP itself rejects unrecognised
types such as CPX, so this closure does not promise arbitrary symbolic parsing. Untested
species/releases and phased multi-record haplotypes remain outside the closure. Haplotype
grouping and combined consequence attribution are the next semantic vertical and require
their own executable oracle and performance campaign. Any newly observed fixed-event
mismatch reopens this contract rather than being relabelled as unsupported.

## Latest tested revision per corpus

```{r latest}
latest_table <- data.frame(
  revision = short_revision(latest_runs$source_revision),
  corpus = latest_runs$corpus_label,
  model = latest_runs$model,
  assembly = oracle_field(latest_runs$oracle_build, "assembly"),
  species = oracle_field(latest_runs$oracle_build, "species"),
  oracle_source = oracle_field(latest_runs$oracle_build, "oracle"),
  oracle = paste0("VEP ", latest_runs$oracle_version),
  pairs = latest_runs$n,
  exact = paste0(latest_runs$exact_agree, "/", latest_runs$n),
  unresolved = latest_runs$unresolved,
  resolved_disagreements = latest_runs$resolved_discordant,
  resolved_error_upper_95 = percent(latest_runs$upper95),
  stringsAsFactors = FALSE
)
knitr::kable(latest_table, row.names = FALSE)
```

Each row is the newest tested ancestor of the current source for that named
corpus. Expensive corpora do not inherit evidence from a later run of another
corpus, and a newly tested corpus does not hide older still-applicable evidence.
The SO and impact tables keep the same runs separate so the largest corpus cannot
hide a smaller species- or assembly-specific frontier.

## Independent-event HGVS differential

```{r hgvs}
if (!is.null(latest_hgvs_run_ids) && nrow(latest_hgvs_run_ids)) {
  hgvs_latest <- merge(
    latest_hgvs_run_ids,
    hgvs_history,
    by = c(
      "source_revision", "corpus", "model", "oracle_version", "oracle_build"
    ),
    all.x = TRUE,
    sort = FALSE
  )
  hgvs_table <- do.call(rbind, lapply(
    split(hgvs_latest, interaction(hgvs_latest$corpus, hgvs_latest$metric)),
    function(rows) {
      exact <- sum(rows$n[rows$comparison %in% c("match", "both_absent")])
      total <- sum(rows$n)
      data.frame(
        revision = short_revision(rows$source_revision[[1L]]),
        corpus = rows$corpus[[1L]],
        model = rows$model[[1L]],
        metric = toupper(rows$metric[[1L]]),
        exact = paste0(
          format(exact, big.mark = ","),
          "/",
          format(total, big.mark = ",")
        ),
        match = sum(rows$n[rows$comparison == "match"]),
        both_absent = sum(rows$n[rows$comparison == "both_absent"]),
        discordant = sum(
          rows$n[!rows$comparison %in% c("match", "both_absent")]
        ),
        stringsAsFactors = FALSE
      )
    }
  ))
  hgvs_table <- hgvs_table[order(hgvs_table$corpus, hgvs_table$metric), ]
  knitr::kable(hgvs_table, row.names = FALSE)
} else {
  cat("No revision-keyed HGVS differential has been recorded.\n")
}
```

This is exact string agreement for independent transcript events with VEP 116
invoked using `--hgvs`. A comparison is exact when both engines emit the same
string or both omit that HGVS field. Unresolved, missing, extra, and unequal
strings remain discordant; none is removed from the denominator.

## Paired-breakend differential

```{r paired-breakends}
breakend_run <- latest_runs[
  latest_runs$corpus == "breakend_multichrom_seed31_isolated",
  ,
  drop = FALSE
]
if (nrow(breakend_run)) {
  breakend_table <- data.frame(
    revision = short_revision(breakend_run$source_revision),
    generated_events = format(1004, big.mark = ","),
    transcript_pairs = format(breakend_run$n, big.mark = ","),
    exact = paste0(
      format(breakend_run$exact_agree, big.mark = ","),
      "/",
      format(breakend_run$n, big.mark = ",")
    ),
    unresolved = breakend_run$unresolved,
    extra = breakend_run$engine_extra,
    missing = breakend_run$engine_missing,
    resolved_error_upper_95 = percent(breakend_run$upper95),
    stringsAsFactors = FALSE
  )
  knitr::kable(breakend_table, row.names = FALSE)
}
```

The generated seed-31 corpus spans chromosomes 1, 2, 7, 21, and X; intra- and
interchromosomal mates; all four VCF bracket orientations; and transcript,
exon, intron, CDS, and directional-flank endpoint states. The comparison is the
union of consequences produced by both breakend endpoints for each transcript,
which is VEP 116's transcript-level paired-breakend contract.

VEP 116's buffered BND path inserts mate coordinates into a chromosome-blind
interval tree. A multichromosome batch can therefore omit valid transcript
pairs even when the input chromosomes are contiguous and position-sorted. The
oracle command uses `--buffer_size 1` in one Perl process so every event is
evaluated independently. This is oracle isolation, not a DuckVEP compatibility
rule; the ledger records `breakend_buffer_size=1` and the artifact hash.

## Core regulation and motif differential

```{r regulation}
regulation_corpora <- c(
  regulation_giab_chr21_seed1 = "GIAB chromosome 21",
  regulation_sv_chr21_seed17 = "generated exact SVs"
)
regulation_inputs <- c(
  regulation_giab_chr21_seed1 = 1196,
  regulation_sv_chr21_seed17 = 2700
)
regulation_runs <- latest_runs[
  latest_runs$corpus %in% names(regulation_corpora),
  ,
  drop = FALSE
]
if (nrow(regulation_runs)) {
  regulation_table <- data.frame(
    revision = short_revision(regulation_runs$source_revision),
    workload = unname(regulation_corpora[regulation_runs$corpus]),
    input_events = format(
      unname(regulation_inputs[regulation_runs$corpus]),
      big.mark = ","
    ),
    object_pairs = format(regulation_runs$n, big.mark = ","),
    exact = paste0(
      format(regulation_runs$exact_agree, big.mark = ","),
      "/",
      format(regulation_runs$n, big.mark = ",")
    ),
    unresolved = regulation_runs$unresolved,
    extra = regulation_runs$engine_extra,
    missing = regulation_runs$engine_missing,
    exact_error_upper_95 = percent(regulation_runs$upper95),
    stringsAsFactors = FALSE
  )
  knitr::kable(regulation_table, row.names = FALSE)

  regulation_terms <- c(
    "TFBS_ablation", "TFBS_amplification", "TF_binding_site_variant",
    "regulatory_region_ablation", "regulatory_region_amplification",
    "regulatory_region_variant"
  )
  regulation_revisions <- setNames(
    regulation_runs$source_revision,
    regulation_runs$corpus
  )
  term_rows <- history[
    history$stratum_kind == "so_term" &
      history$consequence_class %in% regulation_terms &
      history$corpus %in% names(regulation_revisions),
    ,
    drop = FALSE
  ]
  term_rows <- term_rows[
    term_rows$source_revision == unname(regulation_revisions[term_rows$corpus]),
    ,
    drop = FALSE
  ]
  if (nrow(term_rows)) {
    term_counts <- aggregate(
      cbind(n, exact_agree) ~ corpus + consequence_class,
      term_rows,
      sum
    )
    term_counts$workload <- unname(regulation_corpora[term_counts$corpus])
    term_counts$memberships <- format(term_counts$n, big.mark = ",")
    term_counts$exact <- paste0(
      format(term_counts$exact_agree, big.mark = ","),
      "/",
      format(term_counts$n, big.mark = ",")
    )
    term_counts <- term_counts[
      order(
        match(term_counts$consequence_class, regulation_terms),
        match(term_counts$corpus, names(regulation_corpora))
      ),
      c("workload", "consequence_class", "memberships", "exact"),
      drop = FALSE
    ]
    knitr::kable(term_counts, row.names = FALSE)
  }
}
```

The GIAB run checks ordinary alleles against transcript and core funcgen objects.
The generated structural run deliberately crosses, contains, exactly matches,
and partially overlaps RegulatoryFeature and MotifFeature intervals under DEL,
DUP, TDUP, INV, INS, and CNV operations. Structural `STR` is covered separately by
the source-derived VEP-116 rule, fixed SQL/R adapter tests, and randomized C oracles;
this generated executable-VEP run does not reconstruct raw repeat metadata. The resident model
contains only VEP-admitted core funcgen objects: VEP 116 removes
`epigenetically_modified_region` rows before constructing RegulatoryFeature
overlap objects, so DuckVEP excludes them during deterministic model preparation
rather than filtering output after candidate traversal.

## Prepared model receipts

```{r model-receipts}
receipt_table <- data.frame(
  revision = short_revision(model_receipts$source_revision),
  species = model_receipts$species,
  release = model_receipts$source_version,
  assembly = model_receipts$assembly,
  regions = format(model_receipts$region_count, big.mark = ","),
  transcripts = format(model_receipts$transcript_count, big.mark = ","),
  coding_backed = format(
    model_receipts$sequence_backed_transcript_count,
    big.mark = ","
  ),
  exons = format(model_receipts$exon_membership_count, big.mark = ","),
  mature_miRNA_segments = format(
    model_receipts$mature_mirna_segment_count,
    big.mark = ","
  ),
  peptide_edits = format(model_receipts$peptide_edit_count, big.mark = ","),
  regulatory_regions = format(
    model_receipts$regulatory_region_count,
    big.mark = ","
  ),
  motif_features = format(model_receipts$motif_feature_count, big.mark = ","),
  codon_tables = model_receipts$codon_table_counts,
  model_sha256 = substr(model_receipts$model_sha256, 1, 12),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
knitr::kable(receipt_table, row.names = FALSE)
```

These are complete model-build receipts, not counts inferred from a differential.
The ledger retains the full source-manifest, reference, and model SHA-256 values,
the exact VEP transcript filter, every count above, CDS/flank base totals, and the
external artifact name. The Plasmodium row is an Ensembl Genomes release-63 cache
paired with the VEP/core-116 executable libraries, which is why both release
numbers are recorded.

## History

```{r history}
timeline <- overall[, c(
  "run_date", "source_revision", "corpus", "model", "n", "exact_agree",
  "unresolved", "resolved_n", "resolved_discordant", "upper95"
)]
timeline$source_revision <- short_revision(timeline$source_revision)
timeline$exact_rate <- percent(timeline$exact_agree / timeline$n)
timeline$resolved_error_upper_95 <- percent(timeline$upper95)
timeline$upper95 <- NULL
knitr::kable(timeline, row.names = FALSE)
```

## Randomized executable-VEP state exploration

This is the anti-overfitting lane against the VEP executable, not an internal
property test. Each seed contains the fixed predicate witnesses plus 100,000
unique alleles. Three quarters of random positions are within six bases of
splice, exon, translation-start, and translation-stop boundaries; one quarter
is uniform across the transcript. SNVs, MNVs, insertions, deletions, and delins
are sampled with equal probability, with differing alleles up to 49 bases.

```{r state-exploration}
if (nrow(latest_state_runs)) {
  state_table <- data.frame(
    revision = short_revision(latest_state_runs$source_revision),
    seed = sub("^state_exploration_seed_", "", latest_state_runs$corpus),
    pairs = latest_state_runs$n,
    exact = latest_state_runs$exact_agree,
    unresolved = latest_state_runs$unresolved,
    resolved_disagreements = latest_state_runs$resolved_discordant,
    exact_error_upper_95_ppm = mapply(
      upper95,
      latest_state_runs$n - latest_state_runs$exact_agree,
      latest_state_runs$n
    ) * 1e6,
    stringsAsFactors = FALSE
  )
  state_total_n <- sum(latest_state_runs$n)
  state_total_exact <- sum(latest_state_runs$exact_agree)
  state_table <- rbind(
    state_table,
    data.frame(
      revision = short_revision(latest_state_revision),
      seed = "combined",
      pairs = state_total_n,
      exact = state_total_exact,
      unresolved = sum(latest_state_runs$unresolved),
      resolved_disagreements = sum(latest_state_runs$resolved_discordant),
      exact_error_upper_95_ppm = upper95(
        state_total_n - state_total_exact,
        state_total_n
      ) * 1e6,
      stringsAsFactors = FALSE
    )
  )
  for (column in c("pairs", "exact", "unresolved", "resolved_disagreements")) {
    state_table[[column]] <- format(
      state_table[[column]],
      big.mark = ",",
      scientific = FALSE
    )
  }
  state_table$exact_error_upper_95_ppm <- sprintf(
    "%.1f",
    state_table$exact_error_upper_95_ppm
  )
  knitr::kable(state_table, row.names = FALSE)
} else {
  cat("No randomized executable-VEP state exploration is recorded.\n")
}
```

The same campaign covered the following SO terms. Counts are term memberships,
not distinct transcript pairs, because one pair may carry several terms and the
four seeds deliberately retain the same fixed witnesses.

```{r state-exploration-so}
if (nrow(latest_state_runs)) {
  state_run_ids <- unique(latest_state_runs[, c(
    "source_revision", "engine", "corpus", "model", "oracle_version", "oracle_build"
  )])
  state_term_rows <- merge(
    history[history$stratum_kind == "so_term", , drop = FALSE],
    state_run_ids,
    by = c(
      "source_revision", "engine", "corpus", "model", "oracle_version", "oracle_build"
    )
  )
  state_term_counts <- aggregate(
    cbind(n, unresolved, term_mismatch, engine_extra, engine_missing) ~
      consequence_class + impact,
    data = state_term_rows,
    FUN = sum
  )
  state_seed_counts <- aggregate(
    corpus ~ consequence_class + impact,
    data = state_term_rows,
    FUN = function(values) length(unique(values))
  )
  names(state_seed_counts)[names(state_seed_counts) == "corpus"] <- "seeds_observed"
  state_term_counts <- merge(
    state_term_counts,
    state_seed_counts,
    by = c("consequence_class", "impact")
  )
  state_term_counts <- state_term_counts[order(
    -state_term_counts$n,
    state_term_counts$consequence_class
  ), c(
    "consequence_class", "impact", "seeds_observed", "n", "unresolved",
    "term_mismatch", "engine_extra", "engine_missing"
  )]
  knitr::kable(state_term_counts, row.names = FALSE)
}
```

This distribution deliberately stresses local allele and boundary states on one
engineered transcript. It does not replace the indexed-cache corpora, which add
real transcript density, imported flags, exceptional peptide edits, codon tables,
assemblies, and species. The revision is shown explicitly because this expensive
campaign is not silently attributed to later code.

## Official Ensembl release corpus in Parquet

The official release consequence VCF is already BGZF-compressed. This table measures its
complete typed DuckHTS reader projection and the narrower `VE` plus CSQ projection used by
the bulk oracle lane. It is a storage comparison, not a claim that the Parquet projection
can reproduce the original VCF byte-for-byte.

The consequence projection is also the natural CI payload: retain deterministic shards
with source URL, Ensembl release/species/assembly, artifact digest, row cardinalities, and
CSQ schema. A scheduled full-release job may pair the complete projection with a published
receipt-hashed DuckDB model. Neither form broadens the supported consequence contract; it
only makes a large known-variant regression cheap to replay.

```{r release-parquet}
last_storage <- release_parquet[nrow(release_parquet), , drop = FALSE]
storage <- release_parquet[
  release_parquet$source_revision == last_storage$source_revision &
    release_parquet$source_sha256 == last_storage$source_sha256,
  ,
  drop = FALSE
]
storage <- storage[match(c("full_typed", "consequence"), storage$projection), ]
storage_table <- data.frame(
  revision = short_revision(storage$source_revision),
  release = storage$ensembl_release,
  assembly = storage$assembly,
  chromosome = storage$chromosome,
  projection = storage$projection,
  columns = storage$columns,
  records = format(storage$records, big.mark = ",", scientific = FALSE),
  ALT_alleles = format(
    storage$alternate_alleles,
    big.mark = ",",
    scientific = FALSE
  ),
  CSQ_entries = format(storage$csq_entries, big.mark = ",", scientific = FALSE),
  source_MiB = sprintf("%.1f", storage$source_bytes / 2^20),
  parquet_MiB = sprintf("%.1f", storage$parquet_bytes / 2^20),
  parquet_of_source = sprintf(
    "%.1f%%",
    100 * storage$parquet_fraction_of_source
  ),
  elapsed_seconds = sprintf("%.1f", storage$elapsed_seconds),
  records_per_second = format(
    round(storage$records_per_second),
    big.mark = ",",
    scientific = FALSE
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
knitr::kable(storage_table, row.names = FALSE)
```

The ledger records the official source URL, SHA-256 of every input and output, DuckHTS and
DuckDB versions, compression, row-group size, thread count, machine, and exact byte sizes.

## Randomized pure-C properties

The property ledger is separate from the VEP differential. It records successful
runs of each randomized oracle, including the exact seed and duplicate count. A
failed suite does not append rows.

```{r property-history}
property_runs <- aggregate(
  cbind(trials, passed, failed, skipped, duplicates) ~
    run_date + source_revision + configured_trials + seed + suite_tests +
      suite_assertions + suite_elapsed_seconds + compiler,
  data = property_history,
  FUN = sum
)
target_counts <- aggregate(
  target ~ run_date + source_revision + configured_trials + seed,
  data = property_history,
  FUN = length
)
names(target_counts)[names(target_counts) == "target"] <- "randomized_targets"
property_runs <- merge(
  property_runs,
  target_counts,
  by = c("run_date", "source_revision", "configured_trials", "seed")
)
property_runs$source_revision <- short_revision(property_runs$source_revision)
property_runs <- property_runs[, c(
  "run_date", "source_revision", "seed", "randomized_targets", "trials",
  "passed", "failed", "duplicates", "suite_tests", "suite_assertions",
  "suite_elapsed_seconds", "compiler"
)]
for (column in c("trials", "passed", "suite_assertions")) {
  property_runs[[column]] <- format(
    property_runs[[column]],
    big.mark = ",",
    scientific = FALSE
  )
}
knitr::kable(property_runs, row.names = FALSE)
```

```{r property-targets}
last_property <- property_history[nrow(property_history), , drop = FALSE]
latest_properties <- property_history[
  property_history$source_revision == last_property$source_revision &
    property_history$seed == last_property$seed &
    property_history$configured_trials == last_property$configured_trials,
  c("target", "trials", "passed", "failed", "skipped", "duplicates"),
  drop = FALSE
]
for (column in c("trials", "passed", "failed", "skipped", "duplicates")) {
  latest_properties[[column]] <- format(
    latest_properties[[column]],
    big.mark = ",",
    scientific = FALSE
  )
}
knitr::kable(latest_properties, row.names = FALSE)
```

Passing the requested number of trials is necessary but does not prove that a
generator visited the states named by its contract. Randomized properties therefore
emit distribution counters, and the recorder stores each counter as a separate
numeric row. The table below is the latest complete run's state distribution; the
long-form CSV remains the machine-readable authority. Zero is evidence too: it
identifies a state that the declared seed did not exercise and must not be hidden by
the suite-level pass count.

```{r property-coverage}
latest_coverage <- property_coverage_history[
  property_coverage_history$source_revision == last_property$source_revision &
    property_coverage_history$seed == last_property$seed &
    property_coverage_history$configured_trials ==
      last_property$configured_trials,
  c("coverage", "metric", "count"),
  drop = FALSE
]
if (nrow(latest_coverage)) {
  latest_coverage$count <- format(
    latest_coverage$count,
    big.mark = ",",
    scientific = FALSE
  )
  latest_coverage$state_count <- paste0(
    latest_coverage$metric,
    "=",
    latest_coverage$count
  )
  coverage_table <- aggregate(
    state_count ~ coverage,
    data = latest_coverage,
    FUN = paste,
    collapse = "; "
  )
  names(coverage_table) <- c("randomized distribution", "observed states")
  knitr::kable(coverage_table, row.names = FALSE)
} else {
  cat("No distribution-counter ledger has been recorded for the latest run yet.\n")
}
```

## Individual Sequence Ontology terms

For each transcript pair, this compares the union of terms emitted by either
engine. A missing or extra term is therefore visible under its own SO name.
Rows must not be summed across terms because one pair can carry several terms.

```{r so-terms}
latest_run_ids <- unique(latest_runs[, c(
  "source_revision", "engine", "corpus", "model", "oracle_version", "oracle_build"
)])
term_rows <- merge(
  history[history$stratum_kind == "so_term", , drop = FALSE],
  latest_run_ids,
  by = c(
    "source_revision", "engine", "corpus", "model", "oracle_version", "oracle_build"
  )
)
term_counts <- aggregate(
  cbind(
    n, exact_agree, unresolved, resolved_n, resolved_agree,
    resolved_discordant, term_mismatch, engine_extra, engine_missing
  ) ~ corpus + consequence_class + impact,
  data = term_rows,
  FUN = sum
)
term_counts$corpus_label <- unname(corpus_labels[term_counts$corpus])
term_counts$corpus_label[is.na(term_counts$corpus_label)] <-
  term_counts$corpus[is.na(term_counts$corpus_label)]

term_coverage <- do.call(rbind, lapply(
  split(term_counts, term_counts$corpus_label),
  function(rows) data.frame(
    corpus = rows$corpus_label[[1L]],
    observed_terms = nrow(rows),
    terms_with_mismatch = sum(rows$term_mismatch > 0),
    term_mismatches = sum(rows$term_mismatch),
    engine_extra = sum(rows$engine_extra),
    engine_missing = sum(rows$engine_missing),
    terms_with_unresolved = sum(rows$unresolved > 0),
    unresolved_term_memberships = sum(rows$unresolved),
    stringsAsFactors = FALSE
  )
))
term_coverage <- term_coverage[match(
  unname(corpus_labels),
  term_coverage$corpus
), ]
term_coverage <- term_coverage[!is.na(term_coverage$corpus), ]
knitr::kable(term_coverage, row.names = FALSE)
```

`term_mismatches` counts an SO term that is missing or extra on a transcript
pair. `unresolved_term_memberships` is reported separately: an unresolved pair
can still carry the exact VEP term set, and a multi-term pair appears once under
each term.

```{r so-frontier, results='asis'}
term_frontier <- term_counts[
  term_counts$term_mismatch > 0 | term_counts$unresolved > 0,
  c(
    "corpus_label", "consequence_class", "impact", "n", "unresolved",
    "term_mismatch", "engine_extra", "engine_missing"
  ),
  drop = FALSE
]
names(term_frontier)[names(term_frontier) == "corpus_label"] <- "corpus"
term_frontier <- term_frontier[order(
  match(term_frontier$corpus, unname(corpus_labels)),
  -term_frontier$term_mismatch,
  -term_frontier$unresolved,
  term_frontier$consequence_class
), ]
if (nrow(term_frontier)) {
  knitr::kable(term_frontier, row.names = FALSE)
} else {
  cat("No SO-term mismatch or unresolved membership remains in the latest declared runs.\n")
}
```

Terms absent from this frontier table were exact and resolved everywhere they
were observed. The complete zero and nonzero strata remain in the CSV ledger.

## VEP impact classes

This table uses full consequence sets, so each transcript pair is counted once
within each corpus.

```{r impact}
set_rows <- merge(
  history[history$stratum_kind == "consequence_set", , drop = FALSE],
  latest_run_ids,
  by = c(
    "source_revision", "engine", "corpus", "model", "oracle_version", "oracle_build"
  )
)
impact_counts <- aggregate(
  cbind(n, exact_agree, unresolved, resolved_n, resolved_discordant) ~ corpus + impact,
  data = set_rows,
  FUN = sum
)
impact_counts$corpus <- unname(corpus_labels[impact_counts$corpus])
impact_counts$upper95 <- mapply(
  upper95,
  impact_counts$resolved_discordant,
  impact_counts$resolved_n
)
impact_counts$exact_rate <- percent(impact_counts$exact_agree / impact_counts$n)
impact_counts$resolved_error_upper_95 <- percent(impact_counts$upper95)
impact_counts$upper95 <- NULL
knitr::kable(impact_counts, row.names = FALSE)
```

The source artifact hash and exact Ensembl core/variation build remain in
`test/duckvep/conformance/data/conformance_history.csv` for audit and reruns.

## Variant-induced NMD

This is a separate executable differential against the pinned VEP Plugins
release/116 `NMD.pm`. It compares `triggering`, `escaping`, and `unresolved`
for every eligible transcript pair; it does not infer NMD from the core
`NMD_transcript_variant` biotype consequence.

```{r nmd-summary}
if (!is.null(latest_nmd_run_ids) && nrow(latest_nmd_run_ids)) {
  latest_nmd <- merge(
    nmd_history,
    latest_nmd_run_ids,
    by = nmd_run_keys
  )
  nmd_summary <- aggregate(
    cbind(
      n, exact_agree, term_mismatch, unresolved, resolved_n,
      resolved_discordant, engine_extra, engine_missing
    ) ~ source_revision + corpus + model,
    data = latest_nmd,
    FUN = sum
  )
  oracle_rows <- latest_nmd[
    latest_nmd$consequence_class == "unresolved",
    ,
    drop = FALSE
  ]
  if (nrow(oracle_rows)) {
    oracle_unresolved <- aggregate(
      n ~ source_revision + corpus + model,
      data = oracle_rows,
      FUN = sum
    )
    names(oracle_unresolved)[names(oracle_unresolved) == "n"] <-
      "oracle_unresolved"
    nmd_summary <- merge(
      nmd_summary,
      oracle_unresolved,
      by = c("source_revision", "corpus", "model"),
      all.x = TRUE
    )
    nmd_summary$oracle_unresolved[is.na(nmd_summary$oracle_unresolved)] <- 0
  } else {
    nmd_summary$oracle_unresolved <- 0
  }
  nmd_summary$resolved_error_upper_95 <- percent(mapply(
    upper95,
    nmd_summary$resolved_discordant,
    nmd_summary$resolved_n
  ))
  nmd_summary$revision <- short_revision(nmd_summary$source_revision)
  nmd_summary$exact <- paste0(nmd_summary$exact_agree, "/", nmd_summary$n)
  nmd_summary$not_comparable <-
    nmd_summary$engine_extra + nmd_summary$engine_missing
  knitr::kable(
    nmd_summary[, c(
      "revision", "corpus", "model", "exact", "term_mismatch",
      "not_comparable", "oracle_unresolved", "unresolved",
      "resolved_error_upper_95"
    )],
    col.names = c(
      "revision", "corpus", "model", "exact", "mismatches",
      "not comparable", "VEP unresolved", "DuckVEP unresolved",
      "resolved error upper 95%"
    ),
    row.names = FALSE
  )
} else {
  cat("No executable NMD-plugin differential is recorded.\n")
}
```

The ledger keeps the prediction confusion matrix rather than only the total:

```{r nmd-predictions}
if (exists("latest_nmd") && nrow(latest_nmd)) {
  nmd_predictions <- aggregate(
    n ~ source_revision + corpus + consequence_class + engine_status,
    data = latest_nmd,
    FUN = sum
  )
  names(nmd_predictions)[names(nmd_predictions) == "consequence_class"] <-
    "VEP_prediction"
  names(nmd_predictions)[names(nmd_predictions) == "engine_status"] <-
    "DuckVEP_prediction"
  nmd_predictions$revision <- short_revision(nmd_predictions$source_revision)
  knitr::kable(
    nmd_predictions[, c(
      "revision", "corpus", "VEP_prediction", "DuckVEP_prediction", "n"
    )],
    row.names = FALSE
  )
}
```

VEP projects the complete uploaded `VariationFeature` for the plugin's CDS and
exon-position rules. DuckVEP therefore retains both geometries: minimized edit
coordinates drive consequence and sequence changes, while the original feature
endpoints drive NMD. A padded and a minimal allele can encode the same sequence
edit but cross the plugin's inclusive positional threshold differently.
