---
title: "DuckVEP conformance against Ensembl VEP 116"
output: github_document
---

<!-- duckvep_conformance.md is generated from duckvep_conformance.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
history_path <- file.path(
  root, "test", "duckvep", "conformance", "data", "conformance_history.csv"
)
history <- read.csv(
  history_path,
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(oracle_version = "character")
)
property_history <- read.csv(
  file.path(
    root, "test", "duckvep", "conformance", "data", "property_history.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character", seed = "character")
)
overall <- history[history$stratum_kind == "all", , drop = FALSE]
latest <- overall[nrow(overall), , drop = FALSE]
same_run <- function(x) {
  x$source_revision == latest$source_revision &
    x$engine == latest$engine &
    x$corpus == latest$corpus &
    x$model == latest$model
}
upper95 <- function(k, n) {
  if (n == 0) NA_real_ else binom.test(k, n, conf.level = 0.95)$conf.int[[2L]]
}
percent <- function(x) {
  ifelse(is.na(x), "not estimable", sprintf("%.2f%%", 100 * x))
}
short_revision <- function(x) substr(x, 1, 8)
```

This report records exact transcript-level consequence agreement with the real
Ensembl VEP 116 `--gff` output. Unresolved DuckVEP rows remain in the denominator;
they are not discarded as unsupported cases. The CSV is append-only by source
revision, corpus, and resident model.

## Latest run

```{r latest}
latest_table <- data.frame(
  revision = short_revision(latest$source_revision),
  corpus = latest$corpus,
  model = latest$model,
  oracle = paste0("VEP ", latest$oracle_version),
  pairs = latest$n,
  exact = paste0(latest$exact_agree, "/", latest$n),
  unresolved = latest$unresolved,
  resolved_disagreements = latest$resolved_discordant,
  resolved_error_upper_95 = percent(latest$upper95),
  stringsAsFactors = FALSE
)
knitr::kable(latest_table, row.names = FALSE)
```

## History

```{r history}
timeline <- overall[, c(
  "run_date", "source_revision", "corpus", "model", "n", "exact_agree",
  "unresolved", "resolved_n", "resolved_discordant", "upper95"
)]
timeline$source_revision <- short_revision(timeline$source_revision)
timeline$exact_rate <- percent(timeline$exact_agree / timeline$n)
timeline$resolved_error_upper_95 <- percent(timeline$upper95)
timeline$upper95 <- NULL
knitr::kable(timeline, row.names = FALSE)
```

## Randomized pure-C properties

The property ledger is separate from the VEP differential. It records successful
runs of each randomized oracle, including the exact seed and duplicate count. A
failed suite does not append rows.

```{r property-history}
property_runs <- aggregate(
  cbind(trials, passed, failed, skipped, duplicates) ~
    run_date + source_revision + configured_trials + seed + suite_tests +
      suite_assertions + suite_elapsed_seconds + compiler,
  data = property_history,
  FUN = sum
)
target_counts <- aggregate(
  target ~ run_date + source_revision + configured_trials + seed,
  data = property_history,
  FUN = length
)
names(target_counts)[names(target_counts) == "target"] <- "randomized_targets"
property_runs <- merge(
  property_runs,
  target_counts,
  by = c("run_date", "source_revision", "configured_trials", "seed")
)
property_runs$source_revision <- short_revision(property_runs$source_revision)
property_runs <- property_runs[, c(
  "run_date", "source_revision", "seed", "randomized_targets", "trials",
  "passed", "failed", "duplicates", "suite_tests", "suite_assertions",
  "suite_elapsed_seconds", "compiler"
)]
for (column in c("trials", "passed", "suite_assertions")) {
  property_runs[[column]] <- format(
    property_runs[[column]],
    big.mark = ",",
    scientific = FALSE
  )
}
knitr::kable(property_runs, row.names = FALSE)
```

```{r property-targets}
last_property <- property_history[nrow(property_history), , drop = FALSE]
latest_properties <- property_history[
  property_history$source_revision == last_property$source_revision &
    property_history$seed == last_property$seed &
    property_history$configured_trials == last_property$configured_trials,
  c("target", "trials", "passed", "failed", "skipped", "duplicates"),
  drop = FALSE
]
for (column in c("trials", "passed", "failed", "skipped", "duplicates")) {
  latest_properties[[column]] <- format(
    latest_properties[[column]],
    big.mark = ",",
    scientific = FALSE
  )
}
knitr::kable(latest_properties, row.names = FALSE)
```

## Individual Sequence Ontology terms

For each transcript pair, this compares the union of terms emitted by either
engine. A missing or extra term is therefore visible under its own SO name.
Rows must not be summed across terms because one pair can carry several terms.

```{r so-terms}
term_rows <- history[
  history$stratum_kind == "so_term" & same_run(history),
  ,
  drop = FALSE
]
term_counts <- aggregate(
  cbind(
    n, exact_agree, unresolved, resolved_n, resolved_agree,
    resolved_discordant, term_mismatch, engine_extra, engine_missing
  ) ~ consequence_class + impact,
  data = term_rows,
  FUN = sum
)
term_counts$upper95 <- mapply(
  upper95,
  term_counts$resolved_discordant,
  term_counts$resolved_n
)
term_counts <- term_counts[order(
  -term_counts$resolved_discordant,
  -term_counts$unresolved,
  -term_counts$n,
  term_counts$consequence_class
), ]
term_counts$resolved_error_upper_95 <- percent(term_counts$upper95)
term_counts$upper95 <- NULL
knitr::kable(term_counts, row.names = FALSE)
```

## VEP impact classes

This table uses full consequence sets, so each transcript pair is counted once.

```{r impact}
set_rows <- history[
  history$stratum_kind == "consequence_set" & same_run(history),
  ,
  drop = FALSE
]
impact_counts <- aggregate(
  cbind(n, exact_agree, unresolved, resolved_n, resolved_discordant) ~ impact,
  data = set_rows,
  FUN = sum
)
impact_counts$upper95 <- mapply(
  upper95,
  impact_counts$resolved_discordant,
  impact_counts$resolved_n
)
impact_counts$exact_rate <- percent(impact_counts$exact_agree / impact_counts$n)
impact_counts$resolved_error_upper_95 <- percent(impact_counts$upper95)
impact_counts$upper95 <- NULL
knitr::kable(impact_counts, row.names = FALSE)
```

The source artifact hash and exact Ensembl core/variation build remain in
`test/duckvep/conformance/data/conformance_history.csv` for audit and reruns.
