---
title: "Phased replay: native stream and public SQL"
output: github_document
---

<!-- duckvep_haplotypes.md is generated from duckvep_haplotypes.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
history <- read.csv(file.path(root, "benchmarks/data/duckvep_haplotypes.csv"),
  colClasses = c(source_revision = "character", xor_hash = "character", sum_hash = "character",
    replay_xor_hash = "character", replay_sum_hash = "character",
    local_so_xor_hash = "character", local_so_sum_hash = "character"))
stopifnot(all(history$extension_build_binding == "htslib_distclean_make_release"),
  !anyDuplicated(history[c("source_revision", "transcripts", "samples", "overlap", "mode", "pass")]),
  all(history$mode %in% c("native", "sql", "sql_records", "sql_singletons", "sql_singletons_hgvs")))
singleton_history <- history[history$mode %in% c("sql_singletons", "sql_singletons_hgvs"), ]
history <- history[history$mode %in% c("native", "sql", "sql_records"), ]
if (!"output_contract" %in% names(history)) history$output_contract <- NA_character_
history$output_contract[is.na(history$output_contract)] <- "literal_replay"
for (field in c("json_bytes", "xor_hash", "sum_hash")) {
  projection <- paste0("replay_", field)
  if (!projection %in% names(history)) history[[projection]] <- history[[field]]
  missing <- is.na(history[[projection]])
  history[[projection]][missing] <- history[[field]][missing]
  local_so <- paste0("local_so_", field)
  if (!local_so %in% names(history)) history[[local_so]] <- NA
  old_contract <- history$output_contract == "local_coding_block_so"
  history[[local_so]][old_contract] <- history[[field]][old_contract]
}
revisions <- unique(history$source_revision)
stopifnot(length(revisions) >= 2L)
revision <- tail(revisions, 1L)
baseline_revision <- tail(revisions, 2L)[1L]
keys <- c("transcripts", "samples", "overlap", "mode")
groups <- split(history, interaction(history[c("source_revision", keys)], drop = TRUE))
summary <- do.call(rbind, lapply(groups, function(x) {
  stopifnot(nrow(x) == 3L, all(sort(x$pass) == 1:3))
  data.frame(x[1L, c("source_revision", keys)], min_s = min(x$seconds), median_s = median(x$seconds),
    max_s = max(x$seconds), max_process_rss_mib = max(x$peak_process_rss_kib) / 1024)
}))
comparison <- merge(
  subset(summary, source_revision == baseline_revision & mode == "sql"),
  subset(summary, source_revision == revision & mode == "sql"),
  by = keys, suffixes = c("_before", "_after"))
comparison$median_change_percent <- 100 * (comparison$median_s_after / comparison$median_s_before - 1)
comparison <- comparison[do.call(order, comparison[keys]), ]
current <- history[history$source_revision == revision, ]
summary <- summary[summary$source_revision == revision, setdiff(names(summary), "source_revision")]
summary <- summary[do.call(order, summary[keys]), ]
core <- current[current$mode == "native" & current$pass == 1L, ]
core <- core[order(core$transcripts, core$samples, core$overlap), ]
public <- current[current$mode == "sql" & current$pass == 1L, ]
public <- public[order(public$transcripts, public$samples, public$overlap), ]
raw <- current[current$mode == "sql_records" & current$pass == 1L, ]
stopifnot(identical(as.list(core[c("transcripts", "samples", "overlap")]),
  as.list(public[c("transcripts", "samples", "overlap")])))
fingerprint_keys <- c("transcripts", "samples", "overlap")
fingerprints <- unique(history[history$mode == "sql", c(fingerprint_keys, "output_contract",
  "output_leaves", "output_carriers", "cds_bytes", "protein_bytes", "physical_edits",
  "contributors", "blocks", "json_bytes", "xor_hash", "sum_hash")])
replay_fingerprints <- unique(history[history$mode == "sql", c(fingerprint_keys,
  "replay_json_bytes", "replay_xor_hash", "replay_sum_hash")])
local_so_fingerprints <- unique(history[history$mode == "sql" & history$output_contract %in%
  c("local_coding_block_so", "local_coding_block_so_hgvs_status"), c(fingerprint_keys,
    "local_so_json_bytes", "local_so_xor_hash", "local_so_sum_hash")])
stopifnot(nrow(comparison) == nrow(public),
  !anyDuplicated(fingerprints[c(fingerprint_keys, "output_contract")]),
  !anyNA(local_so_fingerprints), !anyDuplicated(local_so_fingerprints[fingerprint_keys]),
  !anyDuplicated(replay_fingerprints[fingerprint_keys]))
compared_contracts <- unique(history$output_contract[history$mode == "sql" &
  history$source_revision %in% c(revision, baseline_revision)])
same_contract <- length(compared_contracts) == 1L
raw_fingerprints <- unique(history[history$mode == "sql_records", c(fingerprint_keys,
  "output_contract", "input_physical_records", "input_record_sample_calls",
  "input_candidate_sample_rows", "output_leaves", "output_carriers", "cds_bytes",
  "protein_bytes", "physical_edits", "contributors", "blocks", "json_bytes", "xor_hash", "sum_hash")])
stopifnot(!anyNA(raw_fingerprints),
  !anyDuplicated(raw_fingerprints[c(fingerprint_keys, "output_contract")]))
```

Current source: `r revision`; same-input baseline: `r baseline_revision`.
Native measurements cover literal phased replay. SQL materializes all current
fields, including local coding-block SO. HGVS generation is disabled;
whole-haplotype SO/HGVS is unfinished and its computation is not timed here.
Native and decoded SQL consume standalone ALT events and decoded calls.
The additional source-record SQL lane measures literal GT parsing and replay;
standalone native raw-record execution and conditional source omissions are not timed.
All runs use one thread pinned to CPU 2 on an Intel Core i5-13500, DuckDB 1.5.3,
and a source-bound clean extension build. The native bridge uses the same kernel
sources compiled with `-O3 -DNDEBUG`; receipts retain compiler/version, binary and
fixture hashes, worker jobs, GNU-time logs and complete result fingerprints.

## Workload and verification

The [registry](../r/duckhtsbench/inst/benchmark_registry.tsv) stages two committed,
checksum-verified fixtures without network access: a 180-base CDS and four ALT
events. Two substitutions share a codon; an insertion/deletion pair displaces
and restores the frame. Four diploid sample templates repeat across the cohort,
producing exactly three occupied paths per transcript. Strict phase uses PS=10.
Transcript groups share genomic intervals, with non-overlapping groups 1,000
bases apart; `overlap` is the number of simultaneously active transcripts.

This deliberately favourable prefix-sharing workload is not a real population,
a long-transcript benchmark, a reader benchmark, or evidence for arbitrary phase
and structural configurations. Native input is a factorized, already-ordered
feed; SQL input is its complete event/candidate/sample relation stored in hash
order. Both retain every source identity and called carrier.

Before timing, native validation checks every transcript, complete CDS/protein
against independently rebuilt/Biostrings-translated sequences, contributors and
sample/lane/PS/ploidy membership. SQL validation compares the full carrier
multiset with decoded input, checks sequences and edit provenance, and rejects an
injected duplicate. Native validation also rejects a corrupted expected sequence.
Every measured decoded pass must retain all common native/SQL denominators and SQL's full,
canonicalized nested-row fingerprint; failed runs are not promoted. Full fingerprints
must also agree across revisions sharing an output contract. Local coding-block
and literal-replay projections provide separate comparisons with their recorded
output contracts; full current output is always checked. The HGVS-status schema
adds `hgvsp` (NULL) and `hgvsp_status` (`not_requested`) in this workload. Only the
separate prior-schema projections exclude these columns.

```{r denominators}
knitr::kable(core[c("transcripts", "samples", "overlap", "input_records",
  "projected_events", "input_calls", "output_leaves", "output_carriers",
  "prefixes_created", "translated_bases")], row.names = FALSE)
```

`input_records` counts distinct source ALT events, not candidate/sample rows.
`projected_events` counts event/transcript pairs. `output_carriers` counts
memberships of sample/lane keys in completed leaves; `output_leaves` counts
unique occupied prefixes, not unique peptide strings. Prefix creation and active
pool peaks below are observed in the native stream, not inferred SQL counters.

## Timing and process memory

```{r timing}
knitr::kable(summary, row.names = FALSE, digits = 3)
```

Both compared revisions return each block's local SO mask, coding status and
position relative to the first stop. A shared coding context evaluates physical
blocks on completed replay. `r if (same_contract) "Full-output fingerprints and input/output denominators match across the compared configurations." else "The current output adds HGVS columns. All prior local-coding-block fields and input/output denominators match, but complete rows have different schemas and byte counts. This is a same-input comparison, not identical-output work."`
The native count sink omits local SO evaluation. HGVS-enabled performance
requires a separate workload.

```{r sql-comparison}
knitr::kable(comparison[c("transcripts", "samples", "overlap", "median_s_before",
  "median_s_after", "median_change_percent", "max_process_rss_mib_before",
  "max_process_rss_mib_after")], row.names = FALSE, digits = 3)
```

Each recorded pass uses a fresh process and a full warm-up. Native timing starts
after workspace initialization and includes ordered-feed construction, GT routing,
projection, replay/translation, block construction and a count/length sink. It
does **not** include alignment, DuckDB, or nested output materialization.

SQL timing covers `CREATE OR REPLACE TABLE ... AS SELECT * FROM duckvep_haplotypes(...)`:
preparation, phase-domain discovery, mandatory sort, workspace initialization,
replay, both sequence alignments, local coding-block SO and **all** nested output
columns. Input/model
staging is outside timing. The warm-up output table is replaced by the measured
table. JSON sizing and fingerprint aggregation occur afterward, not in the timer;
no file-output throughput is claimed.

GNU `/usr/bin/time -v` measures whole-process maximum RSS, including the R host,
model/input tables, warm-up, measured materialization and post-run aggregates.
The heavier SQL correctness audit runs separately and is excluded from recorded
RSS. RSS is not an isolated operator allocation measurement. These two execution
scopes are intentionally different: their times are not a speedup comparison.
Post-run fingerprint aggregates depend on the output schema, so a schema change
also prevents interpreting process RSS as an identical-work memory comparison.

```{r output-size}
knitr::kable(public[c("transcripts", "samples", "overlap", "output_leaves",
  "cds_bytes", "protein_bytes", "json_bytes")], row.names = FALSE)
```

CDS/protein bytes count the sequence views returned once per occupied leaf.
JSON bytes are the measured UTF-8 size of complete canonicalized SQL rows,
including carriers, contributors, blocks and differences; they are not bytes
written to disk.

## Sparse native state

```{r state}
knitr::kable(core[c("transcripts", "samples", "overlap", "peak_transcripts",
  "peak_carriers", "peak_prefixes", "peak_events", "peak_projections",
  "peak_allele_bytes", "workspace_bytes", "model_bytes")], row.names = FALSE)
```

Native workspace bytes count actual preallocated buffer capacities and state,
separately from immutable model allocations. The fixed fixture is borrowed;
validation additionally owns `transcripts + 2 * samples` bytes outside the timed
kernel workspace. Peak pool occupancy is measured by the production stream.
Increasing total transcripts at fixed cohort size and overlap does not increase
that workspace; immutable model storage still grows with transcript count.

Both paths use the same active-pool limits. SQL additionally allows 65,536
alignment cells and 64 differing runs per axis, with a 64 MiB native workspace
ceiling and a 4 GB DuckDB memory limit. SQL input, sorting and output memory belong
to DuckDB and may grow with the complete relation. The process RSS table must not
be presented as a constant-total-memory guarantee.

This matched comparison retains three repeated passes per workload and revision.
Sanitizer and conformance jobs did not overlap either benchmark campaign. Both
ran at different times on a shared machine; CPU pinning does not isolate the
cause of timing differences between campaigns.
One machine and this deliberately shared synthetic cohort do not establish
production throughput, statistical significance or a general absence of regression.
Whole-haplotype SO/HGVS and typed structural composition require their own measurements.

## Reproduction

Run [the R driver](duckvep_haplotypes.R) with a clean source-bound extension receipt
from the shared evidence builder; it stages inputs through `duckhtsbench`.
All six configurations above use `--passes 3 --cpu 2`. To prepare the receipt
from a clean checkout (no network access in the build):

```{r clean-build, eval=FALSE, echo=TRUE}
source("scripts/duckvep_evidence.R")
root <- normalizePath(".")
revision <- duckvep_evidence_revision(root)
extension <- duckvep_evidence_build_extension(root,
  "build/release/duckhts.duckdb_extension", revision)
duckvep_evidence_write_extension_receipt("/tmp/haplotype-extension.tsv", revision, extension)
```

Pass `--extension-receipt /tmp/haplotype-extension.tsv` to the R driver. For a
network-free smoke test without a clean-build certificate:

```{bash smoke, eval=FALSE, echo=TRUE}
Rscript benchmarks/duckvep_haplotypes.R \
  --transcripts 16 --samples 4 --overlap 4 --passes 1 --diagnostic
```

Diagnostic results are retained but are not accepted into the recorded baseline.

### Raw source-record SQL lane

`--modes sql_records` measures the public `input_mode := 'source_records'` path
under `vep116_compat`; the default `--modes native,sql` selects the decoded lanes
reported above. All modes use the same registered reference and four biallelic
events. Source rows carry the ordered ALT list and literal diploid GT text.
The timed query includes source validation, replay-order planning, raw-GT parsing,
allele expansion, sorting, replay and full nested output. File decoding and
model/input staging are outside timing. This is not a standalone raw native
kernel benchmark or a rare-configuration conformance campaign.

The source lane returns four paths per transcript and two carriers per
sample/transcript, including the occupied reference path. Decoded lanes return
three paths per transcript and 7/4 carriers per sample/transcript.
Complete CDS/protein strings, carrier
multisets, contributor IDs, ALT ordinals and block/edit counts are checked
against the fixture. Eleven corruption controls cover changed CDS, protein,
missing CDS, carrier identity, missing ploidy, contributor identity, missing
ALT ordinal, invented HGVS, changed HGVS status, invalid source GT and duplicate output. They run in the separate
correctness worker.

`input_physical_records`, `input_record_sample_calls` and
`input_candidate_sample_rows` are separately counted from the staged source
relation. REF, ALT and undefined-slot descriptors are execution work, not extra
physical records. For this biallelic workload the raw lane reserves 13 active
event slots, `12 * overlap + 1` projection slots and 192 allele bytes; other
limits match the decoded SQL lane. Whole-process RSS includes both the decoded
fixture table and the derived source table. Output fingerprints include the
raw lane's complete schema and reference paths; they are not substituted for
decoded-output fingerprints.

```{bash source-record-smoke, eval=FALSE, echo=TRUE}
Rscript benchmarks/duckvep_haplotypes.R \
  --transcripts 16 --samples 4 --overlap 4 --passes 1 \
  --modes native,sql,sql_records --diagnostic
```

The source-bound raw lane uses three fresh-process passes at the current source,
with one thread pinned to CPU 2. Its timing and process RSS appear in the timing
table above; its input and output denominators are:

```{r raw-denominators}
knitr::kable(raw[c("transcripts", "samples", "overlap", "input_physical_records",
  "input_record_sample_calls", "input_candidate_sample_rows", "output_leaves", "output_carriers")],
  row.names = FALSE)
knitr::kable(raw[c("output_leaves", "cds_bytes", "protein_bytes", "json_bytes")], row.names = FALSE)
```

There is no earlier source-bound raw-lane measurement for a same-contract
comparison. Different output denominators preclude treating raw versus decoded
time as an implementation speedup.

## Singleton HGVS materialization

`--modes sql_singletons,sql_singletons_hgvs` compares the same decoded input with
protein HGVS disabled and enabled. Each sample carries one of the four registered
ALT events heterozygously. Four paths per transcript retain all non-reference
carriers; reference paths remain implicit. The two substitutions and two
frame-changing indels exercise the independent-event HGVS consumer, not compound
HGVS. This is a different carrier distribution from the compound replay workload
above; their timings are not interchangeable.

The driver repeats the registered reference CDS at the model's genomic coordinates,
pads the inter-transcript sequence with A, and indexes that per-run FASTA. Both
modes load exactly the same model and reference. A transcript has no UTR: explicitly
empty pre-/post-CDS sequences distinguish a complete zero-length flank from an
unavailable flank. Reference/model/input preparation is outside timing. The timed
query includes sorting, bounded query initialization, sequence replay, differences,
local coding facts and every output column. HGVS-enabled execution also opens its
query-local faidx handle, retrieves genomic context and computes protein HGVS.

Each mode has a separate correctness process, followed by three fresh-process
measurement passes with a full warm-up. Direct sequence replacement and Biostrings
translation check every expected CDS/protein. Multiset comparisons check complete
carrier identities, phase sets, ploidies and source-event membership. The enabled
path also compares every event/transcript HGVSp with independent annotation, allowing
only its documented prediction-parenthesis difference. That internal consistency
check is not an independent VEP oracle or compound-HGVS certificate. The four fixed
event shapes do not measure broad HGVS conformance.

Nine shared output corruptions and duplicate-output insertion must fail the
correctness process. HGVS-enabled checks additionally reject missing HGVS and
malformed internal parentheses. Whole-output byte counts and fingerprints must match every
pass of the same mode. All non-HGVS fields must match between modes. HGVS-on/off
complete outputs intentionally differ and are never reported as identical work.
The worker receipts retain the derived FASTA/index, actual versus independent HGVS,
source/binary/input hashes, jobs and process logs. Process RSS includes setup,
warm-up and post-query aggregates, not just native workspace.

```{r singleton-hgvs}
if (!nrow(singleton_history)) {
  cat("No source-bound singleton HGVS measurement is recorded.\n")
} else {
  singleton_revision <- tail(unique(singleton_history$source_revision), 1L)
  measured <- singleton_history[singleton_history$source_revision == singleton_revision, ]
  stopifnot(all(measured$threads == 1L), all(measured$cpu_affinity == 2L),
    all(measured$duckdb_version == "1.5.3"),
    all(measured$output_leaves == measured$transcripts * 4),
    all(measured$output_carriers == measured$transcripts * measured$samples),
    all(measured$input_candidate_sample_rows == measured$transcripts * measured$samples * 4),
    all(measured$input_physical_records == measured$transcripts / measured$overlap * 4),
    all(measured$input_record_sample_calls ==
      measured$transcripts / measured$overlap * measured$samples * 4))
  configuration <- c("transcripts", "samples", "overlap")
  mode_keys <- c(configuration, "mode")
  non_hgvs <- unique(measured[c(configuration, "output_leaves", "output_carriers",
    "cds_bytes", "protein_bytes", "physical_edits", "contributors", "blocks",
    "local_so_json_bytes", "local_so_xor_hash", "local_so_sum_hash")])
  stopifnot(!anyNA(non_hgvs), !anyDuplicated(non_hgvs[configuration]))
  full <- unique(measured[c(mode_keys, "json_bytes", "xor_hash", "sum_hash")])
  stopifnot(!anyNA(full), !anyDuplicated(full[mode_keys]))
  timings <- do.call(rbind, lapply(split(measured, interaction(measured[mode_keys], drop = TRUE)),
    function(x) {
      stopifnot(nrow(x) == 3L, identical(sort(x$pass), 1:3))
      data.frame(x[1L, mode_keys], median_s = median(x$seconds),
        min_s = min(x$seconds), max_s = max(x$seconds),
        peak_process_rss_mib = max(x$peak_process_rss_kib) / 1024)
    }))
  paired <- merge(timings[timings$mode == "sql_singletons", ],
    timings[timings$mode == "sql_singletons_hgvs", ], by = configuration,
    suffixes = c("_disabled", "_enabled"))
  stopifnot(nrow(paired) * 2L == nrow(timings))
  cat("Measured source: `", singleton_revision,
    "`. One thread, CPU 2, Intel i5-13500, DuckDB 1.5.3.\n\n", sep = "")
  knitr::kable(timings, row.names = FALSE)
  knitr::kable(measured[measured$pass == 1L, c(mode_keys, "input_physical_records",
    "input_record_sample_calls", "input_candidate_sample_rows", "output_leaves",
    "output_carriers", "cds_bytes", "protein_bytes", "json_bytes")], row.names = FALSE)
}
```

This is the first source-bound measurement of this HGVS-enabled workload, not a
before/after implementation comparison. Compound HGVS, raw-input omissions,
multi-exon models and heterogeneous genomic contexts still need their own matched
measurements. The compound replay and raw-record baselines above retain their
original source revisions and denominators.

```{bash singleton-hgvs-smoke, eval=FALSE, echo=TRUE}
Rscript benchmarks/duckvep_haplotypes.R \
  --transcripts 16 --samples 4 --overlap 4 --passes 1 \
  --modes sql_singletons,sql_singletons_hgvs --diagnostic
```
