---
title: "DuckVEP conformance against Ensembl VEP 116"
output: github_document
---

<!-- duckvep_conformance.md is generated from duckvep_conformance.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
history_path <- file.path(
  root, "test", "duckvep", "conformance", "data", "conformance_history.csv"
)
history <- read.csv(
  history_path,
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(oracle_version = "character")
)
property_history <- read.csv(
  file.path(
    root, "test", "duckvep", "conformance", "data", "property_history.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character", seed = "character")
)
release_parquet <- read.csv(
  file.path(
    root, "benchmarks", "data", "duckvep_release_parquet.csv"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character", chromosome = "character")
)
model_receipts <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_model_receipts.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character", source_version = "character")
)
overall <- history[history$stratum_kind == "all", , drop = FALSE]
tested_revisions <- unique(overall$source_revision)
revision_ancestry <- suppressWarnings(system2(
  "git",
  c("-C", shQuote(root), "rev-list", "--topo-order", "HEAD"),
  stdout = TRUE,
  stderr = FALSE
))
tested_ancestry <- revision_ancestry[revision_ancestry %in% tested_revisions]
latest_revision <- if (length(tested_ancestry)) {
  tested_ancestry[[1L]]
} else {
  overall$source_revision[[nrow(overall)]]
}
latest_runs <- overall[
  overall$source_revision == latest_revision,
  ,
  drop = FALSE
]
latest <- latest_runs[which.max(latest_runs$n), , drop = FALSE]
state_overall <- overall[
  grepl("^state_exploration_seed_[0-9]+$", overall$corpus),
  ,
  drop = FALSE
]
state_revisions <- unique(state_overall$source_revision)
state_ancestry <- revision_ancestry[revision_ancestry %in% state_revisions]
latest_state_revision <- if (length(state_ancestry)) {
  state_ancestry[[1L]]
} else if (nrow(state_overall)) {
  state_overall$source_revision[[nrow(state_overall)]]
} else {
  ""
}
latest_state_runs <- state_overall[
  state_overall$source_revision == latest_state_revision,
  ,
  drop = FALSE
]
same_run <- function(x) {
  x$source_revision == latest$source_revision &
    x$engine == latest$engine &
    x$corpus == latest$corpus &
    x$model == latest$model
}
upper95 <- function(k, n) {
  if (n == 0) NA_real_ else binom.test(k, n, conf.level = 0.95)$conf.int[[2L]]
}
percent <- function(x) {
  ifelse(is.na(x), "not estimable", sprintf("%.2f%%", 100 * x))
}
short_revision <- function(x) substr(x, 1, 8)
oracle_field <- function(x, field) {
  pattern <- paste0("(^|;)", field, "=([^;]+)")
  found <- regexec(pattern, x)
  values <- regmatches(x, found)
  vapply(values, function(value) {
    if (length(value) >= 3L) value[[3L]] else ""
  }, character(1L))
}
```

This report records exact transcript-level consequence agreement with the real
Ensembl VEP 116 executable, using either its declared indexed cache or a staged
GFF oracle. Unresolved DuckVEP rows remain in the denominator; they are not
discarded as unsupported cases. The CSV is append-only by source revision,
corpus, and resident model. Independent frozen distributions and seeds are kept
separate so a fix cannot improve its own hand-picked witnesses and hide a
regression elsewhere.

## Latest tested revision

```{r latest}
latest_table <- data.frame(
  revision = short_revision(latest_runs$source_revision),
  corpus = latest_runs$corpus,
  model = latest_runs$model,
  assembly = oracle_field(latest_runs$oracle_build, "assembly"),
  species = oracle_field(latest_runs$oracle_build, "species"),
  oracle_source = oracle_field(latest_runs$oracle_build, "oracle"),
  oracle = paste0("VEP ", latest_runs$oracle_version),
  pairs = latest_runs$n,
  exact = paste0(latest_runs$exact_agree, "/", latest_runs$n),
  unresolved = latest_runs$unresolved,
  resolved_disagreements = latest_runs$resolved_discordant,
  resolved_error_upper_95 = percent(latest_runs$upper95),
  stringsAsFactors = FALSE
)
knitr::kable(latest_table, row.names = FALSE)
```

The SO and impact tables below keep all corpora at this tested revision separate;
they do not let the largest corpus hide a smaller species- or assembly-specific
frontier.

## Prepared model receipts

```{r model-receipts}
receipt_table <- data.frame(
  revision = short_revision(model_receipts$source_revision),
  species = model_receipts$species,
  release = model_receipts$source_version,
  assembly = model_receipts$assembly,
  regions = format(model_receipts$region_count, big.mark = ","),
  transcripts = format(model_receipts$transcript_count, big.mark = ","),
  coding_backed = format(
    model_receipts$sequence_backed_transcript_count,
    big.mark = ","
  ),
  exons = format(model_receipts$exon_membership_count, big.mark = ","),
  mature_miRNA_segments = format(
    model_receipts$mature_mirna_segment_count,
    big.mark = ","
  ),
  peptide_edits = format(model_receipts$peptide_edit_count, big.mark = ","),
  codon_tables = model_receipts$codon_table_counts,
  model_sha256 = substr(model_receipts$model_sha256, 1, 12),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
knitr::kable(receipt_table, row.names = FALSE)
```

These are complete model-build receipts, not counts inferred from a differential.
The ledger retains the full source-manifest, reference, and model SHA-256 values,
the exact VEP transcript filter, every count above, CDS/flank base totals, and the
external artifact name. The Plasmodium row is an Ensembl Genomes release-63 cache
paired with the VEP/core-116 executable libraries, which is why both release
numbers are recorded.

## History

```{r history}
timeline <- overall[, c(
  "run_date", "source_revision", "corpus", "model", "n", "exact_agree",
  "unresolved", "resolved_n", "resolved_discordant", "upper95"
)]
timeline$source_revision <- short_revision(timeline$source_revision)
timeline$exact_rate <- percent(timeline$exact_agree / timeline$n)
timeline$resolved_error_upper_95 <- percent(timeline$upper95)
timeline$upper95 <- NULL
knitr::kable(timeline, row.names = FALSE)
```

## Randomized executable-VEP state exploration

This is the anti-overfitting lane against the VEP executable, not an internal
property test. Each seed contains the fixed predicate witnesses plus 100,000
unique alleles. Three quarters of random positions are within six bases of
splice, exon, translation-start, and translation-stop boundaries; one quarter
is uniform across the transcript. SNVs, MNVs, insertions, deletions, and delins
are sampled with equal probability, with differing alleles up to 49 bases.

```{r state-exploration}
if (nrow(latest_state_runs)) {
  state_table <- data.frame(
    revision = short_revision(latest_state_runs$source_revision),
    seed = sub("^state_exploration_seed_", "", latest_state_runs$corpus),
    pairs = latest_state_runs$n,
    exact = latest_state_runs$exact_agree,
    unresolved = latest_state_runs$unresolved,
    resolved_disagreements = latest_state_runs$resolved_discordant,
    exact_error_upper_95_ppm = mapply(
      upper95,
      latest_state_runs$n - latest_state_runs$exact_agree,
      latest_state_runs$n
    ) * 1e6,
    stringsAsFactors = FALSE
  )
  state_total_n <- sum(latest_state_runs$n)
  state_total_exact <- sum(latest_state_runs$exact_agree)
  state_table <- rbind(
    state_table,
    data.frame(
      revision = short_revision(latest_state_revision),
      seed = "combined",
      pairs = state_total_n,
      exact = state_total_exact,
      unresolved = sum(latest_state_runs$unresolved),
      resolved_disagreements = sum(latest_state_runs$resolved_discordant),
      exact_error_upper_95_ppm = upper95(
        state_total_n - state_total_exact,
        state_total_n
      ) * 1e6,
      stringsAsFactors = FALSE
    )
  )
  for (column in c("pairs", "exact", "unresolved", "resolved_disagreements")) {
    state_table[[column]] <- format(
      state_table[[column]],
      big.mark = ",",
      scientific = FALSE
    )
  }
  state_table$exact_error_upper_95_ppm <- sprintf(
    "%.1f",
    state_table$exact_error_upper_95_ppm
  )
  knitr::kable(state_table, row.names = FALSE)
} else {
  cat("No randomized executable-VEP state exploration is recorded.\n")
}
```

The same campaign covered the following SO terms. Counts are term memberships,
not distinct transcript pairs, because one pair may carry several terms and the
four seeds deliberately retain the same fixed witnesses.

```{r state-exploration-so}
if (nrow(latest_state_runs)) {
  state_run_ids <- unique(latest_state_runs[, c(
    "source_revision", "engine", "corpus", "model", "oracle_version", "oracle_build"
  )])
  state_term_rows <- merge(
    history[history$stratum_kind == "so_term", , drop = FALSE],
    state_run_ids,
    by = c(
      "source_revision", "engine", "corpus", "model", "oracle_version", "oracle_build"
    )
  )
  state_term_counts <- aggregate(
    cbind(n, unresolved, term_mismatch, engine_extra, engine_missing) ~
      consequence_class + impact,
    data = state_term_rows,
    FUN = sum
  )
  state_seed_counts <- aggregate(
    corpus ~ consequence_class + impact,
    data = state_term_rows,
    FUN = function(values) length(unique(values))
  )
  names(state_seed_counts)[names(state_seed_counts) == "corpus"] <- "seeds_observed"
  state_term_counts <- merge(
    state_term_counts,
    state_seed_counts,
    by = c("consequence_class", "impact")
  )
  state_term_counts <- state_term_counts[order(
    -state_term_counts$n,
    state_term_counts$consequence_class
  ), c(
    "consequence_class", "impact", "seeds_observed", "n", "unresolved",
    "term_mismatch", "engine_extra", "engine_missing"
  )]
  knitr::kable(state_term_counts, row.names = FALSE)
}
```

This distribution deliberately stresses local allele and boundary states on one
engineered transcript. It does not replace the indexed-cache corpora, which add
real transcript density, imported flags, exceptional peptide edits, codon tables,
assemblies, and species. The revision is shown explicitly because this expensive
campaign is not silently attributed to later code.

## Official Ensembl release corpus in Parquet

The official release consequence VCF is already BGZF-compressed. This table measures its
complete typed DuckHTS reader projection and the narrower `VE` plus CSQ projection used by
the bulk oracle lane. It is a storage comparison, not a claim that the Parquet projection
can reproduce the original VCF byte-for-byte.

```{r release-parquet}
last_storage <- release_parquet[nrow(release_parquet), , drop = FALSE]
storage <- release_parquet[
  release_parquet$source_revision == last_storage$source_revision &
    release_parquet$source_sha256 == last_storage$source_sha256,
  ,
  drop = FALSE
]
storage <- storage[match(c("full_typed", "consequence"), storage$projection), ]
storage_table <- data.frame(
  revision = short_revision(storage$source_revision),
  release = storage$ensembl_release,
  assembly = storage$assembly,
  chromosome = storage$chromosome,
  projection = storage$projection,
  columns = storage$columns,
  records = format(storage$records, big.mark = ",", scientific = FALSE),
  ALT_alleles = format(
    storage$alternate_alleles,
    big.mark = ",",
    scientific = FALSE
  ),
  CSQ_entries = format(storage$csq_entries, big.mark = ",", scientific = FALSE),
  source_MiB = sprintf("%.1f", storage$source_bytes / 2^20),
  parquet_MiB = sprintf("%.1f", storage$parquet_bytes / 2^20),
  parquet_of_source = sprintf(
    "%.1f%%",
    100 * storage$parquet_fraction_of_source
  ),
  elapsed_seconds = sprintf("%.1f", storage$elapsed_seconds),
  records_per_second = format(
    round(storage$records_per_second),
    big.mark = ",",
    scientific = FALSE
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
knitr::kable(storage_table, row.names = FALSE)
```

The ledger records the official source URL, SHA-256 of every input and output, DuckHTS and
DuckDB versions, compression, row-group size, thread count, machine, and exact byte sizes.

## Randomized pure-C properties

The property ledger is separate from the VEP differential. It records successful
runs of each randomized oracle, including the exact seed and duplicate count. A
failed suite does not append rows.

```{r property-history}
property_runs <- aggregate(
  cbind(trials, passed, failed, skipped, duplicates) ~
    run_date + source_revision + configured_trials + seed + suite_tests +
      suite_assertions + suite_elapsed_seconds + compiler,
  data = property_history,
  FUN = sum
)
target_counts <- aggregate(
  target ~ run_date + source_revision + configured_trials + seed,
  data = property_history,
  FUN = length
)
names(target_counts)[names(target_counts) == "target"] <- "randomized_targets"
property_runs <- merge(
  property_runs,
  target_counts,
  by = c("run_date", "source_revision", "configured_trials", "seed")
)
property_runs$source_revision <- short_revision(property_runs$source_revision)
property_runs <- property_runs[, c(
  "run_date", "source_revision", "seed", "randomized_targets", "trials",
  "passed", "failed", "duplicates", "suite_tests", "suite_assertions",
  "suite_elapsed_seconds", "compiler"
)]
for (column in c("trials", "passed", "suite_assertions")) {
  property_runs[[column]] <- format(
    property_runs[[column]],
    big.mark = ",",
    scientific = FALSE
  )
}
knitr::kable(property_runs, row.names = FALSE)
```

```{r property-targets}
last_property <- property_history[nrow(property_history), , drop = FALSE]
latest_properties <- property_history[
  property_history$source_revision == last_property$source_revision &
    property_history$seed == last_property$seed &
    property_history$configured_trials == last_property$configured_trials,
  c("target", "trials", "passed", "failed", "skipped", "duplicates"),
  drop = FALSE
]
for (column in c("trials", "passed", "failed", "skipped", "duplicates")) {
  latest_properties[[column]] <- format(
    latest_properties[[column]],
    big.mark = ",",
    scientific = FALSE
  )
}
knitr::kable(latest_properties, row.names = FALSE)
```

## Individual Sequence Ontology terms

For each transcript pair, this compares the union of terms emitted by either
engine. A missing or extra term is therefore visible under its own SO name.
Rows must not be summed across terms because one pair can carry several terms.

```{r so-terms}
latest_run_ids <- unique(latest_runs[, c(
  "source_revision", "engine", "corpus", "model", "oracle_version", "oracle_build"
)])
term_rows <- merge(
  history[history$stratum_kind == "so_term", , drop = FALSE],
  latest_run_ids,
  by = c(
    "source_revision", "engine", "corpus", "model", "oracle_version", "oracle_build"
  )
)
term_counts <- aggregate(
  cbind(
    n, exact_agree, unresolved, resolved_n, resolved_agree,
    resolved_discordant, term_mismatch, engine_extra, engine_missing
  ) ~ corpus + consequence_class + impact,
  data = term_rows,
  FUN = sum
)
corpus_labels <- c(
  final_dbsnp157_windows_seed29 = "GRCh38 dbSNP",
  final_giab_grch38_seed71 = "GRCh38 GIAB",
  final_clinvar_coding_seed113 = "GRCh38 ClinVar coding",
  final_clinvar_crosschrom_seed17 = "GRCh38 ClinVar cross-chromosome",
  final_grch37_cache_seed37 = "GRCh37",
  `plasmodium-falciparum-vep63-seed11663` = "P. falciparum"
)
term_counts$corpus_label <- unname(corpus_labels[term_counts$corpus])
term_counts$corpus_label[is.na(term_counts$corpus_label)] <-
  term_counts$corpus[is.na(term_counts$corpus_label)]

term_coverage <- do.call(rbind, lapply(
  split(term_counts, term_counts$corpus_label),
  function(rows) data.frame(
    corpus = rows$corpus_label[[1L]],
    observed_terms = nrow(rows),
    terms_with_mismatch = sum(rows$term_mismatch > 0),
    term_mismatches = sum(rows$term_mismatch),
    engine_extra = sum(rows$engine_extra),
    engine_missing = sum(rows$engine_missing),
    terms_with_unresolved = sum(rows$unresolved > 0),
    unresolved_term_memberships = sum(rows$unresolved),
    stringsAsFactors = FALSE
  )
))
term_coverage <- term_coverage[match(
  unname(corpus_labels),
  term_coverage$corpus
), ]
term_coverage <- term_coverage[!is.na(term_coverage$corpus), ]
knitr::kable(term_coverage, row.names = FALSE)
```

`term_mismatches` counts an SO term that is missing or extra on a transcript
pair. `unresolved_term_memberships` is reported separately: an unresolved pair
can still carry the exact VEP term set, and a multi-term pair appears once under
each term.

```{r so-frontier}
term_frontier <- term_counts[
  term_counts$term_mismatch > 0 | term_counts$unresolved > 0,
  c(
    "corpus_label", "consequence_class", "impact", "n", "unresolved",
    "term_mismatch", "engine_extra", "engine_missing"
  ),
  drop = FALSE
]
names(term_frontier)[names(term_frontier) == "corpus_label"] <- "corpus"
term_frontier <- term_frontier[order(
  match(term_frontier$corpus, unname(corpus_labels)),
  -term_frontier$term_mismatch,
  -term_frontier$unresolved,
  term_frontier$consequence_class
), ]
knitr::kable(term_frontier, row.names = FALSE)
```

Terms absent from this frontier table were exact and resolved everywhere they
were observed. The complete zero and nonzero strata remain in the CSV ledger.

## VEP impact classes

This table uses full consequence sets, so each transcript pair is counted once
within each corpus.

```{r impact}
set_rows <- merge(
  history[history$stratum_kind == "consequence_set", , drop = FALSE],
  latest_run_ids,
  by = c(
    "source_revision", "engine", "corpus", "model", "oracle_version", "oracle_build"
  )
)
impact_counts <- aggregate(
  cbind(n, exact_agree, unresolved, resolved_n, resolved_discordant) ~ corpus + impact,
  data = set_rows,
  FUN = sum
)
impact_counts$corpus <- unname(corpus_labels[impact_counts$corpus])
impact_counts$upper95 <- mapply(
  upper95,
  impact_counts$resolved_discordant,
  impact_counts$resolved_n
)
impact_counts$exact_rate <- percent(impact_counts$exact_agree / impact_counts$n)
impact_counts$resolved_error_upper_95 <- percent(impact_counts$upper95)
impact_counts$upper95 <- NULL
knitr::kable(impact_counts, row.names = FALSE)
```

The source artifact hash and exact Ensembl core/variation build remain in
`test/duckvep/conformance/data/conformance_history.csv` for audit and reruns.
