---
title: "DuckVEP sorted-stream throughput"
output: github_document
---

<!-- duckvep_throughput.md is generated from duckvep_throughput.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
history <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_throughput.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
history$revision <- substr(history$source_revision, 1, 8)
resources <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_model_resources.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
resources$revision <- substr(resources$source_revision, 1, 8)

nearest_tested_ancestor <- function(revisions) {
  revisions <- unique(revisions[nzchar(revisions)])
  distance <- vapply(
    revisions,
    function(revision) {
      is_ancestor <- system2(
        "git",
        c("-C", root, "merge-base", "--is-ancestor", revision, "HEAD"),
        stdout = FALSE,
        stderr = FALSE
      )
      if (is_ancestor != 0L) {
        return(Inf)
      }
      as.numeric(system2(
        "git",
        c("-C", root, "rev-list", "--count", paste0(revision, "..HEAD")),
        stdout = TRUE,
        stderr = FALSE
      ))
    },
    numeric(1)
  )
  if (!length(distance) || all(!is.finite(distance))) {
    return(NA_character_)
  }
  revisions[[which.min(distance)]]
}
```

This benchmark drives either the rich `duckvep_annotate` result or the numeric
`duckvep_annotate_compact` result through DuckDB's stable extension API with
nondecreasing sequence-region and position keys. It records candidate sweep,
consequence classification, list materialization, `unnest`, and final aggregation
together. Model loading and input staging are outside the timed pass.

The checked-in one-transcript workload is a repeatable adapter floor. The final
production-density workload uses the receipt-hashed Ensembl 116 GRCh38 model:
194 sequence regions, 644,427 transcripts, 5,068,416 exon memberships, and
369,631 sequence-backed coding transcripts. The regulation-enabled form adds
380,818 RegulatoryFeature and 1,002,762 MotifFeature intervals in an independent
resident SoA. Its ordinary topology workload is a deterministic 1-in-40 hash
sample of 100,957 sorted GIAB sites. The coding
workloads deliberately repeat a frozen set to stabilize one-core kernel timing:
200,000 SNVs repeat 1,000 distinct alleles; the 36,000 non-SNV workload repeats
1,000 MNVs and 8,000 length-changing alleles four times; the 200,000 mixed
workload is 10% SNV, 10% MNV, 40% lengthening, and 40% shortening. Repetition
favours warm model and sequence caches, so these rows measure the resident
execution lanes, not file ingestion or an end-to-end genome run.

The paired-breakend workload repeats the 1,004-event multichromosome seed-31
VEP differential to 200,000 sorted endpoint pairs. Both endpoints participate
in candidate discovery and each transcript receives the union of the two
endpoint consequence sets. Its much larger transcript fan-out makes output
rows/s as important as paired events/s.

Earlier chromosome-only and complete-primary staging rows remain in the ledger
as historical measurements. They have different transcript inventories and
checksums and must not be treated as revisions of the final receipt.
The production query restates `ORDER BY` in the timed input CTE, so its result
includes the cost of enforcing the kernel's nondecreasing-input contract even
though the staging table was already written in that order. The `79cfaaf2`
rows predate that review correction and relied on the single-thread physical
scan preserving insertion order; they remain only as historical measurements.

## Recorded runs

```{r runs}
table <- history[, c(
  "run_date", "revision", "workload", "output_mode", "threads", "variants",
  "transcripts", "exons", "regulation_features", "annotated_rows", "passes", "min_seconds",
  "median_seconds", "max_seconds", "variants_per_second", "ns_per_variant",
  "cpu"
)]
table$output_rows_per_second <- round(
  table$annotated_rows / table$median_seconds
)
table$variants <- format(table$variants, big.mark = ",", scientific = FALSE)
table$transcripts <- format(table$transcripts, big.mark = ",", scientific = FALSE)
table$exons <- format(table$exons, big.mark = ",", scientific = FALSE)
table$regulation_features <- format(
  table$regulation_features,
  big.mark = ",",
  scientific = FALSE
)
table$annotated_rows <- format(
  table$annotated_rows,
  big.mark = ",",
  scientific = FALSE
)
table$variants_per_second <- round(table$variants_per_second)
table$ns_per_variant <- round(table$ns_per_variant, 1)
table$output_rows_per_second <- format(
  table$output_rows_per_second,
  big.mark = ",",
  scientific = FALSE
)
knitr::kable(table, row.names = FALSE)
```

Each pass consumes every staged input and checks output cardinality plus either
the rendered consequence-byte total or the numeric consequence-mask sum. Rows
with different workload, output mode, thread count, host, or variant count are
separate measurements and should not be compared as if only the engine changed.

## Core regulation and motif sweep

```{r regulation-throughput}
regulation_runs <- history[
  history$workload == "ensembl116_grch38_final_giab_sites_hash40_regulation" &
    history$output_mode == "compact" & history$threads == 1L,
  ,
  drop = FALSE
]
if (nrow(regulation_runs)) {
  transcript_controls <- history[
    history$workload == "ensembl116_grch38_final_giab_sites_hash40" &
      history$output_mode == "compact" & history$threads == 1L,
    ,
    drop = FALSE
  ]
  comparison_revision <- nearest_tested_ancestor(intersect(
    regulation_runs$source_revision,
    transcript_controls$source_revision
  ))
  regulation <- regulation_runs[
    regulation_runs$source_revision == comparison_revision,
    ,
    drop = FALSE
  ]
  transcript_control <- transcript_controls[
    transcript_controls$source_revision == comparison_revision,
    ,
    drop = FALSE
  ]
  comparison <- rbind(
    transcript_control[nrow(transcript_control), , drop = FALSE],
    regulation[nrow(regulation), , drop = FALSE]
  )
  if (nrow(comparison) == 2L) {
    regulation_table <- data.frame(
      lane = c("transcripts", "transcripts + regulation/motif"),
      revision = comparison$revision,
      input_variants = format(comparison$variants, big.mark = ","),
      resident_interval_features = format(
        comparison$regulation_features,
        big.mark = ","
      ),
      output_rows = format(comparison$annotated_rows, big.mark = ","),
      median_seconds = comparison$median_seconds,
      variants_per_second = format(
        round(comparison$variants_per_second),
        big.mark = ","
      ),
      output_rows_per_second = format(
        round(comparison$annotated_rows / comparison$median_seconds),
        big.mark = ","
      ),
      stringsAsFactors = FALSE
    )
    knitr::kable(regulation_table, row.names = FALSE)
  }
}
```

This is one pinned-core comparison at one source revision. Both lanes load the
same transcript model and consume the same coordinate-sorted GIAB sample. The
second lane additionally loads every VEP-admitted GRCh38 RegulatoryFeature and
MotifFeature interval and emits their compact consequence rows. The comparison
therefore measures candidate traversal and the extra output together; it is not
an isolated interval-lookup microbenchmark.

## Paired breakends

```{r paired-breakend-throughput}
breakend_runs <- history[
  history$workload == "ensembl116_grch38_final_breakend_200k" &
    history$output_mode == "compact" &
    history$threads == 1L,
  ,
  drop = FALSE
]
if (nrow(breakend_runs)) {
  breakend <- breakend_runs[nrow(breakend_runs), , drop = FALSE]
  breakend_table <- data.frame(
    revision = breakend$revision,
    paired_events = format(
      breakend$variants,
      big.mark = ",",
      scientific = FALSE
    ),
    consequence_rows = format(
      breakend$annotated_rows,
      big.mark = ",",
      scientific = FALSE
    ),
    rows_per_event = round(
      breakend$annotated_rows / breakend$variants,
      2
    ),
    median_seconds = breakend$median_seconds,
    paired_events_per_second = format(
      round(breakend$variants_per_second),
      big.mark = ",",
      scientific = FALSE
    ),
    consequence_rows_per_second = format(
      round(breakend$annotated_rows / breakend$median_seconds),
      big.mark = ",",
      scientific = FALSE
    ),
    stringsAsFactors = FALSE
  )
  knitr::kable(breakend_table, row.names = FALSE)
}
```

This measurement used the complete receipt-hashed GRCh38 model, compact output,
one DuckDB thread pinned with `taskset -c 2`, 15 timed passes, and a 10,000-event
warmup. Model loading and staging were outside the timed region. The site-rate
headline alone would be misleading: every paired event produced an average of
93.83 transcript consequence rows, and the checksum covered every emitted
mask.

## Validated CDS projection cache

The `7dd90ce8` extension source is byte-identical to the immediate pre-cache
parent; the intervening commits changed only documentation and benchmark data.
The `f649d724` rows measure the cache with the same final model, input tables,
single pinned CPU, ordered query, compact output, and warmup size. Positive
change means less elapsed time. The non-SNV and mixed lanes avoid repeated
binary searches for both CDS endpoints. The sorted SNV and ordinary GIAB lanes
show the control cost because they rarely or never consume that cache.

```{r projection-cache}
cache_before <- history[
  history$revision == "7dd90ce8" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
cache_after <- history[
  history$revision == "f649d724" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
cache_table <- merge(
  cache_before,
  cache_after,
  by = c("workload", "variants", "annotated_rows"),
  suffixes = c("_before", "_after")
)
cache_order <- c(
  "ensembl116_grch38_final_coding_mixed_200k",
  "ensembl116_grch38_final_coding_nonsnv_36k",
  "ensembl116_grch38_final_coding_snv_200k",
  "ensembl116_grch38_final_giab_sites_hash40"
)
cache_labels <- c(
  "mixed coding",
  "non-SNV coding",
  "coding SNV",
  "GIAB sites"
)
cache_table <- cache_table[
  match(cache_order, cache_table$workload),
  ,
  drop = FALSE
]
cache_table$workload <- cache_labels
cache_table$elapsed_change_percent <- round(
  100 * (cache_table$median_seconds_before -
    cache_table$median_seconds_after) /
    cache_table$median_seconds_before,
  1
)
cache_table$variants <- format(
  cache_table$variants, big.mark = ",", scientific = FALSE
)
cache_table$annotated_rows <- format(
  cache_table$annotated_rows, big.mark = ",", scientific = FALSE
)
cache_table$variants_per_second_before <- round(
  cache_table$variants_per_second_before
)
cache_table$variants_per_second_after <- round(
  cache_table$variants_per_second_after
)
names(cache_table) <- c(
  "workload", "variants", "output rows", "seconds before", "variants/s before",
  "seconds after", "variants/s after", "elapsed change (%)"
)
knitr::kable(cache_table, row.names = FALSE)
```

## Reused intron coordinates for mismatch islands

The `ecb7eaae` rows are the merged projection-cache source immediately before
the splice-classifier change. The `57b5b566` rows reuse the intron coordinates
already resolved for each exon pair, avoid calling the shared predicate when
neither the intron nor a splice or short-intron window can contribute, and keep
that predicate inlined in the sorted point classifier. Both revisions use the
same final model, input tables, pinned CPU, ordered query, compact output, and
10,000-variant warmup. Positive change means less elapsed time.

```{r splice-intron-reuse}
splice_before <- history[
  history$revision == "ecb7eaae" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
splice_after <- history[
  history$revision == "57b5b566" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
splice_table <- merge(
  splice_before,
  splice_after,
  by = c("workload", "variants", "annotated_rows"),
  suffixes = c("_before", "_after")
)
splice_order <- c(
  "ensembl116_grch38_final_coding_nonsnv_36k",
  "ensembl116_grch38_final_coding_mixed_200k",
  "ensembl116_grch38_final_coding_snv_200k"
)
splice_labels <- c("non-SNV coding", "mixed coding", "coding SNV control")
splice_table <- splice_table[
  match(splice_order, splice_table$workload),
  ,
  drop = FALSE
]
splice_table$workload <- splice_labels
splice_table$elapsed_change_percent <- round(
  100 * (splice_table$median_seconds_before -
    splice_table$median_seconds_after) /
    splice_table$median_seconds_before,
  1
)
splice_table$variants <- format(
  splice_table$variants, big.mark = ",", scientific = FALSE
)
splice_table$annotated_rows <- format(
  splice_table$annotated_rows, big.mark = ",", scientific = FALSE
)
splice_table$variants_per_second_before <- round(
  splice_table$variants_per_second_before
)
splice_table$variants_per_second_after <- round(
  splice_table$variants_per_second_after
)
names(splice_table) <- c(
  "workload", "variants", "output rows", "seconds before", "variants/s before",
  "seconds after", "variants/s after", "elapsed change (%)"
)
knitr::kable(splice_table, row.names = FALSE)
```

The coding-SNV row is a control: that route does not use mismatch-island
classification. Its 5 ms movement is reported as measurement noise rather than
attributed to this change.

## Reused coding projection fact for NMD

The `1903f7fb` rows are the exact merged source before this change. The
`a346a7aa` rows record the first projection-cache optimization: retain the
`CDS end <= 101` comparison already established by the sequence classifier and
consume it when VEP 116 NMD prediction needs the same fact. That revision is a
measured historical intermediate. The later executable-plugin audit proved
that the fact is interchangeable only when the minimized edit and VEP feature
cover exactly the same genomic bases. Both revisions below were rebuilt and
measured adjacently on the same final model, input tables, pinned CPU, ordered
query, compact output, and 10,000-variant warmup. Positive change means less
elapsed time.

```{r nmd-projection-reuse}
nmd_before <- history[
  history$revision == "1903f7fb" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
nmd_after <- history[
  history$revision == "a346a7aa" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
nmd_table <- merge(
  nmd_before,
  nmd_after,
  by = c("workload", "variants", "annotated_rows"),
  suffixes = c("_before", "_after")
)
nmd_order <- c(
  "ensembl116_grch38_final_coding_mixed_200k",
  "ensembl116_grch38_final_coding_nonsnv_36k",
  "ensembl116_grch38_final_coding_snv_200k"
)
nmd_labels <- c("mixed coding", "non-SNV coding", "coding SNV control")
nmd_table <- nmd_table[
  match(nmd_order, nmd_table$workload),
  ,
  drop = FALSE
]
nmd_table$workload <- nmd_labels
nmd_table$elapsed_change_percent <- round(
  100 * (nmd_table$median_seconds_before - nmd_table$median_seconds_after) /
    nmd_table$median_seconds_before,
  1
)
nmd_table$variants <- format(
  nmd_table$variants, big.mark = ",", scientific = FALSE
)
nmd_table$annotated_rows <- format(
  nmd_table$annotated_rows, big.mark = ",", scientific = FALSE
)
nmd_table$variants_per_second_before <- round(
  nmd_table$variants_per_second_before
)
nmd_table$variants_per_second_after <- round(
  nmd_table$variants_per_second_after
)
names(nmd_table) <- c(
  "workload", "variants", "output rows", "seconds before", "variants/s before",
  "seconds after", "variants/s after", "elapsed change (%)"
)
knitr::kable(nmd_table, row.names = FALSE)
```

The coding-SNV route does not carry the reused projection fact. Its 5 ms
movement is reported as a control rather than attributed to the NMD change.

## Cost of exact VEP feature projection

The `c361346f` correction retains the cached fact only for identical genomic
spans. Wider equal-length features and insertions instead project the same
parent `TranscriptVariation` state read by `NMD.pm`. These rows compare that
exact implementation with the earlier over-broad cache revision. Output row
counts and consequence-mask checksums are identical for every workload.
Negative elapsed change means the exact implementation took longer; the SNV
lane is a control because it does not use this NMD path.

```{r nmd-feature-projection}
nmd_feature_before <- history[
  history$revision == "a346a7aa" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "checksum_value",
    "median_seconds", "variants_per_second")
]
nmd_feature_after <- history[
  history$revision == "c361346f" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "checksum_value",
    "median_seconds", "variants_per_second")
]
nmd_feature_table <- merge(
  nmd_feature_before,
  nmd_feature_after,
  by = c("workload", "variants", "annotated_rows", "checksum_value"),
  suffixes = c("_before", "_exact")
)
nmd_feature_order <- c(
  "ensembl116_grch38_final_coding_mixed_200k",
  "ensembl116_grch38_final_coding_nonsnv_36k",
  "ensembl116_grch38_final_coding_snv_200k"
)
nmd_feature_labels <- c(
  "mixed coding", "non-SNV coding", "coding SNV control"
)
nmd_feature_table <- nmd_feature_table[
  match(nmd_feature_order, nmd_feature_table$workload),
  ,
  drop = FALSE
]
nmd_feature_table$workload <- nmd_feature_labels
nmd_feature_table$elapsed_change_percent <- round(
  100 * (
    nmd_feature_table$median_seconds_before -
      nmd_feature_table$median_seconds_exact
  ) / nmd_feature_table$median_seconds_before,
  1
)
nmd_feature_table$variants <- format(
  nmd_feature_table$variants, big.mark = ",", scientific = FALSE
)
nmd_feature_table$annotated_rows <- format(
  nmd_feature_table$annotated_rows, big.mark = ",", scientific = FALSE
)
nmd_feature_table$variants_per_second_before <- round(
  nmd_feature_table$variants_per_second_before
)
nmd_feature_table$variants_per_second_exact <- round(
  nmd_feature_table$variants_per_second_exact
)
nmd_feature_table$checksum_value <- NULL
names(nmd_feature_table) <- c(
  "workload", "variants", "output rows", "seconds before", "variants/s before",
  "seconds exact", "variants/s exact", "elapsed change (%)"
)
knitr::kable(nmd_feature_table, row.names = FALSE)
```

## Last measured transcript-only GRCh38 model resources

```{r resources}
last_resource <- resources[nrow(resources), , drop = FALSE]
resource_table <- data.frame(
  measure = c(
    "revision", "model artifact", "model load", "process RSS before load",
    "process RSS after load", "peak process RSS", "process RSS after drop",
    "DuckDB base-table cache", "planned scratch per worker",
    "observed RSS change on first workspace", "largest transcript run",
    "largest prepared CDS"
  ),
  value = c(
    last_resource$revision,
    sprintf("%.3f GiB", last_resource$model_artifact_bytes / 1024^3),
    sprintf("%.3f s", last_resource$model_load_seconds),
    sprintf("%.3f GiB", last_resource$rss_baseline_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$rss_model_loaded_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$rss_peak_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$rss_after_drop_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$duckdb_base_table_cache_bytes / 1024^3),
    sprintf("%.3f MiB", last_resource$workspace_planned_buffer_bytes / 1024^2),
    sprintf("%.3f MiB", last_resource$workspace_rss_delta_kib / 1024),
    format(last_resource$max_transcripts_per_region, big.mark = ","),
    sprintf("%s bases", format(last_resource$max_cds_bytes, big.mark = ","))
  ),
  stringsAsFactors = FALSE
)
knitr::kable(resource_table, row.names = FALSE)
```

This resource row predates the resident regulation/motif relation and remains a
transcript-only baseline; the throughput table above records the feature count
for regulation-enabled runs. The artifact size is a compact DuckDB containing
only `model_regions`, `model_transcripts`, and `model_receipt`; benchmark inputs
are excluded. RSS is
for one R/DuckDB process, not a C-heap estimate. The load peak includes both the
DuckDB scan/cache and construction of the second immutable C representation.
Dropping the model releases the C representation, but DuckDB retains its table
cache in the process. The planned per-worker number is the exact sum of the C
buffer capacities (active and candidate transcript indices, one exon cursor per
transcript, and worst-case edit/CDS/peptide scratch); allocator metadata is not
included. The much smaller first-workspace RSS change reflects lazy or reused
pages, not a smaller capacity.

The historical standalone model ABI 0.12 cache adds three `uint32_t` arrays and one
`uint8_t` array per transcript. Its exact element payload for this model is
`r format(13 * last_resource$transcripts[[1L]], big.mark = ",", scientific = FALSE)`
bytes (`r sprintf("%.3f MiB", 13 * last_resource$transcripts[[1L]] / 1024^2)`),
plus four allocator records. This deterministic allocation is not retroactively
included in the older `b204dd49` RSS measurement above.

The production form is run with:

```sh
Rscript benchmarks/duckvep_throughput.R \
  --database /path/to/staged.duckdb \
  --variants-table bench_variants \
  --workload-name ensembl116_grch38_final_giab_sites_hash40 \
  --variants 100957 --passes 9 --warmup 10000 --threads 1 \
  --output compact
```

Add `--regulatory` and use the
`ensembl116_grch38_final_giab_sites_hash40_regulation` workload name to load
`duckvep_regulation_features` and record the integrated core-feature lane.
