---
title: "DuckVEP sorted-stream throughput"
output: github_document
---

<!-- duckvep_throughput.md is generated from duckvep_throughput.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
history <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_throughput.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
history$revision <- substr(history$source_revision, 1, 8)
resources <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_model_resources.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
resources$revision <- substr(resources$source_revision, 1, 8)
```

This benchmark drives either the rich `duckvep_annotate` result or the numeric
`duckvep_annotate_compact` result through DuckDB's stable extension API with
nondecreasing sequence-region and position keys. It records candidate sweep,
consequence classification, list materialization, `unnest`, and final aggregation
together. Model loading and input staging are outside the timed pass.

The checked-in one-transcript workload is a repeatable adapter floor. The final
production-density workload uses the receipt-hashed Ensembl 116 GRCh38 model:
194 sequence regions, 644,427 transcripts, 5,068,416 exon memberships, and
369,631 sequence-backed coding transcripts. Its ordinary topology workload is
a deterministic 1-in-40 hash sample of 100,957 sorted GIAB sites. The coding
workloads deliberately repeat a frozen set to stabilize one-core kernel timing:
200,000 SNVs repeat 1,000 distinct alleles; the 36,000 non-SNV workload repeats
1,000 MNVs and 8,000 length-changing alleles four times; the 200,000 mixed
workload is 10% SNV, 10% MNV, 40% lengthening, and 40% shortening. Repetition
favours warm model and sequence caches, so these rows measure the resident
execution lanes, not file ingestion or an end-to-end genome run.

Earlier chromosome-only and complete-primary staging rows remain in the ledger
as historical measurements. They have different transcript inventories and
checksums and must not be treated as revisions of the final receipt.
The production query restates `ORDER BY` in the timed input CTE, so its result
includes the cost of enforcing the kernel's nondecreasing-input contract even
though the staging table was already written in that order. The `79cfaaf2`
rows predate that review correction and relied on the single-thread physical
scan preserving insertion order; they remain only as historical measurements.

## Recorded runs

```{r runs}
table <- history[, c(
  "run_date", "revision", "workload", "output_mode", "threads", "variants",
  "transcripts", "exons", "annotated_rows", "passes", "min_seconds",
  "median_seconds", "max_seconds", "variants_per_second", "ns_per_variant",
  "cpu"
)]
table$output_rows_per_second <- round(
  table$annotated_rows / table$median_seconds
)
table$variants <- format(table$variants, big.mark = ",", scientific = FALSE)
table$transcripts <- format(table$transcripts, big.mark = ",", scientific = FALSE)
table$exons <- format(table$exons, big.mark = ",", scientific = FALSE)
table$annotated_rows <- format(
  table$annotated_rows,
  big.mark = ",",
  scientific = FALSE
)
table$variants_per_second <- round(table$variants_per_second)
table$ns_per_variant <- round(table$ns_per_variant, 1)
table$output_rows_per_second <- format(
  table$output_rows_per_second,
  big.mark = ",",
  scientific = FALSE
)
knitr::kable(table, row.names = FALSE)
```

Each pass consumes every staged input and checks output cardinality plus either
the rendered consequence-byte total or the numeric consequence-mask sum. Rows
with different workload, output mode, thread count, host, or variant count are
separate measurements and should not be compared as if only the engine changed.

## Validated CDS projection cache

The `7dd90ce8` extension source is byte-identical to the immediate pre-cache
parent; the intervening commits changed only documentation and benchmark data.
The `f649d724` rows measure the cache with the same final model, input tables,
single pinned CPU, ordered query, compact output, and warmup size. Positive
change means less elapsed time. The non-SNV and mixed lanes avoid repeated
binary searches for both CDS endpoints. The sorted SNV and ordinary GIAB lanes
show the control cost because they rarely or never consume that cache.

```{r projection-cache}
cache_before <- history[
  history$revision == "7dd90ce8" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
cache_after <- history[
  history$revision == "f649d724" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
cache_table <- merge(
  cache_before,
  cache_after,
  by = c("workload", "variants", "annotated_rows"),
  suffixes = c("_before", "_after")
)
cache_order <- c(
  "ensembl116_grch38_final_coding_mixed_200k",
  "ensembl116_grch38_final_coding_nonsnv_36k",
  "ensembl116_grch38_final_coding_snv_200k",
  "ensembl116_grch38_final_giab_sites_hash40"
)
cache_labels <- c(
  "mixed coding",
  "non-SNV coding",
  "coding SNV",
  "GIAB sites"
)
cache_table <- cache_table[
  match(cache_order, cache_table$workload),
  ,
  drop = FALSE
]
cache_table$workload <- cache_labels
cache_table$elapsed_change_percent <- round(
  100 * (cache_table$median_seconds_before -
    cache_table$median_seconds_after) /
    cache_table$median_seconds_before,
  1
)
cache_table$variants <- format(
  cache_table$variants, big.mark = ",", scientific = FALSE
)
cache_table$annotated_rows <- format(
  cache_table$annotated_rows, big.mark = ",", scientific = FALSE
)
cache_table$variants_per_second_before <- round(
  cache_table$variants_per_second_before
)
cache_table$variants_per_second_after <- round(
  cache_table$variants_per_second_after
)
names(cache_table) <- c(
  "workload", "variants", "output rows", "seconds before", "variants/s before",
  "seconds after", "variants/s after", "elapsed change (%)"
)
knitr::kable(cache_table, row.names = FALSE)
```

## Final GRCh38 model resources

```{r resources}
last_resource <- resources[nrow(resources), , drop = FALSE]
resource_table <- data.frame(
  measure = c(
    "revision", "model artifact", "model load", "process RSS before load",
    "process RSS after load", "peak process RSS", "process RSS after drop",
    "DuckDB base-table cache", "planned scratch per worker",
    "observed RSS change on first workspace", "largest transcript run",
    "largest prepared CDS"
  ),
  value = c(
    last_resource$revision,
    sprintf("%.3f GiB", last_resource$model_artifact_bytes / 1024^3),
    sprintf("%.3f s", last_resource$model_load_seconds),
    sprintf("%.3f GiB", last_resource$rss_baseline_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$rss_model_loaded_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$rss_peak_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$rss_after_drop_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$duckdb_base_table_cache_bytes / 1024^3),
    sprintf("%.3f MiB", last_resource$workspace_planned_buffer_bytes / 1024^2),
    sprintf("%.3f MiB", last_resource$workspace_rss_delta_kib / 1024),
    format(last_resource$max_transcripts_per_region, big.mark = ","),
    sprintf("%s bases", format(last_resource$max_cds_bytes, big.mark = ","))
  ),
  stringsAsFactors = FALSE
)
knitr::kable(resource_table, row.names = FALSE)
```

The artifact size is a compact DuckDB containing only `model_regions`,
`model_transcripts`, and `model_receipt`; benchmark inputs are excluded. RSS is
for one R/DuckDB process, not a C-heap estimate. The load peak includes both the
DuckDB scan/cache and construction of the second immutable C representation.
Dropping the model releases the C representation, but DuckDB retains its table
cache in the process. The planned per-worker number is the exact sum of the C
buffer capacities (active and candidate transcript indices, one exon cursor per
transcript, and worst-case edit/CDS/peptide scratch); allocator metadata is not
included. The much smaller first-workspace RSS change reflects lazy or reused
pages, not a smaller capacity.

The standalone model ABI 0.12 cache adds three `uint32_t` arrays and one
`uint8_t` array per transcript. Its exact element payload for this model is
`r format(13 * last_resource$transcripts[[1L]], big.mark = ",", scientific = FALSE)`
bytes (`r sprintf("%.3f MiB", 13 * last_resource$transcripts[[1L]] / 1024^2)`),
plus four allocator records. This deterministic allocation is not retroactively
included in the older `b204dd49` RSS measurement above.

The production form is run with:

```sh
Rscript benchmarks/duckvep_throughput.R \
  --database /path/to/staged.duckdb \
  --variants-table bench_variants \
  --workload-name ensembl116_grch38_final_giab_sites_hash40 \
  --variants 100957 --passes 9 --warmup 10000 --threads 1 \
  --output compact
```
