---
title: "DuckVEP sorted-stream throughput"
output: github_document
---

<!-- duckvep_throughput.md is generated from duckvep_throughput.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
history <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_throughput.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(
    source_revision = "character",
    checksum_value = "character",
    extension_sha256 = "character",
    model_database_sha256 = "character",
    model_logical_sha256 = "character",
    model_source_manifest_sha256 = "character",
    model_reference_sha256 = "character",
    model_region_ordinal_sha256 = "character",
    variants_database_sha256 = "character",
    corpus_source_sha256 = "character",
    reference_fasta_sha256 = "character",
    reference_fai_sha256 = "character",
    reference_gzi_sha256 = "character",
    fingerprint_xor_hash = "character",
    fingerprint_sum_hash = "character"
  )
)
history$revision <- substr(history$source_revision, 1, 8)
resources <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_model_resources.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
resources$revision <- substr(resources$source_revision, 1, 8)
dense_threading <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_dense_threading.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(
    source_revision = "character",
    xor_hash = "character",
    sum_hash = "character"
  )
)
dense_threading$revision <- substr(dense_threading$source_revision, 1, 8)
stopifnot(
  length(unique(dense_threading$rss_measurement)) == 1L,
  identical(
    unique(dense_threading$rss_measurement),
    "GNU /usr/bin/time -v process maximum resident set size"
  )
)

nearest_tested_ancestor <- function(revisions) {
  revisions <- unique(revisions[nzchar(revisions)])
  distance <- vapply(
    revisions,
    function(revision) {
      is_ancestor <- system2(
        "git",
        c("-C", root, "merge-base", "--is-ancestor", revision, "HEAD"),
        stdout = FALSE,
        stderr = FALSE
      )
      if (is_ancestor != 0L) {
        return(Inf)
      }
      as.numeric(system2(
        "git",
        c("-C", root, "rev-list", "--count", paste0(revision, "..HEAD")),
        stdout = TRUE,
        stderr = FALSE
      ))
    },
    numeric(1)
  )
  if (!length(distance) || all(!is.finite(distance))) {
    return(NA_character_)
  }
  revisions[[which.min(distance)]]
}
```

This benchmark drives the internal rich `_duckvep_annotate_small_rich`, numeric
`_duckvep_annotate_small_compact`, independent-event `_duckvep_annotate_small_hgvs`,
or fused rich-plus-HGVS result through DuckDB's stable extension API with
nondecreasing sequence-region and position keys. It records candidate sweep,
consequence classification, list materialization, `unnest`, and final aggregation
together. Model loading and input staging are outside the timed pass.

The checked-in one-transcript workload is a repeatable adapter floor. The final
production-density workload uses the receipt-hashed Ensembl 116 GRCh38 model:
194 sequence regions, 644,427 transcripts, 5,068,416 exon memberships, and
369,631 sequence-backed coding transcripts. The regulation-enabled form adds
380,818 RegulatoryFeature and 1,002,762 MotifFeature intervals in an independent
resident SoA. Its ordinary topology workload is a deterministic 1-in-40 hash
sample of 100,957 sorted GIAB sites. The coding
workloads deliberately repeat a frozen set to stabilize one-core kernel timing:
200,000 SNVs repeat 1,000 distinct alleles; the 36,000 non-SNV workload repeats
1,000 MNVs and 8,000 length-changing alleles four times; the 200,000 mixed
workload is 10% SNV, 10% MNV, 40% lengthening, and 40% shortening. Repetition
favours warm model and sequence caches, so these rows measure the resident
execution lanes, not file ingestion or an end-to-end genome run.

The paired-breakend workload repeats the 1,004-event multichromosome seed-31
VEP differential to 200,000 sorted endpoint pairs. Both endpoints participate
in candidate discovery and each transcript receives the union of the two
endpoint consequence sets. Its much larger transcript fan-out makes output
rows/s as important as paired events/s.

Earlier chromosome-only and complete-primary staging rows remain in the ledger
as historical measurements. They have different transcript inventories and
checksums and must not be treated as revisions of the final receipt.
The production query restates `ORDER BY` in the timed input CTE, so its result
includes the cost of enforcing the kernel's nondecreasing-input contract even
though the staging table was already written in that order. The `79cfaaf2`
rows predate that review correction and relied on the single-thread physical
scan preserving insertion order; they remain only as historical measurements.

## Recorded runs

```{r runs}
table <- history[, c(
  "run_date", "revision", "workload", "output_mode", "threads",
  "input_partitions", "transcript_distance", "variants", "transcripts", "exons",
  "regulation_features", "annotated_rows", "passes", "min_seconds",
  "median_seconds", "max_seconds", "variants_per_second", "ns_per_variant", "cpu",
  "cpu_affinity"
)]
table$output_rows_per_second <- round(
  table$annotated_rows / table$median_seconds
)
table$variants <- format(table$variants, big.mark = ",", scientific = FALSE)
table$transcripts <- format(table$transcripts, big.mark = ",", scientific = FALSE)
table$exons <- format(table$exons, big.mark = ",", scientific = FALSE)
table$regulation_features <- format(
  table$regulation_features,
  big.mark = ",",
  scientific = FALSE
)
table$annotated_rows <- format(
  table$annotated_rows,
  big.mark = ",",
  scientific = FALSE
)
table$variants_per_second <- round(table$variants_per_second)
table$ns_per_variant <- round(table$ns_per_variant, 1)
table$output_rows_per_second <- format(
  table$output_rows_per_second,
  big.mark = ",",
  scientific = FALSE
)
knitr::kable(table, row.names = FALSE)
```

Each pass consumes every staged input and checks output cardinality plus either
the rendered consequence-byte total or the numeric consequence-mask sum. Rows
with different workload, output mode, thread count, host, or variant count are
separate measurements and should not be compared as if only the engine changed.
`cpu_affinity` is read from the benchmark process itself; a single integer
records an externally pinned run, while a range records a scheduler-visible CPU
set.

## Borrowed-CDS projection bounds check

The `--fixture-indels` workload alternates `T>TAC` insertions and `TA>T`
deletions at position 124 of the checked-in one-transcript model and FASTA.
Unlike the single-SNV adapter control, these length-changing alleles exercise
CDS edit projection and HGVS construction. They remain independent events,
not one compound haplotype. Model loading and input generation are untimed.

```{r cds-slice-guard}
cds_guard <- history[
  history$workload == "fixture_one_transcript_sorted_indels" &
    history$source_revision %in% c(
      "e5f1f2690264fe76d0cc9b9ee6134dd76c77dab8",
      "10320db65b6494de2753fcd766d1754fbcc3a647"
    ),
]
cds_guard <- cds_guard[match(c("e5f1f26", "10320db"), substr(cds_guard$revision, 1L, 7L)), ]
stopifnot(nrow(cds_guard) == 2L, !anyNA(cds_guard$source_revision))
same_workload <- c(
  "host", "cpu", "cpu_affinity", "duckdb_version", "workload", "output_mode",
  "variants", "annotated_rows", "threads", "input_partitions", "transcript_distance",
  "passes", "warmup_variants", "reference_fasta_sha256", "reference_fai_sha256",
  "checksum_kind", "checksum_value"
)
stopifnot(all(vapply(cds_guard[same_workload], function(x) length(unique(x)) == 1L, TRUE)))
knitr::kable(cds_guard[, c(
  "revision", "variants", "annotated_rows", "threads", "passes",
  "min_seconds", "median_seconds", "max_seconds", "checksum_value"
)], row.names = FALSE)
```

Both clean in-tree builds use DuckDB `r cds_guard$duckdb_version[[1L]]` on
`r cds_guard$cpu[[1L]]`, pinned to CPU `r cds_guard$cpu_affinity[[1L]]`, with
100,000 warm-up events and a 5,000-base transcript halo. The baseline commit
adds only the benchmark case; the second commit adds the borrowed-slice checks
and genomic carrier differential. Their median difference is
`r sprintf("%+.2f%%", 100 * (cds_guard$median_seconds[[2L]] / cds_guard$median_seconds[[1L]] - 1))`.
This narrow valid-model workload does not measure malformed-model rejection,
file ingestion, cohort state, sort-plus-phased execution, or peak memory.

```{r terminal-codon-window-control}
terminal_window <- history[history$source_revision ==
  "35bfb85896200b9dd5319ca49147076de511f878" &
  history$workload == "fixture_one_transcript_sorted_indels", ]
stopifnot(nrow(terminal_window) == 1L,
  all(vapply(same_workload, function(field)
    identical(terminal_window[[field]], cds_guard[[field]][2L]), TRUE)))
```

The codon-window correction at `r terminal_window$revision` reran this identical
one-million-input/one-million-output indel/HGVS workload: five passes, one DuckDB
thread pinned to CPU 2, `r terminal_window$checksum_value` rendered HGVS/status
bytes. Its median was `r sprintf("%.3f", terminal_window$median_seconds)` seconds
(range `r sprintf("%.3f", terminal_window$min_seconds)`–`r sprintf("%.3f", terminal_window$max_seconds)`),
`r sprintf("%+.2f%%", 100 * (terminal_window$median_seconds / cds_guard$median_seconds[2L] - 1))`
versus the nearest identical recorded workload at `10320db`. This is an observed
single-run comparison, not a no-regression claim or attribution to one intervening
change. It exercises CDS edits and HGVS but not the newly corrected terminal
partial-codon/UTR sites; a targeted throughput measurement for those sites is still
missing. The byte total is not a full-row fingerprint.

## Annotation-dense transcript distance and ordered parallel partitions

```{r dense-threading}
dense_keys <- paste(
  dense_threading$xor_hash,
  dense_threading$sum_hash,
  sep = ":"
)
fingerprints_per_distance <- vapply(
  split(dense_keys, dense_threading$transcript_distance),
  function(keys) length(unique(keys)),
  integer(1L)
)
stopifnot(all(fingerprints_per_distance == 1L))

single_seconds <- setNames(
  dense_threading$median_seconds[dense_threading$threads == 1L],
  dense_threading$transcript_distance[dense_threading$threads == 1L]
)
single_rss <- setNames(
  dense_threading$maximum_rss_kib[dense_threading$threads == 1L],
  dense_threading$transcript_distance[dense_threading$threads == 1L]
)
dense_table <- data.frame(
  revision = dense_threading$revision,
  transcript_distance = format(
    dense_threading$transcript_distance,
    big.mark = ",",
    scientific = FALSE
  ),
  threads = dense_threading$threads,
  ordered_partitions = dense_threading$input_partitions,
  input_alleles = format(
    dense_threading$variants,
    big.mark = ",",
    scientific = FALSE
  ),
  output_rows = format(
    dense_threading$annotated_rows,
    big.mark = ",",
    scientific = FALSE
  ),
  median_seconds = dense_threading$median_seconds,
  input_alleles_per_second = format(
    round(dense_threading$variants / dense_threading$median_seconds),
    big.mark = ",",
    scientific = FALSE
  ),
  output_rows_per_second = format(
    round(dense_threading$annotated_rows / dense_threading$median_seconds),
    big.mark = ",",
    scientific = FALSE
  ),
  speedup = round(
    unname(single_seconds[as.character(dense_threading$transcript_distance)]) /
      dense_threading$median_seconds,
    2L
  ),
  peak_RSS_GiB = round(dense_threading$maximum_rss_kib / 1024^2, 3L),
  RSS_above_one_thread_MiB = round(
    (
      dense_threading$maximum_rss_kib -
        unname(single_rss[as.character(dense_threading$transcript_distance)])
    ) / 1024,
    2L
  ),
  stringsAsFactors = FALSE
)
knitr::kable(dense_table, row.names = FALSE)
```

This is the release-116 GRCh38 model with all 644,427 admitted transcripts,
5,068,416 exon memberships, and 1,383,580 regulation/motif intervals. The
517,097 ClinVar alleles were selected from 318 annotation-dense genomic tiles,
not diluted with transcript-sparse whole-genome positions. The largest REF and
ALT alleles are 9,974 and 9,971 bases. Each row is the median of five complete
passes after a 100,000-allele warmup. Model loading and corpus staging are
outside the timed pass; compact list production, `unnest`, and final aggregation
are included.

One-thread runs are pinned to physical P core 2. Four-thread runs are pinned to
physical P cores 2, 4, 6, and 8 and explicitly split the already ordered input
into four disjoint coordinate ranges. Merely setting DuckDB `threads = 4` does
not split one scalar-function branch. Each concurrent scalar callback borrows an
exclusive mutable workspace from the shared immutable model; a partition is not
permanently bound to one workspace. `input_variant_index` is the global input
ordinal, while `annotation_index` is the one-based position inside each variant's
result list. Consumers that require a canonical materialized row order use
`ORDER BY input_variant_index, annotation_index` after the parallel union.

The one- and four-partition runs have identical XOR and sum fingerprints over
every `(input_variant_index, annotation_index, public row)` at each tested
distance. These commutative fingerprints prove row-multiset and per-input-list
equality, not global emission order. The 50,000-base case emits 88,784,213 rows
and still scales from `r format(round(517097 / 4.161), big.mark = ",")` to
`r format(round(517097 / 1.342), big.mark = ",")` input alleles/s. Peak RSS is
GNU `/usr/bin/time -v` maximum resident set size for the complete one-process
R/DuckDB invocation, including model load, warmup, and all five passes. Across
the matrix, the largest observed four-thread premium over its one-thread pair is
`r sprintf("%.2f MiB", max(dense_table$RSS_above_one_thread_MiB))`. Source
ownership establishes that callbacks share the immutable model; the RSS result
shows no model-sized process increase for these four-branch runs but is not
allocation attribution. Wider transcript distance increases candidate pairs and
output cardinality, not model ownership. These observations apply to the tested
0, 5,000, 10,000, and 50,000-base configurations; arbitrary partition counts
still require measurement.

The benchmark source accepts every transcript distance from zero through
`UINT32_MAX`; no classifier or sweep buffer is sized for a 5,000-base special
case. VEP's separate fixed 5,000-base paired-breakend close-to-feature rule is a
biological compatibility predicate applied after transcript discovery and must
not be confused with the configurable transcript-distance search.

## Routine optimization backlog

The declared fixed-event consequence contract no longer depends on another
semantic rewrite. Future speed work should preserve the conformance and
full-row-fingerprint gates while attacking measured costs in this order:

1. write compact SoA fields directly into DuckDB result vectors, avoiding the
   compact-list-to-`unnest` materialization round trip;
2. make the common coding-SNV route leaner while retaining one generated effect
   composition authority;
3. reuse genomic-to-transcript, exon, splice-window, and coding projection facts
   that are currently rediscovered by adjacent predicates;
4. separate HGVS fact production from string rendering and reuse one normalized
   shift result across HGVSc/HGVSn/HGVSp;
5. expose a persistent sorted table-function sweep so one logical input stream
   does not reseed at every DuckDB vector edge; and
6. make disjoint ordered partitions a first-class planner surface, keeping
   stable identifiers numeric and joining presentation strings late.

Exploratory local profiling suggested low branch-miss and cache-miss rates in the
candidate sweep and a small candidate-discovery share of dense runtime. That profile
is not retained as a checked comparable benchmark, so it is an avenue-selection hint,
not a quantitative claim. Broad SIMD work is therefore not the first optimization
target. Each item above needs an adjacent exact-revision comparison on the same model,
corpus, transcript distance, output denominator, core affinity, and fingerprint before
it can be called an improvement.

## Public relation API cost and scaling

```{r public-relation-api}
public_workload <-
  "ensembl116_grch38_giab_hg002_v4_2_1_full_literal_public_relation"
surface_modes <- c("compact", "rich", "hgvs", "rich_hgvs")
surface_threads <- c(1L, 4L)
surface_expected <- as.vector(outer(
  surface_modes,
  surface_threads,
  FUN = function(mode, threads) paste(mode, threads, sep = ":")
))
surface_runs <- history[
  history$workload == public_workload &
    history$output_mode %in% surface_modes &
    history$threads %in% surface_threads &
    history$input_partitions == 1L &
    history$transcript_distance == 5000 &
    history$regulation_features == 1383580L,
  ,
  drop = FALSE
]
surface_groups <- split(surface_runs, surface_runs$source_revision)
surface_complete_revisions <- names(Filter(
  function(x) {
    nrow(x) == length(surface_expected) &&
      setequal(
        paste(x$output_mode, x$threads, sep = ":"),
        surface_expected
      )
  },
  surface_groups
))
surface_revision <- nearest_tested_ancestor(
  surface_complete_revisions
)
surface_current <- surface_runs[
  surface_runs$source_revision == surface_revision,
  ,
  drop = FALSE
]
stopifnot(
  nrow(surface_current) == length(surface_expected),
  setequal(
    paste(surface_current$output_mode, surface_current$threads, sep = ":"),
    surface_expected
  ),
  length(unique(surface_current$variants)) == 1L,
  length(unique(surface_current$annotated_rows)) == 1L,
  length(unique(surface_current$model_database_sha256)) == 1L,
  length(unique(surface_current$variants_database_sha256)) == 1L,
  length(unique(surface_current$corpus_source_sha256)) == 1L,
  all(surface_current$passes == 5L),
  all(surface_current$warmup_variants == 100000L)
)
surface_by_mode <- split(surface_current, surface_current$output_mode)
stopifnot(all(vapply(
  surface_by_mode,
  function(x) {
    length(unique(x$checksum_kind)) == 1L &&
      length(unique(x$checksum_value)) == 1L
  },
  logical(1)
)))
surface_fingerprint_complete <- all(
  !is.na(surface_current$fingerprint_schema_version) &
    nzchar(surface_current$fingerprint_schema_version) &
    !is.na(surface_current$fingerprint_xor_hash) &
    nzchar(surface_current$fingerprint_xor_hash) &
    !is.na(surface_current$fingerprint_sum_hash) &
    nzchar(surface_current$fingerprint_sum_hash)
)
if (surface_fingerprint_complete) {
  stopifnot(all(vapply(
    surface_by_mode,
    function(x) {
      length(unique(x$fingerprint_schema_version)) == 1L &&
        length(unique(x$fingerprint_xor_hash)) == 1L &&
        length(unique(x$fingerprint_sum_hash)) == 1L
    },
    logical(1)
  )))
}
surface_current <- surface_current[order(
  match(surface_current$output_mode, surface_modes),
  surface_current$threads
), , drop = FALSE]
surface_one_core <- setNames(
  surface_current$median_seconds[surface_current$threads == 1L],
  surface_current$output_mode[surface_current$threads == 1L]
)
surface_compact <- setNames(
  surface_current$median_seconds[surface_current$output_mode == "compact"],
  surface_current$threads[surface_current$output_mode == "compact"]
)
surface_table <- data.frame(
  revision = surface_current$revision,
  output = surface_current$output_mode,
  threads = surface_current$threads,
  input_alleles = format(
    surface_current$variants,
    big.mark = ",",
    scientific = FALSE
  ),
  output_rows = format(
    surface_current$annotated_rows,
    big.mark = ",",
    scientific = FALSE
  ),
  median_seconds = surface_current$median_seconds,
  alleles_per_second = format(
    round(surface_current$variants_per_second),
    big.mark = ",",
    scientific = FALSE
  ),
  output_rows_per_second = format(
    round(surface_current$annotated_rows / surface_current$median_seconds),
    big.mark = ",",
    scientific = FALSE
  ),
  scaling_vs_one_core = round(
    unname(surface_one_core[surface_current$output_mode]) /
      surface_current$median_seconds,
    2L
  ),
  elapsed_vs_compact = round(
    surface_current$median_seconds /
      unname(surface_compact[as.character(surface_current$threads)]),
    2L
  ),
  stringsAsFactors = FALSE
)
knitr::kable(surface_table, row.names = FALSE)
```

### Adjacent revision comparison

```{r public-relation-adjacent}
surface_previous_revision <- nearest_tested_ancestor(
  setdiff(surface_complete_revisions, surface_revision)
)
surface_previous <- surface_runs[
  surface_runs$source_revision == surface_previous_revision,
  ,
  drop = FALSE
]
surface_comparison <- merge(
  surface_previous,
  surface_current,
  by = c("output_mode", "threads"),
  suffixes = c("_previous", "_current"),
  sort = FALSE
)
stopifnot(
  nrow(surface_comparison) == length(surface_expected),
  surface_comparison$variants_previous == surface_comparison$variants_current,
  surface_comparison$annotated_rows_previous ==
    surface_comparison$annotated_rows_current,
  surface_comparison$model_logical_sha256_previous ==
    surface_comparison$model_logical_sha256_current,
  surface_comparison$model_source_manifest_sha256_previous ==
    surface_comparison$model_source_manifest_sha256_current,
  surface_comparison$model_reference_sha256_previous ==
    surface_comparison$model_reference_sha256_current,
  surface_comparison$model_region_ordinal_sha256_previous ==
    surface_comparison$model_region_ordinal_sha256_current,
  surface_comparison$corpus_source_sha256_previous ==
    surface_comparison$corpus_source_sha256_current,
  surface_comparison$checksum_kind_previous ==
    surface_comparison$checksum_kind_current,
  surface_comparison$checksum_value_previous ==
    surface_comparison$checksum_value_current
)
surface_comparison <- surface_comparison[order(
  match(surface_comparison$output_mode, surface_modes),
  surface_comparison$threads
), , drop = FALSE]
surface_delta_table <- data.frame(
  output = surface_comparison$output_mode,
  threads = surface_comparison$threads,
  previous_revision = substr(surface_previous_revision, 1L, 8L),
  current_revision = substr(surface_revision, 1L, 8L),
  previous_median_seconds =
    surface_comparison$median_seconds_previous,
  current_median_seconds =
    surface_comparison$median_seconds_current,
  elapsed_change_percent = round(
    100 * (
      surface_comparison$median_seconds_current /
        surface_comparison$median_seconds_previous - 1
    ),
    2L
  ),
  previous_alleles_per_second = format(
    round(surface_comparison$variants_per_second_previous),
    big.mark = ",",
    scientific = FALSE
  ),
  current_alleles_per_second = format(
    round(surface_comparison$variants_per_second_current),
    big.mark = ",",
    scientific = FALSE
  ),
  throughput_change_percent = round(
    100 * (
      surface_comparison$variants_per_second_current /
        surface_comparison$variants_per_second_previous - 1
    ),
    2L
  ),
  stringsAsFactors = FALSE
)
knitr::kable(surface_delta_table, row.names = FALSE)
```

The adjacent table compares the nearest complete ancestor on the same host,
corpus, model semantics, output denominator, core affinity, and pass count.
Positive elapsed change means slower execution; negative throughput change
means fewer input alleles per second. Identical output checksums are required
before a pair is admitted to this table.

This exact-revision comparison uses the complete Ensembl 116 GRCh38 model:
644,427 transcripts, 5,068,416 exon memberships, and all 1,383,580 resident
RegulatoryFeature and MotifFeature intervals. It annotates every one of the
4,095,611 model-addressable literal alleles staged from GIAB HG002 v4.2.1 and
emits 47,835,851 rows at a 5,000-base transcript distance after a 100,000-allele
warmup and five complete passes. One-thread runs are pinned to CPU 2; four-thread
runs are restricted to CPUs 2, 4, 6, and 8. Input VCF decoding, coordinate
sorting, canonical event-table staging, and model loading are outside the timed
pass. Each public row times event validation and family dispatch through
`duckvep_annotate(...)`, native consequence/HGVS work, fixed-schema row expansion,
and checksum aggregation. The current one-thread/four-thread pair requires the
same aggregate checksum for each output contract.
`r if (surface_fingerprint_complete) {
  "The current rows additionally retain an untimed full-row fingerprint."
} else {
  "The current rows do not retain the optional untimed full-row fingerprint."
}`
`elapsed_vs_compact` compares projections at the same thread count; it is not a
comparison against a private kernel lane.

## Historical full-corpus transcript-only diagnostic

```{r full-literal-corpora}
full_workload_labels <- c(
  ensembl116_grch38_clinvar_20260706_full_literal =
    "ClinVar 2026-07-06",
  ensembl116_grch38_giab_hg002_v4_2_1_full_literal =
    "GIAB HG002 v4.2.1"
)
full_runs <- history[
  history$workload %in% names(full_workload_labels) &
    history$output_mode %in% c("compact", "rich", "hgvs") &
    history$threads == 1L &
    history$input_partitions == 1L &
    history$transcript_distance == 5000,
  ,
  drop = FALSE
]
full_revision_groups <- split(full_runs, full_runs$source_revision)
full_complete_revisions <- names(Filter(
  function(rows) {
    nrow(rows) == 6L &&
      setequal(rows$workload, names(full_workload_labels)) &&
      all(vapply(
        split(rows$output_mode, rows$workload),
        function(x) setequal(x, c("compact", "rich", "hgvs")),
        logical(1L)
      ))
  },
  full_revision_groups
))
full_revision <- nearest_tested_ancestor(full_complete_revisions)
full_current <- full_runs[
  full_runs$source_revision == full_revision,
  ,
  drop = FALSE
]
stopifnot(
  nrow(full_current) == 6L,
  all(vapply(
    split(full_current$output_mode, full_current$workload),
    function(x) setequal(x, c("compact", "rich", "hgvs")),
    logical(1L)
  ))
)
mode_order <- c("compact", "rich", "hgvs")
full_current <- full_current[order(
  match(full_current$workload, names(full_workload_labels)),
  match(full_current$output_mode, mode_order)
), , drop = FALSE]
compact_seconds <- setNames(
  full_current$median_seconds[full_current$output_mode == "compact"],
  full_current$workload[full_current$output_mode == "compact"]
)
full_table <- data.frame(
  revision = full_current$revision,
  corpus = unname(full_workload_labels[full_current$workload]),
  output = full_current$output_mode,
  input_alleles = format(
    full_current$variants,
    big.mark = ",",
    scientific = FALSE
  ),
  output_rows = format(
    full_current$annotated_rows,
    big.mark = ",",
    scientific = FALSE
  ),
  rows_per_allele = round(
    full_current$annotated_rows / full_current$variants,
    2L
  ),
  median_seconds = full_current$median_seconds,
  alleles_per_second = format(
    round(full_current$variants_per_second),
    big.mark = ",",
    scientific = FALSE
  ),
  output_rows_per_second = format(
    round(full_current$annotated_rows / full_current$median_seconds),
    big.mark = ",",
    scientific = FALSE
  ),
  elapsed_vs_compact = round(
    full_current$median_seconds /
      unname(compact_seconds[full_current$workload]),
    2L
  ),
  projected_700M_allele_hours = round(
    700e6 / full_current$variants_per_second / 3600,
    2L
  ),
  stringsAsFactors = FALSE
)
knitr::kable(full_table, row.names = FALSE)
```

These are full prepared literal-allele relations, not annotation-density samples.
They are retained historical diagnostics from before core regulation/motif loading;
they are not the production comparison.
The ClinVar 2026-07-06 source contains 4,439,617 records and 4,438,515 source
ALTs; 4,438,467 literal, model-addressable alleles enter the benchmark
(4,134,694 SNVs and 303,773 non-SNVs). The longest retained REF and ALT are
9,983 and 9,971 bases. The GIAB HG002 GRCh38 v4.2.1 benchmark source contains
4,048,342 records and 4,096,123 source ALTs; 4,095,611 literal,
model-addressable alleles enter the benchmark (3,463,000 SNVs and 632,611
non-SNVs). Its longest retained REF and ALT are 9,704 and 3,686 bases.

Every row uses the complete Ensembl 116 GRCh38 transcript model, a 5,000-base
transcript distance, one ordered input partition, one thread pinned to CPU 2,
a 100,000-allele warmup, and five complete timed passes. Model loading, VCF
staging, and the untimed corpus receipt are excluded. Regulatory and motif
features are not loaded in these rows. The timed query includes candidate
discovery, consequence calculation, list materialization, `unnest`, and the
mode-specific checksum aggregation.

The 126,733,707 ClinVar rows are 28.55 transcript rows per input allele; the
47,629,345 GIAB rows are 11.63 per allele. Consequently, the same HGVS engine
reports 90,594 input alleles/s on ClinVar and 244,222/s on GIAB while sustaining
2.59 and 2.84 million expanded rows/s respectively. At those measured rates,
700 million input alleles project to 2.15 or 0.80 single-core hours; that range
is an input-density sensitivity, not a gnomAD runtime claim. The full ClinVar
HGVS pass also supplied the 1,406-base-ALT regression that found the former
scratch-buffer over-read; all five recorded HGVS passes complete after the
exact-size retry fix.

## Full-corpus core VEP model

```{r full-core-vep-corpora}
core_workload_labels <- c(
  ensembl116_grch38_clinvar_20260706_full_literal_regulation =
    "ClinVar 2026-07-06",
  ensembl116_grch38_giab_hg002_v4_2_1_full_literal_regulation =
    "GIAB HG002 v4.2.1"
)
core_runs <- history[
  history$workload %in% names(core_workload_labels) &
    history$output_mode %in% c("compact", "rich", "hgvs") &
    history$threads == 1L &
    history$input_partitions == 1L &
    history$transcript_distance == 5000,
  ,
  drop = FALSE
]
core_revision <- nearest_tested_ancestor(core_runs$source_revision)
core_current <- core_runs[
  core_runs$source_revision == core_revision,
  ,
  drop = FALSE
]
stopifnot(
  nrow(core_current) == 6L,
  all(core_current$regulation_features == 1383580),
  all(
    core_current$extension_build_binding == "htslib_distclean_make_release"
  ),
  all(nchar(core_current$extension_sha256) == 64L),
  all(nchar(core_current$model_database_sha256) == 64L),
  all(nchar(core_current$model_logical_sha256) == 64L),
  all(nchar(core_current$model_source_manifest_sha256) == 64L),
  all(nchar(core_current$model_reference_sha256) == 64L),
  all(nchar(core_current$model_region_ordinal_sha256) == 64L),
  all(nchar(core_current$variants_database_sha256) == 64L),
  all(nchar(core_current$corpus_source_sha256) == 64L),
  all(core_current$fingerprint_schema_version == 2L),
  all(nzchar(core_current$fingerprint_xor_hash)),
  all(nzchar(core_current$fingerprint_sum_hash)),
  all(core_current$model_source_version == "116"),
  all(core_current$model_assembly == "GRCh38"),
  all(core_current$variants == core_current$corpus_supported_alt_count),
  length(unique(core_current$extension_sha256)) == 1L,
  length(unique(core_current$model_database_sha256)) == 1L,
  length(unique(core_current$model_logical_sha256)) == 1L,
  all(vapply(
    split(core_current$output_mode, core_current$workload),
    function(x) setequal(x, c("compact", "rich", "hgvs")),
    logical(1L)
  ))
)
core_current <- core_current[order(
  match(core_current$workload, names(core_workload_labels)),
  match(core_current$output_mode, mode_order)
), , drop = FALSE]
core_value <- function(workload, output_mode, column) {
  value <- core_current[
    core_current$workload == workload &
      core_current$output_mode == output_mode,
    column
  ]
  stopifnot(length(value) == 1L)
  value[[1L]]
}
format_rate <- function(value) {
  format(round(value), big.mark = ",", scientific = FALSE)
}
clinvar_core <- names(core_workload_labels)[[1L]]
giab_core <- names(core_workload_labels)[[2L]]
core_rates <- c(
  clinvar_compact = format_rate(core_value(
    clinvar_core, "compact", "variants_per_second"
  )),
  clinvar_rich = format_rate(core_value(
    clinvar_core, "rich", "variants_per_second"
  )),
  clinvar_hgvs = format_rate(core_value(
    clinvar_core, "hgvs", "variants_per_second"
  )),
  giab_compact = format_rate(core_value(
    giab_core, "compact", "variants_per_second"
  )),
  giab_rich = format_rate(core_value(
    giab_core, "rich", "variants_per_second"
  )),
  giab_hgvs = format_rate(core_value(
    giab_core, "hgvs", "variants_per_second"
  ))
)
core_hgvs_hours <- c(
  clinvar = 700e6 / core_value(
    clinvar_core, "hgvs", "variants_per_second"
  ) / 3600,
  giab = 700e6 / core_value(
    giab_core, "hgvs", "variants_per_second"
  ) / 3600
)
giab_core_compact_range <- c(
  core_value(giab_core, "compact", "min_seconds"),
  core_value(giab_core, "compact", "max_seconds")
)
core_compact_seconds <- setNames(
  core_current$median_seconds[core_current$output_mode == "compact"],
  core_current$workload[core_current$output_mode == "compact"]
)
core_table <- data.frame(
  revision = core_current$revision,
  corpus = unname(core_workload_labels[core_current$workload]),
  output = core_current$output_mode,
  resident_regulation_motif = format(
    core_current$regulation_features,
    big.mark = ",",
    scientific = FALSE
  ),
  input_alleles = format(
    core_current$variants,
    big.mark = ",",
    scientific = FALSE
  ),
  output_rows = format(
    core_current$annotated_rows,
    big.mark = ",",
    scientific = FALSE
  ),
  rows_per_allele = round(
    core_current$annotated_rows / core_current$variants,
    2L
  ),
  median_seconds = core_current$median_seconds,
  alleles_per_second = format(
    round(core_current$variants_per_second),
    big.mark = ",",
    scientific = FALSE
  ),
  output_rows_per_second = format(
    round(core_current$annotated_rows / core_current$median_seconds),
    big.mark = ",",
    scientific = FALSE
  ),
  elapsed_vs_compact = round(
    core_current$median_seconds /
      unname(core_compact_seconds[core_current$workload]),
    2L
  ),
  projected_700M_allele_hours = round(
    700e6 / core_current$variants_per_second / 3600,
    2L
  ),
  stringsAsFactors = FALSE
)
knitr::kable(core_table, row.names = FALSE)

core_receipts <- data.frame(
  revision = core_current$revision,
  corpus = unname(core_workload_labels[core_current$workload]),
  output = core_current$output_mode,
  extension_build = core_current$extension_build_binding,
  extension_binary = substr(core_current$extension_sha256, 1L, 12L),
  physical_model = substr(core_current$model_database_sha256, 1L, 12L),
  logical_model = substr(core_current$model_logical_sha256, 1L, 12L),
  corpus_source = substr(core_current$corpus_source_sha256, 1L, 12L),
  staged_corpus = substr(core_current$variants_database_sha256, 1L, 12L),
  public_row_xor = core_current$fingerprint_xor_hash,
  public_row_sum = core_current$fingerprint_sum_hash,
  stringsAsFactors = FALSE
)
knitr::kable(core_receipts, row.names = FALSE)
```

This is the complete resident core-VEP model used for independent small
variants: 644,427 transcripts plus 380,818 RegulatoryFeature and 1,002,762
MotifFeature intervals. It is a production core-consequence workload, not an
end-to-end product benchmark: VCF/Parquet ingestion and external exact,
positional, gene, interval, frequency, and clinical-resource joins remain
separate DuckDB stages.

The core model emits 126,930,027 rows for ClinVar and 47,835,851 for GIAB,
including 196,320 and 206,506 regulation/motif rows respectively. The pinned
single-core rates are `r core_rates[["clinvar_compact"]]` compact,
`r core_rates[["clinvar_rich"]]` rich, and
`r core_rates[["clinvar_hgvs"]]` cumulative-HGVS alleles/s for ClinVar; GIAB
reaches `r core_rates[["giab_compact"]]`, `r core_rates[["giab_rich"]]`, and
`r core_rates[["giab_hgvs"]]`/s. At the measured HGVS rates, 700 million input
alleles project to `r sprintf("%.2f", core_hgvs_hours[["clinvar"]])` or
`r sprintf("%.2f", core_hgvs_hours[["giab"]])` single-core hours before
supplementary SQL joins.

Every row in the production-core table is tied to exact source revision
`r core_revision`,
a vendored-htslib `distclean` followed by an in-tree `make release` rebuild and
its extension-binary digest, the
physical and logical model digests,
the Ensembl source-manifest and logical-reference digests, the model-region
ordinal digest, the staged-corpus digest, and the original VCF digest. The HGVS
rows additionally record the physical FASTA and FAI digests. The receipt table
shows shortened artifact digests for readability and the complete schema-version-2
public-row XOR/sum fingerprints; the full SHA-256 values are retained in
`benchmarks/data/duckvep_throughput.csv`.

`_duckvep_annotate_small_hgvs(...)` preserves interval-feature rows from the same
candidate sweep and leaves their transcript/protein HGVS fields NULL. The HGVS
byte checksum is therefore unchanged while output cardinality includes every
regulatory/motif row; SQL and R regressions pin that pass-through. The earlier
transcript-only GIAB run ranged from 3.592 to 6.477 seconds across five passes,
whereas this core-model run ranged from
`r sprintf("%.3f", giab_core_compact_range[[1L]])` to
`r sprintf("%.3f", giab_core_compact_range[[2L]])` seconds. The apparently
faster compact result is not attributed to loading more intervals; use the
stable absolute core-model result rather than treating non-adjacent noisy runs
as an optimization comparison.

## Independent-event HGVS materialization

```{r hgvs-throughput, results='asis'}
hgvs_runs <- history[
  history$output_mode %in% c("rich", "hgvs") &
    history$workload == "ensembl116_grch38_final_coding_mixed_200k" &
    history$threads == 1L,
  ,
  drop = FALSE
]
if (nrow(hgvs_runs)) {
  hgvs_keys <- interaction(
    hgvs_runs$source_revision,
    hgvs_runs$host,
    hgvs_runs$workload,
    hgvs_runs$variants,
    ifelse(
      is.na(hgvs_runs$cpu_affinity),
      "(unrecorded)",
      hgvs_runs$cpu_affinity
    ),
    drop = TRUE
  )
  complete_keys <- names(which(vapply(
    split(hgvs_runs$output_mode, hgvs_keys),
    function(x) all(c("rich", "hgvs") %in% x),
    logical(1L)
  )))
  comparable <- hgvs_runs[hgvs_keys %in% complete_keys, , drop = FALSE]
  if (nrow(comparable)) {
    comparison_revision <- nearest_tested_ancestor(comparable$source_revision)
    comparable <- comparable[
      comparable$source_revision == comparison_revision,
      ,
      drop = FALSE
    ]
    hgvs_table <- comparable[, c(
      "revision", "workload", "output_mode", "threads", "cpu_affinity",
      "variants", "annotated_rows", "median_seconds", "variants_per_second",
      "checksum_kind"
    )]
    comparison_group <- paste(
      comparable$host,
      comparable$workload,
      comparable$variants,
      comparable$threads,
      comparable$cpu_affinity,
      sep = "\034"
    )
    rich_seconds <- setNames(
      comparable$median_seconds[comparable$output_mode == "rich"],
      comparison_group[comparable$output_mode == "rich"]
    )
    hgvs_table$output_rows_per_second <- round(
      comparable$annotated_rows / comparable$median_seconds
    )
    hgvs_table$elapsed_vs_rich <- round(
      comparable$median_seconds / unname(rich_seconds[comparison_group]),
      2L
    )
    hgvs_table$variants <- format(
      hgvs_table$variants,
      big.mark = ",",
      scientific = FALSE
    )
    hgvs_table$annotated_rows <- format(
      hgvs_table$annotated_rows,
      big.mark = ",",
      scientific = FALSE
    )
    hgvs_table$variants_per_second <- round(hgvs_table$variants_per_second)
    hgvs_table$output_rows_per_second <- format(
      hgvs_table$output_rows_per_second,
      big.mark = ",",
      scientific = FALSE
    )
    cat(knitr::kable(hgvs_table, row.names = FALSE), sep = "\n")
    cat("\n\n")

    current_hgvs <- comparable[comparable$output_mode == "hgvs", , drop = FALSE]
    if (nrow(current_hgvs) == 1L) {
      same_workload <- hgvs_runs$output_mode == "hgvs" &
        hgvs_runs$source_revision != comparison_revision &
        hgvs_runs$host == current_hgvs$host &
        hgvs_runs$workload == current_hgvs$workload &
        hgvs_runs$variants == current_hgvs$variants &
        hgvs_runs$threads == current_hgvs$threads &
        hgvs_runs$cpu_affinity == current_hgvs$cpu_affinity
      previous_revision <- nearest_tested_ancestor(
        hgvs_runs$source_revision[same_workload]
      )
      previous_hgvs <- hgvs_runs[
        same_workload & hgvs_runs$source_revision == previous_revision,
        ,
        drop = FALSE
      ]
      if (nrow(previous_hgvs) == 1L) {
        change <- data.frame(
          baseline_revision = previous_hgvs$revision,
          current_revision = current_hgvs$revision,
          input_variants = format(
            current_hgvs$variants,
            big.mark = ",",
            scientific = FALSE
          ),
          output_rows = format(
            current_hgvs$annotated_rows,
            big.mark = ",",
            scientific = FALSE
          ),
          baseline_seconds = previous_hgvs$median_seconds,
          current_seconds = current_hgvs$median_seconds,
          speedup = round(
            previous_hgvs$median_seconds / current_hgvs$median_seconds,
            2L
          ),
          current_input_variants_per_second = round(
            current_hgvs$variants_per_second
          ),
          current_output_rows_per_second = format(
            round(current_hgvs$annotated_rows / current_hgvs$median_seconds),
            big.mark = ",",
            scientific = FALSE
          )
        )
        cat(knitr::kable(change, row.names = FALSE), sep = "\n")
        cat("\n")
      }
    }
  }
}
```

[Ferro-hgvs v0.9.0](https://github.com/fulcrumgenomics/ferro-hgvs/tree/278e2c11134e3b49067d0c334f650c7c29db9cbe)
reports 5.1 million patterns/s for parsing already rendered HGVS and
a 77,000 patterns/s full-population normalization peak on an Apple M2 Max. Those
are useful independent reference points but not the operation timed here:
DuckVEP begins with genomic alleles, discovers every overlapping transcript,
computes consequences, shifts and renders transcript/protein HGVS, and emits
28.78 transcript rows per input in this workload. The table therefore reports
both input alleles/s and expanded transcript-HGVS rows/s; it does not present a
direct cross-machine speed ranking.

The HGVS row materializes transcript/protein strings and status fields and
checks their total byte count. Compare it only with the rich consequence row at
the same source revision, host, workload, variant count, thread count, and CPU
affinity. Model loading, FASTA validation, and input staging remain outside the
timed pass; lazy per-worker faidx setup occurs during warmup.

## Core regulation and motif sweep

```{r regulation-throughput}
regulation_runs <- history[
  history$workload == "ensembl116_grch38_final_giab_sites_hash40_regulation" &
    history$output_mode == "compact" & history$threads == 1L,
  ,
  drop = FALSE
]
if (nrow(regulation_runs)) {
  transcript_controls <- history[
    history$workload == "ensembl116_grch38_final_giab_sites_hash40" &
      history$output_mode == "compact" & history$threads == 1L,
    ,
    drop = FALSE
  ]
  comparison_revision <- nearest_tested_ancestor(intersect(
    regulation_runs$source_revision,
    transcript_controls$source_revision
  ))
  regulation <- regulation_runs[
    regulation_runs$source_revision == comparison_revision,
    ,
    drop = FALSE
  ]
  transcript_control <- transcript_controls[
    transcript_controls$source_revision == comparison_revision,
    ,
    drop = FALSE
  ]
  comparison <- rbind(
    transcript_control[nrow(transcript_control), , drop = FALSE],
    regulation[nrow(regulation), , drop = FALSE]
  )
  if (nrow(comparison) == 2L) {
    regulation_table <- data.frame(
      lane = c("transcripts", "transcripts + regulation/motif"),
      revision = comparison$revision,
      input_variants = format(comparison$variants, big.mark = ","),
      resident_interval_features = format(
        comparison$regulation_features,
        big.mark = ","
      ),
      output_rows = format(comparison$annotated_rows, big.mark = ","),
      median_seconds = comparison$median_seconds,
      variants_per_second = format(
        round(comparison$variants_per_second),
        big.mark = ","
      ),
      output_rows_per_second = format(
        round(comparison$annotated_rows / comparison$median_seconds),
        big.mark = ","
      ),
      stringsAsFactors = FALSE
    )
    knitr::kable(regulation_table, row.names = FALSE)
  }
}
```

This is one pinned-core comparison at one source revision. Both lanes load the
same transcript model and consume the same coordinate-sorted GIAB sample. The
second lane additionally loads every VEP-admitted GRCh38 RegulatoryFeature and
MotifFeature interval and emits their compact consequence rows. The comparison
therefore measures candidate traversal and the extra output together; it is not
an isolated interval-lookup microbenchmark.

## Paired breakends

### Raw ALT preparation and endpoint-union correction

The 2026-09-06 fixture comparison uses source
`60a15f2121c5674fcae7bba6a8ff22d0298988f8` (benchmark-only baseline) and
`9ed772f71376b9ee47fcac03ba3e010fb2bfc973` (native ALT parser and endpoint-union fix).
Each run repeats **1,000,000 input BND events into 1,000,000 compact transcript
rows**, using one DuckDB thread pinned to logical CPU 2 of the same i5-13500,
DuckDB 1.5.3, five timed passes, a 100,000-event warmup, and a 5,000-base
transcript window. Both revisions are clean, in-tree release rebuilds.

```{r breakend-fixture-preparation}
bnd_fixture <- history[
  history$source_revision %in% c(
    "60a15f2121c5674fcae7bba6a8ff22d0298988f8",
    "9ed772f71376b9ee47fcac03ba3e010fb2bfc973"
  ) & grepl("^fixture_one_transcript_sorted_breakend", history$workload),
  , drop = FALSE
]
stopifnot(
  nrow(bnd_fixture) == 3L,
  all(bnd_fixture$variants == 1000000),
  all(bnd_fixture$annotated_rows == 1000000),
  all(bnd_fixture$checksum_value == "268435460000000"),
  length(unique(bnd_fixture$fingerprint_xor_hash)) == 1L,
  length(unique(bnd_fixture$fingerprint_sum_hash)) == 1L
)
knitr::kable(bnd_fixture[, c(
  "revision", "workload", "min_seconds", "median_seconds", "max_seconds"
)], row.names = FALSE, digits = 3)
```

The identical typed-endpoint control changes from **0.267 s to 0.270 s**
(+1.1% median; both observed ranges 0.266–0.272 s). Adding
`--fixture-breakend-alt` measures **0.396 s** (0.394–0.402 s), including all four
raw bracket forms, `duckvep_breakend_geometry()`, the explicit mate-name join,
ordering, annotation, unnesting and aggregation. It is additional preparation
work, not a before/after implementation comparison. The untimed full compact-row
fingerprints match for all three runs; no consequence, status, coordinate or
input-identity field is dropped from those fingerprints.

The model is the checked-in one-transcript, two-exon fixture, with local position
159 and mate position 170. This is a paired-adapter floor, not a measurement of
large raw records, high-fan-out models, fusion reconstruction, or phased work.
It does not measure the newly corrected empty-endpoint predicate state; fixed
native/SQL/R witnesses and executable VEP differentials cover that state. No
prior checked-in workload measured raw ALT preparation. The production BND
baseline below (200,000 events / 18,766,240 transcript rows) has a different model
and fan-out and is not a same-workload comparator for this fixture.

```{r paired-breakend-throughput}
breakend_controls <- history[
  history$workload == "ensembl116_grch38_final_breakend_200k" &
    history$output_mode == "compact" &
    history$threads == 1L,
  ,
  drop = FALSE
]
breakend_regulation <- history[
  history$workload ==
    "ensembl116_grch38_final_breakend_regulation_200k" &
    history$output_mode == "compact" &
    history$threads == 1L,
  ,
  drop = FALSE
]
comparison_revision <- nearest_tested_ancestor(intersect(
  breakend_controls$source_revision,
  breakend_regulation$source_revision
))
pre_literal_revision <- paste0(
  "e7c3623dddc44a21c29d2be93faa7e0e5d8158e5"
)
breakend <- rbind(
  breakend_controls[
    breakend_controls$revision == "360619ed",
    ,
    drop = FALSE
  ],
  breakend_controls[
    breakend_controls$revision == "f97101e1",
    ,
    drop = FALSE
  ],
  breakend_controls[
    breakend_controls$source_revision == pre_literal_revision,
    ,
    drop = FALSE
  ],
  breakend_regulation[
    breakend_regulation$source_revision == pre_literal_revision,
    ,
    drop = FALSE
  ],
  breakend_controls[
    breakend_controls$source_revision == comparison_revision,
    ,
    drop = FALSE
  ],
  breakend_regulation[
    breakend_regulation$source_revision == comparison_revision,
    ,
    drop = FALSE
  ]
)
if (nrow(breakend) == 6L) {
  breakend_pre_literal_median <- breakend$median_seconds[[3L]]
  breakend_pre_literal_regulation_median <- breakend$median_seconds[[4L]]
  breakend_current_median <- breakend$median_seconds[[5L]]
  breakend_regulation_median <- breakend$median_seconds[[6L]]
  breakend_current_change <- 100 * (
    breakend_current_median / breakend_pre_literal_median - 1
  )
  breakend_regulation_change <- 100 * (
    breakend_regulation_median / breakend_pre_literal_regulation_median - 1
  )
  breakend_regulation_overhead <- 100 * (
    breakend_regulation_median / breakend_current_median - 1
  )
  breakend_regulation_rows_per_second <-
    breakend$annotated_rows[[6L]] / breakend_regulation_median
  breakend_table <- data.frame(
    lane = c(
      "transcript baseline",
      "rejected full-result sort",
      "pre-literal transcript A/B",
      "pre-literal transcript + regulation A/B",
      "post-literal transcript A/B",
      "post-literal transcript + regulation A/B"
    ),
    revision = breakend$revision,
    paired_events = format(
      breakend$variants,
      big.mark = ",",
      scientific = FALSE
    ),
    consequence_rows = format(
      breakend$annotated_rows,
      big.mark = ",",
      scientific = FALSE
    ),
    rows_per_event = round(
      breakend$annotated_rows / breakend$variants,
      2
    ),
    median_seconds = breakend$median_seconds,
    paired_events_per_second = format(
      round(breakend$variants_per_second),
      big.mark = ",",
      scientific = FALSE
    ),
    consequence_rows_per_second = format(
      round(breakend$annotated_rows / breakend$median_seconds),
      big.mark = ",",
      scientific = FALSE
    ),
    stringsAsFactors = FALSE
  )
  knitr::kable(breakend_table, row.names = FALSE)
}
```

This measurement used the complete receipt-hashed GRCh38 model, compact output,
one DuckDB thread, 15 timed passes, and a 10,000-event warmup. Model loading and
staging were outside the timed region. The historical baseline and rejected-sort
rows used `taskset -c 2`; the direct pre-literal/post-literal A/B rows were rebuilt
from their exact revisions and run consecutively with `taskset -c 4`. Here
"post-literal" is revision `r substr(comparison_revision, 1L, 8L)`, not the report's
current source; source `e25c151` has not rerun this identical BND workload. The
site-rate headline alone would be misleading: the transcript-only workload emits
93.83 rows per event, and the checksum covers every emitted mask. The
`f97101e1` intermediate sorted all 18.8 million already variant-major transcript
rows and was rejected. The accepted comparison revision instead sorts only the
smaller feature stream and merges the two sorted streams linearly.

The direct transcript-only A/B is
`r sprintf("%.3f", breakend_current_median)` seconds versus
`r sprintf("%.3f", breakend_pre_literal_median)` seconds before the long-literal
predicate fix (`r sprintf("%+.1f%%", breakend_current_change)`). The corresponding
regulation A/B is `r sprintf("%.3f", breakend_regulation_median)` versus
`r sprintf("%.3f", breakend_pre_literal_regulation_median)` seconds
(`r sprintf("%+.1f%%", breakend_regulation_change)`). Loading 1,383,580
regulation/motif intervals and emitting 110,880 additional rows costs
`r sprintf("%+.1f%%", breakend_regulation_overhead)` relative to the post-literal
transcript-only lane and preserves
`r sprintf("%.2f", breakend_regulation_rows_per_second / 1e6)` million output
rows/s.

An earlier run of the comparison revision pinned to CPU 2 reported a 3.386-second median while
its fastest pass still completed in 2.712 seconds. Host inspection found three
long-running unpinned CPU-bound workers, and a clean CPU-4 rerun returned to the
2.6-second range. That contaminated row is not retained in the ledger. The
remaining 1--2% direct A/B difference is reported without a no-regression claim.

## Validated CDS projection cache

The `7dd90ce8` extension source is byte-identical to the immediate pre-cache
parent; the intervening commits changed only documentation and benchmark data.
The `f649d724` rows measure the cache with the same final model, input tables,
single pinned CPU, ordered query, compact output, and warmup size. Positive
change means less elapsed time. The non-SNV and mixed lanes avoid repeated
binary searches for both CDS endpoints. The sorted SNV and ordinary GIAB lanes
show the control cost because they rarely or never consume that cache.

```{r projection-cache}
cache_before <- history[
  history$revision == "7dd90ce8" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
cache_after <- history[
  history$revision == "f649d724" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
cache_table <- merge(
  cache_before,
  cache_after,
  by = c("workload", "variants", "annotated_rows"),
  suffixes = c("_before", "_after")
)
cache_order <- c(
  "ensembl116_grch38_final_coding_mixed_200k",
  "ensembl116_grch38_final_coding_nonsnv_36k",
  "ensembl116_grch38_final_coding_snv_200k",
  "ensembl116_grch38_final_giab_sites_hash40"
)
cache_labels <- c(
  "mixed coding",
  "non-SNV coding",
  "coding SNV",
  "GIAB sites"
)
cache_table <- cache_table[
  match(cache_order, cache_table$workload),
  ,
  drop = FALSE
]
cache_table$workload <- cache_labels
cache_table$elapsed_change_percent <- round(
  100 * (cache_table$median_seconds_before -
    cache_table$median_seconds_after) /
    cache_table$median_seconds_before,
  1
)
cache_table$variants <- format(
  cache_table$variants, big.mark = ",", scientific = FALSE
)
cache_table$annotated_rows <- format(
  cache_table$annotated_rows, big.mark = ",", scientific = FALSE
)
cache_table$variants_per_second_before <- round(
  cache_table$variants_per_second_before
)
cache_table$variants_per_second_after <- round(
  cache_table$variants_per_second_after
)
names(cache_table) <- c(
  "workload", "variants", "output rows", "seconds before", "variants/s before",
  "seconds after", "variants/s after", "elapsed change (%)"
)
knitr::kable(cache_table, row.names = FALSE)
```

## Reused intron coordinates for mismatch islands

The `ecb7eaae` rows are the merged projection-cache source immediately before
the splice-classifier change. The `57b5b566` rows reuse the intron coordinates
already resolved for each exon pair, avoid calling the shared predicate when
neither the intron nor a splice or short-intron window can contribute, and keep
that predicate inlined in the sorted point classifier. Both revisions use the
same final model, input tables, pinned CPU, ordered query, compact output, and
10,000-variant warmup. Positive change means less elapsed time.

```{r splice-intron-reuse}
splice_before <- history[
  history$revision == "ecb7eaae" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
splice_after <- history[
  history$revision == "57b5b566" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
splice_table <- merge(
  splice_before,
  splice_after,
  by = c("workload", "variants", "annotated_rows"),
  suffixes = c("_before", "_after")
)
splice_order <- c(
  "ensembl116_grch38_final_coding_nonsnv_36k",
  "ensembl116_grch38_final_coding_mixed_200k",
  "ensembl116_grch38_final_coding_snv_200k"
)
splice_labels <- c("non-SNV coding", "mixed coding", "coding SNV control")
splice_table <- splice_table[
  match(splice_order, splice_table$workload),
  ,
  drop = FALSE
]
splice_table$workload <- splice_labels
splice_table$elapsed_change_percent <- round(
  100 * (splice_table$median_seconds_before -
    splice_table$median_seconds_after) /
    splice_table$median_seconds_before,
  1
)
splice_table$variants <- format(
  splice_table$variants, big.mark = ",", scientific = FALSE
)
splice_table$annotated_rows <- format(
  splice_table$annotated_rows, big.mark = ",", scientific = FALSE
)
splice_table$variants_per_second_before <- round(
  splice_table$variants_per_second_before
)
splice_table$variants_per_second_after <- round(
  splice_table$variants_per_second_after
)
names(splice_table) <- c(
  "workload", "variants", "output rows", "seconds before", "variants/s before",
  "seconds after", "variants/s after", "elapsed change (%)"
)
knitr::kable(splice_table, row.names = FALSE)
```

The coding-SNV row is a control: that route does not use mismatch-island
classification. Its 5 ms movement is reported as measurement noise rather than
attributed to this change.

## Reused coding projection fact for NMD

The `1903f7fb` rows are the exact merged source before this change. The
`a346a7aa` rows record the first projection-cache optimization: retain the
`CDS end <= 101` comparison already established by the sequence classifier and
consume it when VEP 116 NMD prediction needs the same fact. That revision is a
measured historical intermediate. The later executable-plugin audit proved
that the fact is interchangeable only when the minimized edit and VEP feature
cover exactly the same genomic bases. Both revisions below were rebuilt and
measured adjacently on the same final model, input tables, pinned CPU, ordered
query, compact output, and 10,000-variant warmup. Positive change means less
elapsed time.

```{r nmd-projection-reuse}
nmd_before <- history[
  history$revision == "1903f7fb" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
nmd_after <- history[
  history$revision == "a346a7aa" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "median_seconds",
    "variants_per_second")
]
nmd_table <- merge(
  nmd_before,
  nmd_after,
  by = c("workload", "variants", "annotated_rows"),
  suffixes = c("_before", "_after")
)
nmd_order <- c(
  "ensembl116_grch38_final_coding_mixed_200k",
  "ensembl116_grch38_final_coding_nonsnv_36k",
  "ensembl116_grch38_final_coding_snv_200k"
)
nmd_labels <- c("mixed coding", "non-SNV coding", "coding SNV control")
nmd_table <- nmd_table[
  match(nmd_order, nmd_table$workload),
  ,
  drop = FALSE
]
nmd_table$workload <- nmd_labels
nmd_table$elapsed_change_percent <- round(
  100 * (nmd_table$median_seconds_before - nmd_table$median_seconds_after) /
    nmd_table$median_seconds_before,
  1
)
nmd_table$variants <- format(
  nmd_table$variants, big.mark = ",", scientific = FALSE
)
nmd_table$annotated_rows <- format(
  nmd_table$annotated_rows, big.mark = ",", scientific = FALSE
)
nmd_table$variants_per_second_before <- round(
  nmd_table$variants_per_second_before
)
nmd_table$variants_per_second_after <- round(
  nmd_table$variants_per_second_after
)
names(nmd_table) <- c(
  "workload", "variants", "output rows", "seconds before", "variants/s before",
  "seconds after", "variants/s after", "elapsed change (%)"
)
knitr::kable(nmd_table, row.names = FALSE)
```

The coding-SNV route does not carry the reused projection fact. Its 5 ms
movement is reported as a control rather than attributed to the NMD change.

## Cost of exact VEP feature projection

The `c361346f` correction retains the cached fact only for identical genomic
spans. Wider equal-length features and insertions instead project the same
parent `TranscriptVariation` state read by `NMD.pm`. These rows compare that
exact implementation with the earlier over-broad cache revision. Output row
counts and consequence-mask checksums are identical for every workload.
Negative elapsed change means the exact implementation took longer; the SNV
lane is a control because it does not use this NMD path.

```{r nmd-feature-projection}
nmd_feature_before <- history[
  history$revision == "a346a7aa" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "checksum_value",
    "median_seconds", "variants_per_second")
]
nmd_feature_after <- history[
  history$revision == "c361346f" &
    history$output_mode == "compact" &
    history$threads == 1L,
  c("workload", "variants", "annotated_rows", "checksum_value",
    "median_seconds", "variants_per_second")
]
nmd_feature_table <- merge(
  nmd_feature_before,
  nmd_feature_after,
  by = c("workload", "variants", "annotated_rows", "checksum_value"),
  suffixes = c("_before", "_exact")
)
nmd_feature_order <- c(
  "ensembl116_grch38_final_coding_mixed_200k",
  "ensembl116_grch38_final_coding_nonsnv_36k",
  "ensembl116_grch38_final_coding_snv_200k"
)
nmd_feature_labels <- c(
  "mixed coding", "non-SNV coding", "coding SNV control"
)
nmd_feature_table <- nmd_feature_table[
  match(nmd_feature_order, nmd_feature_table$workload),
  ,
  drop = FALSE
]
nmd_feature_table$workload <- nmd_feature_labels
nmd_feature_table$elapsed_change_percent <- round(
  100 * (
    nmd_feature_table$median_seconds_before -
      nmd_feature_table$median_seconds_exact
  ) / nmd_feature_table$median_seconds_before,
  1
)
nmd_feature_table$variants <- format(
  nmd_feature_table$variants, big.mark = ",", scientific = FALSE
)
nmd_feature_table$annotated_rows <- format(
  nmd_feature_table$annotated_rows, big.mark = ",", scientific = FALSE
)
nmd_feature_table$variants_per_second_before <- round(
  nmd_feature_table$variants_per_second_before
)
nmd_feature_table$variants_per_second_exact <- round(
  nmd_feature_table$variants_per_second_exact
)
nmd_feature_table$checksum_value <- NULL
names(nmd_feature_table) <- c(
  "workload", "variants", "output rows", "seconds before", "variants/s before",
  "seconds exact", "variants/s exact", "elapsed change (%)"
)
knitr::kable(nmd_feature_table, row.names = FALSE)
```

## Last measured transcript-only GRCh38 model resources

```{r resources}
last_resource <- resources[nrow(resources), , drop = FALSE]
resource_table <- data.frame(
  measure = c(
    "revision", "model artifact", "model load", "process RSS before load",
    "process RSS after load", "peak process RSS", "process RSS after drop",
    "DuckDB base-table cache", "planned scratch per worker",
    "observed RSS change on first workspace", "largest transcript run",
    "largest prepared CDS"
  ),
  value = c(
    last_resource$revision,
    sprintf("%.3f GiB", last_resource$model_artifact_bytes / 1024^3),
    sprintf("%.3f s", last_resource$model_load_seconds),
    sprintf("%.3f GiB", last_resource$rss_baseline_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$rss_model_loaded_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$rss_peak_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$rss_after_drop_kib / 1024^2),
    sprintf("%.3f GiB", last_resource$duckdb_base_table_cache_bytes / 1024^3),
    sprintf("%.3f MiB", last_resource$workspace_planned_buffer_bytes / 1024^2),
    sprintf("%.3f MiB", last_resource$workspace_rss_delta_kib / 1024),
    format(last_resource$max_transcripts_per_region, big.mark = ","),
    sprintf("%s bases", format(last_resource$max_cds_bytes, big.mark = ","))
  ),
  stringsAsFactors = FALSE
)
knitr::kable(resource_table, row.names = FALSE)
```

This resource row predates the resident regulation/motif relation and remains a
transcript-only baseline; the throughput table above records the feature count
for regulation-enabled runs. The artifact size is a compact DuckDB containing
only `model_regions`, `model_transcripts`, and `model_receipt`; benchmark inputs
are excluded. RSS is
for one R/DuckDB process, not a C-heap estimate. The load peak includes both the
DuckDB scan/cache and construction of the second immutable C representation.
Dropping the model releases the C representation, but DuckDB retains its table
cache in the process. The planned per-worker number is the exact sum of the C
buffer capacities (active and candidate transcript indices, one exon cursor per
transcript, and worst-case edit/CDS/peptide scratch); allocator metadata is not
included. The much smaller first-workspace RSS change reflects lazy or reused
pages, not a smaller capacity.

The historical standalone model ABI 0.12 cache adds three `uint32_t` arrays and one
`uint8_t` array per transcript. Its exact element payload for this model is
`r format(13 * last_resource$transcripts[[1L]], big.mark = ",", scientific = FALSE)`
bytes (`r sprintf("%.3f MiB", 13 * last_resource$transcripts[[1L]] / 1024^2)`),
plus four allocator records. This deterministic allocation is not retroactively
included in the older `b204dd49` RSS measurement above.

The production form is run with an externally pinned process when a stable CPU
comparison is required:

```sh
taskset -c 2 Rscript benchmarks/duckvep_throughput.R \
  --database /path/to/model.duckdb \
  --variants-database /path/to/staged-corpus.duckdb \
  --variants-table bench_variants \
  --corpus-source /path/to/source.vcf.gz \
  --workload-name ensembl116_grch38_full_corpus_regulation \
  --regulatory --variants 4095611 --passes 5 --warmup 100000 \
  --threads 1 --input-partitions 1 --transcript-distance 5000 \
  --output compact --fingerprint /tmp/full-public-row-fingerprint.csv
```

For cumulative HGVS, add `--reference-fasta` and change `--output` to `hgvs`.
The source revision must be the current checkout; the history row records the
extension binary, physical/logical model, model source, reference, staged
corpus, original source, and public-row fingerprint receipts.

## Real whole-genome composition run

The resident-engine tables above isolate comparable annotation surfaces. The
following separate run verifies the complete relational composition shown in
the root README on real human data. It scans the public HG002 40x PCR-free
DeepVariant GRCh38 WGS callset sequentially, retains literal alleles on
chromosomes 1--22, X, Y, and MT, loads the complete Ensembl 116 GRCh38
transcript plus regulation/motif model, requests rich consequence and
HGVSc/HGVSp, and joins dated ClinVar, ClinvArbitration, AlphaMissense, gnomAD
v2.1.1 gene constraint, and an Ensembl regulatory interval Parquet relation.
The final relation is written as ZSTD Parquet rather than reduced to a scalar.

```{r human-composition}
human_composition <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_human_annotation.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character")
)
stopifnot(
  nrow(human_composition) == 13L,
  length(unique(human_composition$input_alleles)) == 1L,
  length(unique(human_composition$threads)) == 1L
)
selected_stages <- c(
  "model_load",
  "vcf_decode_canonicalize_sort",
  "ClinVar_exact_collision_safe",
  "ClinvArbitration_exact_collision_safe",
  "AlphaMissense_exact",
  "Ensembl_regulatory_RegionKey_IEJoin_stream",
  "Ensembl_regulatory_cgranges_bulk_stream",
  "DuckVEP_rich_HGVS_regulatory_parallel_stream_4GB",
  "compose_all_providers_parallel_stream_4GB"
)
human_table <- human_composition[
  match(selected_stages, human_composition$stage),
  c("stage", "result_rows", "seconds", "threads", "max_rss_kb")
]
stopifnot(!anyNA(human_table$stage))
human_table$stage <- c(
  "load Ensembl 116 model",
  "read VCF, canonicalize and sort",
  "ClinVar exact join",
  "ClinvArbitration exact join",
  "AlphaMissense exact join",
  "regulatory RegionKey IEJoin + write",
  "regulatory cgranges build/query + write",
  "rich + HGVS + core regulation, four writers",
  "rich + HGVS + all providers, four writers"
)
human_table$result_rows <- format(
  human_table$result_rows, big.mark = ",", scientific = FALSE
)
human_table$max_rss_kb <- ifelse(
  is.na(human_table$max_rss_kb),
  "",
  sprintf("%.2f", human_table$max_rss_kb / 1024^2)
)
names(human_table) <- c(
  "stage", "result rows", "seconds", "threads", "peak RSS (GiB)"
)
knitr::kable(human_table, row.names = FALSE)
```

These are warm-page-cache, unpinned integration measurements, not replacements
for the pinned resident-engine comparisons above. DuckDB CLI `.timer on` records
the statements and GNU `/usr/bin/time -v` records process high-water RSS. The
corrected production query sets a 4 GB DuckDB memory limit and writes one Parquet
file per worker so the complete consequence relation is never retained or
globally sorted.

```{r human-interval-validation}
interval_validation <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_human_interval_validation.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = c(source_revision = "character")
)
stopifnot(
  nrow(interval_validation) == 3L,
  length(unique(interval_validation$matched_events)) == 1L,
  length(unique(interval_validation$overlap_pairs)) == 1L,
  length(unique(interval_validation$stable_ids)) == 1L,
  length(unique(interval_validation$so_terms)) == 1L,
  all(interval_validation$mismatched_event_rows == 0L)
)
interval_table <- interval_validation[, c(
  "method", "matched_events", "overlap_pairs", "seconds", "max_rss_kb"
)]
interval_table$method <- c(
  "chromosome hash join + range residuals",
  "two packed RegionKey inequalities (IEJoin)",
  "cgranges bulk index + query"
)
interval_table$matched_events <- format(
  interval_table$matched_events, big.mark = ",", scientific = FALSE
)
interval_table$overlap_pairs <- format(
  interval_table$overlap_pairs, big.mark = ",", scientific = FALSE
)
interval_table$max_rss_kb <- ifelse(
  is.na(interval_table$max_rss_kb),
  "",
  sprintf("%.0f", interval_table$max_rss_kb / 1024)
)
names(interval_table) <- c(
  "method", "matched alleles", "overlap pairs", "seconds", "peak RSS (MiB)"
)
knitr::kable(interval_table, row.names = FALSE)
```

The first query was mislabeled as IEJoin in the initial receipt. `EXPLAIN`
showed a chromosome `HASH_JOIN` with all four range predicates applied to the
same-chromosome candidate products. The receipt remains above rather than being
hidden. Removing the redundant string equality lets the two packed RegionKey
inequalities encode both chromosome and half-open overlap and selects
`IE_JOIN`. It returns the same
`r format(unique(interval_validation$overlap_pairs), big.mark = ",")` pairs and
`r format(unique(interval_validation$matched_events), big.mark = ",")` matched
alleles in
`r sprintf("%.2fx", interval_validation$seconds[1] / interval_validation$seconds[2])`
less wall time. The cgranges path is slightly slower here but uses roughly half
the process memory and supports arbitrary literal contig names.

The bounded four-writer composition emits
`r format(human_composition$result_rows[human_composition$stage == "compose_all_providers_parallel_stream_4GB"], big.mark = ",")`
rows in
`r sprintf("%.3f", human_composition$seconds[human_composition$stage == "compose_all_providers_parallel_stream_4GB"])`
seconds and writes
`r format(human_composition$output_bytes[human_composition$stage == "compose_all_providers_parallel_stream_4GB"], big.mark = ",")`
bytes. Peak RSS is
`r sprintf("%.2f GiB", human_composition$max_rss_kb[human_composition$stage == "compose_all_providers_parallel_stream_4GB"] / 1024^2)`;
the high-water mark includes transient construction of the complete immutable
Ensembl model, whose C allocations are outside DuckDB's 4 GB buffer-manager
limit.

```{r human-composition-fingerprint, include=FALSE}
human_fingerprint <- read.csv(
  file.path(root, "benchmarks", "data", "duckvep_human_annotation_fingerprint.csv"),
  stringsAsFactors = FALSE,
  colClasses = "character"
)
stopifnot(
  nrow(human_fingerprint) == 2L,
  length(unique(human_fingerprint$rows)) == 1L,
  length(unique(human_fingerprint$xor64)) == 1L,
  length(unique(human_fingerprint$sum64)) == 1L
)
```

The earlier retained-intermediate/single-file run is also preserved in the CSV:
it took
`r sprintf("%.3f", human_composition$seconds[human_composition$stage == "whole_process_warm_page_cache"])`
seconds and peaked at
`r sprintf("%.2f GiB", human_composition$max_rss_kb[human_composition$stage == "whole_process_warm_page_cache"] / 1024^2)`.
The bounded output has the same row count and the same order-independent XOR and
sum fingerprints over every projected column. Controlled resident and FastVEP
comparisons remain the authorities above.
