---
title: "Phased replay: native stream and public SQL"
output: github_document
---

<!-- duckvep_haplotypes.md is generated from duckvep_haplotypes.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
history <- read.csv(file.path(root, "benchmarks/data/duckvep_haplotypes.csv"),
  colClasses = c(source_revision = "character", xor_hash = "character", sum_hash = "character",
    replay_xor_hash = "character", replay_sum_hash = "character"))
stopifnot(all(history$extension_build_binding == "htslib_distclean_make_release"),
  !anyDuplicated(history[c("source_revision", "transcripts", "samples", "overlap", "mode", "pass")]))
if (!"output_contract" %in% names(history)) history$output_contract <- NA_character_
history$output_contract[is.na(history$output_contract)] <- "literal_replay"
for (field in c("json_bytes", "xor_hash", "sum_hash")) {
  projection <- paste0("replay_", field)
  if (!projection %in% names(history)) history[[projection]] <- history[[field]]
  missing <- is.na(history[[projection]])
  history[[projection]][missing] <- history[[field]][missing]
}
revisions <- unique(history$source_revision)
stopifnot(length(revisions) >= 2L)
revision <- tail(revisions, 1L)
baseline_revision <- tail(revisions, 2L)[1L]
keys <- c("transcripts", "samples", "overlap", "mode")
groups <- split(history, interaction(history[c("source_revision", keys)], drop = TRUE))
summary <- do.call(rbind, lapply(groups, function(x) {
  stopifnot(nrow(x) == 3L, all(sort(x$pass) == 1:3))
  data.frame(x[1L, c("source_revision", keys)], min_s = min(x$seconds), median_s = median(x$seconds),
    max_s = max(x$seconds), max_process_rss_mib = max(x$peak_process_rss_kib) / 1024)
}))
comparison <- merge(
  subset(summary, source_revision == baseline_revision & mode == "sql"),
  subset(summary, source_revision == revision & mode == "sql"),
  by = keys, suffixes = c("_before", "_after"))
comparison$median_change_percent <- 100 * (comparison$median_s_after / comparison$median_s_before - 1)
comparison <- comparison[do.call(order, comparison[keys]), ]
current <- history[history$source_revision == revision, ]
summary <- summary[summary$source_revision == revision, setdiff(names(summary), "source_revision")]
summary <- summary[do.call(order, summary[keys]), ]
core <- current[current$mode == "native" & current$pass == 1L, ]
core <- core[order(core$transcripts, core$samples, core$overlap), ]
public <- current[current$mode == "sql" & current$pass == 1L, ]
public <- public[order(public$transcripts, public$samples, public$overlap), ]
stopifnot(identical(as.list(core[c("transcripts", "samples", "overlap")]),
  as.list(public[c("transcripts", "samples", "overlap")])))
fingerprint_keys <- c("transcripts", "samples", "overlap")
fingerprints <- unique(history[history$mode == "sql", c(fingerprint_keys, "output_contract",
  "output_leaves", "output_carriers", "cds_bytes", "protein_bytes", "physical_edits",
  "contributors", "blocks", "json_bytes", "xor_hash", "sum_hash")])
replay_fingerprints <- unique(history[history$mode == "sql", c(fingerprint_keys,
  "replay_json_bytes", "replay_xor_hash", "replay_sum_hash")])
stopifnot(nrow(comparison) == nrow(public),
  !anyDuplicated(fingerprints[c(fingerprint_keys, "output_contract")]),
  !anyDuplicated(replay_fingerprints[fingerprint_keys]))
```

Current source: `r revision`; identical-workload baseline: `r baseline_revision`.
Native measurements cover literal phased replay. SQL materializes all current
fields, including local coding-block SO; whole-haplotype SO/HGVS is unfinished.
Both paths consume standalone ALT events and decoded calls; native raw-record
parsing and conditional replay are not timed by this workload.
All runs use one thread pinned to CPU 2 on an Intel Core i5-13500, DuckDB 1.5.3,
and a source-bound clean extension build. The native bridge uses the same kernel
sources compiled with `-O3 -DNDEBUG`; receipts retain compiler/version, binary and
fixture hashes, worker jobs, GNU-time logs and complete result fingerprints.

## Workload and verification

The [registry](../r/duckhtsbench/inst/benchmark_registry.tsv) stages two committed,
checksum-verified fixtures without network access: a 180-base CDS and four ALT
events. Two substitutions share a codon; an insertion/deletion pair displaces
and restores the frame. Four diploid sample templates repeat across the cohort,
producing exactly three occupied paths per transcript. Strict phase uses PS=10.
Transcript groups share genomic intervals, with non-overlapping groups 1,000
bases apart; `overlap` is the number of simultaneously active transcripts.

This deliberately favourable prefix-sharing workload is not a real population,
a long-transcript benchmark, a reader benchmark, or evidence for arbitrary phase
and structural configurations. Native input is a factorized, already-ordered
feed; SQL input is its complete event/candidate/sample relation stored in hash
order. Both retain every source identity and called carrier.

Before timing, native validation checks every transcript, complete CDS/protein
against independently rebuilt/Biostrings-translated sequences, contributors and
sample/lane/PS/ploidy membership. SQL validation compares the full carrier
multiset with decoded input, checks sequences and edit provenance, and rejects an
injected duplicate. Native validation also rejects a corrupted expected sequence.
Every measured pass must retain all common native/SQL denominators and SQL's full,
canonicalized nested-row fingerprint; failed runs are not promoted. Full fingerprints
must also agree across revisions sharing an output contract. The literal-replay
projection provides a separate comparison with the narrower recorded output
contract; full current output is always checked.

```{r denominators}
knitr::kable(core[c("transcripts", "samples", "overlap", "input_records",
  "projected_events", "input_calls", "output_leaves", "output_carriers",
  "prefixes_created", "translated_bases")], row.names = FALSE)
```

`input_records` counts distinct source ALT events, not candidate/sample rows.
`projected_events` counts event/transcript pairs. `output_carriers` counts
memberships of sample/lane keys in completed leaves; `output_leaves` counts
unique occupied prefixes, not unique peptide strings. Prefix creation and active
pool peaks below are observed in the native stream, not inferred SQL counters.

## Timing and process memory

```{r timing}
knitr::kable(summary, row.names = FALSE, digits = 3)
```

Both compared revisions return each block's local SO mask, coding status and
position relative to the first stop. A shared coding context evaluates physical
blocks on completed replay. Every full-output fingerprint and input/output
denominator matches across the six configurations. The native count sink omits
local SO evaluation.

```{r sql-comparison}
knitr::kable(comparison[c("transcripts", "samples", "overlap", "median_s_before",
  "median_s_after", "median_change_percent", "max_process_rss_mib_before",
  "max_process_rss_mib_after")], row.names = FALSE, digits = 3)
```

Each recorded pass uses a fresh process and a full warm-up. Native timing starts
after workspace initialization and includes ordered-feed construction, GT routing,
projection, replay/translation, block construction and a count/length sink. It
does **not** include alignment, DuckDB, or nested output materialization.

SQL timing covers `CREATE OR REPLACE TABLE ... AS SELECT * FROM duckvep_haplotypes(...)`:
preparation, phase-domain discovery, mandatory sort, workspace initialization,
replay, both sequence alignments, local coding-block SO and **all** nested output
columns. Input/model
staging is outside timing. The warm-up output table is replaced by the measured
table. JSON sizing and fingerprint aggregation occur afterward, not in the timer;
no file-output throughput is claimed.

GNU `/usr/bin/time -v` measures whole-process maximum RSS, including the R host,
model/input tables, warm-up, measured materialization and post-run aggregates.
The heavier SQL correctness audit runs separately and is excluded from recorded
RSS. RSS is not an isolated operator allocation measurement. These two execution
scopes are intentionally different: their times are not a speedup comparison.

```{r output-size}
knitr::kable(public[c("transcripts", "samples", "overlap", "output_leaves",
  "cds_bytes", "protein_bytes", "json_bytes")], row.names = FALSE)
```

CDS/protein bytes count the sequence views returned once per occupied leaf.
JSON bytes are the measured UTF-8 size of complete canonicalized SQL rows,
including carriers, contributors, blocks and differences; they are not bytes
written to disk.

## Sparse native state

```{r state}
knitr::kable(core[c("transcripts", "samples", "overlap", "peak_transcripts",
  "peak_carriers", "peak_prefixes", "peak_events", "peak_projections",
  "peak_allele_bytes", "workspace_bytes", "model_bytes")], row.names = FALSE)
```

Native workspace bytes count actual preallocated buffer capacities and state,
separately from immutable model allocations. The fixed fixture is borrowed;
validation additionally owns `transcripts + 2 * samples` bytes outside the timed
kernel workspace. Peak pool occupancy is measured by the production stream.
Increasing total transcripts at fixed cohort size and overlap does not increase
that workspace; immutable model storage still grows with transcript count.

Both paths use the same active-pool limits. SQL additionally allows 65,536
alignment cells and 64 differing runs per axis, with a 64 MiB native workspace
ceiling and a 4 GB DuckDB memory limit. SQL input, sorting and output memory belong
to DuckDB and may grow with the complete relation. The process RSS table must not
be presented as a constant-total-memory guarantee.

This matched comparison retains three repeated passes per workload and revision.
Sanitizer and conformance jobs did not overlap either benchmark campaign. Both
ran at different times on a shared machine; CPU pinning does not isolate the
cause of timing differences between campaigns.
One machine and this deliberately shared synthetic cohort do not establish
production throughput, statistical significance or a general absence of regression.
Whole-haplotype SO/HGVS and typed structural composition require their own measurements.

## Reproduction

Run [the R driver](duckvep_haplotypes.R) with a clean source-bound extension receipt
from the shared evidence builder; it stages inputs through `duckhtsbench`.
All six configurations above use `--passes 3 --cpu 2`. To prepare the receipt
from a clean checkout (no network access in the build):

```{r clean-build, eval=FALSE}
source("scripts/duckvep_evidence.R")
root <- normalizePath(".")
revision <- duckvep_evidence_revision(root)
extension <- duckvep_evidence_build_extension(root,
  "build/release/duckhts.duckdb_extension", revision)
duckvep_evidence_write_extension_receipt("/tmp/haplotype-extension.tsv", revision, extension)
```

Pass `--extension-receipt /tmp/haplotype-extension.tsv` to the R driver. For a
network-free smoke test without a clean-build certificate:

```{bash smoke, eval=FALSE}
Rscript benchmarks/duckvep_haplotypes.R \
  --transcripts 16 --samples 4 --overlap 4 --passes 1 --diagnostic
```

Diagnostic results are retained but are not accepted into the recorded baseline.
