---
title: "Indel checkpoint: replay benchmark with unequal outputs"
output: github_document
---

<!-- duckvep_haplotypes_indel.md is generated from duckvep_haplotypes_indel.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", getwd()), mustWork = TRUE)
current <- read.csv(file.path(root, "benchmarks/data/duckvep_haplotypes_indel.csv"),
  colClasses = "character")
history <- read.csv(file.path(root, "benchmarks/data/duckvep_haplotypes.csv"),
  colClasses = "character")
revision <- unique(current$source_revision)
baseline_revision <- "e3ec6d231cc7e5c769685761a4739c8fcffc30ff"
baseline <- history[history$source_revision == baseline_revision, ]
keys <- c("transcripts", "samples", "overlap", "mode", "pass")
stopifnot(identical(revision, "20efcf2af33b38c5be7596e2db7b243a3be53c47"),
  identical(names(current), names(baseline)), nrow(current) == 51L, nrow(baseline) == 51L,
  !anyDuplicated(current[keys]), !anyDuplicated(baseline[keys]),
  all(current$extension_build_binding == "htslib_distclean_make_release"),
  all(current$threads == "1"), all(current$cpu_affinity == "2"),
  all(current$duckdb_version[current$mode != "native"] == "1.5.3"),
  all(is.finite(as.numeric(current$seconds))), all(as.numeric(current$seconds) > 0),
  all(is.finite(as.numeric(current$peak_process_rss_kib))),
  all(as.numeric(current$peak_process_rss_kib) > 0))
current <- current[do.call(order, current[keys]), ]
baseline <- baseline[do.call(order, baseline[keys]), ]
rownames(current) <- rownames(baseline) <- NULL
stopifnot(identical(current[keys], baseline[keys]))
fingerprints <- grep("(^|_)(xor|sum)_hash$", names(current), value = TRUE)
measurement_fields <- c("source_revision", "seconds", "peak_process_rss_kib",
  "artifact_directory", "receipt_sha256", "extension_sha256", "native_binary_sha256")
invariants <- setdiff(names(current), c(measurement_fields, fingerprints))
stopifnot(identical(current[invariants], baseline[invariants]))
equal <- function(a, b) (is.na(a) & is.na(b)) | (!is.na(a) & !is.na(b) & a == b)
differences <- do.call(rbind, lapply(fingerprints, function(field) {
  at <- which(!equal(current[[field]], baseline[[field]]))
  data.frame(current[at, keys], field = rep(field, length(at)),
    baseline_value = baseline[[field]][at], current_value = current[[field]][at])
}))
groups <- split(seq_len(nrow(current)), interaction(current[keys[-5L]], drop = TRUE))
summary <- do.call(rbind, lapply(groups, function(at) {
  stopifnot(identical(sort(current$pass[at]), c("1", "2", "3")))
  for (field in setdiff(names(current), c("pass", "seconds", "peak_process_rss_kib"))) {
    stopifnot(length(unique(current[[field]][at])) == 1L,
      length(unique(baseline[[field]][at])) == 1L)
  }
  data.frame(current[at[1L], keys[-5L]],
    baseline_median_s = median(as.numeric(baseline$seconds[at])),
    current_median_s = median(as.numeric(current$seconds[at])),
    current_max_rss_mib = max(as.numeric(current$peak_process_rss_kib[at])) / 1024,
    full_output_equal = if (current$mode[at[1L]] == "native") NA else
      all(equal(current$xor_hash[at], baseline$xor_hash[at])) &&
      all(equal(current$sum_hash[at], baseline$sum_hash[at])),
    replay_equal = if (current$mode[at[1L]] == "native") NA else
      all(equal(current$replay_xor_hash[at], baseline$replay_xor_hash[at])) &&
      all(equal(current$replay_sum_hash[at], baseline$replay_sum_hash[at])))
}))
summary <- summary[do.call(order, summary[keys[-5L]]), ]
witness_path <- file.path(root, "benchmarks/data/duckvep_haplotypes_indel_witness.jsonl.gz")
stopifnot(digest::digest(file = witness_path, algo = "sha256") ==
  "753dfd26db78bd02b9662ed857840c64e2742f71d4e34f1dd74623cbdf52e1a7")
witness_input <- gzfile(witness_path, "rt")
witness <- lapply(readLines(witness_input), jsonlite::fromJSON)
close(witness_input)
witness_keys <- vapply(witness, function(x) paste(x$origin, x$name, sep = "/"), "")
stopifnot(length(witness) == 38L, !anyDuplicated(witness_keys))
for (record in witness) stopifnot(digest::digest(jsonlite::base64_dec(record$bytes_base64),
  algo = "sha256", serialize = FALSE) == record$sha256)
read_witness <- function(key) {
  at <- which(witness_keys == key)
  stopifnot(length(at) == 1L)
  strsplit(rawToChar(jsonlite::base64_dec(witness[[at]]$bytes_base64)), "\n", fixed = TRUE)[[1L]]
}
before_rows <- lapply(read_witness("native/baseline.jsonl"), jsonlite::fromJSON, simplifyVector = FALSE)
after_rows <- lapply(read_witness("native/current.jsonl"), jsonlite::fromJSON, simplifyVector = FALSE)
stopifnot(length(before_rows) == 4L, length(after_rows) == 4L,
  before_rows[[4L]]$contributors[[1L]]$event_index == 4L,
  before_rows[[4L]]$coding_blocks[[1L]]$local_consequence_mask == 8388608L,
  after_rows[[4L]]$coding_blocks[[1L]]$local_consequence_mask == 8396800L)
before_rows[[4L]]$coding_blocks[[1L]]$local_consequence_mask <- 8396800L
stopifnot(identical(before_rows, after_rows))
oracle <- jsonlite::fromJSON(read_witness("oracle/tva.stdout"))
stopifnot(oracle$source$position1 == 67L, oracle$source$reference == "GC",
  oracle$source$alternate == "G", oracle$ref_codon == "GCT", oracle$alt_codon == "GT",
  oracle$ref_peptide == "A", oracle$alt_peptide == "X",
  oracle$raw_predicates$frameshift == 1L, oracle$raw_predicates$missense_variant == 1L,
  length(oracle$raw_predicates) == 13L, sum(unlist(oracle$raw_predicates)) == 2L,
  identical(oracle$consequences, "frameshift_variant"),
  identical(oracle$hgvsp, "bench_protein.1:p.Ala23ValfsTer?"))

capture_input_paths <- function(receipt) {
  hashes <- receipt$input_sha256
  stopifnot(is.list(hashes), !is.null(names(hashes)), !anyDuplicated(names(hashes)))
  driver <- names(hashes)[endsWith(names(hashes), "/benchmarks/duckvep_haplotypes.R")]
  stopifnot(length(driver) == 1L, is.character(receipt$original_job),
    length(receipt$original_job) == 1L, !is.na(receipt$original_job))
  job_directory <- dirname(receipt$original_job)
  stopifnot(dirname(job_directory) == file.path(dirname(dirname(driver)),
    "test/duckvep/conformance/results"))
  c(driver = driver, reference = file.path(job_directory, "reference.fa"),
    index = file.path(job_directory, "reference.fa.fai"))
}

recorded_hash <- function(hashes, path) {
  stopifnot(is.list(hashes), !is.null(names(hashes)), !anyDuplicated(names(hashes)),
    is.character(path), length(path) == 1L, !is.na(path))
  value <- hashes[[path]]
  stopifnot(is.character(value), length(value) == 1L, !is.na(value),
    grepl("^[0-9a-f]{64}$", value))
  value
}

check_original_job <- function(job, hash, pass, capture, original, ledger, fixture) {
  configuration <- capture$configuration
  directory <- dirname(capture$original_job)
  stem <- paste0(configuration$mode, "_", pass)
  path <- file.path(directory, paste0(stem, "_job.rds"))
  stopifnot(hash == recorded_hash(original$sha256, path))
  if (pass == 0L) stopifnot(hash == recorded_hash(capture$input_sha256, path))
  stopifnot(identical(names(job), c("transcripts", "samples", "overlap", "mode", "verify",
      "input", "extension", "shared", "reference_fasta", "reference_length", "result")),
    identical(job[names(configuration)], configuration), identical(job$verify, pass == 0L),
    identical(job$reference_length, 99L + 1000L * job$transcripts %/% job$overlap),
    identical(job$input$reference, fixture$reference), identical(job$input$events, fixture$events),
    identical(job$reference_fasta, file.path(directory, "reference.fa")),
    identical(job$shared, file.path(directory, "stream.so")),
    identical(job$result, file.path(directory, paste0(stem, ".rds"))),
    all(ledger$extension_sha256 == recorded_hash(original$sha256, job$extension)),
    all(ledger$native_binary_sha256 == recorded_hash(original$sha256, job$shared)))
  job[setdiff(names(job), c("extension", "shared", "reference_fasta", "result"))]
}

capture_original_jobs <- function(capture, ledger, directory) {
  sha256 <- function(path) digest::digest(file = path, algo = "sha256")
  job_directory <- dirname(capture$original_job)
  retained <- file.path(directory, "original", basename(job_directory))
  receipt_path <- file.path(retained, "receipt.json")
  hash <- sha256(receipt_path)
  stopifnot(hash == recorded_hash(capture, "benchmark_receipt_sha256"),
    all(ledger$receipt_sha256 == hash),
    hash == recorded_hash(capture$input_sha256, file.path(job_directory, "receipt.json")))
  original <- jsonlite::fromJSON(receipt_path)
  stopifnot(original$source_revision == capture$benchmark_source_revision)
  fixture_paths <- file.path(directory, "fixtures", c("haplotype_benchmark.fa", "haplotype_benchmark.tsv"))
  for (i in seq_along(fixture_paths)) {
    suffix <- c("/duckvep-haplotypes/reference.fa", "/duckvep-haplotypes/events.tsv")[i]
    path <- names(original$sha256)[endsWith(names(original$sha256), suffix)]
    field <- c("reference_sha256", "events_sha256")[i]
    stopifnot(length(path) == 1L, sha256(fixture_paths[i]) == recorded_hash(original$sha256, path),
      all(ledger[[field]] == sha256(fixture_paths[i])))
  }
  fasta <- readLines(fixture_paths[1L])
  stopifnot(length(fasta) == 2L, nchar(fasta[2L]) == 180L)
  fixture <- list(reference = fasta[2L], events = read.delim(fixture_paths[2L]))
  for (name in c("reference.fa", "reference.fa.fai")) {
    path <- file.path(job_directory, name)
    stopifnot(sha256(file.path(retained, name)) == recorded_hash(original$sha256, path),
      sha256(file.path(retained, name)) == recorded_hash(capture$input_sha256, path))
  }
  stopifnot(identical(readLines(file.path(retained, "reference.fa")), c(">bench",
    paste0(strrep("A", 99L), strrep(paste0(fixture$reference, strrep("A", 820L)),
      capture$configuration$transcripts / capture$configuration$overlap)))))
  rejected <- function(expression) tryCatch({ force(expression); FALSE }, error = function(e) TRUE)
  jobs <- lapply(0:3, function(pass) {
    path <- file.path(job_directory, paste0(capture$configuration$mode, "_", pass, "_job.rds"))
    retained_path <- file.path(retained, basename(path))
    job <- readRDS(retained_path)
    hash <- sha256(retained_path)
    semantic <- check_original_job(job, hash, pass, capture, original, ledger, fixture)
    for (value in list(strrep("0", 64L), NULL, "malformed")) {
      changed <- original
      changed$sha256[[path]] <- value
      stopifnot(rejected(check_original_job(job, hash, pass, capture, changed, ledger, fixture)))
      if (pass == 0L) {
        changed <- capture
        changed$input_sha256[[path]] <- value
        stopifnot(rejected(check_original_job(job, hash, pass, changed, original, ledger, fixture)))
      }
    }
    if (pass == 0L) for (field in c("events", "samples")) {
      changed <- job
      if (field == "events") changed$input$events$alternate[1L] <- "T" else changed$samples <- 65L
      changed_hash <- digest::digest(serialize(changed, NULL), algo = "sha256", serialize = FALSE)
      changed_original <- original
      changed_capture <- capture
      changed_original$sha256[[path]] <- changed_hash
      changed_capture$input_sha256[[path]] <- changed_hash
      stopifnot(rejected(check_original_job(changed, changed_hash, pass,
        changed_capture, changed_original, ledger, fixture)))
    }
    semantic
  })
  for (pass in 1:3) {
    compared <- jobs[[pass + 1L]]
    compared$verify <- TRUE
    stopifnot(identical(jobs[[1L]], compared))
  }
  jobs
}

compare_capture_inputs <- function(before, after, mode, transcripts) {
  identities <- lapply(list(before, after), function(receipt) {
    stopifnot(identical(receipt$configuration,
      list(mode = mode, transcripts = transcripts, samples = 64L, overlap = 16L)),
      basename(receipt$original_job) == paste0(mode, "_0_job.rds"))
    paths <- capture_input_paths(receipt)
    hashes <- vapply(paths, function(path) {
      value <- receipt$input_sha256[[path]]
      stopifnot(is.character(value), length(value) == 1L, !is.na(value),
        grepl("^[0-9a-f]{64}$", value))
      value
    }, "")
    stopifnot(hashes[["driver"]] ==
      "b53ba0692c077953884e3c435d4be42094e1f682a4ea3f250e8366485bc04775")
    hashes
  })
  stopifnot(identical(identities[[1L]], identities[[2L]]))
  invisible(TRUE)
}

extract_audit_bundle <- function(bundle, expected_hash, count, directory) {
  sha256 <- function(path) digest::digest(file = path, algo = "sha256")
  stopifnot(sha256(bundle) == expected_hash)
  connection <- gzfile(bundle, "rt")
  records <- lapply(readLines(connection), jsonlite::fromJSON)
  close(connection)
  paths <- vapply(records, function(x) paste(x$origin, x$name, sep = "/"), "")
  stopifnot(length(records) == count, !anyDuplicated(paths),
    all(grepl("^[A-Za-z0-9_.-]+(/[A-Za-z0-9_.-]+)*$", paths)),
    !any(unlist(strsplit(paths, "/", fixed = TRUE)) %in% c(".", "..")))
  for (i in seq_along(records)) {
    bytes <- jsonlite::base64_dec(records[[i]]$bytes_base64)
    stopifnot(digest::digest(bytes, algo = "sha256", serialize = FALSE) == records[[i]]$sha256)
    path <- file.path(directory, paths[i])
    if (!dir.exists(dirname(path))) stopifnot(dir.create(dirname(path), recursive = TRUE))
    writeBin(bytes, path)
  }
  invisible(TRUE)
}

audit_full_outputs <- function() {
  sha256 <- function(path) digest::digest(file = path, algo = "sha256")
  directory <- tempfile("duckhts-singleton-audit-")
  stopifnot(dir.create(directory))
  on.exit(unlink(directory, recursive = TRUE), add = TRUE)
  extract_audit_bundle(file.path(root, "benchmarks/data/duckvep_haplotypes_indel_full_audit.jsonl.gz"),
    "47c307544fc12d5c3430a7b9359eff68cf299e371e710111a04671807fc8c8c7", 82L, directory)
  inputs <- file.path(directory, "inputs")
  extract_audit_bundle(file.path(root, "benchmarks/data/duckvep_haplotypes_indel_inputs.jsonl.gz"),
    "c5293f3ba2072a82290075e9476c79b1a28810b554f19b4ea3273d25b7ca95f1", 46L, inputs)
  comparator <- file.path(directory, "audit/audit_pairs.R")
  invoke <- function(label, before, after, transcripts, expected_error = NULL) {
    result <- file.path(directory, paste0(label, ".json"))
    log <- file.path(directory, paste0(label, ".log"))
    status <- system2(file.path(R.home("bin"), "Rscript"),
      shQuote(c(comparator, before, after, transcripts, result)), stdout = log, stderr = log)
    if (!is.null(expected_error)) {
      stopifnot(status != 0L, !file.exists(result),
        any(grepl(expected_error, readLines(log), fixed = TRUE)))
      return(invisible(NULL))
    }
    if (status != 0L) stop(paste(readLines(log), collapse = "\n"))
    jsonlite::fromJSON(result)
  }
  comparisons <- list()
  for (transcripts in c(1024L, 10240L)) {
    for (mode in c("sql_singletons", "sql_singletons_hgvs")) {
      label <- paste(transcripts, mode, sep = "_")
      captures <- setNames(file.path(directory, paste(c("baseline", "current"), label,
        sep = "_")), c("baseline", "current"))
      receipts <- lapply(captures, function(capture)
        jsonlite::fromJSON(file.path(capture, "receipt.json")))
      jobs <- list()
      for (side in names(captures)) {
        capture <- captures[[side]]
        receipt <- receipts[[side]]
        metrics <- read.csv(file.path(capture, "metrics.csv"), colClasses = "character")
        ledger <- if (side == "baseline") baseline else current
        ledger <- ledger[ledger$transcripts == transcripts & ledger$samples == "64" &
          ledger$overlap == "16" & ledger$mode == mode, ]
        stopifnot(receipt$status == "complete", nrow(ledger) == 3L, nrow(metrics) == 1L,
          length(receipt$verified_non_timing_fields) == 27L,
          identical(names(metrics), receipt$verified_non_timing_fields),
          identical(receipt$verified_timed_passes, 1:3), receipt$parquet_roundtrip_equal,
          all(ledger$source_revision == receipt$benchmark_source_revision),
          all(ledger$receipt_sha256 == receipt$benchmark_receipt_sha256),
          all(ledger$extension_sha256 == receipt$input_sha256[[receipt$loaded_extension]]))
        for (name in names(receipt$output_sha256)) {
          stopifnot(sha256(file.path(capture, name)) == receipt$output_sha256[[name]])
        }
        for (field in names(metrics)) stopifnot(all(ledger[[field]] == metrics[[field]]))
        jobs[[side]] <- capture_original_jobs(receipt, ledger, inputs)
      }
      stopifnot(identical(jobs$baseline, jobs$current))
      compare_capture_inputs(receipts$baseline, receipts$current, mode, transcripts)
      rejected_input <- function(candidate) tryCatch({
        compare_capture_inputs(receipts$baseline, candidate, mode, transcripts)
        FALSE
      }, error = function(e) TRUE)
      for (path in capture_input_paths(receipts$current)) {
        changed <- receipts$current
        hash <- changed$input_sha256[[path]]
        changed$input_sha256[[path]] <- paste0(if (startsWith(hash, "0")) "1" else "0",
          substring(hash, 2L))
        stopifnot(rejected_input(changed))
        changed$input_sha256[[path]] <- NULL
        stopifnot(rejected_input(changed))
      }
      changed <- receipts$current
      changed$configuration$samples <- 65L
      stopifnot(rejected_input(changed))
      observed <- invoke(label, file.path(captures[["baseline"]], "rows.parquet"),
        file.path(captures[["current"]], "rows.parquet"), transcripts)
      retained <- jsonlite::fromJSON(file.path(directory, "audit", paste0("full_", label, ".json")))
      stopifnot(identical(observed$comparisons, retained$comparisons))
      comparisons[[label]] <- data.frame(transcripts, mode, observed$comparisons)
    }
  }
  controls <- read.csv(file.path(directory, "controls/summary.csv"), colClasses = "character")
  stopifnot(nrow(controls) == 8L, all(controls$passed == "TRUE"),
    sum(controls$expected_outcome == "accept") == 1L,
    sum(controls$expected_outcome == "reject") == 7L)
  for (i in which(controls$expected_outcome == "reject")) {
    control <- controls$control[i]
    invoke(control, file.path(directory, "controls", control, "baseline.parquet"),
      file.path(directory, "controls", control, "current.parquet"), 1024L,
      controls$expected_assertion[i])
  }
  do.call(rbind, comparisons)
}
full_audit <- audit_full_outputs()
```

Source: `r revision`. Nearest same-input baseline: `r baseline_revision`.
All 51 timed observations are retained in
[the checkpoint data](data/duckvep_haplotypes_indel.csv), including every full-output
fingerprint. **The cross-revision full-output equality check failed.** None of
these rows was appended to the existing benchmark history, and its equality
guard remains unchanged. Rendering this diagnostic report is not a passing
benchmark-promotion verdict.

## Workload and execution

[The benchmark driver](duckvep_haplotypes.R) uses the same registered 180-base
canonical-DNA CDS, four original ALT events and diploid sample templates as
[the baseline report](duckvep_haplotypes.md). It checks complete carrier multisets,
sequences, edit provenance, corrupted inputs and full output fingerprints within
each mode. The singleton HGVS check compares independent and phased DuckHTS
paths; it is not an independent VEP oracle.

Both campaigns ran on the Intel i5-13500 with DuckDB 1.5.3, one thread pinned to
CPU 2, three fresh-process timed passes and separate correctness workers.
The current extension was rebuilt from a clean tracked checkout with HTSlib
distclean; its SHA-256 is
`a5dceecf83f96b484339fec50372953465dc0543199508bb368df34a9c647f13`.
Retained receipts identify source files, native bridge, extension, registered
inputs, derived FASTA/index and every worker artifact. They are local
source-bound records, not signed CI attestations.

Native timing covers initialized ordered-feed replay and a count sink. SQL timing
covers full table materialization, including preparation, sort, replay, sequence
differences, local coding facts and all nested output. Model/input staging and
post-query fingerprints are outside timing. Whole-process maximum RSS includes
the R/DuckDB host, setup, warm-up, output and post-query checks.

## All measured configurations

```{r timing}
knitr::kable(summary, row.names = FALSE, digits = 4)
```

`NA` marks the native count sink, which has no serialized SQL-row fingerprint.
Counts, input identities, output contracts, sequence bytes and JSON byte lengths
match between revisions. The unequal singleton fingerprints mean those timings
are **same-input, different-output measurements**, not identical-output work.
Matching byte lengths do not establish matching values.

```{r denominators}
denominators <- c(keys[-5L], "input_records", "projected_events", "input_calls",
  "input_physical_records", "input_record_sample_calls", "input_candidate_sample_rows",
  "output_leaves", "output_carriers", "cds_bytes", "protein_bytes", "json_bytes")
knitr::kable(current[current$pass == "1", denominators], row.names = FALSE)
```

## Retained output differences

All three measured passes agree within each mode. Across revisions,
`r nrow(unique(differences[keys]))` of 51 timed rows have unequal fingerprints:
the two singleton modes at both transcript counts, with three passes each.
Literal replay fingerprints agree. The complete observations below retain the
unequal hash values; no field was normalized away to obtain a match.

```{r unequal-fingerprints}
knitr::kable(differences[differences$pass == "1", ], row.names = FALSE)
```

A [retained four-event, one-transcript reproduction](data/duckvep_haplotypes_indel_witness.jsonl.gz) isolates the deletion
`GC>G` at CDS position 67: its block `local_consequence_mask` changes from
8,388,608 (`frameshift`) to 8,396,800 (`frameshift` plus raw `missense`). Every
other field in those four complete rows is unchanged. The physical deletion
changes local `GCT` to partial `GT`, giving peptide operands `A` and `X`.
The documented local-block contract evaluates raw coding predicates without an
independent source event's uploaded-feature class gates. See
[ERRATA.md](../ERRATA.md) for the pinned predicate evidence and its limits.
The compact witness preserves both complete native outputs and all original
files from the exact pinned-VEP observation, including source, observation
scripts, commands, receipts and failed-attempt diagnostics. Its 38 files are
stored as byte-exact base64 payloads with per-file hashes. Rendering verifies
the bundle identity, the complete four-row difference and the independent
raw-predicate observation. It is unsigned diagnostic evidence.

The [complete-output audit](data/duckvep_haplotypes_indel_full_audit.jsonl.gz)
retains all eight baseline/current Parquet relations for the two singleton modes
at both transcript counts. Each untimed capture matches all 27 non-timing result
fields from its original correctness worker and all three timed workers, including
every full-output fingerprint; Parquet round trips preserve those metrics and types.
Rendering verifies the pinned bundle and receipt identities, binds the retained
metrics to both ledgers, and checks the exact workload configuration. The
[original-input capsule](data/duckvep_haplotypes_indel_inputs.jsonl.gz) preserves
32 original worker jobs, four benchmark receipts, their four FASTA/index pairs,
and both registered fixtures. Each original receipt is bound to its capture and
ledger; every job's actual bytes must match that receipt, and the correctness
job must also match its capture's input hash. Decoded jobs must contain the exact
registered reference and event table. Their complete semantic fields agree
across revisions and passes, with each pass's correctness flag checked separately;
only named binary, output and reference-file paths are excluded from that equality.
Extension identities are checked against their own revision, not equated across
revisions. The driver remains independently pinned, and retained FASTA/index bytes
must match both receipt layers. Original paths are identifiers only: rendering
does not reopen historical files. Input controls reject missing, malformed or
altered job hashes and self-consistently rehashed jobs with changed events or
configuration. Rendering then reruns the complete typed, exact-key comparison.

```{r complete-output-audit}
knitr::kable(full_audit, row.names = FALSE)
```

Across `r sum(full_audit$joined_rows)` keyed row pairs,
`r sum(full_audit$changed_rows)` differ only in the declared deletion mask above.
There are no missing keys, changes to other events or unexplained field changes.
The audit checks the full expected transcript/event inventory and deletion geometry,
not only aggregate hashes. Its seven corruption controls reject NULL blocks,
count-preserving key shifts, altered integer types, changed protein sequence,
omitted or duplicate rows and an incorrect deletion mask; rendering reruns them.
The bundle retains the observation/comparison scripts, receipts, complete corrupted
inputs and original logs. These remain unsigned local observations. The outputs
are still unequal: explaining their difference does not turn these measurements
into identical-output work or change the historical promotion guard.

Three passes on one shared machine do not establish statistical significance,
a causal performance effect or a general absence of regression. This fixture
does not measure N-rich annotation, whole-haplotype consequences or compound HGVS.

## Reproduction

Use the clean extension-receipt procedure in the baseline report, then run
`benchmarks/duckvep_haplotypes.R` with `--passes 3 --cpu 2` and these six workloads:

```{r workloads}
workloads <- unique(current[c("transcripts", "samples", "overlap")])
workloads$modes <- vapply(seq_len(nrow(workloads)), function(i) {
  selected <- current$transcripts == workloads$transcripts[i] &
    current$samples == workloads$samples[i] & current$overlap == workloads$overlap[i]
  paste(unique(current$mode[selected]), collapse = ",")
}, "")
knitr::kable(workloads, row.names = FALSE)
```

Pass `--extension-receipt PATH` and the listed `--transcripts`, `--samples`,
`--overlap` and `--modes` to each invocation. Preserve every result and receipt;
do not append a campaign that fails the historical full-output check.

## Canonical workload at the translation checkpoints

```{r translation-checkpoint, include=FALSE}
translation <- read.csv(file.path(root, "benchmarks/data/duckvep_haplotypes_indel_translation.csv"),
  colClasses = "character")
translation_previous <- current[current$transcripts == "1024" & current$samples == "64" &
  current$overlap == "16", ]
translation_modes <- c("native", "sql", "sql_records", "sql_singletons", "sql_singletons_hgvs")
translation_revisions <- unique(translation$source_revision)
stopifnot(identical(names(translation), names(current)),
  nrow(translation) == 15L * length(translation_revisions),
  nrow(translation_previous) == 15L, !anyDuplicated(translation[c("source_revision", keys)]),
  identical(translation_revisions, c("52ce78513b3f5f7c2a46fe82dc44dd458333092b",
    "de3d008004f615c198a422a42a773bd7c9de0b70", "d6d188da54f0a0cf97ec74244243ebad251194f5")),
  all(translation$extension_sha256[translation$source_revision == translation_revisions[1L]] ==
    "0217585b75ac7277794590b7f714b85eb942667a1705ea87a31021faa7833ab5"),
  all(translation$extension_sha256[translation$source_revision == translation_revisions[2L]] ==
    "cb34f65f738f5d26ce5b849bbcf9628e95db180e62a199a153aacbdac44f25b0"),
  all(translation$extension_sha256[translation$source_revision == translation_revisions[3L]] ==
    "2563543e63566c24872a9f8c5a6db4a12871bd8b58e9f99e9f5a87672b3dd565"))

check_translation_job <- function(job, mode, pass, original_directory, receipt, fixture) {
  stopifnot(identical(names(job), c("transcripts", "samples", "overlap", "mode", "verify",
      "input", "extension", "shared", "reference_fasta", "reference_length", "result")),
    identical(job$transcripts, 1024L), identical(job$samples, 64L), identical(job$overlap, 16L),
    identical(job$mode, mode), identical(job$verify, pass == 0L),
    identical(job$input$reference, fixture$reference), identical(job$input$events, fixture$events),
    identical(job$reference_length, 64099L),
    identical(job$reference_fasta, file.path(original_directory, "reference.fa")),
    identical(job$shared, file.path(original_directory, "stream.so")),
    identical(job$result, file.path(original_directory, paste0(mode, "_", pass, ".rds"))))
  recorded_hash(receipt$sha256, job$extension)
  job[setdiff(names(job), c("verify", "extension", "shared", "reference_fasta", "result"))]
}

audit_translation_checkpoint <- function() {
  sha256 <- function(path) digest::digest(file = path, algo = "sha256")
  directory <- tempfile("duckhts-translation-benchmark-")
  stopifnot(dir.create(directory))
  on.exit(unlink(directory, recursive = TRUE), add = TRUE)
  extract_audit_bundle(file.path(root, "benchmarks/data/duckvep_haplotypes_indel_translation.jsonl.gz"),
    "2f189dcf4e6d0e1ccb6d3e2fb606610ccd002dfb966ea47480ead23319c8819e", 351L, directory)
  fixture_paths <- file.path(directory, "fixtures", c("reference.fa", "events.tsv"))
  fasta <- readLines(fixture_paths[1L])
  stopifnot(length(fasta) == 2L, nchar(fasta[2L]) == 180L)
  fixture <- list(reference = fasta[2L], events = read.delim(fixture_paths[2L]))
  driver_hash <- sha256(file.path(directory, "source/duckvep_haplotypes.R"))
  stopifnot(driver_hash == "b53ba0692c077953884e3c435d4be42094e1f682a4ea3f250e8366485bc04775")
  jobs <- receipts <- list()
  origins <- list.dirs(directory, recursive = FALSE, full.names = FALSE)
  origins <- origins[file.exists(file.path(directory, origins, "receipt.json"))]
  expected_revisions <- c(unique(translation_previous$source_revision), translation_revisions)
  stopifnot(!anyDuplicated(expected_revisions), length(origins) == length(expected_revisions))
  for (side in origins) {
    path <- file.path(directory, side)
    receipt <- jsonlite::fromJSON(file.path(path, "receipt.json"))
    source_revision <- receipt$source_revision
    stopifnot(is.character(source_revision), length(source_revision) == 1L,
      source_revision %in% expected_revisions, !source_revision %in% names(receipts))
    ledger <- if (source_revision == expected_revisions[1L]) translation_previous else
      translation[translation$source_revision == source_revision, ]
    stopifnot(nrow(ledger) == 15L)
    receipts[[source_revision]] <- receipt
    stopifnot(all(ledger$receipt_sha256 == sha256(file.path(path, "receipt.json"))),
      all(ledger$source_revision == receipt$source_revision),
      all(ledger$extension_build_binding == "htslib_distclean_make_release"),
      identical(receipt$extension_build_binding, "htslib_distclean_make_release"),
      identical(receipt$options[c("transcripts", "samples", "overlap", "passes", "cpu", "modes",
        "diagnostic")], list(transcripts = 1024L, samples = 64L, overlap = 16L, passes = 3L,
          cpu = 2L, modes = paste(translation_modes, collapse = ","), diagnostic = FALSE)))
    result_path <- names(receipt$sha256)[endsWith(names(receipt$sha256), "/results.csv")]
    stopifnot(length(result_path) == 1L)
    original_directory <- dirname(result_path)
    stopifnot(all(endsWith(original_directory, paste0("/", ledger$artifact_directory))))
    original_files <- names(receipt$sha256)[startsWith(names(receipt$sha256),
      paste0(original_directory, "/"))]
    stopifnot(!anyDuplicated(basename(original_files)),
      setequal(list.files(path), c("receipt.json", setdiff(basename(original_files), "stream.so"))))
    for (name in setdiff(list.files(path), "receipt.json")) {
      stopifnot(sha256(file.path(path, name)) ==
        recorded_hash(receipt$sha256, file.path(original_directory, name)))
    }
    for (i in seq_along(fixture_paths)) {
      original <- names(receipt$sha256)[endsWith(names(receipt$sha256),
        paste0("/duckvep-haplotypes/", basename(fixture_paths[i])))]
      field <- c("reference_sha256", "events_sha256")[i]
      stopifnot(length(original) == 1L,
        sha256(fixture_paths[i]) == recorded_hash(receipt$sha256, original),
        all(ledger[[field]] == sha256(fixture_paths[i])))
    }
    original_driver <- names(receipt$sha256)[endsWith(names(receipt$sha256),
      "benchmarks/duckvep_haplotypes.R")]
    stopifnot(length(original_driver) == 1L,
      recorded_hash(receipt$sha256, original_driver) == driver_hash,
      identical(readLines(file.path(path, "reference.fa")),
        c(">bench", paste0(strrep("A", 99L), strrep(paste0(fixture$reference, strrep("A", 820L)), 64L)))))
    build <- read.delim(file.path(directory, "build", paste0(side, ".tsv")), colClasses = "character")
    stopifnot(identical(names(build), c("field", "value")),
      identical(build$field, c("source_revision", "path", "binding", "sha256")))
    build <- setNames(build$value, build$field)
    stopifnot(build[["source_revision"]] == receipt$source_revision,
      build[["binding"]] == receipt$extension_build_binding,
      build[["sha256"]] == recorded_hash(receipt$sha256, build[["path"]]),
      all(ledger$extension_sha256 == build[["sha256"]]),
      all(ledger$native_binary_sha256 == recorded_hash(receipt$sha256,
        file.path(original_directory, "stream.so"))))
    measured <- read.csv(file.path(path, "results.csv"), colClasses = "character")
    stopifnot(nrow(measured) == 15L, !anyDuplicated(measured[keys]))
    measured <- measured[do.call(order, measured[keys]), ]
    ledger <- ledger[do.call(order, ledger[keys]), ]
    rownames(measured) <- rownames(ledger) <- NULL
    stopifnot(identical(measured, ledger[names(measured)]))
    jobs[[source_revision]] <- list()
    for (mode in translation_modes) {
      first <- NULL
      for (pass in 0:3) {
        stem <- paste0(mode, "_", pass)
        job <- readRDS(file.path(path, paste0(stem, "_job.rds")))
        semantic <- check_translation_job(job, mode, pass, original_directory, receipt, fixture)
        stopifnot(identical(job$extension, build[["path"]]))
        jobs[[source_revision]][[stem]] <- semantic
        value <- readRDS(file.path(path, paste0(stem, ".rds")))
        non_timing <- value[setdiff(names(value), "seconds")]
        if (pass == 0L) first <- non_timing else stopifnot(identical(non_timing, first))
        if (pass != 0L) {
          row <- ledger[ledger$mode == mode & ledger$pass == pass, ]
          stopifnot(nrow(row) == 1L)
          for (field in names(value)) stopifnot(identical(as.character(value[[field]]), row[[field]]))
          timing <- readLines(file.path(path, paste0(stem, "_time.txt")))
          rss <- sub(".*: ", "", timing[grepl("Maximum resident set size", timing, fixed = TRUE)])
          stopifnot(identical(rss, row$peak_process_rss_kib))
        }
        if (pass == 0L) for (field in c("samples", "events")) {
          changed <- job
          if (field == "samples") changed$samples <- 65L else changed$input$events$alternate[1L] <- "T"
          rejected <- tryCatch({
            check_translation_job(changed, mode, pass, original_directory, receipt, fixture)
            FALSE
          }, error = function(e) TRUE)
          stopifnot(rejected)
        }
      }
    }
  }
  stopifnot(setequal(names(receipts), expected_revisions))
  baseline_revision <- expected_revisions[1L]
  for (source_revision in translation_revisions) {
    stopifnot(identical(jobs[[baseline_revision]], jobs[[source_revision]]),
      identical(receipts[[baseline_revision]]$compiler, receipts[[source_revision]]$compiler),
      identical(receipts[[baseline_revision]]$compiler_flags, receipts[[source_revision]]$compiler_flags))
  }
  invisible(TRUE)
}
audit_translation_checkpoint()
translation_previous <- translation_previous[do.call(order, translation_previous[keys]), ]
rownames(translation_previous) <- NULL
for (source_revision in translation_revisions) {
  observed <- translation[translation$source_revision == source_revision, ]
  observed <- observed[do.call(order, observed[keys]), ]
  rownames(observed) <- NULL
  stopifnot(identical(observed[setdiff(names(observed), measurement_fields)],
    translation_previous[setdiff(names(translation_previous), measurement_fields)]))
}
stopifnot(all(translation$threads == "1"), all(translation$cpu_affinity == "2"),
  all(translation$duckdb_version[translation$mode != "native"] == "1.5.3"),
  all(is.finite(as.numeric(translation$seconds))), all(as.numeric(translation$seconds) > 0))
translation_summary <- do.call(rbind, lapply(translation_revisions, function(source_revision) {
  revision_index <- match(source_revision, translation_revisions)
  previous <- if (revision_index == 1L) translation_previous else
    translation[translation$source_revision == translation_revisions[revision_index - 1L], ]
  do.call(rbind, lapply(translation_modes, function(mode) {
    after <- translation[translation$source_revision == source_revision & translation$mode == mode, ]
    before <- previous[previous$mode == mode, ]
    stopifnot(identical(sort(after$pass), c("1", "2", "3")))
    data.frame(revision = substr(source_revision, 1L, 7L),
      baseline_revision = substr(unique(before$source_revision), 1L, 7L), mode,
      input_candidate_sample_rows = 262144L, output_leaves = after$output_leaves[1L],
      output_carriers = after$output_carriers[1L],
      baseline_median_s = median(as.numeric(before$seconds)),
      checkpoint_median_s = median(as.numeric(after$seconds)),
      checkpoint_max_rss_mib = max(as.numeric(after$peak_process_rss_kib)) / 1024)
  }))
}))
```

Sources `52ce78513b3f5f7c2a46fe82dc44dd458333092b`,
`de3d008004f615c198a422a42a773bd7c9de0b70` and
`d6d188da54f0a0cf97ec74244243ebad251194f5` each use the nearest recorded identical
workload as their timing baseline: `20efcf2af33b38c5be7596e2db7b243a3be53c47`
for `52ce785`, `52ce785` for `de3d008`, and `de3d008` for `d6d188d`.
The workload has 1,024 transcripts, 64 samples,
16 overlapping transcripts, 256 physical events and 262,144 candidate/sample input rows.
All four runs use one thread on CPU 2, DuckDB 1.5.3 and three fresh-process timed passes.
No DuckHTS test or conformance jobs ran concurrently with the `52ce785` timers.
No DuckHTS build, test, conformance or diagnostic jobs overlapped the `d6d188d` timers.
The `de3d008` campaign ran from 00:59:19 to 00:59:59 +0200 on September 11, 2026;
a previously launched small read-only classification/oracle diagnostic completed
at 01:00:01, so concurrent diagnostic activity cannot be excluded. This measured
run is retained without replacement. Other shared-host activity was not controlled.

```{r translation-timing}
knitr::kable(translation_summary, digits = 6, row.names = FALSE)
```

Each checkpoint's 15 matched rows have identical non-timing result fields, including every SQL
full-output fingerprint, output denominator and native workspace/count metric.
The [45 checkpoint observations](data/duckvep_haplotypes_indel_translation.csv)
retain individual passes. The [351-file evidence capsule](data/duckvep_haplotypes_indel_translation.jsonl.gz)
retains all four runs' original worker jobs, results, logs, process-time reports,
benchmark receipts and FASTA/index pairs, all four clean-build receipts, the registered
fixtures and the unchanged driver. Rendering verifies every retained file against
its receipt before comparing actual jobs and results. It requires complete job
semantics to agree across revisions, excluding only the explicitly checked
correctness flag and named binary/output/reference-file paths. All 80 result
objects are checked for within-mode repeatability; all timed values and RSS
measurements are bound to the corresponding ledger rows. Forty job-mutation
controls reject changed sample counts or event alleles. Original paths remain
identifiers only; rendering does not reopen historical files.

Compiled extension and native-bridge payloads are not included in this capsule;
their identities are retained and checked through the original receipts. The
`d6d188d` extension SHA-256 is
`2563543e63566c24872a9f8c5a6db4a12871bd8b58e9f99e9f5a87672b3dd565`.
These are unsigned local source-bound measurements. The historical unequal-output
comparison and its complete-output controls remain separate and unchanged.
The checkpoint comparisons establish fingerprint equality, not complete typed-row equality.

This canonical table-1 workload measures shared replay and singleton-HGVS paths.
It does not measure the newly accepted N-bearing REF or nonstandard-codon-table
branches. Three passes on a shared host do not establish statistical significance,
causality or a general absence of regression. Reproduce with the clean-build
receipt procedure above and this command:

```{bash translation-reproduction, eval=FALSE, echo=TRUE}
Rscript benchmarks/duckvep_haplotypes.R \
  --transcripts 1024 --samples 64 --overlap 16 --passes 3 --cpu 2 \
  --modes native,sql,sql_records,sql_singletons,sql_singletons_hgvs \
  --extension-receipt /tmp/duckhts-indel-d6d188d-evidence.0aBmKK/extension.tsv
```
