---
title: "Draft upstream GFFBase strict-mode issue"
output:
  github_document:
    html_preview: false
---

<!-- gffbase_strict_issue_draft.md is generated from gffbase_strict_issue_draft.Rmd. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
repo_root <- normalizePath(Sys.getenv("DUCKHTS_REPO_ROOT", unset = getwd()), mustWork = FALSE)
if (basename(repo_root) == "benchmarks") repo_root <- dirname(repo_root)
bench_csv <- file.path(repo_root, ".tmp", "gffbase_conformance", "performance.csv")
meta_json <- file.path(repo_root, ".tmp", "gffbase_conformance", "metadata.json")

safe_cmd <- function(cmd, args = character(), env = character()) {
  out <- tryCatch(system2(cmd, args, stdout = TRUE, stderr = TRUE, env = env), error = function(e) NA_character_)
  status <- attr(out, "status")
  if (length(out) == 0 || anyNA(out) || (!is.null(status) && status != 0)) return("N/A")
  paste(out, collapse = " ")
}

gffbase_site <- file.path(repo_root, ".tmp", "gffbase_site")
python_env <- character()
if (dir.exists(gffbase_site)) python_env <- sprintf("PYTHONPATH=%s", gffbase_site)
py <- function(code) safe_cmd("python3", c("-c", shQuote(code)), env = python_env)

os_name <- if (file.exists("/etc/os-release")) {
  os_release <- readLines("/etc/os-release", warn = FALSE)
  sub('^PRETTY_NAME="?([^"\n]+)"?$', "\\1", os_release[grepl("^PRETTY_NAME=", os_release)][1])
} else Sys.info()[["sysname"]]

meta <- NULL
if (file.exists(meta_json) && requireNamespace("jsonlite", quietly = TRUE)) {
  meta <- tryCatch(jsonlite::fromJSON(meta_json), error = function(e) NULL)
}

fallback_perf <- data.frame(
  dataset = c(
    rep("synthetic_200000.gff3", 7),
    rep("gencode.v49.basic.annotation.gff3.gz", 7)
  ),
  tool = rep(c(rep("DuckHTS", 6), "GFFBase"), 2),
  variant = rep(c(
    "read_gff COUNT(*)",
    "read_gff strict COUNT(*)",
    "read_gff projected filter/sum",
    "read_gff attributes_map",
    "read_gff attributes_list",
    "read_gff attributes_pairs",
    "parse_gff strict (rust)"
  ), 2),
  rows = c(rep(200000, 7), rep(5866158, 7)),
  passes = 3,
  median_sec = c(0.0109, 0.0672, 0.0357, 0.1050, 0.1790, 0.1246, 0.5540,
                 1.9906, 5.1975, 2.9470, 10.1395, 17.2256, 13.7290, 40.8327),
  vs_gffbase_parse = c(50.69, 8.25, 15.53, 5.28, 3.09, 4.45, 1.00,
                       20.51, 7.86, 13.86, 4.03, 2.37, 2.97, 1.00),
  stringsAsFactors = FALSE
)
perf <- if (file.exists(bench_csv)) utils::read.csv(bench_csv, check.names = FALSE) else fallback_perf
if (!"vs_gffbase_parse" %in% names(perf)) {
  perf$vs_gffbase_parse <- NA_real_
  for (ds in unique(perf$dataset)) {
    idx <- perf$dataset == ds
    base <- perf$median_sec[idx & perf$tool == "GFFBase" & grepl("parse_gff", perf$variant)][1]
    perf$vs_gffbase_parse[idx] <- round(base / perf$median_sec[idx], 2)
  }
}
show_variants <- c("read_gff strict COUNT(*)", "read_gff attributes_list", "read_gff attributes_pairs", "read_gff attributes_map+list+pairs", "parse_gff strict (rust)")
perf_issue <- perf[perf$variant %in% show_variants, c("dataset", "tool", "variant", "rows", "passes", "median_sec", "vs_gffbase_parse")]
perf_issue$rows <- format(perf_issue$rows, big.mark = ",", scientific = FALSE)
perf_issue$median_sec <- round(perf_issue$median_sec, 4)
perf_issue$vs_gffbase_parse <- round(perf_issue$vs_gffbase_parse, 2)
```

# Copy-ready issue body

## Summary

`parse_gff(..., strict=True)` accepts several GFF3 rows that look structurally malformed: rows with a 10th tab-separated field, rows with a partially malformed attribute segment such as `ID=ok;broken`, and rows with an unknown start but concrete `end=0`.

## Environment

| | |
|---|---|
| **OS** | `r os_name` |
| **Python version** | `r py("import sys; print(sys.version.split()[0])")` |
| **GFFBase version** | `r py("import gffbase; print(gffbase.__version__)")` |
| **Install method** | `pip install --target .tmp/gffbase_site gffbase==0.1.0` for this reproduction; also cross-checked against the local source mirror at commit `78714cf30a9d799eab544e00a79a4da9754987ca`. |
| **DuckDB version** | `r py("import duckdb; print(duckdb.__version__)")` |
| **PyArrow version** | `r py("import pyarrow; print(pyarrow.__version__)")` |
| **Native extension built?** | `r py("from gffbase import native_available; print(native_available())")` |

## Minimal reproducible example

```python
import tempfile
from pathlib import Path
import gffbase

cases = {
    "extra_field": b"chr1\tsrc\texon\t1\t10\t.\t+\t.\tID=ok\textra\n",
    "partial_malformed_attribute": b"chr1\tsrc\texon\t1\t10\t.\t+\t.\tID=ok;broken\n",
    "nonpositive_end_with_unknown_start": b"chr1\tsrc\tregion\t.\t0\t.\t.\t.\tID=x\n",
}

for name, text in cases.items():
    path = Path(tempfile.mkdtemp()) / f"{name}.gff3"
    path.write_bytes(text)
    try:
        rows = list(gffbase.parse_gff(str(path), strict=True, engine="rust"))
        first = rows[0] if rows else None
        print(name, "ACCEPTED", "rows=", len(rows))
        if first is not None:
            print("  start/end:", first.start, first.end)
            print("  attributes_pairs:", first.attributes_pairs)
            print("  extra:", first.extra)
    except Exception as exc:
        print(name, "REJECTED", type(exc).__name__, str(exc))
```

Minimal GFF3 snippets:

```text
##gff-version 3
chr1    src    exon      1    10    .    +    .    ID=ok    extra
chr1    src    exon      1    10    .    +    .    ID=ok;broken
chr1    src    region    .    0     .    .    .    ID=x
```

## What you expected

I expected `strict=True` either to reject these rows or for the documentation to explicitly describe this as intentionally permissive / gffutils-compatible parsing.

My reading is:

- GFF3 feature rows are structurally nine tab-separated columns; a 10th field is not part of the core GFF3 row shape.
- Non-empty GFF3 attribute segments are `tag=value` pairs separated by semicolons; `ID=ok;broken` has one valid pair plus one malformed segment.
- Concrete coordinates are 1-based positive integers. If a parser allows unknown `start='.'`, a concrete `end=0` still seems nonpositive.
- The current `parse_gff()` docstring says `strict=True` raises `GFFFormatError` on the first malformed line.

## What actually happened

On `gffbase==0.1.0` with the native Rust parser available, all three rows were accepted:

```text
extra_field ACCEPTED rows= 1
  start/end: 1 10
  attributes_pairs: [('ID', 'ok', 0)]
  extra: ['extra']
partial_malformed_attribute ACCEPTED rows= 1
  start/end: 1 10
  attributes_pairs: [('ID', 'ok', 0), ('broken', '', 0)]
  extra: []
nonpositive_end_with_unknown_start ACCEPTED rows= 1
  start/end: None 0
  attributes_pairs: [('ID', 'x', 0)]
  extra: []
```

## Additional context

We found this while implementing strict GFF3 validation in DuckHTS. GFFBase's parser and NCBI/GFF3-oriented tests were useful inspiration for DuckHTS' own strict mode, so this report is meant as cross-validation feedback rather than a criticism. DuckHTS now intentionally rejects the three cases above under `read_gff(..., strict := true)`.

For context, we also ran direct parser/scan benchmarks against GFFBase. These numbers compare DuckHTS direct DuckDB scans with `gffbase.parse_gff(strict=True, engine="rust")`; they do **not** compare GFFBase's heavier `create_db()` annotation-database ingest path.

```{r benchmark-table, echo=FALSE}
knitr::kable(perf_issue, row.names = FALSE)
```

Benchmark machine:

```{r server-specs, echo=FALSE}
if (!is.null(meta$server_specs)) {
  ss <- meta$server_specs
  knitr::kable(data.frame(
    parameter = c("OS", "kernel / machine", "CPU", "logical / affinity CPUs", "memory", "DuckDB threads"),
    value = c(
      if (!is.null(ss$os_pretty_name)) ss$os_pretty_name else ss$platform,
      paste(ss$release, ss$machine, sep = " / "),
      ss$cpu_model,
      paste(ss$cpu_logical, ss$cpu_affinity, sep = " / "),
      paste0(ss$mem_total_gib, " GiB"),
      meta$duckdb_threads
    )
  ), row.names = FALSE)
} else {
  knitr::kable(data.frame(
    parameter = c("OS", "CPU", "memory", "DuckDB threads"),
    value = c("Ubuntu 24.04.3 LTS", "13th Gen Intel(R) Core(TM) i5-13500", "62.58 GiB", "4")
  ), row.names = FALSE)
}
```

One performance observation: GFFBase's Arrow staging path is a good Python-to-DuckDB ingestion strategy, especially for `create_db()`. For maximum DuckDB throughput, however, a future Rust-native path could avoid the `Rust parser -> Python objects/lists -> PyArrow Table -> DuckDB INSERT` boundary. If `duckdb-rs` exposes the needed appender/table-function/vector APIs for the supported DuckDB versions, writing parsed records directly into DuckDB vectors from Rust may be a simpler high-performance path. If not, Arrow remains the pragmatic Python boundary.
